### 2017-03-22 random_projection/draft/1dproj.tex
random projections

\documentclass[aps,preprint,nofootinbib,floatfix]{revtex4-1}

%\usepackage{cite}
\usepackage[pdftex]{graphicx}
\graphicspath{{./figures/}}
%\graphicspath{{./}}
\usepackage{amsmath,amssymb,amsfonts,amsthm,amscd,bm}
\usepackage{algorithm}
\usepackage{algorithmic}
%\usepackage{epstopdf}
%\usepackage{fixltx2e}
%\usepackage{stfloats}

\hyphenation{op-tical net-works semi-conduc-tor}

\newtheorem{theorem}{Theorem}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{assumption}[theorem]{Assumption}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{example}{Example}

%% our definitions %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\DeclareMathOperator{\aff}{aff}
\DeclareMathOperator{\st}{s.t.}
\DeclareMathOperator{\LC}{LC}
\DeclareMathOperator{\affnot}{aff_0}
\DeclareMathOperator{\conv}{conv}
\DeclareMathOperator{\relint}{relint}
\DeclareMathOperator{\vol}{vol}
\DeclareMathOperator{\range}{range}
\DeclareMathOperator{\image}{im}
\DeclareMathOperator{\nullspace}{null}
\DeclareMathOperator{\area}{area}
\DeclareMathOperator{\vspan}{span}
\DeclareMathOperator{\id}{Id}
\DeclareMathOperator{\cond}{cond}
\DeclareMathOperator{\prox}{prox}
\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator*{\minimize}{minimize}
\DeclareMathOperator{\diag}{diag}
\DeclareMathOperator{\Tr}{Tr}

\graphicspath{{figs/}}

\begin{document}

\title{One Dimensional Random Projection and Clustering}

%\author{}
%\email{guifranca@gmail.com}
%\affiliation{Johns Hopkins University, Center for Imaging Science}

\begin{figure}
\begin{minipage}{.49\textwidth}
\includegraphics[scale=.45]{2d_gauss_separate.pdf}
\end{minipage}
\begin{minipage}{.49\textwidth}
\begin{tabular}{ c | c | c | c }
~$n$~ & $k$-means & PCA $k$-means & RAN $k$-means \\
\hline
1 & 0.9735 & 0.9735 & 0.9745 \\
2 & 0.982 & 0.9815 & 0.9825 \\
3 & 0.9785 & 0.9785 & 0.982 \\
\hline
\end{tabular}
\end{minipage}
\caption{\label{fig:2d_gauss_sep}
We have $x \sim \tfrac{1}{2}\left( \mathcal{N}(\mu_1, I) +
\mathcal{N}(\mu_2, I)\right)$ where $\mu_1 = (0,0)^T$ and $\mu_2=(4,0)^T$,
and $1000$ points on each cluster.
}
\end{figure}

\begin{figure}
\begin{minipage}{.49\textwidth}
\includegraphics[scale=.45]{2d_cigar.pdf}
\end{minipage}
\begin{minipage}{.49\textwidth}
\begin{tabular}{ c | c | c | c }
~$n$~ & $k$-means & PCA $k$-means & RAN $k$-means \\
\hline
1 & 0.5265 & 0.5215 & 1.0 \\
2 & 0.5285 & 0.5285 & 1.0 \\
3 & 0.5285 & 0.513 & 1.0 \\
\hline
\end{tabular}
\end{minipage}
\caption{\label{fig:2d_gauss_sep}
We have $x \sim \tfrac{1}{2}\left( \mathcal{N}(\mu_1, \Sigma) +
\mathcal{N}(\mu_2, \Sigma)\right)$ where $\mu_1 = (0,0)^T$, $\mu_2=(5,0)^T$,
and $\Sigma = \left( \begin{smallmatrix} 1/2 & 0 \\ 0 & 15 \end{smallmatrix}
\right)$,
and $1000$ points on each cluster.
}
\end{figure}

\begin{figure}
\begin{minipage}{.49\textwidth}
\includegraphics[scale=.45]{30d_gauss.pdf}
\end{minipage}
\begin{minipage}{.49\textwidth}
\renewcommand*{\arraystretch}{.3}
\begin{tabular}{ c | c | c | c }
~$D$~ & $k$-means & PCA $k$-means & RAN $k$-means \\
\hline
5 & 0.84 & 0.8415 & 0.841 \\
10 & 0.844 & 0.846 & 0.7985 \\
15 & 0.8325 & 0.8365 & 0.7465 \\
20 & 0.846 & 0.85 & 0.764 \\
25 & 0.8555 & 0.8505 & 0.714 \\
30 & 0.825 & 0.8225 & 0.728 \\
50 & 0.8485 & 0.8465 & 0.6775 \\
100 & 0.832 & 0.8325 & 0.644 \\
200 & 0.8085 & 0.8145 & 0.592 \\
300 & 0.7645 & 0.794 & 0.5755 \\
500 & 0.6915 & 0.756 & 0.5615 \\
1000 & 0.5075 & 0.6965 & 0.562 \\
2000 & 0.5495 & 0.662 & 0.543 \\
5000 & 0.531 & 0.5135 & 0.539 \\
\hline
\end{tabular}
\end{minipage}
\caption{\label{fig:2d_gauss_sep}
High dimensions.
We have $x \sim \tfrac{1}{2}\left( \mathcal{N}(\mu_1, I_D) +
\mathcal{N}(\mu_2, I_D)\right)$ where $\mu_1 = (0,0,\dotsc,0)^T$,
$\mu_2=(1,0,\dots,0)^T$,
and $1000$ points on each cluster. We show only the two principal components
of the data in the plot above.
}
\end{figure}

\end{document}

### 2017-03-22 random_projection/evaluation.py

@@ -1,0 +1,32 @@
+
+from __future__ import division
+
+import numpy as np
+import scipy.optimize
+
+
+def accuracy(z, zh):
+    """Compute misclassification error, or better the accuracy which is
+    1 - error. Use Hungarian algorithm which is O(n^3) instead
+    of O(n!). z and zh are vectors with the dimension
+    being the number of points, and each entry is the cluster label assigned
+    to that point.
+
+    """
+    n, k = len(z), len(np.unique(z))
+    m, q = len(zh), len(np.unique(zh))
+
+    assert m == n
+    
+    Q = np.zeros((n,k))
+    for i in range(n):
+        Q[i, z[i]] = 1
+    Qh = np.zeros((n,k))
+    for i in range(m):
+        Qh[i, zh[i]] = 1
+    
+    cost_matrix = Qh.T.dot(Q)
+    row_ind, col_ind = scipy.optimize.linear_sum_assignment(-cost_matrix)
+    
+    return cost_matrix[row_ind, col_ind].sum()/n
+

### 2017-03-22 random_projection/draft/1dproj.tex
random

@@
 \usepackage{amsmath,amssymb,amsfonts,amsthm,amscd,bm}
+\usepackage{bbm}
 \usepackage{algorithm}
@@
 
-\title{One Dimensional Random Projection and Clustering}
+\title{Discussion about $k$-Means and $1D$ Random Projections}
 
-%\author{}
+\author{Guilherme Fran\c ca}
 %\email{guifranca@gmail.com}
 %\affiliation{Johns Hopkins University, Center for Imaging Science}
+
+\maketitle
+
+\section{Procedure}
+
+Given data $X=\{ x_i \}_{i=1}^{n}$, where $x_i \in \mathbb{R}^{D}$, 
+and the number of clusters $k$, we perform the following experiments:
+\begin{enumerate}
+\item Run $k$-means$++$ 
+on the original data space. This is the column named 
+``$k$-means'' in the following tables.
+\item Use PCA to 
+project the data in the first principal component, 
+$Y=\{ y_i \}_{i=1}^n$ where $y_i = u_{1}\cdot x_i \in \mathbb{R}$, then
+apply $k$-means in this $1$-dimensional space. This is the column named
+``PCA'' in the following tables.
+\item We randomly project the data in one dimension by picking a vector
+$w$ such that $w_i \sim \mathcal{N}(0,1)$ and normalize it $\| w \|=1$.
+Thus $Y = \{y_i\}_{i=1}^n $ where $y_i = w\cdot x_i \in \mathbb{R}$. 
+We apply $k$-means in this randomly projected
+$1$-dimensional space. We do this several times and pick the best answer.
+This is the column named ``Random'' in the fllowing tables.
+\end{enumerate}
+
+The evaluation of the clustering procedure 
+will be based on the true labels by the following quantity, called
+accuracy:
+\begin{equation}
+\label{eq:accuracy}
+a(z,\hat{z}) = 
+\dfrac{1}{n}\sum_{i=1}^{n} \mathbbm{1}\left(z_i = \pi(\hat{z}_i) \right)
+\end{equation}
+where $z$ is an $n$-dimensional vector containing the true labels, entry
+$z_i$ corresponds to point $x_i$, and $\hat{z}$ is the estimated labels
+through the clustering procedure. $\pi$ is a permutation of the labels.
+Thus the above formula gives $1$ if all points were correctly classified and
+$0$ if all points were wrongly classified. In a two class problem with
+the same number of points, $a=1/2$ corresponds to picking the points in each
+cluster at random. This quantity $a$ is the number shown in the following tables.
+
+\paragraph*{How to choose the best answer for random projections?}
+Consider $k$-means objective function
+\begin{equation}
+\label{eq:J}
+J(X) = \dfrac{1}{2} \sum_{j=1}^k \sum_{x \in \mathcal{C}_j} \| x - m_j \|^2,
+\qquad
+m_j = \dfrac{1}{n_j}\sum_{x\in\mathcal{C}_j} x.
+\end{equation}
+Above $m_j$ is the center
+of cluster $\mathcal{C}_j$, and 
+$n_j$ is the number of elements in cluster $\mathcal{C}_j$. $k$-means
+algorithm solves the problem $\min_{\{C_j\}} J(X)$.
+
+One might think
+that a good criteria to choose the best answer is to pick the minimum $J(Y)$
+computed in the $1D$ randomly projected space. This doesn't work because
+each random projection gives different values for the data $Y$ which in general
+does not preserve the structure in the original
+distribution. For instance, considering two random projections yielding
+$Y_1$ and $Y_2$. It can happen that the clustering on $Y_2$ gives a better
+accuracy than the clustering on $Y_1$ and even though $J(Y_1) < J(Y_2)$.
+So we are really comparing apples and bananas here.
+
+Another approach would be to cluster on $Y$ yielding labels $\hat{z}$, then
+compute $J(X)$ based on these labels, and pick the smallest $J(X)$  computed
+on the original data. This only works in case where $k$-means in the original
+space itself provides a good answer. For cases where $k$-means is problematic,
+e.g. the data are not spherical gaussians, then $J(X)$ itself is not a good
+function to detect the clustering. This happens for parallel cigars in the
+example below.
+
+Thus if one still desires to follow this approach, a good function which
+does not depend on the true labels must be chosen. I tried to use the energy
+function for this, and still didn't quite work. Also, observe that the 
+energy function on the original space is expensive to compute, 
+$O(n^2)$.
+
+In the following experiments I will \emph{cheat} and use \eqref{eq:accuracy}
+to pick the best answer. The purpose of this is to see if the ``truth'' can
+be captured by random projections.
+
+\section{Experiments}
+
+In the first experiment shown in Fig.~\ref{fig:2d_gauss_sep} we choose
+two well separated gaussians in $2D$. All of these procedures give good
+results.
 
@@
 \begin{minipage}{.49\textwidth}
-\begin{tabular}{ c | c | c | c }
-~$n$~ & $k$-means & PCA $k$-means & RAN $k$-means \\
+\begin{tabular}{ l  l  l }
 \hline
-1 & 0.9735 & 0.9735 & 0.9745 \\
-2 & 0.982 & 0.9815 & 0.9825 \\
-3 & 0.9785 & 0.9785 & 0.982 \\
+$k$-means~~ & PCA~~~~ & Random \\
+\hline
+0.9735 & 0.9735 & 0.9755 \\
+0.978 & 0.978 & 0.979 \\
+0.977 &  0.9775 & 0.978 \\
 \hline
@@
 \mathcal{N}(\mu_2, I)\right)$ where $\mu_1 = (0,0)^T$ and $\mu_2=(4,0)^T$,
-and $1000$ points on each cluster.
+and $1000$ points on each cluster. We run the experiment three times.
 }
 \end{figure}
+
+In the experiment of Fig.~\ref{fig:cigar}, still in $2D$, we choose
+parallel cigars. Both $k$-means and PCA cannot perform well, however random
+projections can do well. This is not surprising because this data
+can be linearly separable in $1D$. After many tries random projections will
+find the correct line. Probably, LDA can perform as well on this example.
 
@@
 \begin{minipage}{.49\textwidth}
-\begin{tabular}{ c | c | c | c }
-~$n$~ & $k$-means & PCA $k$-means & RAN $k$-means \\
+\begin{tabular}{  l  l  l }
 \hline
-1 & 0.5265 & 0.5215 & 1.0 \\
-2 & 0.5285 & 0.5285 & 1.0 \\
-3 & 0.5285 & 0.513 & 1.0 \\
+$k$-means~~ & PCA~~~~ & Random \\
+\hline
+0.501 & 0.5075 & 0.9995 \\
+0.522 & 0.515 & 1.0 \\
+0.505 & 0.5055 & 0.9995 \\
 \hline
@@
 \end{minipage}
-\caption{\label{fig:2d_gauss_sep}
+\caption{\label{fig:cigar}
 We have $x \sim \tfrac{1}{2}\left( \mathcal{N}(\mu_1, \Sigma) +
@@
 \right)$,
-and $1000$ points on each cluster.
+and $1000$ points on each cluster. We run the experiment three times.
 }
 \end{figure}
+
+In the experiment of Fig.~\ref{fig:highd} we increase the number of
+dimensions of the gaussian distributions. Both $k$-means and PCA perform well
+if the dimension is not too high, while $1D$ random projections provide
+poor results. This is also expected since randomly projecting high dimensional
+data in a very low dimensional space practically destroy any information
+about the original distribution.
 
@@
 \renewcommand*{\arraystretch}{.3}
-\begin{tabular}{ c | c | c | c }
-~$D$~ & $k$-means & PCA $k$-means & RAN $k$-means \\
+\begin{tabular}{ l l l l }
+\hline
+$D$ & $k$-means~~ & PCA~~~~  & Random \\
 \hline
@@
 \end{minipage}
-\caption{\label{fig:2d_gauss_sep}
+\caption{\label{fig:highd}
 High dimensions.
@@
 $\mu_2=(1,0,\dots,0)^T$,
-and $1000$ points on each cluster. We show only the two principal components
+and $1000$ points on each cluster. We show the two principal components
 of the data in the plot above.

### 2017-03-24 random_projection/draft/1dproj.tex
random projections

@@
 \item Run $k$-means$++$ 
-on the original data space. This is the column named 
+on the original data. This is the column named 
 ``$k$-means'' in the following tables.
@@
 $1$-dimensional space. We do this several times and pick the best answer.
-This is the column named ``Random'' in the fllowing tables.
+This is the column named ``$k$-random'' in the following tables.
+We will choose 100 random projections.
+\item We use random projections as in the previous item, but use the $T$-test
+from energy statistics. In $1D$ we can compute the energy distance in an
+efficient manner if we sort the data $Y$. This is the column called
+``$\mathcal{E}$-random'' in the following tables.
+We will choose 100 random projections.
 \end{enumerate}
@@
 through the clustering procedure. $\pi$ is a permutation of the labels.
-Thus the above formula gives $1$ if all points were correctly classified and
-$0$ if all points were wrongly classified. In a two class problem with
+Thus the above formula gives $a=1$ if all points were correctly classified and
+$a=0$ if all points were wrongly classified. In a two class problem with
 the same number of points, $a=1/2$ corresponds to picking the points in each
-cluster at random. This quantity $a$ is the number shown in the following tables.
+cluster at random. 
+This quantity $a$ is the number shown in the following tables.
 
 \paragraph*{How to choose the best answer for random projections?}
-Consider $k$-means objective function
-\begin{equation}
-\label{eq:J}
-J(X) = \dfrac{1}{2} \sum_{j=1}^k \sum_{x \in \mathcal{C}_j} \| x - m_j \|^2,
-\qquad
-m_j = \dfrac{1}{n_j}\sum_{x\in\mathcal{C}_j} x.
-\end{equation}
-Above $m_j$ is the center
-of cluster $\mathcal{C}_j$, and 
-$n_j$ is the number of elements in cluster $\mathcal{C}_j$. $k$-means
-algorithm solves the problem $\min_{\{C_j\}} J(X)$.
+In the following experiments we are going to cheat and pick the best answer
+according to the highest value of \eqref{eq:accuracy}. This is cheating because
+we are using the true labels, however, this test allows us to easily
+check if the ``truth'' can be captured by these procedures involving
+random projections. If the results
+are poor in this way, no matter what criteria we choose, optimizing 
+$\mathcal{E}$ or any other function will also not work.
 
-One might think
-that a good criteria to choose the best answer is to pick the minimum $J(Y)$
-computed in the $1D$ randomly projected space. This doesn't work because
-each random projection gives different values for the data $Y$ which in general
-does not preserve the structure in the original
-distribution. For instance, considering two random projections yielding
-$Y_1$ and $Y_2$. It can happen that the clustering on $Y_2$ gives a better
-accuracy than the clustering on $Y_1$ and even though $J(Y_1) < J(Y_2)$.
-So we are really comparing apples and bananas here.
+\paragraph*{Note:} The procedures involving random projections are much
+slower than simple $k$-means in the original data. It is also much slower
+than PCA$+$$k$-means. Moreover, computing random projections in high
+dimensions is also a little expensive. Doing this many times can make
+the algorithm really slow.
 
-Another approach would be to cluster on $Y$ yielding labels $\hat{z}$, then
-compute $J(X)$ based on these labels, and pick the smallest $J(X)$  computed
-on the original data. This only works in case where $k$-means in the original
-space itself provides a good answer. For cases where $k$-means is problematic,
-e.g. the data are not spherical gaussians, then $J(X)$ itself is not a good
-function to detect the clustering. This happens for parallel cigars in the
-example below.
+%Consider $k$-means objective function
+%\begin{equation}
+%\label{eq:J}
+%J(X) = \dfrac{1}{2} \sum_{j=1}^k \sum_{x \in \mathcal{C}_j} \| x - m_j \|^2,
+%\qquad
+%m_j = \dfrac{1}{n_j}\sum_{x\in\mathcal{C}_j} x.
+%\end{equation}
+%Above $m_j$ is the center
+%of cluster $\mathcal{C}_j$, and 
+%$n_j$ is the number of elements in cluster $\mathcal{C}_j$. $k$-means
+%algorithm solves the problem $\min_{\{C_j\}} J(X)$. 
+%
+%One might think
+%that a good criteria to choose the best answer is to pick the minimum $J(Y)$
+%computed in the $1D$ randomly projected space. This doesn't work because
+%each random projection gives different values for the data $Y$ which in general
+%does not preserve the structure in the original
+%distribution. For instance, considering two random projections yielding
+%$Y_1$ and $Y_2$. It can happen that the clustering on $Y_2$ gives a better
+%accuracy than the clustering on $Y_1$ and even though $J(Y_1) < J(Y_2)$.
+%So we are really comparing apples and bananas here.
+%
+%Another approach would be to cluster on $Y$ yielding labels $\hat{z}$, then
+%compute $J(X)$ based on these labels, and pick the smallest $J(X)$  computed
+%on the original data. This only works in case where $k$-means in the original
+%space itself provides a good answer. For cases where $k$-means is problematic,
+%e.g. the data are not spherical gaussians, then $J(X)$ itself is not a good
+%function to detect the clustering. This happens for parallel cigars in the
+%example below.
+%
+%Thus if one still desires to follow this approach, a good function which
+%does not depend on the true labels must be chosen. I tried to use the energy
+%function for this, and still didn't quite work. Also, observe that the 
+%energy function on the original space is expensive to compute, 
+%$O(n^2)$.
+%
+%In the following experiments I will \emph{cheat} and use \eqref{eq:accuracy}
+%to pick the best answer. The purpose of this is to see if the ``truth'' can
+%be captured by random projections.
 
-Thus if one still desires to follow this approach, a good function which
-does not depend on the true labels must be chosen. I tried to use the energy
-function for this, and still didn't quite work. Also, observe that the 
-energy function on the original space is expensive to compute, 
-$O(n^2)$.
-
-In the following experiments I will \emph{cheat} and use \eqref{eq:accuracy}
-to pick the best answer. The purpose of this is to see if the ``truth'' can
-be captured by random projections.
-
-\section{Experiments}
+\section{First Experiments}
 
@@
 \begin{minipage}{.49\textwidth}
-\begin{tabular}{ l  l  l }
+\begin{tabular}{ l  l  l l}
 \hline
-$k$-means~~ & PCA~~~~ & Random \\
+$k$-means~ & PCA~~~ & $k$-random~ & $\mathcal{E}$-random \\
 \hline
-0.9735 & 0.9735 & 0.9755 \\
-0.978 & 0.978 & 0.979 \\
-0.977 &  0.9775 & 0.978 \\
+0.9735 &
+0.9735 &
+0.9745 &
+0.976 \\
+%
+0.9745 &
+0.975 &
+0.976 &
+0.975 \\
+%
+0.973 &
+0.9735 & 
+0.976 &
+0.976 \\
+%
 \hline
@@
 \begin{minipage}{.49\textwidth}
-\begin{tabular}{  l  l  l }
+\begin{tabular}{ l l l l}
 \hline
-$k$-means~~ & PCA~~~~ & Random \\
+$k$-means~ & PCA~~~ & $k$-random~ & $\mathcal{E}$-random \\
 \hline
-0.501 & 0.5075 & 0.9995 \\
-0.522 & 0.515 & 1.0 \\
-0.505 & 0.5055 & 0.9995 \\
+0.5085 &
+0.5025 &
+1.0 &
+1.0 \\
+%
+0.5025 &
+0.5035 &
+1.0 &
+1.0 \\
+%
+0.503 &
+0.5 &
+1.0 &
+1.0 \\
+%
 \hline
@@
 and $\Sigma = \left( \begin{smallmatrix} 1/2 & 0 \\ 0 & 15 \end{smallmatrix}
+\right)$,
+and $1000$ points on each cluster. We run the experiment three times.
+}
+\end{figure}
+
+In the experiment of Fig.~\ref{fig:weird1} notice that the $k$-means with
+random projections performs better than any other method.
+
+\begin{figure}
+\begin{minipage}{.49\textwidth}
+\includegraphics[scale=.45]{2d_weird1.pdf}
+\end{minipage}
+\begin{minipage}{.49\textwidth}
+\begin{tabular}{ l l l l}
+\hline
+$k$-means~ & PCA~~~ & $k$-random~ & $\mathcal{E}$-random \\
+\hline
+0.6315 &
+0.6295 &
+0.7235 & 
+0.66 \\
+%
+0.725 &
+0.7065 &
+0.726 &
+0.678 \\
+%
+0.6405 &
+0.6465 &
+0.724 & 
+0.679 \\
+%
+\hline
+\end{tabular}
+\end{minipage}
+\caption{\label{fig:weird1}
+We have $x \sim \tfrac{1}{2}\left( \mathcal{N}(\mu_1, \Sigma_1) +
+\mathcal{N}(\mu_2, \Sigma_2)\right)$ where $\mu_1 = (0,0)^T$, $\mu_2=(2,1)^T$,
+$\Sigma_1 = \left( \begin{smallmatrix} 0.5 & -0.8 \\ -0.8 & 15 
+\end{smallmatrix}
+\right)$,
+$\Sigma_2 = \left( \begin{smallmatrix} 15 & 1 \\ 1 & 1 \end{smallmatrix}
 \right)$,
@@
 \begin{minipage}{.49\textwidth}
+\centering
 \includegraphics[scale=.45]{30d_gauss.pdf}
 \end{minipage}
-\begin{minipage}{.49\textwidth}
+\begin{minipage}{.5\textwidth}
 \renewcommand*{\arraystretch}{.3}
-\begin{tabular}{ l l l l }
+\begin{tabular}{l l l l l}
 \hline
-$D$ & $k$-means~~ & PCA~~~~  & Random \\
+$D$ & $k$-means & PCA & $k$-random & $\mathcal{E}$-random \\
 \hline
-5 & 0.84 & 0.8415 & 0.841 \\
-10 & 0.844 & 0.846 & 0.7985 \\
-15 & 0.8325 & 0.8365 & 0.7465 \\
-20 & 0.846 & 0.85 & 0.764 \\
-25 & 0.8555 & 0.8505 & 0.714 \\
-30 & 0.825 & 0.8225 & 0.728 \\
-50 & 0.8485 & 0.8465 & 0.6775 \\
-100 & 0.832 & 0.8325 & 0.644 \\
-200 & 0.8085 & 0.8145 & 0.592 \\
-300 & 0.7645 & 0.794 & 0.5755 \\
-500 & 0.6915 & 0.756 & 0.5615 \\
-1000 & 0.5075 & 0.6965 & 0.562 \\
-2000 & 0.5495 & 0.662 & 0.543 \\
-5000 & 0.531 & 0.5135 & 0.539 \\
+%
+5 & 0.934500 & 0.934500 & 0.919000 & 0.924000 \\
+10 & 0.939500 & 0.941500 & 0.912000 & 0.856000 \\
+15 & 0.933000 & 0.935000 & 0.882000 & 0.816000 \\
+20 & 0.940000 & 0.942000 & 0.762000 & 0.817000 \\
+25 & 0.931500 & 0.933500 & 0.767000 & 0.759000 \\
+30 & 0.936000 & 0.934000 & 0.754000 & 0.725000 \\
+50 & 0.946000 & 0.942500 & 0.718500 & 0.741000 \\
+100 & 0.926500 & 0.927000 & 0.658500 & 0.647000 \\
+200 & 0.940000 & 0.940000 & 0.604000 & 0.647000 \\
+300 & 0.918000 & 0.921500 & 0.592500 & 0.602000 \\
+500 & 0.915000 & 0.922000 & 0.596500 & 0.576000 \\
+1000 & 0.874000 & 0.890500 & 0.552000 & 0.574000 \\
+2000 & 0.597000 & 0.863500 & 0.542000 & 0.554000 \\
+5000 & 0.509000 & 0.788000 & 0.543000 & 0.540000 \\
+%
 \hline
@@
 \mathcal{N}(\mu_2, I_D)\right)$ where $\mu_1 = (0,0,\dotsc,0)^T$,
-$\mu_2=(1,0,\dots,0)^T$,
+$\mu_2=(3,0,\dots,0)^T$,
 and $1000$ points on each cluster. We show the two principal components
@@
 
+\section{Second Experiments}
+
+We repeated the experiments above using the objective function as a criteria
+instead of the accuracy \eqref{eq:accuracy} so we do not use the true labels.
+We computed the objective function in the $1D$ projected space.
+
+For $k$-means
+the function is the sum of squared distances from the points 
+to the centers of the clusters,
+and for energy statistics we compute the ``energy'' $\mathcal{E}$ in the
+$1D$ space, which can be done in $O(n \log n)$. The results are just slightly
+worse than before because not always the optimal function coincides with
+the best clustering. Notice that we are comparing these functions on 
+different data sets, one for each projection. Thus this is not a very
+consistent way of doing the test.
+
+\section{Conclusions}
+
+These are just my thoughts about why random projections in $1D$ 
+does not work
+for \emph{clustering}---it surely
+must work with random projections in higher dimensions though, since
+we have Johnson-Lindenstrauss lemma that guarantees that distances
+are preserved in lower dimensional spaces, but they are not so 
+low-dimensional---regardless of 
+the function you are trying to optimize,
+e.g. $k$-means objective function, energy statistics, whatever, etc.
+
+In $1D$, random projections just seems to me a more expensive way of doing
+LDA. In $1D$ the best line which separates the data (I believe) is the line
+passing through the means of each cluster (I'm thinking about two clusters
+only). If data is not linearly separable
+in $1D$, neither of both methods work.
+
+Although $\mathcal{E}$ seems a nice function, which can distinguish different
+distributions in a non-parametric way, random projections in general destroy
+a lot of information from the original distribution. Therefore, it does not
+matter which function you try to optimize, the information was already lost.
+Moreover, projecting data in $1D$ clumps data together, which is the opposite
+of separating the data, which can be achieved only in higher dimensions.
+
+Also, $1D$ random projections are not necessarily cheap. Suppose we
+have $n$ points in $D$ dimensions, 
+$k$ clusters, and we compute $p$ random projections.
+The cost of each $1D$ random projection is $O(d n)$. Suppose the cost
+of optimizing the function is $O(n)$ (which is the best possible scenario).
+The total cost would be $O(p D n)$ in the best possible case. Roughly,
+$k$-means on the original data can be $O(k D n)$. Since $p \gg k$ usually
+random projections may be more expensive than pure $k$-means.
+In our experiment this was always the case.
+
+One may think about this criteria as an initialization procedure. I still
+believe this is not so good (hope to be wrong though).
+I think a way better proposal would be the following.
+Suppose we have $N$ data points in $D$-dimensions where both can be really
+large, so $k$-means++ initialization is not feasible. Then sample
+$M \lll N$ points uniformly. Now do a few random projections in
+$O(\log(M / \epsilon^2)$ dimensions, thus obeying the Johnson-Lindenstrauss
+lemma which assures that in this case we preserve distances in the original
+space. Now apply initialization procedure in this reduced number of points and
+in lower dimensions. Also, applying the whole clustering procedure in this
+new data set should give a descent estimate for clustering in the entire
+data set.
+
 \end{document}

### 2017-03-24 random_projection/projection.py

@@ -1,171 +1,0 @@
-
-from __future__ import division
-
-import numpy as np
-import matplotlib.pyplot as plt
-from sklearn import random_projection
-from sklearn.cluster import KMeans
-from sklearn import decomposition
-from sklearn.metrics import silhouette_score
-
-from evaluation import accuracy
-
-import sys
-
-def kmeans(X):
-    kmeans = KMeans(n_clusters=2).fit(X)
-    z = kmeans.labels_
-    #mu = kmeans.cluster_centers_
-    #J = -kmeans.score(X)
-    return z
-
-def kmeans_function(X, z):
-    J = 0
-    for k in np.unique(z):
-        idx = np.where(z==k)
-        Xks = X[idx]
-        muk = Xks.mean(axis=0)
-        J += (((Xks - muk)**2).sum())
-    n = X.shape[0]
-    d = muk.shape[0]
-    return J
-
-def mean1D(A, B):
-    """Assume A and B are a sorted list of elements. Compute the energy
-    distance in 1D in O(N).
-
-    """
-    nA = len(A)
-    nB = len(B)
-    sumA = 0
-    sumB = 0
-    i = j = 0
-    while i < nA and j < nB:
-        if A[i] <= B[j]:
-            sumA += A[i]*(2*j-nB)/nB
-            i += 1
-        else:
-            sumB += B[j]*(2*i-nA)/nA
-            j += 1
-    if i >= nA:
-        sumB += B[j:nB].sum()
-    else:
-        sumA += A[i:nA].sum()
-    return sumA/nA + sumB/nB
-
-def energy1D(A, B):
-    return 2*mean1D(A, B) - mean1D(A, A) - mean1D(B, B)
-
-def mean(A, B):
-    """Compute energy distance from standard formula."""
-    sum = 0
-    for x in A:
-        for y in B:
-           sum += np.linalg.norm(x-y)
-    return sum/(len(A)*len(B))
-
-def energy(A, B):
-    return 2*mean(A, B) - mean(A, A) - mean(B, B)
-
-def two_gaussians(d):
-    """"Generate data from two Gaussians."""
-
-    n1 = n2 = 1000
-
-    m1 = np.zeros(d)
-    s1 = np.eye(d)
-    X1 = np.random.multivariate_normal(m1, s1, n1)
-    z1 = [0]*n1
-
-    m2 = m1
-    m2[0] = 4
-    s2 = np.eye(d)
[…]

### 2017-04-05 discussion/discussion.tex
new initialization

--- 2017-03-22 random_projection/draft/1dproj.tex
@@
 
-\title{One Dimensional Random Projection and Clustering}
+\title{Notes about Statistics, Clustering, Graphs, \ldots}
 
-%\author{}
-%\email{guifranca@gmail.com}
-%\affiliation{Johns Hopkins University, Center for Imaging Science}
+\author{Guilherme Fran\c ca}
+
+\begin{abstract}
+In this document we collect notes about our experiments and theories
+so we can discuss together with the group in an organized way.
+\end{abstract}
+
+\maketitle
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Experiments using Energy Statistics and One-Dimensional 
+Random Projections (Gui 04-01-2017)}
+
+Given data $X=\{ x_i \}_{i=1}^{n}$, where $x_i \in \mathbb{R}^{D}$, 
+and the number of clusters $k$, we perform the following experiments:
+\begin{enumerate}
+\item Run $k$-means++ 
+on the original data. This is the column named 
+``$k$-means'' in the following tables.
+\item Use PCA to 
+project the data in the first principal component, 
+$Y=\{ y_i \}_{i=1}^n$ where $y_i = u_{1}\cdot x_i \in \mathbb{R}$, then
+apply $k$-means++ in this $1$-dimensional space. This is the column named
+``PCA'' in the following tables.
+\item We randomly project the data in one dimension by picking a vector
+$w$ such that $w_i \sim \mathcal{N}(0,1)$ and normalize it $\| w \|=1$.
+Thus $Y = \{y_i\}_{i=1}^n $ where $y_i = w\cdot x_i \in \mathbb{R}$. 
+We apply $k$-means++ in this randomly projected
+$1$-dimensional space. We do this several times and pick the best answer,i.e.
+the one which has minumim objective function in the \emph{original} space,
+since it is cheap to compute this objective function.
+This is the column named ``$k$-random'' in the following tables.
+\item We use random projections as in the previous item, but use the $T$-test
+from energy statistics. In $1D$ we can compute the energy distance in an
+efficient manner if we sort the data $Y$. This is accomplished in
+$O(n\log n)$. This is the column called
+``$\mathcal{E}$-random'' in the following tables. We pick the best answer
+by computing the largest $T$, in the one-dimensional space.
+\end{enumerate}
+
+The evaluation of the clustering procedure 
+will be based on the true labels by the following quantity, called
+accuracy:
+\begin{equation}
+\label{eq:accuracy}
+a(z,\hat{z}) = 
+\dfrac{1}{n}\sum_{i=1}^{n} \mathbbm{1}\left(z_i = \pi(\hat{z}_i) \right)
+\end{equation}
+where $z$ is an $n$-dimensional vector containing the true labels, entry
+$z_i$ corresponds to point $x_i$, and $\hat{z}$ is the estimated labels
+through the clustering procedure. $\pi$ is a permutation of the labels.
+Thus the above formula gives $a=1$ if all points were correctly classified and
+$a=0$ if all points were wrongly classified. In a two class problem with
+the same number of points, $a=1/2$ corresponds to picking the points in each
+cluster at random. 
+This quantity $a$ is the number shown in the following tables.
+
+In the first experiment shown in Fig.~\ref{fig:2d_gauss_sep} we choose
+two well separated gaussians in $2D$. All of these procedures give good
+results.
 
@@
 \begin{minipage}{.49\textwidth}
-\begin{tabular}{ c | c | c | c }
-~$n$~ & $k$-means & PCA $k$-means & RAN $k$-means \\
+\begin{tabular}{ l  l  l l}
 \hline
-1 & 0.9735 & 0.9735 & 0.9745 \\
-2 & 0.982 & 0.9815 & 0.9825 \\
-3 & 0.9785 & 0.9785 & 0.982 \\
+$k$-means~ & PCA~~~ & $k$-random~ & $\mathcal{E}$-random \\
+\hline
+0.97775 &
+0.97825 & 
+0.79925 &
+0.977 \\
+0.975 &
+0.975 &
+0.7305 &
+0.976 \\
+0.9765 &
+0.9765 &
+0.9655 &
+0.9605 \\
+%
 \hline
@@
 \mathcal{N}(\mu_2, I)\right)$ where $\mu_1 = (0,0)^T$ and $\mu_2=(4,0)^T$,
-and $1000$ points on each cluster.
+and $1000$ points on each cluster. We run the experiment three times.
+We choose only $20$ random projections.
 }
@@
 \begin{minipage}{.49\textwidth}
-\begin{tabular}{ c | c | c | c }
-~$n$~ & $k$-means & PCA $k$-means & RAN $k$-means \\
+\begin{tabular}{ l l l l}
 \hline
-1 & 0.5265 & 0.5215 & 1.0 \\
-2 & 0.5285 & 0.5285 & 1.0 \\
-3 & 0.5285 & 0.513 & 1.0 \\
+$k$-means~ & PCA~~~ & $k$-random~ & $\mathcal{E}$-random \\
+\hline
+0.51325 &
+0.506 &
+0.70325 &
+0.977 \\
+0.52175 &
+0.5105 &
+0.54725 &
+0.7295 \\
+0.53325 &
+0.53175 &
+0.9785 &
+0.998 \\
+%
 \hline
@@
 \end{minipage}
-\caption{\label{fig:2d_gauss_sep}
+\caption{\label{fig:cigar}
 We have $x \sim \tfrac{1}{2}\left( \mathcal{N}(\mu_1, \Sigma) +
-\mathcal{N}(\mu_2, \Sigma)\right)$ where $\mu_1 = (0,0)^T$, $\mu_2=(5,0)^T$,
-and $\Sigma = \left( \begin{smallmatrix} 1/2 & 0 \\ 0 & 15 \end{smallmatrix}
+\mathcal{N}(\mu_2, \Sigma)\right)$ where $\mu_1 = (0,0)^T$, $\mu_2=(6,0)^T$,
+and $\Sigma = \left( \begin{smallmatrix} 1 & 0 \\ 0 & 20 \end{smallmatrix}
 \right)$,
-and $1000$ points on each cluster.
+and $2000$ points on each cluster. We run the experiment three times.
+We use $30$ random projections.
 }
@@
 \begin{minipage}{.49\textwidth}
-\includegraphics[scale=.45]{30d_gauss.pdf}
+\includegraphics[scale=.45]{2d_weird1.pdf}
 \end{minipage}
 \begin{minipage}{.49\textwidth}
-\renewcommand*{\arraystretch}{.3}
-\begin{tabular}{ c | c | c | c }
-~$D$~ & $k$-means & PCA $k$-means & RAN $k$-means \\
+\begin{tabular}{ l l l l}
 \hline
-5 & 0.84 & 0.8415 & 0.841 \\
-10 & 0.844 & 0.846 & 0.7985 \\
-15 & 0.8325 & 0.8365 & 0.7465 \\
-20 & 0.846 & 0.85 & 0.764 \\
-25 & 0.8555 & 0.8505 & 0.714 \\
-30 & 0.825 & 0.8225 & 0.728 \\
-50 & 0.8485 & 0.8465 & 0.6775 \\
-100 & 0.832 & 0.8325 & 0.644 \\
-200 & 0.8085 & 0.8145 & 0.592 \\
-300 & 0.7645 & 0.794 & 0.5755 \\
-500 & 0.6915 & 0.756 & 0.5615 \\
-1000 & 0.5075 & 0.6965 & 0.562 \\
-2000 & 0.5495 & 0.662 & 0.543 \\
-5000 & 0.531 & 0.5135 & 0.539 \\
+$k$-means~ & PCA~~~ & $k$-random~ & $\mathcal{E}$-random \\
+\hline
+0.71275 &
+0.68675 &
+0.51275 &
+0.6505 \\
+0.7235 & 
+0.6645 & 
+0.6095 & 
+0.641 \\
+0.72575 &
+0.68475 &
+0.60975 &
+0.509 \\
 \hline
@@
 \end{minipage}
-\caption{\label{fig:2d_gauss_sep}
+\caption{\label{fig:weird1}
+We have $x \sim \tfrac{1}{2}\left( \mathcal{N}(\mu_1, \Sigma_1) +
+\mathcal{N}(\mu_2, \Sigma_2)\right)$ where $\mu_1 = (0,0)^T$, $\mu_2=(2,1)^T$,
+$\Sigma_1 = \left( \begin{smallmatrix} 0.5 & -0.8 \\ -0.8 & 15 
+\end{smallmatrix}
+\right)$,
+$\Sigma_2 = \left( \begin{smallmatrix} 15 & 1 \\ 1 & 1 \end{smallmatrix}
+\right)$,
+and $2000$ points on each cluster. We run the experiment three times,
+with 30 random projections.
+}
+\end{figure}
+
+In the experiment of Fig.~\ref{fig:highd} we increase the number of
+dimensions of the gaussian distributions. Both $k$-means and PCA perform well
+if the dimension is not too high, while $1D$ random projections provide
+poor results. This is also expected since randomly projecting high dimensional
+data in a very low dimensional space practically destroy any information
+about the original distribution.
+
+\begin{figure}
+\begin{minipage}{.49\textwidth}
+\centering
+\includegraphics[scale=.45]{30d_gauss.pdf}
+\end{minipage}
+\begin{minipage}{.5\textwidth}
+\renewcommand*{\arraystretch}{.3}
+\begin{tabular}{l l l l l}
+\hline
+$D$ & $k$-means & PCA & $k$-random & $\mathcal{E}$-random \\
+\hline
+%
+5 & 0.928000 & 0.928000 & 0.689500 & 0.859000 \\
+10 & 0.935500 & 0.935500 & 0.551000 & 0.890000 \\
+15 & 0.929000 & 0.930000 & 0.580500 & 0.839000 \\
+20 & 0.938000 & 0.940000 & 0.647000 & 0.741000 \\
+25 & 0.937500 & 0.936500 & 0.576500 & 0.801000 \\
+30 & 0.939000 & 0.938500 & 0.545000 & 0.565000 \\
+50 & 0.933500 & 0.930000 & 0.505500 & 0.701000 \\
+100 & 0.928500 & 0.932000 & 0.624500 & 0.565000 \\
+200 & 0.935000 & 0.935000 & 0.502000 & 0.528000 \\
+300 & 0.922500 & 0.924500 & 0.508500 & 0.527000 \\
+500 & 0.916500 & 0.923500 & 0.509000 & 0.513000 \\
+1000 & 0.881000 & 0.889500 & 0.503000 & 0.501000 \\
+2000 & 0.615000 & 0.867000 & 0.509500 & 0.505000 \\
+5000 & 0.534500 & 0.780000 & 0.514000 & 0.511000 \\
+%
+\hline
+\end{tabular}
+\end{minipage}
+\caption{\label{fig:highd}
 High dimensions.
@@
 \mathcal{N}(\mu_2, I_D)\right)$ where $\mu_1 = (0,0,\dotsc,0)^T$,
-$\mu_2=(1,0,\dots,0)^T$,
-and $1000$ points on each cluster. We show only the two principal components
+$\mu_2=(3,0,\dots,0)^T$,
+and $1000$ points on each cluster. We do 100 random projections.
+We show the two principal components
 of the data in the plot above.
@@
 
+\begin{figure}
+\begin{minipage}{.49\textwidth}
+\centering
+\includegraphics[scale=.45]{30d_more_signal_guass.pdf}
+\end{minipage}
+\begin{minipage}{.5\textwidth}
+\renewcommand*{\arraystretch}{.3}
+\begin{tabular}{l l l l l}
+\hline
+$D$ & $k$-means & PCA & $k$-random & $\mathcal{E}$-random \\
+\hline
+%
+5 & 0.963000 & 0.963000 & 0.615000 & 0.921000 \\
+10 & 0.985500 & 0.985000 & 0.758000 & 0.957000 \\
+15 & 0.998000 & 0.998000 & 0.537500 & 0.919000 \\
+20 & 0.998500 & 0.998500 & 0.505500 & 0.889000 \\
+25 & 1.000000 & 1.000000 & 0.530000 & 0.968000 \\
+30 & 1.000000 & 1.000000 & 0.733500 & 0.978000 \\
+50 & 1.000000 & 1.000000 & 0.640000 & 0.957000 \\
+100 & 1.000000 & 1.000000 & 0.887500 & 0.972000 \\
+200 & 1.000000 & 1.000000 & 0.708500 & 0.964000 \\
+300 & 1.000000 & 1.000000 & 0.549000 & 0.967000 \\
+500 & 1.000000 & 1.000000 & 0.653000 & 0.968000 \\
+1000 & 1.000000 & 1.000000 & 0.604500 & 0.983000 \\
+2000 & 1.000000 & 1.000000 & 0.535000 & 0.967000 \\
+5000 & 1.000000 & 1.000000 & 0.727500 & 0.949000 \\
+%
+\hline
+\end{tabular}
+\end{minipage}
+\caption{\label{fig:highd}
+High dimensions.
+We have $x \sim \tfrac{1}{2}\left( \mathcal{N}(\mu_1, I_D) +
+\mathcal{N}(\mu_2, I_D)\right)$ where 
+$\mu_1 = (1,0,1,\dotsc,0,1)^T$,
+$\mu_2 = (-1,0,-1,\dotsc,0,-1)^T$, i.e. the even dimensions are
+shifted at $+1$ and $-1$, respectivelly.
+We have $1000$ points on each cluster. We do 100 random projections.
+We show the two principal components
+of the data in the plot above. When we have more signal we can see that
+random projections with $\mathcal{E}$ is much superior to random projections
+and $k$-means++.
+}
+\end{figure}
+
+\subsection{Conclusions}
+If we have enough signal, even in high dimensions random projections with
+energy statistics seems like a good thing. I believe that at least for
+a initialization procedure this can be usefull. It will be interesting to
+have a benchmark of how many random projections we need. Too many is expensive,
+however, if this is an initialization procedure, it will be required only
+once. Maybe we can compare just $\mathcal{E}$-random with the initialization
+provided by $k$-means++.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Energy Statistics Based Clustering (Gui 04-05-2017)}
+
+We show some experiments regarding our method for clustering using energy
+statistics. We have a QCQP approximation to the original problem, and we solve
+this QCQP through a relaxation, thus there are two approximations going on
+here. We solve this through a kernel approach. The function has a maximum, but
+it is really sensitive to initialization, since apparently there are lots of
+local minima. We use energy statistics in $1D$ with random projections for
+initialization. This seems to provide a better initialization than pure random
+or even k-means.
+
+\begin{figure}
+\begin{minipage}{.49\textwidth}
+\includegraphics[scale=.45]{two_gaussians_separated.pdf}
+\end{minipage}
+\begin{minipage}{.49\textwidth}
+\begin{tabular}{ l l l l l}
+\hline
+Ker/$\mathcal{E}$ & Ker/Gauss & Ker/Poly & GMM~~ & $k$-means \\
+\hline
+0.97 &
+0.965 &
+0.863 &
+0.969 & 
+0.971 \\
+0.98 &
+0.98 &
+0.878 &
+0.982 &
+0.98 \\
+0.982 &
+0.982 &
+0.877 &
+0.983 &
+0.983 \\
+\hline
+\end{tabular}
+\end{minipage}
+\caption{\label{fig:weird1}
+We have $x \sim \tfrac{1}{2}\left( \mathcal{N}(\mu_1, \Sigma_1) +
+\mathcal{N}(\mu_2, \Sigma_2)\right)$ where $\mu_1 = (0,0)^T$, $\mu_2=(4,0)^T$,
+$\Sigma_1 = \Sigma_2 = I$,
+and $500$ points on each cluster. We run the experiment three times,
+with 20 random projections as initialization.
+}
+\end{figure}
+
+\begin{figure}
+\begin{minipage}{.49\textwidth}
+\includegraphics[scale=.45]{two_gaussians_closer.pdf}
+\end{minipage}
+\begin{minipage}{.49\textwidth}
+\begin{tabular}{ l l l l l}
+\hline
+Ker/$\mathcal{E}$ & Ker/Gauss & Ker/Poly & GMM~~ & $k$-means \\
+\hline
+0.844 &
+0.829 &
+0.688 &
+0.841 &
+0.843 \\
+0.834 &
+0.836 &
+0.695 &
+0.835 &
+0.834 \\
+0.843 &
+0.813 &
+0.68 &
+0.805 &
+0.841 \\
+\hline
+\end{tabular}
+\end{minipage}
+\caption{\label{fig:weird1}
+We have $x \sim \tfrac{1}{2}\left( \mathcal{N}(\mu_1, \Sigma_1) +
+\mathcal{N}(\mu_2, \Sigma_2)\right)$ where $\mu_1 = (0,0)^T$, $\mu_2=(2,0)^T$,
+$\Sigma_1 = \Sigma_2 = I$,
+and $500$ points on each cluster. We run the experiment three times,
+with 20 random projections as initialization.
+}
+\end{figure}
+
+\begin{figure}
+\begin{minipage}{.49\textwidth}
+\includegraphics[scale=.45]{blob_cigar.pdf}
+\end{minipage}
+\begin{minipage}{.49\textwidth}
+\begin{tabular}{ l l l l l}
+\hline
+Ker/$\mathcal{E}$ & Ker/Gauss & Ker/Poly & GMM~~ & $k$-means \\
+\hline
+0.988 &
+0.807 &
+0.603 &
+0.988 & 
+0.7 \\
+0.981 & 
+0.831 &
+0.603 &
+0.983 & 
+0.698 \\
+0.731 & 
+0.789 &
+0.626 & 
+0.986 &
+0.717 \\
+\\
+\hline
+\end{tabular}
+\end{minipage}
+\caption{\label{fig:weird1}
+We have $x \sim \tfrac{1}{2}\left( \mathcal{N}(\mu_1, \Sigma_1) +
+\mathcal{N}(\mu_2, \Sigma_2)\right)$ where $\mu_1 = (0,0)^T$, $\mu_2=(4,0)^T$,
+$\Sigma_1 = I$, 
+$\Sigma_2 = \left( \begin{smallmatrix} 
+1 & 0 \\ 1 & 20
+\end{smallmatrix}\right)$
+and $500$ points on each cluster. We run the experiment three times,
+with 20 random projections as initialization.
+}
+\end{figure}
+
 \end{document}

### 2017-04-05 qcqp_clustering/draft/energy_clustering.tex
new initialization

\documentclass[aps,preprint,nofootinbib,floatfix]{revtex4-1}

%\usepackage{cite}
\usepackage[pdftex]{graphicx}
\usepackage{amsmath,amssymb,amsfonts,amsthm,amscd,bm}
\usepackage{bbm}
\usepackage{algorithm}
\usepackage{algorithmic}

\graphicspath{{./figs}}

\hyphenation{op-tical net-works semi-conduc-tor}

\newtheorem{theorem}{Theorem}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{assumption}[theorem]{Assumption}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{example}{Example}

%% our definitions %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\DeclareMathOperator{\aff}{aff}
\DeclareMathOperator{\st}{s.t.}
\DeclareMathOperator{\LC}{LC}
\DeclareMathOperator{\affnot}{aff_0}
\DeclareMathOperator{\conv}{conv}
\DeclareMathOperator{\relint}{relint}
\DeclareMathOperator{\vol}{vol}
\DeclareMathOperator{\range}{range}
\DeclareMathOperator{\image}{im}
\DeclareMathOperator{\nullspace}{null}
\DeclareMathOperator{\area}{area}
\DeclareMathOperator{\vspan}{span}
\DeclareMathOperator{\id}{Id}
\DeclareMathOperator{\cond}{cond}
\DeclareMathOperator{\prox}{prox}
\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator*{\minimize}{minimize}
\DeclareMathOperator{\diag}{diag}
\DeclareMathOperator{\Tr}{Tr}

\def\Energy{\mathcal{E}}
\def\E{\mathbb{E}}

\begin{document}

\title{Nonparametric Clustering Based on Energy Distance}

\author{Guilherme Fran\c ca}
\email{guifranca@gmail.com}
\affiliation{Johns Hopkins University, Center for Imaging Science}

\author{Joshua Vogelstein}
\email{jovo@jhu.edu}
\affiliation{Johns Hopkins University, Center for Imaging Science}

\begin{abstract}
blabla
\end{abstract}

\maketitle

\section{Introduction}

Mention why energy is important, main results, where it was applied, etc.
Motivate how this can be used for clustering. Mention most important
papers on this \ldots Explain main results of this paper and give a brief
outline.

\section{Energy Statistics and RKHS}

In this section we briefly review the main concepts from energy
statistics and reproducing kernel Hilbert spaces (RKHS) used in this paper. 
For more details we refer the reader
to \cite{Szkely2013} (and references therein) and also \cite{Sejdinovic2013}.

Consider random variables in $\mathbb{R}^D$ 
such that $X,X' \stackrel{iid}{\sim} P$ and 
$Y,Y' \stackrel{iid}{\sim} Q$, where $P$ and $Q$ are cumulative
distribution functions with finite first moments. 
The quantity \cite{Szkely2013}
\begin{equation}\label{eq:energy}
\Energy(P, Q) = 2 \E \| X - Y\| - \| X - X' \| - \| Y - Y' \|,
\end{equation}
called energy distance, 
is rotational invariant and nonnegative, $\Energy(P,Q) \ge 0$, where
equality
to zero holds if and only if $P = Q$.
Above, $\| \cdot \|$ denotes the
Euclidean norm in $\mathbb{R}^D$. 
Energy distance
provides a characterization of equality of distributions, and
$\Energy^{1/2}$ is
a metric on the space of distributions.

The energy distance can be generalized as 
\begin{equation}\label{eq:energy2}
\Energy_\alpha(P, Q) = 2 \E \| X - Y\|^{\alpha} - \| X - X' \|^{\alpha} - 
\| Y - Y' \|^{\alpha}
\end{equation}
where $0<\alpha\le 2$. This quantity is also nonnegative,
$\Energy_\alpha(P,Q) \ge 0$. Furthermore, for $0<\alpha<2$ we have that
$\Energy_\alpha(P,Q) = 0$ if and only if $P=Q$, while for $\alpha=2$ 
we have $\Energy_2(P,Q) = 2\| \E(X) - \E(Y) \|^2$, showing that
equality to zero only requires
equality of the means and thus it does not imply equality of distributions.
It is important to 
mention that \eqref{eq:energy2} can be further generalized by
replacing
$\| X - Y\| \to \rho(X,Y)$, where $\rho$ is a so-called semimetric
of negative type \cite{Sejdinovic2013}. 
In this case there is a Hilbert space $\mathcal{H}$ and
a map $\varphi: \mathbb{R}^D \to
\mathcal{H}$ such that
$\rho(X, Y) = \| \varphi(X) - \varphi(Y) \|_{\mathcal{H}}^2$. 
Even though the semimetric 
$\rho$ may not satisfy the triangle inequality, its square
root $\rho^{1/2}$ does, since it is an actual metric.

There is an equivalence 
between energy distance, commonly used in statistics,
and distances between embeddings of distributions in reproducible
kernel Hilbert spaces (RKHS),
commonly used in machine learning. This equivalence was established
in \cite{Sejdinovic2013}. We recall the definition of
RKHS. Let $\mathcal{H}$ be a Hilbert space over real-valued functions
over $\mathbb{R}^D$. A function $k : \mathbb{R}^D \times \mathbb{R}^D \to 
\mathbb{R}$ is a reproducing kernel of $\mathcal{H}$ if it satisfies
the following two conditions:
\begin{enumerate}
\item $K_x \equiv k(\cdot, x) \in \mathcal{H}$ for any $x \in \mathbb{R}^D$.
\item $\langle K_x, f \rangle_{\mathcal{H}} = f(x)$ for
any $x\in\mathbb{R}^D$ and $f\in \mathcal{H}$.
\end{enumerate}
In other words, for any $x \in \mathbb{R}^D$ there is a unique function
$K_x \in \mathcal{H}$ that reproduces $f(x)$ through the inner product
of $\mathcal{H}$.
If such a kernel function $k$ exists, then $\mathcal{H}$ is called a RKHS.
From this we have $\langle K_x, K_y \rangle = K_y(x) = k(x,y)$. This implies
that $k(x,y)$ is symmetric and positive definite, $\sum_{i,j=1}^n c_i c_j
k(x_i,x_j) \ge 0$ for $c_i,c_j \in \mathbb{R}$.

The Moore-Aronszajn theorem establishes the converse. For every symmetric
and positive definite function $k: \mathbb{R}^D\times \mathbb{R}^D \to
\mathbb{R}$, there is an associated RKHS, $\mathcal{H}_k$, with reproducing
kernel $k$. The map $\varphi: x \mapsto K_x \in \mathcal{H}_k$ is called
the canonical feature map. Given a kernel $k$,
this theorem enables us to define an embedding of a probability measure
$P$ into the RKHS: $P \mapsto K_P \in
\mathcal{H}_k$ such that 
$\int f(x) d P(x) = \langle f, K_P \rangle$ for all $f \in \mathcal{H}_k$,
or alternatively $K_P = \int k(\cdot, x)  d P(x)$. 
We can now  introduce the 
notion of distance between two probability measures using the inner product
of $\mathcal{H}_k$. This is called the maximum mean discrepancy (MMD) and
it is given by
\begin{equation}\label{eq:mmd}
\gamma_k(P,Q) = \| K_P - K_Q \|_{\mathcal{H}_k},
\end{equation}
which can also be written as \cite{Gretton2012}
\begin{equation}\label{eq:mmd2}
\gamma_k^2(P,Q) = \E k(X,X') + \E k(Y,Y') - 2 \E k(X, Y)
\end{equation}
where $X,X' \stackrel{iid}{\sim} P$ and $Y,Y'\stackrel{iid}{\sim} Q$.
From the equality between \eqref{eq:mmd} and \eqref{eq:mmd2} we also
have that
\begin{equation}\label{eq:inner_data}
\langle K_P, K_Q \rangle_{\mathcal{H}_k} = \E \, k(X, Y).
\end{equation}
Thus we can compute the inner product between the embedded distributions 
by averaging the kernel function over sampled data.

The following important result shows that semimetrics of negative
type and symmetric positive definite kernels are closely related
\cite{Berg1984}. Let $\rho: \mathbb{R}^D \times \mathbb{R}^D \to \mathbb{R}$
be a semimetric, 
and $x_0 \in \mathbb{R}^D$ an arbitrary but fixed point.
Define
\begin{equation}\label{eq:kernel_semimetric}
k(x,y) = \tfrac{1}{2} \left\{  \rho(x,x_0) + \rho(y,x_0) - \rho(x,y)\right\}.
\end{equation}
Then $k$ is positive definite if and only if $\rho$ is of negative type.
Thus we have a family of kernels, one for each choice of $x_0$. Conversely,
if $\rho$ is a semimetric of negative type and $k$ is a kernel in this
family, then 
\begin{equation}\label{eq:gen_kernel}
\rho(x,y) = k(x,x) + k(y,y) -2k(x,y) = \| K_x - K_y
\|^2_{\mathcal{H}_k},
\end{equation}
and the canonical feature map 
$\varphi: x \mapsto K_x$ is injective \cite{Sejdinovic2013}.
We say that the kernel $k$ generates the semimetric $\rho$. 
If two different kernels generate the same $\rho$, they are
equivalent kernels.

Now we can state the equivalence between energy distance $\Energy$ and
inner products on RKHS, which is one of the main results of
\cite{Sejdinovic2013}. If $\rho$ is a semimetric
of negative type and $k$ a kernel that generates $\rho$, then
\begin{align}
\Energy(P, Q) &\equiv 2\E \rho(X,Y) - \E \rho(X,X') - \E \rho(Y,Y') 
\label{eq:EphoDef}\\
&= 2 \left[ \E \, k(X, X') + \E \, k(Y, Y') - 2\E \, k(X, Y)\right] \\
&=2 \gamma_k^2(P,Q) 
\end{align}
This result follows simply by substituting \eqref{eq:gen_kernel} into
\eqref{eq:EphoDef}, and using \eqref{eq:mmd2}.
Since $\gamma_k^2(P, Q) = \| K_P - K_Q \|^2_{\mathcal{H}_k}$, we
can compute the energy distance using the inner product in the RKHS. Moreover,
this can be computed from the data according to \eqref{eq:inner_data}.

\section{Clustering using Energy Distance}

\section{Numerical Experiments}

\section{Conclusion}

\subsection*{Acknowledgements}
We thank \ldots

\bibliographystyle{unsrt}
\bibliography{biblio.bib}

\end{document}

### 2017-04-05 qcqp_clustering/code/data.py

@@ -1,0 +1,93 @@
+from __future__ import division
+
+import numpy as np
+import matplotlib.pyplot as plt
+
+from sklearn.decomposition import PCA
+
+
+def multi_gaussians(means, sigmas, ns):
+    """Generate several normal distributed data points.
+    
+    Parameters: 
+        means = [m1, m2, ...]
+        sigmas = [s1, s2, ...]
+        ns = [n1, n2, ...]
+
+        Each parameter in these lists are associated to one multivariate
+        normal distribution. The data is shuffled so points are not ordered.
+    
+    Returns: data X and labels z.
+    
+    """
+    n = len(means)
+    X = []
+    z = []
+    k = 0
+    for mu, sigma, n in zip(means, sigmas, ns):
+        X.append(np.random.multivariate_normal(mu, sigma, n))
+        z.append([k]*n)
+        k += 1
+    X = np.concatenate(X)
+    z = np.concatenate(z)
+    idx = np.random.permutation(sum(ns))
+    return X[idx], z[idx]
+
+def circles(rs, eps, ns):
+    """Generate concentric circles with radius in rs and gaussian noise
+    in eps.
+    
+    """
+    k = 0
+    X = []
+    z = []
+    for r, e, n in zip(rs, eps, ns):
+        x = np.array([[r*np.cos(a), r*np.sin(a)] 
+                            for a in np.random.uniform(0, 2*np.pi, n)])
+        x += e*np.random.multivariate_normal([0,0],np.eye(2), n)
+        X.append(x)
+        z.append([k]*n)
+        k += 1
+    X = np.concatenate(X)
+    z = np.concatenate(z)
+    idx = np.random.permutation(sum(ns))
+    return X[idx], z[idx]
+
+def plot(X, z, fname='plot.pdf'):
+    """Plot clusters according to labels in z.
+    If data is multidimensional, pick the first two principal components.
+    
+    """
+
+    z_unique = np.unique(z)
+
+    if len(X[0]) > 2:
+        pca = PCA(n_components=2)
+        X_new = pca.fit_transform(X)
+    else:
+        X_new = X
+
+    fig = plt.figure()
+    ax = fig.add_subplot(111)
+    colors = iter(plt.cm.rainbow(np.linspace(0,1,len(z_unique))))
+    for k in z_unique:
+        idx = np.where(z==k)
[…]

### 2017-04-05 qcqp_clustering/code/energy.py

@@ -1,0 +1,57 @@
+
+from __future__ import division
+
+import numpy as np
+import matplotlib.pyplot as plt
+
+
+def fast_mean(X, Y):
+    """We can compute the mean in O(n) in 1D for sorted X and Y."""
+    sx = 0
+    sy = 0
+    i = 1
+    j = 1
+    nx = len(X)
+    ny = len(Y)
+
+    while i <= nx and j <= ny:
+        if X[i-1] <= Y[j-1]:
+            sx += ((j-1)-(ny-(j-1)))/ny*X[i-1]
+            i += 1
+        else:
+            sy += ((i-1)-(nx-(i-1)))/nx*Y[j-1]
+            j += 1
+    if i > nx:
+        sy += Y[j-1:ny].sum()
+    else:
+        sx += X[i-1:nx].sum()
+    return (sx/nx) + (sy/ny)
+            
+def fast_energy(X, Y):
+    #X = np.sort(A)
+    #Y = np.sort(B)
+    return 2*fast_mean(X,Y) - fast_mean(X,X) - fast_mean(Y,Y)
+
+def fastT(X, Y):
+    nx = len(X)
+    ny = len(Y)
+    return nx*ny/(nx+ny)*fast_energy(X, Y)
+
+def mean(X, Y):
+    """Compute mean of X and Y. This is O(n^2)."""
+    return sum([np.linalg.norm(x-y) for x in X for y in Y])/len(X)/len(Y)
+
+def energy(X, Y):
+    return 2*mean(X,Y) - mean(X,X) - mean(Y,Y)
+
+def T(X, Y):
+    nx = len(X)
+    ny = len(Y)
+    return nx*ny/(nx+ny)*energy(X, Y)
+
+
+if __name__ == "__main__":
+    X = np.random.normal(0,1,100)
+    Y = np.random.normal(2,1,100)
+    print T(X, Y)
+    print fastT(X, Y)

### 2017-04-05 qcqp_clustering/code/kkmeans.py

@@ -1,0 +1,123 @@
+
+import numpy as np
+
+from sklearn.base import BaseEstimator, ClusterMixin
+from sklearn.metrics.pairwise import pairwise_kernels
+from sklearn.utils import check_random_state
+from sklearn.cluster import KMeans
+
+from randproj import energy_multi_random
+
+
+class KernelKMeans(BaseEstimator, ClusterMixin):
+
+    def __init__(self, n_clusters=3, max_iter=50, tol=1e-3, random_state=None,
+                 kernel="linear", gamma=None, degree=3, coef0=1,
+                 kernel_params=None, verbose=0, init='kmeans', labels=None):
+        self.n_clusters = n_clusters
+        self.max_iter = max_iter
+        self.tol = tol
+        self.random_state = random_state
+        self.kernel = kernel
+        self.gamma = gamma
+        self.degree = degree
+        self.coef0 = coef0
+        self.kernel_params = kernel_params
+        self.verbose = verbose
+        self.init = init
+        self.labels = labels
+        
+    @property
+    def _pairwise(self):
+        return self.kernel == "precomputed"
+
+    def _get_kernel(self, X, Y=None):
+        if callable(self.kernel):
+            params = self.kernel_params or {}
+        else:
+            params = {"gamma": self.gamma,
+                      "degree": self.degree,
+                      "coef0": self.coef0}
+        return pairwise_kernels(X, Y, metric=self.kernel,
+                                filter_params=True, **params)
+
+    def fit(self, X, y=None, sample_weight=None):
+        n_samples = X.shape[0]
+
+        K = self._get_kernel(X)
+
+        sw = sample_weight if sample_weight else np.ones(n_samples)
+        self.sample_weight_ = sw
+
+        if self.init == 'kmeans':
+            km = KMeans(n_clusters=self.n_clusters)
+            self.labels_ = km.fit_predict(X)
+        elif self.init == 'randproj':
+            self.labels_ = energy_multi_random(X, 20)
+        elif self.labels:
+            self.labels_ = self.labels
+        else:
+            rs = check_random_state(self.random_state)
+            self.labels_ = rs.randint(self.n_clusters, size=n_samples)
+
+        dist = np.zeros((n_samples, self.n_clusters))
+        self.within_distances_ = np.zeros(self.n_clusters)
+
+        for it in xrange(self.max_iter):
+            dist.fill(0)
+            self._compute_dist(K, dist, self.within_distances_,
[…]

### 2017-04-05 qcqp_clustering/code/randproj.py

@@ -64,7 +64,7 @@ def pca_projection(X):
     pca.fit(X)
     return pca.transform(X)
 
-def kmeans_multi_random(X, z, n=10):
+def kmeans_multi_random(X, n=10):
     bestJ = np.inf
     for i in range(n):
         Y = rand_projection(X)
@@ -75,7 +75,7 @@ def kmeans_multi_random(X, z, n=10):
             bestJ = J
     return bestz
 
-def energy_multi_random(X, z, n=10):
+def energy_multi_random(X, n=10):
     bestT = 0
     for i in range(n):
         Y = rand_projection(X).flatten()

### 2017-04-05 qcqp_clustering/code/spectral.py

@@ -1,0 +1,69 @@
+from __future__ import division
+
+import numpy as np
+from numpy import linalg as LA
+from sklearn.cluster import KMeans
+
+
+class EnergySpectral:
+
+    def __init__(self, n_clusters, kernel=None):
+        self.k = n_clusters
+
+    def fit_predict(self, X):
+        self.X = X
+        self.n = len(X)
+        self.G = self.gram_matrix()
+        self.U = self.eigen_matrix()
+        self.labels = self.k_means()
+        return self.labels
+
+    def energy_kernel(self, x, y):
+        """Energy stats kernel."""
+        return np.linalg.norm(x) + np.linalg.norm(y) - np.linalg.norm(x-y)
+
+    def gram_matrix(self):
+        """Generate Gram matrix from kernel function."""
+        G = np.zeros((self.n, self.n))
+        for i in range(self.n):
+            for j in range(i+1, self.n):
+                G[i,j] = G[j,i] = self.energy_kernel(self.X[i], self.X[j])
+        return G
+
+    def eigen_matrix(self):
+        """Form a normalized matrix with first k eigenvector of Gram matrix."""
+        eigenvals, eigenvecs = LA.eigh(self.G)
+        # V is a matrix with first k eigenvectors as columns
+        V = eigenvecs.T[:,:k]
+        # U contains the normalized rows of V
+        C = np.diag(1./np.sqrt((V**2).sum(axis=1)))
+        U = C.dot(V)
+        return U
+
+    def k_means(self):
+        """Apply k-means to matrix of eigenvectors."""
+        km = KMeans(n_clusters=self.k)
+        z = km.fit_predict(self.U)
+        return z
+
+
+if __name__ == "__main__":
+    import data
+    from evaluation import accuracy
+    
+    m1 = np.array([0,0])
+    s1 = np.array([[1,0],[0,1]])
+    n1 = 300
+
+    m2 = np.array([4,0])
+    s2 = np.array([[4,1],[1,2]])
+    n2 = 300
+
+    X, z = data.multi_gaussians([m1,m2], [s1,s2], [n1,n2])
+
+    k = 2
+
+    ec = EnergySpectral(n_clusters=k)
+    zh = ec.fit_predict(X)
+    print "Energy Clustering:", accuracy(z, zh)
+

### 2017-04-05 random_projection/gmm.py

@@ -1,108 +1,0 @@
-
-"""
-EM algorithm for GMMM
-
-We implement the Expectation Maximization (EM) algorithm for Gaussian
-Mixture Models (GMM). The algorithm is described as follows:
-
-1. Initialize \mu_k, \Sigma_k, and \pi_k. Evaluate log likelihood function.
-
-2. E Step: evaluate responsabilities using the current values of parameters.
-
-3. M Step: re-estimate \mu_k, \Sigma_k, and \pi_k using the updated
-   responsabilities.
-
-4. Evaluate log likelihood function and check for convergence. 
-   Repeat from step 2 if not converged.
-
-
-"""
-
-from __future__ import division
-
-import numpy as np
-import matplotlib.pyplot as plt
-from scipy.stats import multivariate_normal
-from matplotlib.patches import Ellipse
-
-
-def gaussian(x, mu, sigma):
-    return multivariate_normal.pdf(x, mu, sigma)
-
-
-class GMM:
-    """Fit a GMM using EM algorithm."""
-
-    def __init__(self, data, means, covs, pi, tol=0.001):
-        self.x = data
-        self.mu = means
-        self.sigma = covs
-        self.pi = pi
-        self.N = len(data)
-        self.K = len(pi)
-        self.tol = tol
-        self.gamma = np.zeros((self.N, self.K), dtype=float) # respons.
-    
-    def gauss(self, n, k):
-        """Return multidimensional gaussian at point x_n for component k, 
-        i.e. N(x_n | \mu_k, \sigma_k).
-
-        """
-        return multivariate_normal.pdf(self.x[n], self.mu[k], self.sigma[k])
-
-    def log_likelihood(self):
-        """Compute log likelihood function."""
-        f = 0
-        for n in range(self.N):
-            f += np.array([self.pi[k]*self.gauss(n, k) 
-                                for k in range(self.K)]).sum()
-        return f
-
-    def xxT(self, n, k):
-        x = self.x[n] - self.mu[k]
-        return np.outer(x, x)
-
-    def gmm_em(self):
-        """Implementation of EM algorithm."""
-        converged = False
-        old_loglh = self.log_likelihood()
-
-        while not converged:
-            
-            # compute responsabilities, E-step
-            for n in range(self.N):
-                total = sum([self.pi[j]*self.gauss(n, j) 
-                                        for j in range(self.K)])
-                for k in range(self.K):
-                    self.gamma[n, k] = self.pi[k]*self.gauss(n, k)/total
-            
-            # compute parameters, M-step
[…]

### 2017-04-19 discussion/discussion.tex
new stuff,experiments,notes

@@
 In this document we collect notes about our experiments and theories
-so we can discuss together with the group in an organized way.
+for discussion.
 \end{abstract}
@@
 \section{Experiments using Energy Statistics and One-Dimensional 
-Random Projections (Gui 04-01-2017)}
+Random Projections}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Energy Statistics Based Clustering (Gui 04-05-2017)}
+\section{Energy Statistics Based Clustering}
 
@@
 
+\clearpage
+
+\section{1$D$ Energy Clustering}
+
+Now we do energy clustering in $1D$ ($\mathcal{E}1D$), 
+which in this case we can optimize
+the function exactly. We also compare with the Kernel approach 
+(Ker/$\mathcal{E}$)to energy
+statistics, and also the graph partitioning approach (EClust). One can see that
+these results agree between themselves.
+
+\begin{figure}[h]
+\begin{minipage}{.49\textwidth}
+\includegraphics[scale=.45]{hist_gauss1.pdf}
+\end{minipage}
+\begin{minipage}{.49\textwidth}
+\begin{tabular}{ l l l l l}
+\hline
+Ker/$\mathcal{E}$ & $\mathcal{E} 1D$ & EClust & GMM~~ & $k$-means \\
+\hline
+0.823 &
+0.823 &
+0.823 &
+0.798 &
+0.826 \\
+0.852 & 
+0.852 &
+0.852 &
+0.852 & 
+0.853 \\
+0.842 &
+0.841 &
+0.841 & 
+0.842 &
+0.839 \\
+\hline
+\end{tabular}
+\end{minipage}
+\caption{\label{fig:1dgaus1}
+We have $x \sim \tfrac{1}{2}\left( \mathcal{N}(\mu_1, \Sigma_1) +
+\mathcal{N}(\mu_2, \Sigma_2)\right)$ where $\mu_1 = 0$, $\mu_2=2$,
+$\Sigma_1 = \Sigma_2 = 1$,
+and $500$ points on each cluster. We run the experiment three times,
+using $k$-means++ as initialization in kernel.
+}
+\end{figure}
+
+\begin{figure}[h]
+\begin{minipage}{.49\textwidth}
+\includegraphics[scale=.45]{hist_gauss2.pdf}
+\end{minipage}
+\begin{minipage}{.49\textwidth}
+\begin{tabular}{ l l l l l}
+\hline
+Ker/$\mathcal{E}$ & $\mathcal{E} 1D$ & EClust & GMM~~ & $k$-means \\
+\hline
+0.704 &
+0.704 &
+0.704 &
+0.703 &
+0.704 \\
+0.693 &
+0.686 &
+0.686 &
+0.688 &
+0.686 \\
+0.687 &
+0.688 &
+0.688 &
+0.691 &
+0.691 \\
+\hline
+\end{tabular}
+\end{minipage}
+\caption{\label{fig:1dgaus1}
+We have $x \sim \tfrac{1}{2}\left( \mathcal{N}(\mu_1, \Sigma_1) +
+\mathcal{N}(\mu_2, \Sigma_2)\right)$ where $\mu_1 = 0$, $\mu_2=1$,
+$\Sigma_1 = \Sigma_2 = 1$,
+and $500$ points on each cluster. We run the experiment three times,
+using $k$-means++ as initialization in kernel.
+}
+\end{figure}
+
+\begin{figure}[h]
+\begin{minipage}{.49\textwidth}
+\includegraphics[scale=.45]{hist_loggauss1.pdf}
+\end{minipage}
+\begin{minipage}{.49\textwidth}
+\begin{tabular}{ l l l l l}
+\hline
+Ker/$\mathcal{E}$ & $\mathcal{E} 1D$ & EClust & GMM~~ & $k$-means \\
+\hline
+0.914 &
+0.921 &
+0.921 &
+0.978 &
+0.762 \\
+0.894 &
+0.894 &
+0.894 &
+0.979 &
+0.8 \\
+0.877 &
+0.877 &
+0.877 &
+0.972 &
+0.794 \\
+\hline
+\end{tabular}
+\end{minipage}
+\caption{\label{fig:1dloggaus1}
+We have $x \sim \tfrac{1}{2}\left( \log \mathcal{N}(\mu_1, \Sigma_1) +
+\log \mathcal{N}(\mu_2, \Sigma_2)\right)$ where $\mu_1 = 0$, $\mu_2=1.5$,
+$\Sigma_1 = 0.25$, $ \Sigma_2 = 0.5$,
+and $500$ points on each cluster. We run the experiment three times,
+using $k$-means++ as initialization in kernel.
+}
+\end{figure}
+
+\begin{figure}[h]
+\begin{minipage}{.49\textwidth}
+\includegraphics[scale=.45]{hist1d_gauss2.pdf}
+\end{minipage}
+\begin{minipage}{.49\textwidth}
+\begin{tabular}{ l l l l l}
+\hline
+Ker/$\mathcal{E}$ & $\mathcal{E} 1D$ & GMM~~ & $k$-means \\
+\hline
+0.663&
+0.653&
+0.606&
+0.606\\
+0.726&
+0.723&
+0.793&
+0.71 \\
+0.663&
+0.666&
+0.593&
+0.66 \\
+0.653&
+0.62 &
+0.653&
+0.603\\
+0.676&
+0.63 &
+0.72 & 
+0.62 \\
+\hline
+\end{tabular}
+\end{minipage}
+\caption{\label{fig:1dloggaus1}
+We have $x \sim \tfrac{1}{2}\left( \mathcal{N}(\mu_1, \Sigma_1) +
+\mathcal{N}(\mu_2, \Sigma_2)\right)$ where $\mu_1 = 0$, $\mu_2=2$,
+$\Sigma_1 = 1$, $ \Sigma_2 = 3$, $n_1 = 100$, $n_2=200$.
+For each experiment we run several times and pick the answer with
+optimal objective function.
+}
+\end{figure}
+
+\clearpage
+
+\section{Energy Clustering Higher $D$}
+
+Same as before but in higher dimensions.
+
+\begin{figure}[h]
+\begin{minipage}{.49\textwidth}
+\includegraphics[scale=.45]{gauss_4d1.pdf}
+\end{minipage}
+\begin{minipage}{.49\textwidth}
+\begin{tabular}{ l l l l}
+\hline
+Ker/$\mathcal{E}$ & Spectral/$\mathcal{E}$ & GMM~~ & $k$-means \\
+\hline
+0.7825 &
+0.7175 &
+0.6 &
+0.7675 \\
+0.815 &
+0.77 &
+0.665 &
+0.8175 \\
+0.8325 &
+0.7825 & 
+0.685 & 
+0.8175 \\
+0.835 &
+0.7675 &
+0.8 & 
+0.83 \\
+\hline
+\end{tabular}
+\end{minipage}
+\caption{\label{fig:1dloggaus1}
+We have $x \sim \tfrac{1}{2}\left( \mathcal{N}(\mu_1, \Sigma_1) +
+\mathcal{N}(\mu_2, \Sigma_2)\right)$ where $\mu_1 = (0,0,0,0)^T$, 
+$\mu_2=(1.5, 0,0,0)^T$,
+$\Sigma_1 = 
+\left( \begin{smallmatrix} 
+0.6 & 0.2 & 0.2 & 0.2 \\ 
+0.2 & 0.6 & 0.2 & 0.2 \\ 
+0.2 & 0.2 & 0.6 & 0.2 \\ 
+0.2 & 0.2 & 0.2 & 0.6 \\ 
+\end{smallmatrix} \right)$, 
+$\Sigma_2 = 
+\left( \begin{smallmatrix} 
+0.5 & 0 & 0 & 0 \\ 
+0 & 0.5 & 0 & 0 \\ 
+0 & 0 & 0.5 & 0 \\ 
+0 & 0 & 0 & 0.5 \\ 
+\end{smallmatrix} \right)$,
+$n_1=200$, $n_2=200$.
+We run the experiments few times and pick the answer with best
+objective function.
+}
+\end{figure}
+
+\begin{figure}[h]
+\begin{minipage}{.49\textwidth}
+\includegraphics[scale=.45]{loggauss1.pdf}
+\end{minipage}
+\begin{minipage}{.49\textwidth}
+\begin{tabular}{ l l l l}
+\hline
+Ker/$\mathcal{E}$ & EClust & GMM~~ & $k$-means \\
+\hline
+0.755 &
+0.755 &
+0.71 &
+0.585 \\
+0.74 &
+0.745 &
+0.81 &
+0.575 \\
+0.775 &
+0.775 &
+0.805 &
+0.635 \\
+\hline
+\end{tabular}
+\end{minipage}
+\caption{\label{fig:1dloggaus1}
+We have $x \sim \tfrac{1}{2}\left( \log \mathcal{N}(\mu_1, \Sigma_1) +
+\log \mathcal{N}(\mu_2, \Sigma_2)\right)$ where $\mu_1 = (0,0)^T$, 
+$\mu_2=(1.5, 0)^T$,
+$\Sigma_1 = 
+\left( \begin{smallmatrix} 
+0.25 & 0 \\ 0 & 0.25 
+\end{smallmatrix} \right)$, 
+$\Sigma_2 = 
+\left( \begin{smallmatrix} 
+0.5 & 0 \\ 0 & 0.5 
+\end{smallmatrix} \right)$, 
+and $100$ points on each cluster. We run the experiment three times,
+using $k$-means++ as initialization in kernel. EClust is extremelly slow.
+}
+\end{figure}
+
+\begin{figure}[h]
+\begin{minipage}{.49\textwidth}
+\includegraphics[scale=.45]{loggauss2.pdf}
+\end{minipage}
+\begin{minipage}{.49\textwidth}
+\begin{tabular}{ l l l l}
+\hline
+Ker/$\mathcal{E}$ & EClust & GMM~~ & $k$-means \\
+\hline
+0.895 &
+0.895 &
+0.95 &
+0.715 \\
+0.91 &
+0.91 &
+0.955 &
+0.8 \\
+0.815 &
+0.815 &
+0.93 &
+0.69 \\
+\hline
+\end{tabular}
+\end{minipage}
+\caption{\label{fig:1dloggaus1}
+We have $x \sim \tfrac{1}{2}\left( \log \mathcal{N}(\mu_1, \Sigma_1) +
+\log \mathcal{N}(\mu_2, \Sigma_2)\right)$ where $\mu_1 = (0,0,0,0)^T$, 
+$\mu_2=(1.5, 1.5,0,0)^T$,
+$\Sigma_1 = 
+\left( \begin{smallmatrix} 
+0.25 & 0 & 0 & 0 \\ 
+0 & 0.25 & 0 & 0 \\ 
+0 & 0 & 0.25 & 0 \\ 
+0 & 0 & 0 & 0.25 \\ 
+\end{smallmatrix} \right)$, 
+$\Sigma_2 = 
+\left( \begin{smallmatrix} 
+0.5 & 0 & 0 & 0 \\ 
+0 & 0.5 & 0 & 0 \\ 
+0 & 0 & 0.5 & 0 \\ 
+0 & 0 & 0 & 0.5 \\ 
+\end{smallmatrix} \right)$, 
+and $100$ points on each cluster. We run the experiment three times,
+using $k$-means++ as initialization in kernel. EClust is extremelly slow.
+}
+\end{figure}
+
+\begin{figure}[h]
+\begin{minipage}{.49\textwidth}
+\includegraphics[scale=.45]{circles.pdf}
+\end{minipage}
+\begin{minipage}{.49\textwidth}
+\begin{tabular}{ l l l l}
+\hline
+Ker/$\mathcal{E}$ & EClust & GMM~~ & $k$-means \\
+\hline
+0.74 &
+0.516 &
+0.516 &
+0.536 \\
+0.583 &
+0.513 &
+0.52 & 
+0.556 \\
+0.706 &
+0.503 &
+0.516 &
+0.503 \\
+0.633 &
+0.5 &
+0.553 &
+0.52 \\
+\hline
+\end{tabular}
+\end{minipage}
+\caption{\label{fig:circles}
+We generate two circles with random noise.
+First circle has radius $1$ and a gaussian noise of size $0.2$.
+Second circle has radius $.2$ and a gaussian noise of size $0.2$.
+We have 200 points on each class.
+}
+\end{figure}
+
 \end{document}

### 2017-04-19 qcqp_clustering/code/eclust/data.py

@@ -1,0 +1,228 @@
+"""Functions used to generate or manipulate data."""
+
+# Guilherme Franca <guifranca@gmail.com>
+# Johns Hopkins Unversity, Neurodata
+
+
+from __future__ import division
+
+import numpy as np
+import matplotlib.pyplot as plt
+import numbers
+
+from sklearn.decomposition import PCA
+
+
+def multivariate_normal(means, sigmas, ns):
+    """Generate several normal distributed data points.
+    
+    Parameters: 
+        means = [m1, m2, ...]
+        sigmas = [s1, s2, ...]
+        ns = [n1, n2, ...]
+
+        Each parameter in these lists are associated to one multivariate
+        normal distribution. The data is shuffled so points are not ordered.
+    
+    Returns: data X and labels z.
+    
+    """
+    n = len(means)
+    X = []
+    z = []
+    k = 0
+    for mu, sigma, n in zip(means, sigmas, ns):
+        X.append(np.random.multivariate_normal(mu, sigma, n))
+        z.append([k]*n)
+        k += 1
+    X = np.concatenate(X)
+    z = np.concatenate(z)
+    idx = np.random.permutation(sum(ns))
+    return X[idx], z[idx]
+
+def univariate_normal(means, sigmas, ns):
+    """Generate several univariate normal distributed data points.
+    
+    Parameters: 
+        means = [m1, m2, ...]
+        sigmas = [s1, s2, ...]
+        ns = [n1, n2, ...]
+    
+    Returns: data X and labels z.
+    
+    """
+    n = len(means)
+    X = []
+    z = []
+    k = 0
+    for mu, sigma, n in zip(means, sigmas, ns):
+        X.append(np.random.normal(mu, sigma, n))
+        z.append([k]*n)
+        k += 1
+    X = np.concatenate(X)
+    z = np.concatenate(z)
+    idx = np.random.permutation(sum(ns))
+    return X[idx], z[idx]
+
+def univariate_lognormal(means, sigmas, ns):
+    """Return data according to log normal distribtion.
+    Input: means = [mu1, mu2, ...], 
+           sigmas = [sigma1, sigma2, ...], 
+           ns = [n1, n2, ...]
+    Ouput: Data matrix X with corresponding labels z
+    """
+    x, z = univariate_normal(means, sigmas, ns)
+    return np.exp(x), z
+    
[…]

### 2017-04-19 qcqp_clustering/code/eclust/kkmeans.py

@@ -1,3 +1,10 @@
+"""Kernel k-means with different initializations."""
+
+# Guilherme Franca <guifranca@gmail.com>
+# Johns Hopkins University, Neurodata
+
+# code updated from Mathieu Blondel
+# License: BSD 3 clause
 
 import numpy as np
 
@@ -6,14 +13,15 @@ from sklearn.metrics.pairwise import pairwise_kernels
 from sklearn.utils import check_random_state
 from sklearn.cluster import KMeans
 
-from randproj import energy_multi_random
+import kmeanspp
 
 
 class KernelKMeans(BaseEstimator, ClusterMixin):
+    """Kernel k-means with Energy Statistics."""
 
     def __init__(self, n_clusters=3, max_iter=50, tol=1e-3, random_state=None,
                  kernel="linear", gamma=None, degree=3, coef0=1,
-                 kernel_params=None, verbose=0, init='kmeans', labels=None):
+                 kernel_params=None, verbose=0, init='kmeans++', labels=None):
         self.n_clusters = n_clusters
         self.max_iter = max_iter
         self.tol = tol
@@ -25,7 +33,7 @@ class KernelKMeans(BaseEstimator, ClusterMixin):
         self.kernel_params = kernel_params
         self.verbose = verbose
         self.init = init
-        self.labels = labels
+        self.labels_ = labels
         
     @property
     def _pairwise(self):
@@ -38,24 +46,27 @@ class KernelKMeans(BaseEstimator, ClusterMixin):
             params = {"gamma": self.gamma,
                       "degree": self.degree,
                       "coef0": self.coef0}
-        return pairwise_kernels(X, Y, metric=self.kernel,
+        n_samples = X.shape[0]
+        #return np.array([self.kernel(x, y) 
+        #        for x in X for y in X]).reshape((n_samples, n_samples))
+        K = pairwise_kernels(X, Y, metric=self.kernel,
                                 filter_params=True, **params)
+        self.kernel_matrix_ = K
+        return K
 
     def fit(self, X, y=None, sample_weight=None):
         n_samples = X.shape[0]
 
         K = self._get_kernel(X)
-
+        
         sw = sample_weight if sample_weight else np.ones(n_samples)
         self.sample_weight_ = sw
 
-        if self.init == 'kmeans':
-            km = KMeans(n_clusters=self.n_clusters)
-            self.labels_ = km.fit_predict(X)
-        elif self.init == 'randproj':
[…]

### 2017-04-19 qcqp_clustering/code/eclust/randproj.py

@@ -1,0 +1,123 @@
+"""We test clustering using 1D random projections. Only 2 clusters."""
+
+# Guilherme Franca <guifranca@gmail.com>
+# Johns Hopkins University, Neurodata
+
+
+from __future__ import division
+
+import numpy as np
+import matplotlib.pyplot as plt
+from sklearn import random_projection
+from sklearn.cluster import KMeans
+from sklearn import decomposition
+from sklearn.metrics import silhouette_score
+
+from metric import accuracy
+import energy1d
+import data
+
+
+def kmeans(X):
+    kmeans = KMeans(n_clusters=2).fit(X)
+    z = kmeans.labels_
+    score = -kmeans.score(X)
+    return z, score
+
+def rand_proj(X):
+    transf = random_projection.GaussianRandomProjection(n_components=1)
+    return transf.fit_transform(X)
+
+def pca_proj(X):
+    pca = decomposition.PCA(n_components=1)
+    pca.fit(X)
+    return pca.transform(X)
+
+def kmeans_random(X, n=50):
+    best_score = np.inf
+    for i in range(n):
+        Y = rand_proj(X)
+        zh, score = kmeans(Y)
+        if score < best_score:
+            best_z = zh
+            best_score = score
+    return best_z
+
+def energy_random(X, n=50):
+    best_score = np.inf
+    for i in range(n):
+        Y = rand_proj(X).flatten()
+        zh, score = energy1d.two_clusters1D(Y)
+        if score < best_score:
+            best_z = zh
+            best_score = score
+    return best_z
+    
+def test(X):
+    zh, score = kmeans(X)
+    a = accuracy(z, zh)
+    print "k-means original space:", a
+
+    Y = pca_proj(X)
+    zh, score = kmeans(Y)
+    a = accuracy(z, zh)
+    print "k-means/1D PCA:", a
+
+    zh = kmeans_random(X, 20)
+    a = accuracy(z, zh)
+    print "k-means/random projection 1D:", a
+
+    zh = energy_random(X, 20)
+    a = accuracy(z, zh)
+    print "energy/random projection 1D", a
+
+
+###############################################################################
+if __name__ == '__main__':
+
+    m1 = np.array([0,0])
[…]

### 2017-04-19 qcqp_clustering/code/eclust/withingraph.py

@@ -1,0 +1,121 @@
+"""Optimize Energy Clustering through iterative procedure. This idea
+is the same as graph partitioning. This algorithm is extremelly slow.
+
+"""
+
+# Guilherme Franca <guifranca@gmail.com>
+# Johns Hopkins University, Neurodata
+
+from __future__ import division
+
+import numpy as np
+from sklearn.cluster import KMeans
+import numbers
+
+from energy import energy
+import kmeanspp
+
+
+class EClust:
+
+    def __init__(self, n_clusters, max_iter=50, init='kmeans++',
+                    labels=None):
+        self.n_clusters = n_clusters
+        self.max_iter= max_iter
+        self.init = init
+        self.labels_ = labels
+
+    def initial_labels(self):
+        if self.labels_ is not None:
+            pass
+        elif self.init == 'kmeans++':
+            if isinstance(self.X[0], numbers.Number):
+                Y = np.array([[x] for x in self.X])
+            else:
+                Y = self.X
+            self.labels_ = kmeanspp.kpp(self.n_clusters, Y)
+        elif self.init == 'random':
+            self.labels_ = np.random.choice(range(self.n_clusters), len(self.X))
+        else:
+            self.labels_ = np.random.choice(range(self.n_clusters), len(X))
+
+    def fit(self, X):
+        """Fit data X against our algorithm. It will set self.labels_"""
+        # main algorithm
+        # go through each point and compute the cost with its current
+        # partition and minimal cost with other partitions
+        # move label in case there is a smaller one
+        # repeat until convergence
+        self.X = X
+        self.initial_labels()
+        count = 0
+        converged = False
+        while not converged and count < self.max_iter:
+            converged = True 
+            for i, k in enumerate(self.labels_):
+                point = self.X[i]
+                curr_cost = self.cost_current_partition(point, k)
+                min_cost, j = self.min_cost_other_partitions(point, k)
+                if not (curr_cost < min_cost):
+                    self.labels_[i] = j
+                    converged = False
+            count += 1
+        self.count = count
+        if count >= self.max_iter:
+            print "Warning: didn't converge in %i iterations." % self.max_iter
+
+    def fit_predict(self, X):
+        """Fit and return predicted labels."""
+        self.fit(X)
+        return self.labels_
+
+    def cost_current_partition(self, point, label):
+        """Compute the cost of point with its current partition."""
[…]

### 2017-04-19 qcqp_clustering/code/spectral.py

@@ -1,69 +1,0 @@
-from __future__ import division
-
-import numpy as np
-from numpy import linalg as LA
-from sklearn.cluster import KMeans
-
-
-class EnergySpectral:
-
-    def __init__(self, n_clusters, kernel=None):
-        self.k = n_clusters
-
-    def fit_predict(self, X):
-        self.X = X
-        self.n = len(X)
-        self.G = self.gram_matrix()
-        self.U = self.eigen_matrix()
-        self.labels = self.k_means()
-        return self.labels
-
-    def energy_kernel(self, x, y):
-        """Energy stats kernel."""
-        return np.linalg.norm(x) + np.linalg.norm(y) - np.linalg.norm(x-y)
-
-    def gram_matrix(self):
-        """Generate Gram matrix from kernel function."""
-        G = np.zeros((self.n, self.n))
-        for i in range(self.n):
-            for j in range(i+1, self.n):
-                G[i,j] = G[j,i] = self.energy_kernel(self.X[i], self.X[j])
-        return G
-
-    def eigen_matrix(self):
-        """Form a normalized matrix with first k eigenvector of Gram matrix."""
-        eigenvals, eigenvecs = LA.eigh(self.G)
-        # V is a matrix with first k eigenvectors as columns
-        V = eigenvecs.T[:,:k]
-        # U contains the normalized rows of V
-        C = np.diag(1./np.sqrt((V**2).sum(axis=1)))
-        U = C.dot(V)
-        return U
-
-    def k_means(self):
-        """Apply k-means to matrix of eigenvectors."""
-        km = KMeans(n_clusters=self.k)
-        z = km.fit_predict(self.U)
-        return z
-
-
-if __name__ == "__main__":
-    import data
-    from evaluation import accuracy
-    
-    m1 = np.array([0,0])
-    s1 = np.array([[1,0],[0,1]])
-    n1 = 300
-
-    m2 = np.array([4,0])
-    s2 = np.array([[4,1],[1,2]])
-    n2 = 300
-
-    X, z = data.multi_gaussians([m1,m2], [s1,s2], [n1,n2])
-
-    k = 2
-
-    ec = EnergySpectral(n_clusters=k)
-    zh = ec.fit_predict(X)
-    print "Energy Clustering:", accuracy(z, zh)
-

### 2017-04-19 qcqp_clustering/code/tests_nd.py

@@ -1,0 +1,106 @@
+"""
+Several tests and comparison for clustering.
+
+"""
+
+import numpy as np
+import matplotlib.pyplot as plt
+from sklearn.cluster import KMeans
+from sklearn.cluster import SpectralClustering
+from sklearn.mixture import GMM
+
+from eclust.kkmeans import KernelKMeans
+from eclust.metric import accuracy
+import eclust.data as data
+from eclust.withingraph import EClust
+from eclust.energy1d import two_clusters1D
+from eclust.energy import energy_kernel
+from eclust.objectives import kernel_score
+
+
+def kernel_energy(k, X, alpha=1, cutoff=0, num_times=10):
+    best_score = 0
+    for i in range(num_times):
+        km = KernelKMeans(n_clusters=2, max_iter=300, kernel=energy_kernel,
+                      kernel_params={'alpha':alpha, 'cutoff':cutoff})
+        zh = km.fit_predict(X)
+        score = kernel_score(km.kernel_matrix_, zh)
+        if score > best_score:
+            best_score = score
+            best_z = zh
+    return best_z
+
+def spectral_energy(k, X, alpha=1, cutoff=0, n_neighbors=10, num_times=10):
+    best_score = 0
+    for i in range(num_times):
+        sc = SpectralClustering(n_clusters=k, affinity=energy_kernel,
+                        assign_labels='kmeans', n_init=20, 
+                        kernel_params={'alpha':1, 'cutoff':0},
+                        n_neighbors=n_neighbors)
+        sc.fit(X)
+        zh = sc.labels_
+        score = kernel_score(sc.affinity_matrix_, zh)
+        if score > best_score:
+            best_z = zh
+            best_score = score
+    return best_z
+
+
+###############################################################################
+if __name__ == '__main__':
+    
+    m1 = np.array([0, 0, 0, 0])
+    s1 = np.array([
+        [0.6, .2, .2, .2], 
+        [.2, 0.6, 0, .2],
+        [.2, 0, 0.6, .2],
+        [.2, .2, .2, 0.6],
+    ])
+    #s1 = np.eye(4)
+    n1 = 300
+    
+    m2 = np.array([1.5, 0, 0, 0])
+    s2 = np.array([
+        [0.5, 0, 0, 0], 
[…]

### 2017-05-04 qcqp_clustering/draft/energy_clustering.tex
draft

@@
-\documentclass[aps,preprint,nofootinbib,floatfix]{revtex4-1}
+%\documentclass[aps,preprint,nofootinbib,floatfix]{revtex4-1}
+\documentclass{article}
+
+%\usepackage{nips_2017}
+
+% to compile a camera-ready version, add the [final] option, e.g.:
+\usepackage[final]{nips_2017}
+
+\usepackage{nicefrac}       % compact symbols for 1/2, etc.
+\usepackage{microtype}      % microtypography
+\usepackage{amsmath,amssymb,amsfonts,amsthm,amscd,bm}
+\usepackage{algorithm}
+\usepackage{algorithmic}
+\usepackage[inline]{enumitem}
+\usepackage[pdftex]{graphicx}
+\graphicspath{{./figs/}}
 
 %\usepackage{cite}
-\usepackage[pdftex]{graphicx}
-\usepackage{amsmath,amssymb,amsfonts,amsthm,amscd,bm}
 \usepackage{bbm}
@@
 
+\title{Nonparametric Clustering Based on Energy Distance}
+
+\author{
+Guilherme Fran\c ca \\
+Johns Hopkins University\\
+\texttt{guifranca@gmail.com} \\
+%% examples of more authors
+\And
+Joshua Vogelstein \\
+Johns Hopkins University\\
+\texttt{jovo@jhu.edu}
+}
+
+%\author{Guilherme Fran\c ca}
+%\email{guifranca@gmail.com}
+%\affiliation{Johns Hopkins University, Center for Imaging Science}
+
+%\author{Joshua Vogelstein}
+%\email{jovo@jhu.edu}
+%\affiliation{Johns Hopkins University, Center for Imaging Science}
+
 \begin{document}
 
-\title{Nonparametric Clustering Based on Energy Distance}
-
-\author{Guilherme Fran\c ca}
-\email{guifranca@gmail.com}
-\affiliation{Johns Hopkins University, Center for Imaging Science}
-
-\author{Joshua Vogelstein}
-\email{jovo@jhu.edu}
-\affiliation{Johns Hopkins University, Center for Imaging Science}
+\maketitle
 
@@
 \end{abstract}
-
-\maketitle
 
@@
 In this section we briefly review the main concepts from energy
-statistics and reproducing kernel Hilbert spaces (RKHS) used in this paper. 
+statistics and reproducing kernel Hilbert spaces (RKHS) used through paper. 
 For more details we refer the reader
-to \cite{Szkely2013} (and references therein) and also \cite{Sejdinovic2013}.
+to \cite{Szkely2013} 
+(and references therein) and also \cite{Sejdinovic2013}.
 
@@
 between energy distance, commonly used in statistics,
-and distances between embeddings of distributions in reproducible
+and distances between embeddings of distributions in reproducing
 kernel Hilbert spaces (RKHS),
@@
 
-\section{Clustering using Energy Distance}
+\section{Energy Distance based Clustering}
+
+Now we formulate an optimization problem for clustering based on 
+energy statistics
+and RKHS.
+Assume we have data $\mathcal{X} = \{ x_1,x_2,\dotsc, x_N \}$,
+and a partition $\mathcal{X} = \cup_{k=1}^K \mathcal{A}_k$, where
+$\mathcal{A}_k \cap \mathcal{A}_{k'} = \emptyset$.
+Each expectation in \eqref{eq:energy2} can be computed 
+through the function
+\begin{equation}
+\label{eq:g_def}
+g_\alpha(\mathcal{A}_k, \mathcal{A}_{k'}) \equiv 
+\dfrac{1}{N_k N_{k'}}\sum_{x \in \mathcal{A}_k} 
+\sum_{y \in \mathcal{A}_{k'}} \| x - y\|^\alpha
+\end{equation}
+where $N_k = |\mathcal{A}_k|$ is the number of elements in partition
+$\mathcal{A}_k$. In energy statistics 
+\cite{Szkely2013} we have the within energy dispersion
+\begin{equation}
+\mathcal{W}_\alpha = 
+\sum_{k=1}^{K} \dfrac{N_k}{2} g_\alpha(\mathcal{A}_k, \mathcal{A}_k),
+\end{equation}
+and also the between-sample energy statistic
+\begin{equation}
+\mathcal{S}_\alpha = 
+\sum_{1 \le  k < k' \le K } \dfrac{N_k N_{k'}}{2 N} \left[
+2 g_{\alpha}(\mathcal{A}_k, \mathcal{A}_{k'}) - 
+g_{\alpha}(\mathcal{A}_k, \mathcal{A}_{k}) - 
+g_{\alpha}(\mathcal{A}_{k'}, \mathcal{A}_{k'})
+\right].
+\end{equation}
+This is a test statistic for equality of distributions, which is small
+if all datapoints comes from the same distribution, and diverges otherwise,
+in the limit $N \to \infty$. Therefore, our criteria for 
+clustering data is to 
+maximize $\mathcal{S}_\alpha$.
+It can be shown that the total dispersion of the data obeys \cite{Szkely2013}
+\begin{equation}
+\mathcal{T}_\alpha(\mathcal{X}) 
+= \mathcal{W}_\alpha + \mathcal{S}_\alpha = \dfrac{N}{2}
+g_{\alpha}(\mathcal{X}, \mathcal{X}). 
+\end{equation}
+Note that $\mathcal{T}$ only depends on the pooled data, so it 
+does not depend on how we partition $\mathcal{X}$. Therefore, maximizing
+$\mathcal{S}_\alpha$ is equivalent to minimizing $\mathcal{W}_\alpha$, which
+has a simpler form. Thus, our clustering problem corresponds to
+find the best partition of $\mathcal{X}$ such that
+\begin{equation}\label{eq:minimize}
+\min_{\{ \mathcal{A}_k \} } \mathcal{W}_\alpha(\{ 
+\mathcal{A}_1, \dotsc, \mathcal{A}_{K}
+\})
+\end{equation}
+where each datapoint belongs to one and only one partition (hard assignments).
+
+Based on the equivalence between semimetrics of negative type
+and kernel functions \eqref{eq:gen_kernel}, and choosing $x_0 = 0$ for
+simplicity (one might choose any other point in $\mathbb{R}^D$),
+in the case of \eqref{eq:g_def} we have the associated kernel
+\begin{equation}\label{eq:kernel}
+k_\alpha(x,y) = \tfrac{1}{2} \left( 
+\| x \|^\alpha + \| y \|^\alpha - \| x-y \|^\alpha 
+\right) .
+\end{equation}
+Now we can write
+\begin{equation}\label{eq:W2}
+\mathcal{W}_\alpha(\{ \mathcal{A}_1,\dotsc,\mathcal{A}_K \})
+= - \sum_{k=1}^{K} \dfrac{N_k}{2} \, \mathbb{E} \, 
+k_\alpha(\mathcal{A}_k, \mathcal{A}_k) = 
+- \sum_{k=1}^{K} \dfrac{1}{2 N_k} \sum_{x, x' \in \mathcal{A}_k} 
+k_\alpha(x, x').
+\end{equation}
+
+Now let $Z \in \{ 0,1 \}^{N\times K}$ be a binary matrix such
+that 
+\begin{equation}
+Z_{nk} = \begin{cases}
+1 & \mbox{if $x_n \in \mathcal{A}_k$ } \\
+0 & \mbox{otherwise.}
+\end{cases}
+\end{equation}
+Notice that $D = Z^T Z = \diag( N_1, N_2, \dotsc, N_K )$ contains the number
+of elements in each partition. Introducing the kernel matrix
+$K^{\alpha} \in \mathbb{R}^{N\times N}$ such that
+\begin{equation}
+K^{\alpha}_{ij} = k_\alpha(x_i, x_j),
+\end{equation}
+we can write \eqref{eq:W2} as
+$- \tfrac{1}{2} \Tr D^{-1} Z^\top K^\alpha Z $. Thus our optimization problem
+\eqref{eq:minimize} can be written as
+\begin{equation}\label{eq:qcqp}
+\begin{aligned}
+& \max_{Z} \Tr\left\{ \big( Z D^{-1/2}\big)^\top K^{\alpha} 
+\big( ZD^{-1/2} \big) 
+\right\} \\
+&\mbox{s.t. $Z_{ij} \in \{0,1\}$, $\sum_{k=1}^K Z_{nk} = 1$, 
+$\sum_{n=1}^N Z_{nk} = N_k$, and $D = Z^\top Z$}.
+\end{aligned}
+\end{equation}
+This is a quadratic problem with integer constraints, which is usually
+NP-hard. This problem has the same formulation as Kernel $K$-means, but
+with our non-parametric kernel function given by \eqref{eq:kernel}.
 

### 2017-05-04 qcqp_clustering/code/tests2.py

@@ -1,0 +1,124 @@
+"""
+Several tests and comparison for clustering.
+
+"""
+
+import numpy as np
+import matplotlib.pyplot as plt
+from sklearn.cluster import KMeans
+from sklearn.cluster import SpectralClustering
+from sklearn.mixture import GMM
+
+from eclust.kkmeans import KernelKMeans
+from eclust.metric import accuracy
+import eclust.data as data
+from eclust.withingraph import EClust
+from eclust.energy1d import two_clusters1D
+from eclust.energy import energy_kernel
+from eclust.objectives import kernel_score
+
+
+def kernel_energy(k, X, alpha=.5, cutoff=0, num_times=8):
+    best_score = 0
+    for i in range(num_times):
+        km = KernelKMeans(n_clusters=2, max_iter=300, kernel=energy_kernel,
+                      kernel_params={'alpha':alpha, 'cutoff':cutoff})
+        zh = km.fit_predict(X)
+        score = kernel_score(km.kernel_matrix_, zh)
+        if score > best_score:
+            best_score = score
+            best_z = zh
+    return best_z
+
+def spectral_energy(k, X, alpha=.5, cutoff=0, n_neighbors=12, num_times=8):
+    best_score = 0
+    for i in range(num_times):
+        sc = SpectralClustering(n_clusters=k, affinity=energy_kernel,
+                        assign_labels='kmeans', n_init=20, 
+                        kernel_params={'alpha':1, 'cutoff':0},
+                        n_neighbors=n_neighbors)
+        sc.fit(X)
+        zh = sc.labels_
+        score = kernel_score(sc.affinity_matrix_, zh)
+        if score > best_score:
+            best_z = zh
+            best_score = score
+    return best_z
+
+def kmeans(k, X):
+    km = KMeans(n_clusters=k)
+    zh = km.fit_predict(X)
+    return zh
+
+def gmm(k, X):
+    gmm = GMM(n_components=k)
+    gmm.fit(X)
+    zh = gmm.predict(X)
+    return zh
+
+def cluster_many(k, X, z, clustering_function, num_times):
+    accuracies = []
+    for i in range(num_times):
+        zh = clustering_function(k, X)
+        a = accuracy(z, zh)
+        accuracies.append(a)
+    return np.array(accuracies)
+
+def compare_algos(k, X, z, num_times=10):
[…]

### 2017-05-16 qcqp_clustering/draft/energy_clustering.tex
draft

--- 2017-04-05 qcqp_clustering/draft/energy_clustering.tex
@@
-\documentclass[aps,preprint,nofootinbib,floatfix]{revtex4-1}
+%\documentclass[aps,preprint,nofootinbib,floatfix]{revtex4-1}
+\documentclass{article}
+
+%\usepackage{nips_2017}
+
+% to compile a camera-ready version, add the [final] option, e.g.:
+\usepackage[final]{nips_2017}
+
+\usepackage{nicefrac}       % compact symbols for 1/2, etc.
+\usepackage{microtype}      % microtypography
+\usepackage{amsmath,amssymb,amsfonts,amsthm,amscd,bm,bbm}
+\usepackage{algorithm}
+\usepackage{algorithmic}
+\usepackage[inline]{enumitem}
+\usepackage[pdftex]{graphicx}
+\graphicspath{{./figs/}}
 
 %\usepackage{cite}
-\usepackage[pdftex]{graphicx}
-\usepackage{amsmath,amssymb,amsfonts,amsthm,amscd,bm}
 \usepackage{bbm}
@@
 
-\def\Energy{\mathcal{E}}
-\def\E{\mathbb{E}}
+\newcommand\Energy{\mathcal{E}}
+\newcommand\E{\mathbb{E}}
+\newcommand\kk{K}
+\newcommand\kkk{h}
+\newcommand\Hk{{\mathcal{H}}_{\kk}}
+\newcommand\HH{\mathcal{H}}
+\newcommand\C{{\mathcal{C}}}
+\newcommand\Zt{\widetilde{Z}}
+
+\title{Nonparametric Clustering Based on Energy Statistics}
+
+\author{
+Guilherme Fran\c ca \\
+Johns Hopkins University\\
+\texttt{guifranca@gmail.com} \\
+%% examples of more authors
+\And
+Joshua T. Vogelstein \\
+Johns Hopkins University\\
+\texttt{jovo@jhu.edu}
+}
+
+%\author{Guilherme Fran\c ca}
+%\email{guifranca@gmail.com}
+%\affiliation{Johns Hopkins University, Center for Imaging Science}
+
+%\author{Joshua Vogelstein}
+%\email{jovo@jhu.edu}
+%\affiliation{Johns Hopkins University, Center for Imaging Science}
 
@@
 
-\title{Nonparametric Clustering Based on Energy Distance}
-
-\author{Guilherme Fran\c ca}
-\email{guifranca@gmail.com}
-\affiliation{Johns Hopkins University, Center for Imaging Science}
-
-\author{Joshua Vogelstein}
-\email{jovo@jhu.edu}
-\affiliation{Johns Hopkins University, Center for Imaging Science}
+\maketitle
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Energy Statistics and RKHS}
+\label{sec:background}
 
 In this section we briefly review the main concepts from energy
-statistics and reproducing kernel Hilbert spaces (RKHS) used in this paper. 
+statistics and its relation to reproducing kernel Hilbert spaces 
+(RKHS), which form the basis of our work.
 For more details we refer the reader
@@
 \end{equation}
-called energy distance, 
+called \emph{energy distance}, 
 is rotational invariant and nonnegative, $\Energy(P,Q) \ge 0$, where
@@
 
-The energy distance can be generalized as 
+The energy distance can be generalized as, for instance,
 \begin{equation}\label{eq:energy2}
 \Energy_\alpha(P, Q) = 2 \E \| X - Y\|^{\alpha} - \| X - X' \|^{\alpha} - 
-\| Y - Y' \|^{\alpha}
+\| Y - Y' \|^{\alpha},
 \end{equation}
@@
 equality to zero only requires
-equality of the means and thus it does not imply equality of distributions.
+equality of the means, and thus it does not imply equality of distributions.
+
 It is important to 
-mention that \eqref{eq:energy2} can be further generalized by
-replacing
+mention that \eqref{eq:energy2} can be even further generalized.
+Let $X, Y \in \mathcal{X}$ and replace the Euclidean norm by
+$\rho: \mathcal{X}\times\mathcal{X} \to \mathbb{R}^+$, i.e.
 $\| X - Y\| \to \rho(X,Y)$, where $\rho$ is a so-called semimetric
-of negative type \cite{Sejdinovic2013}. 
-In this case there is a Hilbert space $\mathcal{H}$ and
-a map $\varphi: \mathbb{R}^D \to
+of negative type \cite{Sejdinovic2013}, which satisfy
+\begin{equation}
+\label{eq:negative_type}
+\sum_{i,j=1}^n \alpha_i \alpha_j \rho(X_i, X_j) \le 0,
+\end{equation}
+where $X_i \in \mathcal{X}$, and $\alpha_i \in \mathbb{R}$ such that
+$\sum_{i=1}^n \alpha_i = 0$.
+In this case, there is a Hilbert space $\mathcal{H}$ and
+a map $\varphi: \mathcal{X} \to
 \mathcal{H}$ such that
@@
 Even though the semimetric 
-$\rho$ may not satisfy the triangle inequality, its square
-root $\rho^{1/2}$ does, since it is an actual metric.
+$\rho$ may not satisfy the triangle inequality, 
+$\rho^{1/2}$ does since it can be shown to be a legit metric.
 
-There is an equivalence 
-between energy distance, commonly used in statistics,
-and distances between embeddings of distributions in reproducible
-kernel Hilbert spaces (RKHS),
-commonly used in machine learning. This equivalence was established
-in \cite{Sejdinovic2013}. We recall the definition of
-RKHS. Let $\mathcal{H}$ be a Hilbert space over real-valued functions
-over $\mathbb{R}^D$. A function $k : \mathbb{R}^D \times \mathbb{R}^D \to 
-\mathbb{R}$ is a reproducing kernel of $\mathcal{H}$ if it satisfies
+There is an equivalence between energy distance, 
+commonly used in statistics,
+and distances between embeddings of distributions in 
+RKHS, commonly used in machine learning. 
+This equivalence was established
+in \cite{Sejdinovic2013}. Let us first recall the definition of
+RKHS. Let $\HH$ be a Hilbert space of real-valued functions
+over $\mathcal{X}$. A function 
+$\kk : \mathcal{X} \times \mathcal{X} \to 
+\mathbb{R}$ is a reproducing kernel of $\HH$ if it satisfies
 the following two conditions:
 \begin{enumerate}
-\item $K_x \equiv k(\cdot, x) \in \mathcal{H}$ for any $x \in \mathbb{R}^D$.
-\item $\langle K_x, f \rangle_{\mathcal{H}} = f(x)$ for
-any $x\in\mathbb{R}^D$ and $f\in \mathcal{H}$.
+\item $\kkk_x \equiv \kk(\cdot, x) \in \HH$ 
+for any $x \in \mathcal{X}$.
+\item $\langle \kkk_x, f \rangle_{\HH} = f(x)$ for
+any $x\in\mathcal{X}$ and $f\in \HH$.
 \end{enumerate}
-In other words, for any $x \in \mathbb{R}^D$ there is a unique function
-$K_x \in \mathcal{H}$ that reproduces $f(x)$ through the inner product
-of $\mathcal{H}$.
-If such a kernel function $k$ exists, then $\mathcal{H}$ is called a RKHS.
-From this we have $\langle K_x, K_y \rangle = K_y(x) = k(x,y)$. This implies
-that $k(x,y)$ is symmetric and positive definite, $\sum_{i,j=1}^n c_i c_j
-k(x_i,x_j) \ge 0$ for $c_i,c_j \in \mathbb{R}$.
+In other words, for any $x \in \mathcal{X}$ there is a unique function
+$\kkk_x \in \HH$ that reproduces $f(x)$ through the inner product
+of $\HH$.
+If such a \emph{kernel} 
+function $\kk$ exists, then $\HH$ is called a RKHS.
+From this we have $\langle \kkk_x, \kkk_y \rangle = \kkk_y(x) = \kk(x,y)$. 
+This implies
+that $\kk(x,y)$ is symmetric and positive definite, 
+$\sum_{i,j=1}^n c_i c_j
+\kk(x_i,x_j) \ge 0$ for $c_i,c_j \in \mathbb{R}$.
 
-The Moore-Aronszajn theorem establishes the converse. For every symmetric
-and positive definite function $k: \mathbb{R}^D\times \mathbb{R}^D \to
-\mathbb{R}$, there is an associated RKHS, $\mathcal{H}_k$, with reproducing
-kernel $k$. The map $\varphi: x \mapsto K_x \in \mathcal{H}_k$ is called
-the canonical feature map. Given a kernel $k$,
+The Moore-Aronszajn theorem establishes the converse \cite{Aronszajn}.
+For every symmetric
+and positive definite function $\kk: \mathcal{X}\times \mathcal{X} \to
+\mathbb{R}$, there is an associated RKHS, $\Hk$, 
+with reproducing
+kernel $\kk$. The map $\varphi: x \mapsto \kkk_x \in \Hk$ is called
+the canonical feature map. Given a kernel $\kk$,
 this theorem enables us to define an embedding of a probability measure
-$P$ into the RKHS: $P \mapsto K_P \in
-\mathcal{H}_k$ such that 
-$\int f(x) d P(x) = \langle f, K_P \rangle$ for all $f \in \mathcal{H}_k$,
-or alternatively $K_P = \int k(\cdot, x)  d P(x)$. 
+$P$ into the RKHS: $P \mapsto \kkk_P \in
+\Hk$ such that 
+$\int f(x) d P(x) = \langle f, \kkk_P \rangle$ for all $f \in \Hk$,
+or alternatively $\kkk_P = \int \kk(\cdot, x)  d P(x)$. 
 We can now  introduce the 
 notion of distance between two probability measures using the inner product
-of $\mathcal{H}_k$. This is called the maximum mean discrepancy (MMD) and
-it is given by
+of $\Hk$. This is called the maximum mean discrepancy (MMD) and
+is given by
 \begin{equation}\label{eq:mmd}
-\gamma_k(P,Q) = \| K_P - K_Q \|_{\mathcal{H}_k},
+\gamma_\kk(P,Q) = \| \kkk_P - \kkk_Q \|_{\Hk},
 \end{equation}
@@
 \begin{equation}\label{eq:mmd2}
-\gamma_k^2(P,Q) = \E k(X,X') + \E k(Y,Y') - 2 \E k(X, Y)
+\gamma_\kk^2(P,Q) = \E \kk(X,X') + \E \kk(Y,Y') - 2 \E \kk(X, Y)
 \end{equation}
@@
 From the equality between \eqref{eq:mmd} and \eqref{eq:mmd2} we also
-have that
+have 
 \begin{equation}\label{eq:inner_data}
-\langle K_P, K_Q \rangle_{\mathcal{H}_k} = \E \, k(X, Y).
+\langle \kkk_P, \kkk_Q \rangle_{\Hk} = \E \, \kk(X, Y).
 \end{equation}
-Thus we can compute the inner product between the embedded distributions 
+Therefore, in practice, we can estimate the inner product between the 
+embedded distributions 
 by averaging the kernel function over sampled data.
@@
 type and symmetric positive definite kernels are closely related
-\cite{Berg1984}. Let $\rho: \mathbb{R}^D \times \mathbb{R}^D \to \mathbb{R}$
+\cite{Berg1984}. Let $\rho: \mathcal{X} \times \mathcal{X} \to \mathbb{R}$
 be a semimetric, 
-and $x_0 \in \mathbb{R}^D$ an arbitrary but fixed point.
+and $x_0 \in \mathcal{X}$ an arbitrary but fixed point.
 Define
-\begin{equation}\label{eq:kernel_semimetric}
-k(x,y) = \tfrac{1}{2} \left\{  \rho(x,x_0) + \rho(y,x_0) - \rho(x,y)\right\}.
+\begin{equation}
+\label{eq:kernel_semimetric}
+\kk(x,y) = \tfrac{1}{2} \left\{  \rho(x,x_0) + \rho(y,x_0) - \rho(x,y)\right\}.
 \end{equation}
-Then $k$ is positive definite if and only if $\rho$ is of negative type.
-Thus we have a family of kernels, one for each choice of $x_0$. Conversely,
-if $\rho$ is a semimetric of negative type and $k$ is a kernel in this
+Then, $\kk$ is positive definite if and only if $\rho$ is of negative type
+\eqref{eq:negative_type}.
+Here we have a family of kernels, one for each choice of $x_0$. Conversely,
+if $\rho$ is a semimetric of negative type and $\kk$ is a kernel in this
 family, then 
-\begin{equation}\label{eq:gen_kernel}
-\rho(x,y) = k(x,x) + k(y,y) -2k(x,y) = \| K_x - K_y
-\|^2_{\mathcal{H}_k},
+\begin{equation}
+\label{eq:gen_kernel}
+\rho(x,y) = \kk(x,x) + \kk(y,y) -2\kk(x,y) = \| \kkk_x - \kkk_y
+\|^2_{\Hk},
 \end{equation}
 and the canonical feature map 
-$\varphi: x \mapsto K_x$ is injective \cite{Sejdinovic2013}.
-We say that the kernel $k$ generates the semimetric $\rho$. 
+$\varphi: x \mapsto \kkk_x$ is injective \cite{Sejdinovic2013}.
+We say that the kernel $\kk$ generates the semimetric $\rho$. 
 If two different kernels generate the same $\rho$, they are
@@
 \cite{Sejdinovic2013}. If $\rho$ is a semimetric
-of negative type and $k$ a kernel that generates $\rho$, then
+of negative type and $\kk$ a kernel that generates $\rho$, then
 \begin{align}
-\Energy(P, Q) &\equiv 2\E \rho(X,Y) - \E \rho(X,X') - \E \rho(Y,Y') 
+\Energy(P, Q) &\equiv 2\E \, \rho(X,Y) - \E \, \rho(X,X') - \E \, \rho(Y,Y') 
 \label{eq:EphoDef}\\
-&= 2 \left[ \E \, k(X, X') + \E \, k(Y, Y') - 2\E \, k(X, Y)\right] \\
-&=2 \gamma_k^2(P,Q) 
+&= 2 \left[ \E \, \kk(X, X') + \E \, \kk(Y, Y') - 2\E \, \kk(X, Y)\right] \\
+&=2 \gamma_\kk^2(P,Q) 
 \end{align}
@@
 \eqref{eq:EphoDef}, and using \eqref{eq:mmd2}.
-Since $\gamma_k^2(P, Q) = \| K_P - K_Q \|^2_{\mathcal{H}_k}$, we
-can compute the energy distance using the inner product in the RKHS. Moreover,
+Since $\gamma_k^2(P, Q) = \| \kkk_P - \kkk_Q \|^2_{\Hk}$, we
+can compute the energy distance using the inner product of $\Hk$. Moreover,
 this can be computed from the data according to \eqref{eq:inner_data}.
 
-\section{Clustering using Energy Distance}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Energy Distance based Clustering}
+\label{sec:clustering_theory}
+
+Now we formulate an optimization problem for clustering based on 
+energy statistics
+and RKHS, introduced in the previous section.
+Assume we have data $\mathbb{X} = \{ x_1,x_2,\dotsc, x_n \}$, where
+$x_i \in \mathcal{X}$ is in some space of 
+negative type \eqref{eq:negative_type},
+and a partition $\mathbb{X} = \cup_{j=1}^k \C_j$, where
+$\C_i \cap \C_j = \emptyset$.
+Each expectation in 
+\eqref{eq:EphoDef}
+can be computed 
+through the function
+\begin{equation}
+\label{eq:g_def}
+g (\C_i, \C_j) \equiv 
+\dfrac{1}{n_i n_j}
+\sum_{x \in \C_i} 
+\sum_{y \in \C_j} \rho(x, y)
+\end{equation}
+where $n_i = |\C_i|$ is the number of elements in partition
+$\C_i$. In energy statistics 
+\cite{Szkely2013} we have the within energy dispersion
+\begin{equation}
+\label{eq:within}
+W = 
+\sum_{j=1}^{k} \dfrac{n_j}{2} g(\C_j, \C_j),
+\end{equation}
+and also the between-sample energy statistic
+\begin{equation}
+S = 
+\sum_{1 \le  j < l \le k } \dfrac{n_j n_{l}}{2 n} \left[
+2 g(\C_j, \C_l) - 
+g(\C_j, \C_j) - 
+g(\C_l, \C_l)
+\right],
+\end{equation}
+where $n=\sum_{j=1}^{k} n_j$. Given a set of distributions
+$\{ P_j\}_{j=1}^k$ where $x \in \C_j \sim P_j$, the quantity
+$S_\alpha$ provides
+a \emph{nonparametric} test statistic for equality of distributions.
+Under the null hypothesis $H_0: P_1=P_2=\dotsm=P_k$, $S_\alpha$ is small, 
+and under
+the alternative hypothesis $H_1: P_i \ne P_j$ for at least two $i\ne j$, 
+$S_\alpha \to \infty$ as the sample size is large, $n \to \infty$.
+This test is nonparametric in the sense that it does not make any assumptions
+about the distributions $P_j$.
+
+Based on this test statistic for equality of distributions, a possible
+criteria for clustering data is to 
+maximize $S$. However, 
+it can be shown that the total dispersion of the data obeys \cite{Szkely2013}
+\begin{equation}
+T(\mathbb{X}) 
+= W + S = \dfrac{n}{2}
+g(\mathbb{X}, \mathbb{X}). 
+\end{equation}
+Note that $T$ only depends on the pooled data, so it 
+does not depend on how we partition $\mathbb{X}$. Therefore, maximizing
+$S$ is equivalent to minimizing $W$, which
+has a simpler form. Thus, our clustering problem corresponds to
+find the best partition of $\mathbb{X}$ to solve the optimization problem
+\begin{equation}
+\label{eq:minimize}
+\min_{\{ \C_j \} } W(\{ 
+\C_1, \dotsc, \C_k
+\})
+\end{equation}
+where each datapoint belongs to one and only one partition (hard assignments).
+
+Now fix an arbitrary point $x_0 \in \mathcal{X}$, and 
+suppose that the kernel $K$ generates $\rho$, such that 
+\eqref{eq:kernel_semimetric} and \eqref{eq:gen_kernel} hold.
+Then we can write \eqref{eq:within} as
+\begin{equation}\label{eq:W2}
+W(\{ \C_j \} )
+= \dfrac{1}{2} \sum_{j=1}^k \dfrac{1}{n_j} \sum_{x,y \in \C_j} \rho(x,y)
+= \sum_{j=1}^k \sum_{x \in \C_j}  \bigg(
+\kk(x,x) - \dfrac{1}{n_j} \sum_{y \in \C_j} \kk(x,y) \bigg)
+\end{equation}
+When minimizing $W$, the first term is just a constant and does
+not contribute,
+therefore the problem reduces to
+\begin{equation}
+\label{eq:max_prob}
+\max_{ \{ \C_j \} } \sum_{j=1}^k \dfrac{1}{n_j} \sum_{x,y\in C_j} \kk(x,y) .
+\end{equation}
+Let us introduce the binary matrix $Z \in \{ 0,1 \}^{n\times k}$ 
+such that 
+\begin{equation}
+Z_{ij} = \begin{cases}
+1 & \mbox{if $x_i \in \C_j$ } \\
+0 & \mbox{otherwise.}
+\end{cases}
+\end{equation}
+Notice that $D = Z^T Z = \diag( n_1, n_2, \dotsc, n_k )$ contains the number
+of elements in each partition. Introducing the kernel matrix
+$G \in \mathbb{R}^{n\times n}$ such that
+\begin{equation}
+\label{eq:kernel_matrix}
+G_{ij} = \kk(x_i, x_j),
+\end{equation}
+then \eqref{eq:max_prob} is equal to
+$\max \Tr \left\{ D^{-1} Z^\top G Z \right\}$. 
+Therefore, our clustering optimization problem based on energy statistics
+can be formulated as
+\begin{equation}\label{eq:qcqp}
+\begin{aligned}
+& \max_{Z} \Tr\left\{ \big( Z D^{-1/2}\big)^\top G 
+\big( ZD^{-1/2} \big) 
+\right\} \\
+&\mbox{s.t. $Z_{ij} \in \{0,1\}$, $\sum_{j=1}^k Z_{ij} = 1$, 
+$\sum_{i=1}^n Z_{ij} = n_j$, and $D = Z^\top Z$}.
+\end{aligned}
+\end{equation}
+This is a quadratic problem with integer constraints, which is 
+NP-hard. Let us now write this in terms of 
+$\Zt \equiv Z D^{-1/2}$, which componentwise is 
+\begin{equation}
+\Zt_{ij} = \begin{cases}
+\tfrac{1}{\sqrt{n_j}} & \mbox{if $x_i \in \C_j$ } \\
+0 & \mbox{otherwise.}
+\end{cases}
+\end{equation}
+We thus have
+\begin{equation}
+\label{eq:qcqp2}
+\max_{\Zt} \Tr \left\{ \Zt^\top G \Zt \right\}  \qquad
+\mbox{s.t. $\Zt \ge 0$, $\Zt^\top Z = I$, 
+$\Zt \Zt^\top \bm{e} = \bm{e}$},
+\end{equation}
+where $\bm{e} = (1,1,\dots,1) \in \mathbb{R}^n$ is the all-ones vector,
+and $G$ is the pairwise kernel matrix \eqref{eq:kernel_matrix} obtained
+from \eqref{eq:kernel_semimetric}.
+Therefore, to cluster data $\{ x_i \}_{i=1}^n \in \mathcal{X}$ into 
+$k$ partitions,
+assuming that $k$ is given, we first compute $G$  --- 
+which is  defined by an arbitrary semimetric of negative 
+type on $\mathcal{X}$  --- and then 
+solve the optimization problem \eqref{eq:qcqp2} for $\Zt \in
+\mathbb{R}^{n\times k}$. The $i$th row
+of $\Zt$ will contain a single nonzero element in some $j$th column,
+indicating that $x_i \in \C_j$.
+
+In general, 
+problem \eqref{eq:qcqp2} is NP-hard since it is a quadratically
+constrained quadratic problem (QCQP). There are  
+few methods available to tackle this kind of problem directly,
+which is computational prohibitive even for relatively small datasets.
+However, one can find an approximate solution by relaxing some 
+of the constraints. For instance, an approximation can be given
+by 
+$\max_{Y} \Tr\left\{ Y^\top G \, Y \right\}$ subject to $Y^\top Y = I$,
+and requiring that the rows of $Y$ are normalized. 
+It is possible to find a global solution to this problem by
+choosing $Y$ as the top $k$ eigenvectors of $G$, which results
+in 
+$\max \Tr \left\{ Y^\top G \, Y \right\}  = \sum_{i=1}^k \lambda_i(G)$, 
+which is the
+sum of the top $k$ eigenvalues of $G$.
+
+It is important to note that \eqref{eq:qcqp2} has the same
+formulation as kernel $k$-means, spectral clustering, and the maximum
+cut problem on graphs \cite{Dhillon}. The 
+result \eqref{eq:qcqp2} brings energy statistics based clustering
+into this broad picture, and \eqref{eq:qcqp2} should have 
+interesting applications in
+graph partitioning problems and unsupervised learning in general. 
+Furthermore, and most importantly, our analysis is valid
+for any space $\mathcal{X}$ equiped with a semimetric of negative type.
+This method is
+nonparametric since it does not assume any form of the 
+distribution of the data,
+contrary to $k$-means and gaussian mixture models (GMM), for example.
+Also, there is no concept of cluster center involved in this approach.
+
+\subsection{A Simple Algorithm}
+
+We can formulate an iterative algorithm to find
+an approximate solution to \eqref{eq:qcqp2} on the same lines
+as kernel $k$-means.
+Let $t$ be the iteration time. First, precompute the kernel
+matrix $G$, fix the number of clusters $k$, then
+perform the following steps:
+\begin{enumerate}
+\item \label{initialize} 
+Initialize clusters $\{ \C_1^{(0)},\dotsc,\C_k^{(0)} \}$, which
+determines the label matrix $\Zt^{(0)}$.
+\item \label{assignment} For each datapoint $x_i$ compute
+its cluster assignment through
+\begin{equation}
+\label{eq:algo}
+\Zt^{(t+1)}_{ij} = 
+\begin{cases}
+\tfrac{1}{\sqrt{n_j}} & \mbox{if $j = \argmax_\ell \tfrac{1}{n_\ell} 
+\sum_{m=1}^n G_{i m} \Zt^{(t)}_{m\ell}$ } \\
+0 & \mbox{otherwise.}
+\end{cases}
+\end{equation}
+\item If converged return $\Zt^{(t+1)}$, otherwise
+set $t = t+1$ and repeat step~\ref{assignment}.
+\end{enumerate}
+If data is $D$-dimensional, computing $G$ has complexity
+$O(n^2 D)$. Step \ref{assignment} above has complexity $O(n)$ for
+each point, thus total complexity $O(n^2)$. Assuming we perform $T$ 
+iterations,
+the total complexity of the algorithm is $O(n^2(D + T))$.
+We can initialize the algorithm in step~\ref{initialize} with any method
+we want, a good alternative is the initialization from $k$-means++.
+
+\subsection{Two-Class Problem in One Dimension}
+
+If data is one-dimensional and we choose
+$\rho(x,y) = |x - y|$, we can actually compute 
+\eqref{eq:g_def} in $O(n \log n)$ instead of $O(n^2 D)$ and find
+a direct solution to \eqref{eq:minimize}. 
+This is done by noticing that
+\begin{equation}
+|x - y|  = \mathbbm{1}(x \ge y ) (x - y) -
+\mathbbm{1}(x < y) (x - y).
+\end{equation}
+Denote
+\begin{equation}
+n_j^{-} (x) = \sum_{y \in \C_j} \mathbbm{1}(y \le x), \quad  
+n_j^{+} (x) = \sum_{y \in C_j} \mathbbm{1}(y > x), \quad
+\mathcal{D}_j(x) = \dfrac{n_j^-(x) - n_j^+(x)}{n_j}
+\end{equation}
+where $n_j^+(x)$ is the number of elements in $\C_j$ which are larger
+than $x \in \C_i$, and so on.
+Now \eqref{eq:g_def} can be written as
+\begin{equation}
+\label{eq:g1d}
+g(\C_i, \C_j) = 
+\dfrac{1}{n_i} \sum_{x \in \C_i} \mathcal{D}_j(x) \, x - 
+\dfrac{1}{n_j} \sum_{y \in \C_j} \mathcal{D}_i(y) \, y.
+\end{equation}
+If we merge $\C_i$ and $\C_j$ and sort the data, this expression 
+can be computed in $O(n)$, where $n=n_i+n_j$.
+Now for a two-class problem, we have to loop through each point
+and compute \eqref{eq:within} to find the best split. Since sorting
+sorting usually takes $O(n \log n)$, the complexity of
+this procedure is $O(n(\log n + n)) = O(n^2)$.
 

### 2017-05-16 qcqp_clustering/code/eclust/data.py

@@ -154,7 +154,7 @@ def plot(X, z, fname='plot.pdf'):
     plt.axes().set_aspect('equal', 'datalim')
     fig.savefig(fname)
 
-def hist(X, z, fname='plot.pdf'):
+def histogram(X, z, fname='plot.pdf'):
     """Plot histograms of 1-dimensional data."""
     z_unique = np.unique(z)
     fig = plt.figure()

### 2017-05-16 qcqp_clustering/code/eclust/energy.py

@@ -1,4 +1,4 @@
-"""Energy Statistics Functions"""
+"""Energy Statistics Functions."""
 
 # Guilherme Franca <guifranca@gmail.com>
 # Johns Hopkinks University, Neurodata
@@ -103,12 +103,14 @@ def total_dispersion(A):
     else:
         return len(X)/2*mean(X, X)
 
-def energy_kernel(x, y, alpha=1, cutoff=0):
+def energy_kernel(x, y, alpha=1):
     w = np.power(np.linalg.norm(x), alpha) + \
         np.power(np.linalg.norm(y), alpha) - \
         np.power(np.linalg.norm(x-y), alpha)
-    if w <= cutoff:
-        w = 0
+    return w
+
+def distance_kernel(x, y, alpha=1):
+    w = - np.power(np.linalg.norm(x-y), alpha)
     return w
 
 

### 2017-05-16 qcqp_clustering/code/eclust/kmeans.py

@@ -1,0 +1,52 @@
+"""K-means++ clustering."""
+
+from __future__ import division
+
+import numpy as np
+import itertools
+import scipy.optimize
+
+import kmeanspp
+
+
+def euclidean(a, b):
+    return np.linalg.norm(a-b)**2
+
+def kmeans(K, X, distance=euclidean, max_iter=50):
+    """K is the number of clusters. X is the dataset, which is a matrix
+    with one data point per row. distance is a function
+    which accepts two data points and returns a real number.
+    
+    """
+    N = X.shape[0] # number of points
+    mus = kmeanspp.kpp(K, X, ret='centers')
+    labels = np.empty(N, dtype=np.int8)
+    
+    converged = False
+    count = 0
+    while not converged and count < max_iter:
+        
+        converged = True
+        
+        # for each vector label it according to the closest centroid
+        for n in range(N):
+            for k in range(K):
+                D = np.array([distance(X[n], mus[k]) for k in range(K)])
+            labels[n] = np.argmin(D)
+
+        # update the centroids based on the vectors in the cluster
+        for k in range(K):
+            old_mu = mus[k]
+            new_mu = np.mean(X[np.where(labels==k)], axis=0)
+            if not np.array_equal(new_mu, old_mu):
+                mus[k] = new_mu
+                converged = False
+        
+        count += 1
+
+    if count >= max_iter:
+        print "Warning: K-means didn't converge after %i iterations." % count
+    
+    return labels
+    #return labels, mus, J, count
+

### 2017-05-16 qcqp_clustering/code/eclust/kmeanspp.py

@@ -6,13 +6,10 @@
 import numpy as np
 
 
-def energy_kernel(x, y):
-    return np.linalg.norm(x) + np.linalg.norm(y) - np.linalg.norm(x-y)
-
 def euclidean_distance(x, y):
     return np.linalg.norm(x-y)**2
 
-def kpp(k, X):
+def kpp(k, X, ret='labels'):
     """This is the k-means++ initialization proposed by Arthur and
     Vassilvitskii (2007). We label the points according to closest
     distance to the centers.
@@ -31,12 +28,18 @@ def kpp(k, X):
         C.append(X[j])
     mus = np.array(C)
     
-    labels = np.empty(N, dtype=np.int8)
-    for n in range(N):
-        for k in range(K):
-            D = np.array([distance(X[n], mus[k]) for k in range(K)])
-        labels[n] = np.argmin(D)
-    return labels
+    if ret == 'labels' or ret == 'both':
+        labels = np.empty(N, dtype=np.int8)
+        for n in range(N):
+            for k in range(K):
+                D = np.array([distance(X[n], mus[k]) for k in range(K)])
+            labels[n] = np.argmin(D)
+        if ret == 'both':
+            return mus, labels
+        else:
+            return labels
+    else:
+        return mus
 
 def discrete_rv(p):
     """Return an integer according to probability function p."""

### 2017-05-16 qcqp_clustering/draft/energy_clustering.tex
draft

@@
 \newcommand\C{{\mathcal{C}}}
-\newcommand\Zt{\widetilde{Z}}
+%\newcommand\Zt{\widetilde{Z}}
+\newcommand\Zt{Y}
 
@@
 \max_{\Zt} \Tr \left\{ \Zt^\top G \Zt \right\}  \qquad
-\mbox{s.t. $\Zt \ge 0$, $\Zt^\top Z = I$, 
+\mbox{s.t. $\Zt \ge 0$, $\Zt^\top \Zt = I$, 
 $\Zt \Zt^\top \bm{e} = \bm{e}$},
 \end{equation}
-where $\bm{e} = (1,1,\dots,1) \in \mathbb{R}^n$ is the all-ones vector,
+where $\bm{e} = (1,1,\dots,1)^\top \in \mathbb{R}^n$ is the all-ones vector,
 and $G$ is the pairwise kernel matrix \eqref{eq:kernel_matrix} obtained
@@
 Furthermore, and most importantly, our analysis is valid
-for any space $\mathcal{X}$ equiped with a semimetric of negative type.
+for any space $\mathcal{X}$ equiped with a semimetric of negative type $\rho$.
 This method is
@@
 as kernel $k$-means.
-Let $t$ be the iteration time. First, precompute the kernel
+Let $t$ be the iteration time. First precompute the kernel
 matrix $G$, fix the number of clusters $k$, then
@@
 can be computed in $O(n)$, where $n=n_i+n_j$.
-Now for a two-class problem, we have to loop through each point
-and compute \eqref{eq:within} to find the best split. Since sorting
+For a two-class problem, we have to loop through each point
+and compute \eqref{eq:within} to find the best split of the data. 
+Since sorting
 sorting usually takes $O(n \log n)$, the complexity of

### 2017-05-17 qcqp_clustering/draft/energy_clustering.tex
draft again

@@
 \newcommand\Zt{Y}
+\newcommand{\Ind}[1]{\delta_{#1}}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Energy Statistics and RKHS}
+\section{Background on Energy Statistics and RKHS}
 \label{sec:background}
@@
 For more details we refer the reader
-to \cite{Szkely2013} (and references therein) and also \cite{Sejdinovic2013}.
+to 
+\cite{Szkely2013} 
+(and references therein) and also \cite{Sejdinovic2013}.
 
@@
 Let $X, Y \in \mathcal{X}$ and replace the Euclidean norm by
-$\rho: \mathcal{X}\times\mathcal{X} \to \mathbb{R}^+$, i.e.
-$\| X - Y\| \to \rho(X,Y)$, where $\rho$ is a so-called semimetric
-of negative type \cite{Sejdinovic2013}, which satisfy
+$\rho: \mathcal{X}\times\mathcal{X} \to \mathbb{R}$, i.e.
+$\| X - Y\| \to \rho(X,Y)$, where $\rho$ is a so-called \emph{semimetric
+of negative type},
+which satisfy 
 \begin{equation}
@@
 This implies
-that $\kk(x,y)$ is symmetric and positive definite, 
+that $\kk(x,y)$ is symmetric and positive semi-definite, 
 $\sum_{i,j=1}^n c_i c_j
@@
 and positive definite function $\kk: \mathcal{X}\times \mathcal{X} \to
-\mathbb{R}$, there is an associated RKHS, $\Hk$, 
+\mathbb{R}$, there is an associated RKHS $\Hk$ 
 with reproducing
@@
 The following important result shows that semimetrics of negative
-type and symmetric positive definite kernels are closely related
+type and symmetric positive semi-definite kernels are closely related
 \cite{Berg1984}. Let $\rho: \mathcal{X} \times \mathcal{X} \to \mathbb{R}$
@@
 \end{equation}
-Then, $\kk$ is positive definite if and only if $\rho$ is of negative type
+Then, $\kk$ is positive semi-definite if and only if $\rho$ is of negative type
 \eqref{eq:negative_type}.
@@
 &= 2 \left[ \E \, \kk(X, X') + \E \, \kk(Y, Y') - 2\E \, \kk(X, Y)\right] \\
-&=2 \gamma_\kk^2(P,Q) 
+&=2 \gamma_\kk^2(P,Q) .
 \end{align}
@@
 \eqref{eq:EphoDef}, and using \eqref{eq:mmd2}.
-Since $\gamma_k^2(P, Q) = \| \kkk_P - \kkk_Q \|^2_{\Hk}$, we
+Since $\gamma_k^2(P, Q) = \| \kkk_P - \kkk_Q \|^2_{\Hk}$ we
 can compute the energy distance using the inner product of $\Hk$. Moreover,
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Energy Distance based Clustering}
-\label{sec:clustering_theory}
-
-Now we formulate an optimization problem for clustering based on 
-energy statistics
-and RKHS, introduced in the previous section.
+Finally, let us recall the main formulas for test statistics
+\cite{Szkely2013}. 
 Assume we have data $\mathbb{X} = \{ x_1,x_2,\dotsc, x_n \}$, where
@@
 negative type \eqref{eq:negative_type},
-and a partition $\mathbb{X} = \cup_{j=1}^k \C_j$, where
+and a partition $\mathbb{X} = \cup_{j=1}^k \C_j$ where
 $\C_i \cap \C_j = \emptyset$.
@@
 \end{equation}
-where $n_i = |\C_i|$ is the number of elements in partition
-$\C_i$. In energy statistics 
-\cite{Szkely2013} we have the within energy dispersion
+where $n_i = |\C_i|$ is the number of elements in 
+$\C_i$. 
+The \emph{within energy dispersion} is defined by
 \begin{equation}
 \label{eq:within}
-W = 
+W \equiv
 \sum_{j=1}^{k} \dfrac{n_j}{2} g(\C_j, \C_j),
 \end{equation}
-and also the between-sample energy statistic
+and the \emph{between-sample energy statistic} is defined by
 \begin{equation}
-S = 
+\label{eq:between}
+S \equiv
 \sum_{1 \le  j < l \le k } \dfrac{n_j n_{l}}{2 n} \left[
@@
 g(\C_l, \C_l)
-\right],
+\right].
 \end{equation}
-where $n=\sum_{j=1}^{k} n_j$. Given a set of distributions
+Given a set of distributions
 $\{ P_j\}_{j=1}^k$ where $x \in \C_j \sim P_j$, the quantity
-$S_\alpha$ provides
+$S$ provides
 a \emph{nonparametric} test statistic for equality of distributions.
-Under the null hypothesis $H_0: P_1=P_2=\dotsm=P_k$, $S_\alpha$ is small, 
+Under the null hypothesis $H_0: P_1=P_2=\dotsm=P_k$, $S$ is small, 
 and under
 the alternative hypothesis $H_1: P_i \ne P_j$ for at least two $i\ne j$, 
-$S_\alpha \to \infty$ as the sample size is large, $n \to \infty$.
+$S \to \infty$ as the sample size is large, $n \to \infty$.
 This test is nonparametric in the sense that it does not make any assumptions
@@
 
-Based on this test statistic for equality of distributions, a possible
-criteria for clustering data is to 
-maximize $S$. However, 
-it can be shown that the total dispersion of the data obeys \cite{Szkely2013}
-\begin{equation}
-T(\mathbb{X}) 
-= W + S = \dfrac{n}{2}
-g(\mathbb{X}, \mathbb{X}). 
-\end{equation}
-Note that $T$ only depends on the pooled data, so it 
-does not depend on how we partition $\mathbb{X}$. Therefore, maximizing
-$S$ is equivalent to minimizing $W$, which
-has a simpler form. Thus, our clustering problem corresponds to
-find the best partition of $\mathbb{X}$ to solve the optimization problem
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Clustering Based on Energy Statistics}
+\label{sec:clustering_theory}
+
+This section contains the main results of this paper, where 
+we propose an optimization problem for clustering data
+based on energy statistics and RKHS introduced in the previous section.
+
+Based on the test statistic \eqref{eq:between}, the obvious
+criterion for clustering data is to 
+maximize $S$, which intuitivelly makes the distribution of each
+cluster $\C_j$ as different
+as possible from the other ones. However, the following 
+simple result
+shows that this is equivalent to minimizing
+\eqref{eq:within}, which has a more convenient form.
+
+\begin{proposition}
+\label{th:minimize}
+Let $\mathbb{X} = \{x_i\}_{i=1}^{n}$ where 
+$x_i$ lives in a space $\mathcal{X}$ endowed with a semi-metric $\rho$ of
+negative type. For a fixed integer $k$,
+the partition
+$\mathbb{X} = \cup_{j=1}^k \C_j$, where $\C_i \cap C_j = \emptyset$ for
+all $i\ne j$, maximizes $S$ if and only if
 \begin{equation}
@@
 \C_1, \dotsc, \C_k
-\})
+\}).
 \end{equation}
-where each datapoint belongs to one and only one partition (hard assignments).
+\end{proposition}
+\begin{proof}
+It can be shown that the total dispersion of the data obeys \cite{Szkely2013}
+\begin{equation}
+W + S = \dfrac{n}{2} g(\mathbb{X}, \mathbb{X}). 
+\end{equation}
+Note, however, that the right hand side of this equation 
+only depends on the pooled data, so it is a constant
+independent on the choice of partition. Therefore, maximizing
+$S$ is equivalent to minimizing~$W$.
+\end{proof}
 
-Now fix an arbitrary point $x_0 \in \mathcal{X}$, and 
-suppose that the kernel $K$ generates $\rho$, such that 
-\eqref{eq:kernel_semimetric} and \eqref{eq:gen_kernel} hold.
-Then we can write \eqref{eq:within} as
-\begin{equation}\label{eq:W2}
-W(\{ \C_j \} )
-= \dfrac{1}{2} \sum_{j=1}^k \dfrac{1}{n_j} \sum_{x,y \in \C_j} \rho(x,y)
-= \sum_{j=1}^k \sum_{x \in \C_j}  \bigg(
-\kk(x,x) - \dfrac{1}{n_j} \sum_{y \in \C_j} \kk(x,y) \bigg)
-\end{equation}
-When minimizing $W$, the first term is just a constant and does
-not contribute,
-therefore the problem reduces to
-\begin{equation}
-\label{eq:max_prob}
-\max_{ \{ \C_j \} } \sum_{j=1}^k \dfrac{1}{n_j} \sum_{x,y\in C_j} \kk(x,y) .
-\end{equation}
-Let us introduce the binary matrix $Z \in \{ 0,1 \}^{n\times k}$ 
-such that 
-\begin{equation}
-Z_{ij} = \begin{cases}
-1 & \mbox{if $x_i \in \C_j$ } \\
-0 & \mbox{otherwise.}
-\end{cases}
-\end{equation}
-Notice that $D = Z^T Z = \diag( n_1, n_2, \dotsc, n_k )$ contains the number
-of elements in each partition. Introducing the kernel matrix
+Thus, given $k$, our clustering problem ammounts to
+finding the best partition of the data through solving~\eqref{eq:minimize}.
+In this setting, each datapoint belongs to one and only one cluster
+(hard assignments).
+
+Based on 
+\eqref{eq:kernel_semimetric} and \eqref{eq:gen_kernel}, assume that 
+the kernel $\kk: \mathcal{X} \times \mathcal{X} \to \mathbb{R}$ 
+generates $\rho$. 
+Let us define  the kernel matrix
 $G \in \mathbb{R}^{n\times n}$ such that
@@
 \label{eq:kernel_matrix}
-G_{ij} = \kk(x_i, x_j),
+G_{ij} \equiv \kk(x_i, x_j).
 \end{equation}
-then \eqref{eq:max_prob} is equal to
-$\max \Tr \left\{ D^{-1} Z^\top G Z \right\}$. 
-Therefore, our clustering optimization problem based on energy statistics
-can be formulated as
-\begin{equation}\label{eq:qcqp}
-\begin{aligned}
-& \max_{Z} \Tr\left\{ \big( Z D^{-1/2}\big)^\top G 
-\big( ZD^{-1/2} \big) 
-\right\} \\
-&\mbox{s.t. $Z_{ij} \in \{0,1\}$, $\sum_{j=1}^k Z_{ij} = 1$, 
-$\sum_{i=1}^n Z_{ij} = n_j$, and $D = Z^\top Z$}.
-\end{aligned}
+Let $Z$ be the label matrix, $Z \in \{ 0,1 \}^{n\times k}$, 
+with only one nonvanishing entry per row indicating to which cluster (columns)
+each point (rows) belongs to. This matrix satisfy
+$Z^\top Z = D$ where $D = \diag( n_1,\dotsc, n_k )$  contains
+the number of points in each cluster. We also introduce the rescaled
+matrix  $Y \equiv Z D^{-1/2}$. In component form they are given by
+\begin{equation}
+\label{eq:label_matrix}
+Z_{ij} \equiv \begin{cases}
+1 & \mbox{if $x_i \in \C_j$ } \\
+0 & \mbox{otherwise}
+\end{cases} \qquad
+\Zt_{ij} \equiv \begin{cases}
+\tfrac{1}{\sqrt{n_j}} & \mbox{if $x_i \in \C_j$ } \\
+0 & \mbox{otherwise}
+\end{cases} .
 \end{equation}
-This is a quadratic problem with integer constraints, which is 
-NP-hard. Let us now write this in terms of 
-$\Zt \equiv Z D^{-1/2}$, which componentwise is 
-\begin{equation}
-\Zt_{ij} = \begin{cases}
-\tfrac{1}{\sqrt{n_j}} & \mbox{if $x_i \in \C_j$ } \\
-0 & \mbox{otherwise.}
-\end{cases}
-\end{equation}
-We thus have
+
+Our next result reveals the optimization problem behind \eqref{eq:minimize},
+which is in general NP-hard since
+it is a quadratically constrained quadratic problem (QCQP).
+
+\begin{proposition} 
+\label{th:qcqp2}
+The problem \eqref{eq:minimize} is equivalent to
 \begin{equation}
@@
 where $\bm{e} = (1,1,\dots,1)^\top \in \mathbb{R}^n$ is the all-ones vector,
-and $G$ is the pairwise kernel matrix \eqref{eq:kernel_matrix} obtained
-from \eqref{eq:kernel_semimetric}.
+and $G$ is the pairwise kernel matrix \eqref{eq:kernel_matrix}, assumed
+to be given.
+\end{proposition}
+\begin{proof}
+Replacing \eqref{eq:gen_kernel}  into
+\eqref{eq:g_def},  
+we can write \eqref{eq:within} as
+\begin{equation}
+\label{eq:W2}
+W(\{ \C_j \} )
+= \dfrac{1}{2} \sum_{j=1}^k \dfrac{1}{n_j} \sum_{x,y \in \C_j} \rho(x,y)
+= \sum_{j=1}^k \sum_{x \in \C_j}  \bigg(
+\kk(x,x) - \dfrac{1}{n_j} \sum_{y \in \C_j} \kk(x,y) \bigg).
+\end{equation}
+The first term is a global constant 
+which does not contribute,
+so minimizing \eqref{eq:W2} is the same as
+\begin{equation}
+\label{eq:max_prob}
+\max_{ \{ \C_j \} } \sum_{j=1}^k \dfrac{1}{n_j} \sum_{x,y\in C_j} \kk(x,y) .
+\end{equation}
+But 
+\begin{equation}
+\sum_{x, y \in \C_j} \kk(x, y) =
+\sum_{l=1}^{n} \sum_{m=1}^{n} Z_{lj} Z_{mj} G_{lm} = 
+(Z^\top G Z)_{jj}
+\end{equation}
+where we used the definitions \eqref{eq:kernel_matrix} and
+\eqref{eq:label_matrix}. Thus \eqref{eq:max_prob} is equal to
+$\max \Tr \left\{ D^{-1} Z^\top G Z \right\}$. Now we can
+use the cyclic property
+of the trace, and by the own definition of the matrix $Z$
+in \eqref{eq:label_matrix} we obtain the following integer
+programing problem:
+\begin{equation}\label{eq:qcqp}
+\begin{aligned}
+& \max_{Z} \Tr\left\{ \big( Z D^{-1/2}\big)^\top G 
+\big( ZD^{-1/2} \big) 
+\right\} \\
+&\mbox{s.t. $Z_{ij} \in \{0,1\}$, $\sum_{j=1}^k Z_{ij} = 1$, 
+$\sum_{i=1}^n Z_{ij} = n_j$.}
+\end{aligned}
+\end{equation}
+Now we write this in terms of $Y = Z D^{-1/2}$, given by
+\eqref{eq:label_matrix}. The objective function immediately becomes
+$\Tr Y^\top G Y$. Notice that the above constraints
+immediatly imply that $Z^T Z = D$, which in turn gives
+$D^{-1/2} Y^T Y D^{-1/2} = D$, or $Y^\top Y = I$. 
+Also, every entry of $Y$ is positive by definition,
+$Y \ge 0$. Now it only remains to show the last 
+constraint in \eqref{eq:qcqp2}, which comes from the last
+constraint in \eqref{eq:qcqp}. In matrix form this reads
+$Z^T \bm{e} = D \bm{e}$. Replacing by $Z=YD^{1/2}$ we have
+$Y^\top \bm{e} = D^{1/2} \bm{e}$. Multiplying this last equation
+on the left by $Y$, and noticing
+that $Y D^{1/2} \bm{e} = Z \bm{e} = \bm{e}$, we finally obtain
+$Y Y^\top \bm{e} = \bm{e}$ as desired.
+\end{proof}
+
 Therefore, to cluster data $\{ x_i \}_{i=1}^n \in \mathcal{X}$ into 
 $k$ partitions,
-assuming that $k$ is given, we first compute $G$  --- 
+assuming that $k$ is given, we first compute the kernel matrix
+$G$  --- 
 which is  defined by an arbitrary semimetric of negative 
@@
 In general, 
-problem \eqref{eq:qcqp2} is NP-hard since it is a quadratically
-constrained quadratic problem (QCQP). There are  
-few methods available to tackle this kind of problem directly,
+problem \eqref{eq:qcqp2} is NP-hard and there
+are few methods
+available to tackle this kind of problem directly,
 which is computational prohibitive even for relatively small datasets.
 However, one can find an approximate solution by relaxing some 
-of the constraints. For instance, an approximation can be given
+of the constraints, or obtaining a relaxed SDP version of it. 
+For instance, a simple approximation can be given
 by 
-$\max_{Y} \Tr\left\{ Y^\top G \, Y \right\}$ subject to $Y^\top Y = I$,
+$\max_{Y} \Tr Y^\top G \, Y $ subject to $Y^\top Y = I$,
 and requiring that the rows of $Y$ are normalized. 
@@
 which is the
-sum of the top $k$ eigenvalues of $G$.
+sum of the top $k$ eigenvalues of $G$. This is what is usually done
+in spectral clustering. However, we will propose a simple iterative
+algorithm in the next section.
 
-It is important to note that \eqref{eq:qcqp2} has the same
+An interesting point is that \eqref{eq:qcqp2} has the same
 formulation as kernel $k$-means, spectral clustering, and the maximum
-cut problem on graphs \cite{Dhillon}. The 
-result \eqref{eq:qcqp2} brings energy statistics based clustering
+cut problem on graphs \cite{Dhillon}. 
+Proposition~\ref{th:qcqp2} brings energy statistics based clustering
 into this broad picture, and \eqref{eq:qcqp2} should have 
@@
 graph partitioning problems and unsupervised learning in general.
+Notice that it is straightforward to generalize \eqref{eq:qcqp2} to
+a weighted version, where weights $W = \diag(w(x_1), \dotsc, w(x_n))^\top $
+are associated to each data point $x_i$. One just replace the objective
+function in \eqref{eq:qcqp2} by 
+\begin{equation}
+\max_Y \Tr \left\{ Y^\top W^{1/2} G W^{1/2} Y \right\}.
+\end{equation}
 Furthermore, and most importantly, our analysis is valid
@@
 contrary to $k$-means and gaussian mixture models (GMM), for example.
-Also, there is no concept of cluster center involved in this approach.
+Moreover, this approach does not require the concept of the cluster mean, or
+median, which can be ill-defined in some types of data, such as images for
+instance.
 
-\subsection{A Simple Algorithm}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{A Simple Iterative Algorithm}
+\label{sec:iterative_algo}
 
@@
 
-\subsection{Two-Class Problem in One Dimension}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Two-Class Problem in One Dimension}
 
+Here we consider an alternative approach in the simplest case possible, which
+is one-dimensional data and a two-class problem.
 If data is one-dimensional and we choose
@@
 \begin{equation}
-|x - y|  = \mathbbm{1}(x \ge y ) (x - y) -
-\mathbbm{1}(x < y) (x - y).
+\begin{aligned}
+|x - y|  &= \Ind{x \ge y} (x - y) -
+\Ind{x < y} (x - y) \\
+&= 
+\left( \Ind{x \ge y} - \Ind{x < y} \right) x + 
+\left( \Ind{y > x} - \Ind{y \le x} \right) y
+\end{aligned}
 \end{equation}
-Denote
+where $\Ind{}$ is the indicator function.
+Let $\C$ be a partition with
+$n$ elements. Using the above distance in \eqref{eq:g_def} we have
 \begin{equation}
-n_j^{-} (x) = \sum_{y \in \C_j} \mathbbm{1}(y \le x), \quad  
-n_j^{+} (x) = \sum_{y \in C_j} \mathbbm{1}(y > x), \quad
-\mathcal{D}_j(x) = \dfrac{n_j^-(x) - n_j^+(x)}{n_j}
+\label{eq:g_ind}
+g(\C,\C) = \dfrac{1}{n^2} \sum_{x \in \C} 
+\sum_{y \in \C} 
+\left(
+\Ind{x \ge y} + \Ind{y > x} - 
+\Ind{x \ge y}-\Ind{x < y} \right) x.
 \end{equation}
-where $n_j^+(x)$ is the number of elements in $\C_j$ which are larger
-than $x \in \C_i$, and so on.
-Now \eqref{eq:g_def} can be written as
+The sum over $y$ can be eliminated since the term in
+parenthesis is simply counting the number of elements in $\C$ that satisfy
+the conditions of the indicator functions. Therefore, assume
+that we order the data as 
+$\bar{\C} = [ x_j \in \C: x_1 \le x_2 \le \dotsm \le x_{n}]$. Thus, we
+can write \eqref{eq:g_ind} as
 \begin{equation}
 \label{eq:g1d}
-g(\C_i, \C_j) = 
-\dfrac{1}{n_i} \sum_{x \in \C_i} \mathcal{D}_j(x) \, x - 
-\dfrac{1}{n_j} \sum_{y \in \C_j} \mathcal{D}_i(y) \, y.
+g(\C, \C) = \dfrac{2}{n^2} \sum_{j=1}^n (2j - 1 - n) x_j .
 \end{equation}
-If we merge $\C_i$ and $\C_j$ and sort the data, this expression 
-can be computed in $O(n)$, where $n=n_i+n_j$.
-For a two-class problem, we have to loop through each point
-and compute \eqref{eq:within} to find the best split of the data. 
-Since sorting
-sorting usually takes $O(n \log n)$, the complexity of
-this procedure is $O(n(\log n + n)) = O(n^2)$.
+Note that the cost of computing this is $O(n)$, once the data
+is sorted which costs at the most $O(n\log n)$.
+Now suppose we order each partition, obtaining $\mathbb{X} = \cup_{j=1}^k
+\bar{\C}_j$. Then \eqref{eq:within} is given by
+\begin{equation}
+\label{eq:w1d}
+W(\{ \bar{\C}_i \}) = 2
+\sum_{j=1}^k \dfrac{1}{n_j} \sum_{\ell=1}^{n_j} (2\ell - 1 - n_j) x_\ell.
+\end{equation}
+For a two-class problem, we can cluster the data by computing \eqref{eq:w1d}
+for each split and pick the one with minimum value. This is done
+as follows:
+\begin{enumerate}
+\item Sort the pooled data, obtaining the sorted set $\bar{\mathbb{X}}$ with
+$n$ elements.
+\item For each $t=1,2,\dotsc,n$, make two partitions 
+$\bar{\C}_1^{(t)} = [x_i]_{i=1}^t$ and 
+$\bar{\C}_2^{(t)} = [x_i]_{i=t+1}^n$. 
+\item Compute $W^{(t)} = W(\{ \bar{\C}_1^{(t)},\bar{\C}_2^{(t)}  \})$ 
+from \eqref{eq:w1d}.
+\item The best split is $t^\star = \argmin_t W^{(t)}$, which
+determines the final clusters.
+\end{enumerate}
+Notice that this procedure does not require any initialization, however,
+it only works for one-dimensional data and for Euclidean distance.
 

### 2017-05-17 qcqp_clustering/code/eclust/energy1d.py

@@ -42,6 +42,11 @@ def within_sample(A):
     """
     return sum([len(a)/2*mean1D(a,a)  for a in A])
 
+def mean2(A):
+    """Another version of the above. Assume A is sorted."""
+    n = len(A)
+    return (2./n**2)*sum([A[i]*(2.*i +1- n) for i in range(n)])
+
 def two_clusters1D(X):
     """Optimize energy statistics for two clusters in 1D."""
     sorted_indices = np.argsort(X)
@@ -71,7 +76,8 @@ if __name__ == "__main__":
     from sklearn.cluster import KMeans
     from sklearn.mixture import GMM
     import matplotlib.pyplot as plt
-    
+   
+    """
     X1 = np.random.normal(0, 1, 2000)
     #X1 = np.random.gamma(2, 2, 1000)
     #X1 = np.random.uniform(0, 1, 1000)
@@ -96,4 +102,15 @@ if __name__ == "__main__":
     gmm.fit(X)
     zh = gmm.predict(X)
     print accuracy(z, zh)
+    """
+    
+    import energy
+    
+    X = np.random.normal(0, 1, 100)
+    
+    print energy.mean(X, X)
+    
+    Y = np.sort(X)
+    print  mean1D(Y, Y)
+    print mean2(Y)
 

### 2017-05-18 qcqp_clustering/draft/energy_clustering.tex
figures draft

@@
 \usepackage[inline]{enumitem}
-\usepackage[pdftex]{graphicx}
-\graphicspath{{./figs/}}
 
@@
 
-\graphicspath{{./figs}}
+\usepackage[pdftex]{graphicx}
+\graphicspath{{./figs/}}
 
@@
 \begin{equation}\label{eq:energy}
-\Energy(P, Q) = 2 \E \| X - Y\| - \| X - X' \| - \| Y - Y' \|,
+\Energy(P, Q) = 2 \E \| X - Y\| - \E \| X - X' \| - \E \| Y - Y' \|,
 \end{equation}
@@
 \begin{equation}\label{eq:energy2}
-\Energy_\alpha(P, Q) = 2 \E \| X - Y\|^{\alpha} - \| X - X' \|^{\alpha} - 
-\| Y - Y' \|^{\alpha},
+\Energy_\alpha(P, Q) = 2 \E \| X - Y\|^{\alpha} - \E \| X - X' \|^{\alpha} - 
+\E \| Y - Y' \|^{\alpha},
 \end{equation}
@@
 \end{cases}
+\nonumber
 \end{equation}
@@
 is one-dimensional data and a two-class problem.
-If data is one-dimensional and we choose
+If we choose
 $\rho(x,y) = |x - y|$, we can actually compute 
@@
 $\bar{\C} = [ x_j \in \C: x_1 \le x_2 \le \dotsm \le x_{n}]$. Thus, we
-can write \eqref{eq:g_ind} as
+can write \eqref{eq:g_ind} in the following very simple form:
 \begin{equation}
 \label{eq:g1d}
-g(\C, \C) = \dfrac{2}{n^2} \sum_{j=1}^n (2j - 1 - n) x_j .
+g(\C, \C) = \dfrac{2}{n^2} \sum_{\ell=1}^n (2\ell - 1 - n) x_\ell .
 \end{equation}
@@
 \label{eq:w1d}
-W(\{ \bar{\C}_i \}) = 2
-\sum_{j=1}^k \dfrac{1}{n_j} \sum_{\ell=1}^{n_j} (2\ell - 1 - n_j) x_\ell.
+W(\{ \bar{\C}_i \}) = 
+\sum_{j=1}^k \sum_{\ell=1}^{n_j} \dfrac{2\ell - 1 - n_j}{n_j} \, x_\ell.
 \end{equation}
 For a two-class problem, we can cluster the data by computing \eqref{eq:w1d}
-for each split and pick the one with minimum value. This is done
+for each possible split, and pick the one with minimum value. This is done
 as follows:
@@
 
+\begin{figure}
+\begin{minipage}{0.33\textwidth}
+\centering
+\includegraphics[width=\textwidth]{gauss_means.pdf}\\[-.8em]
+(a)
+\end{minipage}
+\begin{minipage}{0.33\textwidth}
+\centering
+\includegraphics[width=\textwidth]{gauss_covs.pdf}\\[-.8em]
+(b)
+\end{minipage}
+\begin{minipage}{0.33\textwidth}
+\centering
+\includegraphics[width=\textwidth]{gauss_pis.pdf}\\[-.8em]
+(c)
+\end{minipage}
+\caption{
+\label{fig:dimension}
+gadsga
+}
+\end{figure}
+
+\begin{figure}
+\centering
+\caption{
+\label{fig:unbalanced}
+gadsga
+}
+\end{figure}
+
+\begin{figure}
+\begin{minipage}{0.33\textwidth}
+\hspace{.7cm}\includegraphics[width=.82\textwidth]{cigar_data.pdf}
+\end{minipage}
+\begin{minipage}{0.33\textwidth}
+\hspace{.7cm}\includegraphics[width=.82\textwidth]{circles_data.pdf}
+\end{minipage}
+\begin{minipage}{0.33\textwidth}
+\hspace{.7cm}\includegraphics[width=.82\textwidth]{spirals_data.pdf}
+\end{minipage}
+\begin{minipage}{0.33\textwidth}
+\centering
+\includegraphics[width=\textwidth]{cigars.pdf}\\[-.8em]
+(a)
+\end{minipage}
+\begin{minipage}{0.33\textwidth}
+\centering
+\includegraphics[width=\textwidth]{circles.pdf}\\[-.8em]
+(b)
+\end{minipage}
+\begin{minipage}{0.33\textwidth}
+\centering
+\includegraphics[width=\textwidth]{spirals.pdf}\\[-.8em]
+(c)
+\end{minipage}
+\caption{
+\label{fig:dimension}
+gadsga
+}
+\end{figure}
+
 \section{Conclusion}

### 2017-05-18 qcqp_clustering/code/eclust/data.py

@@ -9,8 +9,31 @@ from __future__ import division
 import numpy as np
 import matplotlib.pyplot as plt
 import numbers
+import pylab
 
 from sklearn.decomposition import PCA
+
+
+# pyplot customization
+pylab.rc('lines', linewidth=.5, antialiased=True, markeredgewidth=0.1)
+pylab.rc('font', family='computer modern roman', style='normal',
+         weight='normal', serif='computer modern sans serif', size=10)
+pylab.rc('text', usetex=True)
+pylab.rc('text.latex', preamble=[
+        '\usepackage{amsmath,amsfonts,amssymb,relsize,cancel}'])
+pylab.rc('axes', linewidth=0.5, labelsize=10)
+pylab.rc('xtick', labelsize=10)
+pylab.rc('ytick', labelsize=10)
+#pylab.rc('legend', numpoints=1, fontsize=10, handlelength=0.5)
+pylab.rc('legend', numpoints=1, fontsize=10)
+fig_width_pt = 455.0 / 1.5 # take this from LaTeX \textwidth in points
+inches_per_pt = 1.0/72.27
+golden_mean = (np.sqrt(5.0)-1.0)/2.0
+fig_width = fig_width_pt*inches_per_pt
+fig_height = fig_width_pt*inches_per_pt
+#fig_height = fig_width*golden_mean
+pylab.rc('figure', figsize=(fig_width, fig_height))
+
 
 
 def multivariate_normal(means, sigmas, ns):
@@ -126,7 +149,7 @@ def spirals(rs, ns, noise=0.2):
     idx = np.random.permutation(sum(ns))
     return X[idx], z[idx]
 
-def plot(X, z, fname='plot.pdf'):
+def plot(X, z, colors=['b', 'r', 'g'], fname='plot.pdf'):
     """Plot data according to labels in z.
     If data is multidimensional, pick the first two principal components.
     
@@ -142,16 +165,21 @@ def plot(X, z, fname='plot.pdf'):
 
     fig = plt.figure()
     ax = fig.add_subplot(111)
-    colors = iter(plt.cm.rainbow(np.linspace(0,1,len(z_unique))))
+    #colors = iter(plt.cm.rainbow(np.linspace(0,1,len(z_unique))))
+    colors = iter(colors)
+    
     for k in z_unique:
         idx = np.where(z==k)
         x = X_new[idx][:,0]
[…]

### 2017-05-18 qcqp_clustering/code/eclust/kenergy.py

@@ -1,0 +1,275 @@
+"""Iterative algorithm to optimize QCQP for energy statistics clustering."""
+
+# Guilherme Franca <guifranca@gmail.com>
+# Johns Hopkins University, Neurodata
+
+
+from __future__ import division
+
+import numpy as np
+from sklearn.metrics.pairwise import pairwise_distances
+from sklearn.base import BaseEstimator, ClusterMixin
+from sklearn.utils import check_random_state
+
+import kmeanspp
+
+
+def ztoZ(z):
+    """Convert label vector to label matrix."""
+    n = z.shape[0]
+    k = np.unique(z).shape[0]
+    Z = np.zeros((n,k))
+    for i in range(n):
+        Z[i, int(z[i])] = 1
+    return Z
+
+def Ztoz(Z):
+    """Convert label matrix to label vector."""
+    n, k = Z.shape
+    z = np.zeros(n)
+    points, labels = np.where(Z==1)
+    for i in range(len(points)):
+        z[points[i]] = labels[i]
+    return z
+
+def objective(Z, G):
+    """Compute objective function."""
+    Zt = Z.T
+    D = np.dot(Zt, Z)
+    Dinv = np.linalg.inv(D)
+    return np.trace(Dinv.dot(Zt.dot(G.dot(Z))))
+
+def euclidean_rho(x, y, alpha=2):
+    return np.power(np.linalg.norm(x-y), alpha)
+
+def kernel_function(x, y, rho, x0=None):
+    """Return kernel function based on rho."""
+    if x0 == None:
+        x0 = np.zeros(x.shape)
+    return 0.5*(rho(x,x0) + rho(y,x0) - rho(x,y))
+
+def kernel_matrix(X, rho, x0=None, diag=True):
+    """Compute Kernel matrix based on kernel function K(x,y)."""
+    kfunc = lambda x, y: kernel_function(x, y, rho, x0)
+    G = pairwise_distances(X, metric=kfunc, n_jobs=4)
+    if not diag:
+        i = range(len(X))
+        G[i,i] = 0
+    return G
+
+def kenergy_brute(k, G, Z0, max_iter=50):
+    """Optimize energy QCQP in a brute force way. Kind of expensive,
+    but actually works slightly better than Kernel Kmeans, at least
+    it seems like that.
+    """
+    
+    n = G.shape[0]
+    Z = Z0
+    count = 0
+    converged = False
+    while not converged and count < max_iter:
+        converged = True
[…]

### 2017-05-18 qcqp_clustering/code/eclust/kmeans.py

@@ -12,15 +12,20 @@ import kmeanspp
 def euclidean(a, b):
     return np.linalg.norm(a-b)**2
 
-def kmeans(K, X, distance=euclidean, max_iter=50):
+def kmeans(K, X, distance=euclidean, max_iter=50, labels_=False, mus_=False):
     """K is the number of clusters. X is the dataset, which is a matrix
     with one data point per row. distance is a function
     which accepts two data points and returns a real number.
     
     """
     N = X.shape[0] # number of points
-    mus = kmeanspp.kpp(K, X, ret='centers')
-    labels = np.empty(N, dtype=np.int8)
+
+    if not (labels_.any() and  mus_.any()):
+        mus = kmeanspp.kpp(K, X, ret='centers')
+        labels = np.zeros(N)
+    else:
+        labels = labels_
+        mus = mus_
     
     converged = False
     count = 0

### 2017-05-18 qcqp_clustering/code/eclust/metric.py

@@ -24,10 +24,10 @@ def accuracy(z, zh):
     
     Q = np.zeros((n,k))
     for i in range(n):
-        Q[i, z[i]] = 1
+        Q[i, int(z[i])] = 1
     Qh = np.zeros((n,k))
     for i in range(m):
-        Qh[i, zh[i]] = 1
+        Qh[i, int(zh[i])] = 1
     
     cost_matrix = Qh.T.dot(Q)
     row_ind, col_ind = scipy.optimize.linear_sum_assignment(-cost_matrix)

### 2017-06-21 qcqp_clustering/draft/energy_clustering.tex
working on draft

@@
-%\documentclass[aps,preprint,nofootinbib,floatfix]{revtex4-1}
-\documentclass{article}
+\documentclass[aps,preprint,nofootinbib,floatfix]{revtex4-1}
+%\documentclass{article}
 
@@
 % to compile a camera-ready version, add the [final] option, e.g.:
-\usepackage[final]{nips_2017}
+%\usepackage[final]{nips_2017}
 
-\usepackage{nicefrac}       % compact symbols for 1/2, etc.
-\usepackage{microtype}      % microtypography
+%\usepackage{nicefrac}       % compact symbols for 1/2, etc.
+%\usepackage{microtype}      % microtypography
 \usepackage{amsmath,amssymb,amsfonts,amsthm,amscd,bm,bbm}
-\usepackage{algorithm}
-\usepackage{algorithmic}
-\usepackage[inline]{enumitem}
-
+%\usepackage{mathtools}
+%\usepackage{printlen}
+%\usepackage[inline]{enumitem}
 %\usepackage{cite}
-\usepackage{bbm}
-\usepackage{algorithm}
-\usepackage{algorithmic}
+%\usepackage{bbm}
 
@@
 \graphicspath{{./figs/}}
+
+\usepackage{algorithmic}
+\usepackage{algorithm}
 
@@
 \newcommand\C{{\mathcal{C}}}
+\newcommand\OO{{\mathcal{O}}}
 %\newcommand\Zt{\widetilde{Z}}
 \newcommand\Zt{Y}
-\newcommand{\Ind}[1]{\delta_{#1}}
-
-\title{Nonparametric Clustering Based on Energy Statistics}
-
-\author{
-Guilherme Fran\c ca \\
-Johns Hopkins University\\
-\texttt{guifranca@gmail.com} \\
-%% examples of more authors
-\And
-Joshua T. Vogelstein \\
-Johns Hopkins University\\
-\texttt{jovo@jhu.edu}
-}
-
-%\author{Guilherme Fran\c ca}
-%\email{guifranca@gmail.com}
-%\affiliation{Johns Hopkins University, Center for Imaging Science}
-
-%\author{Joshua Vogelstein}
-%\email{jovo@jhu.edu}
-%\affiliation{Johns Hopkins University, Center for Imaging Science}
+%\newcommand{\Ind}[1]{\delta_{#1}}
+\newcommand{\Ind}[1]{\mathbbm{1}_{#1}}
 
@@
 
-\maketitle
+\title{Nonparametric Clustering from  Energy Statistics}
+
+\author{Guilherme Fran\c ca}
+\email{guifranca@gmail.com} 
+%\affiliation{Johns Hopkins University}
+
+\author{Joshua T. Vogelstein}
+\email{jovo@jhu.edu}
+
+\affiliation{Johns Hopkins University}
 
 \begin{abstract}
-blabla
+Energy statistics provides a nonparametric test for equality of distributions.
+It was proposed by 
+Sz\' ekely in the 80's,
+inspired by Newton's potential energy between massive bodies. The idea
+is to associate a statistical potential energy to observations such that 
+minimum energy is achieved under the null hypothesis of equality of 
+distributions. This 
+was further generalized for probability 
+distributions on arbitrary metric spaces,
+and more recently, a connection with kernels in RKHS was established.
+Nevertheless, although extensively used by the statistics community, it was
+not incorporated in machine learning problems.
+In this paper, we consider the problem of clustering data based
+energy statistics theory. 
+We provide a precise mathematical formulation, obtaining
+a quadratically constrained optimization problem (QCQP). We show that
+this optimization problem is equivalent to kernel $k$-means
+optimization problem, however,
+energy statistics is able to fix the kernel choice. 
+This method is nonparametric, and if prior information is available
+it can be easily incorporated in the kernel construction.
+We propose an algorithm
+to find local optimizers of this QCQP problem based on the energy cost
+of moving points to different clusters. We then compare this algorithm with
+kernel $k$-means, standard $k$-means, and GMM. We test our method
+on synthetic and real data experiments.
+Our results show that energy statistics based clustering outperforms
+these most used clustering algorithms.
 \end{abstract}
+
+\maketitle
 
@@
 statistics and its relation to reproducing kernel Hilbert spaces 
-(RKHS), which form the basis of our work.
+(RKHS) which form the basis of our work.
 For more details we refer the reader
-to 
-\cite{Szkely2013} 
-(and references therein) and also \cite{Sejdinovic2013}.
+to \cite{Szkely2013} (and references therein) and 
+also \cite{Sejdinovic2013}.
 
@@
 \begin{equation}\label{eq:energy}
-\Energy(P, Q) = 2 \E \| X - Y\| - \E \| X - X' \| - \E \| Y - Y' \|,
+\Energy(P, Q) \equiv 2 \E \| X - Y\| - \E \| X - X' \| - \E \| Y - Y' \|,
 \end{equation}
@@
 to zero holds if and only if $P = Q$.
-Above, $\| \cdot \|$ denotes the
+Above $\| \cdot \|$ denotes the
 Euclidean norm in $\mathbb{R}^D$. 
 Energy distance
-provides a characterization of equality of distributions, and
+provides a characterization of equality of distributions and
 $\Energy^{1/2}$ is
@@
 \begin{equation}\label{eq:energy2}
-\Energy_\alpha(P, Q) = 2 \E \| X - Y\|^{\alpha} - \E \| X - X' \|^{\alpha} - 
+\Energy_\alpha(P, Q) \equiv 
+2 \E \| X - Y\|^{\alpha} - \E \| X - X' \|^{\alpha} - 
 \E \| Y - Y' \|^{\alpha},
@@
 where $0<\alpha\le 2$. This quantity is also nonnegative,
-$\Energy_\alpha(P,Q) \ge 0$. Furthermore, for $0<\alpha<2$ we have that
+$\Energy_\alpha(P,Q) \ge 0$. Furthermore, for $0<\alpha<2$ we have 
 $\Energy_\alpha(P,Q) = 0$ if and only if $P=Q$, while for $\alpha=2$ 
-we have $\Energy_2(P,Q) = 2\| \E(X) - \E(Y) \|^2$, showing that
+we have $\Energy_2(P,Q) = 2\| \E(X) - \E(Y) \|^2$ showing that
 equality to zero only requires
-equality of the means, and thus it does not imply equality of distributions.
+equality of the means and thus $\Energy_2(P,Q)=0$ does 
+not imply equality of distributions.
 
@@
 mention that \eqref{eq:energy2} can be even further generalized.
-Let $X, Y \in \mathcal{X}$ and replace the Euclidean norm by
-$\rho: \mathcal{X}\times\mathcal{X} \to \mathbb{R}$, i.e.
-$\| X - Y\| \to \rho(X,Y)$, where $\rho$ is a so-called \emph{semimetric
-of negative type},
-which satisfy 
+Let $X, Y \in \mathcal{X}$,  where $\mathcal{X}$ is a space endowed
+with a \emph{semimetric of negative type}
+$\rho: \mathcal{X}\times\mathcal{X} \to \mathbb{R}$, which satisfies
 \begin{equation}
@@
 \end{equation}
-where $X_i \in \mathcal{X}$, and $\alpha_i \in \mathbb{R}$ such that
-$\sum_{i=1}^n \alpha_i = 0$.
-In this case, there is a Hilbert space $\mathcal{H}$ and
+where $X_i \in \mathcal{X}$ and $\alpha_i \in \mathbb{R}$ such that
+$\sum_{i=1}^n \alpha_i = 0$. Then $\mathcal{X}$ is called a space of
+negative type.
+We can thus replace $\mathbb{R}^D \to \mathcal{X}$ and 
+$\| X - Y \| \to \rho(X , Y)$ in the definition \eqref{eq:energy} obtaining
+the energy distance as
+\begin{equation}\label{eq:energy3}
+\Energy(P, Q) \equiv 2 \E \rho(X,Y) - \E \rho(X, X') - \E \rho(Y,Y').
+\end{equation}
+For spaces of negative type exists a Hilbert space $\mathcal{H}$ and
 a map $\varphi: \mathcal{X} \to
@@
 This equivalence was established
-in \cite{Sejdinovic2013}. Let us first recall the definition of
+in \cite{Sejdinovic2013}. We first recall the definition of
 RKHS. Let $\HH$ be a Hilbert space of real-valued functions
@@
 \item $\kkk_x \equiv \kk(\cdot, x) \in \HH$ 
-for any $x \in \mathcal{X}$.
+for all $x \in \mathcal{X}$.
 \item $\langle \kkk_x, f \rangle_{\HH} = f(x)$ for
-any $x\in\mathcal{X}$ and $f\in \HH$.
+all $x\in\mathcal{X}$ and $f\in \HH$.
 \end{enumerate}
-In other words, for any $x \in \mathcal{X}$ there is a unique function
+In other words, for any $x \in \mathcal{X}$ and any function $f \in \HH$ 
+there is a unique 
 $\kkk_x \in \HH$ that reproduces $f(x)$ through the inner product
@@
 If such a \emph{kernel} 
-function $\kk$ exists, then $\HH$ is called a RKHS.
-From this we have $\langle \kkk_x, \kkk_y \rangle = \kkk_y(x) = \kk(x,y)$. 
-This implies
-that $\kk(x,y)$ is symmetric and positive semi-definite, 
-$\sum_{i,j=1}^n c_i c_j
-\kk(x_i,x_j) \ge 0$ for $c_i,c_j \in \mathbb{R}$.
+function $\kk$ exists then $\HH$ is called a RKHS. The above two 
+properties immediately imply that $\kk$ is symmetric and positive
+definite. Indeed, notice that
+$\langle \kkk_x, \kkk_y \rangle = \kkk_y(x) = \kk(x,y)$, and since
+this inner product is real,
+$\langle \kkk_x, \kkk_y \rangle^* = \langle \kkk_y, \kkk_x \rangle = 
+\langle \kkk_x, \kkk_y \rangle$, we immediately have that
+the kernel is symmetric,
+$\kk(y,x) = \kk(y,x)$. Moreover, for any $w \in
+\HH$ we can write $w = \sum_{i=1}^n c_i \kkk_{x_i}$, where
+$\{ \kkk_{x_i} \}_{i=1}^n$ is a basis of $\HH$. It follows that
+$\langle w, w \rangle_{\HH}  = \sum_{i,j=1}^n c_i c_j \kk(x_i,x_j) \ge 0$
+showing that the kernel is positive definite. If $G$ is a matrix with
+elements $G_{ij} = \kk(x_i,x_j)$, this is equivalent to $G$ being
+positive semi-definite, $\bm{w}^\top G \bm{w} \ge 0$ for any vector
+$\bm{w} \in \mathbb{R}^n$.
 
@@
 $\int f(x) d P(x) = \langle f, \kkk_P \rangle$ for all $f \in \Hk$,
-or alternatively $\kkk_P = \int \kk(\cdot, x)  d P(x)$. 
+or alternatively, $\kkk_P = \int \kk( \, \cdot \,, x)  d P(x)$. 
 We can now  introduce the 
@@
 \begin{equation}\label{eq:mmd}
-\gamma_\kk(P,Q) = \| \kkk_P - \kkk_Q \|_{\Hk},
+\gamma_\kk(P,Q) \equiv \| \kkk_P - \kkk_Q \|_{\Hk},
 \end{equation}
@@
 The following important result shows that semimetrics of negative
-type and symmetric positive semi-definite kernels are closely related
-\cite{Berg1984}. Let $\rho: \mathcal{X} \times \mathcal{X} \to \mathbb{R}$
-be a semimetric, 
+type and symmetric positive semidefinite kernels are closely related
+\cite{Berg1984}. Let $\rho: \mathcal{X} \times \mathcal{X} \to \mathbb{R}$,
 and $x_0 \in \mathcal{X}$ an arbitrary but fixed point.
@@
 \end{equation}
-Then, $\kk$ is positive semi-definite if and only if $\rho$ is of negative type
+Thus, $\kk$ is positive semidefinite if and only if $\rho$ is a semimetric
+of negative type
 \eqref{eq:negative_type}.
@@
 \label{eq:gen_kernel}
-\rho(x,y) = \kk(x,x) + \kk(y,y) -2\kk(x,y) = \| \kkk_x - \kkk_y
-\|^2_{\Hk},
+\begin{split}
+\rho(x,y) &= \kk(x,x) + \kk(y,y) -2\kk(x,y) \\
+&=  \| \kkk_x - \kkk_y \|^2_{\Hk},
+\end{split}
 \end{equation}
@@
 We say that the kernel $\kk$ generates the semimetric $\rho$. 
-If two different kernels generate the same $\rho$, they are
+If two different kernels generate the same $\rho$ they are
 equivalent kernels.
@@
 of negative type and $\kk$ a kernel that generates $\rho$, then
-\begin{align}
-\Energy(P, Q) &\equiv 2\E \, \rho(X,Y) - \E \, \rho(X,X') - \E \, \rho(Y,Y') 
-\label{eq:EphoDef}\\
-&= 2 \left[ \E \, \kk(X, X') + \E \, \kk(Y, Y') - 2\E \, \kk(X, Y)\right] \\
-&=2 \gamma_\kk^2(P,Q) .
-\end{align}
-This result follows simply by substituting \eqref{eq:gen_kernel} into
-\eqref{eq:EphoDef}, and using \eqref{eq:mmd2}.
+replacing \eqref{eq:gen_kernel} into
+\eqref{eq:energy3} and using \eqref{eq:mmd2} yields
+\begin{equation} \label{eq:Erho}
+\Energy(P, Q) = 
+2 \left[ \E \, \kk(X, X') + \E \, \kk(Y, Y') - 2\E \, \kk(X, Y)\right] 
+= 2 \gamma_\kk^2(P,Q) .
+\end{equation}
 Since $\gamma_k^2(P, Q) = \| \kkk_P - \kkk_Q \|^2_{\Hk}$ we
@@
 Finally, let us recall the main formulas for test statistics
-\cite{Szkely2013}. 
-Assume we have data $\mathbb{X} = \{ x_1,x_2,\dotsc, x_n \}$, where
-$x_i \in \mathcal{X}$ is in some space of 
-negative type \eqref{eq:negative_type},
-and a partition $\mathbb{X} = \cup_{j=1}^k \C_j$ where
+of equality of distributions \cite{Szkely2013}. 
+Assume we have data $\mathbb{X} = \{ x_1,\dotsc, x_n \}$, where
+$x_i \in \mathcal{X}$ and $\mathcal{X}$ is a space of negative type.
+Consider a partition $\mathbb{X} = \bigcup_{j=1}^k \C_j$, with
 $\C_i \cap \C_j = \emptyset$.
 Each expectation in 
-\eqref{eq:EphoDef}
+\eqref{eq:energy3}
 can be computed 
@@
 S \equiv
-\sum_{1 \le  j < l \le k } \dfrac{n_j n_{l}}{2 n} \left[
-2 g(\C_j, \C_l) - 
-g(\C_j, \C_j) - 
-g(\C_l, \C_l)
+\sum_{1 \le  i < j \le k } \dfrac{n_i n_{j}}{2 n} \left[
+2 g(\C_i, \C_j) - 
+g(\C_i, \C_i) - 
+g(\C_j, \C_j)
 \right].
@@
 Given a set of distributions
-$\{ P_j\}_{j=1}^k$ where $x \in \C_j \sim P_j$, the quantity
-$S$ provides
-a \emph{nonparametric} test statistic for equality of distributions.
-Under the null hypothesis $H_0: P_1=P_2=\dotsm=P_k$, $S$ is small, 
+$\{ P_j\}_{j=1}^k$, where $x \in \C_j$ if and only if $x \sim P_j$, 
+the quantity $S$ provides
+a \emph{nonparametric} test statistic for equality of distributions
+\cite{Szkely2013}.
+When the sample size is large enough, $n\to \infty$,
+under the null hypothesis $H_0: P_1=P_2=\dotsm=P_k$ we have that
+$S\to 0$, 
 and under
 the alternative hypothesis $H_1: P_i \ne P_j$ for at least two $i\ne j$, 
-$S \to \infty$ as the sample size is large, $n \to \infty$.
+we have that $S \to \infty$.
 This test is nonparametric in the sense that it does not make any assumptions
-about the distributions $P_j$.
+about the distributions $P_j$. Next we show how we can use $S$ for clustering
+data.
 
@@
 This section contains the main results of this paper, where 
-we propose an optimization problem for clustering data
+we formulate an optimization problem for clustering 
 based on energy statistics and RKHS introduced in the previous section.
 
-Based on the test statistic \eqref{eq:between}, the obvious
+Due to the test statistic \eqref{eq:between} for equality of distributions,
+the obvious
 criterion for clustering data is to 
-maximize $S$, which intuitivelly makes the distribution of each
-cluster $\C_j$ as different
-as possible from the other ones. However, the following 
-simple result
-shows that this is equivalent to minimizing
+maximize $S$, which intuitively makes 
+each cluster as different
+as possible from the other ones.
+In other words, given a set of points coming from different probability
+distributions, $S$ should attain a maximum when the points are correctly
+classified since $S$ measures a cost between different clusters.
+However, the following 
+straightforward result
+shows that maximizing \eqref{eq:between} is equivalent to minimizing
 \eqref{eq:within}, which has a more convenient form.
@@
 \label{th:minimize}
-Let $\mathbb{X} = \{x_i\}_{i=1}^{n}$ where 
-$x_i$ lives in a space $\mathcal{X}$ endowed with a semi-metric $\rho$ of
-negative type. For a fixed integer $k$,
+Let $\mathbb{X} = \{x_1,\dotsc,x_n\}$ where each data point
+$x_i$ lives in a space $\mathcal{X}$ endowed with a semimetric $\rho:
+\mathcal{X}\times\mathcal{X} \to \mathbb{R}$ of
+negative type \eqref{eq:negative_type}. For a fixed integer $k$,
 the partition
-$\mathbb{X} = \cup_{j=1}^k \C_j$, where $\C_i \cap C_j = \emptyset$ for
-all $i\ne j$, maximizes $S$ if and only if
+$\mathbb{X} = \bigcup_{j=1}^k \C_j$, where $\C_i \cap C_j = \emptyset$ for
+all $i\ne j$, maximizes \eqref{eq:between} if and only if
 \begin{equation}
 \label{eq:minimize}
-\min_{\{ \C_j \} } W(\{ 
-\C_1, \dotsc, \C_k
-\}).
+\min_{\C_1,\dotsc,C_k  } W(
+\C_1, \dotsc, \C_k),
 \end{equation}
+where $W$ is given by \eqref{eq:within}.
 \end{proposition}
@@
 \end{equation}
-Note, however, that the right hand side of this equation 
+Note that the right hand side of this equation 
 only depends on the pooled data, so it is a constant
-independent on the choice of partition. Therefore, maximizing
+independent of the choice of partition. Therefore, maximizing
 $S$ is equivalent to minimizing~$W$.
@@
 
-Thus, given $k$, our clustering problem ammounts to
-finding the best partition of the data through solving~\eqref{eq:minimize}.
-In this setting, each datapoint belongs to one and only one cluster
-(hard assignments).
+Thus, for a given $k$, the clustering problem amounts to
+finding the best partition of the data by solving~\eqref{eq:minimize}.
+Notice that this is a hard assignment clustering problem.
 
+Now we show how to formulate problem \eqref{eq:minimize} in the corresponding
+RKHS.
 Based on 
@@
 generates $\rho$. 
-Let us define  the kernel matrix
-$G \in \mathbb{R}^{n\times n}$ such that
+Let us define  the Gram matrix
 \begin{equation}
 \label{eq:kernel_matrix}
-G_{ij} \equiv \kk(x_i, x_j).
+G \equiv \begin{pmatrix}
+\kk(x_1,x_1) & \kk(x_1,x_2) & \dotsm & \kk(x_1,x_n) \\
+\kk(x_2,x_1) & \kk(x_2,x_2) & \dotsm & \kk(x_2,x_n) \\
+\vdots & \vdots & \ddots  & \vdots \\
+\kk(x_n,x_1) & \kk(x_n,x_2) & \dotsm & \kk(x_n,x_n) 
+\end{pmatrix} .
 \end{equation}
-Let $Z$ be the label matrix, $Z \in \{ 0,1 \}^{n\times k}$, 
-with only one nonvanishing entry per row indicating to which cluster (columns)
-each point (rows) belongs to. This matrix satisfy
+Let $Z \in \{ 0,1 \}^{n\times k}$ be the label matrix, 
+with only one nonvanishing entry per row, 
+indicating to which cluster (column)
+each point (row) belongs to. This matrix satisfy
 $Z^\top Z = D$ where $D = \diag( n_1,\dotsc, n_k )$  contains
@@
 \end{equation}
+Throughout the paper, we use the notation $M_{i\bullet}$ to denote
+the $i$th row of a matrix $M$, and $M_{\bullet j}$ denotes its $j$th column.
 
 Our next result reveals the optimization problem behind \eqref{eq:minimize},
-which is in general NP-hard since
+which is NP-hard since
 it is a quadratically constrained quadratic problem (QCQP).
@@
 \label{eq:qcqp2}
-\max_{\Zt} \Tr \left\{ \Zt^\top G \Zt \right\}  \qquad
+\max_{\Zt} \Tr \left( \Zt^\top G \, \Zt \right)  \qquad
 \mbox{s.t. $\Zt \ge 0$, $\Zt^\top \Zt = I$, 
@@
 where $\bm{e} = (1,1,\dots,1)^\top \in \mathbb{R}^n$ is the all-ones vector,
-and $G$ is the pairwise kernel matrix \eqref{eq:kernel_matrix}, assumed
-to be given.
+and $G$ is the kernel matrix \eqref{eq:kernel_matrix}.
 \end{proposition}
 \begin{proof}
-Replacing \eqref{eq:gen_kernel}  into
-\eqref{eq:g_def},  
-we can write \eqref{eq:within} as
+From 
+\eqref{eq:gen_kernel},
+\eqref{eq:g_def}, and
+\eqref{eq:within}
+we have
 \begin{equation}
 \label{eq:W2}
-W(\{ \C_j \} )
+W(\C_1,\dotsc,\C_k  )
 = \dfrac{1}{2} \sum_{j=1}^k \dfrac{1}{n_j} \sum_{x,y \in \C_j} \rho(x,y)
@@
 \end{equation}
-The first term is a global constant 
-which does not contribute,
-so minimizing \eqref{eq:W2} is the same as
+Note that the first term does not contribute to the optimization problem,
+since it is a global term that does not depend which partition is chosen. 
+Therefore, minimizing \eqref{eq:W2} is equivalent to
 \begin{equation}
 \label{eq:max_prob}
-\max_{ \{ \C_j \} } \sum_{j=1}^k \dfrac{1}{n_j} \sum_{x,y\in C_j} \kk(x,y) .
+\max_{ \C_1,\dotsc,\C_k } 
+\sum_{j=1}^k \dfrac{1}{n_j} \sum_{x,y\in C_j} \kk(x,y) .
 \end{equation}
@@
 \sum_{x, y \in \C_j} \kk(x, y) =
-\sum_{l=1}^{n} \sum_{m=1}^{n} Z_{lj} Z_{mj} G_{lm} = 
-(Z^\top G Z)_{jj}
+\sum_{p=1}^{n} \sum_{q=1}^{n} Z_{pj} Z_{qj} G_{pq} = 
+(Z^\top G \, Z)_{jj},
 \end{equation}
 where we used the definitions \eqref{eq:kernel_matrix} and
-\eqref{eq:label_matrix}. Thus \eqref{eq:max_prob} is equal to
-$\max \Tr \left\{ D^{-1} Z^\top G Z \right\}$. Now we can
+\eqref{eq:label_matrix}. Thus the objective function in 
+\eqref{eq:max_prob} is equal to
+$\Tr \left( D^{-1} Z^\top G Z \right)$. Now we can
 use the cyclic property
@@
 \begin{equation}\label{eq:qcqp}
-\begin{aligned}
-& \max_{Z} \Tr\left\{ \big( Z D^{-1/2}\big)^\top G 
+\max_{Z} \Tr\Big( \big( Z D^{-1/2}\big)^\top G 
 \big( ZD^{-1/2} \big) 
-\right\} \\
-&\mbox{s.t. $Z_{ij} \in \{0,1\}$, $\sum_{j=1}^k Z_{ij} = 1$, 
+\Big) \quad
+\mbox{s.t. $Z_{ij} \in \{0,1\}$, $\sum_{j=1}^k Z_{ij} = 1$, 
 $\sum_{i=1}^n Z_{ij} = n_j$.}
-\end{aligned}
 \end{equation}
-Now we write this in terms of $Y = Z D^{-1/2}$, given by
-\eqref{eq:label_matrix}. The objective function immediately becomes
-$\Tr Y^\top G Y$. Notice that the above constraints
-immediatly imply that $Z^T Z = D$, which in turn gives
+Now we write this in terms of the matrix $Y = Z D^{-1/2}$.
+The objective function immediately becomes
+$\Tr\left( Y^\top G \, Y\right)$. Notice that the above constraints
+imply that $Z^T Z = D$, where $D=\diag(n_1,\dotsc,n_k)$, which in turn gives
 $D^{-1/2} Y^T Y D^{-1/2} = D$, or $Y^\top Y = I$. 
@@
 constraint in \eqref{eq:qcqp}. In matrix form this reads
-$Z^T \bm{e} = D \bm{e}$. Replacing by $Z=YD^{1/2}$ we have
+$Z^T \bm{e} = D \bm{e}$. Replacing $Z=YD^{1/2}$ we have
 $Y^\top \bm{e} = D^{1/2} \bm{e}$. Multiplying this last equation
@@
 that $Y D^{1/2} \bm{e} = Z \bm{e} = \bm{e}$, we finally obtain
-$Y Y^\top \bm{e} = \bm{e}$ as desired.
+$Y Y^\top \bm{e} = \bm{e}$. Thus, the optimization 
+problem \eqref{eq:qcqp} is equivalent
+to \eqref{eq:qcqp2} .
 \end{proof}
 
-Therefore, to cluster data $\{ x_i \}_{i=1}^n \in \mathcal{X}$ into 
-$k$ partitions,
-assuming that $k$ is given, we first compute the kernel matrix
-$G$  --- 
-which is  defined by an arbitrary semimetric of negative 
-type on $\mathcal{X}$  --- and then 
+Based on Proposition~\ref{th:qcqp2}, to group data $\{ x_1,\dotsc,x_n \}$
+into  $k$ clusters, we first compute the Gram matrix
+$G$ and then 
 solve the optimization problem \eqref{eq:qcqp2} for $\Zt \in
@@
 indicating that $x_i \in \C_j$. 
+Problem \eqref{eq:qcqp2} is NP-hard and there
+are few methods
+available to solve it directly,
+which is computational prohibitive even for small datasets.
+However, one can find approximate solutions by relaxing some 
+of the constraints, or obtaining a relaxed SDP version of the problem. 
+For instance, the relaxed problem
+\begin{equation}
+\max_{Y} \Tr \left( Y^\top G \, Y \right) \quad \mbox{s.t. $Y^\top Y = I$}
+\end{equation}
+has a well-known closed form solution given by $Y^\star = U R$, where the
+columns of $U$ contain the leading $k$ eigenvectors of $G$ corresponding
+to the $k$ largest eigenvalues $\{ \lambda_1,\dotsc,\lambda_k \}$, and
+$R \in \mathbb{R}^{k\times k}$ is an arbitrary orthogonal matrix. 
+The resulting
+optimal objective function is thus given by
+$\max \Tr \left( {Y^\star}^\top G \, Y^\star \right)  = 
+\sum_{i=1}^k \lambda_i$. One might then normalize and threshold the rows
+of $Y^\star$, or even better, apply standard $k$-means to the rows of
+$Y$ to provide the final clusters. This procedure is known as spectral
+clustering. Nonetheless, computing eigenvectors of a very large matrix
+can be problematic, and usually iterative methods are preferred.
 
-In general, 
-problem \eqref{eq:qcqp2} is NP-hard and there
-are few methods
-available to tackle this kind of problem directly,
-which is computational prohibitive even for relatively small datasets.
-However, one can find an approximate solution by relaxing some 
-of the constraints, or obtaining a relaxed SDP version of it. 
-For instance, a simple approximation can be given
-by 
-$\max_{Y} \Tr Y^\top G \, Y $ subject to $Y^\top Y = I$,
-and requiring that the rows of $Y$ are normalized. 
-It is possible to find a global solution to this problem by
-choosing $Y$ as the top $k$ eigenvectors of $G$, which results
-in 
-$\max \Tr \left\{ Y^\top G \, Y \right\}  = \sum_{i=1}^k \lambda_i(G)$, 
-which is the
-sum of the top $k$ eigenvalues of $G$. This is what is usually done
-in spectral clustering. However, we will propose a simple iterative
-algorithm in the next section.
+It is important to note 
+that clustering based on energy statistics
+holds for data living in an \emph{arbitrary space of negative type}.
+This clustering method is
+\emph{nonparametric} since it does not make any assumptions
+about 
+the distribution of the data,
+contrary to $k$-means and gaussian mixture models (GMM), for example.
+Moreover, this approach \emph{does not require} the concept of the 
+\emph{cluster mean}
+which can be ill-defined for some types of data, such as images for
+instance. 
+Thus, this method is quite general and makes very few
+assumptions about the data.
 
-An interesting point is that \eqref{eq:qcqp2} has the same
-formulation as kernel $k$-means, spectral clustering, and the maximum
-cut problem on graphs \cite{Dhillon}. 
-Proposition~\ref{th:qcqp2} brings energy statistics based clustering
-into this broad picture, and \eqref{eq:qcqp2} should have 
-interesting applications in
-graph partitioning problems and unsupervised learning in general.
-Notice that it is straightforward to generalize \eqref{eq:qcqp2} to
-a weighted version, where weights $W = \diag(w(x_1), \dotsc, w(x_n))^\top $
-are associated to each data point $x_i$. One just replace the objective
-function in \eqref{eq:qcqp2} by 
+Although \eqref{eq:qcqp2} is nonparametric, in practice,
+the clustering quality strongly depend on the choice of a suitable
+$\rho$ which is what measures the similarity between different data points,
+and is equivalent to choosing an appropriate kernel.
+Nevertheless, if prior knowledge is available for choosing $\rho$, 
+or equivalently $K$,
+this can easily be taken into account.
+
+Now let us consider the relation to the well-known kernel $k$-means 
+optimization problem.
+For positive semidefinite $G$, there exists a map
+$\varphi: \mathcal{X} \to \HH_\kk$ such that
+$\kk(x,y) = \varphi(x)^\top \varphi(y)$. The kernel $k$-means problem,
+in the feature space,
+is given by
 \begin{equation}
-\max_Y \Tr \left\{ Y^\top W^{1/2} G W^{1/2} Y \right\}.
+\label{eq:kernel_kmeans}
+\min_{\C_1,\dotsc,\C_k}\bigg\{ 
+J(\C_1,\dots,\C_k) \equiv \dfrac{1}{2} \sum_{j=1}^k
+\sum_{x \in \C_j} \| \varphi(x) - \varphi(\mu_j) \|^2
+\bigg\},
 \end{equation}
-Furthermore, and most importantly, our analysis is valid
-for any space $\mathcal{X}$ equiped with a semimetric of negative type $\rho$.
-This method is
-nonparametric since it does not assume any form of the 
-distribution of the data,
-contrary to $k$-means and gaussian mixture models (GMM), for example.
-Moreover, this approach does not require the concept of the cluster mean, or
-median, which can be ill-defined in some types of data, such as images for
-instance.
+where $\mu_j = \tfrac{1}{n_j} \sum_{x \in \C_j} x$ is the  mean of cluster
+$\C_j$.
+It is well-known \cite{Dhillon} that problem \eqref{eq:kernel_kmeans} 
+is equivalent to the QCQP
+\eqref{eq:qcqp2}. The next result makes this explicit, showing its
+relation with the energy statistics based clustering considered thus far.
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{A Simple Iterative Algorithm}
-\label{sec:iterative_algo}
+\begin{proposition}
+\label{th:kernel_kmeans}
+The clustering problem
+\eqref{eq:minimize} based on energy statistics 
+is equivalent to the kernel $k$-means problem
+\eqref{eq:kernel_kmeans}, and both are equivalent to \eqref{eq:qcqp2}.
+\end{proposition}
+\begin{proof}
+Notice that $\| \varphi(x) - \varphi(\mu_j) \|^2 = \varphi(x)^\top \varphi(x)
+- 2 \varphi(x)^\top \varphi(\mu_j) + \varphi(\mu_j)^\top \varphi(\mu_j)$,
+therefore
+\begin{equation}
+\label{eq:J}
+J = \dfrac{1}{2} \sum_{j=1}^k \sum_{x\in\C_j} \bigg(
+\kk(x,x) - 
+\dfrac{2}{n_j} \sum_{y\in \C_j} \kk(x,y) + \dfrac{1}{n_j^2}
+\sum_{y,z \in \C_j} \kk(y,z) \bigg).
+\end{equation}
+The first term is global so it does not contribute to the optimization
+problem. Notice that the third term gives
+$\sum_{x\in\C_j} \tfrac{1}{n_j^2} \sum_{y,z\in\C_j} \kk(y,z) =
+\tfrac{1}{n_j}\sum_{y,z\in\C_j} \kk(y,z)$, which is the same as
+the second term. Thus the optimization problem is
+\begin{equation}
+\min_{\C_1,\dotsc,\C_k} J(\C_1,\dotsc,\C_k) = \max_{\C_1,\dotsc,\C_k}
+\sum_{j=1}^k \dfrac{1}{n_j} \sum_{x,y \in\C_j} \kk(x,y)
+\end{equation}
+which is exactly the same as 
+\eqref{eq:max_prob}. The remaining of the proof proceeds as 
+already shown in the proof of Proposition~\ref{th:qcqp2}.
+\end{proof}
 
-We can formulate an iterative algorithm to find
-an approximate solution to \eqref{eq:qcqp2} on the same lines
-as kernel $k$-means.
-Let $t$ be the iteration time. First precompute the kernel
-matrix $G$, fix the number of clusters $k$, then
-perform the following steps:
-\begin{enumerate}
-\item \label{initialize} 
-Initialize clusters $\{ \C_1^{(0)},\dotsc,\C_k^{(0)} \}$, which
-determines the label matrix $\Zt^{(0)}$.
-\item \label{assignment} For each datapoint $x_i$ compute
-its cluster assignment through
-\begin{equation}
-\label{eq:algo}
-\Zt^{(t+1)}_{ij} = 
-\begin{cases}
-\tfrac{1}{\sqrt{n_j}} & \mbox{if $j = \argmax_\ell \tfrac{1}{n_\ell} 
-\sum_{m=1}^n G_{i m} \Zt^{(t)}_{m\ell}$ } \\
-0 & \mbox{otherwise.}
-\end{cases}
-\nonumber
-\end{equation}
-\item If converged return $\Zt^{(t+1)}$, otherwise
-set $t = t+1$ and repeat step~\ref{assignment}.
-\end{enumerate}
-If data is $D$-dimensional, computing $G$ has complexity
-$O(n^2 D)$. Step \ref{assignment} above has complexity $O(n)$ for
-each point, thus total complexity $O(n^2)$. Assuming we perform $T$ 
-iterations,
-the total complexity of the algorithm is $O(n^2(D + T))$.
-We can initialize the algorithm in step~\ref{initialize} with any method
-we want, a good alternative is the initialization from $k$-means++.
+It was shown \cite{Dhillon} that kernel $k$-means, spectral clustering,
+and graph partitioning problems such as ratio association, ratio cut, and
+normalized cut are all equivalent to a QCQP of the form \eqref{eq:qcqp2}.
+Actually, in general, 
+this corresponds to a weighted version of \eqref{eq:qcqp2} which reads
+$ \Tr \left( Y^\top W^{1/2} G \, W^{1/2} Y \right)$, where 
+$W = 
+\diag(w(x_1),\dotsc, w(x_n))$ and $w(\cdot)$ is a
+weight attributed to each data point. Our previous 
+results show that energy statistics
+based clustering is also equivalent to these problems. The advantage
+of energy statistics is that the semimetric $\rho$ fixes the kernel
+through \eqref{eq:kernel_semimetric}.
 
@@
 \section{Two-Class Problem in One Dimension}
+\label{sec:twoclass}
 
-Here we consider an alternative approach in the simplest case possible, which
-is one-dimensional data and a two-class problem.
+Before stating a general algorithm to solve \eqref{eq:qcqp2}, 
+let us first consider the simplest possible case which
+is one-dimensional data and a two-class problem. This will also 
+be useful later
+for comparison with the more general iterative algorithm.
+
 If we choose
-$\rho(x,y) = |x - y|$, we can actually compute 
-\eqref{eq:g_def} in $O(n \log n)$ instead of $O(n^2 D)$ and find
+$\rho(x,y) = |x - y|$ we can actually compute 
+\eqref{eq:g_def} in $\OO(n \log n)$ and find
 a direct solution to \eqref{eq:minimize}. 
@@
 \left( \Ind{x \ge y} - \Ind{x < y} \right) x + 
-\left( \Ind{y > x} - \Ind{y \le x} \right) y
+\left( \Ind{y > x} - \Ind{y \le x} \right) y ,
 \end{aligned}
 \end{equation}
-where $\Ind{}$ is the indicator function.
+where we have the indicator function defined as
+$\Ind{A}=1$ if $A$ is true, and $\Ind{A}=0$ otherwise. 
 Let $\C$ be a partition with
@@
 \label{eq:g_ind}
-g(\C,\C) = \dfrac{1}{n^2} \sum_{x \in \C} 
+g\left(\C,\C\right) = \dfrac{1}{n^2} \sum_{x \in \C} 
 \sum_{y \in \C} 
@@
 parenthesis is simply counting the number of elements in $\C$ that satisfy
-the conditions of the indicator functions. Therefore, assume
-that we order the data as 
-$\bar{\C} = [ x_j \in \C: x_1 \le x_2 \le \dotsm \le x_{n}]$. Thus, we
-can write \eqref{eq:g_ind} in the following very simple form:
+the conditions of the indicator functions. Assuming
+that we first order the data in the partition, obtaining
+$\bar{\C} = [ x_j \in \C: x_1 \le x_2 \le \dotsm \le x_{n}]$, we
+can write \eqref{eq:g_ind} in the following simple form:
 \begin{equation}
 \label{eq:g1d}
-g(\C, \C) = \dfrac{2}{n^2} \sum_{\ell=1}^n (2\ell - 1 - n) x_\ell .
+g\left(\bar{\C}, \bar{\C}\right) = 
+\dfrac{2}{n^2} \sum_{\ell=1}^n (2\ell - 1 - n) x_\ell .
 \end{equation}
-Note that the cost of computing this is $O(n)$, once the data
-is sorted which costs at the most $O(n\log n)$.
-Now suppose we order each partition, obtaining $\mathbb{X} = \cup_{j=1}^k
-\bar{\C}_j$. Then \eqref{eq:within} is given by
+Note that the cost of computing this is $\OO(n)$, and the cost of
+sorting the data
+is at the most $\OO(n\log n)$.
+Assuming that each partition is ordered  $\mathbb{X} = \bigcup_{j=1}^k
+\bar{\C}_j$, but notice that the entire data set $\mathbb{X}$ does not 
+need to be necessarily ordered, the within energy dispersion
+\eqref{eq:within} can be written as
 \begin{equation}
 \label{eq:w1d}
-W(\{ \bar{\C}_i \}) = 
+W\left( \bar{\C}_1,\dotsc,\bar{\C}_k \right) = 
 \sum_{j=1}^k \sum_{\ell=1}^{n_j} \dfrac{2\ell - 1 - n_j}{n_j} \, x_\ell.
 \end{equation}
-For a two-class problem, we can cluster the data by computing \eqref{eq:w1d}
-for each possible split, and pick the one with minimum value. This is done
-as follows:
-\begin{enumerate}
-\item Sort the pooled data, obtaining the sorted set $\bar{\mathbb{X}}$ with
-$n$ elements.
-\item For each $t=1,2,\dotsc,n$, make two partitions 
-$\bar{\C}_1^{(t)} = [x_i]_{i=1}^t$ and 
-$\bar{\C}_2^{(t)} = [x_i]_{i=t+1}^n$. 
-\item Compute $W^{(t)} = W(\{ \bar{\C}_1^{(t)},\bar{\C}_2^{(t)}  \})$ 
-from \eqref{eq:w1d}.
-\item The best split is $t^\star = \argmin_t W^{(t)}$, which
-determines the final clusters.
-\end{enumerate}
-Notice that this procedure does not require any initialization, however,
-it only works for one-dimensional data and for Euclidean distance.
 
-\section{Numerical Experiments}
+For a two-class problem we can use \eqref{eq:w1d} to cluster data
+through a simple algorithm 
+as follows. We first order
+the entire dataset $\mathbb{X} \to \bar{\mathbb{X}}$. Then 
+we compute \eqref{eq:w1d} for each possible split of $\bar{\mathbb{X}}$
+and pick the point which gives the minimum value of $W$. 
+This procedure is described in Algorithm~\ref{algo1d} below. 
+Notice that
+this method does not require any initialization,
+however,
+it only works for one-dimensional data with Euclidean distance. The total
+complexity of the algorithm is $\OO(n\log n + n^2) = \OO(n^2)$.
+
+\begin{algorithm}[H]\vspace{.5em}
+\begin{algorithmic}[1]
+\INPUT Data $\mathbb{X}$
+\OUTPUT Label matrix $Z$
+\STATE Sort $\mathbb{X}$ obtaining 
+$\bar{\mathbb{X}}= [ x_1,\dotsc,x_n ]$
+    \FOR{$j\in [ 1,\dotsc,n ]$}
+        \STATE Let $\bar{\C}_1^{(j)} = [x_i: i=1,\dotsc,j]$ and 
+                $\bar{\C}_2^{(j)} = [x_i : i=j+1,\dotsc,n]$
+        \STATE  
+            $W^{(j)} \leftarrow W \big( \bar{\C}_1^{(j)},\bar{\C}_2^{(j)}  
+            \big)$ from \eqref{eq:w1d}
+    \ENDFOR
+    \STATE $j^\star \leftarrow \argmin_j W^{(j)}$ determines the best split
+    \STATE $Z_{j\bullet} = (1,0) $ if $j\le j^\star$, and
+           $Z_{j\bullet} \leftarrow (0,1)$ otherwise, for $j=1,\dotsc,n$
+\end{algorithmic}
+\caption{\label{algo1d}
+Procedure to find an approximate solution to \eqref{eq:qcqp2} 
+for a two-class problem in one dimension and with Euclidean distance.}
+\end{algorithm}
+
+Assuming the true label matrix $Z$ is available, a direct
+measure of how different the estimated matrix $\hat{Z}$ 
+is from $Z$, up to label
+permutations, is given by
+\begin{equation}
+\label{eq:accuracy}
+\textnormal{accuracy}(\hat{Z}) = \max_\sigma 
+\dfrac{1}{n}\sum_{i=1}^n\sum_{j=1}^k \hat{Z}_{i \sigma(j)} Z_{ij}
+\end{equation}
+where $\sigma$ is a permutation
+of the $k$ cluster groups. 
+The accuracy is always between $[0,1]$, where
+$1$ corresponds to all points correctly clustered, and 
+$0$ to all points wrongly clustered.
+For a two-class problem with equal
+number of points in each cluster, the value $1/2$ correspond
+to chance.
+
+Before proposing a more general iterative algorithm to \eqref{eq:qcqp2},
+let us consider two simple experiments with equal number of points
+in each cluster. 
+We keep increasing the number of points in the clusters for each experiment, 
+and cluster the data using Algorithm~\ref{algo1d}. 
+We also cluster the same data set
+with GMM, through EM algorithm, and with $k$-means. In both of these
+cases we use the initialization from $k$-means++ and we run the algorithms
+a couple of times with different initializations and choose the answer
+with best objective function value. We use \eqref{eq:accuracy} to measure
+the clustering quality. The results are shown in 
+Fig.~\ref{fig:1d}. In  Fig.~\ref{fig:1d}a 
+we have data from normal distributions,
+where we can see that all the three methods
+perform closely, with a slight advantage of GMM, as expected since
+it is the right model for the data. However, as shown in Fig~\ref{fig:1d}b,
+for lognormal distributions Algorithm~\ref{algo1d} provides a huge improvement
+compared to both GMM and $k$-means, which basically cluster at chance.
+The zero accuracy values for GMM happened when EM algorithm was unable
+to estimate the parameters. These two simple experiments illustrate
+how energy statistics based clustering is nonparametric.
 
 \begin{figure}
-\begin{minipage}{0.33\textwidth}
-\centering
-\includegraphics[width=\textwidth]{gauss_means.pdf}\\[-.8em]
+\begin{minipage}{0.49\textwidth}
+\includegraphics[width=\textwidth]{normal.pdf}\\[-1em]
 (a)
 \end{minipage}
-\begin{minipage}{0.33\textwidth}
-\centering
-\includegraphics[width=\textwidth]{gauss_covs.pdf}\\[-.8em]
+\begin{minipage}{0.49\textwidth}
+\includegraphics[width=\textwidth]{lognormal.pdf}\\[-1em]
 (b)
 \end{minipage}
-\begin{minipage}{0.33\textwidth}
-\centering
-\includegraphics[width=\textwidth]{gauss_pis.pdf}\\[-.8em]
-(c)
-\end{minipage}
 \caption{
-\label{fig:dimension}
-gadsga
+\label{fig:1d}
+We cluster data using Algorithm~\ref{algo1d} ($1D$-Energy in the plots), 
+GMM, and $k$-means. We use
+\eqref{eq:accuracy} to evaluate cluster quality. Both clusters
+have the same number of points, which are increased in each experiment.
+(a) 
+$x\sim \tfrac{1}{2}\left( \mathcal{N}(\mu_1,\sigma_1) +
+\mathcal{N}(\mu_2,\sigma_2)  \right)$ with 
+$\mu_1 = 0$,
+$\mu_2 = 5$,
+$\sigma_1 = 1$, and
+$\sigma_2 = 2$.
+(b) 
+$x\sim \tfrac{1}{2}\left( e^{\mathcal{N}(\mu_1,\sigma_1)} +
+e^{\mathcal{N}(\mu_2,\sigma_2)}  \right)$ with 
+$\mu_1 = 0$,
+$\mu_2 = -1.5$,
+$\sigma_1 = 0.3$, and
+$\sigma_2 = 1.5$.
 }
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Iterative Algorithm for Energy Clustering}
+\label{sec:algo}
+
+In this section we will introduce a new iterative algorithm to find a local
+maximizer of \eqref{eq:qcqp2}, however, due to 
+Proposition~\ref{th:kernel_kmeans} we can also find an approximate
+solution by the well-known kernel $k$-means algorithm. Thus, to be
+self-contained let us first recall kernel $k$-means algorithm in this
+context.
+
+\subsection{Kernel $\bm{k}$-Means Algorithm}
+
+Consider this optimization problem as
+written in the form \eqref{eq:max_prob},
+\begin{equation}
+\label{eq:maxQ}
+\max_{\{ \C_1,\dotsc,\C_k \}} 
+\bigg\{ Q = \sum_{j=1}^k \dfrac{Q_j}{n_j}  \bigg\},
+\qquad Q_j = \sum_{x,y\in\C_j} \kk(x,y),
+\end{equation}
+where $Q_j$ represents the internal cost of cluster $\C_j$, and
+$Q$ is the total cost where each cluster cost 
+is weighted by the inverse
+of the number of its elements. For a data point $x_i$ its cost
+with cluster $\C_j$ is denoted by
+\begin{equation}
+\label{eq:costxij}
+Q_j(x_i) \equiv \sum_{y\in\C_j} \kk(x_i, y) = 
+G_{i \bullet} \cdot Z_{\bullet j}.
+\end{equation}
+
+Now for kernel $k$-means 
+consider \eqref{eq:J} where we define
+the function
+\begin{equation}
+\label{eq:Jell}
+J^{(\ell)}(x_i) \equiv -\dfrac{2}{n_\ell} Q_\ell(x_i) + \dfrac{1}{n_\ell^2}
+Q_\ell
+\end{equation}
+which represents the cost of $x_i$ with cluster $\C_\ell$. Thus,
+one assigns point $x_i$ to the cluster $\C_{j^\star}$ according
+to $j^\star = \argmin_\ell J^{(\ell)}(x_i)$, for $\ell = 1,\dotsc,k$.
+This procedure is performed for every data point, and repeated until
+convergence, i.e. until no new assignments are made.
+The complete algorithm is shown in Algorithm~\ref{kmeans_algo}.
+Although our formulation looks a little bit different than the standard
+kernel $k$-means found in the literature \cite{Dhillon}, this is precisely
+the same algorithm.
+
+Notice that to compute the first term in \eqref{eq:Jell} requires
+$\OO(n_\ell)$ operations, and although the second term requires
+$\OO(n_\ell^2)$, it only needs to be computed once outside the loop through
+data points. Therefore, the time complexity of this algorithm is
+$\OO(n k \max_\ell n_\ell) = \OO(k n^2)$. If $G$ is sparse with 
+$n'$ nonzero elements, this complexity can be further reduced
+to $\OO(k n')$.
+
+\begin{algorithm}[H]
+\vspace{.5em}
+\begin{algorithmic}[1]
+    \INPUT number of clusters $k$, Gram matrix $G$, initial label
+    matrix $Z = Z_0$
+    \OUTPUT label matrix $Z$ 
+  \STATE Let $\bm{q} = (Q_1, \dotsc, Q_k)^\top$ 
+            have the costs of each cluster, according to \eqref{eq:maxQ}
+  \STATE Let $\bm{n} = (n_1,\dotsc,n_k)^\top$ have the number of elements
+        in each cluster
+  \REPEAT
+    \FOR{ $i=1,\dotsc,n$}
+        \STATE Let $j$ be such that $x_i \in \C_j$
+        \STATE $j^\star \leftarrow \argmin_{\ell} J^{(\ell)}(x_i)$
+            according to \eqref{eq:Jell}, for $\ell=1,2,\dots,k$
+        \IF{ $j^\star \ne j$} 
+            \STATE Move $x_i$ to $\C_{j^\star}$: $Z_{ij} \leftarrow 0$,
+            $Z_{ij^\star} \leftarrow 1$
+            \STATE Update $\bm{n}$: $n_j \leftarrow n_j - 1$,
+                    $n_{j^\star} \leftarrow n_{j^\star} + 1$
+            \STATE Update $\bm{q}$: $q_j \leftarrow q_j - 2Q_j(x_i)$, 
+    $q_{j^\star} \leftarrow q_{j^\star} + 2Q_{j^\star}(x_i)$
+    %    \ELSE
+    %        \STATE Do nothing;
+        \ENDIF
+    \ENDFOR
+  \UNTIL{convergence}
+\end{algorithmic}
+\caption{\label{kmeans_algo}
+Kernel $k$-means algorithm 
+to find an approximate solution to \eqref{eq:qcqp2}.}
+\end{algorithm}
+
+\subsection{Energy Cost Algorithm}
+
+Now let us consider a different algorithm, which is based on computing
+the difference in the energy cost when assigning a given data point to
+a different partition.
+Suppose we have a data point $x_i \in \mathcal{X}$
+which is currently classified as being in cluster $\C_j$, yielding
+a total cost function \eqref{eq:maxQ} denoted by $Q^{(j)}$.
+Let us consider the change in the total cost by moving
+$x_i$ to cluster $\C_\ell$. 
+Denote the new total cost after moving $x_i$ to $\C_\ell$ by $Q^{(\ell)}$.
+It is straightforward to see that
+\begin{equation}
+\label{eq:changeQ}
+\begin{split}
+\Delta Q^{j \to \ell}(x_i) &\equiv Q^{(\ell)} - Q^{(j)} \\ 
+&= 
+\dfrac{1}{n_j - 1}\left[ \dfrac{Q_j}{n_j} - 2 Q_j(x_i) \right]
+- \dfrac{1}{n_\ell + 1}\left[ \dfrac{Q_\ell}{n_\ell} - 2 \big(Q_\ell(x_i) + 
+\kk(x_i,x_i)\big) 
+\right].
+\end{split}
+\end{equation}
+Thus, if $\Delta Q^{j\to \ell}(x_i) > 0$ we get closer to a 
+maximum of \eqref{eq:maxQ} by
+moving $x_i$ to $\C_\ell$, otherwise we keep $x_i$ in $\C_j$. Based on
+this we propose an algorithm where
+the iterates are performed as follows.
+We start with an initial configuration for the label matrix $Z$, 
+then for each
+point $x_i$ 
+we compute the cost of moving it to another cluster,
+$\Delta Q^{j\to \ell}(x_i)$ for 
+$\ell=1,\dots,k$ with $\ell \ne j$. 
+We then choose $j^\star = \argmax_\ell \Delta^{j \to \ell}(x_i)$.
+If $\Delta Q^{j \to j^\star}(x_i) > 0$ 
+we move $x_i$ to cluster $\C_{j^\star}$, otherwise 
+we keep $x_i$ in its original cluster $\C_j$. We update $Z$ accordingly.
+The process is repeated
+until convergence, i.e. until no points are assigned to new clusters. 
+This procedure is described in Algorithm~\ref{algo} below.
+
+\begin{algorithm}[H]
+\vspace{.5em}
+\begin{algorithmic}[1]
+    \INPUT number of clusters $k$, Gram matrix $G$, 
+                initial label matrix $Z=Z_0$
+    \OUTPUT label matrix $Z$
+  \STATE Let $\bm{q} = (Q_1, \dotsc, Q_k)^\top$ 
+            have the energy costs of each cluster, according to \eqref{eq:maxQ}
+  \STATE Let $\bm{n} = (n_1,\dotsc,n_k)^\top$ have the number of elements
+        in each cluster
+  \REPEAT
+    \FOR{ $i=1,\dotsc,n$}
+        \STATE Let $j$ be such that $x_i \in \C_j$
+        \STATE $j^\star \leftarrow \argmax_{\ell} \Delta Q^{j\to \ell}(x_i)$, 
+            for $\ell=1,2,\dots,k$ and $\ell \ne j$ \label{stepmove}
+        \IF{ $\Delta Q^{j \to j^\star}(x_i) > 0$ }
+            \STATE Move $x_i$ to $\C_{j^\star}$: $Z_{ij} \leftarrow 0$,
+            $Z_{ij^\star} \leftarrow 1$
+            \STATE Update $\bm{n}$: $n_j \leftarrow n_j - 1$,
+                    $n_{j^\star} \leftarrow n_{j^\star} + 1$
+            \STATE Update $\bm{q}$: $q_j \leftarrow q_j - 2Q_j(x_i)$, 
+    $q_{j^\star} \leftarrow q_{j^\star} + 2\left(Q_{j^\star}(x_i)+
+    G_{ii}\right)$
+    %    \ELSE
+    %        \STATE Do nothing;
+        \ENDIF
+    \ENDFOR
+  \UNTIL{convergence}
+\end{algorithmic}
+\caption{\label{algo}
+Energy cost algorithm to find an approximate solution to \eqref{eq:qcqp2}.}
+\end{algorithm}
+
+Notice that computing $G$ requires $\OO( D n^2)$ operations, where 
+$D$ is the dimension of each data point and $n$ is the data size. However,
+both previous algorithms assume that $G$ is given. There are more efficient
+methods to compute $G$, specially if it is sparse. We will not consider
+this further, and just assume that $G$ is given.
+The computation of each cluster cost
+$Q_j$ has complexity $\OO(n_j^2)$, and overall to compute $\bm{q}$
+we have $\OO(n_1^2+\dots + n_k^2) = \OO(k \max_j n_j^2)$. 
+These operations, however, only need to be performed a single time. Now for
+each point $x_i$ we need to compute $Q_j(x_i)$ once, which is
+$\OO(n_j)$, and we need to compute $Q_\ell(x_i)$ for each $\ell\ne j$. 
+The cost of computing 
+\eqref{eq:costxij} is $\OO(n_j)$, thus the cost of step~$8$ in
+Algorithm~\ref{algo} is $\OO(k \max_j n_j)$ for $j=1,\dotsc,k$.
+For the 
+entire dataset this gives a time-complexity
+of $\OO(n k  \max_j n_j) =\OO(k n^2)$. This is the same cost as
+in kernel $k$-means, Algorithm~\ref{kmeans_algo}. Again, if $G$ is sparse
+this can be reduced to $\OO(k n')$, where $n'$ is the number of nonzero
+entries of $G$.
+In the numerical experiments below
+we choose the initial $Z$ from the initialization 
+procedure of $k$-means++ \cite{Vassilvitskii}.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Numerical Experiments}
+\label{sec:numerics}
+
+In the following experiments we fix the semimetric as $\rho(x,y) = \| x-y\|$,
+with induced kernel 
+\eqref{eq:kernel_semimetric}. We will compare Algorithm~\ref{algo} with
+kernel $k$-means Algorithm~\ref{kmeans_algo}, and also with
+standard $k$-means and GMM through the Expectation Maximization (EM) algorithm
+as well. For synthetic data our measure of clustering performance
+will be \eqref{eq:accuracy}. Moreover, for all the algorithms, we always
+choose the initialization from $k$-means++ \cite{Vassilvitskii}.
+
+We first consider clustering in high dimensions and analyze 
+how the algorithms degrade as the number of dimensions increase, while
+keeping the number of points in each cluster fixed. 
+In Figure~\ref{fig:gauss}a we have data generated from normal 
+distributions in 
+$D$-dimensions:
+\begin{equation}
+\label{eq:gauss1}
+\begin{split}
+&x \sim \tfrac{1}{2}\left\{ 
+\mathcal{N}(\mu_1,\Sigma_1) + \mathcal{N}(\mu_2, \Sigma_2)\right\}, \\
+&\mu_1 = (\underbrace{0,\dotsc,0}_{\times D})^\top , \quad
+\mu_2 = 0.7 \times (\underbrace{1,\dots,1}_{\times 10},
+\underbrace{0,\dots,0}_{\times (D-10)})^\top, \quad
+\Sigma_1 = \Sigma_2 = I_D.
+\end{split}
+\end{equation}
+For each experiment we only keep signal in
+$\mu_2$ in the first $10$
+dimensions, and keep increasing the ambient dimension $D$. For each
+data set, we run the algorithms $10$ times and pick the result with
+the best objective function value.
+In Figure~\ref{fig:gauss}a one sees that
+GMM is not able to estimate the covariance matrix 
+when the number
+of dimensions exceeds the number of points in each cluster. 
+One can see a slightly better performance of Algorithm~\ref{algo} compared
+to the other ones.
+In Figure~\ref{fig:gauss}b we have the same type of experiment but 
+with 
+\begin{equation}
+\label{eq:gauss2}
+\begin{split}
+&x \sim \tfrac{1}{2}\left( 
+\mathcal{N}(\mu_1,\Sigma_1) + \mathcal{N}(\mu_2, \Sigma_2)\right), \\
+&\mu_1 = (\underbrace{0,\dotsc,0}_{\times D})^\top , \,
+\mu_2 = 0.7 \times (\underbrace{1,\dots,1}_{\times 10},
+\underbrace{0,\dots,0}_{\times (D-10)})^\top, \,
+\Sigma_1 = I_D, \, 
+\Sigma_2 = \left( \begin{smallmatrix} \tfrac{1}{2} I_{10} & 0 \\ 0 & I_{D-10}
+\end{smallmatrix}\right). \quad
+\end{split}
+\end{equation}
+
 \begin{figure}
+\begin{minipage}{0.49\textwidth}
 \centering
+\includegraphics[width=1\textwidth]{gauss_dim.pdf}\\[-.8em]
+(a)
+\end{minipage}
+\begin{minipage}{0.49\textwidth}
+\centering
+\includegraphics[width=1\textwidth]{gauss_cov.pdf}\\[-.8em]
+(b)
+\end{minipage}
+\caption{
+\label{fig:gauss}
+Experiments where data is normally distributed.
+(a) We keep $100$ points on each cluster
+and increase the ambient dimension, as described in \eqref{eq:gauss1}. The blue
+dots correspond to Algorithm~\ref{algo}, while
+the red dots 
+corresponds to Algorithm~\ref{kmeans_algo}.
+(b) The same but with data following \eqref{eq:gauss2}.
+In both experiments one notice a slightly better accuracy
+of Algorithm~\ref{algo} compared to the other ones.
+}
+\end{figure}
+
+In Figure~\ref{fig:unbalanced}a we consider how the previous algorithms
+behave
+for unbalanced clusters. We generate data as
+\begin{equation}
+\label{eq:gauss3}
+\begin{split}
+&x \sim  
+\dfrac{n_1}{N} \mathcal{N}(\mu_1,\Sigma_1) + 
+\dfrac{n_2}{N} \mathcal{N}(\mu_2, \Sigma_2), 
+\quad \mu_1 = (0,0,0,0)^\top , \,
+\mu_2 = 1.5\times (1,1,0,0)^\top, \\
+&
+\Sigma_1 = I_4, \quad
+\Sigma_2 = \left( 
+\begin{smallmatrix} 
+1/2 & 0 & 0 & 0\\
+0 & 1/2 & 0 & 0 \\
+0 & 0 & 1 & 0 \\
+0 & 0 & 0 & 1 
+\end{smallmatrix}\right), \quad
+n_1 = N - m, \quad  n_2 = N + m, \quad N=200,
+\end{split}
+\end{equation}
+and we increase $m$, i.e. we make the clusters progressively more unbalanced.
+As well-known, one an see that GMM
+works well for unbalanced clusters, mostly because 
+it is a soft clustering algorithm.
+We can see that all the other kernel based methods
+degrade more rapidly than GMM for
+highly unbalanced clusters.
+This is
+unsurprising since Algorithms~\ref{algo}~and~\ref{kmeans_algo} both 
+make hard assignments. Finally, Figure~\ref{fig:unbalanced}b considers
+the same experiment as in Figure~\ref{fig:gauss}a but for lognormal
+data. With $\rho(x,y) = \| x-y\|^{1/2}$ we even gain a slight improvement.
+Energy statistics clustering outperforms all the other methods since
+it is nonparametric.
+
+\begin{figure}
+\begin{minipage}{0.49\textwidth}
+\centering
+\includegraphics[width=1\textwidth]{gauss_pi.pdf}\\[-.8em]
+(a)
+\end{minipage}
+\begin{minipage}{0.49\textwidth}
+\centering
+\includegraphics[width=1\textwidth]{loggauss_dim.pdf}\\[-.8em]
+(b)
+\end{minipage}
 \caption{
 \label{fig:unbalanced}
-gadsga
+(a) Previous algorithms for unbalanced clusters, 
+according to \eqref{eq:gauss3}.
+(b) The same experiment as in Figure~\ref{fig:gauss}a but for lognormal
+data as 
+$x\sim \tfrac{1}{2}\left( e^{\mathcal{N}(\mu_1,\Sigma_1)} +
+e^{\mathcal{N}(\mu_2,\Sigma_2)} \right)$ with
+$\mu_1 = (0,\dotsc,0)^\top$,
+$\mu_2 = 0.5\times(1,\dotsc,1,0,\dotsc,0)^\top$ with signal in $d=10$, 
+$\Sigma_1 = 0.3 \times I_D$, and $\Sigma_2=I_D$. 
+One see a clear advantage of
+energy statistics clustering. The extra green line correspond
+to $\rho(x,y) = \| x-y\|^{1/2}$.
 }
 \end{figure}
+
+In Figure~\ref{fig:weird} we consider other types of data in two dimensions.
+In each case we perform 10 experiments and compute the accuracy
+based on the true labels. The plots show the average results of 10 runs
+of each experiment (dots), and the maximum and minimum values as well (shaded
+area). Although energy clustering is nonparametric, these datasets are
+very specific so we can take advantage and make a suitable choice of
+the semimetric $\rho$. This illustrates that energy clustering is also
+flexible enough to incorporate prior information about the data.
+In Figure~\ref{fig:weird}a we have parallel cigars from
+\begin{equation}
+\label{eq:cigar}
+\begin{split}
+&x \sim \tfrac{1}{2}\left( 
+\mathcal{N}(\mu_1, \Sigma_1) + 
+\mathcal{N}(\mu_2, \Sigma_2)
+\right), \\
+& \mu_1 = (0,0)^\top, \quad \mu_2 = (6.5, 0)^\top, \quad
+\Sigma_1 = \Sigma_2 = \left( \begin{smallmatrix} 
+1 & 0 \\ 0 & 20
+\end{smallmatrix}\right)
+\end{split}
+\end{equation}
+with $100$ points in each cluster. Notice that $k$-means is unable to
+cluster this data, while energy clustering performs pretty much as well
+as GMM, which is the optimal algorithm suited to this kind of data.
+We used the semimetric $\rho(x,y) = \| x -  y\|^{1/2}$ in this case.
+In Figure~\ref{fig:weird}b we generate two concentric circles with a small
+gaussian noise. To make the semimetric a little bit more localized we
+include a decaying exponential in the form:
+\begin{equation}
+\label{eq:rho_loc}
+\rho(x,y) = \| x-y \|^2 e^{-\tfrac{1}{2}\| x-y \|}.
+\end{equation}
+Notice that energy clustering is able to cluster this data almost perfectly
+in most cases.
+In Figure~\ref{fig:weird}c we consider two spirals, also with a small
+gaussian noise. Due to the geometry of the data
+we consider
+\begin{equation}
+\label{eq:rho_sin}
+\rho(x,y) = \| x - y\|^2 \theta \sin \left(  \tfrac{\| x-y \|}{\theta}\right),
+\qquad \theta = 0.9.
+\end{equation}
+Again, energy clustering performs much better than $k$-means and GMM which
+are unable to cluster this kind of dataset.
 
@@
 \caption{
-\label{fig:dimension}
-gadsga
+\label{fig:weird}
+(a) Parallel cigars generated from \eqref{eq:cigar} with $100$ points
+in each class. Use use energy clustering
+with $\rho = \| x - y\|^{1/2}$. (b) Two concentric circles with $150$ points
+in each class. We use the semimetric \eqref{eq:rho_loc}. 
+(c) Two concentric spirals with
+$150$ points in each class. We use the semimetric \eqref{eq:rho_sin}.
 }

### 2017-06-21 qcqp_clustering/code/eclust/data.py

@@ -13,28 +13,7 @@ import pylab
 
 from sklearn.decomposition import PCA
 
-
-# pyplot customization
-pylab.rc('lines', linewidth=.5, antialiased=True, markeredgewidth=0.1)
-pylab.rc('font', family='computer modern roman', style='normal',
-         weight='normal', serif='computer modern sans serif', size=10)
-pylab.rc('text', usetex=True)
-pylab.rc('text.latex', preamble=[
-        '\usepackage{amsmath,amsfonts,amssymb,relsize,cancel}'])
-pylab.rc('axes', linewidth=0.5, labelsize=10)
-pylab.rc('xtick', labelsize=10)
-pylab.rc('ytick', labelsize=10)
-#pylab.rc('legend', numpoints=1, fontsize=10, handlelength=0.5)
-pylab.rc('legend', numpoints=1, fontsize=10)
-fig_width_pt = 455.0 / 1.5 # take this from LaTeX \textwidth in points
-inches_per_pt = 1.0/72.27
-golden_mean = (np.sqrt(5.0)-1.0)/2.0
-fig_width = fig_width_pt*inches_per_pt
-fig_height = fig_width_pt*inches_per_pt
-#fig_height = fig_width*golden_mean
-pylab.rc('figure', figsize=(fig_width, fig_height))
-
-
+from pplotcust import *
 
 def multivariate_normal(means, sigmas, ns):
     """Generate several normal distributed data points.
@@ -182,17 +161,23 @@ def plot(X, z, colors=['b', 'r', 'g'], fname='plot.pdf'):
     fig.tight_layout()
     fig.savefig(fname)
 
-def histogram(X, z, fname='plot.pdf'):
+def histogram(X, z, colors=['b', 'r'], fname='plot.pdf'):
     """Plot histograms of 1-dimensional data."""
     z_unique = np.unique(z)
     fig = plt.figure()
-    bins=np.arange(min(X), max(X) + 0.2, 0.2)
+    #bins=np.arange(min(X), max(X) + 0.2, 0.2)
     ax = fig.add_subplot(111)
-    colors = iter(plt.cm.rainbow(np.linspace(0,1,len(z_unique))))
+    colors = iter(colors)
     for k in z_unique:
         idx = np.where(z==k)[0]
         x = X[idx]
-        ax.hist(x, bins=bins, facecolor=next(colors), alpha=.6, normed=1)
[…]

### 2017-06-21 qcqp_clustering/code/eclust/energy1d.py

@@ -1,4 +1,4 @@
-"""Energy Statistics in 1D"""
+"""Energy Statistics Clustering in 1D"""
 
 # Guilherme Franca <guifranca@gmail.com>
 # Johns Hopkinks University, Neurodata
@@ -40,12 +40,13 @@ def within_sample(A):
     Output: between sample energy statistics (S)
 
     """
-    return sum([len(a)/2*mean1D(a,a)  for a in A])
+    #return sum([len(a)/2*mean1D(a,a)  for a in A])
+    return sum([len(a)/2*mean2(a)  for a in A])
 
 def mean2(A):
     """Another version of the above. Assume A is sorted."""
     n = len(A)
-    return (2./n**2)*sum([A[i]*(2.*i +1- n) for i in range(n)])
+    return (2./n**2)*sum([A[i]*(2.*i +1 - n) for i in range(n)])
 
 def two_clusters1D(X):
     """Optimize energy statistics for two clusters in 1D."""
@@ -76,21 +77,27 @@ if __name__ == "__main__":
     from sklearn.cluster import KMeans
     from sklearn.mixture import GMM
     import matplotlib.pyplot as plt
+    import sys
    
-    """
-    X1 = np.random.normal(0, 1, 2000)
+    #X1 = np.random.normal(0, 1, 100)
     #X1 = np.random.gamma(2, 2, 1000)
     #X1 = np.random.uniform(0, 1, 1000)
-    #X2 = np.random.uniform(0.5, 1.5, 1000)
-    X2 = np.random.normal(2, 1, 100)
+    
+    #X2 = np.random.uniform(0.5, 1.5, 100)
+    #X2 = np.random.normal(2, 1, 100)
     #X2 = np.random.gamma(2, 2, 2000)
-    X, z = data.shuffle_data([X1, X2])
     
-    plt.hist(X1, 80, facecolor='blue', alpha=0.6, normed=1)
-    plt.hist(X2, 80, facecolor='red', alpha=0.6, normed=1)
-    plt.show()
-    
-    zh = two_clusters1D(X)
+    #X, z = data.shuffle_data([X1, X2])
+
+    #X, z = data.univariate_lognormal([0, -1.5], [0.25, 1], [4000, 4000])
+    #X, z = data.univariate_normal([0, 5], [1, 2], [4000, 4000])
[…]

### 2017-06-21 qcqp_clustering/code/eclust/kenergy.py

@@ -1,275 +1,0 @@
-"""Iterative algorithm to optimize QCQP for energy statistics clustering."""
-
-# Guilherme Franca <guifranca@gmail.com>
-# Johns Hopkins University, Neurodata
-
-
-from __future__ import division
-
-import numpy as np
-from sklearn.metrics.pairwise import pairwise_distances
-from sklearn.base import BaseEstimator, ClusterMixin
-from sklearn.utils import check_random_state
-
-import kmeanspp
-
-
-def ztoZ(z):
-    """Convert label vector to label matrix."""
-    n = z.shape[0]
-    k = np.unique(z).shape[0]
-    Z = np.zeros((n,k))
-    for i in range(n):
-        Z[i, int(z[i])] = 1
-    return Z
-
-def Ztoz(Z):
-    """Convert label matrix to label vector."""
-    n, k = Z.shape
-    z = np.zeros(n)
-    points, labels = np.where(Z==1)
-    for i in range(len(points)):
-        z[points[i]] = labels[i]
-    return z
-
-def objective(Z, G):
-    """Compute objective function."""
-    Zt = Z.T
-    D = np.dot(Zt, Z)
-    Dinv = np.linalg.inv(D)
-    return np.trace(Dinv.dot(Zt.dot(G.dot(Z))))
-
-def euclidean_rho(x, y, alpha=2):
-    return np.power(np.linalg.norm(x-y), alpha)
-
-def kernel_function(x, y, rho, x0=None):
-    """Return kernel function based on rho."""
-    if x0 == None:
-        x0 = np.zeros(x.shape)
-    return 0.5*(rho(x,x0) + rho(y,x0) - rho(x,y))
-
-def kernel_matrix(X, rho, x0=None, diag=True):
-    """Compute Kernel matrix based on kernel function K(x,y)."""
-    kfunc = lambda x, y: kernel_function(x, y, rho, x0)
-    G = pairwise_distances(X, metric=kfunc, n_jobs=4)
-    if not diag:
-        i = range(len(X))
-        G[i,i] = 0
-    return G
-
-def kenergy_brute(k, G, Z0, max_iter=50):
-    """Optimize energy QCQP in a brute force way. Kind of expensive,
-    but actually works slightly better than Kernel Kmeans, at least
-    it seems like that.
-    """
-    
-    n = G.shape[0]
-    Z = Z0
-    count = 0
-    converged = False
-    while not converged and count < max_iter:
-        converged = True
[…]

### 2017-06-21 qcqp_clustering/code/old_tests/tests.py

@@ -1,0 +1,463 @@
+"""
+Tests and comparison between Energy, K-means, and GMM.
+
+We run Energy, K-means, and GMM a few times, and show average accuracy 
+and variability for several settings.
+
+"""
+
+# Guilhere Franca <guifranca@gmail.com>
+# Johns Hopkins University, Neurodata
+
+import numpy as np
+import matplotlib.pyplot as plt
+import pylab
+
+import eclust.data as data
+import eclust.kernel as kernel
+import eclust.kmeans as km
+import eclust.gmm as gm
+from eclust.metric import accuracy
+from eclust import objectives
+
+
+# pyplot customization
+pylab.rc('lines', linewidth=.5, antialiased=True, markeredgewidth=0.1)
+pylab.rc('font', family='computer modern roman', style='normal',
+         weight='normal', serif='computer modern sans serif', size=10)
+pylab.rc('text', usetex=True)
+pylab.rc('text.latex', preamble=[
+        '\usepackage{amsmath,amsfonts,amssymb,relsize,cancel}'])
+pylab.rc('axes', linewidth=0.5, labelsize=10)
+pylab.rc('xtick', labelsize=10)
+pylab.rc('ytick', labelsize=10)
+#pylab.rc('legend', numpoints=1, fontsize=10, handlelength=0.5)
+pylab.rc('legend', numpoints=1, fontsize=10)
+fig_width_pt = 455.0 / 1.5 # take this from LaTeX \textwidth in points
+inches_per_pt = 1.0/72.27
+golden_mean = (np.sqrt(5.0)-1.0)/2.0
+fig_width = fig_width_pt*inches_per_pt
+fig_height = fig_width*golden_mean
+pylab.rc('figure', figsize=(fig_width, fig_height))
+
+
+###############################################################################
+# clustering functions
+
+def kernel_energy(k, X, z, alpha=1, run_times=3):
+    """Run few times and pick the best objective function value. This
+    stabilizes a little bit the algorithm.
+    
+    """
+    best_score = 0
+    for i in range(run_times):
+        ke = kernel.KernelEnergy(n_clusters=k, max_iter=300,
+                                 kernel_params={'alpha':1})
+        zh = ke.fit_predict(X)
+        score = objectives.kernel_score(ke.kernel_matrix_, zh)
+        if score > best_score:
+            best_score = score
+            best_z = zh
+    return accuracy(best_z, z)
+
+def kmeans(k, X, z, run_times=3):
[…]

### 2017-06-21 qcqp_clustering/draft/energy_clustering.tex
working draft

@@
 We provide a precise mathematical formulation, obtaining
-a quadratically constrained optimization problem (QCQP). We show that
-this optimization problem is equivalent to kernel $k$-means
+a quadratically constrained quadratic optimization problem (QCQP). 
+We show that
+this is equivalent to kernel $k$-means
 optimization problem, however,

### 2017-06-23 qcqp_clustering/draft/energy_clustering.tex
draft update

@@
 It was proposed by 
-Sz\' ekely in the 80's,
-inspired by Newton's potential energy between massive bodies. The idea
-is to associate a statistical potential energy to observations such that 
+Sz\' ekely in the 80's
+inspired by the Newtonian gravitational potential from classical mechanics. 
+The idea
+is to associate a statistical potential energy to observations, such that 
 minimum energy is achieved under the null hypothesis of equality of 
-distributions. This 
-was further generalized for probability 
+distributions. Energy statistics
+was further generalized to probability 
 distributions on arbitrary metric spaces,
 and more recently, a connection with kernels in RKHS was established.
-Nevertheless, although extensively used by the statistics community, it was
-not incorporated in machine learning problems.
-In this paper, we consider the problem of clustering data based
-energy statistics theory. 
-We provide a precise mathematical formulation, obtaining
+Nevertheless, although extensively used by the statistics community, energy
+statistics has not been
+previously incorporated in machine learning problems.
+In this paper, we consider the problem of clustering data from
+an energy statistics theory perspective.
+We provide a precise mathematical formulation, yielding
 a quadratically constrained quadratic optimization problem (QCQP). 
@@
 energy statistics is able to fix the kernel choice. 
-This method is nonparametric, and if prior information is available
+This clustering 
+method is nonparametric, and if prior information is available
 it can be easily incorporated in the kernel construction.
-We propose an algorithm
-to find local optimizers of this QCQP problem based on the energy cost
-of moving points to different clusters. We then compare this algorithm with
-kernel $k$-means, standard $k$-means, and GMM. We test our method
-on synthetic and real data experiments.
-Our results show that energy statistics based clustering outperforms
+We propose an iterative algorithm
+to find local optimizers of this QCQP problem, based on the change
+in the statistical energy 
+of moving points to different clusters. This algorithm 
+is different but has the same computational cost as kernel $k$-means 
+algorithm.
+We then compare this algorithm with
+well-known kernel $k$-means, standard $k$-means, and GMM algorithms by 
+providing carefully designed numerical experiments. The results
+show that, in general, this method outperforms
 these most used clustering algorithms.
@@
 The quantity \cite{Szkely2013}
-\begin{equation}\label{eq:energy}
+\begin{equation}
+\label{eq:energy}
 \Energy(P, Q) \equiv 2 \E \| X - Y\| - \E \| X - X' \| - \E \| Y - Y' \|,
@@
 Energy distance
-provides a characterization of equality of distributions and
+provides a characterization of equality of distributions, and
 $\Energy^{1/2}$ is
@@
 The energy distance can be generalized as, for instance,
-\begin{equation}\label{eq:energy2}
+\begin{equation}
+\label{eq:energy2}
 \Energy_\alpha(P, Q) \equiv 
@@
 where $0<\alpha\le 2$. This quantity is also nonnegative,
-$\Energy_\alpha(P,Q) \ge 0$. Furthermore, for $0<\alpha<2$ we have 
+$\Energy_\alpha(P,Q) \ge 0$. Furthermore, for $0<\alpha<2$ we have that
 $\Energy_\alpha(P,Q) = 0$ if and only if $P=Q$, while for $\alpha=2$ 
-we have $\Energy_2(P,Q) = 2\| \E(X) - \E(Y) \|^2$ showing that
+we have $\Energy_2(P,Q) = 2\| \E(X) - \E(Y) \|^2$, showing that
 equality to zero only requires
-equality of the means and thus $\Energy_2(P,Q)=0$ does 
+equality of the means, and thus $\Energy_2(P,Q)=0$ does 
 not imply equality of distributions.
@@
 with a \emph{semimetric of negative type}
-$\rho: \mathcal{X}\times\mathcal{X} \to \mathbb{R}$, which satisfies
+$\rho: \mathcal{X}\times\mathcal{X} \to \mathbb{R}$, which is required
+to satisfy
 \begin{equation}
@@
 where $X_i \in \mathcal{X}$ and $\alpha_i \in \mathbb{R}$ such that
-$\sum_{i=1}^n \alpha_i = 0$. Then $\mathcal{X}$ is called a space of
-negative type.
+$\sum_{i=1}^n \alpha_i = 0$. Then, $\mathcal{X}$ is called a \emph{space of
+negative type}.
 We can thus replace $\mathbb{R}^D \to \mathcal{X}$ and 
-$\| X - Y \| \to \rho(X , Y)$ in the definition \eqref{eq:energy} obtaining
+$\| X - Y \| \to \rho(X , Y)$ in the definition \eqref{eq:energy}, obtaining
 the energy distance as
-\begin{equation}\label{eq:energy3}
+\begin{equation}
+\label{eq:energy3}
 \Energy(P, Q) \equiv 2 \E \rho(X,Y) - \E \rho(X, X') - \E \rho(Y,Y').
 \end{equation}
-For spaces of negative type exists a Hilbert space $\mathcal{H}$ and
+For spaces of negative type, there exists a Hilbert space $\mathcal{H}$ and
 a map $\varphi: \mathcal{X} \to
 \mathcal{H}$ such that
-$\rho(X, Y) = \| \varphi(X) - \varphi(Y) \|_{\mathcal{H}}^2$. 
+$\rho(X, Y) = \| \varphi(X) - \varphi(Y) \|_{\mathcal{H}}^2$, which
+allows us to compute quantities related to probability distributions over
+$\mathcal{X}$ in the Hilbert space $\mathcal{H}$.
 Even though the semimetric 
@@
 \end{enumerate}
-In other words, for any $x \in \mathcal{X}$ and any function $f \in \HH$ 
+In other words, for any $x \in \mathcal{X}$ and any function $f \in \HH$,
 there is a unique 
@@
 If such a \emph{kernel} 
-function $\kk$ exists then $\HH$ is called a RKHS. The above two 
+function $\kk$ exists, then $\HH$ is called a RKHS. The above two 
 properties immediately imply that $\kk$ is symmetric and positive
@@
 $\{ \kkk_{x_i} \}_{i=1}^n$ is a basis of $\HH$. It follows that
-$\langle w, w \rangle_{\HH}  = \sum_{i,j=1}^n c_i c_j \kk(x_i,x_j) \ge 0$
+$\langle w, w \rangle_{\HH}  = \sum_{i,j=1}^n c_i c_j \kk(x_i,x_j) \ge 0$,
 showing that the kernel is positive definite. If $G$ is a matrix with
 elements $G_{ij} = \kk(x_i,x_j)$, this is equivalent to $G$ being
-positive semi-definite, $\bm{w}^\top G \bm{w} \ge 0$ for any vector
-$\bm{w} \in \mathbb{R}^n$.
-
-The Moore-Aronszajn theorem establishes the converse \cite{Aronszajn}.
+positive semi-definite: $\bm{v}^\top G \, \bm{v} \ge 0$ for any vector
+$\bm{v} \in \mathbb{R}^n$.
+
+The Moore-Aronszajn theorem 
+\cite{Aronszajn}
+establishes the converse of the above paragraph. 
 For every symmetric
@@
 kernel $\kk$. The map $\varphi: x \mapsto \kkk_x \in \Hk$ is called
-the canonical feature map. Given a kernel $\kk$,
+the canonical \emph{feature map}. Given a kernel $\kk$,
 this theorem enables us to define an embedding of a probability measure
@@
 $\int f(x) d P(x) = \langle f, \kkk_P \rangle$ for all $f \in \Hk$,
-or alternatively, $\kkk_P = \int \kk( \, \cdot \,, x)  d P(x)$. 
+or alternatively, $\kkk_P \equiv \int \kk( \, \cdot \,, x)  d P(x)$. 
 We can now  introduce the 
@@
 \end{equation}
-Therefore, in practice, we can estimate the inner product between the 
+Thus, in practice, we can estimate the inner product between the 
 embedded distributions 
@@
 \label{eq:kernel_semimetric}
-\kk(x,y) = \tfrac{1}{2} \left\{  \rho(x,x_0) + \rho(y,x_0) - \rho(x,y)\right\}.
-\end{equation}
-Thus, $\kk$ is positive semidefinite if and only if $\rho$ is a semimetric
+\kk(x,y) = \tfrac{1}{2} \left[  \rho(x,x_0) + \rho(y,x_0) - \rho(x,y)\right].
+\end{equation}
+Thus, it can be shown that 
+$\kk$ is positive definite if and only if $\rho$ is a semimetric
 of negative type
@@
 $\varphi: x \mapsto \kkk_x$ is injective \cite{Sejdinovic2013}.
-We say that the kernel $\kk$ generates the semimetric $\rho$. 
+When these conditions are satisfied, we say that the kernel $\kk$ 
+generates the semimetric $\rho$. 
 If two different kernels generate the same $\rho$ they are
@@
 replacing \eqref{eq:gen_kernel} into
-\eqref{eq:energy3} and using \eqref{eq:mmd2} yields
+\eqref{eq:energy3}, and using \eqref{eq:mmd2}, yields
 \begin{equation} \label{eq:Erho}
@@
 This test is nonparametric in the sense that it does not make any assumptions
-about the distributions $P_j$. Next we show how we can use $S$ for clustering
-data.
+about the distributions $P_j$.
+
+One can make the analogy 
+that every data point $ x \in \C_j$ form a massive body, 
+whose total mass is characterized by the distribution function $P_j$. 
+Then $S$ is a potential
+energy of the from $S(P_1,\dotsc,P_k)$ which measures how different
+are these
+mass distributions, and achieves the ground state
+$S=0$ when all bodies have the same mass distribution. The potential energy
+$S$ increases as each body has different mass distribution than the other 
+ones.
 
@@
 criterion for clustering data is to 
-maximize $S$, which intuitively makes 
+maximize $S$, which makes 
 each cluster as different
@@
 In other words, given a set of points coming from different probability
-distributions, $S$ should attain a maximum when the points are correctly
-classified since $S$ measures a cost between different clusters.
-However, the following 
+distributions, $S$ should attain a maximum when each point is correctly
+classified as belonging to the cluster associated to its probability
+The following 
 straightforward result
@@
 \begin{proof}
-It can be shown that the total dispersion of the data obeys \cite{Szkely2013}
-\begin{equation}
-W + S = \dfrac{n}{2} g(\mathbb{X}, \mathbb{X}). 
+From \eqref{eq:within} and \eqref{eq:between}, and recall
+that $n=\sum_{i=1}^k n_i$, we have
+\begin{equation}
+\begin{split}
+S + W &= 
+\dfrac{1}{2n} \sum_{\substack{i,j=1 \\ i\ne j}}^k n_i n_j g(\C_i, \C_j)
++ \dfrac{1}{2n} \sum_{i=1}^{k} n_i g(\C_i, \C_i) \left( n - 
+\sum_{j\ne i = 1}^k n_j \right) \\
+& = \dfrac{1}{2n} \sum_{i,j=1}^k n_i n_j g(\C_i, \C_j)
+= \dfrac{1}{2n} \sum_{x \in \mathbb{X}} \sum_{y \in \mathbb{X}} \rho(x,y)
+= \dfrac{n}{2} g(\mathbb{X}, \mathbb{X}).
+\end{split}
 \end{equation}
@@
 independent of the choice of partition. Therefore, maximizing
-$S$ is equivalent to minimizing~$W$.
+$S$ over the choice of partition is equivalent to minimizing~$W$.
 \end{proof}
@@
 the $i$th row of a matrix $M$, and $M_{\bullet j}$ denotes its $j$th column.
-
 Our next result reveals the optimization problem behind \eqref{eq:minimize},
@@
 where $\bm{e} = (1,1,\dots,1)^\top \in \mathbb{R}^n$ is the all-ones vector,
-and $G$ is the kernel matrix \eqref{eq:kernel_matrix}.
+and $G$ is the Gram matrix \eqref{eq:kernel_matrix}.
 \end{proposition}
@@
 \sum_{i=1}^k \lambda_i$. One might then normalize and threshold the rows
-of $Y^\star$, or even better, apply standard $k$-means to the rows of
-$Y$ to provide the final clusters. This procedure is known as spectral
-clustering. Nonetheless, computing eigenvectors of a very large matrix
+of $Y^\star$, or even better, following \cite{NgJordan} we can normalize the
+rows of $Y$ and apply standard $k$-means on this matrix where each
+row is considered as a datapoint.
+This same 
+procedure usually done in spectral clustering on the (normalized) Laplacian
+of the graph defined by a similarity matrix.
+However, computing eigenvectors of a very large matrix
 can be problematic, and usually iterative methods are preferred.
@@
 It is important to note 
-that clustering based on energy statistics
-holds for data living in an \emph{arbitrary space of negative type}.
+that the previous theory for clustering based on energy statistics
+hold for data living in an \emph{arbitrary space of negative type}.
 This clustering method is
@@
 which can be ill-defined for some types of data, such as images for
-instance. 
-Thus, this method is quite general and makes very few
-assumptions about the data.
-
-Although \eqref{eq:qcqp2} is nonparametric, in practice,
+instance, and thus it should also be robust to outliers.
+This method is quite general and makes very few
+assumptions about the data. If one uses the traditional energy distance
+\eqref{eq:energy}, then $\rho(x,y) = \| x - y\|$ and this fixes the kernel
+choice through \eqref{eq:kernel_semimetric}.
+In practice, however,
 the clustering quality strongly depend on the choice of a suitable
@@
 
-Now let us consider the relation to the well-known kernel $k$-means 
-optimization problem.
-For positive semidefinite $G$, there exists a map
+One may wonder how energy statistics clustering 
+relates to the well-known kernel $k$-means problem.
+We now address this question.
+For a positive semidefinite $G$, there exists a map
 $\varphi: \mathcal{X} \to \HH_\kk$ such that
-$\kk(x,y) = \varphi(x)^\top \varphi(y)$. The kernel $k$-means problem,
-in the feature space,
-is given by
+$\kk(x,y) = \varphi(x)^\top \varphi(y)$. The kernel $k$-means optimization
+problem,
+in feature space,
+is defined by
 \begin{equation}
@@
 \min_{\C_1,\dotsc,\C_k}\bigg\{ 
-J(\C_1,\dots,\C_k) \equiv \dfrac{1}{2} \sum_{j=1}^k
+J(\C_1,\dots,\C_k) \equiv  \sum_{j=1}^k
 \sum_{x \in \C_j} \| \varphi(x) - \varphi(\mu_j) \|^2
@@
 where $\mu_j = \tfrac{1}{n_j} \sum_{x \in \C_j} x$ is the  mean of cluster
-$\C_j$.
-It is well-known \cite{Dhillon} that problem \eqref{eq:kernel_kmeans} 
-is equivalent to the QCQP
-\eqref{eq:qcqp2}. The next result makes this explicit, showing its
-relation with the energy statistics based clustering considered thus far.
+$\C_j$ in the ambient space.
+It is known \cite{Dhillon} that problem \eqref{eq:kernel_kmeans} 
+is equivalent to a QCQP in the same form as
+\eqref{eq:qcqp2}. The next result makes this explicit, showing that
+\eqref{eq:minimize} and \eqref{eq:kernel_kmeans} are actually equivalent,
+once the kernel is fixed.
 
@@
 \label{eq:J}
-J = \dfrac{1}{2} \sum_{j=1}^k \sum_{x\in\C_j} \bigg(
+J = \sum_{j=1}^k \sum_{x\in\C_j} \bigg(
 \kk(x,x) - 
@@
 \tfrac{1}{n_j}\sum_{y,z\in\C_j} \kk(y,z)$, which is the same as
-the second term. Thus the optimization problem is
+the second term. Thus the kernel $k$-means optimization problem is
 \begin{equation}
@@
 which is exactly the same as 
-\eqref{eq:max_prob}. The remaining of the proof proceeds as 
-already shown in the proof of Proposition~\ref{th:qcqp2}.
+\eqref{eq:max_prob} from the energy statistics formulation. Therefore,
+once the kernel $\kk$ is fixed, the function 
+$W$ given by \eqref{eq:within} is the same
+as $J$ in \eqref{eq:kernel_kmeans}.
+The remaining of the proof proceeds as 
+already shown in the proof of Proposition~\ref{th:qcqp2}, leading to
+the optimization problem in the form \eqref{eq:qcqp2}.
 \end{proof}
@@
 based clustering is also equivalent to these problems. The advantage
-of energy statistics is that the semimetric $\rho$ fixes the kernel
-through \eqref{eq:kernel_semimetric}.
+of energy statistics compared to kernel $k$-means 
+is that the semimetric $\rho$ is supposed to be fixed. Moreover, from a
+theoretical perspective, it brings the clustering problem into a formal
+statistical theory based on distances between probability distributions
+and embeddings in RKHS. There was no a priori reason to expect that clustering
+based on energy statistics would be equivalent to the kernel $k$-means
+problem~\eqref{eq:kernel_kmeans}.
 
@@
 
-If we choose
-$\rho(x,y) = |x - y|$ we can actually compute 
+Fixing
+$\rho(x,y) = |x - y|$ according to \eqref{eq:energy}, we can actually compute 
 \eqref{eq:g_def} in $\OO(n \log n)$ and find
@@
 \begin{aligned}
-|x - y|  &= \Ind{x \ge y} (x - y) -
-\Ind{x < y} (x - y) \\
+|x - y|  &= (x-y)\Ind{x \ge y} -
+(x-y) \Ind{x < y}  \\
 &= 
-\left( \Ind{x \ge y} - \Ind{x < y} \right) x + 
-\left( \Ind{y > x} - \Ind{y \le x} \right) y ,
+x \left( \Ind{x \ge y} - \Ind{x < y} \right)  + 
+y \left( \Ind{y > x} - \Ind{y \le x} \right)  ,
 \end{aligned}
@@
 \sum_{y \in \C} 
-\left(
+x \left(
 \Ind{x \ge y} + \Ind{y > x} - 
-\Ind{x \ge y}-\Ind{x < y} \right) x.
+\Ind{x \ge y}-\Ind{x < y} \right) .
 \end{equation}
@@
 
-For a two-class problem we can use \eqref{eq:w1d} to cluster data
+For a two-class problem, we can use \eqref{eq:w1d} to cluster data
 through a simple algorithm 
@@
 and pick the point which gives the minimum value of $W$. 
-This procedure is described in Algorithm~\ref{algo1d} below. 
+This procedure is described in Algorithm~\ref{algo1d}. 
 Notice that
@@
 
+\begin{figure}
 \begin{algorithm}[H]\vspace{.5em}
 \begin{algorithmic}[1]
-\INPUT Data $\mathbb{X}$
-\OUTPUT Label matrix $Z$
-\STATE Sort $\mathbb{X}$ obtaining 
+\INPUT data $\mathbb{X}$
+\OUTPUT label matrix $Z$
+\STATE sort $\mathbb{X}$ obtaining 
 $\bar{\mathbb{X}}= [ x_1,\dotsc,x_n ]$
@@
     \ENDFOR
-    \STATE $j^\star \leftarrow \argmin_j W^{(j)}$ determines the best split
-    \STATE $Z_{j\bullet} = (1,0) $ if $j\le j^\star$, and
+    \STATE $j^\star \leftarrow \argmin_j W^{(j)}$ 
+    \STATE $Z_{j\bullet} \leftarrow (1,0) $ if $j\le j^\star$, and
            $Z_{j\bullet} \leftarrow (0,1)$ otherwise, for $j=1,\dotsc,n$
 \end{algorithmic}
-\caption{\label{algo1d}
-Procedure to find an approximate solution to \eqref{eq:qcqp2} 
-for a two-class problem in one dimension and with Euclidean distance.}
+\caption{
+\label{algo1d}
+Approximate solution to \eqref{eq:minimize} 
+for a two-class problem in one dimension. \hspace{\fill}
+}
 \end{algorithm}
+\end{figure}
 
@@
 with GMM, through EM algorithm, and with $k$-means. In both of these
-cases we use the initialization from $k$-means++ and we run the algorithms
-a couple of times with different initializations and choose the answer
+cases we use the initialization from $k$-means++ \cite{Vassilvitskii} 
+and we run the algorithms
+few times with different initializations and choose the answer
 with best objective function value. We use \eqref{eq:accuracy} to measure
-the clustering quality. The results are shown in 
-Fig.~\ref{fig:1d}. In  Fig.~\ref{fig:1d}a 
+the clustering quality. 
+In  Fig.~\ref{fig:1d}a 
 we have data from normal distributions,
 where we can see that all the three methods
-perform closely, with a slight advantage of GMM, as expected since
+perform closely, with a slight advantage of GMM, as expected, since
 it is the right model for the data. However, as shown in Fig~\ref{fig:1d}b,
-for lognormal distributions Algorithm~\ref{algo1d} provides a huge improvement
-compared to both GMM and $k$-means, which basically cluster at chance.
+for lognormal distributions, Algorithm~\ref{algo1d} provides a huge improvement
+compared to both GMM and $k$-means which basically cluster at chance.
 The zero accuracy values for GMM happened when EM algorithm was unable
 to estimate the parameters. These two simple experiments illustrate
-how energy statistics based clustering is nonparametric.
+how energy statistics based clustering is nonparametric, since it is able
+to provide high quality clustering in settings where data comes
+from very different distribtuions.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Iterative Algorithm for Energy Clustering}
+\section{Iterative Algorithm for Energy Statistics Clustering}
 \label{sec:algo}
@@
 In this section we will introduce a new iterative algorithm to find a local
-maximizer of \eqref{eq:qcqp2}, however, due to 
+maximizer of the QCQP \eqref{eq:qcqp2}, however, due to 
 Proposition~\ref{th:kernel_kmeans} we can also find an approximate
-solution by the well-known kernel $k$-means algorithm. Thus, to be
-self-contained let us first recall kernel $k$-means algorithm in this
-context.
-
-\subsection{Kernel $\bm{k}$-Means Algorithm}
-
-Consider this optimization problem as
-written in the form \eqref{eq:max_prob},
+solution by the well-known kernel $k$-means algorithm, which 
+for convenience
+will also be reestated in the present context.
+First, let us introduce some base notation.
+
+Consider the optimization problem 
+written in the form \eqref{eq:max_prob} as follows:
 \begin{equation}
@@
 \end{equation}
-where $Q_j$ represents the internal cost of cluster $\C_j$, and
-$Q$ is the total cost where each cluster cost 
+where $Q_j$ represents an internal energy cost of cluster $\C_j$, and
+$Q$ is the total energy cost where each individual cluster cost 
 is weighted by the inverse
-of the number of its elements. For a data point $x_i$ its cost
-with cluster $\C_j$ is denoted by
+of the number of its elements. For a data point $x_i$ we denote
+its own energy cost
+with the entire cluster $\C_\ell$ by
 \begin{equation}
 \label{eq:costxij}
-Q_j(x_i) \equiv \sum_{y\in\C_j} \kk(x_i, y) = 
-G_{i \bullet} \cdot Z_{\bullet j}.
-\end{equation}
-
-Now for kernel $k$-means 
-consider \eqref{eq:J} where we define
-the function
+Q_\ell(x_i) \equiv \sum_{y\in\C_\ell} \kk(x_i, y) = 
+G_{i \bullet} \cdot Z_{\bullet \ell},
+\end{equation}
+where, we recall, $G_{i\bullet}$ ($G_{\bullet i}$) denotes
+the $i$th row (column) of matrix $G$.
+
+\subsection{Kernel $\bm{k}$-Means Algorithm}
+
+To optimize kernel $k$-means objective function
+\eqref{eq:J}, we remove the global term and define the function
 \begin{equation}
@@
 J^{(\ell)}(x_i) \equiv -\dfrac{2}{n_\ell} Q_\ell(x_i) + \dfrac{1}{n_\ell^2}
-Q_\ell
-\end{equation}
-which represents the cost of $x_i$ with cluster $\C_\ell$. Thus,
-one assigns point $x_i$ to the cluster $\C_{j^\star}$ according
+Q_\ell,
+\end{equation}
+which represents a cost depending on point $x_i$ and cluster $\C_\ell$. One
+thus assigns  $x_i$ to cluster $\C_{j^\star}$ according
 to $j^\star = \argmin_\ell J^{(\ell)}(x_i)$, for $\ell = 1,\dotsc,k$.
@@
 The complete algorithm is shown in Algorithm~\ref{kmeans_algo}.
-Although our formulation looks a little bit different than the standard
+It can be shown that this algorithm converges provided $G$ is positive
+semidefinite.
+Although our notation looks a little different than the standard
 kernel $k$-means found in the literature \cite{Dhillon}, this is precisely
-the same algorithm.
+the same algorithm but written in a more concise and explicit way.
 
@@
 $\OO(n_\ell)$ operations, and although the second term requires
-$\OO(n_\ell^2)$, it only needs to be computed once outside the loop through
-data points. Therefore, the time complexity of this algorithm is
-$\OO(n k \max_\ell n_\ell) = \OO(k n^2)$. If $G$ is sparse with 
+$\OO(n_\ell^2)$, it only needs to be computed once outside the data
+points loop in Algorithm~\ref{kmeans_algo} (step 1).
+Therefore, the time complexity Algorithm~\ref{kmeans_algo} is
+$\OO(n k \max_\ell n_\ell) = \OO(k n^2)$. For a sparse
+Gram matrix $G$ having
 $n'$ nonzero elements, this complexity can be further reduced
@@
 
+\begin{figure}
 \begin{algorithm}[H]
@@
     \OUTPUT label matrix $Z$ 
-  \STATE Let $\bm{q} = (Q_1, \dotsc, Q_k)^\top$ 
+  \STATE $\bm{q} \leftarrow (Q_1, \dotsc, Q_k)^\top$ 
             have the costs of each cluster, according to \eqref{eq:maxQ}
-  \STATE Let $\bm{n} = (n_1,\dotsc,n_k)^\top$ have the number of elements
-        in each cluster
+  \STATE $\bm{n} \leftarrow (n_1,\dotsc,n_k)^\top$ 
+        have the number of points in each cluster, obtained 
+        from $D = Z^\top Z$
   \REPEAT
     \FOR{ $i=1,\dotsc,n$}
-        \STATE Let $j$ be such that $x_i \in \C_j$
+        \STATE let $j$ be such that $x_i \in \C_j$
         \STATE $j^\star \leftarrow \argmin_{\ell} J^{(\ell)}(x_i)$
@@
         \IF{ $j^\star \ne j$} 
-            \STATE Move $x_i$ to $\C_{j^\star}$: $Z_{ij} \leftarrow 0$,
+            \STATE move $x_i$ to $\C_{j^\star}$: $Z_{ij} \leftarrow 0$ and
             $Z_{ij^\star} \leftarrow 1$
-            \STATE Update $\bm{n}$: $n_j \leftarrow n_j - 1$,
+            \STATE update $\bm{n}$: $n_j \leftarrow n_j - 1$ and
                     $n_{j^\star} \leftarrow n_{j^\star} + 1$
-            \STATE Update $\bm{q}$: $q_j \leftarrow q_j - 2Q_j(x_i)$, 
+            \STATE update $\bm{q}$: $q_j \leftarrow q_j - 2Q_j(x_i)$ and
     $q_{j^\star} \leftarrow q_{j^\star} + 2Q_{j^\star}(x_i)$
@@
 Kernel $k$-means algorithm 
-to find an approximate solution to \eqref{eq:qcqp2}.}
+to find an approximate solution to \eqref{eq:qcqp2}.
+\hspace{\fill}
+}
 \end{algorithm}
+\end{figure}
 
@@
 
-Now let us consider a different algorithm, which is based on computing
-the difference in the energy cost when assigning a given data point to
-a different partition.
+Now let us consider a different algorithm, which is based on the change
+in the within energy statistics when moving a given data point to
+a different cluster.
 Suppose we have a data point $x_i \in \mathcal{X}$
-which is currently classified as being in cluster $\C_j$, yielding
-a total cost function \eqref{eq:maxQ} denoted by $Q^{(j)}$.
-Let us consider the change in the total cost by moving
+which is currently assigned to  cluster $\C_j$, yielding
+a total energy cost function \eqref{eq:maxQ} denoted by $Q^{(j)}$.
+Let us consider the change in the total energy cost by moving
 $x_i$ to cluster $\C_\ell$. 
-Denote the new total cost after moving $x_i$ to $\C_\ell$ by $Q^{(\ell)}$.
+Denote the new energy cost after moving $x_i$ to $\C_\ell$ by $Q^{(\ell)}$.
 It is straightforward to see that
@@
 maximum of \eqref{eq:maxQ} by
-moving $x_i$ to $\C_\ell$, otherwise we keep $x_i$ in $\C_j$. Based on
+moving $x_i$ to $\C_\ell$, otherwise we better keep $x_i$ in $\C_j$. Based on
 this we propose an algorithm where
@@
 until convergence, i.e. until no points are assigned to new clusters. 
-This procedure is described in Algorithm~\ref{algo} below.
-
+This whole procedure is described in Algorithm~\ref{algo}.
+Note that \eqref{eq:changeQ} assures that the objective function is
+monotonically increasing at each iteration.
+
+\begin{figure}
 \begin{algorithm}[H]
@@
     \OUTPUT label matrix $Z$
-  \STATE Let $\bm{q} = (Q_1, \dotsc, Q_k)^\top$ 
+  \STATE $\bm{q} \leftarrow (Q_1, \dotsc, Q_k)^\top$ 
             have the energy costs of each cluster, according to \eqref{eq:maxQ}
-  \STATE Let $\bm{n} = (n_1,\dotsc,n_k)^\top$ have the number of elements
-        in each cluster
+  \STATE $\bm{n} \leftarrow (n_1,\dotsc,n_k)^\top$ have the number of points 
+        in each cluster, obtained from $D=Z^\top Z$
   \REPEAT
     \FOR{ $i=1,\dotsc,n$}
-        \STATE Let $j$ be such that $x_i \in \C_j$
+        \STATE let $j$ be such that $x_i \in \C_j$
         \STATE $j^\star \leftarrow \argmax_{\ell} \Delta Q^{j\to \ell}(x_i)$, 
@@
         \IF{ $\Delta Q^{j \to j^\star}(x_i) > 0$ }
-            \STATE Move $x_i$ to $\C_{j^\star}$: $Z_{ij} \leftarrow 0$,
+            \STATE move $x_i$ to $\C_{j^\star}$: $Z_{ij} \leftarrow 0$ and 
             $Z_{ij^\star} \leftarrow 1$
-            \STATE Update $\bm{n}$: $n_j \leftarrow n_j - 1$,
+            \STATE update $\bm{n}$: $n_j \leftarrow n_j - 1$ and
                     $n_{j^\star} \leftarrow n_{j^\star} + 1$
-            \STATE Update $\bm{q}$: $q_j \leftarrow q_j - 2Q_j(x_i)$, 
+            \STATE update $\bm{q}$: $q_j \leftarrow q_j - 2Q_j(x_i)$ and
     $q_{j^\star} \leftarrow q_{j^\star} + 2\left(Q_{j^\star}(x_i)+
@@
 \caption{\label{algo}
-Energy cost algorithm to find an approximate solution to \eqref{eq:qcqp2}.}
+Energy cost algorithm to find an approximate solution to \eqref{eq:qcqp2}.
+\hspace{\fill}
+}
 \end{algorithm}
+\end{figure}
 
@@
 entries of $G$.
-In the numerical experiments below
-we choose the initial $Z$ from the initialization 
-procedure of $k$-means++ \cite{Vassilvitskii}.
 
@@
 
-In the following experiments we fix the semimetric as $\rho(x,y) = \| x-y\|$,
-with induced kernel 
-\eqref{eq:kernel_semimetric}. We will compare Algorithm~\ref{algo} with
+In the experiments below we fix the semimetric 
+according to the traditional energy distance \eqref{eq:energy}, and
+the fixed point $x_0=0$ is chosen in the associated kernel  
+\eqref{eq:kernel_semimetric}:
+\begin{equation}
+\rho(x,y) = \| x-y\|, \qquad \kk(x,y) = 
+\tfrac{1}{2}\left( \| x \| + \| y \| - \| x-y \| \right).
+\end{equation}
+We may consider other semimetrics/kernels as well, 
+but the above will be considered the
+standard kernel for energy statistics and will always be present in every
+experiment. Another usual choice would be
+to use $\alpha\ne 1$ according to \eqref{eq:energy2}.
+
+We will compare Algorithm~\ref{algo} with
 kernel $k$-means Algorithm~\ref{kmeans_algo}, and also with
-standard $k$-means and GMM through the Expectation Maximization (EM) algorithm
-as well. For synthetic data our measure of clustering performance
-will be \eqref{eq:accuracy}. Moreover, for all the algorithms, we always
+standard $k$-means and GMM through the expectation maximization (EM) 
+algorithm
+as well. Since for synthetic data the true labels are available,
+our measure of clustering quality
+will be \eqref{eq:accuracy}. Moreover, for all algorithms, we always
 choose the initialization from $k$-means++ \cite{Vassilvitskii}.
@@
 \begin{split}
-&x \sim \tfrac{1}{2}\left\{ 
-\mathcal{N}(\mu_1,\Sigma_1) + \mathcal{N}(\mu_2, \Sigma_2)\right\}, \\
+&x \sim \tfrac{1}{2}\left[ 
+\mathcal{N}(\mu_1,\Sigma_1) + \mathcal{N}(\mu_2, \Sigma_2)\right], \\
 &\mu_1 = (\underbrace{0,\dotsc,0}_{\times D})^\top , \quad
@@
 \begin{split}
-&x \sim \tfrac{1}{2}\left( 
-\mathcal{N}(\mu_1,\Sigma_1) + \mathcal{N}(\mu_2, \Sigma_2)\right), \\
+&x \sim \tfrac{1}{2}\left[ 
+\mathcal{N}(\mu_1,\Sigma_1) + \mathcal{N}(\mu_2, \Sigma_2)\right], \\
 &\mu_1 = (\underbrace{0,\dotsc,0}_{\times D})^\top , \,
@@
 \label{fig:gauss}
-Experiments where data is normally distributed.
-(a) We keep $100$ points on each cluster
-and increase the ambient dimension, as described in \eqref{eq:gauss1}. The blue
-dots correspond to Algorithm~\ref{algo}, while
-the red dots 
+Two clusters with normally distributed data, with $100$ points in each
+cluster. We analyse the effect of ambient dimension while keeping Bayes
+error fixed.
+(a) 
+We increase $D$ as described in \eqref{eq:gauss1}. The blue
+line correspond to Algorithm~\ref{algo}, while
+the magenta line 
 corresponds to Algorithm~\ref{kmeans_algo}.

### 2017-06-23 qcqp_clustering/code/tests2.py

@@ -144,136 +144,156 @@ def circles_or_spirals(num_points=[200,200], num_times=5, run_times=3):
     return table
 
 def gauss_dimensions_mean(dimensions=range(2,100,20), num_points=[200, 200],
-                          delta=0.7, run_times=5, d=None):
+                          delta=0.7, run_times=5, num_experiments=10, d=None):
+    """Here we keep signal in one dimension and increase the ambient dimension.
+    The covariances are kept fixed.
+    
+    """
     k = 2
     if not d:
         d = dimensions[0]
     n1, n2 = num_points
-    table = np.zeros((len(dimensions), 5))
+    table = np.zeros((num_experiments*len(dimensions), 5))
     count = 0
     
     for D in dimensions:
-        
-        # generate data
-        m1 = np.zeros(D)
-        s1 = np.eye(D)
-        m2 = np.concatenate((delta*np.ones(d), np.zeros(D-d)))
-        s2 = np.eye(D)
-        
-        X, z = data.multivariate_normal([m1, m2], [s1, s2], [n1, n2])
-        rho = lambda x, y: np.linalg.norm(x-y)
-        G = ke.kernel_matrix(X, rho)
-        
-        table[count, 0] = D
-        table[count, 1] = cost_energy(k, X, G, z, run_times=run_times)
-        table[count, 2] = kernel_energy(k, X, G, z, run_times=run_times)
-        table[count, 3] = kmeans(k, X, z, run_times=run_times)
-        table[count, 4] = gmm(k, X, z, run_times=run_times)
 
-        count += 1
+        for i in range(num_experiments):
+        
+            # generate data
+            m1 = np.zeros(D)
+            s1 = np.eye(D)
+            m2 = np.concatenate((delta*np.ones(d), np.zeros(D-d)))
+            s2 = np.eye(D)
+        
+            X, z = data.multivariate_normal([m1, m2], [s1, s2], [n1, n2])
+            rho = lambda x, y: np.linalg.norm(x-y)
+            G = ke.kernel_matrix(X, rho)
+        
+            table[count, 0] = D
+            table[count, 1] = cost_energy(k, X, G, z, run_times=run_times)
+            table[count, 2] = kernel_energy(k, X, G, z, run_times=run_times)
+            table[count, 3] = kmeans(k, X, z, run_times=run_times)
+            table[count, 4] = gmm(k, X, z, run_times=run_times)
+
+            count += 1
 
     return table
 
[…]

### 2017-06-26 qcqp_clustering/draft/energy_clustering.tex
first finished draft

@@
 
-Mention why energy is important, main results, where it was applied, etc.
-Motivate how this can be used for clustering. Mention most important
-papers on this \ldots Explain main results of this paper and give a brief
-outline.
+Energy statistics is based on the energy distance between probability
+distributions, which provides a notion of statistical potential energy
+to statistical observations, in close analogy to Newton's gravitational
+potential energy in classical mechanics. We refer the reader to
+\cite{Szkely2013}, and references therein, for an overview.
+It has been applied to several goodness-of-fit hypothesis tests, 
+multi-sample tests of equality of distributions, analysis of variance
+\cite{RizzoVariance}, and (nonlinear)
+dependence tests through
+distance covariance and distance correlation, which generalizes the Pearson
+correlation coefficient. 
+Moreover, energy statistics  was applied
+to hierarchical clustering \cite{RizzoClustering} by extending Ward's
+method of minimum variance.
+
+More recently, the distance covariance was 
+generalized from Euclidean
+spaces to metric spaces of strong negative type \cite{Lyons}.
+Furthermore, a unifying framework establishing an equivalence
+between generalized energy distances to maximum
+mean discrepancies (MMD), which are distances between embeddings of
+distributions in reproducing kernel Hilbert spaces (RKHS), was
+provided \cite{Sejdinovic2013}. This important work establishes the link
+between techniques commonly used in the statistics literature, regarding
+energy statistics, and techniques
+commonly used in machine learning.
+
+Given a dataset $\mathbb{X} = \{x_1, x_2,\dotsc,x_n \}$, where each
+datapoint lives in a generic space $x_i \in \mathcal{X}$, the $k$-clustering
+problem consists in grouping these points into $k$ groups $\{
+\C_1,\C_2,\dotsc,\C_k \}$, such that
+points belonging to the same group are more ``similar'' to each other
+than to points in  other groups. This intuitive notion already assumes
+a ``metric'' able to measure the similarity between datapoints. Clustering
+is an unsupervised method, and one of the most important problems in machine
+learning since it provides the first step towards automatically constructing
+labels for previously unseen data. The most used algorithms are
+by far $k$-means and gaussian mixture models (GMM) through expectation
+maximization (EM) algorithm. Both are parametric, making 
+strong assumptions about
+the distribution of the data (normality), and the 
+space where data lie is Euclidean, $\mathcal{X} = \mathbb{R}^D$, hence
+the similarity is based on Euclidean metric, 
+$\| x_i - x_j \|^2 = (x_i - x_j)^\top(x_i-x_j) = 
+\sum_{\ell=1}^D (x_{i,\ell} - x_{j,\ell})^2$. These algorithms provide
+a good clustering when data is linearly separable in Euclidean space.
+
+To account for nonlinearly separable data, which may possibly live
+in an arbitrary non-Euclidean space, kernel methods are usually employed.
+If a Mercer kernel $\kk: \mathcal{X}\times \mathcal{X} \to \mathbb{R}$ 
+is available \cite{Mercer}, which is a symmetric and positive definite 
+function, it guarantees that there exists a map $\varphi : \mathcal{X} \to
+\mathcal{H}$, where $\mathcal{H}$ is a Hilbert space. Thus, one can
+compute similarities through the inner product and 
+perform clustering in the feature space $\mathcal{H}$. The nice thing
+about this is that the map $\varphi$ is only used implicitly, i.e. one
+does not need to know $\varphi$, since the inner
+product can be computed only based on the kernel $\kk(\cdot, \cdot)$, 
+a convenient technique known as kernel trick. The well-known 
+kernel $k$-means problem 
+is exactly $k$-means in the feature space
+\cite{Smola,Girolami} (see also \cite{Filippone} for a survey of clustering
+methods). 
+It exploits the nonlinear structure provided by
+a kernel function to perform clustering in datasets that are not linearly
+separable in their original space $\mathcal{X}$.
+As it stands, kernel $k$-means is an heuristic approach in the sense that
+it is not derived from a statistical theory. Moreover, the choice of
+kernel is crucial, and there is no systematic theory for this. One
+must rely on ad hoc methods.
+
+In this paper, we consider how to perform clustering based on energy
+statistics theory. The original goal is to provide a nonparametric
+clustering method, since energy distance is nonparametric.
+The basis of our work is the theory developed in 
+\cite{Sejdinovic2013}. Based on this and on the multi-sample test
+for equality of distributions from energy statistics, we show that
+there is only one function that provides the clustering optimization
+problem. We then show that this clustering
+optimization problem reduces to a quadratically constrained quadratic
+problem (QCQP). This optimization problem has the same form as kernel $k$-means,
+spectral clustering, and some well-known graph partitioning problems
+\cite{Dhillon}. This lead us to show that energy statistics based
+clustering is actually equivalent to kernel $k$-means optimization problem.
+However, energy statistics fixes a kernel to begin with.
+Furthermore, these results can be seen as a first-principle derivation
+of kernel $k$-means from energy statistics, which thus bring
+these kernel based clustering methods into a unifying mathematical theory.
+We also provide an iterative algorithm, different but with the same
+complexity as kernel $k$-means algorithm. The numerical results indicate
+that this method is more accurate and robust.
+
+Our work is organized as follows. In section~\ref{sec:background} we review
+the necessary background on energy statistics and RKHS.
+Section~\ref{sec:clustering_theory} contains the main results of this paper
+where we consider a clustering theory based on energy statistics, leading
+to a QCQP which is NP-hard.
+In section~\ref{sec:twoclass} we consider a simple example in one dimension,
+where we propose an algorithm which requires no initialization.
+In section~\ref{sec:algo} we briefly review kernel $k$-means algorithm,
+and propose a new iterative algorithm to solve this QCQP.
+Section~\ref{sec:numerics} contains some carefully designed numerical
+experiments indicating that this algorithm outperforms kernel
+$k$-means, standard $k$-means, and GMM algorithms.
+Our conclusions are in section~\ref{sec:conclusion}.
 
@@
 to provide high quality clustering in settings where data comes
-from very different distribtuions.
+from very different distributions.
 
@@
 for convenience
-will also be reestated in the present context.
+will also be restated in the present context.
 First, let us introduce some base notation.
@@
 according to the traditional energy distance \eqref{eq:energy}, and
-the fixed point $x_0=0$ is chosen in the associated kernel  
-\eqref{eq:kernel_semimetric}:
-\begin{equation}
+the point $x_0=0$ is chosen in the associated kernel  
+\eqref{eq:kernel_semimetric}. We thus have
+\begin{equation}
+\label{eq:standard_metric}
 \rho(x,y) = \| x-y\|, \qquad \kk(x,y) = 
@@
 \end{equation}
-We may consider other semimetrics/kernels as well, 
+We will consider other semimetrics/kernels as well, 
 but the above will be considered the
 standard kernel for energy statistics and will always be present in every
-experiment. Another usual choice would be
-to use $\alpha\ne 1$ according to \eqref{eq:energy2}.
-
-We will compare Algorithm~\ref{algo} with
-kernel $k$-means Algorithm~\ref{kmeans_algo}, and also with
-standard $k$-means and GMM through the expectation maximization (EM) 
-algorithm
-as well. Since for synthetic data the true labels are available,
+experiment as a reference. Notice that this is a convention, 
+we could have chosen
+any other semimetric as the standard.
+One of the main goals of the following 
+experiments is to compare Algorithm~\ref{algo} to
+kernel $k$-means algorithm, described in Algorithm~\ref{kmeans_algo}. Thus,
+for every kernel used in Algorithm~\ref{algo}, we also use the same 
+kernel in Algorithm~\ref{kmeans_algo}. Another goal is to compare
+Algorithm~\ref{algo} with $k$-means and GMM (through expectation maximization
+algorithm), as these are the most used
+clustering algorithms in practice.
+Since for synthetic data the true labels are available,
 our measure of clustering quality
@@
 how the algorithms degrade as the number of dimensions increase, while
-keeping the number of points in each cluster fixed. 
-In Figure~\ref{fig:gauss}a we have data generated from normal 
-distributions in 
-$D$-dimensions:
+keeping the number of points in each cluster fixed. The Bayes error
+is also kept fixed as ambient dimension increases.
+In Figure~\ref{fig:gauss}a we have data generated from $D$-variate normal 
+distributions as follows: 
 \begin{equation}
@@
 \end{equation}
-For each experiment we only keep signal in
-$\mu_2$ in the first $10$
-dimensions, and keep increasing the ambient dimension $D$. For each
-data set, we run the algorithms $10$ times and pick the result with
-the best objective function value.
-In Figure~\ref{fig:gauss}a one sees that
+We only keep signal in
+in the first $10$
+dimensions of $\mu_2$, and keep increasing the ambient dimension $D$. For each
+$D$, we perform $100$ experiments, obtaining the clustering accuracy
+for each algorithm.
+We can see that
 GMM is not able to estimate the covariance matrix 
 when the number
-of dimensions exceeds the number of points in each cluster. 
-One can see a slightly better performance of Algorithm~\ref{algo} compared
-to the other ones.
+of dimensions exceeds the number of points in each cluster, so it gives
+zero accuracy for $D \gtrsim 100$.
 In Figure~\ref{fig:gauss}b we have the same type of experiment but 
@@
 \end{equation}
+Therefore, for both experiments shown in Figure~\ref{fig:gauss}
+we can see a better performance of Algorithm~\ref{algo} compared
+to the other ones, in particular compared to kernel $k$-means algorithm,
+where we recall that both
+aim at optimizing the same problem (see Proposition~\ref{th:kernel_kmeans}). 
+Also, notice that $k$-means
+and GMM are consistently the right model for this dataset, so it is hard
+to perform better than these algorithms in this current setting.
+Notice that Algorithm~\ref{algo} is more robust as the ambient
+dimension increases.
 
@@
 \centering
-\includegraphics[width=1\textwidth]{gauss_dim.pdf}\\[-.8em]
+\includegraphics[width=1\textwidth]{gauss_dim.pdf}\\[-1.0em]
 (a)
@@
 \centering
-\includegraphics[width=1\textwidth]{gauss_cov.pdf}\\[-.8em]
+\includegraphics[width=1\textwidth]{gauss_cov.pdf}\\[-1.0em]
 (b)
@@
 \label{fig:gauss}
-Two clusters with normally distributed data, with $100$ points in each
-cluster. We analyse the effect of ambient dimension while keeping Bayes
-error fixed.
+Effect of increasing the
+ambient dimension while keeping Bayes error fixed, for 
+two clusters with normally distributed data with $100$ points in each
+cluster. 
 (a) 
@@
 the magenta line 
-corresponds to Algorithm~\ref{kmeans_algo}.
+corresponds to kernel $k$-means, Algorithm~\ref{kmeans_algo}.
 (b) The same but with data following \eqref{eq:gauss2}.
-In both experiments one notice a slightly better accuracy
-of Algorithm~\ref{algo} compared to the other ones.
+One notices that
+Algorithm~\ref{algo} is more robust than the other ones.
 }
@@
 
-In Figure~\ref{fig:unbalanced}a we consider how the previous algorithms
-behave
-for unbalanced clusters. We generate data as
+In Figure~\ref{fig:unbalanced} we consider the effect of having 
+unbalanced clusters. We generate data as
 \begin{equation}
@@
 \end{smallmatrix}\right), \quad
-n_1 = N - m, \quad  n_2 = N + m, \quad N=200,
+n_1 = N - m, \quad  n_2 = N + m, \quad N=200.
 \end{split}
 \end{equation}
-and we increase $m$, i.e. we make the clusters progressively more unbalanced.
-As well-known, one an see that GMM
-works well for unbalanced clusters, mostly because 
-it is a soft clustering algorithm.
-We can see that all the other kernel based methods
-degrade more rapidly than GMM for
-highly unbalanced clusters.
-This is
-unsurprising since Algorithms~\ref{algo}~and~\ref{kmeans_algo} both 
-make hard assignments. Finally, Figure~\ref{fig:unbalanced}b considers
-the same experiment as in Figure~\ref{fig:gauss}a but for lognormal
-data. With $\rho(x,y) = \| x-y\|^{1/2}$ we even gain a slight improvement.
-Energy statistics clustering outperforms all the other methods since
-it is nonparametric.
+We then increase $m$, i.e. we make the clusters progressively more unbalanced.
+We generate $100$ experiments for each $m$, and plot the clustering accuracy
+versus $m$.
+As expected, GMM
+works better than the other algorithms in the case of unbalanced clusters. 
+This is mostly due to its soft assignments.
+We can see that the other methods based on hard assignments degrade similarly,
+and  more rapidly than GMM. This indicates that a fuzzy version of
+energy statistics clustering should compensate for this effect.
+
+\begin{figure}
+\centering
+\includegraphics[width=0.5\textwidth]{gauss_pi.pdf}
+\caption{
+\label{fig:unbalanced}
+Previous algorithms for unbalanced clusters, 
+according to \eqref{eq:gauss3}.}
+\end{figure}
+
+Now, besides \eqref{eq:standard_metric} we consider two other semimetrics:
+\begin{align}
+\rho_{1/2}(x,y) &= \| x-y \|^{1/2}, & \kk(x,y) &= \tfrac{1}{2} \left( 
+\| x \|^{1/2} + \| y \|^{1/2} 
+- \| x-y \|^{1/2} \right), \label{eq:rhohalf}\\
+\rho_{e}(x,y) &= 2 - 2 e^{-\| x- y\|/2}, & \kk(x,y) &= e^{-\| x-y\|/2}.
+\label{eq:rhoe}
+\end{align}
+In Figure~\ref{fig:consist}a we have data in $20$ dimensions distributed as
+\begin{equation}
+\label{eq:20gauss}
+\begin{split}
+x &\sim \tfrac{1}{2}\left[ \mathcal{N}(\mu_1,\Sigma_1) + \mathcal{N}(\mu_2,
+\Sigma_2) \right], \\
+\mu_1 &= (\underbrace{0,\dotsc,0}_{\times 20})^\top ,\quad
+\mu_2 = \tfrac{1}{2} 
+(\underbrace{1,\dotsc,1}_{5},\underbrace{0,\dotsc,0}_{15})^\top, \quad
+\Sigma_1 = \tfrac{1}{2} I_{20},  \quad
+\Sigma_2 = I_{20}.
+\end{split}
+\end{equation}
+We increase the number of points in each cluster and show the clustering
+accuracy with different algorithms. The new semimetrics \eqref{eq:rhohalf}
+and $\eqref{eq:rhoe}$ are indicated in the legend. One can see that
+Algorithm~\ref{algo} performs better than all the other ones, and in 
+particular \eqref{eq:rhoe} provides better results.
+As the number of datapoints 
+get large enough, GMM starts to be as accurate as clustering
+based on energy statistics, as it should since it is 
+consistent model to the data. In Figure~\ref{fig:consist}b, however, we
+use the same parameters as in \eqref{eq:20gauss} but now with data
+log-normally distributed:
+\begin{equation}
+\label{eq:20loggauss}
+x \sim \tfrac{1}{2}\left[ e^{\mathcal{N}(\mu_1,\Sigma_1)}
++ e^{\mathcal{N}(\mu_2, \Sigma_2)}\right].
+\end{equation}
+We see that clustering based on energy statistics still performs accurately
+for this kind of data, while $k$-means works a little bit better than chance,
+and GMM is not even able to estimate the parameters. Again, \eqref{eq:rhoe}
+provides slightly better results than \eqref{eq:standard_metric} or
+\eqref{eq:rhohalf}. Notice also that Algorithm~\ref{algo} performs better
+than Algorithm~\ref{kmeans_algo}. Both experiments in Figure~\ref{fig:consist}
+shows that energy statistics clustering is nonparametric, since it is able
+to cluster data coming from very different distributions.
 
@@
 \centering
-\includegraphics[width=1\textwidth]{gauss_pi.pdf}\\[-.8em]
+\includegraphics[width=1\textwidth]{gauss.pdf}\\[-1.0em]
 (a)
@@
 \centering
-\includegraphics[width=1\textwidth]{loggauss_dim.pdf}\\[-.8em]
+\includegraphics[width=1\textwidth]{loggauss.pdf}\\[-1.0em]
 (b)
@@
 \caption{
-\label{fig:unbalanced}
-(a) Previous algorithms for unbalanced clusters, 
-according to \eqref{eq:gauss3}.
-(b) The same experiment as in Figure~\ref{fig:gauss}a but for lognormal
-data as 
-$x\sim \tfrac{1}{2}\left( e^{\mathcal{N}(\mu_1,\Sigma_1)} +
-e^{\mathcal{N}(\mu_2,\Sigma_2)} \right)$ with
-$\mu_1 = (0,\dotsc,0)^\top$,
-$\mu_2 = 0.5\times(1,\dotsc,1,0,\dotsc,0)^\top$ with signal in $d=10$, 
-$\Sigma_1 = 0.3 \times I_D$, and $\Sigma_2=I_D$. 
-One see a clear advantage of
-energy statistics clustering. The extra green line correspond
-to $\rho(x,y) = \| x-y\|^{1/2}$.
+\label{fig:consist}
+(a) Data normally distributed as in 
+\eqref{eq:20gauss}. We increase the number of points in each cluster to
+illustrate the statistical consistency of the algorithms.
+(b) The same experiment but for data following \eqref{eq:20loggauss}.
+In both experiments, for each case we run every algorithm $100$ times and show
+the average results. One can see the better performance of energy statistics
+clustering, Algorithm~\ref{algo}, and in particular by using the semimetric
+\eqref{eq:rhoe}. These two figures illustrate that energy statistics clustering
+is nonparametric since it works well for very different distributions.
 }
@@
 
-In Figure~\ref{fig:weird} we consider other types of data in two dimensions.
-In each case we perform 10 experiments and compute the accuracy
-based on the true labels. The plots show the average results of 10 runs
-of each experiment (dots), and the maximum and minimum values as well (shaded
-area). Although energy clustering is nonparametric, these datasets are
-very specific so we can take advantage and make a suitable choice of
-the semimetric $\rho$. This illustrates that energy clustering is also
-flexible enough to incorporate prior information about the data.
-In Figure~\ref{fig:weird}a we have parallel cigars from
-\begin{equation}
-\label{eq:cigar}
-\begin{split}
-&x \sim \tfrac{1}{2}\left( 
-\mathcal{N}(\mu_1, \Sigma_1) + 
-\mathcal{N}(\mu_2, \Sigma_2)
-\right), \\
-& \mu_1 = (0,0)^\top, \quad \mu_2 = (6.5, 0)^\top, \quad
-\Sigma_1 = \Sigma_2 = \left( \begin{smallmatrix} 
-1 & 0 \\ 0 & 20
-\end{smallmatrix}\right)
-\end{split}
-\end{equation}
-with $100$ points in each cluster. Notice that $k$-means is unable to
-cluster this data, while energy clustering performs pretty much as well
-as GMM, which is the optimal algorithm suited to this kind of data.
-We used the semimetric $\rho(x,y) = \| x -  y\|^{1/2}$ in this case.
-In Figure~\ref{fig:weird}b we generate two concentric circles with a small
-gaussian noise. To make the semimetric a little bit more localized we
-include a decaying exponential in the form:
-\begin{equation}
-\label{eq:rho_loc}
-\rho(x,y) = \| x-y \|^2 e^{-\tfrac{1}{2}\| x-y \|}.
-\end{equation}
-Notice that energy clustering is able to cluster this data almost perfectly
-in most cases.
-In Figure~\ref{fig:weird}c we consider two spirals, also with a small
-gaussian noise. Due to the geometry of the data
-we consider
-\begin{equation}
-\label{eq:rho_sin}
-\rho(x,y) = \| x - y\|^2 \theta \sin \left(  \tfrac{\| x-y \|}{\theta}\right),
-\qquad \theta = 0.9.
-\end{equation}
-Again, energy clustering performs much better than $k$-means and GMM which
-are unable to cluster this kind of dataset.
-
-\begin{figure}
-\begin{minipage}{0.33\textwidth}
-\hspace{.7cm}\includegraphics[width=.82\textwidth]{cigar_data.pdf}
-\end{minipage}
-\begin{minipage}{0.33\textwidth}
-\hspace{.7cm}\includegraphics[width=.82\textwidth]{circles_data.pdf}
-\end{minipage}
-\begin{minipage}{0.33\textwidth}
-\hspace{.7cm}\includegraphics[width=.82\textwidth]{spirals_data.pdf}
-\end{minipage}
-\begin{minipage}{0.33\textwidth}
-\centering
-\includegraphics[width=\textwidth]{cigars.pdf}\\[-.8em]
-(a)
-\end{minipage}
-\begin{minipage}{0.33\textwidth}
-\centering
-\includegraphics[width=\textwidth]{circles.pdf}\\[-.8em]
-(b)
-\end{minipage}
-\begin{minipage}{0.33\textwidth}
-\centering
-\includegraphics[width=\textwidth]{spirals.pdf}\\[-.8em]
-(c)
-\end{minipage}
-\caption{
-\label{fig:weird}
-(a) Parallel cigars generated from \eqref{eq:cigar} with $100$ points
-in each class. Use use energy clustering
-with $\rho = \| x - y\|^{1/2}$. (b) Two concentric circles with $150$ points
-in each class. We use the semimetric \eqref{eq:rho_loc}. 
-(c) Two concentric spirals with
-$150$ points in each class. We use the semimetric \eqref{eq:rho_sin}.
-}
-\end{figure}
-
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Conclusion}
-
-\subsection*{Acknowledgements}
-We thank \ldots
+\label{sec:conclusion}
+
+In this paper we have considered clustering from the perspective of energy
+statistics, which provides a nonparametric test for equality of distributions.
+Based on this, we showed that the clustering problem reduces to a quadratically
+constrained optimization problem (QCQP), 
+as described in Proposition~\ref{th:qcqp2}.
+Moreover, we showed that clustering based on energy statistics is equivalent
+to kernel $k$-means optimization problem, once the kernel is fixed; see
+Proposition~\ref{th:kernel_kmeans}. Our results imply that kernel $k$-means
+approach to clustering is actually a consequence of energy statistics
+theory, and thus place this method into a principled statistical basis.
+As already known \cite{Dhillon}, this approach is related to spectral
+clustering, and graph partitioning problems. Therefore, all these problems 
+may be seen as arising naturally from energy statistics clustering.
+It is important to mention that energy statistics clustering, as formulated
+here, is valid for
+arbitrary metric spaces of negative type, and makes no assumptions about
+the distribution of the data. Moreover, it does not rely on the concept of a
+cluster mean, even implicitly.
+
+We also proposed Algorithm~\ref{algo} as an alternative to the well-known
+kernel $k$-means algorithm (see Algorithm~\ref{kmeans_algo}), 
+where both have the same time
+complexity. The numerical results show that Algorithm~\ref{algo}
+might provide better clustering accuracy and is more robust 
+than kernel $k$-means algorithm. Since there exists a huge literature about
+kernel $k$-means, and approximation methods to make it faster, with
+applications to several artificial and real data, 
+we limited ourselves to analyze few but carefully designed experiments, which
+illustrates the advantages of Algorithm~\ref{algo}.
+
+\bigskip
+
+%\subsection*{Acknowledgements}
+%We thank \ldots
 

### 2017-06-26 qcqp_clustering/code/pplotcust.py

@@ -21,7 +21,8 @@ inches_per_pt = 1.0/72.27
 golden_mean = (np.sqrt(5.0)-1.0)/2.0
 fig_width = fig_width_pt*inches_per_pt
 #fig_height = fig_width
-fig_height = fig_width*golden_mean
+#fig_height = fig_width*golden_mean
+fig_height = fig_width
 pylab.rc('figure', figsize=(fig_width, fig_height))
 
 

### 2017-07-18 qcqp_clustering/draft/energy_clustering.tex
still working on draft

@@
 \usepackage{algorithm}
+\usepackage[none]{hyphenat}
 
@@
 inspired by the Newtonian gravitational potential from classical mechanics. 
-The idea
-is to associate a statistical potential energy to observations, such that 
-minimum energy is achieved under the null hypothesis of equality of 
-distributions. Energy statistics
+Energy statistics
 was further generalized to probability 
 distributions on arbitrary metric spaces,
-and more recently, a connection with kernels in RKHS was established.
-Nevertheless, although extensively used by the statistics community, energy
-statistics has not been
-previously incorporated in machine learning problems.
+and more recently a connection with reproducing kernel Hilbert spaces 
+was established.
+Nevertheless, although extensively used by the statistics community, it
+has not been previously incorporated in machine learning problems.
 In this paper, we consider the problem of clustering data from
 an energy statistics theory perspective.
-We provide a precise mathematical formulation, yielding
-a quadratically constrained quadratic optimization problem (QCQP). 
-We show that
-this is equivalent to kernel $k$-means
-optimization problem, however,
-energy statistics is able to fix the kernel choice. 
-This clustering 
-method is nonparametric, and if prior information is available
+We provide a precise mathematical formulation yielding
+a quadratically constrained quadratic program (QCQP).
+We show the equivalence between the energy statistics clustering
+formulation 
+and the kernel $k$-means optimization problem (not to be confused with
+kernel $k$-means algorithm). 
+Therefore, our results imply a first principles derivation of 
+kernel $k$-means problem
+from energy statistics theory, thus bringing this important 
+clustering method into a formal statistical framework.
+Moreover, energy statistics is nonparametric, it usually fixes
+the kernel choice,
+and if prior information is available
 it can be easily incorporated in the kernel construction.
 We propose an iterative algorithm
-to find local optimizers of this QCQP problem, based on the change
-in the statistical energy 
+to find local optimizers of the aforementioned QCQP based on the 
+energy change
 of moving points to different clusters. This algorithm 
@@
 algorithm.
-We then compare this algorithm with
-well-known kernel $k$-means, standard $k$-means, and GMM algorithms by 
+We compare it to
+kernel $k$-means, standard $k$-means, and GMM/EM algorithms by 
 providing carefully designed numerical experiments. The results
-show that, in general, this method outperforms
+show that, in general, this algorithm outperforms
 these most used clustering algorithms.
@@
 to statistical observations, in close analogy to Newton's gravitational
-potential energy in classical mechanics. We refer the reader to
-\cite{Szkely2013}, and references therein, for an overview.
-It has been applied to several goodness-of-fit hypothesis tests, 
+potential in classical mechanics. 
+It provides a nonparametric test for equality of distribution,
+such that minimum energy is achieved under equality of 
+distributions. 
+We refer the reader 
+to \cite{Szkely2013}, and references therein, for an overview.
+Energy statistics has been applied to several goodness-of-fit 
+hypothesis tests, 
 multi-sample tests of equality of distributions, analysis of variance
@@
 
-More recently, the distance covariance was 
+More recently, distance covariance was 
 generalized from Euclidean
@@
 distributions in reproducing kernel Hilbert spaces (RKHS), was
-provided \cite{Sejdinovic2013}. This important work establishes the link
+established \cite{Sejdinovic2013}. This important work provides the missing
+link
 between techniques commonly used in the statistics literature, regarding
@@
 
-Given a dataset $\mathbb{X} = \{x_1, x_2,\dotsc,x_n \}$, where each
-datapoint lives in a generic space $x_i \in \mathcal{X}$, the $k$-clustering
+Given a dataset $\mathbb{X} = \{x_1,\dotsc,x_n \}$, where each
+data point $x_i$ lives in a space $\mathcal{X}$, the clustering
 problem consists in grouping these points into $k$ groups $\{
@@
 than to points in  other groups. This intuitive notion already assumes
-a ``metric'' able to measure the similarity between datapoints. Clustering
+a ``metric'' able to measure the similarity between points. Clustering
 is an unsupervised method, and one of the most important problems in machine
 learning since it provides the first step towards automatically constructing
-labels for previously unseen data. The most used algorithms are
-by far $k$-means and gaussian mixture models (GMM) through expectation
-maximization (EM) algorithm. Both are parametric, making 
+labels for previously unseen data. In practice
+$k$-means is arguably the most used algorithm, and gaussian 
+mixture models (GMM) through the expectation
+maximization (EM) algorithm is also often used. Both are parametric, making 
 strong assumptions about
-the distribution of the data (normality), and the 
-space where data lie is Euclidean, $\mathcal{X} = \mathbb{R}^D$, hence
+the distribution of the data (isotropy and normality), and it assumes
+that data points
+lie in Euclidean space, $\mathcal{X} = \mathbb{R}^D$, hence
 the similarity is based on Euclidean metric, 
@@
 \sum_{\ell=1}^D (x_{i,\ell} - x_{j,\ell})^2$. These algorithms provide
-a good clustering when data is linearly separable in Euclidean space.
-
-To account for nonlinearly separable data, which may possibly live
+a good quality clustering when data is linearly separable in Euclidean space,
+i.e. there is a clear resolution in the cluster centers
+and points from different clusters do not overlap considerably.
+
+To account for nonlinearly separable data, which may live
 in an arbitrary non-Euclidean space, kernel methods are usually employed.
-If a Mercer kernel $\kk: \mathcal{X}\times \mathcal{X} \to \mathbb{R}$ 
-is available \cite{Mercer}, which is a symmetric and positive definite 
+If a so-called Mercer kernel $\kk: \mathcal{X}\times \mathcal{X} \to \mathbb{R}$
+is available, which is a symmetric and positive definite 
 function, it guarantees that there exists a map $\varphi : \mathcal{X} \to
-\mathcal{H}$, where $\mathcal{H}$ is a Hilbert space. Thus, one can
-compute similarities through the inner product and 
-perform clustering in the feature space $\mathcal{H}$. The nice thing
-about this is that the map $\varphi$ is only used implicitly, i.e. one
+\mathcal{H}$ where $\mathcal{H}$ is a Hilbert space \cite{Mercer}. 
+One can thus
+compute similarities between data point through the inner product of
+the feature space $\mathcal{H}$ and 
+perform clustering in $\mathcal{H}$. The convenience of a Mercer
+kernel is that the map $\varphi$ is only used implicitly, i.e. one
 does not need to know $\varphi$, since the inner
-product can be computed only based on the kernel $\kk(\cdot, \cdot)$, 
-a convenient technique known as kernel trick. The well-known 
-kernel $k$-means problem 
+of product of $\mathcal{H}$ can be computed only 
+based on the kernel function $\kk(\cdot, \cdot)$ which acts on the original
+space $\mathcal{X}$. 
+This technique is known as the kernel trick.  
+Kernel $k$-means problem 
 is exactly $k$-means in the feature space
@@
 methods).
-It exploits the nonlinear structure provided by
-a kernel function to perform clustering in datasets that are not linearly
-separable in their original space $\mathcal{X}$.
-As it stands, kernel $k$-means is an heuristic approach in the sense that
-it is not derived from a statistical theory. Moreover, the choice of
-kernel is crucial, and there is no systematic theory for this. One
-must rely on ad hoc methods.
-
-In this paper, we consider how to perform clustering based on energy
-statistics theory. The original goal is to provide a nonparametric
+To cluster data that is not linearly
+separable in $\mathcal{X}$,
+kernel based methods exploit the nonlinear structure provided by
+the kernel function, which implicitly maps data points to a usually higher
+dimensional feature 
+space, with the hope that they becomes separable by 
+forming well-defined groups.
+The choice of
+kernel is crucial, and yet there is no principled method
+for this. One must rely on ad hoc methods to find an appropriate kernel
+which is data dependent.
+As it stands, kernel $k$-means problem is an heuristic approach to clustering
+in the sense that it is not derived from a statistical theory.
+
+The main contribution of this paper is to provide a consistent formulation
+to clustering based on energy statistics theory. 
+The original goal is to provide a nonparametric
 clustering method, since energy distance is nonparametric.
 The basis of our work is the theory developed in 
-\cite{Sejdinovic2013}. Based on this and on the multi-sample test
-for equality of distributions from energy statistics, we show that
-there is only one function that provides the clustering optimization
-problem. We then show that this clustering
-optimization problem reduces to a quadratically constrained quadratic
-problem (QCQP). This optimization problem has the same form as kernel $k$-means,
+\cite{Sejdinovic2013} which establishes an equivalence between energy
+distance and RKHS. 
+Based on this and on the multi-sample test
+for equality of distributions, we show that
+there is only one possible clustering optimization
+problem consistent with the energy statistics formulation. 
+We then show that this 
+optimization problem is equivalent to a quadratically constrained quadratic
+program (QCQP), which has the same trace maximization
+form as kernel $k$-means,
 spectral clustering, and some well-known graph partitioning problems
-\cite{Dhillon}. This lead us to show that energy statistics based
-clustering is actually equivalent to kernel $k$-means optimization problem.
-However, energy statistics fixes a kernel to begin with.
-Furthermore, these results can be seen as a first-principle derivation
-of kernel $k$-means from energy statistics, which thus bring
-these kernel based clustering methods into a unifying mathematical theory.
-We also provide an iterative algorithm, different but with the same
-complexity as kernel $k$-means algorithm. The numerical results indicate
-that this method is more accurate and robust.
+\cite{Dhillon2,Dhillon}. We show the equivalence between this approach
+and kernel $k$-means optimization problem, implying that
+the later can actually be derived from energy statistics, thus
+bringing this important clustering method into a unified statistical theory.
+However, it is important to note that 
+energy statistics fixes a kernel to begin with.
+From an algorithmic perspective, our contribution is to 
+provide an iterative algorithm, which is different but has the same
+complexity as kernel $k$-means algorithm. Our numerical results
+provide compelling evidence that this algorithm 
+is more accurate and robust than kernel $k$-means algorithm.
+Moreover, these experiments 
+puts in evidence the nonparametric aspect of energy
+statistics based clustering, since it is able to perform accurately
+on data having very different distributions, contrary to $k$-means and
+GMM for instance.
 
@@
 the necessary background on energy statistics and RKHS.
-Section~\ref{sec:clustering_theory} contains the main results of this paper
+Section~\ref{sec:clustering_theory} contains the main theoretical 
+results of this paper,
 where we consider a clustering theory based on energy statistics, leading
@@
 experiments indicating that this algorithm outperforms kernel
-$k$-means, standard $k$-means, and GMM algorithms.
-Our conclusions are in section~\ref{sec:conclusion}.
+$k$-means, standard $k$-means, and GMM/EM algorithms.
+Our final conclusions are presented in section~\ref{sec:conclusion}.
 
@@
 For more details we refer the reader
-to \cite{Szkely2013} (and references therein) and 
-also \cite{Sejdinovic2013}.
+to \cite{Szkely2013} and \cite{Sejdinovic2013}.
 
@@
 to zero holds if and only if $P = Q$.
-Above $\| \cdot \|$ denotes the
+Above, $\| \cdot \|$ denotes the
 Euclidean norm in $\mathbb{R}^D$. 
@@
 mention that \eqref{eq:energy2} can be even further generalized.
-Let $X, Y \in \mathcal{X}$,  where $\mathcal{X}$ is a space endowed
+Let $X, Y \in \mathcal{X}$,  where $\mathcal{X}$ is an arbitrary space endowed
 with a \emph{semimetric of negative type}
@@
 \label{eq:negative_type}
-\sum_{i,j=1}^n \alpha_i \alpha_j \rho(X_i, X_j) \le 0,
-\end{equation}
-where $X_i \in \mathcal{X}$ and $\alpha_i \in \mathbb{R}$ such that
-$\sum_{i=1}^n \alpha_i = 0$. Then, $\mathcal{X}$ is called a \emph{space of
+\sum_{i,j=1}^n c_i c_j \rho(X_i, X_j) \le 0,
+\end{equation}
+where $X_i \in \mathcal{X}$, and $c_i \in \mathbb{R}$ such that
+$\sum_{i=1}^n c_i = 0$. Then, $\mathcal{X}$ is called a \emph{space of
 negative type}.
@@
 $\| X - Y \| \to \rho(X , Y)$ in the definition \eqref{eq:energy}, obtaining
-the energy distance as
+the generalized energy distance
 \begin{equation}
@@
 This equivalence was established
-in \cite{Sejdinovic2013}. We first recall the definition of
+in \cite{Sejdinovic2013}. Let us first recall the definition of
 RKHS. Let $\HH$ be a Hilbert space of real-valued functions
@@
 elements $G_{ij} = \kk(x_i,x_j)$, this is equivalent to $G$ being
-positive semi-definite: $\bm{v}^\top G \, \bm{v} \ge 0$ for any vector
-$\bm{v} \in \mathbb{R}^n$.
+positive semidefinite; $v^\top G \, v \ge 0$ for any vector
+$v \in \mathbb{R}^n$.
 
@@
 this theorem enables us to define an embedding of a probability measure
-$P$ into the RKHS: $P \mapsto \kkk_P \in
+$P$ into the RKHS as follows: $P \mapsto \kkk_P \in
 \Hk$ such that 
@@
 \end{equation}
-Thus, in practice, we can estimate the inner product between the 
+Thus, in practice, we can estimate the inner product between  
 embedded distributions 
@@
 \label{eq:kernel_semimetric}
-\kk(x,y) = \tfrac{1}{2} \left[  \rho(x,x_0) + \rho(y,x_0) - \rho(x,y)\right].
-\end{equation}
-Thus, it can be shown that 
+\kk(x,y) \equiv 
+\tfrac{1}{2} \left[  \rho(x,x_0) + \rho(y,x_0) - \rho(x,y)\right].
+\end{equation}
+Then, it can be shown that 
 $\kk$ is positive definite if and only if $\rho$ is a semimetric
@@
 
-Finally, let us recall the main formulas for test statistics
+Finally, let us recall the main formulas for test statistic
 of equality of distributions \cite{Szkely2013}. 
 Assume we have data $\mathbb{X} = \{ x_1,\dotsc, x_n \}$, where
-$x_i \in \mathcal{X}$ and $\mathcal{X}$ is a space of negative type.
+$x_i \in \mathcal{X}$, and $\mathcal{X}$ is a space of negative type.
 Consider a partition $\mathbb{X} = \bigcup_{j=1}^k \C_j$, with
@@
 g(\C_j, \C_j)
-\right].
-\end{equation}
+\right],
+\end{equation}
+where $n = \sum_{j=1}^k n_j$.
 Given a set of distributions
@@
 the quantity $S$ provides
-a \emph{nonparametric} test statistic for equality of distributions
+a \emph{nonparametric test statistic} for equality of distributions
 \cite{Szkely2013}.
@@
 whose total mass is characterized by the distribution function $P_j$.
-Then $S$ is a potential
+The quantity $S$ is thus a potential
 energy of the from $S(P_1,\dotsc,P_k)$ which measures how different
-are these
-mass distributions, and achieves the ground state
+these
+mass distributions are, and achieves the ground state
 $S=0$ when all bodies have the same mass distribution. The potential energy
-$S$ increases as each body has different mass distribution than the other 
-ones.
+$S$ increases as bodies have different mass distributions.
 
@@
 
-This section contains the main results of this paper, where 
+This section contains the main theoretical results of this paper, where 
 we formulate an optimization problem for clustering 
@@
 classified as belonging to the cluster associated to its probability
+distribution.
 The following 
 straightforward result
-shows that maximizing \eqref{eq:between} is equivalent to minimizing
-\eqref{eq:within}, which has a more convenient form.
+shows that maximizing \eqref{eq:between} is, however, equivalent to minimizing
+\eqref{eq:within} which has a more convenient form.
 
@@
 \begin{proof}
-From \eqref{eq:within} and \eqref{eq:between}, and recall
-that $n=\sum_{i=1}^k n_i$, we have
+From \eqref{eq:within} and \eqref{eq:between}
+we have
 \begin{equation}
@@
 \dfrac{1}{2n} \sum_{\substack{i,j=1 \\ i\ne j}}^k n_i n_j g(\C_i, \C_j)
-+ \dfrac{1}{2n} \sum_{i=1}^{k} n_i g(\C_i, \C_i) \left( n - 
-\sum_{j\ne i = 1}^k n_j \right) \\
++ \dfrac{1}{2n} \sum_{i=1}^{k} 
+\bigg[ n - 
+\sum_{j\ne i = 1}^k n_j \bigg] 
+n_i g(\C_i, \C_i) 
+\\
 & = \dfrac{1}{2n} \sum_{i,j=1}^k n_i n_j g(\C_i, \C_j)
@@
 
-Thus, for a given $k$, the clustering problem amounts to
+Therefore, for a given $k$, the clustering problem amounts to
 finding the best partition of the data by solving~\eqref{eq:minimize}.
-Notice that this is a hard assignment clustering problem.
+Notice that this is a hard assignment clustering problem as partitions
+are disjoint.
 
@@
 each point (row) belongs to. This matrix satisfy
-$Z^\top Z = D$ where $D = \diag( n_1,\dotsc, n_k )$  contains
-the number of points in each cluster. We also introduce the rescaled
+$Z^\top Z = D$ where the diagonal matrix 
+$D = \diag( n_1,\dotsc, n_k )$  contains
+the number of points in each cluster. Let us also introduce the rescaled
 matrix  $Y \equiv Z D^{-1/2}$. In component form they are given by
@@
 which is NP-hard since
-it is a quadratically constrained quadratic problem (QCQP).
+it is a quadratically constrained quadratic program (QCQP).
 
@@
 where we used the definitions \eqref{eq:kernel_matrix} and
-\eqref{eq:label_matrix}. Thus the objective function in 
+\eqref{eq:label_matrix}. Thus, the objective function in 
 \eqref{eq:max_prob} is equal to
@@
 of the trace, and by the own definition of the matrix $Z$
-in \eqref{eq:label_matrix} we obtain the following integer
+in \eqref{eq:label_matrix}, we obtain the following integer
 programing problem:
@@
 \end{equation}
+
 Now we write this in terms of the matrix $Y = Z D^{-1/2}$.
@@
 that $Y D^{1/2} \bm{e} = Z \bm{e} = \bm{e}$, we finally obtain
-$Y Y^\top \bm{e} = \bm{e}$. Thus, the optimization 
+$Y Y^\top \bm{e} = \bm{e}$. Therefore, the optimization 
 problem \eqref{eq:qcqp} is equivalent
@@
 However, one can find approximate solutions by relaxing some 
-of the constraints, or obtaining a relaxed SDP version of the problem. 
+of the constraints, or obtaining a relaxed SDP version of it.
 For instance, the relaxed problem
@@
 \end{equation}
-has a well-known closed form solution given by $Y^\star = U R$, where the
+has a well-known closed form solution $Y^\star = U R$, where the
 columns of $U$ contain the leading $k$ eigenvectors of $G$ corresponding
-to the $k$ largest eigenvalues $\{ \lambda_1,\dotsc,\lambda_k \}$, and
+to the $k$ largest eigenvalues $\lambda_1\ge \lambda_2\ge\dotsc\ge\lambda_k$, 
+and
 $R \in \mathbb{R}^{k\times k}$ is an arbitrary orthogonal matrix. 
 The resulting
-optimal objective function is thus given by
+optimal objective function is given by
 $\max \Tr \left( {Y^\star}^\top G \, Y^\star \right)  = 
@@
 of $Y^\star$, or even better, following \cite{NgJordan} we can normalize the
-rows of $Y$ and apply standard $k$-means on this matrix where each
-row is considered as a datapoint.
-This same 
-procedure usually done in spectral clustering on the (normalized) Laplacian
+rows of $Y^\star$ and apply standard $k$-means on this matrix where each
+row is considered as a data point.
+This is the same 
+procedure done in spectral clustering on the (normalized) Laplacian
 of the graph defined by a similarity matrix.
@@
 can be problematic, and usually iterative methods are preferred.
+We will propose an iterative method to find approximate solutions
+to \eqref{eq:qcqp2} in section~\ref{sec:algo}.
 
@@
 that the previous theory for clustering based on energy statistics
-hold for data living in an \emph{arbitrary space of negative type}.
+is valid for data living in an \emph{arbitrary} space of negative type.
 This clustering method is
@@
 about 
-the distribution of the data,
+the distribution of the data, nor the metric space where data belongs
+to,
 contrary to $k$-means and gaussian mixture models (GMM), for example.
-Moreover, this approach \emph{does not require} the concept of the 
-\emph{cluster mean}
+Moreover, this approach \emph{does not} require the concept of the 
+\emph{cluster mean},
 which can be ill-defined for some types of data, such as images for
-instance, and thus it should also be robust to outliers.
-This method is quite general and makes very few
-assumptions about the data. If one uses the traditional energy distance
-\eqref{eq:energy}, then $\rho(x,y) = \| x - y\|$ and this fixes the kernel
-choice through \eqref{eq:kernel_semimetric}.
+instance. Since no mean is involved, the method should also be 
+robust to outliers.
+%This method is quite general and makes very few
+%assumptions about the data. 
+If one uses the traditional energy distance
+defined in \eqref{eq:energy}, 
+then $\rho(x,y) = \| x - y\|$, and this fixes the kernel
+through \eqref{eq:kernel_semimetric}. In the same way, 
+for data living in more general
+metric spaces $(\mathcal{X}, \rho)$, the corresponding semimetric $\rho$ fixes
+the kernel.
 In practice, however,
@@
 Nevertheless, if prior knowledge is available for choosing $\rho$, 
-or equivalently $K$,
 this can easily be taken into account.
@@
 One may wonder how energy statistics clustering 
-relates to the well-known kernel $k$-means problem.
+relates to the well-known kernel $k$-means problem\footnote{
+When we refer to kernel $k$-means problem we mean specifically 
+the optimization
+problem \eqref{eq:kernel_kmeans}, which should not be confused with
+kernel $k$-means algorithm, which is just one possible recipe to solve
+\eqref{eq:kernel_kmeans}.}.
 We now address this question.
@@
 where $\mu_j = \tfrac{1}{n_j} \sum_{x \in \C_j} x$ is the  mean of cluster
-$\C_j$ in the ambient space.
-It is known \cite{Dhillon} that problem \eqref{eq:kernel_kmeans} 
+$\C_j$ in the ambient space. Notice that the above objective function
+is strongly tied to the idea of minimizing distances between points
+and cluster centers, which arises from $k$-means objective function.
+It is known \cite{Dhillon2,Dhillon} that problem \eqref{eq:kernel_kmeans} 
 is equivalent to a QCQP in the same form as
 \eqref{eq:qcqp2}. The next result makes this explicit, showing that
-\eqref{eq:minimize} and \eqref{eq:kernel_kmeans} are actually equivalent,
-once the kernel is fixed.
+\eqref{eq:minimize} and \eqref{eq:kernel_kmeans} are actually equivalent.
 
@@
 \label{th:kernel_kmeans}
-The clustering problem
+The clustering optimization problem
 \eqref{eq:minimize} based on energy statistics 
-is equivalent to the kernel $k$-means problem
+is equivalent to the kernel $k$-means optimization problem
 \eqref{eq:kernel_kmeans}, and both are equivalent to \eqref{eq:qcqp2}.
@@
 \tfrac{1}{n_j}\sum_{y,z\in\C_j} \kk(y,z)$, which is the same as
-the second term. Thus the kernel $k$-means optimization problem is
-\begin{equation}
-\min_{\C_1,\dotsc,\C_k} J(\C_1,\dotsc,\C_k) = \max_{\C_1,\dotsc,\C_k}
-\sum_{j=1}^k \dfrac{1}{n_j} \sum_{x,y \in\C_j} \kk(x,y)
+the second term. Thus, the kernel $k$-means optimization problem
+\eqref{eq:kernel_kmeans} is equivalent to
+\begin{equation}
+\min_{\C_1,\dotsc,\C_k} \bigg\{ J(\C_1,\dotsc,\C_k) = \max_{\C_1,\dotsc,\C_k}
+\sum_{j=1}^k \dfrac{1}{n_j} \sum_{x,y \in\C_j} \kk(x,y) \bigg\}
 \end{equation}
@@
 
-It was shown \cite{Dhillon} that kernel $k$-means, spectral clustering,
+The above result shows that 
+kernel $k$-means problem is actually a consequence
+of energy statistics. In this vein, kernel $k$-means is part of a
+statistical framework where we are maximizing distances between probability
+distributions.
+
+As shown in \cite{Dhillon}, kernel $k$-means, spectral clustering,
 and graph partitioning problems such as ratio association, ratio cut, and
-normalized cut are all equivalent to a QCQP of the form \eqref{eq:qcqp2}.
-Actually, in general, 
-this corresponds to a weighted version of \eqref{eq:qcqp2} which reads
-$ \Tr \left( Y^\top W^{1/2} G \, W^{1/2} Y \right)$, where 
-$W = 
-\diag(w(x_1),\dotsc, w(x_n))$ and $w(\cdot)$ is a
-weight attributed to each data point. Our previous 
-results show that energy statistics
-based clustering is also equivalent to these problems. The advantage
-of energy statistics compared to kernel $k$-means 
-is that the semimetric $\rho$ is supposed to be fixed. Moreover, from a
-theoretical perspective, it brings the clustering problem into a formal
-statistical theory based on distances between probability distributions
-and embeddings in RKHS. There was no a priori reason to expect that clustering
-based on energy statistics would be equivalent to the kernel $k$-means
-problem~\eqref{eq:kernel_kmeans}.
+normalized cut are all equivalent to a QCQP of the form \eqref{eq:qcqp2}, thus
+one can use kernel $k$-means algorithm to solve these problems as well.
+This correspondence involves a weighted version of \eqref{eq:qcqp2}, that
+for the sake of completeness demonstrate next.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Clustering Based on Weighted Energy Statistics}
+
+Now we generalize the formulas from energy statistics to incorporate
+weights associated to each data point.
+Let $w(x)$ be a weight function for the point $x \in \mathbb{X}$.
+We can generalize \eqref{eq:g_def} as follows:
+\begin{equation}
+\label{eq:g_def2}
+g(\C_i, \C_j) \equiv \dfrac{1}{s_i s_j} \sum_{x\in \C_i}\sum_{y\in\C_j}
+w(x)w(y) \rho(x,y), \qquad s_i \equiv \sum_{x\in\C_i} w(x).
+\end{equation}
+Now we replace this function in the formulas \eqref{eq:within} and
+\eqref{eq:between}, with 
+$n_i \to s_i$ and $n \to s$ where $s = \sum_{j=1}^k s_j$,
+to obtain a weighted version of energy test statistic. With these
+changes, Proposition~\ref{th:minimize} remains the unnaltered, so the
+clustering problem becomes
+\begin{equation}
+\label{eq:minimize2}
+\min_{\C_1, \dotsc, \C_k} 
+\bigg\{
+W(\C_1,\dotsc,\C_k) \equiv \sum_{j=1}^k \dfrac{s_j}{2} g(\C_j,\C_j)
+\bigg\}
+\end{equation}
+where now $g$ is given by \eqref{eq:g_def2}. 
+Let us define the following matrices and vector:
+\begin{equation}
+\label{eq:weighted_matrices}
+Y_{ij} \equiv \begin{cases}
+\tfrac{1}{\sqrt{s_j}} & \mbox{if $x_i \in \C_j$} \\
+0 & \mbox{otherwise}
+\end{cases}, \qquad
+\mathcal{W} \equiv \diag(w_1,\dotsc,w_n), \qquad
+H \equiv \mathcal{W}^{1/2} Y, \qquad
+\bm{\omega} \equiv \mathcal{W} \bm{e},
+\end{equation}
+where $w_i = w(x_i)$ and $\bm{e} \in \mathbb{R}^n$ is the all-ones
+vector. Now we can show the analogous of
+Proposition~\ref{th:qcqp2} to the case of \eqref{eq:minimize2}.
+
+\begin{proposition}
+\label{th:qcqp3}
+The weighted version of energy statistics clustering given by
+problem \eqref{eq:minimize2} is equivalent to
+\begin{equation}
+\label{eq:qcqp3}
+\max_H \Tr \left\{ H^\top (\mathcal{W}^{1/2} G \mathcal{W}^{1/2}) H  \right\}
+\qquad \mbox{s.t. $H \ge 0$, $H^\top H = I$, $H H^\top \bm{\omega} =
+\bm{\omega}$,}
+\end{equation}
+where $G$ is the Gram matrix \eqref{eq:kernel_matrix} and the other quantities
+are defined in 
+\eqref{eq:weighted_matrices}.
+\end{proposition}
+\begin{proof}
+Replacing \eqref{eq:gen_kernel} and elimating the global terms which 
+do not contribute, the optimization problem \eqref{eq:minimize2}
+becomes 
+\begin{equation}
+\max_{\C_1,\dotsc,\C_k} \sum_{j=1}^k \dfrac{1}{s_j}
+\sum_{x\in\C_j}\sum_{y\in\C_j} w(x)w(y) \kk(x,y) . 
+\end{equation}
+This 
+objective function can be written as
+\begin{equation}
+\begin{split}
+\sum_{j=1}^k \dfrac{1}{s_j} 
+\sum_{p=1}^n \sum_{q=1}^n 
+w_p w_q Z_{pj} Z_{qj} G_{pq} &= 
+\sum_{j=1}^k 
+\sum_{p=1}^n \sum_{q=1}^n 
+\dfrac{Z^\top_{jp}\sqrt{w_p}}{\sqrt{s_j}} w_p^{1/2} G_{pq} w_q^{1/2} 
+\dfrac{\sqrt{w_q} Z_{qj}}{\sqrt{s_j}} \\
+&= 
+\sum_{j=1}^k \left(H^\top \mathcal{W}^{1/2} G \mathcal{W}^{1/2} H\right)_{jj}
+\\
+&= \Tr\left( H^\top \mathcal{W}^{1/2} G \mathcal{W}^{1/2} H  \right).
+\end{split}
+\end{equation}
+To obtain the constraints, note that $H_{ij} \ge 0$ by definition, and
+\begin{equation}
+(H^\top H)_{ij} = \sum_{\ell=1}^n 
+Y_{\ell i} \mathcal{W}_{\ell \ell} Y_{\ell j } = 
+\dfrac{1}{\sqrt{s_i}\sqrt{s_j}} \sum_{\ell=1}^n w_\ell Z_{\ell i} Z_{\ell j}
+= \dfrac{\delta_{ij}}{s_i} \sum_{\ell=1}^n w_\ell Z_{\ell i} = \delta_{ij},
+\end{equation}
+therefore $H^\top H = I$. This is a constraint on the rows of $H$.
+Now we obtain a condition on the columns
+of $H$. Observe that
+\begin{equation}
+\left(H^\top H\right)_{pq} = \sqrt{w_p w_q}\sum_{j=1}^k \dfrac{Z_{pj}
+Z_{qj}}{s_j} = \begin{cases}
+\dfrac{\sqrt{w_p w_q}}{s_i} & \mbox{if both $x_p,x_q \in \C_i$} \\
+0 & \mbox{otherwise}.
+\end{cases}
+\end{equation}
+Therefore, $(H^\top H \mathcal{W}^{1/2})_{pq} = \sqrt{w_p} \, w_q s_i^{-1}$
+if both points $x_p$ and $x_q$ belong to the same cluster, which
+we denote by $\C_i$ for some $i=1,\dotsc,k$, and 
+$(H^\top H \mathcal{W}^{1/2})_{pq} = 0 $ otherwise. Thus, the $p$th
+line of this matrix is nonzero only on entries corresponding to points
+that are in the same cluster as $x_p$. If we sum over the columns of this
+line we obtain $\sqrt{w_p} s_i^{-1} \sum_{q=1}^n w_q Z_{qi} = \sqrt{w_p}$,
+or equivalently $H H^\top \mathcal{W}^{1/2} \bm{e} = \mathcal{W}^{1/2} \bm{e}$,
+which gives the constraint $H H^\top \bm{\omega} = \bm{\omega}$.
+\end{proof}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Connection with Graph Partioning}
+
+The relation between kernel $k$-means and graph partitioning problems
+was already established \cite{Dhillon}. We repeat a similar analysis
+due to the relation of these problems to
+energy statistics and RKHS, which may provide a different perspective.
+
+Consider a graph $\mathcal{G} = (\mathcal{V}, \mathcal{E}, \mathcal{A})$,
+where $\mathcal{V}$ is the set of vertices, $\mathcal{E}$ the set of edges,
+and $\mathcal{A}$ is an affinity matrix of the graph 
+that measure the 
+similarities between pairs of nodes. Thus, $\mathcal{A}_{ij} \ne 0$
+if $(i,j) \in \mathcal{E}$, and $\mathcal{A}_{ij} = 0$ otherwise.
+We also associate weights to every vertex, 
+$w_i = w(i)$ for $i \in \mathcal{V}$, and let $s_j = \sum_{ i \in \C_j} w_i$,
+where $\C_j \subseteq \mathcal{V}$ is one partition of $\mathcal{V}$.
+Let
+\begin{equation}
+\textnormal{links}(\C_\ell, \C_m) \equiv 
+\sum_{i\in \C_\ell, j\in \C_m } A_{ij} .
+\end{equation}
+We want to partition the set of vertices $\mathcal{V}$ into $k$ disjoint
+subsets, $\mathcal{V} = \bigcup_{j=1}^k \C_j $. 
+The generalized ratio association problem is given by
+\begin{equation}
+\label{eq:assoc}
+\max_{\C_i,\dots,\C_k} \sum_{j=1}^k \dfrac{\textnormal{links}(\C_j,\C_j)}{s_j},
+\end{equation}
+and maximizes the within cluster association.
+The generalized ratio cut problem
+\begin{equation}
+\label{eq:cut}
+\min_{\C_i,\dots,\C_k} \sum_{j=1}^k
+\dfrac{\textnormal{links}(\C_j,\mathcal{V} \backslash \C_j)}{s_j},
+\end{equation}
+minimizes the cut between clusters. These two problems are equivalent,
+in analogous way as minimizing \eqref{eq:within} is equivalent to
+maximizing \eqref{eq:between}, as shown in Proposition~\ref{th:minimize}.
+Here this is due to the equality
+$\textnormal{links}(\C_j,\mathcal{V} \backslash \C_j)=
+\textnormal{links}(\C_j,\mathcal{V}) - \textnormal{links}(\C_j,\C_j)$.
+Several graph partitioning methods 
+\cite{Kernighan,Malik,Chan,Yu}
+can be seen as a particular
+case of \eqref{eq:assoc} or \eqref{eq:cut}.
+
+Consider \eqref{eq:assoc}, whose objective function can be written as
+\begin{equation}
+\sum_{j=1}^k \dfrac{1}{s_j} \sum_{p \in \C_j} \sum_{q \in \C_j}
+\mathcal{A}_{pq} = \sum_{j=1}^k \sum_{p=1}^n \sum_{q=1}^n 
+\dfrac{Z^\top_{jp}}{\sqrt{s_j}} \mathcal{A}_{pq} \dfrac{Z_{qj}}{\sqrt{s_j}}
+= \Tr\left( Y^\top \mathcal{A} Y \right)
+\end{equation}
+with $Z$ defined in \eqref{eq:label_matrix} and $Y$ in
+\eqref{eq:weighted_matrices}. To make the analogy with \eqref{eq:qcqp3}
+explicit,  problem \eqref{eq:assoc} is equivalent to
+\begin{equation}
+\max_H \Tr\left( H^\top \mathcal{W}^{-1/2} \mathcal{A} \mathcal{W}^{-1/2} H 
+\right) \qquad \mbox{s.t. $H\ge 0$, $H^\top H = I$, $H H^\top
+\bm{\omega}=\bm{\omega}$}.
+\end{equation}
+Therefore, this is exactly the same problem as weighted energy statistics
+clustering \eqref{eq:qcqp3}, with 
+$G = \mathcal{W}^{-1} \mathcal{A} \mathcal{W}^{-1}$. Assuming this
+matrix is positive semidefinite, this generates a semimetric
+\eqref{eq:gen_kernel} for graphs given by
+\begin{equation}
+\label{eq:metric_graphs}
+\rho(i,j) = 
+\dfrac{\mathcal{A}_{ii}}{w_i^{2}}
++\dfrac{\mathcal{A}_{jj}}{w_j^{2}}
+-\dfrac{2 \mathcal{A}_{ij}}{w_i w_j} \qquad\mbox{or}\qquad
+\rho(i,j) = -\dfrac{2 \mathcal{A}_{ij}}{w_i w_j}
+\end{equation}
+for vertices $i,j \in \mathcal{V}$, and 
+where in the second equation we assume the graph has no self-loops,
+$\mathcal{A}_{ii} = 0$. Using \eqref{eq:metric_graphs} 
+in \eqref{eq:g_def}--\eqref{eq:between}
+allows one to use energy statistics theory for inference
+over populations of graphs.
 
@@
 let us first consider the simplest possible case which
-is one-dimensional data and a two-class problem. This will also 
-be useful later
-for comparison with the more general iterative algorithm.
+is one-dimensional data and a two-class problem. This will be usefull to test
+energy statistics clustering on a simple setting.
 
@@
 \end{equation}
-The sum over $y$ can be eliminated since the term in
-parenthesis is simply counting the number of elements in $\C$ that satisfy
-the conditions of the indicator functions. Assuming
-that we first order the data in the partition, obtaining
+The sum over $y$ can be eliminated since each term in
+the parenthesis is simply counting the number of elements in $\C$ that satisfy
+the condition of the indicator function. Assuming
+that we first order the data inside the partition, obtaining
 $\bar{\C} = [ x_j \in \C: x_1 \le x_2 \le \dotsm \le x_{n}]$, we
@@
 \end{equation}
-Note that the cost of computing this is $\OO(n)$, and the cost of
+Note that the cost of computing 
+\eqref{eq:g1d}
+is $\OO(n)$, and the cost of
 sorting the data
@@
 
-Before proposing a more general iterative algorithm to \eqref{eq:qcqp2},
-let us consider two simple experiments with equal number of points
-in each cluster. 
-We keep increasing the number of points in the clusters for each experiment, 
-and cluster the data using Algorithm~\ref{algo1d}. 
-We also cluster the same data set
-with GMM, through EM algorithm, and with $k$-means. In both of these
-cases we use the initialization from $k$-means++ \cite{Vassilvitskii} 
-and we run the algorithms
-few times with different initializations and choose the answer
-with best objective function value. We use \eqref{eq:accuracy} to measure
-the clustering quality. 
-In  Fig.~\ref{fig:1d}a 
-we have data from normal distributions,
-where we can see that all the three methods
-perform closely, with a slight advantage of GMM, as expected, since
-it is the right model for the data. However, as shown in Fig~\ref{fig:1d}b,
-for lognormal distributions, Algorithm~\ref{algo1d} provides a huge improvement
-compared to both GMM and $k$-means which basically cluster at chance.
-The zero accuracy values for GMM happened when EM algorithm was unable
-to estimate the parameters. These two simple experiments illustrate
-how energy statistics based clustering is nonparametric, since it is able
-to provide high quality clustering in settings where data comes
-from very different distributions.
-
 \begin{figure}
 \begin{minipage}{0.49\textwidth}
-\includegraphics[width=\textwidth]{normal.pdf}\\[-1em]
+\hspace{1cm}\includegraphics[width=.6\textwidth]{hist_normal.pdf}\\[-.5em]
+\includegraphics[width=\textwidth]{gauss1d.pdf}\\[-1em]
 (a)
@@
 \begin{minipage}{0.49\textwidth}
-\includegraphics[width=\textwidth]{lognormal.pdf}\\[-1em]
+\hspace{1cm}\includegraphics[width=.6\textwidth]{hist_lognormal.pdf}\\[-.5em]
+\includegraphics[width=\textwidth]{loggauss1d.pdf}\\[-1em]
 (b)
@@
 \label{fig:1d}
-We cluster data using Algorithm~\ref{algo1d} ($1D$-Energy in the plots), 
-GMM, and $k$-means. We use
-\eqref{eq:accuracy} to evaluate cluster quality. Both clusters
-have the same number of points, which are increased in each experiment.
+Energy statistics clustering by Algorithm~\ref{algo1d},
+compared to $k$-means and GMM.
+Both clusters
+have the same number of points ($x$-axis), 
+which are increased in each experiment.
+We sample $100$ times from the distributions in the histograms and
+the $y$-axis is the average value of the clustering 
+accuracy \eqref{eq:accuracy} (errorbars are standard error).
 (a) 
-$x\sim \tfrac{1}{2}\left( \mathcal{N}(\mu_1,\sigma_1) +
-\mathcal{N}(\mu_2,\sigma_2)  \right)$ with 
+$x\sim \tfrac{1}{2}\left[ \mathcal{N}(\mu_1,\sigma_1) +
+\mathcal{N}(\mu_2,\sigma_2)  \right]$,  
 $\mu_1 = 0$,
@@
 (b) 
-$x\sim \tfrac{1}{2}\left( e^{\mathcal{N}(\mu_1,\sigma_1)} +
-e^{\mathcal{N}(\mu_2,\sigma_2)}  \right)$ with 
+$x\sim \tfrac{1}{2}\left[ e^{\mathcal{N}(\mu_1,\sigma_1)} +
+e^{\mathcal{N}(\mu_2,\sigma_2)}  \right]$,  
 $\mu_1 = 0$,
@@
 \end{figure}
+
+Let us consider two simple experiments with equal number of points
+in each cluster. 
+We keep increasing the number of points in the clusters for each experiment, 
+and cluster the data using Algorithm~\ref{algo1d}. 
+We also cluster the same data set
+with GMM, through EM algorithm, and with standard $k$-means. In both of these
+cases we use the initialization from $k$-means++ \cite{Vassilvitskii} 
+and we run the algorithms
+few times with different initializations and choose the answer
+with best objective function value. We use \eqref{eq:accuracy} to measure
+the clustering quality. Furthermore, we pick $100$ different samples
+for each configuration, and show the average accuracy with errorbars indicating
+the standard error.
+In  Fig.~\ref{fig:1d}a 
+we have data from normal distributions,
+where we can see that all the three methods
+perform closely, with a slight advantage of GMM, as expected, since
+it is the true model for the data. Energy statistics performs slightly better
+than $k$-means in this experiment. Now in Fig~\ref{fig:1d}b
+we consider one example of 
+data from lognormal distributions. We can see that energy
+statistics clustering through Algorithm~\ref{algo1d} provides a huge 
+improvement
+compared to both GMM and $k$-means, which basically cluster at chance
+in this case.
+GMM is worst than chance since sometimes it 
+is not able to estimate the parameters, thus giving zero
+accuracy.
+These two simple experiments illustrate
+how energy statistics based clustering is nonparametric, being able
+to provide high quality clustering in settings where data comes
+from very different distributions.
 
@@
 convergence, i.e. until no new assignments are made.
-The complete algorithm is shown in Algorithm~\ref{kmeans_algo}.
+The whole procedure shown in Algorithm~\ref{kmeans_algo}.
 It can be shown that this algorithm converges provided $G$ is positive
@@
 Although our notation looks a little different than the standard
-kernel $k$-means found in the literature \cite{Dhillon}, this is precisely
+kernel $k$-means found in the literature \cite{Dhillon2,Dhillon}, 
+this is precisely
 the same algorithm but written in a more concise and explicit way.
@@
 $\OO(n_\ell)$ operations, and although the second term requires
-$\OO(n_\ell^2)$, it only needs to be computed once outside the data
-points loop in Algorithm~\ref{kmeans_algo} (step 1).
+$\OO(n_\ell^2)$ it only needs to be computed once outside loop through
+data points (step 1).
 Therefore, the time complexity Algorithm~\ref{kmeans_algo} is
@@
 a total energy cost function \eqref{eq:maxQ} denoted by $Q^{(j)}$.
-Let us consider the change in the total energy cost by moving
+Let us consider the change in the total energy by moving
 $x_i$ to cluster $\C_\ell$. 
@@
 Note that \eqref{eq:changeQ} assures that the objective function is
-monotonically increasing at each iteration.
+monotonically increasing at each iteration, therefore the algorithm
+converges in a finite number of steps.
 
@@
 For the 
-entire dataset this gives a time-complexity
+entire dataset this gives a time complexity
 of $\OO(n k  \max_j n_j) =\OO(k n^2)$. This is the same cost as

### 2017-07-18 qcqp_clustering/code/eclust/energy1d.py

@@ -1,4 +1,9 @@
-"""Energy Statistics Clustering in 1D"""
+"""Energy Statistics Clustering in 1D.
+
+Use exact formula with no initialization to perform clustering in 1D
+from energy statistics.
+
+"""
 
 # Guilherme Franca <guifranca@gmail.com>
 # Johns Hopkinks University, Neurodata
@@ -11,7 +16,8 @@ import numpy as np
 
 def mean1D(X, Y):
     """Assume X and Y are both sorted lists.
-    Compute the energy mean in O(n) time.
+    Compute the ``energy mean'' in O(n) time.
+    This is the function g(X, Y) from the paper.
     
     """
     sx = 0
@@ -36,7 +42,8 @@ def mean1D(X, Y):
 
 def within_sample(A):
     """Compute within-sample dispersion statistics between k sets.
-    Input: A[A_1, A_2, ..., A_k] where each A_i is a sample
+
+    Input: A = [A_1, A_2, ..., A_k] where each A_i is a sample
     Output: between sample energy statistics (S)
 
     """
@@ -44,12 +51,20 @@ def within_sample(A):
     return sum([len(a)/2*mean2(a)  for a in A])
 
 def mean2(A):
-    """Another version of the above. Assume A is sorted."""
+    """Another version of the above, i.e. compute function g(A_i,A_i). 
+    Assume A is sorted."""
     n = len(A)
     return (2./n**2)*sum([A[i]*(2.*i +1 - n) for i in range(n)])
 
 def two_clusters1D(X):
-    """Optimize energy statistics for two clusters in 1D."""
+    """Optimize within energy statistics for two clusters in 1D.
+
+    This algorithm performs 2-class-clustering in 1D. 
+    
+    Input: X = data matrix
+    Output: labels, objective function value
+    
+    """
     sorted_indices = np.argsort(X)
     Y = X[sorted_indices]
     best_cost = np.inf
@@ -69,15 +84,19 @@ def two_clusters1D(X):
     return final_labels, best_cost
 
 
+
 ###############################################################################
 if __name__ == "__main__":
    
-    import data
-    from metric import accuracy
     from sklearn.cluster import KMeans
     from sklearn.mixture import GMM
[…]

### 2017-08-11 qcqp_clustering/draft/energy_clustering.tex
finished draft

@@
 \usepackage[none]{hyphenat}
+\usepackage{multirow}
+\usepackage{colortbl}
+\usepackage{array,booktabs}
+\usepackage{xcolor}
+\usepackage{makecell}
 
@@
 \newcommand{\Ind}[1]{\mathbbm{1}_{#1}}
+\newcommand\e{e}
+\newcommand\om{\omega}
 
@@
 \begin{abstract}
-Energy statistics provides a nonparametric test for equality of distributions.
-It was proposed by 
+Energy statistics 
+was proposed by 
 Sz\' ekely in the 80's
-inspired by the Newtonian gravitational potential from classical mechanics. 
-Energy statistics
-was further generalized to probability 
-distributions on arbitrary metric spaces,
-and more recently a connection with reproducing kernel Hilbert spaces 
-was established.
-Nevertheless, although extensively used by the statistics community, it
+inspired by the Newtonian gravitational potential from classical mechanics,
+and it provides a nonparametric test for equality of distributions.
+It was generalized to 
+metric spaces of strong negative type,
+and more recently a connection with reproducing kernel Hilbert spaces (RKHS)
+was proposed.
+Although  
+extensively used in statistics, it
 has not been previously incorporated in machine learning problems.
-In this paper, we consider the problem of clustering data from
+Here we consider the problem of clustering data from
 an energy statistics theory perspective.
 We provide a precise mathematical formulation yielding
-a quadratically constrained quadratic program (QCQP).
-We show the equivalence between the energy statistics clustering
-formulation 
-and the kernel $k$-means optimization problem (not to be confused with
-kernel $k$-means algorithm). 
-Therefore, our results imply a first principles derivation of 
-kernel $k$-means problem
-from energy statistics theory, thus bringing this important 
-clustering method into a formal statistical framework.
-Moreover, energy statistics is nonparametric, it usually fixes
-the kernel choice,
-and if prior information is available
-it can be easily incorporated in the kernel construction.
-We propose an iterative algorithm
-to find local optimizers of the aforementioned QCQP based on the 
-energy change
-of moving points to different clusters. This algorithm 
-is different but has the same computational cost as kernel $k$-means 
-algorithm.
-We compare it to
-kernel $k$-means, standard $k$-means, and GMM/EM algorithms by 
-providing carefully designed numerical experiments. The results
-show that, in general, this algorithm outperforms
-these most used clustering algorithms.
+a quadratically constrained quadratic program (QCQP), which we show
+to be equivalent
+to kernel $k$-means optimization problem.
+Thus, our results imply a first principles derivation of 
+kernel $k$-means from energy statistics.
+Moreover, we also consider a weighted version of energy statistics 
+applied to clustering, which is related to 
+graph partitioning problems.
+To find local optimizers of such QCQP we consider an iterative
+algorithm based on Hartigan's method, which in this case
+has the same computational cost 
+as kernel $k$-means algorithm but usually with better clustering
+performance. We provide carefully designed numerical experiments
+showing the superiority of this method compared to
+kernel $k$-means, standard $k$-means, and gaussian mixture models.
 \end{abstract}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Introduction}
 
-Energy statistics is based on the energy distance between probability
-distributions, which provides a notion of statistical potential energy
-to statistical observations, in close analogy to Newton's gravitational
+Energy statistics \cite{Szkely2013}
+is based on a 
+notion of statistical potential energy between probability distributions,
+in close analogy to Newton's gravitational
 potential in classical mechanics. 
-It provides a nonparametric test for equality of distribution,
-such that minimum energy is achieved under equality of 
-distributions. 
-We refer the reader 
-to \cite{Szkely2013}, and references therein, for an overview.
+It provides a nonparametric test for equality of distribution which
+is achieved under minimum energy. When probability distributions
+are different, this statistical potential 
+energy diverges as sample size increases.
 Energy statistics has been applied to several goodness-of-fit 
@@
 multi-sample tests of equality of distributions, analysis of variance
-\cite{RizzoVariance}, and (nonlinear)
+\cite{RizzoVariance}, nonlinear
 dependence tests through
 distance covariance and distance correlation, which generalizes the Pearson
-correlation coefficient. 
-Moreover, energy statistics  was applied
-to hierarchical clustering \cite{RizzoClustering} by extending Ward's
-method of minimum variance.
-
-More recently, distance covariance was 
+correlation coefficient, and  
+hierarchical clustering \cite{RizzoClustering} by extending Ward's
+method of minimum variance. (We refer to \cite{Szkely2013} and 
+references therein for an overview of energy statistics.)
+Moreover, an energy statistics formulation
+to clustering was already proposed in \cite{Kgroups} 
+which greatly motivated this paper.
+
+Recently, distance covariance was 
 generalized from Euclidean
@@
 distributions in reproducing kernel Hilbert spaces (RKHS), was
-established \cite{Sejdinovic2013}. This important work provides the missing
-link
-between techniques commonly used in the statistics literature, regarding
-energy statistics, and techniques
-commonly used in machine learning.
-
-Given a dataset $\mathbb{X} = \{x_1,\dotsc,x_n \}$, where each
-data point $x_i$ lives in a space $\mathcal{X}$, the clustering
-problem consists in grouping these points into $k$ groups $\{
-\C_1,\C_2,\dotsc,\C_k \}$, such that
-points belonging to the same group are more ``similar'' to each other
-than to points in  other groups. This intuitive notion already assumes
-a ``metric'' able to measure the similarity between points. Clustering
-is an unsupervised method, and one of the most important problems in machine
-learning since it provides the first step towards automatically constructing
-labels for previously unseen data. In practice
-$k$-means is arguably the most used algorithm, and gaussian 
-mixture models (GMM) through the expectation
-maximization (EM) algorithm is also often used. Both are parametric, making 
-strong assumptions about
-the distribution of the data (isotropy and normality), and it assumes
-that data points
-lie in Euclidean space, $\mathcal{X} = \mathbb{R}^D$, hence
-the similarity is based on Euclidean metric, 
-$\| x_i - x_j \|^2 = (x_i - x_j)^\top(x_i-x_j) = 
-\sum_{\ell=1}^D (x_{i,\ell} - x_{j,\ell})^2$. These algorithms provide
-a good quality clustering when data is linearly separable in Euclidean space,
-i.e. there is a clear resolution in the cluster centers
-and points from different clusters do not overlap considerably.
-
-To account for nonlinearly separable data, which may live
-in an arbitrary non-Euclidean space, kernel methods are usually employed.
-If a so-called Mercer kernel $\kk: \mathcal{X}\times \mathcal{X} \to \mathbb{R}$
-is available, which is a symmetric and positive definite 
-function, it guarantees that there exists a map $\varphi : \mathcal{X} \to
-\mathcal{H}$ where $\mathcal{H}$ is a Hilbert space \cite{Mercer}. 
-One can thus
-compute similarities between data point through the inner product of
-the feature space $\mathcal{H}$ and 
-perform clustering in $\mathcal{H}$. The convenience of a Mercer
-kernel is that the map $\varphi$ is only used implicitly, i.e. one
-does not need to know $\varphi$, since the inner
-of product of $\mathcal{H}$ can be computed only 
-based on the kernel function $\kk(\cdot, \cdot)$ which acts on the original
-space $\mathcal{X}$. 
-This technique is known as the kernel trick.  
-Kernel $k$-means problem 
-is exactly $k$-means in the feature space
-\cite{Smola,Girolami} (see also \cite{Filippone} for a survey of clustering
-methods).
-To cluster data that is not linearly
-separable in $\mathcal{X}$,
-kernel based methods exploit the nonlinear structure provided by
-the kernel function, which implicitly maps data points to a usually higher
-dimensional feature 
-space, with the hope that they becomes separable by 
-forming well-defined groups.
-The choice of
-kernel is crucial, and yet there is no principled method
-for this. One must rely on ad hoc methods to find an appropriate kernel
-which is data dependent.
-As it stands, kernel $k$-means problem is an heuristic approach to clustering
-in the sense that it is not derived from a statistical theory.
-
-The main contribution of this paper is to provide a consistent formulation
-to clustering based on energy statistics theory. 
-The original goal is to provide a nonparametric
-clustering method, since energy distance is nonparametric.
-The basis of our work is the theory developed in 
-\cite{Sejdinovic2013} which establishes an equivalence between energy
-distance and RKHS. 
-Based on this and on the multi-sample test
-for equality of distributions, we show that
-there is only one possible clustering optimization
-problem consistent with the energy statistics formulation. 
-We then show that this 
-optimization problem is equivalent to a quadratically constrained quadratic
-program (QCQP), which has the same trace maximization
-form as kernel $k$-means,
-spectral clustering, and some well-known graph partitioning problems
-\cite{Dhillon2,Dhillon}. We show the equivalence between this approach
-and kernel $k$-means optimization problem, implying that
-the later can actually be derived from energy statistics, thus
-bringing this important clustering method into a unified statistical theory.
-However, it is important to note that 
-energy statistics fixes a kernel to begin with.
-From an algorithmic perspective, our contribution is to 
-provide an iterative algorithm, which is different but has the same
-complexity as kernel $k$-means algorithm. Our numerical results
-provide compelling evidence that this algorithm 
-is more accurate and robust than kernel $k$-means algorithm.
-Moreover, these experiments 
-puts in evidence the nonparametric aspect of energy
-statistics based clustering, since it is able to perform accurately
-on data having very different distributions, contrary to $k$-means and
-GMM for instance.
-
-Our work is organized as follows. In section~\ref{sec:background} we review
+established \cite{Sejdinovic2013}. This provides the 
+link between concepts used in energy statistics and concepts
+commonly used in machine learning, and form the basis of our approach.
+
+The clustering problem has a long history in machine learning.
+The most used method is $k$-means \cite{Lloyd,MacQueen,Forgy}, which
+is based on Lloyd's heuristic \cite{Lloyd} of assigning a data point to
+the cluster with closest center. The only statistical 
+information about each cluster
+comes from its mean and thus it is sensitive to outliers. Moreover, $k$-means
+is strongly parametric, assuming that data is isotropically and normally
+distributed since it is a special case of gaussian mixture models (GMM) 
+through the expectation maximization (EM) algorithm. Nevertheless,
+$k$-means works very well when data is linearly separable 
+in Euclidean space, with a clear resolution of cluster centers and without
+significant overlap of the points belonging to different clusters.
+
+To account for nonlinearities, kernel methods were introduced 
+\cite{Smola,Girolami}. A mercer kernel \cite{Mercer} is used to implicitly
+map data points to a RKHS, then clustering can be performed in the associated
+Hilbert
+space by using its inner product. This enables one to cluster way more
+complicated datasets, however, the kernel choice remains the biggest challenge
+since there is no principled theory behind this. The well-known
+kernel $k$-means optimization problem is nothing but $k$-means in the
+feature space \cite{Girolami}. Furthermore, kernel $k$-means algorithm
+\cite{Dhillon2,Dhillon} is still based on Loyd's heuristic \cite{Lloyd} of
+using the mean of each cluster in the feature space. 
+We refer the reader to \cite{Filippone} for a survey of clustering
+methods.
+
+Although clustering from energy statistics was considered
+in \cite{Kgroups}, the precise optimization problem behind this approach
+remains elusive, as well as the connection to kernel methods.
+The main theoretical contribution of this paper is to fill this gap.
+Since the statistical potential energy is minimum when
+distributions are equal, the principle 
+principle behind clustering is to maximize the statistical energy, 
+enforcing probability distributions associated to each cluster
+to be different from one another.
+We provide a precise mathematical formulation to this
+statement, showing that there is only one possible optimization problem
+consistent with energy statistics. Moreover, we show that this is
+a quadratically constrained
+quadratic program (QCQP) in the associated RKHS. Our results immediately 
+establish the connection 
+to kernel methods used in machine learning, 
+by showing that this QCQP is equivalent
+to kernel $k$-means optimization problem. 
+The equivalence between kernel $k$-means, 
+spectral clustering, and graph partitioning problems was already established
+in \cite{Dhillon,Dhillon2}. We show how these connections arise
+from a weighted version of energy statistics.
+
+Our algorithmic contribution is to use Hartigan's method
+\cite{Hartigan} to find local solutions of the above mentioned QCQP.
+This approach was also proposed in \cite{Kgroups}. In this case,
+the resulting algorithm has the same
+time complexity as kernel $k$-means algorithm, however, the numerical 
+results provide compelling evidence that this method
+is more accurate and robust than kernel $k$-means. This evidence
+is also supported by general theoretical results \cite{Telgarsky,Slonin}.
+Moreover, in our experiments we put in
+in evidence the nonparametric aspect of energy
+statistics based clustering, which provides a default kernel, showing that
+it is able to perform accurately on data coming very different 
+distributions, contrary to $k$-means and GMM for instance.
+
+Our work is organized as follows. In section~\ref{sec:background} we introduce
 the necessary background on energy statistics and RKHS.
@@
 results of this paper,
-where we consider a clustering theory based on energy statistics, leading
-to a QCQP which is NP-hard.
-In section~\ref{sec:twoclass} we consider a simple example in one dimension,
+where we consider a clustering theory based on energy statistics leading
+to a QCQP, which is NP-hard. We also show the equivalence to 
+kernel $k$-means problem.
+In Section~\ref{sec:weighted} we generalize these results to a weighted
+version of energy statistics, which provides the connection with graph
+partitioning problems and spectral clustering.
+In Section~\ref{sec:twoclass} we consider a simple example in one dimension,
 where we propose an algorithm which requires no initialization.
 In section~\ref{sec:algo} we briefly review kernel $k$-means algorithm,
-and propose a new iterative algorithm to solve this QCQP.
+and propose a new iterative algorithm based on Hartigan's method
+to solve this QCQP.
 Section~\ref{sec:numerics} contains some carefully designed numerical
@@
 
-In this section we briefly review the main concepts from energy
-statistics and its relation to reproducing kernel Hilbert spaces 
-(RKHS) which form the basis of our work.
+In this section we introduce the main concepts from energy
+statistics and its relation to 
+RKHS which form the basis of our work.
 For more details we refer the reader
@@
 distribution functions with finite first moments. 
-The quantity \cite{Szkely2013}
+The quantity 
 \begin{equation}
 \label{eq:energy}
-\Energy(P, Q) \equiv 2 \E \| X - Y\| - \E \| X - X' \| - \E \| Y - Y' \|,
-\end{equation}
-called \emph{energy distance}, 
+\Energy(P, Q) \equiv 2 \E \| X - Y\| - \E \| X - X' \| - \E \| Y - Y' \|
+\end{equation}
+called \emph{energy distance} \cite{Szkely2013} 
 is rotational invariant and nonnegative, $\Energy(P,Q) \ge 0$, where
@@
 2 \E \| X - Y\|^{\alpha} - \E \| X - X' \|^{\alpha} - 
-\E \| Y - Y' \|^{\alpha},
+\E \| Y - Y' \|^{\alpha}
 \end{equation}
@@
 $\Energy_\alpha(P,Q) = 0$ if and only if $P=Q$, while for $\alpha=2$ 
-we have $\Energy_2(P,Q) = 2\| \E(X) - \E(Y) \|^2$, showing that
+we have $\Energy_2(P,Q) = 2\| \E(X) - \E(Y) \|^2$ which shows that
 equality to zero only requires
@@
 mention that \eqref{eq:energy2} can be even further generalized.
-Let $X, Y \in \mathcal{X}$,  where $\mathcal{X}$ is an arbitrary space endowed
+Let $X, Y \in \mathcal{X}$  where $\mathcal{X}$ is an arbitrary space endowed
 with a \emph{semimetric of negative type}
@@
 \end{equation}
-where $X_i \in \mathcal{X}$, and $c_i \in \mathbb{R}$ such that
+where $X_i \in \mathcal{X}$ and $c_i \in \mathbb{R}$ such that
 $\sum_{i=1}^n c_i = 0$. Then, $\mathcal{X}$ is called a \emph{space of
@@
 \end{equation}
-For spaces of negative type, there exists a Hilbert space $\mathcal{H}$ and
+For spaces of negative type there exists a Hilbert space $\mathcal{H}$ and
 a map $\varphi: \mathcal{X} \to
 \mathcal{H}$ such that
-$\rho(X, Y) = \| \varphi(X) - \varphi(Y) \|_{\mathcal{H}}^2$, which
+$\rho(X, Y) = \| \varphi(X) - \varphi(Y) \|_{\mathcal{H}}^2$. This
 allows us to compute quantities related to probability distributions over
@@
 definite. Indeed, notice that
-$\langle \kkk_x, \kkk_y \rangle = \kkk_y(x) = \kk(x,y)$, and since
-this inner product is real,
-$\langle \kkk_x, \kkk_y \rangle^* = \langle \kkk_y, \kkk_x \rangle = 
-\langle \kkk_x, \kkk_y \rangle$, we immediately have that
-the kernel is symmetric,
-$\kk(y,x) = \kk(y,x)$. Moreover, for any $w \in
+$\langle \kkk_x, \kkk_y \rangle = \kkk_y(x) = \kk(x,y)$, and by definition
+$\langle \kkk_x, \kkk_y \rangle^* = \langle \kkk_y, \kkk_x \rangle$, but
+since the inner product is real we have 
+$\langle \kkk_y, \kkk_x \rangle = \langle \kkk_x, \kkk_y \rangle$, or
+equivalently 
+$\kk(y,x) = \kk(x,y)$. Moreover, for any $w \in
 \HH$ we can write $w = \sum_{i=1}^n c_i \kkk_{x_i}$, where
@@
 elements $G_{ij} = \kk(x_i,x_j)$, this is equivalent to $G$ being
-positive semidefinite; $v^\top G \, v \ge 0$ for any vector
+positive semidefinite, i.e. $v^\top G \, v \ge 0$ for any vector
 $v \in \mathbb{R}^n$.
@@
 $\int f(x) d P(x) = \langle f, \kkk_P \rangle$ for all $f \in \Hk$,
-or alternatively, $\kkk_P \equiv \int \kk( \, \cdot \,, x)  d P(x)$. 
+or alternatively $\kkk_P \equiv \int \kk( \, \cdot \,, x)  d P(x)$. 
 We can now  introduce the 
 notion of distance between two probability measures using the inner product
-of $\Hk$. This is called the maximum mean discrepancy (MMD) and
+of $\Hk$, which is called the maximum mean discrepancy (MMD) and
 is given by
-\begin{equation}\label{eq:mmd}
-\gamma_\kk(P,Q) \equiv \| \kkk_P - \kkk_Q \|_{\Hk},
-\end{equation}
-which can also be written as \cite{Gretton2012}
+\begin{equation}
+\label{eq:mmd}
+\gamma_\kk(P,Q) \equiv \| \kkk_P - \kkk_Q \|_{\Hk}.
+\end{equation}
+This can also be written as \cite{Gretton2012}
 \begin{equation}\label{eq:mmd2}
@@
 The following important result shows that semimetrics of negative
-type and symmetric positive semidefinite kernels are closely related
-\cite{Berg1984}. Let $\rho: \mathcal{X} \times \mathcal{X} \to \mathbb{R}$,
+type and symmetric positive definite kernels are closely related
+\cite{Berg1984}. Let $\rho: \mathcal{X} \times \mathcal{X} \to \mathbb{R}$
 and $x_0 \in \mathcal{X}$ an arbitrary but fixed point.
@@
 \eqref{eq:negative_type}.
-Here we have a family of kernels, one for each choice of $x_0$. Conversely,
+We have a family of kernels, one for each choice of $x_0$. Conversely,
 if $\rho$ is a semimetric of negative type and $\kk$ is a kernel in this
@@
 \rho(x,y) &= \kk(x,x) + \kk(y,y) -2\kk(x,y) \\
-&=  \| \kkk_x - \kkk_y \|^2_{\Hk},
+&=  \| \kkk_x - \kkk_y \|^2_{\Hk}
 \end{split}
@@
 $\varphi: x \mapsto \kkk_x$ is injective \cite{Sejdinovic2013}.
-When these conditions are satisfied, we say that the kernel $\kk$ 
+When these conditions are satisfied we say that the kernel $\kk$ 
 generates the semimetric $\rho$. 
@@
 \end{equation}
-Since $\gamma_k^2(P, Q) = \| \kkk_P - \kkk_Q \|^2_{\Hk}$ we
-can compute the energy distance using the inner product of $\Hk$. Moreover,
-this can be computed from the data according to \eqref{eq:inner_data}.
-
-Finally, let us recall the main formulas for test statistic
-of equality of distributions \cite{Szkely2013}. 
-Assume we have data $\mathbb{X} = \{ x_1,\dotsc, x_n \}$, where
-$x_i \in \mathcal{X}$, and $\mathcal{X}$ is a space of negative type.
-Consider a partition $\mathbb{X} = \bigcup_{j=1}^k \C_j$, with
+Due to \eqref{eq:mmd} we can compute the energy distance using the inner 
+product of $\Hk$. Moreover, this can be estimated from a sample
+according to \eqref{eq:inner_data}.
+
+Finally, let us recall the main formulas from energy statistics
+for the test statistic of equality of distributions \cite{Szkely2013}. 
+Assume we have data $\mathbb{X} = \{ x_1,\dotsc, x_n \}$ where
+$x_i \in \mathcal{X}$ and $\mathcal{X}$ is a space of negative type.
+Consider a disjoint partition $\mathbb{X} = \bigcup_{j=1}^k \C_j$, with
 $\C_i \cap \C_j = \emptyset$.
@@
 W \equiv
-\sum_{j=1}^{k} \dfrac{n_j}{2} g(\C_j, \C_j),
+\sum_{j=1}^{k} \dfrac{n_j}{2} g(\C_j, \C_j)
 \end{equation}
@@
 g(\C_j, \C_j)
-\right],
+\right]
 \end{equation}
@@
 One can make the analogy 
-that every data point $ x \in \C_j$ form a massive body, 
+that points $ x \in \C_j$ form a massive body 
 whose total mass is characterized by the distribution function $P_j$.
@@
 energy of the from $S(P_1,\dotsc,P_k)$ which measures how different
-these
-mass distributions are, and achieves the ground state
+the distribution of these masses are,  and achieves the ground state
 $S=0$ when all bodies have the same mass distribution. The potential energy
@@
 independent of the choice of partition. Therefore, maximizing
-$S$ over the choice of partition is equivalent to minimizing~$W$.
+$S$ over the choice of partition is equivalent to minimizing $W$.
 \end{proof}
 
-Therefore, for a given $k$, the clustering problem amounts to
+For a given $k$, the clustering problem amounts to
 finding the best partition of the data by solving~\eqref{eq:minimize}.
-Notice that this is a hard assignment clustering problem as partitions
-are disjoint.
-
-Now we show how to formulate problem \eqref{eq:minimize} in the corresponding
-RKHS.
-Based on 
-\eqref{eq:kernel_semimetric} and \eqref{eq:gen_kernel}, assume that 
-the kernel $\kk: \mathcal{X} \times \mathcal{X} \to \mathbb{R}$ 
-generates $\rho$. 
-Let us define  the Gram matrix
+Notice that this is a hard clustering problem, as partitions
+are disjoint. The problem \eqref{eq:minimize} based on
+energy statistics was already proposed in \cite{Kgroups}. However, it is
+important to note that this is equivalent to maximizing \eqref{eq:between}
+which is the test statistic for equality of distributions. In the following
+we show what is the explicit optimization problem behind \eqref{eq:minimize}.
+
+Let us formulate \eqref{eq:minimize} in the corresponding
+RKHS. Based on \eqref{eq:kernel_semimetric} and \eqref{eq:gen_kernel}, 
+assume that the kernel $\kk: \mathcal{X} \times \mathcal{X} \to \mathbb{R}$ 
+generates $\rho$.  Let us define  the Gram matrix
 \begin{equation}
@@
 the $i$th row of a matrix $M$, and $M_{\bullet j}$ denotes its $j$th column.
-Our next result reveals the optimization problem behind \eqref{eq:minimize},
-which is NP-hard since
+Our next result shows the optimization problem \eqref{eq:minimize}
+is NP-hard since
 it is a quadratically constrained quadratic program (QCQP).
@@
 \mbox{s.t. $\Zt \ge 0$, $\Zt^\top \Zt = I$, 
-$\Zt \Zt^\top \bm{e} = \bm{e}$},
-\end{equation}
-where $\bm{e} = (1,1,\dots,1)^\top \in \mathbb{R}^n$ is the all-ones vector,
+$\Zt \Zt^\top \e = \e$},
+\end{equation}
+where $\e = (1,1,\dots,1)^\top \in \mathbb{R}^n$ is the all-ones vector,
 and $G$ is the Gram matrix \eqref{eq:kernel_matrix}.
@@
 \end{equation}
-Note that the first term does not contribute to the optimization problem,
-since it is a global term that does not depend which partition is chosen. 
+Note that the first term is global so it does not contribute to the 
+optimization problem.
 Therefore, minimizing \eqref{eq:W2} is equivalent to
@@
 constraint in \eqref{eq:qcqp}. In matrix form this reads
-$Z^T \bm{e} = D \bm{e}$. Replacing $Z=YD^{1/2}$ we have
-$Y^\top \bm{e} = D^{1/2} \bm{e}$. Multiplying this last equation
+$Z^T \e = D \e$. Replacing $Z=YD^{1/2}$ we have
+$Y^\top \e = D^{1/2} \e$. Multiplying this last equation
 on the left by $Y$, and noticing
-that $Y D^{1/2} \bm{e} = Z \bm{e} = \bm{e}$, we finally obtain
-$Y Y^\top \bm{e} = \bm{e}$. Therefore, the optimization 
+that $Y D^{1/2} \e = Z \e = \e$, we finally obtain
+$Y Y^\top \e = \e$. Therefore, the optimization 
 problem \eqref{eq:qcqp} is equivalent
@@
 
-Based on Proposition~\ref{th:qcqp2}, to group data $\{ x_1,\dotsc,x_n \}$
+Based on Proposition~\ref{th:qcqp2}, to group data
+$\mathbb{X} = \{ x_1,\dotsc,x_n \}$
 into  $k$ clusters, we first compute the Gram matrix
@@
 \sum_{i=1}^k \lambda_i$. One might then normalize and threshold the rows
-of $Y^\star$, or even better, following \cite{NgJordan} we can normalize the
+of $Y^\star$, or better, following \cite{NgJordan} one can normalize the
 rows of $Y^\star$ and apply standard $k$-means on this matrix where each
 row is considered as a data point.
-This is the same 
-procedure done in spectral clustering on the (normalized) Laplacian
-of the graph defined by a similarity matrix.
+This is the procedure done in spectral clustering on a graph Laplacian
+matrix defined by a similarity matrix.
 However, computing eigenvectors of a very large matrix
-can be problematic, and usually iterative methods are preferred.
-We will propose an iterative method to find approximate solutions
-to \eqref{eq:qcqp2} in section~\ref{sec:algo}.
-
-It is important to note 
-that the previous theory for clustering based on energy statistics
-is valid for data living in an \emph{arbitrary} space of negative type.
-This clustering method is
+can be very expensive, and usually iterative methods are preferred. Moreover,
+this procedure is only guaranteed to converge if the normalized $Y^\star$ is
+positive semidefinite.
+In Section~\ref{sec:algo} we will propose an iterative method to directly 
+find approximate solutions to \eqref{eq:qcqp2} that is guaranteed to
+converge and does not require that the similarities $\kk(x_i,x_j)$ be
+nonnegative.
+
+The clustering problem \eqref{eq:qcqp2} based on energy statistics
+is valid for data living in an \emph{arbitrary} space of negative type, where
+a semimetric $\rho$, and thus the kernel \eqref{eq:kernel_semimetric}, are
+assumed to be known. This clustering method is
 \emph{nonparametric} since it does not make any assumptions
-about 
-the distribution of the data, nor the metric space where data belongs
-to,
+about  the distribution of the data, 
 contrary to $k$-means and gaussian mixture models (GMM), for example.
-Moreover, this approach \emph{does not} require the concept of the 
-\emph{cluster mean},
-which can be ill-defined for some types of data, such as images for
-instance. Since no mean is involved, the method should also be 
-robust to outliers.
+%Moreover, this approach \emph{does not} require the concept of the 
+%\emph{cluster mean},
+%which can be ill-defined for some types of data, such as images for
+%instance. Since no mean is involved, the method should also be 
+%robust to outliers.
 %This method is quite general and makes very few
 %assumptions about the data. 
-If one uses the traditional energy distance
+If one uses the standard energy distance
 defined in \eqref{eq:energy}, 
-then $\rho(x,y) = \| x - y\|$, and this fixes the kernel
+then $\rho(x,y) = \| x - y\|$ and this fixes the kernel
 through \eqref{eq:kernel_semimetric}. In the same way, 
-for data living in more general
-metric spaces $(\mathcal{X}, \rho)$, the corresponding semimetric $\rho$ fixes
+for data living in a more general
+metric space $(\mathcal{X}, \rho)$ the corresponding semimetric $\rho$ fixes
 the kernel.
@@
 the clustering quality strongly depend on the choice of a suitable
-$\rho$ which is what measures the similarity between different data points,
+$\rho$ which measures the similarity between data points,
 and is equivalent to choosing an appropriate kernel.
-Nevertheless, if prior knowledge is available for choosing $\rho$, 
-this can easily be taken into account.
-
+The energy distance \eqref{eq:energy2} fixes a family of choices
+$\rho(x,y) = \| x-y\|^\alpha$ with $0 < \alpha\le 2$. If prior information
+is available one may choose a more appropriate $\rho$.
+
+\subsection*{Relation to Kernel $\bm{k}$-Means}
 One may wonder how energy statistics clustering 
-relates to the well-known kernel $k$-means problem\footnote{
-When we refer to kernel $k$-means problem we mean specifically 
-the optimization
-problem \eqref{eq:kernel_kmeans}, which should not be confused with
-kernel $k$-means algorithm, which is just one possible recipe to solve
-\eqref{eq:kernel_kmeans}.}.
-We now address this question.
-For a positive semidefinite $G$, there exists a map
+relates to the well-known kernel $k$-means problem%
+\footnote{When we refer to kernel $k$-means problem we mean specifically 
+the optimization problem \eqref{eq:kernel_kmeans}, which should not be 
+confused with kernel $k$-means algorithm that is just one possible recipe 
+to solve \eqref{eq:kernel_kmeans}.
+} 
+which is extensively used in machine learning.
+%and was proposed as a kernelized version of $k$-means.
+For a positive semidefinite Gram matrix $G$, as defined in
+\eqref{eq:kernel_matrix},
+there exists a map
 $\varphi: \mathcal{X} \to \HH_\kk$ such that
@@
 problem,
-in feature space,
+in the feature space,
 is defined by
@@
 and cluster centers, which arises from $k$-means objective function.
-It is known \cite{Dhillon2,Dhillon} that problem \eqref{eq:kernel_kmeans} 
+It was shown \cite{Dhillon2,Dhillon} 
+that problem \eqref{eq:kernel_kmeans} 
 is equivalent to a QCQP in the same form as
@@
 \label{th:kernel_kmeans}
-The clustering optimization problem
+For a fixed kernel,
+the clustering optimization problem
 \eqref{eq:minimize} based on energy statistics 
@@
 \tfrac{1}{n_j}\sum_{y,z\in\C_j} \kk(y,z)$, which is the same as
-the second term. Thus, the kernel $k$-means optimization problem
+the second term. Thus, problem
 \eqref{eq:kernel_kmeans} is equivalent to
@@
 kernel $k$-means problem is actually a consequence
-of energy statistics. In this vein, kernel $k$-means is part of a
+of energy statistics. In this vein, kernel $k$-means problem is part of a
 statistical framework where we are maximizing distances between probability
-distributions.
-
-As shown in \cite{Dhillon}, kernel $k$-means, spectral clustering,
+distributions. The kernel function $\kk$ arises from a semimetric $\rho$ of
+negative type which defines the energy distance between distributions.
+
+As shown in \cite{Dhillon2,Dhillon}, kernel $k$-means, spectral clustering,
 and graph partitioning problems such as ratio association, ratio cut, and
-normalized cut are all equivalent to a QCQP of the form \eqref{eq:qcqp2}, thus
+normalized cut are all equivalent to a QCQP of the form \eqref{eq:qcqp2}. Thus
 one can use kernel $k$-means algorithm to solve these problems as well.
-This correspondence involves a weighted version of \eqref{eq:qcqp2}, that
-for the sake of completeness demonstrate next.
+This correspondence involves a weighted version of \eqref{eq:qcqp2} that
+we demonstrate in the following.
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Clustering Based on Weighted Energy Statistics}
-
-Now we generalize the formulas from energy statistics to incorporate
+\section{Clustering Based on Weighted Energy Statistics}
+\label{sec:weighted}
+
+We generalize the formulas from energy statistics to incorporate
 weights associated to each data point.
-Let $w(x)$ be a weight function for the point $x \in \mathbb{X}$.
+Let $w(x)$ be a weight function associated to point $x \in \mathcal{X}$.
 We can generalize \eqref{eq:g_def} as follows:
@@
 to obtain a weighted version of energy test statistic. With these
-changes, Proposition~\ref{th:minimize} remains the unnaltered, so the
+changes, Proposition~\ref{th:minimize} remains the unaltered, so the
 clustering problem becomes
@@
 H \equiv \mathcal{W}^{1/2} Y, \qquad
-\bm{\omega} \equiv \mathcal{W} \bm{e},
-\end{equation}
-where $w_i = w(x_i)$ and $\bm{e} \in \mathbb{R}^n$ is the all-ones
+\om \equiv \mathcal{W} \e,
+\end{equation}
+where $w_i = w(x_i)$ and $\e \in \mathbb{R}^n$ is the all-ones
 vector. Now we can show the analogous of
@@
 \max_H \Tr \left\{ H^\top (\mathcal{W}^{1/2} G \mathcal{W}^{1/2}) H  \right\}
-\qquad \mbox{s.t. $H \ge 0$, $H^\top H = I$, $H H^\top \bm{\omega} =
-\bm{\omega}$,}
+\qquad \mbox{s.t. $H \ge 0$, $H^\top H = I$, $H H^\top \om =
+\om$,}
 \end{equation}
@@
 \begin{proof}
-Replacing \eqref{eq:gen_kernel} and elimating the global terms which 
+Replacing \eqref{eq:gen_kernel} and eliminating the global terms which 
 do not contribute, the optimization problem \eqref{eq:minimize2}
@@
 therefore $H^\top H = I$. This is a constraint on the rows of $H$.
-Now we obtain a condition on the columns
-of $H$. Observe that
+To obtain a condition on the columns
+of $H$ observe that
 \begin{equation}
@@
 line we obtain $\sqrt{w_p} s_i^{-1} \sum_{q=1}^n w_q Z_{qi} = \sqrt{w_p}$,
-or equivalently $H H^\top \mathcal{W}^{1/2} \bm{e} = \mathcal{W}^{1/2} \bm{e}$,
-which gives the constraint $H H^\top \bm{\omega} = \bm{\omega}$.
+or equivalently $H H^\top \mathcal{W}^{1/2} \e = \mathcal{W}^{1/2} \e$,
+which gives the constraint $H H^\top \om = \om$.
 \end{proof}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Connection with Graph Partioning}
+\subsection*{Connection with Graph Partitioning}
 
 The relation between kernel $k$-means and graph partitioning problems
-was already established \cite{Dhillon}. We repeat a similar analysis
+was already established \cite{Dhillon2,Dhillon}. We repeat a similar analysis
 due to the relation of these problems to
@@
 
-Consider a graph $\mathcal{G} = (\mathcal{V}, \mathcal{E}, \mathcal{A})$,
+Consider a graph $\mathcal{G} = (\mathcal{V}, \mathcal{E}, \mathcal{A})$
 where $\mathcal{V}$ is the set of vertices, $\mathcal{E}$ the set of edges,
 and $\mathcal{A}$ is an affinity matrix of the graph 
-that measure the 
+that measures the 
 similarities between pairs of nodes. Thus, $\mathcal{A}_{ij} \ne 0$
@@
 \label{eq:assoc}
-\max_{\C_i,\dots,\C_k} \sum_{j=1}^k \dfrac{\textnormal{links}(\C_j,\C_j)}{s_j},
+\max_{\C_i,\dots,\C_k} \sum_{j=1}^k \dfrac{\textnormal{links}(\C_j,\C_j)}{s_j}
 \end{equation}
@@
 \min_{\C_i,\dots,\C_k} \sum_{j=1}^k
-\dfrac{\textnormal{links}(\C_j,\mathcal{V} \backslash \C_j)}{s_j},
+\dfrac{\textnormal{links}(\C_j,\mathcal{V} \backslash \C_j)}{s_j}
 \end{equation}
@@
 \cite{Kernighan,Malik,Chan,Yu}
-can be seen as a particular
-case of \eqref{eq:assoc} or \eqref{eq:cut}.
+can be seen as a particular case of \eqref{eq:assoc} or \eqref{eq:cut}.
 
@@
 \mathcal{A}_{pq} = \sum_{j=1}^k \sum_{p=1}^n \sum_{q=1}^n 
-\dfrac{Z^\top_{jp}}{\sqrt{s_j}} \mathcal{A}_{pq} \dfrac{Z_{qj}}{\sqrt{s_j}}
-= \Tr\left( Y^\top \mathcal{A} Y \right)
+\dfrac{Z^\top_{jp}}{\sqrt{s_j}} \, \mathcal{A}_{pq} \, 
+\dfrac{Z_{qj}}{\sqrt{s_j}}
+= \Tr\left( Y^\top \mathcal{A} Y \right) ,
 \end{equation}
@@
 \right) \qquad \mbox{s.t. $H\ge 0$, $H^\top H = I$, $H H^\top
-\bm{\omega}=\bm{\omega}$}.
+\om=\om$}.
 \end{equation}
 Therefore, this is exactly the same problem as weighted energy statistics
-clustering \eqref{eq:qcqp3}, with 
+clustering \eqref{eq:qcqp3} with 
 $G = \mathcal{W}^{-1} \mathcal{A} \mathcal{W}^{-1}$. Assuming this
@@
 where in the second equation we assume the graph has no self-loops,
-$\mathcal{A}_{ii} = 0$. Using \eqref{eq:metric_graphs} 
+i.e. $\mathcal{A}_{ii} = 0$. Using \eqref{eq:metric_graphs} 
 in \eqref{eq:g_def}--\eqref{eq:between}
@@
 let us first consider the simplest possible case which
-is one-dimensional data and a two-class problem. This will be usefull to test
+is one-dimensional data and a two-class problem. This will be useful to test
 energy statistics clustering on a simple setting.
@@
 x \left( \Ind{x \ge y} - \Ind{x < y} \right)  + 
-y \left( \Ind{y > x} - \Ind{y \le x} \right)  ,
+y \left( \Ind{y > x} - \Ind{y \le x} \right)  
 \end{aligned}
 \end{equation}
-where we have the indicator function defined as
+where we have the indicator function defined by 
 $\Ind{A}=1$ if $A$ is true, and $\Ind{A}=0$ otherwise. 
@@
 the condition of the indicator function. Assuming
-that we first order the data inside the partition, obtaining
+that we first order the data in $\C$, obtaining
 $\bar{\C} = [ x_j \in \C: x_1 \le x_2 \le \dotsm \le x_{n}]$, we
@@
 \eqref{eq:g1d}
-is $\OO(n)$, and the cost of
+is $\OO(n)$ and the cost of
 sorting the data
@@
 Assuming that each partition is ordered  $\mathbb{X} = \bigcup_{j=1}^k
-\bar{\C}_j$, but notice that the entire data set $\mathbb{X}$ does not 
-need to be necessarily ordered, the within energy dispersion
+\bar{\C}_j$ 
+the within energy dispersion
 \eqref{eq:within} can be written as
@@
 
-For a two-class problem, we can use \eqref{eq:w1d} to cluster data
+For a two-class problem we can use \eqref{eq:w1d} to cluster data
 through a simple algorithm 
@@
 \label{eq:accuracy}
-\textnormal{accuracy}(\hat{Z}) = \max_\sigma 
+\textnormal{accuracy}(\hat{Z}) \equiv \max_\sigma
 \dfrac{1}{n}\sum_{i=1}^n\sum_{j=1}^k \hat{Z}_{i \sigma(j)} Z_{ij}
@@
 \begin{minipage}{0.49\textwidth}
-\hspace{1cm}\includegraphics[width=.6\textwidth]{hist_normal.pdf}\\[-.5em]
+\includegraphics[width=.8\textwidth]{hist_normal.pdf}\\[-.8em]
 \includegraphics[width=\textwidth]{gauss1d.pdf}\\[-1em]
@@
 \begin{minipage}{0.49\textwidth}
-\hspace{1cm}\includegraphics[width=.6\textwidth]{hist_lognormal.pdf}\\[-.5em]
+\includegraphics[width=.8\textwidth]{hist_lognormal.pdf}\\[-.8em]
 \includegraphics[width=\textwidth]{loggauss1d.pdf}\\[-1em]
@@
 \label{fig:1d}
-Energy statistics clustering by Algorithm~\ref{algo1d},
-compared to $k$-means and GMM.
-Both clusters
-have the same number of points ($x$-axis), 
-which are increased in each experiment.
-We sample $100$ times from the distributions in the histograms and
-the $y$-axis is the average value of the clustering 
-accuracy \eqref{eq:accuracy} (errorbars are standard error).
-(a) 
-$x\sim \tfrac{1}{2}\left[ \mathcal{N}(\mu_1,\sigma_1) +
-\mathcal{N}(\mu_2,\sigma_2)  \right]$,  
-$\mu_1 = 0$,
-$\mu_2 = 5$,
-$\sigma_1 = 1$, and
-$\sigma_2 = 2$.
-(b) 
-$x\sim \tfrac{1}{2}\left[ e^{\mathcal{N}(\mu_1,\sigma_1)} +
-e^{\mathcal{N}(\mu_2,\sigma_2)}  \right]$,  
-$\mu_1 = 0$,
-$\mu_2 = -1.5$,
-$\sigma_1 = 0.3$, and
-$\sigma_2 = 1.5$. 
+Energy statistics clustering by Algorithm~\ref{algo1d}
+compared to $k$-means and GMM/EM.
+We have the same number of points in both clusters, and for each case
+we sample $100$ times from the distributions in the histograms. We 
+plot the average value of 
+\eqref{eq:accuracy} versus the total number of points 
+(error bars are standard error). The dashed line indicates the
+best possible classification accuracy computed from Bayes error.
+(a) Data coming from \eqref{eq:two_normal}, where the
+optimal accuracy is $\approx 0.956$.
+(b) Data from \eqref{eq:two_lognormal}, where the optimal
+accuracy is $\approx 0.852$.
 }
@@
 Let us consider two simple experiments with equal number of points
-in each cluster. 
-We keep increasing the number of points in the clusters for each experiment, 
-and cluster the data using Algorithm~\ref{algo1d}. 
-We also cluster the same data set
-with GMM, through EM algorithm, and with standard $k$-means. In both of these
-cases we use the initialization from $k$-means++ \cite{Vassilvitskii} 
+in each cluster. We plot the accuracy \eqref{eq:accuracy} versus the number
+of points in each cluster.
+The data is clustered using 
+Algorithm~\ref{algo1d}, GMM through EM algorithm, and standard $k$-means 
+algorithm. In the last two cases 
+we use the initialization from $k$-means++ \cite{Vassilvitskii} 
 and we run the algorithms
-few times with different initializations and choose the answer
-with best objective function value. We use \eqref{eq:accuracy} to measure
-the clustering quality. Furthermore, we pick $100$ different samples
-for each configuration, and show the average accuracy with errorbars indicating
+few times choosing the result with  best objective function value. 
+Notice that Algorithm~\ref{algo1d} requires no initialization.
+Moreover, for each case we sample $100$ times
+and show the average accuracy with error bars indicating
 the standard error.
 In  Fig.~\ref{fig:1d}a 
-we have data from normal distributions,
-where we can see that all the three methods
-perform closely, with a slight advantage of GMM, as expected, since
+we sample data from two normal distributions, with equal number of
+points in each cluster:
+\begin{equation}
+\label{eq:two_normal}
+x \sim \mathcal{N}\big(\mu_i,\sigma_i^2\big) 
+\quad 
+\mbox{with $\mu_1 = 0$, $\sigma_1=1$ and
+$\mu_2 = 5$, $\sigma_2 = 2$.}
+\end{equation}
+The optimal accuracy obtained from Bayes classification error
+is $ \approx 0.956$, indicated by the dashed line in the plot.
+We see that the three methods
+perform closely, with a slight advantage of GMM as expected since
 it is the true model for the data. Energy statistics performs slightly better
-than $k$-means in this experiment. Now in Fig~\ref{fig:1d}b
-we consider one example of 
-data from lognormal distributions. We can see that energy
-statistics clustering through Algorithm~\ref{algo1d} provides a huge 
-improvement
-compared to both GMM and $k$-means, which basically cluster at chance
-in this case.
-GMM is worst than chance since sometimes it 
-is not able to estimate the parameters, thus giving zero
-accuracy.
+than $k$-means. In Fig~\ref{fig:1d}b
+we consider two clusters with lognormal distributions:
+\begin{equation}
+\label{eq:two_lognormal}
+\log x \sim \mathcal{N}\big( \mu_i,\sigma_i^2\big) \quad
+\mbox{with
+$\mu_1 = 0$,
+$\sigma_1 = 0.3$ and
+$\mu_2 = -1.5$,
+$\sigma_2 = 1.5$.}
+\end{equation}
+The optimal accuracy is $\approx 0.852$.
+In this case,
+Algorithm~\ref{algo1d} provides a very accurate clustering while
+GMM and $k$-means basically cluster at chance.
+Sometimes GMM/EM  was unable to estimate the parameters 
+thus giving zero accuracy.
 These two simple experiments illustrate
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Iterative Algorithm for Energy Statistics Clustering}
+\section{Iterative Algorithms for Energy Statistics Clustering}
 \label{sec:algo}
 
-In this section we will introduce a new iterative algorithm to find a local
-maximizer of the QCQP \eqref{eq:qcqp2}, however, due to 
+In this section we introduce an iterative algorithm to find a local
+maximizer of \eqref{eq:qcqp2}. However, due to 
 Proposition~\ref{th:kernel_kmeans} we can also find an approximate
-solution by the well-known kernel $k$-means algorithm, which 
-for convenience
-will also be restated in the present context.
-First, let us introduce some base notation.
+solution by the well-known kernel $k$-means algorithm 
+(see \cite{Dhillon2,Dhillon}), which 
+for convenience will also be restated in the present context.
 
@@
 \end{equation}
-where, we recall, $G_{i\bullet}$ ($G_{\bullet i}$) denotes
+where we recall that $G_{i\bullet}$ ($G_{\bullet i}$) denotes
 the $i$th row (column) of matrix $G$.
 
-\subsection{Kernel $\bm{k}$-Means Algorithm}
-
-To optimize kernel $k$-means objective function
-\eqref{eq:J}, we remove the global term and define the function
+\subsection*{Kernel $\bm{k}$-Means Algorithm}
+
+To optimize the kernel $k$-means objective function
+\eqref{eq:J} we remove the global term and define the function
 \begin{equation}
 \label{eq:Jell}
-J^{(\ell)}(x_i) \equiv -\dfrac{2}{n_\ell} Q_\ell(x_i) + \dfrac{1}{n_\ell^2}
-Q_\ell,
+J^{(\ell)}(x_i) \equiv 
+\dfrac{1}{n_\ell^2} Q_\ell
+-\dfrac{2}{n_\ell} Q_\ell(x_i) 
 \end{equation}
@@
 to $j^\star = \argmin_\ell J^{(\ell)}(x_i)$, for $\ell = 1,\dotsc,k$.
-This procedure is performed for every data point, and repeated until
+This procedure is performed for every data point and repeated until
 convergence, i.e. until no new assignments are made.
-The whole procedure shown in Algorithm~\ref{kmeans_algo}.
+The entire procedure is in Algorithm~\ref{kmeans_algo}.
 It can be shown that this algorithm converges provided $G$ is positive
 semidefinite.
-Although our notation looks a little different than the standard
-kernel $k$-means found in the literature \cite{Dhillon2,Dhillon}, 
-this is precisely
-the same algorithm but written in a more concise and explicit way.
 
@@
 Gram matrix $G$ having
-$n'$ nonzero elements, this complexity can be further reduced
+$n'$ nonzero elements this complexity can be further reduced
 to $\OO(k n')$.
@@
     \OUTPUT label matrix $Z$ 
-  \STATE $\bm{q} \leftarrow (Q_1, \dotsc, Q_k)^\top$ 
+  \STATE $q \leftarrow (Q_1, \dotsc, Q_k)^\top$ 
             have the costs of each cluster, according to \eqref{eq:maxQ}
-  \STATE $\bm{n} \leftarrow (n_1,\dotsc,n_k)^\top$ 
+  \STATE $n \leftarrow (n_1,\dotsc,n_k)^\top$ 
         have the number of points in each cluster, obtained 
@@
             $Z_{ij^\star} \leftarrow 1$
-            \STATE update $\bm{n}$: $n_j \leftarrow n_j - 1$ and
+            \STATE update $n$: $n_j \leftarrow n_j - 1$ and
                     $n_{j^\star} \leftarrow n_{j^\star} + 1$
-            \STATE update $\bm{q}$: $q_j \leftarrow q_j - 2Q_j(x_i)$ and
+            \STATE update $q$: $q_j \leftarrow q_j - 2Q_j(x_i)$ and
     $q_{j^\star} \leftarrow q_{j^\star} + 2Q_{j^\star}(x_i)$
@@
 
-\subsection{Energy Cost Algorithm}
-
-Now let us consider a different algorithm, which is based on the change
-in the within energy statistics when moving a given data point to
-a different cluster.
-Suppose we have a data point $x_i \in \mathcal{X}$
-which is currently assigned to  cluster $\C_j$, yielding
+\subsection*{Hartigan's Algorithm for Energy Statistics Clustering}
+
+We now consider Hartigan's algorithm \cite{Hartigan} applied to 
+\eqref{eq:maxQ}. This gives approximate solutions to 
+\eqref{eq:qcqp2}. This approach was also proposed in \cite{Kgroups}.
+The method is based on the change
+in the within energy statistic when moving a given data point to
+another cluster.
+Suppose point $x_i \in \mathcal{X}$
+is currently assigned to  cluster $\C_j$, yielding
 a total energy cost function \eqref{eq:maxQ} denoted by $Q^{(j)}$.
 Let us consider the change in the total energy by moving
-$x_i$ to cluster $\C_\ell$. 
-Denote the new energy cost after moving $x_i$ to $\C_\ell$ by $Q^{(\ell)}$.
+$x_i$ to $\C_\ell$. 
+Denote this new energy cost by $Q^{(\ell)}$.
 It is straightforward to see that
@@
 moving $x_i$ to $\C_\ell$, otherwise we better keep $x_i$ in $\C_j$. Based on
-this we propose an algorithm where
-the iterates are performed as follows.
+this the algorithm goes as follows.
 We start with an initial configuration for the label matrix $Z$, 
@@
 The process is repeated
-until convergence, i.e. until no points are assigned to new clusters. 
+until no points are assigned to new clusters. 
 This whole procedure is described in Algorithm~\ref{algo}.
-Note that \eqref{eq:changeQ} assures that the objective function is
-monotonically increasing at each iteration, therefore the algorithm
+Note that this ensures that the objective function is
+monotonically increasing at each iteration, and consequently the algorithm
 converges in a finite number of steps.
@@
     \OUTPUT label matrix $Z$
-  \STATE $\bm{q} \leftarrow (Q_1, \dotsc, Q_k)^\top$ 
+  \STATE $q \leftarrow (Q_1, \dotsc, Q_k)^\top$ 
             have the energy costs of each cluster, according to \eqref{eq:maxQ}
-  \STATE $\bm{n} \leftarrow (n_1,\dotsc,n_k)^\top$ have the number of points 
+  \STATE $n \leftarrow (n_1,\dotsc,n_k)^\top$ have the number of points 
         in each cluster, obtained from $D=Z^\top Z$
@@
             $Z_{ij^\star} \leftarrow 1$
-            \STATE update $\bm{n}$: $n_j \leftarrow n_j - 1$ and
+            \STATE update $n$: $n_j \leftarrow n_j - 1$ and
                     $n_{j^\star} \leftarrow n_{j^\star} + 1$
-            \STATE update $\bm{q}$: $q_j \leftarrow q_j - 2Q_j(x_i)$ and
+            \STATE update $q$: $q_j \leftarrow q_j - 2Q_j(x_i)$ and
     $q_{j^\star} \leftarrow q_{j^\star} + 2\left(Q_{j^\star}(x_i)+
@@
 \caption{\label{algo}
-Energy cost algorithm to find an approximate solution to \eqref{eq:qcqp2}.
+Hartigan's method to find an approximate solution to \eqref{eq:qcqp2}.
 \hspace{\fill}
@@
 
-Notice that computing $G$ requires $\OO( D n^2)$ operations, where 
+Computing $G$ requires $\OO( D n^2)$ operations, where 
 $D$ is the dimension of each data point and $n$ is the data size. However,
-both previous algorithms assume that $G$ is given. There are more efficient
-methods to compute $G$, specially if it is sparse. We will not consider
-this further, and just assume that $G$ is given.
+both Algorithms~\ref{kmeans_algo} and \ref{algo} 
+assume that $G$ is given. There are more efficient
+methods to compute $G$, specially if it is sparse, but we will not consider
+this further and just assume that $G$ is given.
 The computation of each cluster cost
-$Q_j$ has complexity $\OO(n_j^2)$, and overall to compute $\bm{q}$
+$Q_j$ has complexity $\OO(n_j^2)$, and overall to compute $q$
 we have $\OO(n_1^2+\dots + n_k^2) = \OO(k \max_j n_j^2)$. 
-These operations, however, only need to be performed a single time. Now for
+These operations only need to be performed a single time. For
 each point $x_i$ we need to compute $Q_j(x_i)$ once, which is
@@
 in kernel $k$-means, Algorithm~\ref{kmeans_algo}. Again, if $G$ is sparse
-this can be reduced to $\OO(k n')$, where $n'$ is the number of nonzero
+this can be reduced to $\OO(k n')$ where $n'$ is the number of nonzero
 entries of $G$.
+
+Some important properties of Hartigan's method 
+are worth mentioning.
+
+\begin{theorem}[Telgarsky-Vattani \cite{Telgarsky}]
+Hartigan's method has the cost function strictly decreasing in each
+iteration. Moreover, if $n > k$ then 
+\begin{enumerate}
+\item \label{noempty} the resulting partition has no empty clusters, and
+\item \label{diffmean} the resulting partition has distinct means.
+\end{enumerate}
+\end{theorem}
+
+Both condition above are not satisfied 
+by Lloyd's method, and consequently by (kernel) $k$-means algorithm.
+
+\begin{theorem}[Telgarsky-Vattani \cite{Telgarsky}]
+The set of local optima of Hartigan's method is a (possibly strict) subset
+of local optima of Lloyd's method.
+\end{theorem}
+
+This means that Algorithm~\ref{kmeans_algo} cannot
+improve on a local optima of Algorithm~\ref{algo}. On the other hand,
+Algorithm~\ref{algo} might improve on a local optima of 
+Algorithm~\ref{kmeans_algo}. Lloyd's method forms Voronoi partitions,
+while Hartigan's method groups data
+in regions formed by the intersection of spheres called circlonoi cells.
+It can be shown that the circlonoi cells are contained within
+a smaller volume of a Voronoi cell, and this excess volume grows
+exponentially with the dimension of $\mathcal{X}$ 
+\cite[Theorems 2.4 and 3.1]{Telgarsky}. 
+Points in this excess volume
+force Hartigan's method to iterate, contrary
+to Lloyd's method. Therefore, Hartigan's method 
+can potentially escape local
+optima of Lloyd's method. 
+Moreover, this improvement should be more prominent as
+dimension increases. Also, the improvement grows as $k$
+increases.
+The empirical results of \cite{Telgarsky} show that 
+an implementation of Hartigan's method has comparable execution time 
+as an implementation of
+Lloyd's method,
+but no explicit complexity was provided. In our case, we showed that both
+Algorithms~\ref{kmeans_algo} and \ref{algo} have the same time complexity.
+
+In \cite{Slonin} Hartigan's method was applied to $k$-means problem
+with any Bregman divergence. They showed that the number of Hartigan's
+local minima is upper bounded by $O(1/k)$ \cite[Proposition 5.1]{Slonin}. 
+In addition, they provide examples where
+\emph{any} initial partition correspond to a local optima of Lloyd's 
+method, while  the number of local optima in Hartigan's is small and 
+correspond to true partitions of the data. Empirically, the number of
+Hartigan's local optima was considerably smaller than the number of Lloyd's
+local optima.
+These results indicate that Hartigan's method
+provides several advantages over Lloyd's method. 
+We will see this explicitly in the following 
+numerical results where Algorithm~\ref{algo}
+outperforms Algorithm~\ref{kmeans_algo} when using the same kernel.
 
@@
 according to the traditional energy distance \eqref{eq:energy}, and
-the point $x_0=0$ is chosen in the associated kernel  
+the point $x_0=0$ is chosen in the corresponding kernel  
 \eqref{eq:kernel_semimetric}. We thus have
@@
 \end{equation}
-We will consider other semimetrics/kernels as well, 
-but the above will be considered the
+We will consider other kernels as well, 
+but \eqref{eq:standard_metric} will be considered the
 standard kernel for energy statistics and will always be present in every
-experiment as a reference. Notice that this is a convention, 
-we could have chosen
-any other semimetric as the standard.
-One of the main goals of the following 
-experiments is to compare Algorithm~\ref{algo} to
-kernel $k$-means algorithm, described in Algorithm~\ref{kmeans_algo}. Thus,
-for every kernel used in Algorithm~\ref{algo}, we also use the same 
-kernel in Algorithm~\ref{kmeans_algo}. Another goal is to compare
-Algorithm~\ref{algo} with $k$-means and GMM (through expectation maximization
-algorithm), as these are the most used
+experiment for reference. 
+%Notice that this is a convention, 
+%we could have chosen
+%any other semimetric/kernel as the standard.
+
+Let us briefly mention that we compared Algorithms~\ref{algo} and
+\ref{algo1d} for several univariate distributions, and
+both perform almost indistinguishable regarding the clustering quality.
+We omit these results since we will analyse more interesting scenarios.
+
+The main goal of the 
+experiments shown in this section 
+is to compare Algorithm~\ref{algo}, based on Hartigan's method, to
+kernel $k$-means as described in Algorithm~\ref{kmeans_algo}, based on
+Lloyd's method. From the discussion in the previous section we expect 
+Algorithm~\ref{algo} to be superior than Algorithm~\ref{kmeans_algo}.
+Another goal is to illustrate the nonparametric aspect of energy statistics.
+To this end we also compare
+Algorithm~\ref{algo} to standard $k$-means and GMM/EM since 
+these are the most used
 clustering algorithms in practice.
-Since for synthetic data the true labels are available,
-our measure of clustering quality
-will be \eqref{eq:accuracy}. Moreover, for all algorithms, we always
-choose the initialization from $k$-means++ \cite{Vassilvitskii}.
-
-We first consider clustering in high dimensions and analyze 
+Moreover, for every algorithm, we always
+choose the initialization procedure from $k$-means++%
+\footnote{Notice that we just use the initialization
+procedure and not the full $k$-means++ algorithm.} \cite{Vassilvitskii}.
+Our measure of clustering quality will be the accuracy \eqref{eq:accuracy}
+based on the ground truth. Furthermore, in every experiment we sample
+data many times and show the average value of the accuracy, with error bars
+indicating standard error. Whenever possible, for reference, 
+we also show the best possible accuracy obtained from a numerical estimation
+of Bayes classification error, which uses the true probability distribution
+of the data.
+
+From the results of \cite{Telgarsky}, the improvement of Hartigan's method
+over Lloyd's method should be more prominent in high dimensions.
+To analyse this we show 
 how the algorithms degrade as the number of dimensions increase, while
-keeping the number of points in each cluster fixed. The Bayes error
-is also kept fixed as ambient dimension increases.
-In Figure~\ref{fig:gauss}a we have data generated from $D$-variate normal 
-distributions as follows: 
+keeping the number of points in each cluster fixed. Consider
+two clusters with multivariate normal distributions given by
 \begin{equation}
@@
 \begin{split}
-&x \sim \tfrac{1}{2}\left[ 
-\mathcal{N}(\mu_1,\Sigma_1) + \mathcal{N}(\mu_2, \Sigma_2)\right], \\
-&\mu_1 = (\underbrace{0,\dotsc,0}_{\times D})^\top , \quad
+x &\in \C_i  \sim 
+\mathcal{N}(\mu_i,\Sigma_i) \qquad (i=1,2),  \\
+\mu_1 &= (\underbrace{0,\dotsc,0}_{\times D})^\top , \quad
 \mu_2 = 0.7 \times (\underbrace{1,\dots,1}_{\times 10},
@@
 \end{equation}
-We only keep signal in
-in the first $10$
-dimensions of $\mu_2$, and keep increasing the ambient dimension $D$. For each
-$D$, we perform $100$ experiments, obtaining the clustering accuracy
-for each algorithm.
-We can see that
-GMM is not able to estimate the covariance matrix 
-when the number
-of dimensions exceeds the number of points in each cluster, so it gives
-zero accuracy for $D \gtrsim 100$.
-In Figure~\ref{fig:gauss}b we have the same type of experiment but 
-with 
+Note that the Bayes error
+is also kept fixed as $D$ increases, given an optimal classification
+accuracy of $\approx 0.86$.
+We keep signal only the first $10$ dimensions of $\mu_2$ and increase $D$.
+For each $D$ we sample $100$ times, picking $100$ points for each cluster.
+We apply each algorithm to the resulting data set 
+and compute the average value of the accuracy \eqref{eq:accuracy}.
+For both Algorithms~\ref{algo} and \ref{kmeans_algo} we use the standard
+kernel \eqref{eq:standard_metric} from energy statistics.
+The results are shown in Fig.~\ref{fig:gauss}a, where error bars indicate
+the standard error.
+Notice that GMM/EM algorithm is unable to estimate the covariance matrix 
+when the number of dimensions exceeds the number of points in each cluster.
+We see that Algorithm~\ref{algo} performs better than all the other ones,
+and in particular it works better than kernel $k$-means as the number
+of dimensions increase. Note also that although GMM and $k$-means are
+parametric, they are consistent models for the above distributions.
+Now we consider the same experiment but with different covariances:
 \begin{equation}
@@
 \begin{split}
-&x \sim \tfrac{1}{2}\left[ 
-\mathcal{N}(\mu_1,\Sigma_1) + \mathcal{N}(\mu_2, \Sigma_2)\right], \\
-&\mu_1 = (\underbrace{0,\dotsc,0}_{\times D})^\top , \,
-\mu_2 = 0.7 \times (\underbrace{1,\dots,1}_{\times 10},
-\underbrace{0,\dots,0}_{\times (D-10)})^\top, \,
-\Sigma_1 = I_D, \, 
-\Sigma_2 = \left( \begin{smallmatrix} \tfrac{1}{2} I_{10} & 0 \\ 0 & I_{D-10}
-\end{smallmatrix}\right). \quad
+x &\in \C_i  \sim 
+\mathcal{N}(\mu_i,\Sigma_i) \qquad (i=1,2),  \\
+\mu_1 &= (\underbrace{0,\dotsc,0}_{\times D})^\top , \quad
+\mu_2 = (\underbrace{1,\dots,1}_{\times 10},
+\underbrace{0,\dots,0}_{\times (D-10)})^\top, \quad
+\Sigma_1 = I_D, \quad 
+\Sigma_2 = \left( \begin{smallmatrix} \widetilde{\Sigma} & 0 \\ 0 & I_{D-5}
+\end{smallmatrix}\right)
 \end{split}
 \end{equation}
-Therefore, for both experiments shown in Figure~\ref{fig:gauss}
+where $\widetilde{\Sigma}_{jj} = j$ for $j=1,\dotsc,5$.
+The best possible accuracy from Bayes classification error is
+$\approx 0.95$.
+The results are shown In Figure~\ref{fig:gauss}b.
+In the experiments of Figure~\ref{fig:gauss}
 we can see a better performance of Algorithm~\ref{algo} compared
-to the other ones, in particular compared to kernel $k$-means algorithm,
-where we recall that both
-aim at optimizing the same problem (see Proposition~\ref{th:kernel_kmeans}). 
-Also, notice that $k$-means
-and GMM are consistently the right model for this dataset, so it is hard
-to perform better than these algorithms in this current setting.
-Notice that Algorithm~\ref{algo} is more robust as the ambient
-dimension increases.
+to the other ones, and in particular to kernel $k$-means,
+where we recall that both find approximate solutions
+to the same optimization problem \eqref{eq:qcqp2}.
+Notice that Algorithm~\ref{algo} is more robust as the number of dimensions
+increase.
 
@@
 \label{fig:gauss}
-Effect of increasing the
-ambient dimension while keeping Bayes error fixed, for 
-two clusters with normally distributed data with $100$ points in each
-cluster. 
-(a) 
-We increase $D$ as described in \eqref{eq:gauss1}. The blue
-line correspond to Algorithm~\ref{algo}, while
-the magenta line 
-corresponds to kernel $k$-means, Algorithm~\ref{kmeans_algo}.
-(b) The same but with data following \eqref{eq:gauss2}.
-One notices that
-Algorithm~\ref{algo} is more robust than the other ones.
+Comparison of Algorithm~\ref{algo}, Algorithm~\ref{kmeans_algo},
+standard $k$-means and GMM/EM as the number of dimensions increase.
+We sample $100$ times and show the average of \eqref{eq:accuracy} with
+error bars being standard error. We have two clusters with $100$ points
+each. 
+(a) Data distributed as in \eqref{eq:gauss1}. The optimal accuracy
+from Bayes error is indicated by the dashed line which is $\approx 0.86$.
+(b) Data from \eqref{eq:gauss2}. The optimal accuracy is $\approx 0.95$.
 }
@@
 \begin{split}
-&x \sim  
-\dfrac{n_1}{N} \mathcal{N}(\mu_1,\Sigma_1) + 
-\dfrac{n_2}{N} \mathcal{N}(\mu_2, \Sigma_2), 
-\quad \mu_1 = (0,0,0,0)^\top , \,
+x &\in \C_i \sim  
+\dfrac{n_i}{N} \mathcal{N}(\mu_i,\Sigma_i) \quad (i=1,2), \quad 
+\mu_1 = (0,0,0,0)^\top , \quad
 \mu_2 = 1.5\times (1,1,0,0)^\top, \\
-&
-\Sigma_1 = I_4, \quad
+\Sigma_1 &= I_4, \quad
 \Sigma_2 = \left( 
@@
 \end{equation}
-We then increase $m$, i.e. we make the clusters progressively more unbalanced.
-We generate $100$ experiments for each $m$, and plot the clustering accuracy
-versus $m$.
+We increase $m$, that is we make the clusters progressively more unbalanced,
+and plot the mean of \eqref{eq:accuracy} with standard errors, where
+we sample $100$ times for each $m$.
 As expected, GMM
-works better than the other algorithms in the case of unbalanced clusters. 
-This is mostly due to its soft assignments.
+works better than the other algorithms since it is a fuzzy clustering.
 We can see that the other methods based on hard assignments degrade similarly,
 and  more rapidly than GMM. This indicates that a fuzzy version of
-energy statistics clustering should compensate for this effect.
+Algorithm~\ref{algo} should compensate for this effect.
 
@@
 \label{fig:unbalanced}
-Previous algorithms for unbalanced clusters, 
-according to \eqref{eq:gauss3}.}
+Comparison of Algorithm~\ref{algo}, Algorithm~\ref{kmeans_algo}, standard
+$k$-means and GMM/EM for data distributed as in \eqref{eq:gauss3}, where
+we make the two clusters progressively more unbalanced.
+}
 \end{figure}
 
-Now, besides \eqref{eq:standard_metric} we consider two other semimetrics:
+Besides the standard kernel from energy statistics 
+\eqref{eq:standard_metric}, associated to \eqref{eq:energy}, 
+let us consider two other semimetrics with their respective generating
+kernels,
 \begin{align}
-\rho_{1/2}(x,y) &= \| x-y \|^{1/2}, & \kk(x,y) &= \tfrac{1}{2} \left( 
+\rho_{1/2}(x,y) &= \| x-y \|^{1/2} & 
+ \kk_{1/2}(x,y) &= \tfrac{1}{2} \left( 
 \| x \|^{1/2} + \| y \|^{1/2} 
 - \| x-y \|^{1/2} \right), \label{eq:rhohalf}\\
-\rho_{e}(x,y) &= 2 - 2 e^{-\| x- y\|/2}, & \kk(x,y) &= e^{-\| x-y\|/2}.
+\rho_{e}(x,y) &= 
+2 - 2 e^{-\tfrac{1}{2}\| x- y\|} &
+ \kk_{e}(x,y) &= e^{-\tfrac{1}{2}\| x-y\|}.
 \label{eq:rhoe}
 \end{align}
-In Figure~\ref{fig:consist}a we have data in $20$ dimensions distributed as
+The kernel \eqref{eq:rhohalf} corresponds to the energy distance
+\eqref{eq:energy2} with $\alpha=1/2$.
+Now consider data normally distributed in $D=20$ as follows:
 \begin{equation}
@@
 \begin{split}
-x &\sim \tfrac{1}{2}\left[ \mathcal{N}(\mu_1,\Sigma_1) + \mathcal{N}(\mu_2,
-\Sigma_2) \right], \\
+x &\in \C_i \sim \mathcal{N}(\mu_i,\Sigma_i) \qquad (i=1,2), \\
 \mu_1 &= (\underbrace{0,\dotsc,0}_{\times 20})^\top ,\quad
@@
 \end{equation}
-We increase the number of points in each cluster and show the clustering
-accuracy with different algorithms. The new semimetrics \eqref{eq:rhohalf}
-and $\eqref{eq:rhoe}$ are indicated in the legend. One can see that
+We sample an equal number of points for each cluster, which is
+increased. The optimal accuracy of \eqref{eq:20gauss} based on Bayes
+classification error is $\approx 0.90$. We repeat the same experiment as
+before but now using also the kernels \eqref{eq:rhohalf} and
+\eqref{eq:rhoe}. The results are shown in Figure~\ref{fig:consist}a.
 Algorithm~\ref{algo} performs better than all the other ones, and in 
-particular \eqref{eq:rhoe} provides better results.
-As the number of datapoints 
-get large enough, GMM starts to be as accurate as clustering
-based on energy statistics, as it should since it is 
-consistent model to the data. In Figure~\ref{fig:consist}b, however, we
-use the same parameters as in \eqref{eq:20gauss} but now with data
-log-normally distributed:
+particular the kernel \eqref{eq:rhoe} provides better results.
+As the number of points get large enough, GMM starts to approach
+optimal Bayes accuracy, 
+as it should since it is 
+consistent model to the distribution of the data. However, 
+Algorithm~\ref{algo} with kernel \eqref{eq:rhoe} achieves optimal accuracy
+with a much smaller number of points. Moreover, Algorithm~\ref{algo} performs
+better than Algorithm~\ref{kmeans_algo} for any of the kernel choices.
+Standard $k$-means performs poorly on this example.
+
+Consider the same parameters as in \eqref{eq:20gauss} but now with
+lognormal distribution, i.e.
 \begin{equation}
 \label{eq:20loggauss}
-x \sim \tfrac{1}{2}\left[ e^{\mathcal{N}(\mu_1,\Sigma_1)}
-+ e^{\mathcal{N}(\mu_2, \Sigma_2)}\right].
-\end{equation}
-We see that clustering based on energy statistics still performs accurately
-for this kind of data, while $k$-means works a little bit better than chance,
-and GMM is not even able to estimate the parameters. Again, \eqref{eq:rhoe}
-provides slightly better results than \eqref{eq:standard_metric} or
-\eqref{eq:rhohalf}. Notice also that Algorithm~\ref{algo} performs better
-than Algorithm~\ref{kmeans_algo}. Both experiments in Figure~\ref{fig:consist}
-shows that energy statistics clustering is nonparametric, since it is able
-to cluster data coming from very different distributions.
+\log x \in \C_i \sim \mathcal{N}(\mu_i, \Sigma_i) \qquad (i=1,2).
+\end{equation}
+The same previous experiment is now shown in 
+Fig.~\ref{fig:consist}b.
+We see that Algorithm~\ref{algo} still performs accurately, 
+while $k$-means works a little bit better than chance,
+and GMM is not even able to estimate the parameters. Again, the
+kernel \eqref{eq:rhoe}
+provides better results than \eqref{eq:standard_metric} or
+\eqref{eq:rhohalf}. Algorithm~\ref{algo} performs better
+than Algorithm~\ref{kmeans_algo} for every kernel choice. 
+The experiments in Fig.~\ref{fig:consist}
+illustrates that energy statistics clustering is nonparametric.
 
@@
 \label{fig:consist}
-(a) Data normally distributed as in 
-\eqref{eq:20gauss}. We increase the number of points in each cluster to
-illustrate the statistical consistency of the algorithms.
-(b) The same experiment but for data following \eqref{eq:20loggauss}.
-In both experiments, for each case we run every algorithm $100$ times and show
-the average results. One can see the better performance of energy statistics
-clustering, Algorithm~\ref{algo}, and in particular by using the semimetric
-\eqref{eq:rhoe}. These two figures illustrate that energy statistics clustering
-is nonparametric since it works well for very different distributions.
+Comparison of Algorithms~\ref{algo} and \ref{kmeans_algo} with kernels
+\eqref{eq:standard_metric}, \eqref{eq:rhohalf} and \eqref{eq:rhoe}, and
+also with $k$-means and GMM/EM. The optimal accuracy in both cases
+is $\approx 0.9$. We show average of \eqref{eq:accuracy}
+over $100$ samples with standard error.
+(a) Data distributed as in 
+\eqref{eq:20gauss}. 
+(b) Data distributed as in \eqref{eq:20loggauss}.
+%One can see the better performance of energy statistics
+%clustering, Algorithm~\ref{algo}, and in particular by using the semimetric
+%\eqref{eq:rhoe}. 
+These experiments illustrates how energy statistics
+based clustering is nonparametric.
 }
 \end{figure}
+
+\begin{figure}
+\begin{minipage}{0.25\textwidth}
+\includegraphics[width=1\textwidth]{2cigars.pdf}\\[-1em] (a)
+\end{minipage}
+\begin{minipage}{0.25\textwidth}
+\includegraphics[width=1\textwidth]{2circles.pdf}\\[-1em] (b)
+\end{minipage}
+\begin{minipage}{0.25\textwidth}
+\includegraphics[width=1\textwidth]{3circles.pdf}\\[-1em] (c)  
+\end{minipage}
+\begin{minipage}{0.22\textwidth}
+\vspace{-0.3em}
+\includegraphics[width=1\textwidth]{mnist.pdf}\\[-1.8em] (d)  
+\end{minipage}
+\caption{\label{fig:other}
+(a) Parallel cigars, $200$ points in each class. (b) Two  
+concentric
+circles with noise, $400$ points each. (c) Three
+concentric circles, $400$ points each.
+Clustering for (a--c) is shown in Table~\ref{table:other}.
+(d) MNIST handwritten digits. Clustering on Table~\ref{table:mnist}.
+}
+\end{figure}
+
+Now we consider two dimensional data as illustrated in
+Fig~\ref{fig:other}(a--c)
+with the following choices of semimetric and corresponding generating kernel:
+\begin{align}
+\rho_{\alpha}(x,y) &= \| x - y\|^\alpha & 
+\kk_{\alpha}(x,y) &= \tfrac{1}{2}\left(
+\| x \|^\alpha +
+\| y \|^\alpha -
+\| x-y \|^\alpha \right) ,
+\label{eq:rhoalpha} \\
+%
+\widetilde{\rho}_{\sigma}(x,y) &= 2 - 2 e^{-\tfrac{\|x-y\|}{2 \sigma}} &
+\widetilde{\kk}_{\sigma}(x,y) &= e^{-\tfrac{\|x-y\|}{2\sigma}} ,
+\label{eq:rhoesigma} \\
+%
+\widehat{\rho}_{\sigma}(x,y) &= 2 - 2 e^{-\tfrac{\|x-y\|^2}{2 \sigma^2}} &
+\widehat{\kk}_{\sigma}(x,y) &= e^{-\tfrac{\|x-y\|^2}{2\sigma^2}} .
+\label{eq:rhogsigma} 
+\end{align}
+We apply both Algorithms~\ref{algo} and \ref{kmeans_algo}, besides
+$k$-means and GMM, using the above kernels. The results are shown
+in Table~\ref{table:other} where the respective choice of parameters for the
+kernels are indicated. In the data shown in Fig.~\ref{fig:other}a 
+the standard semimetrics from energy statistics, $\rho_1$ and $\rho_{1/2}$, are
+able to provide more accurate results compared to $k$-means. However,
+the gaussian kernel $\widetilde{\rho}_2$ gives results very accurate
+results, similar to GMM,
+which is the true model of the data. In Fig.~\ref{fig:other}b we see
+that the choice of kernel is very sensitive, and only \eqref{eq:rhoesigma}
+was able to cluster accurately. The same kernel choice to the case
+of Fig.~\ref{fig:other} still provides better results than
+the other choices, but the results are less accurate.
+
+\newcolumntype{g}{>{\columncolor{gray!20}}l}
+\begin{table}
+\begin{tabular}{@{}r  l g  l g  l g@{}}
+\toprule[1pt]
+ & & \emph{Fig.~\ref{fig:other}a}
+ & & \emph{Fig.~\ref{fig:other}b}
+ & & \emph{Fig.~\ref{fig:other}c} \\
+\midrule[0.5pt]
+\multirow{4}{*}{\emph{Algorithm \ref{algo}}~~~~} 
+& $\rho_{1}$ & $0.766\pm0.066$
+& $\rho_{1}$ & $0.522\pm0.006$
+& $\rho_{1}$ & $0.437\pm0.030$ \\
+& $\rho_{1/2}$ & $0.859\pm0.062$
+& $\rho_{1/2}$ & $0.524\pm0.007$
+& $\rho_{1/2}$ & $0.547\pm0.026$ \\
+& $\widetilde{\rho}_{2}$ & $0.971\pm0.015$
+& $\widetilde{\rho}_{1}$ & $\bm{0.9999\pm0.0001}$
+& $\widetilde{\rho}_{2}$ & $\bm{0.677\pm0.003}$ \\
+& $\widehat{\rho}_{2}$  & $\bm{0.998\pm0.001}$
+& $\widehat{\rho}_{1}$  & $0.597\pm0.052$
+& $\widehat{\rho}_{2}$ & $0.645\pm0.012$ \\
+\arrayrulecolor{gray!80}\midrule[0.5pt]
+\multirow{4}{*}{\emph{Algorithm \ref{kmeans_algo}}~~~~} 
+& $\rho_{1}$ & $0.758\pm 0.069$
+& $\rho_{1}$ & $0.516\pm 0.002$
+& $\rho_{1}$  & $0.452\pm 0.030$ \\
+& $\rho_{1/2}$ & $0.901\pm 0.060$
+& $\rho_{1/2}$ & $0.524\pm 0.007$
+& $\rho_{1/2}$ & $0.570\pm 0.016$ \\
+& $\widetilde{\rho}_{2}$ & $0.971\pm 0.015$
+& $\widetilde{\rho}_{1}$ & $\bm{0.9999\pm 0.0001}$
+& $\widetilde{\rho}_{2}$ & $\bm{0.673\pm 0.002}$ \\
+& $\widehat{\rho}_{2}$ & $\bm{0.998\pm 0.001}$
+& $\widehat{\rho}_{1}$ & $0.528\pm 0.008$
+& $\widehat{\rho}_{2}$ & $0.640\pm 0.013$ \\
+\arrayrulecolor{gray!80}\midrule[0.5pt]
+\emph{$k$-means}~~~~ 
+& & $0.599\pm0.046$
+& & $0.521\pm0.005$
+& & $0.360\pm0.004$ \\
+\emph{GMM}~~~~
+& & $0.9995\pm0.0003$
+& & $0.598\pm0.018$
+& & $0.479\pm0.021$ \\
+\arrayrulecolor{black}\bottomrule[1pt]
+\end{tabular}
+\caption{\label{table:other}
+Clustering the data shown in Fig.~\ref{fig:other} with Algorithms~\ref{algo}
+and \ref{kmeans_algo}, besides $k$-means and GMM, for the kernels 
+\eqref{eq:rhoalpha},
+\eqref{eq:rhoesigma} and
+\eqref{eq:rhogsigma}. We sample the data $10$ times and show the average
+accuracy \eqref{eq:accuracy} with standard error.
+}
+\end{table}
+
+Next we consider 
+the well-known MNIST handwritten digit dataset 
+as illustrated in Fig.~\ref{fig:other}d.
+Each data point is an $8$-bit gray scale
+image corresponding to one of the digits $\{0,1,\dotsc,9 \}$.
+Each image forms a $784$-dimensional vector.
+We consider subsets of these classes, where we sample $100$ points 
+for each class, and cluster the data using Algorithms~\ref{algo} and
+\ref{kmeans_algo}. Besides the kernel \eqref{eq:rhoalpha}, we
+consider the gaussian kernel \eqref{eq:rhogsigma} with the width
+computed from the data as
+\begin{equation}
+\label{eq:sigma}
+\sigma^2 = \dfrac{1}{n^2} \sum_{i,j=1}^n \| x_i - x_j \|^2
+\end{equation}
+where $\{ x_i \}_{i=1}^n$ is the sampled dataset.
+The results are shown in Table~\ref{table:mnist}.
+Algorithm~\ref{algo} performed slightly better than $k$-means, except
+for the last column where all the methods are comparable. 
+Unsupervised clustering on MNIST dataset without any feature extraction
+or dimensionality reduction is not an easy task. For instance,
+the same experiment was performed in \cite{Sapiro} where a low-rank
+transformation is learned then subsequently used in subspace clustering,
+providing very accurate results. One could explore analogous methods
+for learning a better representation of the data and subsequently apply
+Algorithm~\ref{algo} for clustering.
+
+\newcolumntype{g}{>{\columncolor{gray!20}}l}
+\begin{table}
+\begin{tabular}{@{}r l | l | l | l | l@{}}
+\toprule[1pt]
+& & $\{0,1,2\}$ &
+$\{0,1,\dotsc,4\}$ &
+$\{0,1,\dotsc,6\}$ &
+$\{0,1,\dotsc,8\}$ \\
+\midrule[0.5pt]
+\multirow{3}{*}{\emph{Algorithm \ref{algo}}~~~~} 
+& $\rho_{1}$ 
+&$0.907\pm 0.007$
+&$0.866\pm 0.006$
+&$0.715\pm 0.013$
+&$0.616\pm 0.019$
+\\
+& $\rho_{1/2}$ 
+&$\bm{0.918\pm 0.006}$
+&$0.849\pm 0.025$
+&$0.711\pm 0.010$
+&$0.642\pm 0.009$
+\\
+& $\widetilde{\rho}_{\sigma}$ 
+&$0.900\pm 0.007$
+&$\bm{0.871\pm 0.005}$
+&$\bm{0.719\pm 0.010}$
+&$0.630\pm 0.016$
+\\
+\arrayrulecolor{gray!80}\midrule[0.5pt]
+\multirow{3}{*}{\emph{Algorithm \ref{kmeans_algo}}~~~~} 
+& $\rho_{1}$ 
+&$0.914\pm 0.006$
+&$0.845\pm 0.023$
+&$0.664\pm 0.022$
+&$0.614\pm 0.014$
+\\
+& $\rho_{1/2}$ 
+&$0.895\pm 0.011$
+&$0.822 \pm 0.026$
+&$0.669\pm 0.021$
+&$0.591\pm 0.019$
+\\
+& $\widetilde{\rho}_{\sigma}$ 
+&$0.896\pm 0.007$
+&$0.869\pm 0.006$
+&$0.705\pm 0.016$
+&$\bm{0.646\pm 0.020}$
+\\
+\arrayrulecolor{gray!80}\midrule[0.5pt]
+\emph{$k$-means}~~~~ &
+&$0.871\pm 0.015$
+&$0.840\pm 0.022$
+&$0.707\pm 0.012$
+&$0.634\pm 0.011$
+\\
+\arrayrulecolor{black}\bottomrule[1pt]
+\end{tabular}
+\caption{\label{table:mnist}
+Clustering the data shown in Fig.~\ref{fig:other}d with Algorithms~\ref{algo}
+and \ref{kmeans_algo}, and $k$-means.
+We use the kernel \eqref{eq:rhoalpha} with $\alpha\in\{1,2\}$, and
+the gaussian kernel 
+\eqref{eq:rhogsigma} with $\sigma$ given by \eqref{eq:sigma}. 
+For each subset of digits we sample $10$ times and show the average
+accuracy \eqref{eq:accuracy} with standard error. We pick $100$ points for 
+each class.
+}
+\end{table}
+
+%Energy standard  : 0.991 0.00221108319357
+%Energy half      : 0.993 0.000816496580928
+%Energy rbf       : 0.99 0.00235702260396
+%Kernel standard  : 0.9915 0.00183333333333
+%Kernel half      : 0.992 0.00110554159679
+%Kernel rbf       : 0.99 0.00235702260396
+%k-means          : 0.9865 0.00258736244938
+%
+%
+%%%%%%%%%%%%%%%%%%
+%Energy standard  : 0.907333333333 0.00745024649522
+%Energy half      : 0.918 0.00638671460587
+%Energy rbf       : 0.899666666667 0.00708763485947
+%Kernel standard  : 0.913666666667 0.00648740470089
+%Kernel half      : 0.895333333333 0.0113442213494
+%Kernel rbf       : 0.895666666667 0.00723588451341
+%k-means          : 0.870666666667 0.0152412695107
+%
+%
+%
+%Energy standard  : 0.82975 0.0289550657629
+%Energy half      : 0.857 0.0138232895265
+%Energy rbf       : 0.85475 0.0164335784295
+%Kernel standard  : 0.84875 0.016386181509
+%Kernel half      : 0.79025 0.036198622595
+%Kernel rbf       : 0.8545 0.0145334862377
+%k-means          : 0.83425 0.01634289686
+%
+%
+%%%%%%%%%%%%%%%%%%%%%
+%Energy standard  : 0.8662 0.00636972003571
+%Energy half      : 0.8494 0.0250138628231
+%Energy rbf       : 0.8714 0.00536697721669
+%Kernel standard  : 0.845 0.0234932614452
+%Kernel half      : 0.8224 0.0261921107715
+%Kernel rbf       : 0.8688 0.00640971484892
+%k-means          : 0.8398 0.0224151734323
+%
+%
+%Energy standard  : 0.703333333333 0.0166295883857
+%Energy half      : 0.709833333333 0.0133690493859
+%Energy rbf       : 0.702 0.0158149015524
+%Kernel standard  : 0.688833333333 0.0195426883173
+%Kernel half      : 0.708833333333 0.0222403398366
+%Kernel rbf       : 0.7175 0.0153744416741
+%k-means          : 0.693333333333 0.0132590523468
+%
+%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%Energy standard  : 0.714714285714 0.0132292707827
+%Energy half      : 0.711142857143 0.00970728038727
+%Energy rbf       : 0.719142857143 0.010446277641
+%Kernel standard  : 0.663857142857 0.0220872830507
+%Kernel half      : 0.668571428571 0.0209923860509
+%Kernel rbf       : 0.705 0.0157981169405
+%k-means          : 0.707428571429 0.0115799498313
+%
+%Energy standard  : 0.671625 0.0167819450766
+%Energy half      : 0.652875 0.0162233221183
+%Energy rbf       : 0.686375 0.0142173724913
+%Kernel standard  : 0.674875 0.012649728258
+%Kernel half      : 0.654375 0.0232537706023
+%Kernel rbf       : 0.706 0.0188513925215
+%k-means          : 0.6705 0.0121037184369
+%
+%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%Energy standard  : 0.615555555556 0.0190076152525
+%Energy half      : 0.641888888889 0.00939463992175
+%Energy rbf       : 0.630333333333 0.0158222725065
+%Kernel standard  : 0.614444444444 0.0136113000743
+%Kernel half      : 0.590777777778 0.0188685802226
+%Kernel rbf       : 0.645888888889 0.0202454418315
+%k-means          : 0.634222222222 0.0111083947297
+%
+%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%Energy standard  : 0.5369 0.0136100044902
+%Energy half      : 0.5395 0.0138782403624
+%Energy rbf       : 0.5276 0.00942125257065
+%Kernel standard  : 0.518 0.0162446572413
+%Kernel half      : 0.5193 0.0146674620996
+%Kernel rbf       : 0.5289 0.0144001928999
+%k-means          : 0.5066 0.0120103658932
 
@@
 
-In this paper we have considered clustering from the perspective of energy
-statistics, which provides a nonparametric test for equality of distributions.
-Based on this, we showed that the clustering problem reduces to a quadratically
+We considered clustering from the perspective of energy
+statistics theory which provides a nonparametric test for 
+equality of distributions.
+We showed that the clustering problem reduces to a quadratically
 constrained optimization problem (QCQP), 
 as described in Proposition~\ref{th:qcqp2}.
-Moreover, we showed that clustering based on energy statistics is equivalent
+Moreover, this problem is equivalent
 to kernel $k$-means optimization problem, once the kernel is fixed; see
 Proposition~\ref{th:kernel_kmeans}. Our results imply that kernel $k$-means
-approach to clustering is actually a consequence of energy statistics
-theory, and thus place this method into a principled statistical basis.
-As already known \cite{Dhillon}, this approach is related to spectral
-clustering, and graph partitioning problems. Therefore, all these problems 
-may be seen as arising naturally from energy statistics clustering.
-It is important to mention that energy statistics clustering, as formulated
-here, is valid for
-arbitrary metric spaces of negative type, and makes no assumptions about
-the distribution of the data. Moreover, it does not rely on the concept of a
-cluster mean, even implicitly.
-
-We also proposed Algorithm~\ref{algo} as an alternative to the well-known
-kernel $k$-means algorithm (see Algorithm~\ref{kmeans_algo}), 
-where both have the same time
-complexity. The numerical results show that Algorithm~\ref{algo}
-might provide better clustering accuracy and is more robust 
-than kernel $k$-means algorithm. Since there exists a huge literature about
-kernel $k$-means, and approximation methods to make it faster, with
-applications to several artificial and real data, 
-we limited ourselves to analyze few but carefully designed experiments, which
-illustrates the advantages of Algorithm~\ref{algo}.
+is actually a consequence of energy statistics, and thus place this 
+method into a principled statistical basis.
+We also considered a weighted version of energy statistics clustering
+which establishes connection with spectral
+clustering and graph partitioning problems.
+
+We considered Algorithm~\ref{algo} based on Hartigan's method, and
+compared with 
+kernel $k$-means described in Algorithm~\ref{kmeans_algo} which
+is based on Lloyd's heuristic.
+Both have the same time complexity, however, the numerical 
+results provide evidence that Algorithm~\ref{algo} is more robust with
+a superior clustering performance. This is also theoretically
+supported as described in the end of Section~\ref{sec:algo}.
+
+In the same way that kernel $k$-means can be accelerated by exploring
+sparsity or sampling methods to approximate the Gram matrix, these
+techniques are also applicable to Algorithm~\ref{algo}.
+For instance, it would be interesting explore the methods recently proposed
+in \cite{Mahoney} to this case. 
+Another fruitful avenue of exploration would be to find 
+better methods to tackle the optimization problem \eqref{eq:qcqp2}.
 
@@
 
-%\subsection*{Acknowledgements}
-%We thank \ldots
+\subsection*{Acknowledgements}
+GF would like to thank Carey Priebe for discussions.
+This work was supported by XXX grant.
 

### 2017-08-11 qcqp_clustering/code/eclust/bayes.py

@@ -1,0 +1,186 @@
+"""Functions to compute Bayes errors."""
+
+# Guilherme S. Franca <guifranca@gmail.com>
+# Johns Hopkins, Neurodata
+
+
+from __future__ import division
+
+from functools import partial
+from scipy.optimize import minimize
+from scipy.optimize import root
+import numpy as np
+import scipy.stats as stats
+import scipy.integrate as integrate
+import matplotlib.pyplot as plt
+
+import data
+
+
+def bayes_univariate_normal(mu1, mu2, sigma1, sigma2, 
+                            num_times=10, num_points=1000):
+    """Estimate Bayes' error for two-class normal distributions in 1D."""
+    error = np.zeros(num_times)
+    for i in range(num_times):
+        x1 = np.random.normal(mu1, sigma1, num_points)
+        x2 = np.random.normal(mu2, sigma2, num_points)
+    
+        error1 = sum([1 for x in x1 
+            if stats.norm.pdf(x,mu1,sigma1)-stats.norm.pdf(x,mu2,sigma2)<=0
+        ])
+        error2 = sum([1 for x in x2 
+            if stats.norm.pdf(x,mu1,sigma1)-stats.norm.pdf(x,mu2,sigma2)>=0
+        ])
+        error[i] = error1+error2
+    return 1-error.mean()/(2*num_points)
+
+def integrate_univariate_normal(mu1, mu2, sigma1, sigma2):
+    """Same as above but through numerical integration."""
+    
+    g1 = partial(stats.norm.pdf, loc=mu1, scale=sigma1)
+    g2 = partial(stats.norm.pdf, loc=mu2, scale=sigma2)
+    
+    # make it visual to confirm
+    #xs = np.arange(-10,20, 0.001)
+    #y1s = np.array([g1(x) for x in xs])
+    #y2s = np.array([g2(x) for x in xs])
+    #y3s = np.array([np.abs(g1(x)-g2(x)) for x in xs])
+    #plt.plot(xs, y1s)
+    #plt.plot(xs, y2s)
+    #plt.plot(xs, y3s)
+    #plt.show()
+
+    x0 = root(lambda x: g1(x) - g2(x), 1).x[0]
+    
+    # another way
+    #cons = (
+    #    {'type': 'ineq', 'fun': lambda x: x-1},
+    #    {'type': 'ineq', 'fun': lambda x: -(x-2)},
+    #)
[…]

### 2017-08-11 qcqp_clustering/code/eclust/data.py

@@ -117,7 +117,7 @@ def circles(rs, ms, eps, ns):
     X = []
     z = []
     for r, m, e, n in zip(rs, ms, eps, ns):
-        x = np.array([[r*np.cos(a)+m[0], r*np.sin(a)+m[1]]
+        x = np.array([(r*np.cos(a)+m[0], r*np.sin(a)+m[1])
                         for a in np.random.uniform(0, 2*np.pi, n)])
         x += e*np.random.multivariate_normal([0,0],np.eye(2), n)
         X.append(x)
@@ -153,17 +153,10 @@ def spirals(rs, ms, ns, noise=0.2):
     idx = np.random.permutation(sum(ns))
     return X[idx], z[idx]
 
-def plot(X, z, colors=['b', 'r', 'g'], fname='plot.pdf'):
+def plot(X, z, fname='plot.pdf'):
     """Plot data according to labels in z.
     Use PCA to project in 2D in case data is higher dimensional.
 
-    Input: X = data matrix
-           z = labels of each points
-           colors = list of colors for each class
-           fname = output file
-    
-    Ouput: None
-    
     """
 
     z_unique = np.unique(z)
@@ -176,13 +169,14 @@ def plot(X, z, colors=['b', 'r', 'g'], fname='plot.pdf'):
 
     fig = plt.figure()
     ax = fig.add_subplot(111)
-    colors = iter(colors)
+
+    colors = iter(plt.cm.brg(np.linspace(0,1,5)))
     
     for k in z_unique:
         idx = np.where(z==k)
         x = X_new[idx][:,0]
         y = X_new[idx][:,1]
-        ax.plot(x, y, 'bo', alpha=.8, color=next(colors))
+        ax.plot(x, y, 'bo', markersize=4, alpha=.7, color=next(colors))
     
     ax.set_xticks([])
     ax.set_yticks([])
@@ -211,10 +205,52 @@ def histogram(X, z, colors=['#1F77B4', '#FF7F0E'], fname='plot.pdf',
         x = X[idx]
         ax.hist(x, bins=bins, facecolor=next(colors), 
                 histtype='stepfilled',
-                alpha=.8, normed=1, linewidth=0.0)
+                alpha=.8, normed=1, linewidth=0.5)
     ax.set_xlabel(r'$x$')
     if xlim:
[…]

### 2017-08-11 qcqp_clustering/code/eclust/energy1d.py

@@ -91,6 +91,7 @@ if __name__ == "__main__":
     from sklearn.cluster import KMeans
     from sklearn.mixture import GMM
     from scipy.stats import sem
+    import sys
     
     import data
     from metric import accuracy
@@ -101,12 +102,30 @@ if __name__ == "__main__":
     #X1 = np.random.normal(0, 1, 100)
     #X1 = np.random.gamma(2, 2, 1000)
     #X1 = np.random.uniform(0, 1, 1000)
+    X1 = np.random.lognormal(0, 0.3, 1000)
+    X2 = np.random.lognormal(-1, 1, 1000)
     
     #X2 = np.random.uniform(0.5, 1.5, 100)
     #X2 = np.random.normal(2, 1, 100)
     #X2 = np.random.gamma(2, 2, 2000)
     
-    #X, z = data.shuffle_data([X1, X2])
+    X, z = data.shuffle_data([X1, X2])
+    Y = np.array([[x] for x in X])
+    
+    z0 = kpp(2, Y, ret='labels')
+    Z0 = ke.ztoZ(z0)
+
+    zh, cost = two_clusters1D(X)
+    print accuracy(z, zh)
+        
+    km = KMeans(2)
+    zh = km.fit_predict(Y)
+    print accuracy(z, zh)
+        
+    zh = gmm_em(2, Y)
+    print accuracy(z, zh)
+
+    sys.exit()
 
     #X, z = data.univariate_lognormal([0, -1.5], [0.25, 1], [4000, 4000])
     #X, z = data.univariate_lognormal([0, -1.5], [0.3, 1.5], [1000, 1000])

### 2017-08-11 qcqp_clustering/code/number_clusters.py

@@ -76,16 +76,26 @@ def detect_clusters(num_points, num_permutations):
     """Check if algorithms are detecting clusters, compared to chance."""
 
     # generate data from uniform
-    Y = np.random.uniform(0, 10, num_points)
-    X = np.array([[x] for x in Y])
+    Y1 = np.random.uniform(0, 1, num_points)
+    Y2 = np.random.uniform(2, 3, num_points)
+    X, z = data.shuffle_data([Y1, Y2])
+    X = np.array([[x] for x in X])
     
     # cluster with k=2 and pick objectives values
     rho = lambda x, y: np.linalg.norm(x-y)
     G = ke.kernel_matrix(X, rho)
-    k=2
+    k=3
     z_energy, J_energy = energy(k, X, G, run_times=5)
     z_kmeans, J_kmeans = kmeans(k, X, run_times=5)
     z_gmm, J_gmm = gmm(k, X, run_times=5)
+    #print J_energy, J_kmeans, J_gmm
+    # 
+    #k=4
+    #z_energy, J_energy = energy(k, X, G, run_times=5)
+    #z_kmeans, J_kmeans = kmeans(k, X, run_times=5)
+    #z_gmm, J_gmm = gmm(k, X, run_times=5)
+    #
+    #print J_energy, J_kmeans, J_gmm
 
     # random permute labels and compute objectives
     times_energy = 0
@@ -93,13 +103,15 @@ def detect_clusters(num_points, num_permutations):
     times_gmm = 0
     for i in range(num_permutations):
 
-        #fake_z = np.random.randint(0, 2, num_points)
-        #fake_Z = ke.ztoZ(fake_z)
-        fake_z_energy = np.random.choice(z_energy, len(z_energy))
+        fake_z_energy = np.random.randint(0, 6, 2*num_points)
+        fake_z_kmeans = np.random.randint(0, 6, 2*num_points)
+        fake_z_gmm = np.random.randint(0, 6, 2*num_points)
+        
+        #fake_z_energy = np.random.choice(z_energy, len(z_energy), replace=False)
         fake_Z_energy = ke.ztoZ(fake_z_energy)
 
-        fake_z_kmeans = np.random.choice(z_kmeans, len(z_kmeans))
-        fake_z_gmm = np.random.choice(z_gmm, len(z_gmm))
+        #fake_z_kmeans = np.random.choice(z_kmeans, len(z_kmeans), replace=False)
[…]

### 2017-08-11 qcqp_clustering/code/print_digits.py

@@ -1,0 +1,40 @@
+import matplotlib.pyplot as plt
+import cPickle, gzip
+import matplotlib.pyplot as plt
+import numpy as np
+
+from pplotcust import *
+
+
+f = gzip.open('data/mnist.pkl.gz', 'rb')
+train_set, valid_set, test_set = cPickle.load(f)
+f.close()
+images, labels = train_set
+
+X = []
+for d in range(0,10):
+    x = np.where(labels==d)[0]
+    js = np.random.choice(x, 2, replace=False)
+    for j in js:
+        im = np.array(images[j])
+        X.append(im)
+X = np.array(X)
+idx = range(20)
+np.random.shuffle(idx)
+X = X[idx]
+fig, axes = plt.subplots(nrows=5, ncols=4,
+                        #figsize=(3.85,4.69))
+                        figsize=(4,5))
+i = 0
+for row in axes:
+    for ax in row:
+        ax.imshow(X[i].reshape((28,28)), cmap=plt.cm.gray)
+        ax.axis('off')
+        ax.set_aspect('equal')
+        #ax.set_xticklabels([])
+        #ax.set_yticklabels([])
+        i += 1
+fig.subplots_adjust(wspace=-0.1, hspace=-0.1)
+#fig.tight_layout()
+fig.savefig('mnist.pdf')
+

### 2017-08-11 qcqp_clustering/code/tests2.py

@@ -13,7 +13,9 @@ from __future__ import division
 
 import numpy as np
 import scipy.stats
+import matplotlib as mpl
 import matplotlib.pyplot as plt
+import cPickle, gzip
 
 import eclust.data as data
 import eclust.kmeans as km
@@ -115,6 +117,79 @@ def two_clusters1D(X, z):
 ###############################################################################
 # functions to test different settings
 
+def gauss_dimensions_mean(dimensions=range(2,100,20), num_points=[200, 200],
+                          delta=0.7, run_times=5, num_experiments=10, d=None):
+    """Here we keep signal in one dimension and increase the ambient dimension.
+    The covariances are kept fixed.
+    
+    """
+    k = 2
+    if not d:
+        d = dimensions[0]
+    n1, n2 = num_points
+    table = np.zeros((num_experiments*len(dimensions), 5))
+    count = 0
+    
+    for D in dimensions:
+
+        for i in range(num_experiments):
+        
+            # generate data
+            m1 = np.zeros(D)
+            s1 = np.eye(D)
+            m2 = np.concatenate((delta*np.ones(d), np.zeros(D-d)))
+            s2 = np.eye(D)
+        
+            X, z = data.multivariate_normal([m1, m2], [s1, s2], [n1, n2])
+            rho = lambda x, y: np.linalg.norm(x-y)
+            G = ke.kernel_matrix(X, rho)
+        
+            table[count, 0] = D
+            table[count, 1] = cost_energy(k, X, G, z, run_times=run_times)
+            table[count, 2] = kernel_energy(k, X, G, z, run_times=run_times)
+            table[count, 3] = kmeans(k, X, z, run_times=run_times)
+            table[count, 4] = gmm(k, X, z, run_times=run_times)
+
+            count += 1
+
+    return table
+
+def gauss_dimensions_cov(dimensions=range(2,100,20), num_points=[200, 200],
+                         run_times=3, num_experiments=10, d=None):
+    """High dimensions but with nontrivial covariance."""
+    k = 2
+    if not d:
+        d = dimensions[0]
+    n1, n2 = num_points
+    table = np.zeros((num_experiments*len(dimensions), 5))
+    count = 0
+    
+    for D in dimensions:
+
+        for l in range(num_experiments):
+        
+            m1 = np.zeros(D)
+            s1 = np.eye(D)
[…]

### 2017-08-11 qcqp_clustering/draft/energy_clustering.tex
finished draft
corrected typos, included reference

@@
 %\usepackage{printlen}
-%\usepackage[inline]{enumitem}
+\usepackage[inline]{enumitem}
 %\usepackage{cite}
@@
 Moreover, we also consider a weighted version of energy statistics 
-applied to clustering, which is related to 
+applied to clustering, which makes connections to 
 graph partitioning problems.
@@
 Since the statistical potential energy is minimum when
-distributions are equal, the principle 
+distributions are equal, the
 principle behind clustering is to maximize the statistical energy, 
@@
 Moreover, in our experiments we put in
-in evidence the nonparametric aspect of energy
+evidence the nonparametric aspect of energy
 statistics based clustering, which provides a default kernel, showing that
@@
 Assume we have data $\mathbb{X} = \{ x_1,\dotsc, x_n \}$ where
-$x_i \in \mathcal{X}$ and $\mathcal{X}$ is a space of negative type.
+$x_i \in \mathcal{X}$, and $\mathcal{X}$ is a space of negative type.
 Consider a disjoint partition $\mathbb{X} = \bigcup_{j=1}^k \C_j$, with
@@
 criterion for clustering data is to 
-maximize $S$, which makes 
+maximize $S$ which makes 
 each cluster as different
@@
 finding the best partition of the data by solving~\eqref{eq:minimize}.
-Notice that this is a hard clustering problem, as partitions
+Notice that this is a hard clustering problem as partitions
 are disjoint. The problem \eqref{eq:minimize} based on
@@
 the $i$th row of a matrix $M$, and $M_{\bullet j}$ denotes its $j$th column.
-Our next result shows the optimization problem \eqref{eq:minimize}
+Our next result shows that the optimization problem \eqref{eq:minimize}
 is NP-hard since
@@
 $\mathbb{X} = \{ x_1,\dotsc,x_n \}$
-into  $k$ clusters, we first compute the Gram matrix
+into  $k$ clusters we first compute the Gram matrix
 $G$ and then 
@@
 This is the procedure done in spectral clustering on a graph Laplacian
-matrix defined by a similarity matrix.
-However, computing eigenvectors of a very large matrix
-can be very expensive, and usually iterative methods are preferred. Moreover,
+matrix obtained through a similarity matrix.
+However, computing eigenvectors of very large matrices
+can be prohibitively expensive 
+and usually iterative methods are preferred. Moreover,
 this procedure is only guaranteed to converge if the normalized $Y^\star$ is
@@
 The clustering problem \eqref{eq:qcqp2} based on energy statistics
-is valid for data living in an \emph{arbitrary} space of negative type, where
+is valid for data living in an \emph{arbitrary} space of negative type where
 a semimetric $\rho$, and thus the kernel \eqref{eq:kernel_semimetric}, are
@@
 If one uses the standard energy distance
-defined in \eqref{eq:energy}, 
-then $\rho(x,y) = \| x - y\|$ and this fixes the kernel
+defined in \eqref{eq:energy} 
+we have $\rho(x,y) = \| x - y\|$ and this fixes the kernel
 through \eqref{eq:kernel_semimetric}. In the same way, 
@@
 The energy distance \eqref{eq:energy2} fixes a family of choices
-$\rho(x,y) = \| x-y\|^\alpha$ with $0 < \alpha\le 2$. If prior information
+$\rho(x,y) = \| x-y\|^\alpha$ with $0 < \alpha\le 2$. 
+Nevertheless, if prior information
 is available one may choose a more appropriate $\rho$.
@@
 problem,
-in the feature space,
+in feature space,
 is defined by
@@
 \sum_{x \in \C_j} \| \varphi(x) - \varphi(\mu_j) \|^2
-\bigg\},
+\bigg\}
 \end{equation}
@@
 therefore $H^\top H = I$. This is a constraint on the rows of $H$.
-To obtain a condition on the columns
-of $H$ observe that
+To obtain a condition on its columns
+observe that
 \begin{equation}
@@
 if both points $x_p$ and $x_q$ belong to the same cluster, which
-we denote by $\C_i$ for some $i=1,\dotsc,k$, and 
+we denote by $\C_i$ for some $i\in\{1,\dotsc,k\}$, and 
 $(H^\top H \mathcal{W}^{1/2})_{pq} = 0 $ otherwise. Thus, the $p$th
@@
 The relation between kernel $k$-means and graph partitioning problems
-was already established \cite{Dhillon2,Dhillon}. We repeat a similar analysis
-due to the relation of these problems to
+is known \cite{Dhillon2,Dhillon}. For conciseness, we repeat a similar 
+analysis due to the relation of these problems to
 energy statistics and RKHS, which may provide a different perspective.
@@
 in analogous way as minimizing \eqref{eq:within} is equivalent to
-maximizing \eqref{eq:between}, as shown in Proposition~\ref{th:minimize}.
+maximizing \eqref{eq:between} as shown in Proposition~\ref{th:minimize}.
 Here this is due to the equality
@@
 allows one to use energy statistics theory for inference
-over populations of graphs.
+on graphs.
 
@@
 
-Before stating a general algorithm to solve \eqref{eq:qcqp2}, 
+Before stating a general algorithm to solve \eqref{eq:qcqp2} 
 let us first consider the simplest possible case which
@@
 Fixing
-$\rho(x,y) = |x - y|$ according to \eqref{eq:energy}, we can actually compute 
+$\rho(x,y) = |x - y|$, according to \eqref{eq:energy}, we can actually compute 
 \eqref{eq:g_def} in $\OO(n \log n)$ and find
@@
 is at the most $\OO(n\log n)$.
-Assuming that each partition is ordered  $\mathbb{X} = \bigcup_{j=1}^k
-\bar{\C}_j$ 
+Assuming that each partition is ordered,  
+$\mathbb{X} = \bigcup_{j=1}^k \bar{\C}_j$,
 the within energy dispersion
@@
 
-For a two-class problem we can use \eqref{eq:w1d} to cluster data
+For a two-class problem we can use \eqref{eq:w1d} to cluster the data
 through a simple algorithm 
 as follows. We first order
-the entire dataset $\mathbb{X} \to \bar{\mathbb{X}}$. Then 
+the entire dataset, $\mathbb{X} \to \bar{\mathbb{X}}$. Then 
 we compute \eqref{eq:w1d} for each possible split of $\bar{\mathbb{X}}$
@@
 We have the same number of points in both clusters, and for each case
-we sample $100$ times from the distributions in the histograms. We 
+we sample $100$ times from the distributions shown in the histograms. We 
 plot the average value of 
@@
 few times choosing the result with  best objective function value. 
-Notice that Algorithm~\ref{algo1d} requires no initialization.
+Notice that Algorithm~\ref{algo1d} requires no initialization so we only
+run it once.
 Moreover, for each case we sample $100$ times
@@
 In  Fig.~\ref{fig:1d}a 
-we sample data from two normal distributions, with equal number of
-points in each cluster:
+we have data sampled from two normal distributions with equal number of
+points in each cluster,
 \begin{equation}
@@
 \end{equation}
-The optimal accuracy obtained from Bayes classification error
-is $ \approx 0.956$, indicated by the dashed line in the plot.
+For these distributions the optimal classification accuracy 
+obtained from Bayes error
+is $ \approx 0.956$ which is indicated by the dashed line in the plot.
 We see that the three methods
-perform closely, with a slight advantage of GMM as expected since
-it is the true model for the data. Energy statistics performs slightly better
-than $k$-means. In Fig~\ref{fig:1d}b
-we consider two clusters with lognormal distributions:
+perform closely. As expected, GMM has a slight advantage over the other
+methods since
+it corresponds to the true model of the data. 
+Energy statistics performs slightly better
+than $k$-means. On the other hand, in Fig~\ref{fig:1d}b
+we consider two clusters with lognormal distributions,
 \begin{equation}
@@
 \end{equation}
-The optimal accuracy is $\approx 0.852$.
+The optimal classification accuracy from Bayes error is $\approx 0.852$.
 In this case,
@@
 thus giving zero accuracy.
-These two simple experiments illustrate
-how energy statistics based clustering is nonparametric, being able
+These two simple experiments from Fig.\ref{fig:1d} illustrate
+how energy statistics clustering is nonparametric, being able
 to provide high quality clustering in settings where data comes
@@
 where $Q_j$ represents an internal energy cost of cluster $\C_j$, and
-$Q$ is the total energy cost where each individual cluster cost 
+$Q$ is the total energy cost where each $Q_j$ 
 is weighted by the inverse
@@
 convergence, i.e. until no new assignments are made.
-The entire procedure is in Algorithm~\ref{kmeans_algo}.
+The entire procedure is described in Algorithm~\ref{kmeans_algo}.
 It can be shown that this algorithm converges provided $G$ is positive
@@
 
-\subsection*{Hartigan's Algorithm for Energy Statistics Clustering}
-
-We now consider Hartigan's algorithm \cite{Hartigan} applied to 
-\eqref{eq:maxQ}. This gives approximate solutions to 
+\subsection*{Hartigan's Method for Energy Statistics Clustering}
+
+We now consider an algorithm based on Hartigan's method \cite{Hartigan} 
+applied to \eqref{eq:maxQ}. This gives a local and approximate solution to 
 \eqref{eq:qcqp2}. This approach was also proposed in \cite{Kgroups}.
@@
 a total energy cost function \eqref{eq:maxQ} denoted by $Q^{(j)}$.
-Let us consider the change in the total energy by moving
+Let us consider the change in the total within energy by moving
 $x_i$ to $\C_\ell$. 
 Denote this new energy cost by $Q^{(\ell)}$.
-It is straightforward to see that
+A straightforward computation gives
 \begin{equation}
@@
 maximum of \eqref{eq:maxQ} by
-moving $x_i$ to $\C_\ell$, otherwise we better keep $x_i$ in $\C_j$. Based on
+moving $x_i$ to $\C_\ell$, otherwise we keep $x_i$ in $\C_j$. Based on
 this the algorithm goes as follows.
@@
 point $x_i$ 
-we compute the cost of moving it to another cluster,
+we compute the cost of moving it to another cluster $\C_\ell$,
 $\Delta Q^{j\to \ell}(x_i)$ for 
-$\ell=1,\dots,k$ with $\ell \ne j$. 
-We then choose $j^\star = \argmax_\ell \Delta^{j \to \ell}(x_i)$.
+$\ell=1,\dots,k$ with $\ell \ne j$. Hence, let us choose
+$j^\star = \argmax_\ell \Delta^{j \to \ell}(x_i)$.
 If $\Delta Q^{j \to j^\star}(x_i) > 0$ 
@@
 Note that this ensures that the objective function is
-monotonically increasing at each iteration, and consequently the algorithm
+monotonically increasing at each iteration and consequently the algorithm
 converges in a finite number of steps.
@@
 
-Some important properties of Hartigan's method 
-are worth mentioning.
+Let us mention some important results about Hartigan's method.
 
@@
 by Lloyd's method, and consequently by (kernel) $k$-means algorithm.
+The next result indicates that Hartigan's has the potential to
+escape local optima of Lloyd's.
 
@@
 \label{eq:standard_metric}
-\rho(x,y) = \| x-y\|, \qquad \kk(x,y) = 
+\rho(x,y) = \| x-y\| \qquad \mbox{and} \qquad \kk(x,y) = 
 \tfrac{1}{2}\left( \| x \| + \| y \| - \| x-y \| \right).
@@
 
-Let us briefly mention that we compared Algorithms~\ref{algo} and
-\ref{algo1d} for several univariate distributions, and
+Let us briefly mention that we compared Algorithm~\ref{algo} and
+Algorithm~\ref{algo1d} for several univariate distributions, and
 both perform almost indistinguishable regarding the clustering quality.
-We omit these results since we will analyse more interesting scenarios.
+We omit these results since we will analyse more interesting scenarios
+in higher dimensions and with more than two clusters.
 
@@
 Lloyd's method. From the discussion in the previous section we expect 
-Algorithm~\ref{algo} to be superior than Algorithm~\ref{kmeans_algo}.
+that Algorithm~\ref{algo}  be superior than Algorithm~\ref{kmeans_algo}.
 Another goal is to illustrate the nonparametric aspect of energy statistics.
@@
 indicating standard error. Whenever possible, for reference, 
-we also show the best possible accuracy obtained from a numerical estimation
+we also show the best possible classification 
+accuracy obtained from a numerical estimation
 of Bayes classification error, which uses the true probability distribution
@@
 For each $D$ we sample $100$ times, picking $100$ points for each cluster.
-We apply each algorithm to the resulting data set 
+We apply each algorithm to the resulting dataset 
 and compute the average value of the accuracy \eqref{eq:accuracy}.
-For both Algorithms~\ref{algo} and \ref{kmeans_algo} we use the standard
+For both Algorithm~\ref{algo} and Algorithm~\ref{kmeans_algo} 
+we use the standard
 kernel \eqref{eq:standard_metric} from energy statistics.
-The results are shown in Fig.~\ref{fig:gauss}a, where error bars indicate
-the standard error.
+The results are shown in Fig.~\ref{fig:gauss}a.
 Notice that GMM/EM algorithm is unable to estimate the covariance matrix 
@@
 parametric, they are consistent models for the above distributions.
-Now we consider the same experiment but with different covariances:
+Now we consider the same experiment but with different covariances,
 \begin{equation}
@@
 $\approx 0.95$.
-The results are shown In Figure~\ref{fig:gauss}b.
+The results are shown in Figure~\ref{fig:gauss}b.
+
 In the experiments of Figure~\ref{fig:gauss}
@@
 \begin{figure}
-\centering
-\includegraphics[width=0.5\textwidth]{gauss_pi.pdf}
+\includegraphics[width=0.5\textwidth]{gauss_pi.pdf}\vspace{-1em}
 \caption{
@@
 \end{equation}
-We sample an equal number of points for each cluster, which is
+We sample an equal number of points for each cluster, which is progressively
 increased. The optimal accuracy of \eqref{eq:20gauss} based on Bayes
@@
 before but now using also the kernels \eqref{eq:rhohalf} and
-\eqref{eq:rhoe}. The results are shown in Figure~\ref{fig:consist}a.
+\eqref{eq:rhoe}. The results are shown in Fig.~\ref{fig:consist}a.
 Algorithm~\ref{algo} performs better than all the other ones, and in 
@@
 better than Algorithm~\ref{kmeans_algo} for any of the kernel choices.
-Standard $k$-means performs poorly on this example.
-
-Consider the same parameters as in \eqref{eq:20gauss} but now with
-lognormal distribution, i.e.
-\begin{equation}
-\label{eq:20loggauss}
-\log x \in \C_i \sim \mathcal{N}(\mu_i, \Sigma_i) \qquad (i=1,2).
-\end{equation}
-The same previous experiment is now shown in 
-Fig.~\ref{fig:consist}b.
-We see that Algorithm~\ref{algo} still performs accurately, 
-while $k$-means works a little bit better than chance,
-and GMM is not even able to estimate the parameters. Again, the
-kernel \eqref{eq:rhoe}
-provides better results than \eqref{eq:standard_metric} or
-\eqref{eq:rhohalf}. Algorithm~\ref{algo} performs better
-than Algorithm~\ref{kmeans_algo} for every kernel choice. 
-The experiments in Fig.~\ref{fig:consist}
-illustrates that energy statistics clustering is nonparametric.
+Standard $k$-means performs poorly in this example.
 
@@
 \label{fig:consist}
-Comparison of Algorithms~\ref{algo} and \ref{kmeans_algo} with kernels
-\eqref{eq:standard_metric}, \eqref{eq:rhohalf} and \eqref{eq:rhoe}, and
-also with $k$-means and GMM/EM. The optimal accuracy in both cases
-is $\approx 0.9$. We show average of \eqref{eq:accuracy}
+Algorithm~\ref{algo} and Algorithm~\ref{kmeans_algo} with kernels
+\eqref{eq:standard_metric}, \eqref{eq:rhohalf} and \eqref{eq:rhoe}, 
+$k$-means and GMM. The optimal accuracy in both cases
+is $\approx 0.9$. We show the average of \eqref{eq:accuracy}
 over $100$ samples with standard error.
@@
 \eqref{eq:20gauss}. 
-(b) Data distributed as in \eqref{eq:20loggauss}.
-%One can see the better performance of energy statistics
-%clustering, Algorithm~\ref{algo}, and in particular by using the semimetric
-%\eqref{eq:rhoe}. 
-These experiments illustrates how energy statistics
-based clustering is nonparametric.
+(b) Data as in \eqref{eq:20loggauss}.
 }
@@
 
+Consider the same parameters as in \eqref{eq:20gauss} but now with
+lognormal distributions, 
+\begin{equation}
+\label{eq:20loggauss}
+\log x \in \C_i \sim \mathcal{N}(\mu_i, \Sigma_i) \qquad (i=1,2).
+\end{equation}
+The same previous experiment is shown in 
+Fig.~\ref{fig:consist}b.
+Note that Algorithm~\ref{algo} still performs accurately, 
+while $k$-means works almost at chance,
+and GMM is not even able to estimate the parameters. Again, the
+kernel \eqref{eq:rhoe}
+provides better results than \eqref{eq:standard_metric} or
+\eqref{eq:rhohalf}. Algorithm~\ref{algo} performs better
+than Algorithm~\ref{kmeans_algo} for every kernel choice. 
+The experiments in Fig.~\ref{fig:consist}
+illustrates that energy statistics clustering is nonparametric.
+
 \begin{figure}
-\begin{minipage}{0.25\textwidth}
+\begin{minipage}{0.24\textwidth}
 \includegraphics[width=1\textwidth]{2cigars.pdf}\\[-1em] (a)
 \end{minipage}
-\begin{minipage}{0.25\textwidth}
+\begin{minipage}{0.24\textwidth}
 \includegraphics[width=1\textwidth]{2circles.pdf}\\[-1em] (b)
 \end{minipage}
-\begin{minipage}{0.25\textwidth}
+\begin{minipage}{0.24\textwidth}
 \includegraphics[width=1\textwidth]{3circles.pdf}\\[-1em] (c)  
 \end{minipage}
-\begin{minipage}{0.22\textwidth}
-\vspace{-0.3em}
+\begin{minipage}{0.23\textwidth}
+\vspace{-0.9em}
 \includegraphics[width=1\textwidth]{mnist.pdf}\\[-1.8em] (d)  
@@
 \caption{\label{fig:other}
-(a) Parallel cigars, $200$ points in each class. (b) Two  
-concentric
-circles with noise, $400$ points each. (c) Three
-concentric circles, $400$ points each.
-Clustering for (a--c) is shown in Table~\ref{table:other}.
-(d) MNIST handwritten digits. Clustering on Table~\ref{table:mnist}.
+(a) Parallel cigars. (b) Two  
+concentric circles with noise. (c) Three
+concentric circles. (d) MNIST handwritten digits.
+Clustering results are in Table~\ref{table:other}
+and Table~\ref{table:mnist}.
 }
@@
 
-Now we consider two dimensional data as illustrated in
-Fig~\ref{fig:other}(a--c)
-with the following choices of semimetric and corresponding generating kernel:
+Now consider the following choices of semimetric and 
+corresponding generating kernel:
 \begin{align}
@@
 \end{align}
-We apply both Algorithms~\ref{algo} and \ref{kmeans_algo}, besides
+In Fig.\ref{fig:other} we have examples of
+complex two dimensional datasets. For each class in Fig.\ref{fig:other}a
+we sample $200$ points. In each class of 
+Fig.\ref{fig:other}b and Fig.\ref{fig:other}c we sample $400$ points.
+We apply Algorithm~\ref{algo} and Algorithm~\ref{kmeans_algo}, besides
 $k$-means and GMM, using the above kernels. The results are shown
@@
 \caption{\label{table:other}
-Clustering the data shown in Fig.~\ref{fig:other} with Algorithms~\ref{algo}
-and \ref{kmeans_algo}, besides $k$-means and GMM, for the kernels 
+Clustering the data shown in Fig.~\ref{fig:other} with Algorithm~\ref{algo}
+and Algorithm~\ref{kmeans_algo}, besides $k$-means and GMM, for the kernels 
 \eqref{eq:rhoalpha},
@@
 Each data point is an $8$-bit gray scale
-image corresponding to one of the digits $\{0,1,\dotsc,9 \}$.
-Each image forms a $784$-dimensional vector.
-We consider subsets of these classes, where we sample $100$ points 
-for each class, and cluster the data using Algorithms~\ref{algo} and
-\ref{kmeans_algo}. Besides the kernel \eqref{eq:rhoalpha}, we
+image, forming a $784$-dimensional vector, 
+corresponding to the digits $\{0,1,\dotsc,9 \}$.
+Besides the kernel \eqref{eq:rhoalpha}, we
 consider the gaussian kernel \eqref{eq:rhogsigma} with the width
-computed from the data as
+given by
 \begin{equation}
@@
 \end{equation}
-where $\{ x_i \}_{i=1}^n$ is the sampled dataset.
-The results are shown in Table~\ref{table:mnist}.
+which is computed from the sampled data $\{ x_i \}_{i=1}^n$.
+We consider subsets of the  classes $\{0,1,\dotsc,9 \}$, 
+where we sample $100$ points 
+for each class. We cluster the resulting dataset 
+using Algorithm~\ref{algo},
+Algorithm~\ref{kmeans_algo}, and $k$-means.
+The results are shown in Table~\ref{table:mnist}, where the kernel
+and its parameter for each case is indicated.
 Algorithm~\ref{algo} performed slightly better than $k$-means, except
@@
 \begin{table}
-\begin{tabular}{@{}r l | l | l | l | l@{}}
+\begin{tabular}{@{}rl|l|l|l|l@{}}
 \toprule[1pt]
-& & $\{0,1,2\}$ &
+\emph{Class Subset}~~~~ & & $\{0,1,2\}$ &
 $\{0,1,\dotsc,4\}$ &
@@
 \multirow{3}{*}{\emph{Algorithm \ref{algo}}~~~~} 
-& $\rho_{1}$ 
+& $\rho_{1}$\hspace{1em} 
 &$0.907\pm 0.007$
@@
 Moreover, this problem is equivalent
-to kernel $k$-means optimization problem, once the kernel is fixed; see
-Proposition~\ref{th:kernel_kmeans}. Our results imply that kernel $k$-means
+to kernel $k$-means optimization problem once the kernel is fixed; see
+Proposition~\ref{th:kernel_kmeans}. Energy statistics, however, fixes
+a family of standard kernels.
+Our results imply that kernel $k$-means
 is actually a consequence of energy statistics, and thus place this 
 method into a principled statistical basis.
-We also considered a weighted version of energy statistics clustering
-which establishes connection with spectral
+We also considered a weighted version of energy statistics clustering,
+establishing connections with spectral
 clustering and graph partitioning problems.
 
-We considered Algorithm~\ref{algo} based on Hartigan's method, and
+We considered Algorithm~\ref{algo} based on Hartigan's method and
 compared with 
@@
 
-In the same way that kernel $k$-means can be accelerated by exploring
-sparsity or sampling methods to approximate the Gram matrix, these
-techniques are also applicable to Algorithm~\ref{algo}.
-For instance, it would be interesting explore the methods recently proposed
-in \cite{Mahoney} to this case. 
+Finally, kernel methods can benefit from sparsity and
+fixed-rank approximations of the Gram matrix.
+For instance, an approach to make kernel $k$-means scalable
+was recently proposed \cite{Mahoney}, and a modified technique for
+rank reduction in Nystr\"om method was also recently considered 
+\cite{Becker}. It would be interesting to apply these and related
+approaches to the case of Algorithm~\ref{algo}.
 Another fruitful avenue of exploration would be to find 
 better methods to tackle the optimization problem \eqref{eq:qcqp2}.
-
-\bigskip
 

### 2017-08-13 qcqp_clustering/draft/energy_clustering.tex
new experiments, polished draft

@@
 Moreover, we also consider a weighted version of energy statistics 
-applied to clustering, which makes connections to 
+applied to clustering, which makes connection to 
 graph partitioning problems.
@@
 the point $x_0=0$ is chosen in the corresponding kernel  
-\eqref{eq:kernel_semimetric}. We thus have
+\eqref{eq:kernel_semimetric}. Therefore,
 \begin{equation}
@@
 \end{equation}
-We will consider other kernels as well, 
-but \eqref{eq:standard_metric} will be considered the
-standard kernel for energy statistics and will always be present in every
-experiment for reference. 
-%Notice that this is a convention, 
-%we could have chosen
-%any other semimetric/kernel as the standard.
+We will consider other kernels as well 
+but \eqref{eq:standard_metric} will be the
+standard kernel for energy statistics and will always be present in the
+experiments as a reference.
 
 Let us briefly mention that we compared Algorithm~\ref{algo} and
-Algorithm~\ref{algo1d} for several univariate distributions, and
+Algorithm~\ref{algo1d} for several univariate distributions and
 both perform almost indistinguishable regarding the clustering quality.
-We omit these results since we will analyse more interesting scenarios
-in higher dimensions and with more than two clusters.
+We omit these results since we will analyse more interesting scenarios.
 
 The main goal of the 
-experiments shown in this section 
-is to compare Algorithm~\ref{algo}, based on Hartigan's method, to
-kernel $k$-means as described in Algorithm~\ref{kmeans_algo}, based on
-Lloyd's method. From the discussion in the previous section we expect 
+experiments  to follow
+is to compare Algorithm~\ref{algo} based on Hartigan's method, to
+kernel $k$-means, as described in Algorithm~\ref{kmeans_algo}, which
+is based on Lloyd's method. 
+From the discussion in the previous section we expect 
 that Algorithm~\ref{algo}  be superior than Algorithm~\ref{kmeans_algo}.
@@
 Our measure of clustering quality will be the accuracy \eqref{eq:accuracy}
-based on the ground truth. Furthermore, in every experiment we sample
-data many times and show the average value of the accuracy, with error bars
-indicating standard error. Whenever possible, for reference, 
-we also show the best possible classification 
-accuracy obtained from a numerical estimation
-of Bayes classification error, which uses the true probability distribution
-of the data.
-
-From the results of \cite{Telgarsky}, the improvement of Hartigan's method
+based on the ground truth. Furthermore, in every experiment, we sample
+data many times and show the average value of the accuracy with error bars
+indicating the standard error. Whenever possible, 
+we also indicate the optimal 
+accuracy computed from
+Bayes classification error.
+
+From the results of \cite{Telgarsky}, the improvement of Hartigan's 
 over Lloyd's method should be more prominent in high dimensions.
-To analyse this we show 
+Thus, let us analyse
 how the algorithms degrade as the number of dimensions increase, while
@@
 Note that the Bayes error
-is also kept fixed as $D$ increases, given an optimal classification
+is fixed as $D$ increases, giving an optimal 
 accuracy of $\approx 0.86$.
-We keep signal only the first $10$ dimensions of $\mu_2$ and increase $D$.
 For each $D$ we sample $100$ times, picking $100$ points for each cluster.
 We apply each algorithm to the resulting dataset 
-and compute the average value of the accuracy \eqref{eq:accuracy}.
-For both Algorithm~\ref{algo} and Algorithm~\ref{kmeans_algo} 
-we use the standard
-kernel \eqref{eq:standard_metric} from energy statistics.
+and compute the average of the accuracy \eqref{eq:accuracy} (error bars
+are standard error).
+Algorithm~\ref{algo} and Algorithm~\ref{kmeans_algo} 
+use the standard
+kernel \eqref{eq:standard_metric}.
 The results are shown in Fig.~\ref{fig:gauss}a.
-Notice that GMM/EM algorithm is unable to estimate the covariance matrix 
+GMM/EM is unable to estimate the covariance matrices
 when the number of dimensions exceeds the number of points in each cluster.
 We see that Algorithm~\ref{algo} performs better than all the other ones,
-and in particular it works better than kernel $k$-means as the number
+and in particular it outperforms kernel $k$-means as the number
 of dimensions increase. Note also that although GMM and $k$-means are
-parametric, they are consistent models for the above distributions.
-Now we consider the same experiment but with different covariances,
-\begin{equation}
-\label{eq:gauss2}
-\begin{split}
-x &\in \C_i  \sim 
-\mathcal{N}(\mu_i,\Sigma_i) \qquad (i=1,2),  \\
-\mu_1 &= (\underbrace{0,\dotsc,0}_{\times D})^\top , \quad
-\mu_2 = (\underbrace{1,\dots,1}_{\times 10},
-\underbrace{0,\dots,0}_{\times (D-10)})^\top, \quad
-\Sigma_1 = I_D, \quad 
-\Sigma_2 = \left( \begin{smallmatrix} \widetilde{\Sigma} & 0 \\ 0 & I_{D-5}
-\end{smallmatrix}\right)
-\end{split}
-\end{equation}
-where $\widetilde{\Sigma}_{jj} = j$ for $j=1,\dotsc,5$.
-The best possible accuracy from Bayes classification error is
-$\approx 0.95$.
-The results are shown in Figure~\ref{fig:gauss}b.
-
-In the experiments of Figure~\ref{fig:gauss}
-we can see a better performance of Algorithm~\ref{algo} compared
-to the other ones, and in particular to kernel $k$-means,
-where we recall that both find approximate solutions
-to the same optimization problem \eqref{eq:qcqp2}.
-Notice that Algorithm~\ref{algo} is more robust as the number of dimensions
-increase.
+parametric they are consistent models to the data.
 
@@
 \begin{minipage}{0.49\textwidth}
-\centering
-\includegraphics[width=1\textwidth]{gauss_dim.pdf}\\[-1.0em]
-(a)
+\includegraphics[width=1\textwidth]{gauss_dim.pdf}\\[-1.0em] (a)
 \end{minipage}
 \begin{minipage}{0.49\textwidth}
-\centering
-\includegraphics[width=1\textwidth]{gauss_cov.pdf}\\[-1.0em]
-(b)
+\includegraphics[width=1\textwidth]{gauss_cov_squareroot.pdf}\\[-1.0em] (b)
+\end{minipage}
+\begin{minipage}{0.49\textwidth}
+\includegraphics[width=1\textwidth]{gauss_cov_linear.pdf}\\[-1.0em] (c)
+\end{minipage}
+\begin{minipage}{0.49\textwidth}
+\includegraphics[width=1\textwidth]{gauss_cov_square.pdf}\\[-1.0em] (d)
 \end{minipage}
@@
 standard $k$-means and GMM/EM as the number of dimensions increase.
-We sample $100$ times and show the average of \eqref{eq:accuracy} with
+We compute the average of \eqref{eq:accuracy} over $100$ samples with
 error bars being standard error. We have two clusters with $100$ points
 each. 
-(a) Data distributed as in \eqref{eq:gauss1}. The optimal accuracy
-from Bayes error is indicated by the dashed line which is $\approx 0.86$.
-(b) Data from \eqref{eq:gauss2}. The optimal accuracy is $\approx 0.95$.
+(a) Data as in \eqref{eq:gauss1}, where the optimal accuracy
+from Bayes error is the dashed line equal to $\approx 0.86$.
+(b) Data from \eqref{eq:gauss2} with $q=1/2$ in \eqref{eq:cov}. 
+(c) Data from \eqref{eq:gauss2} with $q=1$ in \eqref{eq:cov}.
+(d) Data from \eqref{eq:gauss2} with $q=2$ in \eqref{eq:cov}.
+The optimal accuracy in these last cases is $1$.
 }
@@
 
-In Figure~\ref{fig:unbalanced} we consider the effect of having 
+Let us now consider the same experiment as before with data distributed as
+\begin{equation}
+\label{eq:gauss2}
+x \in \C_i  \sim 
+\mathcal{N}(\mu_i,\Sigma_i) \quad (i=1,2),  \quad
+\mu_1 = (\underbrace{0,\dotsc,0}_{\times D})^\top , \quad
+\mu_2 = (\underbrace{1,\dots,1}_{\times 10},
+\underbrace{0,\dots,0}_{\times (D-10)})^\top, 
+\end{equation}
+where
+\begin{equation}
+\label{eq:cov}
+(\Sigma_1)_{ij} = \begin{cases}
+i^{-q} \delta_{ij} & \mbox{if $i \le 10$} \\
+\delta_{ij} & \mbox{if $10 < i \le D$}
+\end{cases} \qquad
+(\Sigma_2)_{ij} = \begin{cases}
+i^{\,q} \delta_{ij} & \mbox{if $i \le 10$} \\
+\delta_{ij} & \mbox{if $10 < i \le D$}
+\end{cases} 
+\end{equation}
+and we choose $q \in \{ 1/2, 1, 2\}$. In these cases,
+the best possible accuracy from Bayes classification error is $\approx 1$.
+In Fig.~\ref{fig:gauss}b--d we have $q=1/2$, $q=1$, and $q=2$, respectivelly.
+GMM/EM was unnable to estimate the covariance matrices
+as dimensions get large. Algorithm~\ref{algo} outperforms
+Algorithm~\ref{kmeans_algo}, and $k$-means degrades faster as 
+$q$ increases.
+
+To summarize, in the experiments of Fig.~\ref{fig:gauss}
+we see a better performance of Algorithm~\ref{algo} compared
+to the other ones, and in particular to kernel $k$-means,
+where we recall that both find approximate and local solutions
+to the same optimization problem \eqref{eq:qcqp2}.
+Algorithm~\ref{algo} is more robust as the number of dimensions
+increase.
+
+Let us now
+consider the effect of having 
 unbalanced clusters. We generate data as
@@
 \end{equation}
-We increase $m$, that is we make the clusters progressively more unbalanced,
-and plot the mean of \eqref{eq:accuracy} with standard errors, where
-we sample $100$ times for each $m$.
+We then increase $m$, 
+that is we make the clusters progressively more unbalanced,
+and plot the average of \eqref{eq:accuracy} 
+over
+$100$ samples for each $m$
+(error bars are standard error). The results are in Fig.~\ref{fig:unbalanced}.
 As expected, GMM
 works better than the other algorithms since it is a fuzzy clustering.
-We can see that the other methods based on hard assignments degrade similarly,
+The other methods based on hard assignments degrade similarly
 and  more rapidly than GMM. This indicates that a fuzzy version of
@@
 \label{fig:unbalanced}
-Comparison of Algorithm~\ref{algo}, Algorithm~\ref{kmeans_algo}, standard
-$k$-means and GMM/EM for data distributed as in \eqref{eq:gauss3}, where
-we make the two clusters progressively more unbalanced.
+Comparison of Algorithm~\ref{algo}, Algorithm~\ref{kmeans_algo}, 
+$k$-means, and GMM/EM. The data is distributed as \eqref{eq:gauss3} where
+we make the clusters progressively more unbalanced.
 }
@@
 Besides the standard kernel from energy statistics 
-\eqref{eq:standard_metric}, associated to \eqref{eq:energy}, 
+\eqref{eq:standard_metric}, 
 let us consider two other semimetrics with their respective generating
@@
 \eqref{eq:energy2} with $\alpha=1/2$.
-Now consider data normally distributed in $D=20$ as follows:
+Now consider data normally distributed in $D=20$ as
 \begin{equation}
@@
 We sample an equal number of points for each cluster, which is progressively
-increased. The optimal accuracy of \eqref{eq:20gauss} based on Bayes
-classification error is $\approx 0.90$. We repeat the same experiment as
-before but now using also the kernels \eqref{eq:rhohalf} and
-\eqref{eq:rhoe}. The results are shown in Fig.~\ref{fig:consist}a.
-Algorithm~\ref{algo} performs better than all the other ones, and in 
+increased. The optimal accuracy based on Bayes
+classification error is $\approx 0.90$. 
+Clustering results are shown in Fig.~\ref{fig:consist}a.
+Algorithm~\ref{algo} outperforms the other ones, and in 
 particular the kernel \eqref{eq:rhoe} provides better results.
-As the number of points get large enough, GMM starts to approach
-optimal Bayes accuracy, 
-as it should since it is 
-consistent model to the distribution of the data. However, 
-Algorithm~\ref{algo} with kernel \eqref{eq:rhoe} achieves optimal accuracy
-with a much smaller number of points. Moreover, Algorithm~\ref{algo} performs
-better than Algorithm~\ref{kmeans_algo} for any of the kernel choices.
-Standard $k$-means performs poorly in this example.
+As the number of points get large enough GMM starts to approach
+optimal Bayes, as it should since it is a
+consistent model to the data. However, 
+Algorithm~\ref{algo} with kernel \eqref{eq:rhoe} approach optimal Bayes
+with a much smaller number of points. Moreover, Algorithm~\ref{algo} 
+outperforms
+Algorithm~\ref{kmeans_algo} for any of the kernel choices.
+%Standard $k$-means performs poorly in this example.
+
+Now consider the same parameters as in \eqref{eq:20gauss} but with
+lognormal distributions, 
+\begin{equation}
+\label{eq:20loggauss}
+\log x \in \C_i \sim \mathcal{N}(\mu_i, \Sigma_i) \qquad (i=1,2).
+\end{equation}
+The same previous experiment is shown in 
+Fig.~\ref{fig:consist}b.
+Note that Algorithm~\ref{algo} still performs accurately, 
+while $k$-means works almost at chance,
+and GMM is not even able to estimate the parameters. Again, the
+kernel \eqref{eq:rhoe}
+provides better results than \eqref{eq:standard_metric} or
+\eqref{eq:rhohalf}. 
+The experiments in Fig.~\ref{fig:consist}
+illustrate how energy statistics clustering is nonparametric.
 
@@
 \eqref{eq:standard_metric}, \eqref{eq:rhohalf} and \eqref{eq:rhoe}, 
-$k$-means and GMM. The optimal accuracy in both cases
+$k$-means, and GMM. The optimal accuracy in both cases
 is $\approx 0.9$. We show the average of \eqref{eq:accuracy}
@@
 \eqref{eq:20gauss}. 
-(b) Data as in \eqref{eq:20loggauss}.
+(b) Data distributed as in \eqref{eq:20loggauss}.
 }
 \end{figure}
-
-Consider the same parameters as in \eqref{eq:20gauss} but now with
-lognormal distributions, 
-\begin{equation}
-\label{eq:20loggauss}
-\log x \in \C_i \sim \mathcal{N}(\mu_i, \Sigma_i) \qquad (i=1,2).
-\end{equation}
-The same previous experiment is shown in 
-Fig.~\ref{fig:consist}b.
-Note that Algorithm~\ref{algo} still performs accurately, 
-while $k$-means works almost at chance,
-and GMM is not even able to estimate the parameters. Again, the
-kernel \eqref{eq:rhoe}
-provides better results than \eqref{eq:standard_metric} or
-\eqref{eq:rhohalf}. Algorithm~\ref{algo} performs better
-than Algorithm~\ref{kmeans_algo} for every kernel choice. 
-The experiments in Fig.~\ref{fig:consist}
-illustrates that energy statistics clustering is nonparametric.
 
@@
 concentric circles with noise. (c) Three
-concentric circles. (d) MNIST handwritten digits.
+concentric circles with noise. (d) MNIST handwritten digits.
 Clustering results are in Table~\ref{table:other}
@@
 
-Now consider the following choices of semimetric and 
+Consider the following choices of semimetric and 
 corresponding generating kernel:
@@
 In Fig.\ref{fig:other} we have examples of
-complex two dimensional datasets. For each class in Fig.\ref{fig:other}a
-we sample $200$ points. In each class of 
-Fig.\ref{fig:other}b and Fig.\ref{fig:other}c we sample $400$ points.
+complex two dimensional datasets. The two parallel cigars of 
+Fig.~\ref{fig:other}a have $200$ points each. For the concentric circles
+of Fig.~\ref{fig:other}b and Fig.~\ref{fig:other}c we
+sample $400$ points for each class.
 We apply Algorithm~\ref{algo} and Algorithm~\ref{kmeans_algo}, besides
@@
 in Table~\ref{table:other} where the respective choice of parameters for the
-kernels are indicated. In the data shown in Fig.~\ref{fig:other}a 
-the standard semimetrics from energy statistics, $\rho_1$ and $\rho_{1/2}$, are
+kernels are indicated. For the data in Fig.~\ref{fig:other}a 
+the semimetrics $\rho_1$ and $\rho_{1/2}$ are
 able to provide more accurate results compared to $k$-means. However,
-the gaussian kernel $\widetilde{\rho}_2$ gives results very accurate
+the gaussian kernel $\widetilde{\rho}_2$ gives very accurate
 results, similar to GMM,
-which is the true model of the data. In Fig.~\ref{fig:other}b we see
+which is the true model of the data. For the data shown in
+Fig.~\ref{fig:other}b we see
 that the choice of kernel is very sensitive, and only \eqref{eq:rhoesigma}
@@
 of Fig.~\ref{fig:other} still provides better results than
-the other choices, but the results are less accurate.
+the others, but they are less accurate compared to Fig.~\ref{fig:other}b.
 
 \newcolumntype{g}{>{\columncolor{gray!20}}l}
-\begin{table}
+\begin{table}[h]
+\renewcommand*{\arraystretch}{0.75}
 \begin{tabular}{@{}r  l g  l g  l g@{}}
@@
 Clustering the data shown in Fig.~\ref{fig:other} with Algorithm~\ref{algo}
-and Algorithm~\ref{kmeans_algo}, besides $k$-means and GMM, for the kernels 
-\eqref{eq:rhoalpha},
-\eqref{eq:rhoesigma} and
-\eqref{eq:rhogsigma}. We sample the data $10$ times and show the average
+and Algorithm~\ref{kmeans_algo}, besides $k$-means and GMM, with kernels 
+\eqref{eq:rhoalpha}--\eqref{eq:rhogsigma}. 
+We sample $10$ times and show the average
 accuracy \eqref{eq:accuracy} with standard error.
@@
 Each data point is an $8$-bit gray scale
-image, forming a $784$-dimensional vector, 
+image forming a $784$-dimensional vector 
 corresponding to the digits $\{0,1,\dotsc,9 \}$.
 Besides the kernel \eqref{eq:rhoalpha}, we
-consider the gaussian kernel \eqref{eq:rhogsigma} with the width
-given by
+consider the gaussian kernel \eqref{eq:rhogsigma} with 
 \begin{equation}
 \label{eq:sigma}
-\sigma^2 = \dfrac{1}{n^2} \sum_{i,j=1}^n \| x_i - x_j \|^2
+\sigma^2 = \dfrac{1}{n^2} \sum_{i,j=1}^n \| x_i - x_j \|^2 ,
 \end{equation}
@@
 where we sample $100$ points 
-for each class. We cluster the resulting dataset 
-using Algorithm~\ref{algo},
+for each class. We perform clustering through 
+Algorithm~\ref{algo},
 Algorithm~\ref{kmeans_algo}, and $k$-means.
-The results are shown in Table~\ref{table:mnist}, where the kernel
+The results are shown in Table~\ref{table:mnist} where the kernel
 and its parameter for each case is indicated.
@@
 \newcolumntype{g}{>{\columncolor{gray!20}}l}
-\begin{table}
+\begin{table}[h]
+\renewcommand*{\arraystretch}{0.75}
 \begin{tabular}{@{}rl|l|l|l|l@{}}
@@
 \caption{\label{table:mnist}
-Clustering the data shown in Fig.~\ref{fig:other}d with Algorithms~\ref{algo}
-and \ref{kmeans_algo}, and $k$-means.
-We use the kernel \eqref{eq:rhoalpha} with $\alpha\in\{1,2\}$, and
+Clustering the data shown in Fig.~\ref{fig:other}d with Algorithm~\ref{algo},
+Algorithm~\ref{kmeans_algo}, and $k$-means.
+We use the kernel \eqref{eq:rhoalpha} with $\alpha\in\{1,2\}$ and
 the gaussian kernel 
@@
 For each subset of digits we sample $10$ times and show the average
-accuracy \eqref{eq:accuracy} with standard error. We pick $100$ points for 
+accuracy \eqref{eq:accuracy} with standard error. We sample $100$ points for 
 each class.
@@
 We showed that the clustering problem reduces to a quadratically
-constrained optimization problem (QCQP), 
+constrained quadratic program, 
 as described in Proposition~\ref{th:qcqp2}.
@@
 compared with 
-kernel $k$-means described in Algorithm~\ref{kmeans_algo} which
+kernel $k$-means, described in Algorithm~\ref{kmeans_algo}, which
 is based on Lloyd's heuristic.
@@
 \cite{Becker}. It would be interesting to apply these and related
-approaches to the case of Algorithm~\ref{algo}.
+techniques to the case of Algorithm~\ref{algo}.
 Another fruitful avenue of exploration would be to find 

### 2017-08-13 qcqp_clustering/code/eclust/data.py

@@ -336,14 +336,22 @@ if __name__ == '__main__':
     #X, z = multivariate_normal([m1, m2], [s1, s2], [200, 200])
     #plot(X, z, fname='./2cigars.pdf')
     
+    # ball and cigar
+    m1 = np.zeros(2)
+    m2 = np.array([3,0])
+    s1 = np.array([[.3,0],[0,.3]])
+    s2 = np.array([[1,0],[0,20]])
+    X, z = multivariate_normal([m1, m2], [s1, s2], [200, 200])
+    plot(X, z, fname='./ball_cigar.pdf')
+    
     # 2 circles
-    X, z = circles([1, 3], [[0,0], [0,0]], [0.2, 0.2], [400, 400])
-    plot(X, z, fname='./2circles.pdf')
+    #X, z = circles([1, 3], [[0,0], [0,0]], [0.2, 0.2], [400, 400])
+    #plot(X, z, fname='./2circles.pdf')
     
     # 3 circles
-    X, z = circles([1, 3, 5], [[0,0], [0,0], [0,0]], [0.2, 0.2, 0.2], 
-                    [400, 400, 400])
-    plot(X, z, fname='./3circles.pdf')
+    #X, z = circles([1, 3, 5], [[0,0], [0,0], [0,0]], [0.2, 0.2, 0.2], 
+    #                [400, 400, 400])
+    #plot(X, z, fname='./3circles.pdf')
 
     #X, z = spirals([1,-1], [[0.2,0.0], [-0.2,-0.0]], [400,400], noise=0.2)
     #plot(X, z, fname='./2spiral.pdf')

### 2017-08-13 qcqp_clustering/draft/energy_clustering.tex
more polishing

@@
 metric spaces of strong negative type,
-and more recently a connection with reproducing kernel Hilbert spaces (RKHS)
+and more recently a connection with reproducing kernel Hilbert spaces 
 was proposed.
@@
 as kernel $k$-means algorithm but usually with better clustering
-performance. We provide carefully designed numerical experiments
+quality. We provide carefully designed numerical experiments
 showing the superiority of this method compared to
@@
 hierarchical clustering \cite{RizzoClustering} by extending Ward's
-method of minimum variance. (We refer to \cite{Szkely2013} and 
-references therein for an overview of energy statistics.)
+method of minimum variance. We refer to \cite{Szkely2013} and 
+references therein for an overview.
 Moreover, an energy statistics formulation
@@
 information about each cluster
-comes from its mean and thus it is sensitive to outliers. Moreover, $k$-means
+comes from its mean and it is thus sensitive to outliers. Moreover, $k$-means
 is strongly parametric, assuming that data is isotropically and normally
@@
 complicated datasets, however, the kernel choice remains the biggest challenge
-since there is no principled theory behind this. The well-known
+since there is no principled theory behind this, and it usually introduces
+extra parameters that need to be optimized. The well-known
 kernel $k$-means optimization problem is nothing but $k$-means in the
@@
 spectral clustering, and graph partitioning problems was already established
-in \cite{Dhillon,Dhillon2}. We show how these connections arise
+\cite{Dhillon,Dhillon2}. We show how these connections arise
 from a weighted version of energy statistics.
@@
 Our algorithmic contribution is to use Hartigan's method
-\cite{Hartigan} to find local solutions of the above mentioned QCQP.
-This approach was also proposed in \cite{Kgroups}. In this case,
+\cite{Hartigan} to find local 
+solutions of the above mentioned QCQP.
+This approach was also proposed in \cite{Kgroups}.
+In this case,
 the resulting algorithm has the same
@@
 evidence the nonparametric aspect of energy
-statistics based clustering, which provides a default kernel, showing that
-it is able to perform accurately on data coming very different 
+statistics based clustering, which provides a family of 
+default kernels, showing that
+it is able to perform accurately on datasets coming very different 
 distributions, contrary to $k$-means and GMM for instance.
@@
 definite. Indeed, notice that
-$\langle \kkk_x, \kkk_y \rangle = \kkk_y(x) = \kk(x,y)$, and by definition
+$\langle \kkk_x, \kkk_y \rangle = \kkk_y(x) = \kk(x,y)$, and by 
+definition 
 $\langle \kkk_x, \kkk_y \rangle^* = \langle \kkk_y, \kkk_x \rangle$, but
@@
 $\kk(y,x) = \kk(x,y)$. Moreover, for any $w \in
-\HH$ we can write $w = \sum_{i=1}^n c_i \kkk_{x_i}$, where
+\HH$ we can write $w = \sum_{i=1}^n c_i \kkk_{x_i}$ where
 $\{ \kkk_{x_i} \}_{i=1}^n$ is a basis of $\HH$. It follows that
@@
 showing that the kernel is positive definite. If $G$ is a matrix with
-elements $G_{ij} = \kk(x_i,x_j)$, this is equivalent to $G$ being
+elements $G_{ij} = \kk(x_i,x_j)$ this is equivalent to $G$ being
 positive semidefinite, i.e. $v^\top G \, v \ge 0$ for any vector
@@
 about  the distribution of the data, 
-contrary to $k$-means and gaussian mixture models (GMM), for example.
-%Moreover, this approach \emph{does not} require the concept of the 
-%\emph{cluster mean},
-%which can be ill-defined for some types of data, such as images for
-%instance. Since no mean is involved, the method should also be 
-%robust to outliers.
-%This method is quite general and makes very few
-%assumptions about the data. 
+contrary to $k$-means and GMM, for example.
 If one uses the standard energy distance
@@
 already shown in the proof of Proposition~\ref{th:qcqp2}, leading to
-the optimization problem in the form \eqref{eq:qcqp2}.
+\eqref{eq:qcqp2}.
 \end{proof}
@@
 kernel $k$-means problem is actually a consequence
-of energy statistics. In this vein, kernel $k$-means problem is part of a
-statistical framework where we are maximizing distances between probability
-distributions. The kernel function $\kk$ arises from a semimetric $\rho$ of
+of energy statistics. In this vein, kernel $k$-means is part of a
+statistical framework where distances between probability
+distributions are maximized. 
+The kernel function $\kk$ arises from a semimetric $\rho$ of
 negative type which defines the energy distance between distributions.
@@
 This procedure is described in Algorithm~\ref{algo1d}. 
-Notice that
+Note that
 this method does not require any initialization,
@@
 $0$ to all points wrongly clustered.
-For a two-class problem with equal
-number of points in each cluster, the value $1/2$ correspond
+For a two-class problem with balanced
+clusters, the value $1/2$ correspond
 to chance.
@@
 \end{equation}
-For these distributions the optimal classification accuracy 
-obtained from Bayes error
+For these distributions the optimal accuracy 
+obtained from Bayes classification error
 is $ \approx 0.956$ which is indicated by the dashed line in the plot.
@@
 Energy statistics performs slightly better
-than $k$-means. On the other hand, in Fig~\ref{fig:1d}b
+than $k$-means. On the other hand, in Fig.~\ref{fig:1d}b
 we consider two clusters with lognormal distributions,
@@
 thus giving zero accuracy.
-These two simple experiments from Fig.\ref{fig:1d} illustrate
+The two simple experiments of Fig.~\ref{fig:1d} illustrate
 how energy statistics clustering is nonparametric, being able
@@
 Kernel $k$-means algorithm 
-to find an approximate solution to \eqref{eq:qcqp2}.
+to find a local solution to \eqref{eq:qcqp2}.
 \hspace{\fill}
@@
 We now consider an algorithm based on Hartigan's method \cite{Hartigan} 
-applied to \eqref{eq:maxQ}. This gives a local and approximate solution to 
-\eqref{eq:qcqp2}. This approach was also proposed in \cite{Kgroups}.
+applied to \eqref{eq:maxQ}. This gives a local solution to 
+\eqref{eq:qcqp2}. 
+%This approach was also proposed in \cite{Kgroups}.
 The method is based on the change
@@
 \caption{\label{algo}
-Hartigan's method to find an approximate solution to \eqref{eq:qcqp2}.
+Hartigan's method to find a local solution to \eqref{eq:qcqp2}.
 \hspace{\fill}
@@
 
-Both condition above are not satisfied 
+Both conditions of the  above theorem are not satisfied 
 by Lloyd's method, and consequently by (kernel) $k$-means algorithm.
-The next result indicates that Hartigan's has the potential to
-escape local optima of Lloyd's.
+The next result indicates that Hartigan's can potentially 
+escape local optima of Lloyd's method.
 
@@
 to Lloyd's method. Therefore, Hartigan's method 
-can potentially escape local
+can escape local
 optima of Lloyd's method. 
@@
 with any Bregman divergence. They showed that the number of Hartigan's
-local minima is upper bounded by $O(1/k)$ \cite[Proposition 5.1]{Slonin}. 
+local minima is upper bounded by $\mathcal{O}(1/k)$ 
+\cite[Proposition 5.1]{Slonin}. 
 In addition, they provide examples where
 \emph{any} initial partition correspond to a local optima of Lloyd's 
-method, while  the number of local optima in Hartigan's is small and 
+method, while  the number of local optima in Hartigan's method is small and 
 correspond to true partitions of the data. Empirically, the number of
@@
 both perform almost indistinguishable regarding the clustering quality.
-We omit these results since we will analyse more interesting scenarios.
+However, we omit these results since we will analyse more interesting 
+scenarios in high dimensions and $k > 2$ number of clusters.
 
@@
 experiments  to follow
-is to compare Algorithm~\ref{algo} based on Hartigan's method, to
+is to compare Algorithm~\ref{algo} based on Hartigan's method to
 kernel $k$-means, as described in Algorithm~\ref{kmeans_algo}, which
 is based on Lloyd's method. 
-From the discussion in the previous section we expect 
-that Algorithm~\ref{algo}  be superior than Algorithm~\ref{kmeans_algo}.
+From the discussion in the end of the previous section we can anticipate
+that Algorithm~\ref{algo} will be superior than Algorithm~\ref{kmeans_algo}.
 Another goal is to illustrate the nonparametric aspect of energy statistics.
@@
 From the results of \cite{Telgarsky}, the improvement of Hartigan's 
-over Lloyd's method should be more prominent in high dimensions.
+over Lloyd's method should be more accentuated in high dimensions.
 Thus, let us analyse
@@
 Algorithm~\ref{algo} and Algorithm~\ref{kmeans_algo} 
-use the standard
+both use the standard
 kernel \eqref{eq:standard_metric}.
 The results are shown in Fig.~\ref{fig:gauss}a.
-GMM/EM is unable to estimate the covariance matrices
-when the number of dimensions exceeds the number of points in each cluster.
+Note that GMM/EM is unable to estimate the covariance matrices
+when the number of dimensions exceeds the number of points in each cluster,
+giving zero accuracy when dimensions are $\gtrsim 100$.
 We see that Algorithm~\ref{algo} performs better than all the other ones,
-and in particular it outperforms kernel $k$-means as the number
+and in particular it outperforms kernel $k$-means algorithm as the number
 of dimensions increase. Note also that although GMM and $k$-means are
@@
 Comparison of Algorithm~\ref{algo}, Algorithm~\ref{kmeans_algo},
-standard $k$-means and GMM/EM as the number of dimensions increase.
+standard $k$-means and GMM/EM as the number of dimensions increase
+in gaussian settings.
 We compute the average of \eqref{eq:accuracy} over $100$ samples with
@@
 (d) Data from \eqref{eq:gauss2} with $q=2$ in \eqref{eq:cov}.
-The optimal accuracy in these last cases is $1$.
+The optimal accuracy from Bayes error in (b--d) is $\approx 1$.
 }
@@
 In Fig.~\ref{fig:gauss}b--d we have $q=1/2$, $q=1$, and $q=2$, respectivelly.
-GMM/EM was unnable to estimate the covariance matrices
-as dimensions get large. Algorithm~\ref{algo} outperforms
+Again, GMM/EM is unable to estimate the covariance matrices
+as dimensions get larger than $\gtrsim 100$, and it gives poor results even
+for number of dimensions much lower than this. Note that GMM requires
+a larger number of points to estimate the parameters accurately.
+Algorithm~\ref{algo} outperforms
 Algorithm~\ref{kmeans_algo}, and $k$-means degrades faster as 
@@
 to the other ones, and in particular to kernel $k$-means,
-where we recall that both find approximate and local solutions
+where we recall that both find local solutions
 to the same optimization problem \eqref{eq:qcqp2}.
@@
 \end{align}
-In Fig.\ref{fig:other} we have examples of
+In Fig.~\ref{fig:other} we have examples of
 complex two dimensional datasets. The two parallel cigars of 
@@
 We apply Algorithm~\ref{algo} and Algorithm~\ref{kmeans_algo}, besides
-$k$-means and GMM, using the above kernels. The results are shown
+$k$-means and GMM, using the above kernels 
+\eqref{eq:rhoalpha}--\eqref{eq:rhogsigma}. The results are shown
 in Table~\ref{table:other} where the respective choice of parameters for the
@@
 \end{equation}
-which is computed from the sampled data $\{ x_i \}_{i=1}^n$.
+which is computed from a sample  $\{ x_i \}_{i=1}^n$.
 We consider subsets of the  classes $\{0,1,\dotsc,9 \}$, 
@@
 Proposition~\ref{th:kernel_kmeans}. Energy statistics, however, fixes
-a family of standard kernels.
+a family of standard kernels consistent with \eqref{eq:energy2}, and
+more general kernels related to \eqref{eq:energy3} can be obtained.
 Our results imply that kernel $k$-means
@@
 Both have the same time complexity, however, the numerical 
-results provide evidence that Algorithm~\ref{algo} is more robust with
+results provide compelling evidence that Algorithm~\ref{algo} 
+is more robust with
 a superior clustering performance. This is also theoretically

### 2017-09-07 qcqp_clustering/draft/energy_clustering.tex
comments

@@
 has not been previously incorporated in machine learning problems.
+% @gui: not sure what you mean?
+what counts as machine learning?
+be specific, i think yo mean clustering?
+even then, it is not true, both the master's thesis and the RKHS paper kind of do it.
+be as general as possible about the gap that we are filling, but not more general.
 Here we consider the problem of clustering data from
@@
 kernel $k$-means from energy statistics.
+% @gui: have other people shown a relationship between kernel k-means and say, RKHS, or something else?  
 Moreover, we also consider a weighted version of energy statistics 
@@
 The clustering problem has a long history in machine learning.
-The most used method is $k$-means \cite{Lloyd,MacQueen,Forgy}, which
+Perhaps the most used method is $k$-means \cite{Lloyd,MacQueen,Forgy}, which
 is based on Lloyd's heuristic \cite{Lloyd} of assigning a data point to
@@
 information about each cluster
-comes from its mean and it is thus sensitive to outliers. Moreover, $k$-means
+comes from its mean and it is thus sensitive to outliers. 
+% @gui: algorithms don't make assumptions, theorems do.  also, algorithms cannot be parametric or non-parametric, that is a property only of the statistical model, or an estimator. 
+Moreover, $k$-means
 is strongly parametric, assuming that data is isotropically and normally
 distributed since it is a special case of gaussian mixture models (GMM) 
+% @gui: k-means is NOT a special case of GMM, GMM is a soft clustering (not fuzzy, in stats and machine learning we don't say "fuzzy"), where as k-means is a hard clustering.
+k-means is certainly closely related to GMM though.
 through the expectation maximization (EM) algorithm. Nevertheless,
@@
 significant overlap of the points belonging to different clusters.
+% i@gui:  don't know what "clear resolution of cluster centers" adds to this?
 
@@
 space by using its inner product. This enables one to cluster way more
-complicated datasets, however, the kernel choice remains the biggest challenge
-since there is no principled theory behind this, and it usually introduces
-extra parameters that need to be optimized. The well-known
+complicated datasets, 
+% @gui: algorithms cluster, doesn't matter how complicated the data are.  so, what do you actually mean here?  perform well empirically? what does "complicated" actually mean?
+however, the kernel choice remains the biggest challenge
+since there is no principled theory behind this, 
+% @gui: the words "this" and "it" are often vague and ambiguous, be specific/precise
+and it usually introduces
+extra parameters that need to be optimized. 
+% @gui: not sure what you mean here, in kernel k-means, we don't "optimize" over any parameters typically.  sometimes we select hyperparameters of kernels via cross-validation, is that what you mean?
+The well-known
 kernel $k$-means optimization problem is nothing but $k$-means in the
@@
 statement, showing that there is only one possible optimization problem
-consistent with energy statistics. Moreover, we show that this is
+consistent with energy statistics. 
+% @gui: not sure what you mean "only one possible optimization problem".  we can reformulate it many different ways? also, below, "this" is ambiguous.
+Moreover, we show that this is
 a quadratically constrained
@@
 distributions, contrary to $k$-means and GMM for instance.
-
+% @gui: to me, we did not make explicit one of the most important points.
+in this paper, we make clear exactly when we expect energy clustering to perform better than k-means/GMM, and when we expect it to perform worse (empirically).
+
+% @gui: i find ToC paragraphs useless.
 Our work is organized as follows. In section~\ref{sec:background} we introduce
@@
 called \emph{energy distance} \cite{Szkely2013} 
-is rotational invariant and nonnegative, $\Energy(P,Q) \ge 0$, where
+is rotationally invariant and nonnegative, $\Energy(P,Q) \ge 0$, where
 equality
@@
 
+% @gui: let the reader decide what is important, rather than us telling them :)
 It is important to 
@@
 $\rho$ may not satisfy the triangle inequality, 
-$\rho^{1/2}$ does since it can be shown to be a legit metric.
+$\rho^{1/2}$ does since it can be shown to be a legitamate metric.
 
@@
 according to \eqref{eq:inner_data}.
+% @gui: i don't know what this means.  eqn 7 defines something for a population statistic, not a sample statistic.
 
@@
 \cite{Szkely2013}.
+% @gui: i'm not sure the above is true/meaningful.  nonparametric means implicitly estimating an infinite number of parameters.  without any statistical theory, i'm not quite sure what we can say.
 When the sample size is large enough, $n\to \infty$,
@@
 
+% @gui: consider only using equation numbers for equations that you later refer to? there are so many, it is easy to get lost.
 Due to the test statistic \eqref{eq:between} for equality of distributions,
@@
 \eqref{eq:within} which has a more convenient form.
+% @gui: where possible, i recommend using the "name" of the quantity rather than the number of the equation.  eg, here, write S and W, rather than (13) and (12).
 
@@
 \end{proof}
+% @gui: do other people know this result? if so, cite them, if not, state that.
 
@@
 finding the best partition of the data by solving~\eqref{eq:minimize}.
+% @gui: here and elsewhere is another good example.  why not just say "by minizing W"?
 Notice that this is a hard clustering problem as partitions
@@
 
+% @gui: "Let us" is used a bunch, though i think it never adds anything? i'd remove it everywhere.
 Let us formulate \eqref{eq:minimize} in the corresponding
@@
 indicating to which cluster (column)
-each point (row) belongs to. This matrix satisfy
+each point (row) belongs to. This matrix satisfies
 $Z^\top Z = D$ where the diagonal matrix 
@@
 \end{equation}
+% @gui: i don't understand the notaiton (.)_jj. is a sum missing?
 where we used the definitions \eqref{eq:kernel_matrix} and
@@
 \eqref{eq:max_prob} is equal to
-$\Tr \left( D^{-1} Z^\top G Z \right)$. Now we can
+$\Tr \left( D^{-1} Z^\top G Z \right)$. 
+% @gui: the above step is not obvious
+Now we can
 use the cyclic property
-of the trace, and by the own definition of the matrix $Z$
+of the trace, and by the  definition of the matrix $Z$
 in \eqref{eq:label_matrix}, we obtain the following integer
@@
 nonnegative.
+
+% @jovo: stopped here.
 

### 2017-09-17 qcqp_clustering/draft/energy_clustering.tex
more comments

@@
 showing the superiority of this method compared to
-kernel $k$-means, standard $k$-means, and gaussian mixture models.
+kernel $k$-means, standard $k$-means, and gaussian mixture models in a variety of settings.
 \end{abstract}
@@
 Now we write this in terms of the matrix $Y = Z D^{-1/2}$.
+% @gui: Y is a matrix? or a vector? when introducing new terms, always define what space they are in rigorously
 The objective function immediately becomes
@@
 to \eqref{eq:qcqp2} .
+% @gui: don't use equation numbers here.
 \end{proof}
 
+% @jv: stopped here
 Based on Proposition~\ref{th:qcqp2}, to group data
@@
 contrary to $k$-means and GMM, for example.
+% @gui: algorithms don't make assumptions, proofs do.  we don't prove anything about when we get the right answer, so we can't quite say this.
 If one uses the standard energy distance
@@
 and cluster centers, which arises from $k$-means objective function.
-It was shown \cite{Dhillon2,Dhillon} 
+ \cite{Dhillon2,Dhillon} show
 that problem \eqref{eq:kernel_kmeans} 
@@
 kernel $k$-means problem is actually a consequence
+% @gui: i'm not sure about this.  who is to say the correct direction of causality? maybe energy clustering is a special case of kernel k-means?
 of energy statistics. In this vein, kernel $k$-means is part of a
@@
 analysis due to the relation of these problems to
-energy statistics and RKHS, which may provide a different perspective.
+energy statistics and RKHS, which  provides a different perspective.
 
@@
 on graphs.
+% @gui: can you provide some guidance here, in particular, what do the weights "mean" or "correspond to"? right now, it feels like just math.
 
@@
 complexity of the algorithm is $\OO(n\log n + n^2) = \OO(n^2)$.
+% @gui: "does not require any initiializaton" i think you mean is guaranteed to find the optimal solution in O(n^2) time.
 
@@
 \caption{
+% @gui: to provide intuition to the reader, let's show the kernel density estimate for each of the 3 algorithms for both classes.
+i'm not sure the best way to visualize it, maybe play with a few options.
+% @gui: also, the colors are not very different.  i use the following color pallete:
+%co = [0,255,0; % green
+%0,255,255;  % cyan
+%255,0,255;  % magenta
+%255,127,0;  % orange
+%55,126,184; % blue
+%153,153,153;% gray
+%152,78,163; % purple
+%166,86,40;  % brown
+%247,129,191;% pink
+%77,175,74;  % forest
+%228,26,28;  % red
+%255,255,51];% yellow
 \label{fig:1d}
@@
 The data is clustered using 
-Algorithm~\ref{algo1d}, GMM through EM algorithm, and standard $k$-means 
-algorithm. In the last two cases 
+Algorithm~\ref{algo1d}, 
+% @gui: i see why you call this algorithm 1, but can we call it something with semantic meaning? like 1D-\mathcal{E}-clustering? or maybe \mathcal{E}1 for short? same for the other variants.
+GMM through EM algorithm, and standard $k$-means 
+algorithm. 
+% @gui: there are many k-means algorithms, be specific.
+In the last two cases 
 we use the initialization from $k$-means++ \cite{Vassilvitskii} 
@@
 few times choosing the result with  best objective function value. 
+% @gui: a few? be specific
 Notice that Algorithm~\ref{algo1d} requires no initialization so we only
+% @gui: "no initiazation" is weird.  do any algs require an initialization?
 run it once.
@@
 \end{equation}
+% @gui: why not use the same parameters as in fig 1a?  i feel like if we did, it would seem less like we are cherry picking?  
 The optimal classification accuracy from Bayes error is $\approx 0.852$.
+% @gui: computed numerically?
 In this case,
@@
 In this section we introduce an iterative algorithm to find a local
-maximizer of \eqref{eq:qcqp2}. However, due to 
+maximizer of \eqref{eq:qcqp2}. Due to 
 Proposition~\ref{th:kernel_kmeans} we can also find an approximate
@@
 The entire procedure is described in Algorithm~\ref{kmeans_algo}.
+% @gui: let's call this something more informative
 It can be shown that this algorithm converges provided $G$ is positive
@@
 Let us mention some important results about Hartigan's method.
+% @gui: "let us" again...
 
@@
 Both conditions of the  above theorem are not satisfied 
-by Lloyd's method, and consequently by (kernel) $k$-means algorithm.
+by Lloyd's method, 
+% @gui: do you mean Lloyd's method satisfies neither, or does not satisfy both?  please clarify
+and consequently by (kernel) $k$-means algorithm.
 The next result indicates that Hartigan's can potentially 
@@
 Algorithm~\ref{algo} to standard $k$-means and GMM/EM since 
-these are the most used
+these are reference
 clustering algorithms in practice.
@@
 indicating the standard error. Whenever possible, 
+% @gui: bayes error is always possible to numerically estimate, no? maybe sometimes computationally intractable?
 we also indicate the optimal 
@@
 
-From the results of \cite{Telgarsky}, the improvement of Hartigan's 
-over Lloyd's method should be more accentuated in high dimensions.
-Thus, let us analyse
+From the results of \cite{Telgarsky}, we expect the improvement of Hartigan's 
+over Lloyd's method to be more accentuated in high dimensions.
+Thus, we analyze
 how the algorithms degrade as the number of dimensions increase, while

### 2017-09-17 qcqp_clustering/draft/energy_clustering_v2.tex

--- 2017-09-07 qcqp_clustering/draft/energy_clustering.tex
@@
 was proposed.
-Although  
-extensively used in statistics, it
-has not been previously incorporated in machine learning problems.
+%Although  
+%extensively used in statistics, it
+%has not been previously incorporated in machine learning problems.
 % @gui: not sure what you mean?
@@
 even then, it is not true, both the master's thesis and the RKHS paper kind of do it.
-be as general as possible about the gap that we are filling, but not more general.
+be as general as possible about the gap that we are filling, but not more general
+% @jovo: I think we should just remove the phrase. 
 Here we consider the problem of clustering data from
@@
 % @gui: have other people shown a relationship between kernel k-means and say, RKHS, or something else? 
+% @jovo: yes, any kernel method is defined on a HKHS. 
+% For instance ref [11] is the first one to consider kernel k-means in a 
+% RKHS. The whole problem is how do you formulate a 
+% statistical problem in a RKHS. In the case of
+% energy this was done only by [6], and it's based on this work that we 
+% have the theory presented in this paper
 Moreover, we also consider a weighted version of energy statistics 
@@
 % @gui: algorithms don't make assumptions, theorems do.  also, algorithms cannot be parametric or non-parametric, that is a property only of the statistical model, or an estimator. 
-Moreover, $k$-means
-is strongly parametric, assuming that data is isotropically and normally
-distributed since it is a special case of gaussian mixture models (GMM) 
+%Moreover, $k$-means
+%is strongly parametric, assuming that data is isotropically and normally
+%distributed since it is a special case of gaussian mixture models (GMM) 
 % @gui: k-means is NOT a special case of GMM, GMM is a soft clustering (not fuzzy, in stats and machine learning we don't say "fuzzy"), where as k-means is a hard clustering.
 k-means is certainly closely related to GMM though.
-through the expectation maximization (EM) algorithm. Nevertheless,
+%through the expectation maximization (EM) algorithm. 
+Nevertheless,
 $k$-means works very well when data is linearly separable 
-in Euclidean space, with a clear resolution of cluster centers and without
-significant overlap of the points belonging to different clusters.
-% i@gui:  don't know what "clear resolution of cluster centers" adds to this?
+in Euclidean space.
+%, with a clear resolution of cluster centers and without
+%significant overlap of the points belonging to different clusters.
+% @gui:  don't know what "clear resolution of cluster centers" adds to this?
+% @jovo: commented out some parts above and included new phrase here.
+Gaussian mixture models (GMM) is also very commonly used for 
+clustering,
+however it is strongly parametric, as $k$-means which is closely related.
 
@@
 Hilbert
-space by using its inner product. This enables one to cluster way more
-complicated datasets, 
+space by using its inner product. 
+%This enables one to cluster way more complicated datasets, 
+This enables one to cluster data that is not separable in Euclidean space.
 % @gui: algorithms cluster, doesn't matter how complicated the data are.  so, what do you actually mean here?  perform well empirically? what does "complicated" actually mean?
-however, the kernel choice remains the biggest challenge
-since there is no principled theory behind this, 
+% @jovo: what I mean is not tied to any algorithm. I want to say that kernel
+% methods allows one to cluster data that is not separable in Euclidean space.
+% I modified the phrase, see if it's better
+However, the kernel choice remains the biggest challenge
+since there is no principled theory to construct a kernel for a given dataset,
+%behind this, 
 % @gui: the words "this" and "it" are often vague and ambiguous, be specific/precise
-and it usually introduces
-extra parameters that need to be optimized. 
+and usually a kernel introduces
+hyperparameters that need to be carefully chosen.
 % @gui: not sure what you mean here, in kernel k-means, we don't "optimize" over any parameters typically.  sometimes we select hyperparameters of kernels via cross-validation, is that what you mean?
+% @jovo: yes this is what I mean, tried to fix it
 The well-known
@@
 We provide a precise mathematical formulation to this
-statement, showing that there is only one possible optimization problem
-consistent with energy statistics. 
+statement
+%, showing that there is only one possible optimization problem
+%consistent with energy statistics. 
 % @gui: not sure what you mean "only one possible optimization problem".  we can reformulate it many different ways? also, below, "this" is ambiguous.
-Moreover, we show that this is
+% @jovo: I had in mind that minimizing W is equivalent to maximizing S, 
+% but it's confusing at this point so just removed the phrase
+leading to
 a quadratically constrained
@@
 establish the connection 
-to kernel methods used in machine learning, 
+to kernel methods commonly used in machine learning, 
 by showing that this QCQP is equivalent
@@
 default kernels, showing that
-it is able to perform accurately on datasets coming very different 
+it is able to perform accurately on datasets coming from very different 
 distributions, contrary to $k$-means and GMM for instance.
@@
 in this paper, we make clear exactly when we expect energy clustering to perform better than k-means/GMM, and when we expect it to perform worse (empirically).
+% @jovo: added this phrase, check it
+More specifically, the proposed method performs closely to $k$-means
+and GMM on normally distributed data with balanced clusters, however, it
+performs considerably better on  data that is not normally distributed. 
+It also
+performs better than $k$-means and GMM in higher dimensions, even on 
+gaussian settings, and it performs worse than 
+GMM when we have highly unbalanced clusters.
 
 % @gui: i find ToC paragraphs useless.
+% @jovo: I agree and don't mind removing, however it's a tradition that 
+% most referees like to see, and it doesn't do any harm. I'm fine with
+% whatever choice
 Our work is organized as follows. In section~\ref{sec:background} we introduce
@@
 % @gui: let the reader decide what is important, rather than us telling them :)
-It is important to 
-mention that \eqref{eq:energy2} can be even further generalized.
+%It is important to mention 
+The energy distance \eqref{eq:energy2} can be further generalized.
 Let $X, Y \in \mathcal{X}$  where $\mathcal{X}$ is an arbitrary space endowed
@@
 Due to \eqref{eq:mmd} we can compute the energy distance using the inner 
-product of $\Hk$. Moreover, this can be estimated from a sample
-according to \eqref{eq:inner_data}.
+product of $\Hk$. 
+%Moreover, this can be estimated from a sample
+%according to \eqref{eq:inner_data}.
 % @gui: i don't know what this means.  eqn 7 defines something for a population statistic, not a sample statistic.
+% @jovo: removed. I was thinking that you can always estimate the expected 
+% value by computing a sample mean. but I agree, it's confusing
 
@@
 % @gui: i'm not sure the above is true/meaningful.  nonparametric means implicitly estimating an infinite number of parameters.  without any statistical theory, i'm not quite sure what we can say.
+% @jovo: my understanding is that this hypothesis test does not assume any
+% form of the distribution. For instance, Kolmogorov-Smirnov is also
+% said to be nonparametric
 When the sample size is large enough, $n\to \infty$,
@@
 % @gui: consider only using equation numbers for equations that you later refer to? there are so many, it is easy to get lost.
+% @jovo: ok will do that after we are done revising the paper. This is a matter
+% of taste, some people enumerate all equations, others just the ones cited
 Due to the test statistic \eqref{eq:between} for equality of distributions,
@@
 straightforward result
-shows that maximizing \eqref{eq:between} is, however, equivalent to minimizing
-\eqref{eq:within} which has a more convenient form.
+shows that maximizing $S$ is, however, equivalent to minimizing
+$W$ which has a more convenient form.
 % @gui: where possible, i recommend using the "name" of the quantity rather than the number of the equation.  eg, here, write S and W, rather than (13) and (12).
+% @jovo: i think this is fine for the main symbols, will do
 
@@
 $\mathbb{X} = \bigcup_{j=1}^k \C_j$, where $\C_i \cap C_j = \emptyset$ for
-all $i\ne j$, maximizes \eqref{eq:between} if and only if
+all $i\ne j$, maximizes $S$ if and only if
 \begin{equation}
@@
 \min_{\C_1,\dotsc,C_k  } W(
-\C_1, \dotsc, \C_k),
-\end{equation}
-where $W$ is given by \eqref{eq:within}.
+\C_1, \dotsc, \C_k).
+\end{equation}
 \end{proposition}
@@
 % @gui: do other people know this result? if so, cite them, if not, state that.
+% @jovo: eq (15), i.e. S + W = T is well-known. Gabor and Rizzo wrote that
+% in their paper, however, they didn't write a proof. They should have
+% in some other paper, but I didn't bother looking and just did my own
+% way.
+% However, they mention 
+% that this holds only for 0 < alpha < 2, with |x-y|^alpha, while
+% here I did for a general semimetric. The PhD thesis do not 
+% mention anywhere that
+% minimizing W is equivalent to maximizing S, so I didn't see this written
+% before. Anyway, this proposition is trivial, no big deal, but it's
+% necessary to state it clearly for completeness
 
 For a given $k$, the clustering problem amounts to
-finding the best partition of the data by solving~\eqref{eq:minimize}.
+finding the best partition of the data by minimizing $W$.
 % @gui: here and elsewhere is another good example.  why not just say "by minizing W"?
@@
 % @gui: "Let us" is used a bunch, though i think it never adds anything? i'd remove it everywhere.
-Let us formulate \eqref{eq:minimize} in the corresponding
+% @jovo: will remove them
+We now formulate \eqref{eq:minimize} in the corresponding
 RKHS. Based on \eqref{eq:kernel_semimetric} and \eqref{eq:gen_kernel}, 
 assume that the kernel $\kk: \mathcal{X} \times \mathcal{X} \to \mathbb{R}$ 
-generates $\rho$.  Let us define  the Gram matrix
+generates $\rho$.  Define  the Gram matrix
 \begin{equation}
@@
 $D = \diag( n_1,\dotsc, n_k )$  contains
-the number of points in each cluster. Let us also introduce the rescaled
+the number of points in each cluster. We also define the rescaled
 matrix  $Y \equiv Z D^{-1/2}$. In component form they are given by
@@
 % @gui: i don't understand the notaiton (.)_jj. is a sum missing?
+% @jovo: this is standard.  A_{ij} means the element
+% $(i,j)$ of matrix A as you know. 
+% Since here A = Z^T G Z is a product of 3 matrices
+% we just put a parenthesis to avoid confusion when denoting the (j,j) 
+% element of the product
 where we used the definitions \eqref{eq:kernel_matrix} and
@@
 % @gui: the above step is not obvious
+% @jovo: I thing you got confused with the notation
+% from (21) we have \sum_j 1/n_j (Z^T G Z)_jj. But 1/n_j is just
+% D^{-1}_jj (the diagonal matrix. The sum over j goes over the diagonals of
+% D^{-1} Z^T G Z, which is the own definition of the trace
 Now we can
@@
 constraint in \eqref{eq:qcqp}. In matrix form this reads
-$Z^T \e = D \e$. Replacing $Z=YD^{1/2}$ we have
+$Z^T \e = D \e$, where $e$ is the all-ones vector. 
+Replacing $Z=YD^{1/2}$ we have
 $Y^\top \e = D^{1/2} \e$. Multiplying this last equation
@@
 problem \eqref{eq:qcqp} is equivalent
-to \eqref{eq:qcqp2} .
+to \eqref{eq:qcqp2}.
 \end{proof}
@@
 % @jovo: stopped here.
+% @gui: stopped here as well.
 

### 2017-09-17 qcqp_clustering/draft/energy_clustering.tex
continue commenting

@@
 
-\title{Nonparametric Clustering from  Energy Statistics}
+%\title{Nonparametric Clustering from  Energy Statistics}
+\title{Energy Clustering}
 
@@
 \sum_{i=1}^k \lambda_i$. One might then normalize and threshold the rows
-of $Y^\star$, or better, following \cite{NgJordan} one can normalize the
+of $Y^\star$,
+% @gui: this might be a better initialization than k-means++, which essentially initializes using a different kernel?
+ or better, following \cite{NgJordan} one can normalize the
 rows of $Y^\star$ and apply standard $k$-means on this matrix where each
@@
 nonnegative.
-
-% @jovo: stopped here.
 
@@
 is based on Lloyd's method. 
+% @gui: i might call them \mathcal{E}-H & \mathcal{E}-L, or some such?
 From the discussion in the end of the previous section we can anticipate
@@
 accuracy of $\approx 0.86$.
-For each $D$ we sample $100$ times, picking $100$ points for each cluster.
+For each $D$ we generate $100$ Monte Carlo runs, sampling $100$ points for each cluster.
+% @gui: if you sample 100 points from each cluster, it breaks the GMM model, which assumes you sample n_i points from cluster i with probability \pi_i.
+i don't think it matters that much, but good to know :)
 We apply each algorithm to the resulting dataset 
@@
 and in particular it outperforms kernel $k$-means algorithm as the number
-of dimensions increase. Note also that although GMM and $k$-means are
+of dimensions increase. 
+% @gui: when you say "kernel k-means" here, do you mean algorithm 2?  if so, please clarify, if not, please clarify :)
+Note also that although GMM and $k$-means are
 parametric they are consistent models to the data.
+% @guiL gmm & k-means are not 'models". they are algorithms, or in this context, they are estimators.
 
@@
 The optimal accuracy from Bayes error in (b--d) is $\approx 1$.
+% @gui: titles for each, or subcaptions, would be extremely useful.  in general, titles should provide the "setting" of the experiment, ie, the model, etc.
+% @gui: also helpful here would be pairs plots, say, of the first 5 dimensions, showing the performance of our stuff (color points by Alg 3, maybe symbols denote "truth".
+i have no intuition how difficult these are.
 }
@@
 Let us now consider the same experiment as before with data distributed as
+% @gui: what do you mean "the same experiment as before"? clarify, or don't say that. there are obviously differences. 
 \begin{equation}
@@
 and we choose $q \in \{ 1/2, 1, 2\}$. In these cases,
+% @gui: what is delta_ij??? can you explain the intuition for this setting, i don't get it, maybe because i don't know what delta_ij is.....
 the best possible accuracy from Bayes classification error is $\approx 1$.

### 2017-09-17 qcqp_clustering/draft/energy_clustering.tex
finished round of comments

@@
 in gaussian settings.
+% @gui: gaussian is always captilized
+% @gui: also, i don't understand this plot entirely, are you really starting with D=1?
+or 10?
+from the equations, it would seem that Bayes error changes when D< 10, but your Bayes error lines do not seem to change?
 We compute the average of \eqref{eq:accuracy} over $100$ samples with
@@
 increase.
+% @gui: is there any reason to initialize alg 3 with alg 2? is it even worth trying?
 
@@
 and  more rapidly than GMM. This indicates that a fuzzy version of
+% @gui: in stats, we don't say "fuzzy", we say "soft".
+also, why do you think that when clusters are imbalanced that soft clustering is more important?
+i would say that soft clustering is more important when the cluster centers are close, i don't see how imbalanced matters.
 Algorithm~\ref{algo} should compensate for this effect.
@@
 \eqref{eq:standard_metric}, 
+% @gui: please remove all "let us" from manuscript
 let us consider two other semimetrics with their respective generating
@@
 Algorithm~\ref{kmeans_algo} for any of the kernel choices.
-%Standard $k$-means performs poorly in this example.
+% @gui: this is a separate point, that i would make a separate figure for if you want to make it.
+then, one fig would show E-H with 3 kernels vs.
+kmeans & gmm, and then another would show the difference between accuracy of E-H vs E-L for each of the 3 kernels.
 
@@
 while $k$-means works almost at chance,
-and GMM is not even able to estimate the parameters. Again, the
+and GMM is not even able to estimate the parameters. 
+% @gui: something is weird with the way you care computing accuracy.  GMM should not be significantly worse than chance.  i think you are probably counting the NaN's as *wrong*, rather than nothing?
+Again, the
 kernel \eqref{eq:rhoe}
@@
 \label{fig:consist}
+% @gui: too many lines, you demonstrated that E-H is better than E-L already, i don't think we need to show the E-L variants here. 
 Algorithm~\ref{algo} and Algorithm~\ref{kmeans_algo} with kernels
@@
 sample $400$ points for each class.
-We apply Algorithm~\ref{algo} and Algorithm~\ref{kmeans_algo}, besides
-$k$-means and GMM, using the above kernels 
-\eqref{eq:rhoalpha}--\eqref{eq:rhogsigma}. The results are shown
+We apply Algorithm~\ref{algo} and Algorithm~\ref{kmeans_algo}, using the above kernels 
+\eqref{eq:rhoalpha}--\eqref{eq:rhogsigma}, as well as
+$k$-means and GMM. The results are shown
 in Table~\ref{table:other} where the respective choice of parameters for the
@@
 Fig.~\ref{fig:other}b we see
-that the choice of kernel is very sensitive, and only \eqref{eq:rhoesigma}
+that the choice of kernel is very sensitive, 
+% @gui: you mean performance is sensitive to the kernel?
+and only \eqref{eq:rhoesigma}
 was able to cluster accurately. The same kernel choice to the case
-of Fig.~\ref{fig:other} still provides better results than
+of Fig.~\ref{fig:other} 
+% @gui: Fig 5 a? 5c?
+still provides better results than
 the others, but they are less accurate compared to Fig.~\ref{fig:other}b.
@@
 Clustering the data shown in Fig.~\ref{fig:other} with Algorithm~\ref{algo}
-and Algorithm~\ref{kmeans_algo}, besides $k$-means and GMM, with kernels 
-\eqref{eq:rhoalpha}--\eqref{eq:rhogsigma}. 
+and Algorithm~\ref{kmeans_algo}, with kernels 
+\eqref{eq:rhoalpha}--\eqref{eq:rhogsigma}, as well as $k$-means and GMM. 
 We sample $10$ times and show the average
@@
 which is computed from a sample  $\{ x_i \}_{i=1}^n$.
+% @gui: it is cheating if this sample is part of the data, this should be a sample of "training data", not used for subsequent testing. please clarify
 We consider subsets of the  classes $\{0,1,\dotsc,9 \}$, 
@@
 \end{tabular}
+% @gui: i'm not convinced you need alg 2 in these tables at all.
+if i were to add alg 2 to these tables, however, i would do the same thing i would for the figs, that is, for each simulation, compute the diff between performance on alg 2 and alg 3, and report that.
 \caption{\label{table:mnist}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Conclusion}
+\section{Discussion}
 \label{sec:conclusion}
@@
 Our results imply that kernel $k$-means
-is actually a consequence of energy statistics, and thus place this 
+can be interpreted as  a consequence of energy statistics, and thus place this 
+% @gui: "this"? what is "this:? clarify
 method into a principled statistical basis.
@@
 
+% @gui: your discussion is basically just a summary, which i don't find very helpful to the reader.
+if you want to summarize results, 1 paragraph is sufficient.
+another paragraph connects it to previous work not already mentioned, and a third paragraph on additional possible extensions.
+as you know, one thing i'd like to do next is to theoretically prove when Energy Clustering asymptoticaly gets the right answer.
+you also mentioned soft clustering, which is another potential extension, or incorporating this into subspace clustering.
+
 \subsection*{Acknowledgements}
-GF would like to thank Carey Priebe for discussions.
-This work was supported by XXX grant.
+We would like to thank Carey Priebe for discussions.
+This work was supported by NIH TRA  grant.
 

### 2017-09-18 qcqp_clustering/draft/energy_clustering.tex
added comment from cep
first pass - revision

@@
 \documentclass[aps,preprint,nofootinbib,floatfix]{revtex4-1}
-%\documentclass{article}
-
-%\usepackage{nips_2017}
-
-% to compile a camera-ready version, add the [final] option, e.g.:
-%\usepackage[final]{nips_2017}
 
@@
 %\usepackage{bbm}
-
 \usepackage[pdftex]{graphicx}
 \graphicspath{{./figs/}}
-
 \usepackage{algorithmic}
@@
 
-%\title{Nonparametric Clustering from  Energy Statistics}
 \title{Energy Clustering}
@@
 be as general as possible about the gap that we are filling, but not more general.
+% @jovo: commented out, agree
 Here we consider the problem of clustering data from
@@
 % @gui: have other people shown a relationship between kernel k-means and say, RKHS, or something else? 
+% @jovo: any kernel method is defined on a HKHS. 
+% For instance ref [11] is the first one to consider kernel k-means in a 
+% RKHS. The whole problem is how do you formulate a 
+% statistical problem in a RKHS, or how do you know if there is a kernel
+% related to it. In the case of energy this was done by [6], and it's based 
+% on this work that we have the theory presented in this paper
 Moreover, we also consider a weighted version of energy statistics 
@@
 % @gui: algorithms don't make assumptions, theorems do.  also, algorithms cannot be parametric or non-parametric, that is a property only of the statistical model, or an estimator.
-Moreover, $k$-means
-is strongly parametric, assuming that data is isotropically and normally
-distributed since it is a special case of gaussian mixture models (GMM) 
+% @jovo: commented out below and included a phrase in the end
+%Moreover, $k$-means
+%is strongly parametric, assuming that data is isotropically and normally
+%distributed since it is a special case of gaussian mixture models (GMM) 
 % @gui: k-means is NOT a special case of GMM, GMM is a soft clustering (not fuzzy, in stats and machine learning we don't say "fuzzy"), where as k-means is a hard clustering.
 k-means is certainly closely related to GMM though.
-through the expectation maximization (EM) algorithm. Nevertheless,
+%through the expectation maximization (EM) algorithm. 
+Nevertheless,
 $k$-means works very well when data is linearly separable 
-in Euclidean space, with a clear resolution of cluster centers and without
-significant overlap of the points belonging to different clusters.
-% i@gui:  don't know what "clear resolution of cluster centers" adds to this?
+in Euclidean space.
+%, with a clear resolution of cluster centers and without
+%significant overlap of the points belonging to different clusters.
+% @gui:  don't know what "clear resolution of cluster centers" adds to this?
+% @jovo: commented above and included this phrase instead
+Gaussian mixture models (GMM) is also very commonly used for
+clustering,
+however it is strongly parametric, as $k$-means which is closely related.
 
@@
 Hilbert
-space by using its inner product. This enables one to cluster way more
-complicated datasets, 
+space by using its inner product. 
+%This enables one to cluster way more complicated datasets, 
 % @gui: algorithms cluster, doesn't matter how complicated the data are.  so, what do you actually mean here?  perform well empirically? what does "complicated" actually mean?
-however, the kernel choice remains the biggest challenge
-since there is no principled theory behind this, 
+% @jovo: what I mean is not tied to any algorithm. I want to say that kernel
+% methods allows one to cluster data that is not separable in Euclidean space.
+% I modified the phrase, see if it's better
+However, the kernel choice remains the biggest challenge
+since there is no principled theory to construct a kernel for a given
+dataset,
 % @gui: the words "this" and "it" are often vague and ambiguous, be specific/precise
-and it usually introduces
-extra parameters that need to be optimized. 
+and usually a kernel introduces hyperparameters that need to be carefully
+chosen.
+%extra parameters that need to be optimized. 
 % @gui: not sure what you mean here, in kernel k-means, we don't "optimize" over any parameters typically.  sometimes we select hyperparameters of kernels via cross-validation, is that what you mean?
+% @jovo: tried to fix above, you're right in what I meant
 The well-known
@@
 We provide a precise mathematical formulation to this
-statement, showing that there is only one possible optimization problem
-consistent with energy statistics. 
-% @gui: not sure what you mean "only one possible optimization problem".  we can reformulate it many different ways? also, below, "this" is ambiguous.
-Moreover, we show that this is
+statement 
+%showing that there is only one possible optimization problem
+%consistent with energy statistics. 
+% @gui: not sure what you mean "only one possible optimization problem".  we
+% can reformulate it many different ways? also, below, "this" is ambiguous.
+% @jovo: I had in mind that minimizing W is equivalent to maximizing S, 
+% but it's confusing at this point so just removed the phrase and adapted
+%Moreover, we show that this is
+leading to
 a quadratically constrained
@@
 in this paper, we make clear exactly when we expect energy clustering to perform better than k-means/GMM, and when we expect it to perform worse (empirically).
+% @jovo: added this phrase, check it
+More specifically, the proposed method performs closely to $k$-means
+and GMM on normally distributed data with balanced clusters, however, it
+performs considerably better on  data that is not normally distributed.
+It also
+performs better than $k$-means and GMM in higher dimensions, even on
+gaussian settings, and it performs worse than
+GMM when we have highly unbalanced clusters.
 
 % @gui: i find ToC paragraphs useless.
+% @jovo: I agree and don't mind removing, however it's a tradition that 
+% most referees like to see and it doesn't do any harm. I'm fine with
+% whatever choice
 Our work is organized as follows. In section~\ref{sec:background} we introduce
@@
 % @gui: let the reader decide what is important, rather than us telling them :)
-It is important to 
-mention that \eqref{eq:energy2} can be even further generalized.
+% @jovo: removed that
+The energy distance \eqref{eq:energy2} can be even further generalized.
 Let $X, Y \in \mathcal{X}$  where $\mathcal{X}$ is an arbitrary space endowed
@@
 % @gui: consider only using equation numbers for equations that you later refer to? there are so many, it is easy to get lost.
+% @jovo: ok, will do that after we finish everything.
 Due to the test statistic \eqref{eq:between} for equality of distributions,
@@
 % @gui: where possible, i recommend using the "name" of the quantity rather than the number of the equation.  eg, here, write S and W, rather than (13) and (12).
+% @jovo: ok, I will do that for the main symbols
 
@@
 % @gui: do other people know this result? if so, cite them, if not, state that.
+% @jovo: eq (15), i.e. S + W = T is well-known. Gabor and Rizzo wrote that
+% in their paper, however, they didn't write a proof. They should have
+% in some other paper I believe (will double check)
+% However, they mention 
+% that this holds only for 0 < alpha < 2, with |x-y|^alpha, while
+% here I did for a general semimetric. The PhD thesis do not 
+% mention anywhere that
+% minimizing W is equivalent to maximizing S, so I didn't see this written
+% before. Anyway, this result is trivial, but important.
 
 For a given $k$, the clustering problem amounts to
-finding the best partition of the data by solving~\eqref{eq:minimize}.
+finding the best partition of the data by minimizing $W$.
+%solving~\eqref{eq:minimize}.
 % @gui: here and elsewhere is another good example.  why not just say "by minizing W"?
+% @jovo: ok
 Notice that this is a hard clustering problem as partitions
@@
 % @gui: "Let us" is used a bunch, though i think it never adds anything? i'd remove it everywhere.
-Let us formulate \eqref{eq:minimize} in the corresponding
+% @jovo: ok
+%Let us
+We now formulate \eqref{eq:minimize} in the corresponding
 RKHS. Based on \eqref{eq:kernel_semimetric} and \eqref{eq:gen_kernel}, 
@@
 % @gui: i don't understand the notaiton (.)_jj. is a sum missing?
+% @jovo: this is standard.  A_{ij} denotes the 
+% $(i,j)$th element of matrix A as you know. 
+% Since here A = Z^T G Z is a product of 3 matrices
+% we just put a parenthesis to avoid confusion, that's all
 where we used the definitions \eqref{eq:kernel_matrix} and
@@
 % @gui: the above step is not obvious
+% @jovo: It is obvious. 
+% From (21) we have \sum_j 1/n_j (Z^T G Z)_jj. But 1/n_j is just
+% D^{-1}_jj (the diagonal matrix). The sum over j goes over the diagonals of
+% D^{-1} Z^T G Z which is the definition of the trace
 Now we can
@@
 % @gui: Y is a matrix? or a vector? when introducing new terms, always define what space they are in rigorously
+% @jovo: it's obviously a matrix
+% I didn't want to write Y \in \mathbb{R}^{blabla} because it polutes the
+% text too much and it's dimension is obviously the same as Z. Y was defined
+% before also
 The objective function immediately becomes
@@
 % @gui: don't use equation numbers here.
+% @jovo: in these cases there is no way, I have to. Look at the size
+% of the expressions. I agree to not put reference to eq numbers
+% for the main symbols, like S and W, but for large expressions we have to.
+% I also try to avoid poluting the text with unnecessary math symbols, 
+% it looks pretty bad
 \end{proof}
 
-% @jv: stopped here
 Based on Proposition~\ref{th:qcqp2}, to group data
@@
 % @gui: this might be a better initialization than k-means++, which essentially initializes using a different kernel?
+% @jovo: yes it might be, and I agree, k-means++ is initializing in "Euclidean
+% space" not in the feature space where data might be better separable. Maybe I
+% need test this numerically
 or better, following \cite{NgJordan} one can normalize the
@@
 a semimetric $\rho$, and thus the kernel \eqref{eq:kernel_semimetric}, are
-assumed to be known. This clustering method is
-\emph{nonparametric} since it does not make any assumptions
+assumed to be known. 
+This clustering 
+%method is
+formulation from energy statistics is 
+\emph{nonparametric} in the sense that it does not make assumptions
 about  the distribution of the data, 
-contrary to $k$-means and GMM, for example.
+contrary to $k$-means and GMM formulations, for example.
 % @gui: algorithms don't make assumptions, proofs do.  we don't prove anything about when we get the right answer, so we can't quite say this.
+% @jovo: I adapted above, see if it makes sense, if not I just 
+% remove the phrase
 If one uses the standard energy distance
@@
 
+%The above result shows that 
+%kernel $k$-means problem is actually a consequence
+%of energy statistics. 
 The above result shows that 
-kernel $k$-means problem is actually a consequence
+kernel $k$-means problem is equivalent to the clustering problem
+formulated in the energy statistics framework.
 % @gui: i'm not sure about this.  who is to say the correct direction of causality? maybe energy clustering is a special case of kernel k-means?
-of energy statistics. In this vein, kernel $k$-means is part of a
+% @jovo: Let me describe the way I see it, and see if you agree.
+% I adapted the phrase above.
+% There is a theory about energy statistics. Semimetric spaces, RKHS, test
+% statistic, S + W = T, etc. This theory can be used for statistical inference
+% in general, not only for clustering. kernel k-means is an heuristic 
+% formulation for clustering where
+% one minimizes the objective function J, in eq (24), nothing more than that.
+% It turns out that min J = min W, so the clustering formulation from energy
+% statistics and kernel k-means are equivalent. It could have happened that
+% they are not, energy could possibly provide something different a priori, we
+% didn't know that. Moreover, in energy statistics we know that the kernel
+% must come from a semimetric of negative type, etc. there is a whole
+% construction of it
+In this vein, kernel $k$-means is part of a
 statistical framework where distances between probability
@@
 % @gui: can you provide some guidance here, in particular, what do the weights "mean" or "correspond to"? right now, it feels like just math.
+% @jovo: ok, added the phrase below
+% let me know if not enough. In practice the weights can be anything defined
+% by the "user".
+Above, the weight $w_i = w(i)$ of node $i\in \mathcal{V}$ can be, for instance,
+its degree $w_i = d(i)$.
 
@@
 % @gui: "does not require any initiializaton" i think you mean is guaranteed to find the optimal solution in O(n^2) time.
+% @jovo: it does not require initialization in the sense that we do not
+% assign points to clusters in an initial step, as k-means++ for instance, or
+% random initialization. This algorithm is "exact" in the sense that nothing
+% is random in it, and multiple runs will always give the same answer
 
@@
 %255,255,51];% yellow
+% @jovo: ok, I will change the colors later, will have to regenerate the plots.
+% The problem with generating the kernel density for log-normal is that we need
+% a lot of points to get a curve like the one in the histogram. If I do this
+% with only 3000 points as in the graph, I will get a pretty bad curve that I
+% don't think will look like anything the histogram. Also, this is only a
+% simple example in 1D ... do you think it worth the effort?
+% If you do I can try.
 \label{fig:1d}
@@
 % @gui: i see why you call this algorithm 1, but can we call it something with semantic meaning? like 1D-\mathcal{E}-clustering? or maybe \mathcal{E}1 for short? same for the other variants.
-GMM through EM algorithm, and standard $k$-means 
+% @jovo: will call energy clustering, or $\mathcal{E}-clustering$ is name is to
+% big when I regenerate the plots with new figures
+GMM through EM algorithm, and $k$-means++ 
 algorithm. 
 % @gui: there are many k-means algorithms, be specific.
-In the last two cases 
-we use the initialization from $k$-means++ \cite{Vassilvitskii} 
-and we run the algorithms
-few times choosing the result with  best objective function value. 
+% @jovo: there is only one k-means no? Anything else I would call kernel
+% k-means. I called $k$-means++ though
+We use the initialization from $k$-means++ \cite{Vassilvitskii} also
+for GMM.
+We run the algorithms $100$
+times choosing the result with  best objective function value. 
 % @gui: a few? be specific
+% @jovo: ok
 Notice that Algorithm~\ref{algo1d} requires no initialization so we only
 % @gui: "no initiazation" is weird.  do any algs require an initialization?
+% @jovo: I guess so. In this algorithm we just compute something "exactly",
+% contrary to random initializing and iterating to optimize the solution.
+% There is no random initialization in this algo
 run it once.
@@
 % @gui: why not use the same parameters as in fig 1a?  i feel like if we did, it would seem less like we are cherry picking?  
+% @jovo: yes, and GMM would do great! In 1D is very hard to find setting where
+% GMM and k-means perform poorly, so we have to cherry pick. In 1D you
+% basically do not have much choice, if data has minimal separation k-means and
+% GMM does well.
+% The purpose is to
+% illustrate that there are settings where k-means/GMM do poorly and energy
+% does not. If I choose the same parameters as before all algorithms are doing
+% great, sometimes GMM can even do better than energy.
 The optimal classification accuracy from Bayes error is $\approx 0.852$.
 % @gui: computed numerically?
+% @jovo: yes!
 In this case,
@@
 % @gui: let's call this something more informative
+% @jovo: alright, will take care when re-generating the plots
 It can be shown that this algorithm converges provided $G$ is positive
@@
 
-Let us mention some important results about Hartigan's method.
+In the following mention some important results about Hartigan's method.
 % @gui: "let us" again...
+% @jovo: ok
 
@@
 
-Both conditions of the  above theorem are not satisfied 
-by Lloyd's method, 
+%Both conditions of the  above theorem are not satisfied 
+%by Lloyd's method, 
+Neither of these two conditions are satisfied  by Lloyd's method,
 % @gui: do you mean Lloyd's method satisfies neither, or does not satisfy both?  please clarify
+% @jovo: ok
 and consequently by (kernel) $k$-means algorithm.
@@
 force Hartigan's method to iterate, contrary
-to Lloyd's method. Therefore, Hartigan's method 
+to Lloyd's method. Therefore, Hartigan's 
 can escape local
-optima of Lloyd's method. 
+optima of Lloyd's. 
 Moreover, this improvement should be more prominent as
@@
 % @gui: i might call them \mathcal{E}-H & \mathcal{E}-L, or some such?
+% @jovo: it might be a good idea, let's decide when I re-do the plots
 From the discussion in the end of the previous section we can anticipate
@@
 % @gui: bayes error is always possible to numerically estimate, no? maybe sometimes computationally intractable?
+% @jovo: true. I just didn't indicate Bayes error for Fig. 3 for instance cuz
+% will be very boring, will have to compute for every number of points, and
+% also we don't do that for MNIST and circles, thus the phrase
 we also indicate the optimal 
@@
 accuracy of $\approx 0.86$.
-For each $D$ we generate $100$ Monte Carlo runs, sampling $100$ points for each cluster.
+For each $D$ we generate $100$ Monte Carlo runs, 
+sampling $100$ points for each cluster.
 % @gui: if you sample 100 points from each cluster, it breaks the GMM model, which assumes you sample n_i points from cluster i with probability \pi_i.
 i don't think it matters that much, but good to know :)
+% @jovo: humm ... true, didn't think about that. Do you think I should re-do
+% all experiments in this way? sampling from clusters with prob 1/2 for
+% instance, instead of exactly picking a fixed number of points for each?
 We apply each algorithm to the resulting dataset 
@@
 when the number of dimensions exceeds the number of points in each cluster,
-giving zero accuracy when dimensions are $\gtrsim 100$.
+%giving zero accuracy when dimensions are $\gtrsim 100$.
+i.e. when $D \gtrsim 100$.
 We see that Algorithm~\ref{algo} performs better than all the other ones,
-and in particular it outperforms kernel $k$-means algorithm as the number
+and in particular it outperforms kernel $k$-means, Algorithm~\ref{kmeans_algo},
+as the number
 of dimensions increase. 
 % @gui: when you say "kernel k-means" here, do you mean algorithm 2?  if so, please clarify, if not, please clarify :)
-Note also that although GMM and $k$-means are
-parametric they are consistent models to the data.
-% @guiL gmm & k-means are not 'models". they are algorithms, or in this context, they are estimators.
+% @jovo: will change again when we introduce algorithm's names
+
+%Note also that although GMM and $k$-means are
+%parametric they are consistent models to the data.
+Note that in this case GMM and $k$-means are consistent estimators.
+
+% @gui: gmm & k-means are not 'models". they are algorithms, or in this context, they are estimators.
+% @jovo: see if it's correct now. 
+% Don't know how to say this, what I wanted to say is
+% that GMM and k-means are algorithms that were built based on gaussian
+% mixtures models, and they should perform close to optimal when data
+% is normally distributed.
 
@@
 standard $k$-means and GMM/EM as the number of dimensions increase
-in gaussian settings.
+in Gaussian settings.
 % @gui: gaussian is always captilized
@@
 from the equations, it would seem that Bayes error changes when D< 10, but your Bayes error lines do not seem to change?
+% @jovo: I start with D = 10.
+% @gui: cep points out that in Fig 2a, GMM does very poorly even with d small, which does not make sense to us.  can you explain?
+% @jovo: you're right, didn't realize that. Will double check and maybe repeat
+% the experiment.
 We compute the average of \eqref{eq:accuracy} over $100$ samples with
@@
 % @gui: titles for each, or subcaptions, would be extremely useful.  in general, titles should provide the "setting" of the experiment, ie, the model, etc.
+% @jovo: ok will do when I re-generate the plots.
 % @gui: also helpful here would be pairs plots, say, of the first 5 dimensions, showing the performance of our stuff (color points by Alg 3, maybe symbols denote "truth".
 i have no intuition how difficult these are.
+% @jovo: didn't understand this
 }
@@
 
-Let us now consider the same experiment as before with data distributed as
+%Let us now consider the same experiment as before with data distributed as
+Consider now the following setting:
 % @gui: what do you mean "the same experiment as before"? clarify, or don't say that. there are obviously differences. 
+% @jovo: ok
 \begin{equation}
@@
 \end{equation}
-and we choose $q \in \{ 1/2, 1, 2\}$. In these cases,
+and we choose $q \in \{ 1/2, 1, 2\}$. Above $\delta_{ij} = 1$ if $i=j$ and
+$\delta_{ij}=0$ if $i\ne j$ is the Kronecker delta. In these cases,
 % @gui: what is delta_ij??? can you explain the intuition for this setting, i don't get it, maybe because i don't know what delta_ij is.....
+% @jovo: delta_{ij} is the Kronecker delta, I included the definition
 the best possible accuracy from Bayes classification error is $\approx 1$.
@@
 % @gui: is there any reason to initialize alg 3 with alg 2? is it even worth trying?
-
-Let us now
-consider the effect of having 
-unbalanced clusters. We generate data as
+% @jovo: based on the theorems I cited for Hartigan's this is actually worse,
+% since the set of local minima of Hartigan's is a subset of Lloyd's, thus if
+% you do that Hartigan's may not iterate. However, Hartigan's has possibly less
+% local minima than Lloyd's so if you start at random, or not on a local
+% minima it may converge to a better solution
+
+Consider the effect of having 
+unbalanced clusters accordint to
 \begin{equation}
@@
 (error bars are standard error). The results are in Fig.~\ref{fig:unbalanced}.
-As expected, GMM
-works better than the other algorithms since it is a fuzzy clustering.
-The other methods based on hard assignments degrade similarly
-and  more rapidly than GMM. This indicates that a fuzzy version of
+For highly unbalanced clusters we see that GMM performs better than
+the other methods which have similar performance.
+%As expected, GMM
+%works better than the other algorithms since it is a fuzzy clustering.
+%The other methods based on hard assignments degrade similarly
+%and  more rapidly than GMM. This indicates that a soft version of
+%Algorithm~\ref{algo} should compensate for this effect.
 % @gui: in stats, we don't say "fuzzy", we say "soft".
@@
 i would say that soft clustering is more important when the cluster centers are close, i don't see how imbalanced matters.
-Algorithm~\ref{algo} should compensate for this effect.
+% @jovo: you are probably right, I'll remove this phrase. 
+% I thought that this should be the case because in this setting that's pretty
+% much the only difference between GMM and k-means
 
@@
 Besides the standard kernel from energy statistics 
-\eqref{eq:standard_metric}, 
+\eqref{eq:standard_metric} 
 % @gui: please remove all "let us" from manuscript
-let us consider two other semimetrics with their respective generating
-kernels,
+% @jovo: ok, hunting them
+consider the following two other semimetrics with their respective generating
+kernels:
 \begin{align}
@@
 \end{align}
-The kernel \eqref{eq:rhohalf} corresponds to the energy distance
+The kernel $K_{1/2}(x,y)$ corresponds to the energy distance
 \eqref{eq:energy2} with $\alpha=1/2$.
-Now consider data normally distributed in $D=20$ as
+We sample data from the following normal distribution in $D=20$:
 \begin{equation}
@@
 kmeans & gmm, and then another would show the difference between accuracy of E-H vs E-L for each of the 3 kernels.
+% @jovo: ok, point taken, will come together with other plots
 
@@
 % @gui: something is weird with the way you care computing accuracy.  GMM should not be significantly worse than chance.  i think you are probably counting the NaN's as *wrong*, rather than nothing?
+% @jovo: yes, double checking that
 Again, the
@@
 % @gui: too many lines, you demonstrated that E-H is better than E-L already, i don't think we need to show the E-L variants here.
+% @jovo: will fix that
 Algorithm~\ref{algo} and Algorithm~\ref{kmeans_algo} with kernels
@@
 sample $400$ points for each class.
-We apply Algorithm~\ref{algo} and Algorithm~\ref{kmeans_algo}, using the above kernels 
+We apply Algorithm~\ref{algo} and Algorithm~\ref{kmeans_algo}, 
+using the above kernels 
 \eqref{eq:rhoalpha}--\eqref{eq:rhogsigma}, as well as
@@
 results, similar to GMM,
-which is the true model of the data. For the data shown in
+which is a consistent estimator for this data. For the data shown in
 Fig.~\ref{fig:other}b we see
-that the choice of kernel is very sensitive, 
+that the clustering quality is highly sensitive to the choice of kernel,
 % @gui: you mean performance is sensitive to the kernel?
+% @jovo: fixed
 and only \eqref{eq:rhoesigma}
 was able to cluster accurately. The same kernel choice to the case
-of Fig.~\ref{fig:other} 
+of Fig.~\ref{fig:other}c
 % @gui: Fig 5 a? 5c?
+% @jovo: ok
 still provides better results than
-the others, but they are less accurate compared to Fig.~\ref{fig:other}b.
+the other kernels, but the results are less accurate 
+compared to the data in Fig.~\ref{fig:other}b.
 
@@
 % @gui: it is cheating if this sample is part of the data, this should be a sample of "training data", not used for subsequent testing. please clarify
+% @jovo: I'm not sure what to say here, cuz all I'm doing is, given a data set
+% I can compute \sigma as above. Then I use this in the kernel and perform
+% clustering. How can this be cheating if I'm not using any labels, or any
+% other information besides the data I'm clustering?
 We consider subsets of the  classes $\{0,1,\dotsc,9 \}$, 
@@
 if i were to add alg 2 to these tables, however, i would do the same thing i would for the figs, that is, for each simulation, compute the diff between performance on alg 2 and alg 3, and report that.
+% @jovo: will check how the difference looks and change this
 \caption{\label{table:mnist}
@@
 more general kernels related to \eqref{eq:energy3} can be obtained.
-Our results imply that kernel $k$-means
-can be interpreted as  a consequence of energy statistics, and thus place this 
+Our results imply that kernel $k$-means optimization problem
+can be interpreted as a consequence of energy statistics based clustering.
 % @gui: "this"? what is "this:? clarify
-method into a principled statistical basis.
-We also considered a weighted version of energy statistics clustering,
-establishing connections with spectral
+% @jovo: removed, i think it's enough the way it is already
+%thus placing it this 
+%method into a principled statistical basis.
+We also considered a weighted version of energy statistics whose 
+clustering formulation establishes connections with spectral
 clustering and graph partitioning problems.
@@
 you also mentioned soft clustering, which is another potential extension, or incorporating this into subspace clustering.
+% @jovo: not clear how to make the conclusion?
+% how do you think I should modify? A summary is important, because it
+% condenses the results of the paper in one place in a non-technical manner.
+% Besides a summary I don't know what else to say
 

### 2017-09-30 energy_clustering/draft/energy_clustering.tex
initialization issues

--- 2017-09-18 qcqp_clustering/draft/energy_clustering.tex
@@
 
+\usepackage{amsmath,amssymb,amsfonts,amsthm,amscd,bm,bbm}
 %\usepackage{nicefrac}       % compact symbols for 1/2, etc.
 %\usepackage{microtype}      % microtypography
-\usepackage{amsmath,amssymb,amsfonts,amsthm,amscd,bm,bbm}
 %\usepackage{mathtools}
 %\usepackage{printlen}
+%\usepackage{cite}
 \usepackage[inline]{enumitem}
-%\usepackage{cite}
-%\usepackage{bbm}
+\usepackage[inline]{enumitem}
 \usepackage[pdftex]{graphicx}
@@
 \usepackage{algorithm}
-\usepackage[none]{hyphenat}
+%\usepackage[none]{hyphenat}
 \usepackage{multirow}
@@
 
-\hyphenation{op-tical net-works semi-conduc-tor}
+%\pretolerance=10000
+%\tolerance=2000 
+%\emergencystretch=10pt
 
@@
 
-%% our definitions %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \DeclareMathOperator{\aff}{aff}
 \DeclareMathOperator{\st}{s.t.}
-\DeclareMathOperator{\LC}{LC}
 \DeclareMathOperator{\affnot}{aff_0}
@@
 \newcommand\Energy{\mathcal{E}}
+\newcommand\EnergyH{\mathcal{E}^{H}}
+\newcommand\EnergyL{\mathcal{E}^{L}}
+\newcommand\EnergyOne{\mathcal{E}^{1D}}
 \newcommand\E{\mathbb{E}}
@@
 \newcommand\C{{\mathcal{C}}}
+\newcommand\tC{{\widetilde{\C}}}
 \newcommand\OO{{\mathcal{O}}}
-%\newcommand\Zt{\widetilde{Z}}
 \newcommand\Zt{Y}
-%\newcommand{\Ind}[1]{\delta_{#1}}
 \newcommand{\Ind}[1]{\mathbbm{1}_{#1}}
@@
 \email{guifranca@gmail.com} 
-%\affiliation{Johns Hopkins University}
-
 \author{Joshua T. Vogelstein}
 \email{jovo@jhu.edu}
-
 \affiliation{Johns Hopkins University}
@@
 \begin{abstract}
-Energy statistics 
-was proposed by 
-Sz\' ekely in the 80's
-inspired by the Newtonian gravitational potential from classical mechanics,
-and it provides a nonparametric test for equality of distributions.
-It was generalized to 
-metric spaces of strong negative type,
-and more recently a connection with reproducing kernel Hilbert spaces 
-was proposed.
-%Although  
-%extensively used in statistics, it
-%has not been previously incorporated in machine learning problems.
-% @gui: not sure what you mean?
-what counts as machine learning?
-be specific, i think yo mean clustering?
-even then, it is not true, both the master's thesis and the RKHS paper kind of do it.
-be as general as possible about the gap that we are filling, but not more general.
-% @jovo: commented out, agree
-Here we consider the problem of clustering data from
-an energy statistics theory perspective.
-We provide a precise mathematical formulation yielding
-a quadratically constrained quadratic program (QCQP), which we show
-to be equivalent
-to kernel $k$-means optimization problem.
-Thus, our results imply a first principles derivation of 
-kernel $k$-means from energy statistics.
-% @gui: have other people shown a relationship between kernel k-means and say, RKHS, or something else? 
-% @jovo: any kernel method is defined on a HKHS. 
+Energy statistics was proposed by Sz\' ekely in the 80's inspired by the 
+Newtonian gravitational potential from classical mechanics, and it provides 
+a hypothesis test for equality of distributions. Energy statistics was 
+further generalized to data living metric spaces of strong negative type, 
+and more recently a connection with reproducing kernel Hilbert spaces (RKHS) 
+was proposed. Here we consider the problem of clustering data from an 
+energy statistics 
+theory perspective, called energy clustering for short, by providing a 
+precise mathematical formulation yielding a quadratically constrained 
+quadratic program (QCQP) in the associated RKHS, 
+which we show to be equivalent to kernel $k$-means optimization problem.
+Thus, our results imply a first principles derivation of kernel $k$-means 
+from energy statistics.
+Moreover, we also consider a weighted version of energy statistics 
+applied to clustering, making connection to graph partitioning problems.
+To find local optimizers of such QCQP we consider an iterative algorithm based 
+on Hartigan's method, which in this case has the same computational cost 
+as kernel $k$-means algorithm, based on Lloyd's method, but usually 
+with better clustering quality. 
+We provide carefully designed numerical experiments showing the superiority 
+of the proposed method compared to kernel $k$-means, standard $k$-means, 
+and gaussian mixture models in a variety of settings.
+\end{abstract}
+
+% @gui: have other people shown a relationship between kernel k-means and 
+% say, RKHS, or something else? 
+% @jovo: any kernel method is defined on a RKHS. 
 % For instance ref [11] is the first one to consider kernel k-means in a 
-% RKHS. The whole problem is how do you formulate a 
+% RKHS. The problem is how do you formulate a 
 % statistical problem in a RKHS, or how do you know if there is a kernel
-% related to it. In the case of energy this was done by [6], and it's based 
-% on this work that we have the theory presented in this paper
-Moreover, we also consider a weighted version of energy statistics 
-applied to clustering, which makes connection to 
-graph partitioning problems.
-To find local optimizers of such QCQP we consider an iterative
-algorithm based on Hartigan's method, which in this case
-has the same computational cost 
-as kernel $k$-means algorithm but usually with better clustering
-quality. We provide carefully designed numerical experiments
-showing the superiority of this method compared to
-kernel $k$-means, standard $k$-means, and gaussian mixture models in a variety of settings.
-\end{abstract}
+% related to it. In the case of energy statistics this was done by [6], and
+% is a nontrivial result. It's based on this work that we have the theory 
+% presented in this paper
 
@@
 \section{Introduction}
+
+% @gui: algorithms don't make assumptions, theorems do.  also, algorithms cannot be parametric or non-parametric, that is a property only of the statistical model, or an estimator.
+
+% @gui: k-means is NOT a special case of GMM, GMM is a soft clustering (not fuzzy, in stats and machine learning we don't say "fuzzy"), where as k-means is a hard clustering.
+k-means is certainly closely related to GMM though.
+
+% @gui:  don't know what "clear resolution of cluster centers" adds to this?
+
+% @gui: algorithms cluster, doesn't matter how complicated the data are.  so, what do you actually mean here?  perform well empirically? what does "complicated" actually mean?
+
+% @gui: the words "this" and "it" are often vague and ambiguous, be specific/precise
+
+% @gui: not sure what you mean here, in kernel k-means, we don't "optimize" over any parameters typically.  sometimes we select hyperparameters of kernels via cross-validation, is that what you mean?
+
+% @gui: not sure what you mean "only one possible optimization problem".  we can reformulate it many different ways? also, below, "this" is ambiguous.
+
+% @gui: to me, we did not make explicit one of the most important points.
+in this paper, we make clear exactly when we expect energy clustering to perform better than k-means/GMM, and when we expect it to perform worse (empirically).
+
+% @gui: i find ToC paragraphs useless.
+
+% @jovo: corrected all of these points, and also re-wrote some other stuff.
+Please check.
+About ToC, I agree and wouldn't mind remove it, however it is a tradition that most referees like to see.
+In this case I included some important information about our work too, and it makes a transition between introduction and the remaining of the paper.
+If you want to remove it let me know and I will adapt.
 
@@
 notion of statistical potential energy between probability distributions,
-in close analogy to Newton's gravitational
-potential in classical mechanics. 
-It provides a nonparametric test for equality of distribution which
-is achieved under minimum energy. When probability distributions
-are different, this statistical potential 
-energy diverges as sample size increases.
-Energy statistics has been applied to several goodness-of-fit 
-hypothesis tests, 
-multi-sample tests of equality of distributions, analysis of variance
-\cite{RizzoVariance}, nonlinear
-dependence tests through
+in close analogy to Newton's gravitational potential in classical mechanics. 
+It provides a hypothesis test for equality of distributions which is achieved 
+under minimum energy. When probability distributions are different the 
+statistical potential energy diverges as sample size increases, while tends 
+to a nondegenerate limit distribution, i.e. remains finite, when they are 
+equal. Energy statistics has been applied to several goodness-of-fit 
+hypothesis tests, multi-sample tests of equality of distributions, 
+analysis of variance \cite{RizzoVariance}, nonlinear dependence tests through
 distance covariance and distance correlation, which generalizes the Pearson
-correlation coefficient, and  
-hierarchical clustering \cite{RizzoClustering} by extending Ward's
-method of minimum variance. We refer to \cite{Szkely2013} and 
-references therein for an overview.
-Moreover, an energy statistics formulation
-to clustering was already proposed in \cite{Kgroups} 
-which greatly motivated this paper.
-
-Recently, distance covariance was 
-generalized from Euclidean
-spaces to metric spaces of strong negative type \cite{Lyons}.
-Furthermore, a unifying framework establishing an equivalence
-between generalized energy distances to maximum
+correlation coefficient, and hierarchical clustering \cite{RizzoClustering} 
+by extending Ward's method of minimum variance. Moreover, an application of 
+energy statistics to clustering was already proposed \cite{Kgroups}, 
+which in part motivated this paper. We refer the reader to \cite{Szkely2013}, 
+and references therein, for an overview of energy statistics theory and 
+its applications.
+
+In its original formulation, energy statistics has a compact representation
+in terms of expectations of pairwise Euclidean distances, providing
+straightforward empirical estimates. 
+The notion of distance covariance was further generalized from Euclidean 
+spaces to metric spaces of strong negative type \cite{Lyons}. Moreover, 
+the missing link between energy distance based tests and kernel based tests has 
+been recently resolved \cite{Sejdinovic2013}, where a unifying framework
+establishing an equivalence between generalized energy distances to maximum
 mean discrepancies (MMD), which are distances between embeddings of 
-distributions in reproducing kernel Hilbert spaces (RKHS), was
-established \cite{Sejdinovic2013}. This provides the 
-link between concepts used in energy statistics and concepts
-commonly used in machine learning, and form the basis of our approach.
-
-The clustering problem has a long history in machine learning.
-Perhaps the most used method is $k$-means \cite{Lloyd,MacQueen,Forgy}, which
+distributions in reproducing kernel Hilbert spaces (RKHS), was established. 
+This equivalence provides the link between energy distance based measures and 
+RKHS based measures, commonly used in machine learning, and form the basis 
+of our approach.
+
+The clustering problem has such a long history in machine learning, making it
+impossible to mention all important contributions in a short space. 
+Perhaps, the most used method is $k$-means \cite{Lloyd,MacQueen,Forgy}, which
 is based on Lloyd's heuristic \cite{Lloyd} of assigning a data point to
 the cluster with closest center. The only statistical 
-information about each cluster
-comes from its mean and it is thus sensitive to outliers. 
-% @gui: algorithms don't make assumptions, theorems do.  also, algorithms cannot be parametric or non-parametric, that is a property only of the statistical model, or an estimator.
-% @jovo: commented out below and included a phrase in the end
-%Moreover, $k$-means
-%is strongly parametric, assuming that data is isotropically and normally
-%distributed since it is a special case of gaussian mixture models (GMM) 
-% @gui: k-means is NOT a special case of GMM, GMM is a soft clustering (not fuzzy, in stats and machine learning we don't say "fuzzy"), where as k-means is a hard clustering.
-k-means is certainly closely related to GMM though.
-%through the expectation maximization (EM) algorithm. 
-Nevertheless,
-$k$-means works very well when data is linearly separable 
-in Euclidean space.
-%, with a clear resolution of cluster centers and without
-%significant overlap of the points belonging to different clusters.
-% @gui:  don't know what "clear resolution of cluster centers" adds to this?
-% @jovo: commented above and included this phrase instead
-Gaussian mixture models (GMM) is also very commonly used for
-clustering,
-however it is strongly parametric, as $k$-means which is closely related.
+information about each cluster comes from its mean, and it is thus sensitive 
+to outliers. Nevertheless, $k$-means works very well when data is 
+linearly separable in Euclidean space. Gaussian mixture models (GMM) is also 
+very commonly used for clustering, however, it makes strong assumptions about 
+the distribution of the data, as $k$-means which is closely related to GMM.
 
@@
 map data points to a RKHS, then clustering can be performed in the associated
-Hilbert
-space by using its inner product. 
-%This enables one to cluster way more complicated datasets, 
-% @gui: algorithms cluster, doesn't matter how complicated the data are.  so, what do you actually mean here?  perform well empirically? what does "complicated" actually mean?
-% @jovo: what I mean is not tied to any algorithm. I want to say that kernel
-% methods allows one to cluster data that is not separable in Euclidean space.
-% I modified the phrase, see if it's better
-However, the kernel choice remains the biggest challenge
-since there is no principled theory to construct a kernel for a given
-dataset,
-% @gui: the words "this" and "it" are often vague and ambiguous, be specific/precise
-and usually a kernel introduces hyperparameters that need to be carefully
-chosen.
-%extra parameters that need to be optimized. 
-% @gui: not sure what you mean here, in kernel k-means, we don't "optimize" over any parameters typically.  sometimes we select hyperparameters of kernels via cross-validation, is that what you mean?
-% @jovo: tried to fix above, you're right in what I meant
-The well-known
-kernel $k$-means optimization problem is nothing but $k$-means in the
-feature space \cite{Girolami}. Furthermore, kernel $k$-means algorithm
+Hilbert space by using its inner product. Yet, the kernel choice remains 
+the biggest challenge since there is no principled theory to construct a kernel
+for a given dataset, and usually a kernel introduces hyperparameters that 
+need to be carefully chosen.
+The well-known kernel $k$-means optimization problem is nothing but $k$-means 
+in the feature space \cite{Girolami}. Furthermore, kernel $k$-means algorithm
 \cite{Dhillon2,Dhillon} is still based on Loyd's heuristic \cite{Lloyd} of
@@
 Since the statistical potential energy is minimum when
-distributions are equal, the
-principle behind clustering is to maximize the statistical energy, 
-enforcing probability distributions associated to each cluster
-to be different from one another.
-We provide a precise mathematical formulation to this
-statement 
-%showing that there is only one possible optimization problem
-%consistent with energy statistics. 
-% @gui: not sure what you mean "only one possible optimization problem".  we
-% can reformulate it many different ways? also, below, "this" is ambiguous.
-% @jovo: I had in mind that minimizing W is equivalent to maximizing S, 
-% but it's confusing at this point so just removed the phrase and adapted
-%Moreover, we show that this is
-leading to
-a quadratically constrained
-quadratic program (QCQP) in the associated RKHS. Our results immediately 
-establish the connection 
-to kernel methods used in machine learning, 
-by showing that this QCQP is equivalent
-to kernel $k$-means optimization problem. 
-The equivalence between kernel $k$-means, 
-spectral clustering, and graph partitioning problems was already established
-\cite{Dhillon,Dhillon2}. We show how these connections arise
-from a weighted version of energy statistics.
-
-Our algorithmic contribution is to use Hartigan's method
-\cite{Hartigan} to find local 
-solutions of the above mentioned QCQP.
-This approach was also proposed in \cite{Kgroups}.
-In this case,
-the resulting algorithm has the same
-time complexity as kernel $k$-means algorithm, however, the numerical 
-results provide compelling evidence that this method
-is more accurate and robust than kernel $k$-means. This evidence
-is also supported by general theoretical results \cite{Telgarsky,Slonin}.
-Moreover, in our experiments we put in
-evidence the nonparametric aspect of energy
-statistics based clustering, which provides a family of 
-default kernels, showing that
-it is able to perform accurately on datasets coming very different 
-distributions, contrary to $k$-means and GMM for instance.
-% @gui: to me, we did not make explicit one of the most important points.
-in this paper, we make clear exactly when we expect energy clustering to perform better than k-means/GMM, and when we expect it to perform worse (empirically).
-% @jovo: added this phrase, check it
-More specifically, the proposed method performs closely to $k$-means
-and GMM on normally distributed data with balanced clusters, however, it
-performs considerably better on  data that is not normally distributed.
-It also
-performs better than $k$-means and GMM in higher dimensions, even on
-gaussian settings, and it performs worse than
-GMM when we have highly unbalanced clusters.
-
-% @gui: i find ToC paragraphs useless.
-% @jovo: I agree and don't mind removing, however it's a tradition that 
-% most referees like to see and it doesn't do any harm. I'm fine with
-% whatever choice
-Our work is organized as follows. In section~\ref{sec:background} we introduce
+distributions are equal, the principle behind clustering is to maximize 
+the statistical energy,  enforcing probability distributions associated to 
+each cluster to be different from one another. We provide a precise 
+mathematical formulation to this statement, leading to a quadratically 
+constrained quadratic program (QCQP) in the associated RKHS. Our results
+immediately establish the connection to kernel methods, by showing that this 
+QCQP is equivalent to kernel $k$-means optimization problem. The 
+equivalence between kernel $k$-means, spectral clustering, and graph 
+partitioning problems is well-known \cite{Dhillon,Dhillon2}. We demonstrate 
+how these connections arise from a weighted version of energy statistics.
+
+Our algorithmic contribution is to use Hartigan's method \cite{Hartigan} to 
+find local solutions of the above mentioned QCQP, which is NP-hard in general.
+Hartigan's method was also used in \cite{Kgroups}, and its advantages
+over Lloyd's method was already being demonstrated \cite{Telgarsky,Slonin},
+but apparently it did not receive the desired attention. To the 
+best of our knowledge, Hartigan's method was not previously applied together 
+with kernel methods. We provide an algorithm based on Hartigan's method in 
+terms of the Gram matrix of the data computed from a kernel fixed by 
+energy statistics. We make clear the advantages of this proposal versus 
+Lloyd's method, which kernel $k$-means is based upon and will also be used 
+to locally solve our QCQP. We show that both algorithms  have the same
+time complexity.
+
+Our numerical results provide compelling evidence that 
+Hartigan's method applied to energy statistics based clustering, 
+or \emph{energy clustering} for short, is more accurate and robust than 
+Lloyd's method. Moreover, in our experiments we put in evidence the
+flexibility of energy clustering, which provides a family of default kernels, 
+showing that it is able to perform accurately on data coming from 
+very different distributions, contrary to $k$-means and GMM for instance.
+More specifically, the proposed algorithms for energy clustering perform 
+closely to $k$-means and GMM on normally distributed data. On the other
+hand, it performs considerably better than $k$-means and GMM on  data that 
+is not normally distributed. It also performs better than $k$-means 
+and GMM in high dimensions, even on Gaussian settings. However, it performs 
+worse than GMM, and more closely to $k$-means, for highly unbalanced clusters
+with normal distributions.
+
+Our work is organized as follows. In section~\ref{sec:background}, we introduce
 the necessary background on energy statistics and RKHS.
@@
 where we consider a clustering theory based on energy statistics leading
-to a QCQP, which is NP-hard. We also show the equivalence to 
-kernel $k$-means problem.
-In Section~\ref{sec:weighted} we generalize these results to a weighted
-version of energy statistics, which provides the connection with graph
+to a QCQP, which is NP-hard in general. We also show the equivalence to 
+kernel $k$-means optimization problem.
+In Section~\ref{sec:weighted}, we generalize these results to a weighted
+version of energy statistics, providing connections to graph
 partitioning problems and spectral clustering.
-In Section~\ref{sec:twoclass} we consider a simple example in one dimension,
-where we propose an algorithm which requires no initialization.
-In section~\ref{sec:algo} we briefly review kernel $k$-means algorithm,
-and propose a new iterative algorithm based on Hartigan's method
-to solve this QCQP.
+In Section~\ref{sec:twoclass}, we consider the  simple case of a two-class
+problem in one dimension,
+where we propose an algorithm which requires no random initialization.
+In section~\ref{sec:algo}, we consider Lloyd's and Hartigan's methods
+and propose iterative algorithms to solve the QCQP.
 Section~\ref{sec:numerics} contains some carefully designed numerical
-experiments indicating that this algorithm outperforms kernel
-$k$-means, standard $k$-means, and GMM/EM algorithms.
-Our final conclusions are presented in section~\ref{sec:conclusion}.
+experiments showing that in several settings, energy clustering based
+on Hartigan's algorithm
+outperms kernel $k$-means algorithm, standard $k$-means algorithm 
+and GMM with the expectation maximization algorithm. 
+Our final conclusions are presented in Section~\ref{sec:conclusion}.
 
@@
 \label{sec:background}
+
+% @gui: i don't know what this means.  eqn 7 defines something for a population statistic, not a sample statistic.
+% @jovo: removed. I was thinking that you can always estimate the expected 
+% value by computing a sample mean. but I agree, it's confusing
+
+% @gui: i'm not sure the above is true/meaningful.  nonparametric means implicitly estimating an infinite number of parameters.  without any statistical theory, i'm not quite sure what we can say.
+% @jovo: I stoped saying non-parametric, however, Kolmogorov-Smirnov is also said to be non-parametric, so I thought that in this context it means that it does not assume any form of the distributions.
 
@@
 \label{eq:energy}
-\Energy(P, Q) \equiv 2 \E \| X - Y\| - \E \| X - X' \| - \E \| Y - Y' \|
-\end{equation}
-called \emph{energy distance} \cite{Szkely2013} 
+\Energy(P, Q) \equiv 2 \E \| X - Y\| - \E \| X - X' \| - \E \| Y - Y' \|,
+\end{equation}
+called \emph{energy distance} \cite{Szkely2013}, 
 is rotationally invariant and nonnegative, $\Energy(P,Q) \ge 0$, where
@@
 
-% @gui: let the reader decide what is important, rather than us telling them :)
-% @jovo: removed that
-The energy distance \eqref{eq:energy2} can be even further generalized.
+The energy distance can be even further generalized.
 Let $X, Y \in \mathcal{X}$  where $\mathcal{X}$ is an arbitrary space endowed
@@
 $\rho$ may not satisfy the triangle inequality, 
-$\rho^{1/2}$ does since it can be shown to be a legitamate metric.
+$\rho^{1/2}$ does since it can be shown to be a proper metric.
 
@@
 $\kk$ is positive definite if and only if $\rho$ is a semimetric
-of negative type
-\eqref{eq:negative_type}.
+of negative type.
 We have a family of kernels, one for each choice of $x_0$. Conversely,
@@
 product of $\Hk$.
-%Moreover, this can be estimated from a sample
-%according to \eqref{eq:inner_data}.
-% @gui: i don't know what this means.  eqn 7 defines something for a population statistic, not a sample statistic.
-% @jovo: removed. I was thinking that you can always estimate the expected 
-% value by computing a sample mean. but I agree, it's confusing
 
@@
 $\C_i \cap \C_j = \emptyset$.
-Each expectation in 
+Each expectation in the generalized energy distance
 \eqref{eq:energy3}
@@
 \end{equation}
-where $n_i = |\C_i|$ is the number of elements in 
+where $n_i = |\C_i|$ is the number of elements in partition
 $\C_i$. 
@@
 W \equiv
-\sum_{j=1}^{k} \dfrac{n_j}{2} g(\C_j, \C_j)
+\sum_{j=1}^{k} \dfrac{n_j}{2} g(\C_j, \C_j),
 \end{equation}
@@
 g(\C_j, \C_j)
-\right]
+\right],
 \end{equation}
@@
 the quantity $S$ provides
-a \emph{nonparametric test statistic} for equality of distributions
+a test statistic for equality of distributions
 \cite{Szkely2013}.
-% @gui: i'm not sure the above is true/meaningful.  nonparametric means implicitly estimating an infinite number of parameters.  without any statistical theory, i'm not quite sure what we can say.
-% @jovo: my understanding is that this hypothesis test does not assume any
-% form of the distribution. For instance, Kolmogorov-Smirnov is also
-% said to be nonparametric
 When the sample size is large enough, $n\to \infty$,
@@
 we have that $S \to \infty$.
-This test is nonparametric in the sense that it does not make any assumptions
+This test does not make any assumptions
 about the distributions $P_j$.
 
-One can make the analogy 
+One can make a physical analogy by thinking 
 that points $ x \in \C_j$ form a massive body 
@@
 
+% @gui: consider only using equation numbers for equations that you later refer to? there are so many, it is easy to get lost.
+% @jovo: ok, will do that after we finish everything. There is a package that automatically remove numbers from unreferenced equations
+
+% @gui: where possible, i recommend using the "name" of the quantity rather than the number of the equation.  eg, here, write S and W, rather than (13) and (12).
+% @jovo: ok, I will do that for the main symbols, and refer to equations only when necessary
+
+% @gui: do other people know this result? (about min W = max X) if so, cite them, if not, state that.
+% @jovo: eq (15), i.e. S + W = T is well-known. Gabor and Rizzo wrote that
+% in their paper, however, they didn't write a proof. They should have
+% it in some other paper.  However, they mention 
+% that this holds only for 0 < alpha < 2, with |x-y|^alpha, while
+% here I did for a general semimetric. The PhD thesis do not 
+% mention anywhere that minimizing W is equivalent to maximizing S.
+% Thus, although a trivial result, but important, I didn't see this 
+% written explicitly before.
+
+% @gui: "Let us" is used a bunch, though i think it never adds anything? i'd remove it everywhere.
+% @jovo: ok, remove it all along
+
+% @gui: i don't understand the notaiton (.)_jj. is a sum missing?
+% @jovo: this is standard.  A_{ij} denotes the 
+% $(i,j)$th element of matrix A as you know. 
+% Since here A = Z^T G Z is a product of 3 matrices
+% we just put a parenthesis to avoid confusion, that's all
+
+% @gui: don't use equation numbers here.
+% @jovo: in some cases it is unavoidable. I tried to do that when the context
+% is obvious, by writing the symbols, but when we refer to long 
+% expressions or derivation steps it is way more clear so say, replace equation (1) into (2) to get ...
+than saying, replacing A into B to get ...
+and the reader has to glance through the entire paper to recall what A and B are.
+
+% @gui: this might be a better initialization than k-means++, which essentially initializes using a different kernel?
+% @jovo: yes it might be, and I agree, k-means++ is initializing in "Euclidean
+% space" not in the feature space where data might be better separable. We will do this from now on.
+
+% @gui: i'm not sure about this.  who is to say the correct direction of causality? maybe energy clustering is a special case of kernel k-means?
+% @jovo: tried to not commit to a single view. We show that they are
+% equivalent, when opperating on the same kernel. However, kernel k-means per
+% se does not consider anything about the kernel choice. Energy statistics,
+% however, provides a theory valid on spaces of negative type, which guarantees
+% that a kernel constructed in this formalism will be positive-semidefinite.
+
+In this vein, kernel $k$-means is part of a
 This section contains the main theoretical results of this paper, where 
@@
 
-% @gui: consider only using equation numbers for equations that you later refer to? there are so many, it is easy to get lost.
-% @jovo: ok, will do that after we finish everything.
-Due to the test statistic \eqref{eq:between} for equality of distributions,
+Due to the energy test statistic for equality of distributions,
 the obvious
@@
 straightforward result
-shows that maximizing \eqref{eq:between} is, however, equivalent to minimizing
-\eqref{eq:within} which has a more convenient form.
-% @gui: where possible, i recommend using the "name" of the quantity rather than the number of the equation.  eg, here, write S and W, rather than (13) and (12).
-% @jovo: ok, I will do that for the main symbols
+shows that maximizing $S$ is, however, equivalent to minimizing
+$W$ which has a more convenient form.
 
@@
 \mathcal{X}\times\mathcal{X} \to \mathbb{R}$ of
-negative type \eqref{eq:negative_type}. For a fixed integer $k$,
+negative type. For a fixed integer $k$,
 the partition
 $\mathbb{X} = \bigcup_{j=1}^k \C_j$, where $\C_i \cap C_j = \emptyset$ for
-all $i\ne j$, maximizes \eqref{eq:between} if and only if
+all $i\ne j$, maximizes the between-sample statistic $S$, defined
+in equation \eqref{eq:between}, if and only if
 \begin{equation}
@@
 \end{equation}
-where $W$ is given by \eqref{eq:within}.
+where the within energy dispersion $W$ is defined by \eqref{eq:within}.
 \end{proposition}
@@
 \sum_{j\ne i = 1}^k n_j \bigg] 
-n_i g(\C_i, \C_i) 
-\\
+n_i g(\C_i, \C_i) \\
 & = \dfrac{1}{2n} \sum_{i,j=1}^k n_i n_j g(\C_i, \C_j)
@@
 \end{proof}
-% @gui: do other people know this result? if so, cite them, if not, state that.
-% @jovo: eq (15), i.e. S + W = T is well-known. Gabor and Rizzo wrote that
-% in their paper, however, they didn't write a proof. They should have
-% in some other paper I believe (will double check)
-% However, they mention 
-% that this holds only for 0 < alpha < 2, with |x-y|^alpha, while
-% here I did for a general semimetric. The PhD thesis do not 
-% mention anywhere that
-% minimizing W is equivalent to maximizing S, so I didn't see this written
-% before. Anyway, this result is trivial, but important.
 
@@
 finding the best partition of the data by minimizing $W$.
-%solving~\eqref{eq:minimize}.
-% @gui: here and elsewhere is another good example.  why not just say "by minizing W"?
-% @jovo: ok
 Notice that this is a hard clustering problem as partitions
-are disjoint. The problem \eqref{eq:minimize} based on
+are disjoint. The optimization problem \eqref{eq:minimize} based on
 energy statistics was already proposed in \cite{Kgroups}. However, it is
-important to note that this is equivalent to maximizing \eqref{eq:between}
-which is the test statistic for equality of distributions. In the following
-we show what is the explicit optimization problem behind \eqref{eq:minimize}.
-
-% @gui: "Let us" is used a bunch, though i think it never adds anything? i'd remove it everywhere.
-% @jovo: ok
-%Let us
-We now formulate \eqref{eq:minimize} in the corresponding
-RKHS. Based on \eqref{eq:kernel_semimetric} and \eqref{eq:gen_kernel}, 
+important to note that this is equivalent to maximizing $S$
+which is the test statistic for equality of distributions. In this current
+form, the relation with kernels and to other clustering methods is obscure.
+In the following, we show what is the explicit optimization problem behind 
+in the corresponding RKHS, establishing the connection to kernel methods.
+
+We now formulate the analog of the optimization problem \eqref{eq:minimize} 
+in the corresponding RKHS. 
+Based on the relation between kernels and semimetrics, see 
+\eqref{eq:kernel_semimetric} and \eqref{eq:gen_kernel}, 
 assume that the kernel $\kk: \mathcal{X} \times \mathcal{X} \to \mathbb{R}$ 
-generates $\rho$.  Let us define  the Gram matrix
+generates $\rho$.  Define  the Gram matrix
 \begin{equation}
@@
 each point (row) belongs to. This matrix satisfies
-$Z^\top Z = D$ where the diagonal matrix 
+$Z^\top Z = D$, where the diagonal matrix 
 $D = \diag( n_1,\dotsc, n_k )$  contains
-the number of points in each cluster. Let us also introduce the rescaled
+the number of points in each cluster. We also introduce the rescaled
 matrix  $Y \equiv Z D^{-1/2}$. In component form they are given by
@@
 is NP-hard since
-it is a quadratically constrained quadratic program (QCQP).
+it is a quadratically constrained quadratic program (QCQP) in the RKHS.
 
@@
 \label{th:qcqp2}
-The problem \eqref{eq:minimize} is equivalent to
+The optimization problem \eqref{eq:minimize} is equivalent to
 \begin{equation}
@@
 \end{equation}
-% @gui: i don't understand the notaiton (.)_jj. is a sum missing?
-% @jovo: this is standard.  A_{ij} denotes the 
-% $(i,j)$th element of matrix A as you know. 
-% Since here A = Z^T G Z is a product of 3 matrices
-% we just put a parenthesis to avoid confusion, that's all
 where we used the definitions \eqref{eq:kernel_matrix} and
@@
 Now we write this in terms of the matrix $Y = Z D^{-1/2}$.
-% @gui: Y is a matrix? or a vector? when introducing new terms, always define what space they are in rigorously
-% @jovo: it's obviously a matrix
-% I didn't want to write Y \in \mathbb{R}^{blabla} because it polutes the
-% text too much and it's dimension is obviously the same as Z. Y was defined
-% before also
 The objective function immediately becomes
@@
 to \eqref{eq:qcqp2} .
-% @gui: don't use equation numbers here.
-% @jovo: in these cases there is no way, I have to. Look at the size
-% of the expressions. I agree to not put reference to eq numbers
-% for the main symbols, like S and W, but for large expressions we have to.
-% I also try to avoid poluting the text with unnecessary math symbols, 
-% it looks pretty bad
 \end{proof}
@@
 indicating that $x_i \in \C_j$. 
-Problem \eqref{eq:qcqp2} is NP-hard and there
-are few methods
-available to solve it directly,
-which is computational prohibitive even for small datasets.
+This optimization problem is nonconvex, and also NP-hard,
+thus a direct approach 
+is computational prohibitive even for small datasets.
 However, one can find approximate solutions by relaxing some 
@@
 \begin{equation}
+\label{eq:relaxed}
 \max_{Y} \Tr \left( Y^\top G \, Y \right) \quad \mbox{s.t. $Y^\top Y = I$}
@@
 has a well-known closed form solution $Y^\star = U R$, where the
-columns of $U$ contain the leading $k$ eigenvectors of $G$ corresponding
+columns of $U \in \mathbb{R}^{n\times k}$ 
+contain the top $k$ eigenvectors of $G$ corresponding
 to the $k$ largest eigenvalues $\lambda_1\ge \lambda_2\ge\dotsc\ge\lambda_k$, 
@@
 The resulting
-optimal objective function is given by
+optimal objective function value is thus
 $\max \Tr \left( {Y^\star}^\top G \, Y^\star \right)  = 
-\sum_{i=1}^k \lambda_i$. One might then normalize and threshold the rows
-of $Y^\star$,
-% @gui: this might be a better initialization than k-means++, which essentially initializes using a different kernel?
-% @jovo: yes it might be, and I agree, k-means++ is initializing in "Euclidean
-% space" not in the feature space where data might be better separable. Maybe I
-% need test this numerically
-or better, following \cite{NgJordan} one can normalize the
-rows of $Y^\star$ and apply standard $k$-means on this matrix where each
-row is considered as a data point.
-This is the procedure done in spectral clustering on a graph Laplacian
-matrix obtained through a similarity matrix.
-However, computing eigenvectors of very large matrices
-can be prohibitively expensive 
-and usually iterative methods are preferred. Moreover,
-this procedure is only guaranteed to converge if the normalized $Y^\star$ is
-positive semidefinite.
-In Section~\ref{sec:algo} we will propose an iterative method to directly 
-find approximate solutions to \eqref{eq:qcqp2} that is guaranteed to
-converge and does not require that the similarities $\kk(x_i,x_j)$ be
-nonnegative.
-
-The clustering problem \eqref{eq:qcqp2} based on energy statistics
+\sum_{i=1}^k \lambda_i$.
+
+Spectral clustering is based on the above approach, where
+one further normalize the rows of $Y^\star$, 
+i.e. $Y^\star \to \widetilde{Y}$ where 
+$\widetilde{Y}_{ij} \equiv Y^\star_{ij}/ \| Y^\star_{i\bullet} \|$. 
+The vectors
+$y_i \equiv \widetilde{Y}_{i\bullet} \in \mathbb{R}^k$, which are the
+rows of matrix $\widetilde{Y}$, lie on the surface of the $k$-dimensional 
+unit sphere.
+The set $\{ y_i \}_{i=1}^n$ is a spectral embedding of the original
+set $\{ x_i \}_{i=1}^n$ onto an eigen-subspace of the kernel matrix $G$.
+One then group the points $\{ y_i \}_{i=1}^n$ into $k$ clusters 
+$\{ \tC_1,\dotsc,\tC_k \}$, using any clustering algorithm such as
+$k$-means, yielding the final solution
+where the original point $x_i$ belongs
+to cluster $\C_j$ if and only if $y_i$ belongs to $\tC_j$.
+A procedure on these lines was proposed in the seminal paper
+\cite{NgJordan}, which considers the eigen-decomposition of
+the matrix $D^{-1/2} G D^{-1/2}$ instead of $G$, and also 
+where $G$ is an affinity matrix with the 
+modification that its diagonal elements are set to zero $G_{ii}=0$, 
+and $D=\diag(d_{1},\dotsc,d_{n})$ is the degree matrix, 
+$d_{i} \equiv \sum_{j=1}^n G_{ij}$. 
+Here we consider a slightly different
+prescription which we call $\mathcal{E}$-spectral, using exactly the same 
+approach, but considering the
+eigen-decomposition of $G D^{-1}$ instead\footnote{
+We found empirically that our method is more stable and often more
+accurate than \cite{NgJordan}.}. Our method is described in 
+Algorithm~\ref{spectralalgo}, and will be used as an initialization to the
+iterative algorithms to solve the full optimization problem \eqref{eq:qcqp2}
+that will be proposed later.
+
+\begin{figure}
+\begin{flushleft}
+\begin{algorithm}[H]
+\vspace{.5em}
+\begin{algorithmic}[1]
+\INPUT Gram matrix $G\in\mathbb{R}^{n\times n}$, and number of clusters $k$
+\OUTPUT label matrix $Z \in \{ 0, 1\}^{n \times k}$
+\STATE set the diagonal of $G$ to zero: $G_{ii} \leftarrow 0$ 
+        for $i=1,\dotsc,n$
+\STATE $D\leftarrow\diag(d_1,\dotsc,d_n)$ 
+        where $d_i \leftarrow \sum_{j=1}^n G_{ij}$
+\STATE solve the generalized eigenvalue problem $G v_i = \lambda_i D v_i$,
+        where $i=1,\dotsc,n$ with eigenvalues, and corresponding eigenvectors,
+        obeying the ordering
+        $\lambda_1 \ge \lambda_2 \ge \dotsc \lambda_n$
+\STATE $Y \leftarrow (D v_1| \dotsm | D v_k) 
+        \in \mathbb{R}^{n\times k}$ 
+has the top $k$ eigenvectors (multiplied by $D$) as columns
+\STATE normalize the rows $Y$, i.e. $\widetilde{Y}_{ij} \leftarrow Y_{ij} / 
+\left(\sum_{j=1}^n |Y_{ij}|^2\right)^{1/2}$
+\STATE apply $k$-means on $\{ y_i \}_{i=1}^n$, 
+where $y_i = \widetilde{Y}_{i\bullet}$
+is the $i$th row of $\widetilde{Y}$
+\STATE $Z_{ij} \leftarrow 1$ if $y_i$ is in cluster $j$, 
+        and $Z_{ij} \leftarrow 0$ otherwise
+\end{algorithmic}
+\caption{
+\label{spectralalgo}
+$\mathcal{E}$-spectral is an exact method to solve
+the relaxed problem \eqref{eq:relaxed}.
+\hspace{\fill}
+}
+\end{algorithm}
+\end{flushleft}
+\end{figure}
+
+Note that the energy clustering problem \eqref{eq:qcqp2} 
 is valid for data living in an \emph{arbitrary} space of negative type where
-a semimetric $\rho$, and thus the kernel \eqref{eq:kernel_semimetric}, are
-assumed to be known. 
-This clustering 
-%method is
-formulation from energy statistics is 
-\emph{nonparametric} in the sense that it does not make assumptions
-about  the distribution of the data, 
-contrary to $k$-means and GMM formulations, for example.
-% @gui: algorithms don't make assumptions, proofs do.  we don't prove anything about when we get the right answer, so we can't quite say this.
-% @jovo: I adapted above, see if it makes sense, if not I just 
-% remove the phrase
-If one uses the standard energy distance
-defined in \eqref{eq:energy} 
-we have $\rho(x,y) = \| x - y\|$ and this fixes the kernel
-through \eqref{eq:kernel_semimetric}. In the same way, 
-for data living in a more general
+a semimetric $\rho$, and thus the kernel $\kk$, are
+assumed to be known. For instance, standard energy statistics in
+Euclidean space fixes a family of choices, 
+$\rho(x,y) = \| x - y\|^\alpha$ and the corresponding
+kernel $K(x,y) = \tfrac{1}{2} \left( \| x \|^\alpha + \| y \|^\alpha - 
+\| x-y\|^\alpha \right)$, where we
+set $x_0=0$ in \eqref{eq:kernel_semimetric}, for $0<\alpha\le 2$.
+For data living in a more general
 metric space $(\mathcal{X}, \rho)$ the corresponding semimetric $\rho$ fixes
 the kernel.
+In any case, the energy  clustering 
+formulation does not make assumptions
+about  the distribution of the data, 
+contrary to $k$-means and GMM, for example.
 In practice, however,
 the clustering quality strongly depend on the choice of a suitable
-$\rho$ which measures the similarity between data points,
+$\rho$, which measures the similarity between data points,
 and is equivalent to choosing an appropriate kernel.
-The energy distance \eqref{eq:energy2} fixes a family of choices
-$\rho(x,y) = \| x-y\|^\alpha$ with $0 < \alpha\le 2$. 
-Nevertheless, if prior information
-is available one may choose a more appropriate $\rho$.
+However, if prior information is available to choose a suitable $\rho$,
+the energy clustering conviniently incorporate this measure in the 
+optimization problem \eqref{eq:qcqp2}.
 
 \subsection*{Relation to Kernel $\bm{k}$-Means}
-One may wonder how energy statistics clustering 
+
+One may wonder how energy clustering 
 relates to the well-known kernel $k$-means problem%
@@
 confused with kernel $k$-means algorithm that is just one possible recipe 
-to solve \eqref{eq:kernel_kmeans}.
-} 
+to solve \eqref{eq:kernel_kmeans}.} 
 which is extensively used in machine learning.
-%and was proposed as a kernelized version of $k$-means.
 For a positive semidefinite Gram matrix $G$, as defined in
@@
 is strongly tied to the idea of minimizing distances between points
-and cluster centers, which arises from $k$-means objective function.
- \cite{Dhillon2,Dhillon} show
+and cluster centers, which arises from $k$-means objective function based
+on Lloyd's method \cite{Lloyd}.
+It is known \cite{Dhillon2,Dhillon}
 that problem \eqref{eq:kernel_kmeans} 
-is equivalent to a QCQP in the same form as
+can be cast into a trace maximization problem in the same form as 
 \eqref{eq:qcqp2}. The next result makes this explicit, showing that
@@
 \begin{equation}
-\min_{\C_1,\dotsc,\C_k} \bigg\{ J(\C_1,\dotsc,\C_k) = \max_{\C_1,\dotsc,\C_k}
-\sum_{j=1}^k \dfrac{1}{n_j} \sum_{x,y \in\C_j} \kk(x,y) \bigg\}
+\max_{\C_1,\dotsc,\C_k}
+\sum_{j=1}^k \dfrac{1}{n_j} \sum_{x,y \in\C_j} \kk(x,y) 
 \end{equation}
@@
 already shown in the proof of Proposition~\ref{th:qcqp2}, leading to
-\eqref{eq:qcqp2}.
+the optimization problem \eqref{eq:qcqp2}.
 \end{proof}
 
-%The above result shows that 
-%kernel $k$-means problem is actually a consequence
-%of energy statistics. 
 The above result shows that 
 kernel $k$-means problem is equivalent to the clustering problem
-formulated in the energy statistics framework.
-% @gui: i'm not sure about this.  who is to say the correct direction of causality? maybe energy clustering is a special case of kernel k-means?
-% @jovo: Let me describe the way I see it, and see if you agree.
-% I adapted the phrase above.
-% There is a theory about energy statistics. Semimetric spaces, RKHS, test
-% statistic, S + W = T, etc. This theory can be used for statistical inference
-% in general, not only for clustering. kernel k-means is an heuristic 
-% formulation for clustering where
-% one minimizes the objective function J, in eq (24), nothing more than that.
-% It turns out that min J = min W, so the clustering formulation from energy
-% statistics and kernel k-means are equivalent. It could have happened that
-% they are not, energy could possibly provide something different a priori, we
-% didn't know that. Moreover, in energy statistics we know that the kernel
-% must come from a semimetric of negative type, etc. there is a whole
-% construction of it
-In this vein, kernel $k$-means is part of a
-statistical framework where distances between probability
-distributions are maximized. 
-The kernel function $\kk$ arises from a semimetric $\rho$ of
-negative type which defines the energy distance between distributions.
+formulated in the energy statistics framework, when operating on the same
+kernel. Notice, however, that 
+energy statistics theory is valid for arbitrary semimetric spaces of
+negative type, and the corresponding kernel in the associated RKHS is
+guaranteed to be positive semi-definite.
 
@@
 one can use kernel $k$-means algorithm to solve these problems as well.
-This correspondence involves a weighted version of \eqref{eq:qcqp2} that
-we demonstrate in the following.
+This correspondence involves a weighted version of problem 
+\eqref{eq:qcqp2}, that
+we demonstrate in the following from the perspective of energy statistics.
 
@@
 vector. Now we can show the analogous of
-Proposition~\ref{th:qcqp2} to the case of \eqref{eq:minimize2}.
+Proposition~\ref{th:qcqp2} to the case of problem \eqref{eq:minimize2}.
 
@@
 
-Consider \eqref{eq:assoc}, whose objective function can be written as
+Consider the ratio association problem \eqref{eq:assoc}, 
+whose objective function can be written as
 \begin{equation}
@@
 with $Z$ defined in \eqref{eq:label_matrix} and $Y$ in
-\eqref{eq:weighted_matrices}. To make the analogy with \eqref{eq:qcqp3}
-explicit,  problem \eqref{eq:assoc} is equivalent to
+\eqref{eq:weighted_matrices}. Therefore, the ratio association problem
+can be written in the form \eqref{eq:qcqp3} as
 \begin{equation}
@@
 \end{equation}
-Therefore, this is exactly the same problem as weighted energy statistics
-clustering \eqref{eq:qcqp3} with 
+This is exactly the same problem as weighted energy 
+clustering with 
 $G = \mathcal{W}^{-1} \mathcal{A} \mathcal{W}^{-1}$. Assuming this
@@
 where in the second equation we assume the graph has no self-loops,
-i.e. $\mathcal{A}_{ii} = 0$. Using \eqref{eq:metric_graphs} 
-in \eqref{eq:g_def}--\eqref{eq:between}
-allows one to use energy statistics theory for inference
+i.e. $\mathcal{A}_{ii} = 0$. 
+Using \eqref{eq:metric_graphs} 
+in the energy statistics formulation 
+allows one to make inference
 on graphs.
@@
 % @jovo: ok, added the phrase below
-% let me know if not enough. In practice the weights can be anything defined
-% by the "user".
+% let me know if not enough. In practice the weights can be anything attributed
+% to the nodes of the graph
 Above, the weight $w_i = w(i)$ of node $i\in \mathcal{V}$ can be, for instance,
@@
 
-Before stating a general algorithm to solve \eqref{eq:qcqp2} 
-let us first consider the simplest possible case which
+% @gui: "does not require any initiializaton" i think you mean is guaranteed to find the optimal solution in O(n^2) time.
+% @jovo: it does not require initialization in the sense that we do not
+% assign points to clusters in an initial step, as k-means++ for instance, or
+% random initialization. This algorithm is "exact" since nothing
+% is random in it, and multiple runs will always give the same answer
+
+% @gui: to provide intuition to the reader, let's show the kernel density estimate for each of the 3 algorithms for both classes.
+i'm not sure the best way to visualize it, maybe play with a few options.
+
+Before stating a general algorithm to solve the optimization problem
+\eqref{eq:qcqp2}
+we first consider the simplest possible case which
 is one-dimensional data and a two-class problem. This will be useful to test
-energy statistics clustering on a simple setting.
+energy clustering on a simple setting.
 
 Fixing
-$\rho(x,y) = |x - y|$, according to \eqref{eq:energy}, we can actually compute 
-\eqref{eq:g_def} in $\OO(n \log n)$ and find
-a direct solution to \eqref{eq:minimize}. 
-This is done by noticing that
+$\rho(x,y) = |x - y|$ according to the standard energy distance, we can 
+actually compute the function
+\eqref{eq:g_def} in $\OO(n \log n)$ and minimize
+$W$ directly.
+This is done by noting that
 \begin{equation}
@@
 Let $\C$ be a partition with
-$n$ elements. Using the above distance in \eqref{eq:g_def} we have
+$n$ elements. Using the above distance we have
 \begin{equation}
@@
 that we first order the data in $\C$, obtaining
-$\bar{\C} = [ x_j \in \C: x_1 \le x_2 \le \dotsm \le x_{n}]$, we
-can write \eqref{eq:g_ind} in the following simple form:
+$\tC = [ x_j \in \C: x_1 \le x_2 \le \dotsm \le x_{n}]$, we
+get 
 \begin{equation}
 \label{eq:g1d}
-g\left(\bar{\C}, \bar{\C}\right) = 
+g\big(\tC, \tC \big) = 
 \dfrac{2}{n^2} \sum_{\ell=1}^n (2\ell - 1 - n) x_\ell .
@@
 Note that the cost of computing 
-\eqref{eq:g1d}
+$g\big( \tC, \tC \big)$
 is $\OO(n)$ and the cost of
@@
 Assuming that each partition is ordered,  
-$\mathbb{X} = \bigcup_{j=1}^k \bar{\C}_j$,
+$\mathbb{X} = \bigcup_{j=1}^k \tC_j$,
 the within energy dispersion
-\eqref{eq:within} can be written as
+can be written explicitly as
 \begin{equation}
 \label{eq:w1d}
-W\left( \bar{\C}_1,\dotsc,\bar{\C}_k \right) = 
+W\big( \tC_1,\dotsc,\tC_k \big) = 
 \sum_{j=1}^k \sum_{\ell=1}^{n_j} \dfrac{2\ell - 1 - n_j}{n_j} \, x_\ell.
@@
 
-For a two-class problem we can use \eqref{eq:w1d} to cluster the data
+For a two-class problem we can use the formula
+\eqref{eq:w1d} to cluster the data
 through a simple algorithm 
 as follows. We first order
-the entire dataset, $\mathbb{X} \to \bar{\mathbb{X}}$. Then 
-we compute \eqref{eq:w1d} for each possible split of $\bar{\mathbb{X}}$
+the entire dataset, $\mathbb{X} \to \widetilde{\mathbb{X}}$. Then 
+we compute \eqref{eq:w1d} for each possible split of $\widetilde{\mathbb{X}}$
 and pick the point which gives the minimum value of $W$.
-This procedure is described in Algorithm~\ref{algo1d}. 
-Note that
-this method does not require any initialization,
+This procedure is described in Algorithm~\ref{algo1d} and called
+$\mathcal{E}^{1D}$-clustering.
+Note that this algorithm is exact in the sense that it does not involve
+any random initialization,
 however,
 it only works for one-dimensional data with Euclidean distance. The total
-complexity of the algorithm is $\OO(n\log n + n^2) = \OO(n^2)$.
-% @gui: "does not require any initiializaton" i think you mean is guaranteed to find the optimal solution in O(n^2) time.
-% @jovo: it does not require initialization in the sense that we do not
-% assign points to clusters in an initial step, as k-means++ for instance, or
-% random initialization. This algorithm is "exact" in the sense that nothing
-% is random in it, and multiple runs will always give the same answer
+complexity of $\mathcal{E}^{1D}$-clustering is $\OO(n\log n + n^2) = \OO(n^2)$.
 
 \begin{figure}
-\begin{algorithm}[H]\vspace{.5em}
+\begin{flushleft}
+\begin{algorithm}[H]
+\vspace{.5em}
 \begin{algorithmic}[1]
@@
 \STATE sort $\mathbb{X}$ obtaining 
-$\bar{\mathbb{X}}= [ x_1,\dotsc,x_n ]$
+$\widetilde{\mathbb{X}}= [ x_1,\dotsc,x_n ]$
     \FOR{$j\in [ 1,\dotsc,n ]$}
-        \STATE Let $\bar{\C}_1^{(j)} = [x_i: i=1,\dotsc,j]$ and 
-                $\bar{\C}_2^{(j)} = [x_i : i=j+1,\dotsc,n]$
+        \STATE $\tC_{1,j} \leftarrow [x_i: i=1,\dotsc,j]$, and 
+                $\tC_{2,j} \leftarrow [x_i : i=j+1,\dotsc,n]$
         \STATE  
-            $W^{(j)} \leftarrow W \big( \bar{\C}_1^{(j)},\bar{\C}_2^{(j)}  
-            \big)$ from \eqref{eq:w1d}
+            $W^{(j)} \leftarrow W \big( \tC_{1,j},\tC_{2,j}\big)$, 
+            from \eqref{eq:w1d}
     \ENDFOR
@@
 \label{algo1d}
-Approximate solution to \eqref{eq:minimize} 
+$\mathcal{E}^{1D}$-clustering algorithm to
+find local solutions to the optimization 
+problem \eqref{eq:minimize} 
 for a two-class problem in one dimension. \hspace{\fill}
@@
 \end{algorithm}
+\end{flushleft}
 \end{figure}
@@
 $0$ to all points wrongly clustered.
-For a two-class problem with balanced
-clusters, the value $1/2$ correspond
+For a balanced two-class problem the value $1/2$ correspond
 to chance.
@@
 \caption{
-% @gui: to provide intuition to the reader, let's show the kernel density estimate for each of the 3 algorithms for both classes.
-i'm not sure the best way to visualize it, maybe play with a few options.
-% @gui: also, the colors are not very different.  i use the following color pallete:
-%co = [0,255,0; % green
-%0,255,255;  % cyan
-%255,0,255;  % magenta
-%255,127,0;  % orange
-%55,126,184; % blue
-%153,153,153;% gray
-%152,78,163; % purple
-%166,86,40;  % brown
-%247,129,191;% pink
-%77,175,74;  % forest
-%228,26,28;  % red
-%255,255,51];% yellow
-% @jovo: ok, I will change the colors later, will have to regenerate the plots.
-% The problem with generating the kernel density for log-normal is that we need
-% a lot of points to get a curve like the one in the histogram. If I do this
-% with only 3000 points as in the graph, I will get a pretty bad curve that I
-% don't think will look like anything the histogram. Also, this is only a
-% simple example in 1D ... do you think it worth the effort?
-% If you do I can try.
 \label{fig:1d}
@@
 
-Let us consider two simple experiments with equal number of points
+We now consider two simple experiments with equal number of points
 in each cluster. We plot the accuracy \eqref{eq:accuracy} versus the number
@@
 The data is clustered using 
-Algorithm~\ref{algo1d}, 
+the $\mathcal{E}^{1D}$-clustering algorithm,
 % @gui: i see why you call this algorithm 1, but can we call it something with semantic meaning? like 1D-\mathcal{E}-clustering? or maybe \mathcal{E}1 for short? same for the other variants.
@@
 % k-means. I called $k$-means++ though
-We use the initialization from $k$-means++ \cite{Vassilvitskii} also
-for GMM.
-We run the algorithms $100$
+We use the initialization procedure from 
+$k$-means++ \cite{Vassilvitskii} also for GMM.
+We run the algorithms $10$
 times choosing the result with  best objective function value. 
@@
 % @jovo: ok
-Notice that Algorithm~\ref{algo1d} requires no initialization so we only
+Notice that $\mathcal{E}^{1D}$-clustering does not require any random
+initialization so we only
 % @gui: "no initiazation" is weird.  do any algs require an initialization?
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Iterative Algorithms for Energy Statistics Clustering}
+\section{Iterative Algorithms for Energy Clustering}
 \label{sec:algo}
@@
 Proposition~\ref{th:kernel_kmeans} we can also find an approximate
-solution by the well-known kernel $k$-means algorithm 
-(see \cite{Dhillon2,Dhillon}), which 
+solution by the well-known kernel $k$-means algorithm based
+on Lloyd's heuristic (see \cite{Dhillon2,Dhillon}), which 
 for convenience will also be restated in the present context.
@@
 
-\subsection*{Kernel $\bm{k}$-Means Algorithm}
+%\subsection*{Kernel $\bm{k}$-Means Algorithm}
+\subsection*{Lloyd's Method for Energy Clustering}
 
@@
 \dfrac{1}{n_\ell^2} Q_\ell
--\dfrac{2}{n_\ell} Q_\ell(x_i) 
-\end{equation}
-which represents a cost depending on point $x_i$ and cluster $\C_\ell$. One
-thus assigns  $x_i$ to cluster $\C_{j^\star}$ according
-to $j^\star = \argmin_\ell J^{(\ell)}(x_i)$, for $\ell = 1,\dotsc,k$.
-This procedure is performed for every data point and repeated until
+-\dfrac{2}{n_\ell} Q_\ell(x_i) .
+\end{equation}
+We are thus solving 
+\begin{equation}
+\min_{Z} 
+\sum_{i=1}^n 
+\sum_{\ell=1}^k Z_{i\ell} J^{(\ell)}(x_i).
+\end{equation}
+One possible strategy is to
+assign  $x_i$ to cluster $\C_{j^\star}$ according
+to 
+\begin{equation}
+j^\star = \argmin_{\ell=1,\dotsc,k} J^{(\ell)}(x_i) .
+\end{equation}
+This is done for every data point $x_i$ and repeated until
 convergence, i.e. until no new assignments are made.
-The entire procedure is described in Algorithm~\ref{kmeans_algo}.
+The entire procedure is described in Algorithm~\ref{kmeans_algo}, which
+we call $\mathcal{E}^L$-clustering to emphasize
+that we are optimizing the energy function $W$ from energy statistics
+based on Lloyd's method \cite{Lloyd}.
 % @gui: let's call this something more informative
-% @jovo: alright, will take care when re-generating the plots
+% @jovo: alright
 It can be shown that this algorithm converges provided $G$ is positive
@@
 
-Notice that to compute the first term in \eqref{eq:Jell} requires
-$\OO(n_\ell)$ operations, and although the second term requires
+% @jovo: check this paragraph. I included it to answer your question
+% about having K(x,x) and \ell=1,...,k without exluding \ell=j. Essentially,
+% the term K(x,x) will be hidden inside the mean of the cluster to which
+% x is currently assigned to. This also clarifies the connection to 
+% the standard formulation of kernel k-means.
+Actually, $\mathcal{E}^{L}$-clustering is preciselly
+kernel $k$-means algorithm \cite{Dhillon2,Dhillon} 
+but written in a more concise and clear manner. Indeed, recalling that
+$\kk(x,y)=\langle \varphi(x), \varphi(y) \rangle$ where $\varphi: \mathcal{X}
+\to \mathcal{H}_\kk$ is the feature map, we have from \eqref{eq:Jell} that
+\begin{equation}
+%\begin{split}
+J^{(\ell)}(x_i) 
+%&= \dfrac{1}{n_\ell^2} \sum_{x, y \in \C_\ell} 
+%\langle \varphi(x), \varphi(y) \rangle - \dfrac{2}{n_\ell}
+%\sum_{y\in \C_\ell} \langle \varphi(x_i), \varphi(y)\rangle \\ 
+= 
+\langle \varphi(\mu_\ell), \varphi(\mu_\ell) \rangle
+-2 \langle \varphi(x_i), \varphi(\mu_\ell) \rangle 
+= 
+\| \varphi(x_i) - \varphi(\mu_\ell) \|^2 - \| \varphi(x_i) \|^2,
+%\end{split}
+\end{equation}
+where $\mu_\ell = \tfrac{1}{n_\ell} \sum_{x\in \C_\ell}x$ 
+is the mean (center) of cluster $\C_\ell$. 
+Therefore, $\min_\ell J^{(\ell)}(x_i) = 
+\min_\ell \| \varphi(x_i) - \varphi(\mu_\ell)\|^2$, i.e. we are assigning
+$x_i$ to the cluster with closest center (in feature space),
+which is the familiar Lloyd's
+heuristic approach.
+
+To check the complexity of $\mathcal{E}^L$-clustering, 
+notice that to compute the second term of $J^{(\ell)}(x_i)$
+in \eqref{eq:Jell} requires
+$\OO(n_\ell)$ operations, and although the first term requires
 $\OO(n_\ell^2)$ it only needs to be computed once outside loop through
-data points (step 1).
-Therefore, the time complexity Algorithm~\ref{kmeans_algo} is
+data points (step 1 of Algorithm~\ref{kmeans_algo}).
+Therefore, the time complexity of $\mathcal{E}^L$-clustering 
+is
 $\OO(n k \max_\ell n_\ell) = \OO(k n^2)$. For a sparse
@@
 \begin{figure}
+\begin{flushleft}
 \begin{algorithm}[H]
@@
     \INPUT number of clusters $k$, Gram matrix $G$, initial label
-    matrix $Z = Z_0$
+    matrix $Z \leftarrow Z_0$
     \OUTPUT label matrix $Z$ 
   \STATE $q \leftarrow (Q_1, \dotsc, Q_k)^\top$ 
-            have the costs of each cluster, according to \eqref{eq:maxQ}
+            have the costs of each cluster, defined in \eqref{eq:maxQ}
   \STATE $n \leftarrow (n_1,\dotsc,n_k)^\top$ 
-        have the number of points in each cluster, obtained 
-        from $D = Z^\top Z$
+        have the number of points in each cluster%, obtained 
+        %from $D = Z^\top Z$
   \REPEAT
@@
         \STATE let $j$ be such that $x_i \in \C_j$
-        \STATE $j^\star \leftarrow \argmin_{\ell} J^{(\ell)}(x_i)$
-            according to \eqref{eq:Jell}, for $\ell=1,2,\dots,k$
+        \STATE $j^\star \leftarrow \argmin_{\ell=1,\dotsc,k} J^{(\ell)}(x_i)$,
+            where $J^{(\ell)}(x_i)$ is defined in \eqref{eq:Jell}
+            %, for $\ell=1,2,\dots,k$
         \IF{ $j^\star \ne j$} 
@@
 \caption{\label{kmeans_algo}
-Kernel $k$-means algorithm 
-to find a local solution to \eqref{eq:qcqp2}.
-\hspace{\fill}
+$\mathcal{E}^{L}$-clustering: Lloyd's method for energy clustering, which
+is preciselly kernel $k$-means algorithm. This procedure finds
+local solutions to the problem \eqref{eq:qcqp2}.
 }
 \end{algorithm}
+\end{flushleft}
 \end{figure}
 
-\subsection*{Hartigan's Method for Energy Statistics Clustering}
-
-We now consider an algorithm based on Hartigan's method \cite{Hartigan} 
-applied to \eqref{eq:maxQ}. This gives a local solution to 
-\eqref{eq:qcqp2}. 
-%This approach was also proposed in \cite{Kgroups}.
-The method is based on the change
-in the within energy statistic when moving a given data point to
-another cluster.
-Suppose point $x_i \in \mathcal{X}$
-is currently assigned to  cluster $\C_j$, yielding
-a total energy cost function \eqref{eq:maxQ} denoted by $Q^{(j)}$.
-Let us consider the change in the total within energy by moving
-$x_i$ to $\C_\ell$. 
-Denote this new energy cost by $Q^{(\ell)}$.
-A straightforward computation gives
+\subsection*{Hartigan's Method for Energy Clustering}
+
+We now consider Hartigan's method \cite{Hartigan} 
+applied to the optimization problem in the form \eqref{eq:maxQ}, which gives
+a local solution to the QCQP defined in \eqref{eq:qcqp2}. 
+The method is based in computing the maximum change
+in the total cost function $Q$ when moving each data point to
+another cluster. More specifically, 
+suppose point $x_i$
+is currently assigned to  cluster $\C_j$ yielding
+a total cost function denoted by $Q^{(j)}$.
+Moving $x_i$ to cluster $\C_\ell$ yields another total cost function
+denoted by $Q^{(\ell)}$. We are interested in computing the maximum 
+cost change
+$\Delta Q^{j\to \ell} (x_i) \equiv Q^{(\ell)} - Q^{(j)}$, for $\ell\ne j$. 
+From \eqref{eq:maxQ}, by explicitly writting the costs related to these 
+two cluster we obtain
+\begin{equation}
+\Delta Q^{j\to \ell} (x_i) = \dfrac{Q_\ell^{+}}{n_\ell+1} + 
+\dfrac{Q_j^-}{n_j-1} - \dfrac{Q_j}{n_j} - \dfrac{Q_\ell}{n_\ell}
+\end{equation}
+where $Q^{+}_\ell$ denote the cost of the new $\ell$th cluster
+with the point $x_i$ added to it, and $Q^-_j$ is the cost of new 
+$j$th cluster with $x_i$ removed from it. Noting that 
+$Q_\ell^{+} = Q_\ell + 2 Q_\ell(x_i) + G_{ii}$ and
+$Q_j^{-} = Q_j - 2 Q_j(x_i) + G_{ii}$, we get the formula
 \begin{equation}
 \label{eq:changeQ}
-\begin{split}
-\Delta Q^{j \to \ell}(x_i) &\equiv Q^{(\ell)} - Q^{(j)} \\ 
-&= 
-\dfrac{1}{n_j - 1}\left[ \dfrac{Q_j}{n_j} - 2 Q_j(x_i) \right]
-- \dfrac{1}{n_\ell + 1}\left[ \dfrac{Q_\ell}{n_\ell} - 2 \big(Q_\ell(x_i) + 
-\kk(x_i,x_i)\big) 
-\right].
-\end{split}
-\end{equation}
-Thus, if $\Delta Q^{j\to \ell}(x_i) > 0$ we get closer to a 
+\Delta Q^{j \to \ell}(x_i)  = 
+\dfrac{1}{n_j - 1}\left[ \dfrac{Q_j}{n_j} - 2 Q_j(x_i) + G_{ii} \right]
+- \dfrac{1}{n_\ell + 1}\left[ \dfrac{Q_\ell}{n_\ell} - 2 Q_\ell(x_i) 
+- G_{ii} \right].
+\end{equation}
+Therefore, if $\Delta Q^{j\to \ell}(x_i) > 0$ we get closer to a 
 maximum of \eqref{eq:maxQ} by
 moving $x_i$ to $\C_\ell$, otherwise we keep $x_i$ in $\C_j$. Based on
-this the algorithm goes as follows.
+this the proposed algorithm goes as follows.
 We start with an initial configuration for the label matrix $Z$, 
@@
 point $x_i$ 
-we compute the cost of moving it to another cluster $\C_\ell$,
+we compute the cost of moving it to another cluster $\C_\ell$, i.e.
 $\Delta Q^{j\to \ell}(x_i)$ for 
-$\ell=1,\dots,k$ with $\ell \ne j$. Hence, let us choose
-$j^\star = \argmax_\ell \Delta^{j \to \ell}(x_i)$.
+$\ell=1,\dots,k$ with $\ell \ne j$ and $j$ denotes the index of its current
+partition, $x \in \C_j$. Hence, we choose
+\begin{equation}
+j^\star = \argmax_{\ell=1,\dotsc,k \, | \, \ell\ne j} 
+\Delta^{j \to \ell}(x_i).
+\end{equation}
 If $\Delta Q^{j \to j^\star}(x_i) > 0$ 
 we move $x_i$ to cluster $\C_{j^\star}$, otherwise 
-we keep $x_i$ in its original cluster $\C_j$. We update $Z$ accordingly.
-The process is repeated
+we keep $x_i$ in its original cluster $\C_j$. 
+This process is repeated
 until no points are assigned to new clusters. 
-This whole procedure is described in Algorithm~\ref{algo}.
-Note that this ensures that the objective function is
-monotonically increasing at each iteration and consequently the algorithm
+The entire procedure is explicitly described in Algorithm~\ref{algo}, which we
+coin $\mathcal{E}^H$-clustering to emphasize that it is based on
+Hartigan's method.
+This method automatically ensures that the objective function is
+monotonically increasing at each iteration, and consequently the algorithm
 converges in a finite number of steps.
@@
     \INPUT number of clusters $k$, Gram matrix $G$, 
-                initial label matrix $Z=Z_0$
+                initial label matrix $Z \leftarrow Z_0$
     \OUTPUT label matrix $Z$
   \STATE $q \leftarrow (Q_1, \dotsc, Q_k)^\top$ 
-            have the energy costs of each cluster, according to \eqref{eq:maxQ}
+            have the energy costs of each cluster, defined in \eqref{eq:maxQ}
   \STATE $n \leftarrow (n_1,\dotsc,n_k)^\top$ have the number of points 
-        in each cluster, obtained from $D=Z^\top Z$
+        in each cluster%, obtained from $D=Z^\top Z$
   \REPEAT
@@
         \STATE let $j$ be such that $x_i \in \C_j$
-        \STATE $j^\star \leftarrow \argmax_{\ell} \Delta Q^{j\to \ell}(x_i)$, 
-            for $\ell=1,2,\dots,k$ and $\ell \ne j$ \label{stepmove}
+        \STATE $j^\star \leftarrow \argmax_{\ell=1,\dotsc,k \, | \, \ell\ne j} 
+                \Delta Q^{j\to \ell}(x_i)$
+            %for $\ell=1,2,\dots,k$ and $\ell \ne j$; 
+            using Eq. \eqref{eq:changeQ} \label{stepmove}
         \IF{ $\Delta Q^{j \to j^\star}(x_i) > 0$ }
@@
                     $n_{j^\star} \leftarrow n_{j^\star} + 1$
-            \STATE update $q$: $q_j \leftarrow q_j - 2Q_j(x_i)$ and
-    $q_{j^\star} \leftarrow q_{j^\star} + 2\left(Q_{j^\star}(x_i)+
-    G_{ii}\right)$
+            \STATE update $q$: $q_j \leftarrow q_j - 2Q_j(x_i) + G_{ii}$ and
+    $q_{j^\star} \leftarrow q_{j^\star} + 2Q_{j^\star}(x_i)+ G_{ii}$
     %    \ELSE
@@
 \caption{\label{algo}
-Hartigan's method to find a local solution to \eqref{eq:qcqp2}.
+$\mathcal{E}^H$-clustering: Hartigan's method for energy clustering.
+This algorithm finds local solutions to  
+the optimization problem \eqref{eq:qcqp2}.
 \hspace{\fill}
@@
 \end{algorithm}
+\end{flushleft}
 \end{figure}
 
-Computing $G$ requires $\OO( D n^2)$ operations, where 
+The complexity analysis of $\mathcal{E}^H$-clustering is the following.
+Computing the Gram matrix $G$ requires $\OO( D n^2)$ operations, where 
 $D$ is the dimension of each data point and $n$ is the data size. However,
-both Algorithms~\ref{kmeans_algo} and \ref{algo} 
+both algorithms $\mathcal{E}^L$- and $\mathcal{E}^H$-clustering 
 assume that $G$ is given. There are more efficient
@@
 entire dataset this gives a time complexity
-of $\OO(n k  \max_j n_j) =\OO(k n^2)$. This is the same cost as
-in kernel $k$-means, Algorithm~\ref{kmeans_algo}. Again, if $G$ is sparse
+of $\OO(n k  \max_j n_j) =\OO(k n^2)$. Note that this is the same cost as
+in $\mathcal{E}^L$-clustering, or kernel $k$-means algorithm. 
+Again, if $G$ is sparse
 this can be reduced to $\OO(k n')$ where $n'$ is the number of nonzero
@@
 
-In the following mention some important results about Hartigan's method.
+In the following we mention some important known 
+results about Hartigan's method.
 % @gui: "let us" again...
@@
 %by Lloyd's method, 
-Neither of these two conditions are satisfied  by Lloyd's method,
+Neither of these two conditions are guaranteed to be 
+satisfied  by Lloyd's method,
 % @gui: do you mean Lloyd's method satisfies neither, or does not satisfy both?  please clarify
 % @jovo: ok
-and consequently by (kernel) $k$-means algorithm.
-The next result indicates that Hartigan's can potentially 
+and consequently by $\mathcal{E}^L$-clustering 
+algorithm.
+The next result indicates that Hartigan's method can potentially 
 escape local optima of Lloyd's method.
@@
 
-This means that Algorithm~\ref{kmeans_algo} cannot
-improve on a local optima of Algorithm~\ref{algo}. On the other hand,
-Algorithm~\ref{algo} might improve on a local optima of 
-Algorithm~\ref{kmeans_algo}. Lloyd's method forms Voronoi partitions,
+The above theorem implies that $\mathcal{E}^L$-clustering  cannot
+improve on a local optima of $\mathcal{E}^H$-clustering. On the other hand,
+$\mathcal{E}^H$-clustering might improve on a local optima of 
+$\mathcal{E}^L$-clustering. Lloyd's method forms Voronoi partitions,
 while Hartigan's method groups data
@@
 but no explicit complexity was provided. In our case, we showed that both
-Algorithms~\ref{kmeans_algo} and \ref{algo} have the same time complexity.
+$\mathcal{E}^L$- and $\mathcal{E}^H$-clustering
+have the same time complexity.
 
 In \cite{Slonin} Hartigan's method was applied to $k$-means problem
-with any Bregman divergence. They showed that the number of Hartigan's
+with any Bregman divergence. It was shown that the number of Hartigan's
 local minima is upper bounded by $\mathcal{O}(1/k)$ 
 \cite[Proposition 5.1]{Slonin}. 
-In addition, they provide examples where
+In addition, it was provided examples where
 \emph{any} initial partition correspond to a local optima of Lloyd's 
@@
 local optima.
-These results indicate that Hartigan's method
-provides several advantages over Lloyd's method. 
-We will see this explicitly in the following 
-numerical results where Algorithm~\ref{algo}
-outperforms Algorithm~\ref{kmeans_algo} when using the same kernel.
+
+The above results indicate that Hartigan's method
+provides several advantages over Lloyd's method, a fact that will also
+be supported by our numerical experiments in the next section where 
+$\mathcal{E}^H$-clustering  outperforms
+of $\mathcal{E}^L$-clustering (kernel $k$-means) in several settings.
 
@@
 the point $x_0=0$ is chosen in the corresponding kernel  
-\eqref{eq:kernel_semimetric}. Therefore,
+\eqref{eq:kernel_semimetric}. We therefore use
 \begin{equation}
@@
 
-Let us briefly mention that we compared Algorithm~\ref{algo} and
-Algorithm~\ref{algo1d} for several univariate distributions and
-both perform almost indistinguishable regarding the clustering quality.
+Let us briefly mention that we compared $\mathcal{E}^H$-clustering,
+as described in Algorithm~\ref{algo}, to
+the $1D\,\mathcal{E}$-clutering, described in Algorithm~\ref{algo1d}, 
+for several univariate distributions.
+Both perform almost indistinguishable regarding the clustering quality.
 However, we omit these results since we will analyse more interesting 
@@
 Consider the effect of having 
-unbalanced clusters accordint to
+unbalanced clusters according to
 \begin{equation}
@@
 \end{equation}
-which is computed from a sample  $\{ x_i \}_{i=1}^n$.
+which is computed from a 
+%sample  
+given dataset $\{ x_i \}_{i=1}^n$.
 % @gui: it is cheating if this sample is part of the data, this should be a sample of "training data", not used for subsequent testing. please clarify
-% @jovo: I'm not sure what to say here, cuz all I'm doing is, given a data set
-% I can compute \sigma as above. Then I use this in the kernel and perform
+% @jovo: All I'm doing is, given a data set
+% I compute \sigma as above. Then I use this in the kernel and perform
 % clustering. How can this be cheating if I'm not using any labels, or any
@@
 We considered clustering from the perspective of energy
-statistics theory which provides a nonparametric test for 
+statistics theory, coined energy clustering for short, 
+which provides a nonparametric test for 
 equality of distributions.
-We showed that the clustering problem reduces to a quadratically
+We showed that the mathematical formulation of energy clustering 
+reduces to a quadratically
 constrained quadratic program, 
 as described in Proposition~\ref{th:qcqp2}.
-Moreover, this problem is equivalent
-to kernel $k$-means optimization problem once the kernel is fixed; see
+Moreover, the energy clustering optimization problem is equivalent
+to kernel $k$-means optimization problem, once the kernel is fixed; see
 Proposition~\ref{th:kernel_kmeans}. Energy statistics, however, fixes
-a family of standard kernels consistent with \eqref{eq:energy2}, and
-more general kernels related to \eqref{eq:energy3} can be obtained.
-Our results imply that kernel $k$-means optimization problem
-can be interpreted as a consequence of energy statistics based clustering.
+a family of standard kernels consistent with the energy 
+distance \eqref{eq:energy2}.
+More general kernels 
+%related to the energy distance \eqref{eq:energy3},
+defined on semimetric spaces of negative type can also be obtained.
+%Our results imply that kernel $k$-means optimization problem
+%can be interpreted as a consequence of energy clustering.
 % @gui: "this"? what is "this:? clarify
@@
 clustering and graph partitioning problems.
-
-We considered Algorithm~\ref{algo} based on Hartigan's method and
-compared with 
-kernel $k$-means, described in Algorithm~\ref{kmeans_algo}, which
-is based on Lloyd's heuristic.
-Both have the same time complexity, however, the numerical 
-results provide compelling evidence that Algorithm~\ref{algo} 
+We proposed the $\mathcal{E}^H$-clustering algorithm based on 
+Hartigan's method, and compared with $\mathcal{E}^L$-clustering algorithm
+based on Lloyd's heuristic.
+%, which is the same as kernel $k$-means algorithm.
+Both algorithms have the same time complexity, however, the numerical 
+results provide compelling evidence that $\mathcal{E}^H$-clustering
 is more robust with
-a superior clustering performance. This is also theoretically
-supported as described in the end of Section~\ref{sec:algo}.
-
+a superior clustering performance. The fact that $\mathcal{E}^H$-clustering
+has more desirable properties compared to $\mathcal{E}^L$-clustering
+is also supported by theoretical results, 
+as described in the end of Section~\ref{sec:algo}.
+
+It would be interesting to formally 
+demonstrate cases where energy clustering is a 
+consistent estimator. A soft version of energy clustering is also an
+interesting extension.
 Finally, kernel methods can benefit from sparsity and
-fixed-rank approximations of the Gram matrix.
-For instance, an approach to make kernel $k$-means scalable
-was recently proposed \cite{Mahoney}, and a modified technique for
-rank reduction in Nystr\"om method was also recently considered 
-\cite{Becker}. It would be interesting to apply these and related
-techniques to the case of Algorithm~\ref{algo}.
-Another fruitful avenue of exploration would be to find 
-better methods to tackle the optimization problem \eqref{eq:qcqp2}.
+fixed-rank approximations of the Gram matrix, and there is plenty
+of room to make $\mathcal{E}^H$-clustering algorithm more scalable.
+%For instance, an approach to make kernel $k$-means scalable
+%was recently proposed \cite{Mahoney}, and a modified technique for
+%rank reduction in Nystr\"om method was also recently considered 
+%\cite{Becker}. It would be interesting to apply these and related
+%techniques to the case of $\mathcal{E}^H$-clustering algorithm.
+%Another fruitful avenue of exploration would be to find 
+%better methods to tackle the optimization problem \eqref{eq:qcqp2}.
 

### 2017-09-30 energy_clustering/code/energy/bayes.py

@@ -1,0 +1,200 @@
+"""Functions to compute Bayes errors."""
+
+# Guilherme S. Franca <guifranca@gmail.com>
+# Johns Hopkins, Neurodata
+
+
+from __future__ import division
+
+from functools import partial
+from scipy.optimize import minimize
+from scipy.optimize import root
+import numpy as np
+import scipy.stats as stats
+import scipy.integrate as integrate
+import matplotlib.pyplot as plt
+
+import data
+
+
+def bayes_univariate_normal(mu1, mu2, sigma1, sigma2, 
+                            num_times=10, num_points=1000):
+    """Estimate Bayes' error for two-class normal distributions in 1D."""
+    error = np.zeros(num_times)
+    for i in range(num_times):
+        x1 = np.random.normal(mu1, sigma1, num_points)
+        x2 = np.random.normal(mu2, sigma2, num_points)
+    
+        error1 = sum([1 for x in x1 
+            if stats.norm.pdf(x,mu1,sigma1)-stats.norm.pdf(x,mu2,sigma2)<=0
+        ])
+        error2 = sum([1 for x in x2 
+            if stats.norm.pdf(x,mu1,sigma1)-stats.norm.pdf(x,mu2,sigma2)>=0
+        ])
+        error[i] = error1+error2
+    return 1-error.mean()/(2*num_points)
+
+def integrate_univariate_normal(mu1, mu2, sigma1, sigma2):
+    """Same as above but through numerical integration."""
+    
+    g1 = partial(stats.norm.pdf, loc=mu1, scale=sigma1)
+    g2 = partial(stats.norm.pdf, loc=mu2, scale=sigma2)
+    
+    # make it visual to confirm
+    #xs = np.arange(-10,20, 0.001)
+    #y1s = np.array([g1(x) for x in xs])
+    #y2s = np.array([g2(x) for x in xs])
+    #y3s = np.array([np.abs(g1(x)-g2(x)) for x in xs])
+    #plt.plot(xs, y1s)
+    #plt.plot(xs, y2s)
+    #plt.plot(xs, y3s)
+    #plt.show()
+
+    x0 = root(lambda x: g1(x) - g2(x), 1).x[0]
+    
+    # another way
+    #cons = (
+    #    {'type': 'ineq', 'fun': lambda x: x-1},
+    #    {'type': 'ineq', 'fun': lambda x: -(x-2)},
+    #)
[…]

### 2017-09-30 energy_clustering/code/energy/data.py

@@ -1,0 +1,354 @@
+"""
+Generate Data
+
+Some usefull functions to generate or manipulate data.
+Also some functions to plot data for visualization.
+
+"""
+
+# Guilherme Franca <guifranca@gmail.com>
+# Johns Hopkins Unversity, Neurodata
+
+
+from __future__ import division
+
+import numpy as np
+import matplotlib.pyplot as plt
+import numbers
+import pylab
+
+from sklearn.decomposition import PCA
+
+from customize_plots import *
+
+
+def multivariate_normal(means, sigmas, ns):
+    """Generate several normal distributed data points.
+    
+    Input: 
+        means = [m1, m2, ...]
+        sigmas = [s1, s2, ...]
+        ns = [n1, n2, ...]
+    
+    Output: data X and labels z.
+
+    Each parameter in these lists are associated to one multivariate
+    normal distribution. The data is shuffled so points are not ordered.
+    
+    """
+    n = len(means)
+    X = []
+    z = []
+    k = 0
+    for mu, sigma, n in zip(means, sigmas, ns):
+        X.append(np.random.multivariate_normal(mu, sigma, n))
+        z.append([k]*n)
+        k += 1
+    X = np.concatenate(X)
+    z = np.concatenate(z)
+    idx = np.random.permutation(sum(ns))
+    return X[idx], z[idx]
+
+def univariate_normal(means, sigmas, ns):
+    """Generate several univariate normal distributed data points.
+    
+    Input: 
+        means = [m1, m2, ...]
+        sigmas = [s1, s2, ...]
+        ns = [n1, n2, ...]
+    
+    Output: data X and labels z.
+    
+    """
+    n = len(means)
+    X = []
+    z = []
+    k = 0
+    for mu, sigma, n in zip(means, sigmas, ns):
+        X.append(np.random.normal(mu, sigma, n))
+        z.append([k]*n)
+        k += 1
+    X = np.concatenate(X)
+    z = np.concatenate(z)
+    idx = np.random.permutation(sum(ns))
+    return X[idx], z[idx]
+
+def univariate_lognormal(means, sigmas, ns):
+    """Return data according to log normal distribtion.
+
+    Input: means = [mu1, mu2, ...], 
[…]

### 2017-09-30 energy_clustering/code/energy/eclust.py

@@ -1,0 +1,326 @@
+"""Iterative algorithms to optimize QCQP for energy statistics clustering."""
+
+# Guilherme Franca <guifranca@gmail.com>
+# Johns Hopkins University, Neurodata
+
+
+from __future__ import division
+
+import numpy as np
+from sklearn.metrics.pairwise import pairwise_distances
+from sklearn.base import BaseEstimator, ClusterMixin
+from sklearn.utils import check_random_state
+
+
+def ztoZ(z):
+    """Convert label vector to label matrix."""
+    n = z.shape[0]
+    k = np.unique(z).shape[0]
+    Z = np.zeros((n,k))
+    for i in range(n):
+        Z[i, int(z[i])] = 1
+    return Z
+
+def Ztoz(Z):
+    """Convert label matrix to label vector."""
+    n, k = Z.shape
+    z = np.zeros(n)
+    points, labels = np.where(Z==1)
+    for i in range(len(points)):
+        z[points[i]] = labels[i]
+    return z
+
+def objective(Z, G):
+    """Compute objective function."""
+    Zt = Z.T
+    D = np.dot(Zt, Z)
+    Dinv = np.linalg.inv(D)
+    return np.trace(Dinv.dot(Zt.dot(G.dot(Z))))
+
+def kernel_function(x, y, rho, x0=None):
+    """Return kernel function based on rho."""
+    if x0 == None:
+        x0 = np.zeros(x.shape)
+    val = 0.5*(rho(x,x0) + rho(y,x0) - rho(x,y))
+    return val
+
+def kernel_matrix(X, rho, x0=None):
+    """Compute Kernel matrix based on kernel function K(x,y)."""
+    kfunc = lambda x, y: kernel_function(x, y, rho, x0)
+    G = pairwise_distances(X, metric=kfunc, n_jobs=4)
+    return G
+
+def energy_clustering_brute(k, G, Z0, max_iter=100, tol=1e-4, verbose=False):
+    """Optimize energy QCQP in a brute force way. This is expensive,
+    but it works. We move points to other partitions and compute the change
+    in the cost function.
+
+    """
+    
+    n = G.shape[0]
+    Z = np.copy(Z0)
+    count = 0
+    converged = False
+    while not converged and count < max_iter:
+        
+        n_changed = 0
+        
+        for l in range(n):
+            
+            curr_label = np.where(Z[l]==1)[0][0]
+            Ws = []
+            
+            for j in range(k):
+                # for each point, we remove from its original partition
+                # then append it to a new partition
[…]

### 2017-09-30 energy_clustering/code/experiment3.py

@@ -1,0 +1,96 @@
+"""Third experiment. Normal distributions for unbalanced clusters."""
+
+# Guilhere Franca <guifranca@gmail.com>
+# Johns Hopkins University, Neurodata
+
+import numpy as np
+import multiprocessing as mp
+
+import energy.data as data
+import energy.eclust as eclust
+import energy.initialization as initialization
+import run_clustering
+import plot
+
+def gauss_dimensions_pi(num_points=range(0, 180, 10), num_experiments=10):
+    """Test unbalanced clusters."""
+    k = 2
+    D = 4
+    d = 2
+    N = 210
+    table = np.zeros((num_experiments*len(num_points), 5))
+    count = 0
+
+    for p in num_points:
+        for i in range(num_experiments):
+
+            # generate data
+            m1 = np.zeros(D)
+            s1 = np.eye(D)
+            m2 = np.concatenate((1.5*np.ones(d), np.zeros(D-d)))
+            s2 = np.diag(np.concatenate((.5*np.ones(d), np.ones(D-d))))
+            n1 = N-p
+            n2 = N+p
+
+            X, z = data.multivariate_normal([m1, m2], [s1, s2], [n1, n2])
+            rho = lambda x, y: np.linalg.norm(x-y)
+            G = ke.kernel_matrix(X, rho)
+
+            table[count, 0] = p
+            table[count, 1] = eclust.energy_hartigan(k, X, G, z, 
+                                    init="spectral", run_times=1)
+            table[count, 2] = eclust.energy_lloyd(k, X, G, z, 
+                                    init="spectral", run_times=1)
+            table[count, 3] = eclust.kmeans(k, X, z)
+            table[count, 4] = eclust.gmm(k, X, z)
+
+            count += 1
+
+    return table
+
+def make_plot(*data_files):
+    table = []
+    for f in data_files:
+        t = np.loadtxt(f, delimiter=',')
+        t[:,0] = np.array([int(D) for D in t[:,0]])
+        table.append(t)
+    table = np.concatenate(table)
+
+    ## customize plot below ##
+    p = plot.ErrorBar()
+    p.xlabel = 'number of unbalanced points'
+    p.legends = [r'$\mathcal{E}^{H}$-clustering', 
+                 r'$\mathcal{E}^{L}$-clustering', 
+                 r'$k$-means', 
+                 'GMM']
+    p.colors = ['b', 'r', 'g', 'm']
[…]

### 2017-09-30 energy_clustering/code/plot.py

@@ -1,0 +1,85 @@
+"""Plot accuracy results."""
+
+# Guilhere Franca <guifranca@gmail.com>
+# Johns Hopkins University, Neurodata
+
+from __future__ import division
+
+import numpy as np
+import scipy.stats
+import matplotlib as mpl
+import matplotlib.pyplot as plt
+
+from customize_plots import *
+
+
+class ErrorBar:
+
+    def __init__(self):
+        self.xlabel = 'dimension'
+        self.ylabel = 'accuracy'
+        self.symbols = ['o', 's', 'D', 'v']
+        self.legends = ['energy', r'$k$-means', 'GMM', r'kernel $k$-means']
+        self.colors = ['b', 'r', 'g', 'c']
+        self.xlim = None
+        self.ylim = None
+        self.bayes = None
+        self.loc = 0
+        self.doublex = False
+        self.legcols = 1
+        self.output = 'plot.pdf'
+
+    def _set_params(self, table):
+        self.col0 = col0 = np.array([int(x) for x in table[:,0]])
+        self.xs = np.unique(col0)
+        
+        self.n = len(table[0][1:])
+        #self.colors = iter(plt.cm.brg(np.linspace(0,1,self.n+1)))
+        self.colors = iter(self.colors)
+        self.legends = iter(self.legends)
+        self.symbols = iter(self.symbols)
+    
+        if self.doublex:
+            self.xs2 = 2*self.xs
+        else:
+            self.xs2 = self.xs
+            
+        if self.bayes:
+            if not isinstance(self.bayes, list):
+                self.bayes = [self.bayes]*len(self.xs)
+
+    def make_plot(self, table):
+        self._set_params(table)
+
+        fig = plt.figure()
+        ax = fig.add_subplot(111)
+    
+        for i in range(1, self.n+1):
+            
+            c = next(self.colors)
+            l = next(self.legends)
+            m = next(self.symbols)
+
+            mean_ = np.array([table[np.where(self.col0==x)[0],i].mean() 
+                              for x in self.xs])
+            stderr_ = np.array([scipy.stats.sem(
+                table[np.where(self.col0==x)[0],i]) for x in self.xs])
+            if self.bayes:
+                ax.plot(self.xs2, self.bayes, '--', color='black', linewidth=1, 
+                        zorder=0)
+            ax.errorbar(self.xs2, mean_, yerr=stderr_, 
[…]

### 2017-09-30 energy_clustering/code/run_clustering.py

@@ -1,0 +1,108 @@
+"""Functions to run clustering algorithms. Just need to call these functions
+in the tests.
+"""
+
+# Guilhere Franca <guifranca@gmail.com>
+# Johns Hopkins University, Neurodata
+
+import numpy as np
+from sklearn.cluster import KMeans
+from sklearn.mixture import GMM
+
+import energy.initialization as initialization
+import energy.metric as metric
+import energy.eclust as eclust 
+import energy.energy1d as energy1d
+
+def initialize(init, k, G, X):
+        if init == "spectral":
+            z0 = initialization.topeigen(k, G)
+            Z0 = eclust.ztoZ(z0)
+        elif init == "kmeans++":
+            z0 = initialization.kmeanspp(k, X)
+            Z0 = eclust.ztoZ(z0)
+        elif init == "random":
+            z0 = np.random.randint(0, k, len(X))
+            Z0 = eclust.ztoZ(z0)
+        else:
+            raise ValueError("No initialization method provided")
+        return Z0
+
+def energy_lloyd(k, X, G, z, run_times=1, init="spectral"):
+    """Run few times and pick the best objective function value."""
+    best_score = -np.inf
+    for rt in range(run_times):
+        Z0 = initialize(init, k, G, X)
+        zh = eclust.energy_lloyd(k, G, Z0, max_iter=300)
+        Zh = eclust.ztoZ(zh)
+        score = eclust.objective(Zh, G)
+        
+        if score > best_score:
+            best_score = score
+            best_z = zh
+
+    return metric.accuracy(z, best_z)
+
+def energy_hartigan(k, X, G, z, run_times=1, init="spectral"):
+    """Run few times and pick the best objective function value."""
+    best_score = -np.inf
+    for rt in range(run_times):
+        Z0 = initialize(init, k, G, X)
+        zh = eclust.energy_hartigan(k, G, Z0, max_iter=300)
+        Zh = eclust.ztoZ(zh)
+        score = eclust.objective(Zh, G)
+        
+        if score > best_score:
+            best_score = score
+            best_z = zh
+
+    return metric.accuracy(z, best_z)
+
+def energy1D(X, z):
+    """Energy clustering in 1 dimension. No need to run multiple times since
+    this is exact.
+    
+    """
+    zh, cost = energy1d.two_clusters1D(X)
+    return metric.accuracy(z, zh)
+
+def kmeans(k, X, z):
[…]

### 2017-09-30 qcqp_clustering/code/eclust/bayes.py

@@ -1,201 +1,0 @@
-"""Functions to compute Bayes errors."""
-
-# Guilherme S. Franca <guifranca@gmail.com>
-# Johns Hopkins, Neurodata
-
-
-from __future__ import division
-
-from functools import partial
-from scipy.optimize import minimize
-from scipy.optimize import root
-import numpy as np
-import scipy.stats as stats
-import scipy.integrate as integrate
-import matplotlib.pyplot as plt
-
-import data
-
-
-def bayes_univariate_normal(mu1, mu2, sigma1, sigma2, 
-                            num_times=10, num_points=1000):
-    """Estimate Bayes' error for two-class normal distributions in 1D."""
-    error = np.zeros(num_times)
-    for i in range(num_times):
-        x1 = np.random.normal(mu1, sigma1, num_points)
-        x2 = np.random.normal(mu2, sigma2, num_points)
-    
-        error1 = sum([1 for x in x1 
-            if stats.norm.pdf(x,mu1,sigma1)-stats.norm.pdf(x,mu2,sigma2)<=0
-        ])
-        error2 = sum([1 for x in x2 
-            if stats.norm.pdf(x,mu1,sigma1)-stats.norm.pdf(x,mu2,sigma2)>=0
-        ])
-        error[i] = error1+error2
-    return 1-error.mean()/(2*num_points)
-
-def integrate_univariate_normal(mu1, mu2, sigma1, sigma2):
-    """Same as above but through numerical integration."""
-    
-    g1 = partial(stats.norm.pdf, loc=mu1, scale=sigma1)
-    g2 = partial(stats.norm.pdf, loc=mu2, scale=sigma2)
-    
-    # make it visual to confirm
-    #xs = np.arange(-10,20, 0.001)
-    #y1s = np.array([g1(x) for x in xs])
-    #y2s = np.array([g2(x) for x in xs])
-    #y3s = np.array([np.abs(g1(x)-g2(x)) for x in xs])
-    #plt.plot(xs, y1s)
-    #plt.plot(xs, y2s)
-    #plt.plot(xs, y3s)
-    #plt.show()
-
-    x0 = root(lambda x: g1(x) - g2(x), 1).x[0]
-    
-    # another way
-    #cons = (
-    #    {'type': 'ineq', 'fun': lambda x: x-1},
-    #    {'type': 'ineq', 'fun': lambda x: -(x-2)},
-    #)
[…]

### 2017-09-30 qcqp_clustering/code/eclust/data.py

@@ -1,358 +1,0 @@
-"""
-Generate Data
-=============
-
-Some usefull functions to generate or manipulate data.
-For instance, generate data from multivariate normal distribution,
-circles, lognormal, etc. Also some functions to plot data for
-visualization.
-
-"""
-
-# Guilherme Franca <guifranca@gmail.com>
-# Johns Hopkins Unversity, Neurodata
-
-
-from __future__ import division
-
-import numpy as np
-import matplotlib.pyplot as plt
-import numbers
-import pylab
-
-from sklearn.decomposition import PCA
-
-from pplotcust import *
-
-
-def multivariate_normal(means, sigmas, ns):
-    """Generate several normal distributed data points.
-    
-    Input: 
-        means = [m1, m2, ...]
-        sigmas = [s1, s2, ...]
-        ns = [n1, n2, ...]
-    
-    Output: data X and labels z.
-
-    Each parameter in these lists are associated to one multivariate
-    normal distribution. The data is shuffled so points are not ordered.
-    
-    """
-    n = len(means)
-    X = []
-    z = []
-    k = 0
-    for mu, sigma, n in zip(means, sigmas, ns):
-        X.append(np.random.multivariate_normal(mu, sigma, n))
-        z.append([k]*n)
-        k += 1
-    X = np.concatenate(X)
-    z = np.concatenate(z)
-    idx = np.random.permutation(sum(ns))
-    return X[idx], z[idx]
-
-def univariate_normal(means, sigmas, ns):
-    """Generate several univariate normal distributed data points.
-    
-    Input: 
-        means = [m1, m2, ...]
-        sigmas = [s1, s2, ...]
-        ns = [n1, n2, ...]
-    
-    Output: data X and labels z.
-    
-    """
-    n = len(means)
-    X = []
-    z = []
-    k = 0
-    for mu, sigma, n in zip(means, sigmas, ns):
-        X.append(np.random.normal(mu, sigma, n))
-        z.append([k]*n)
-        k += 1
-    X = np.concatenate(X)
-    z = np.concatenate(z)
-    idx = np.random.permutation(sum(ns))
-    return X[idx], z[idx]
-
-def univariate_lognormal(means, sigmas, ns):
[…]

### 2017-09-30 qcqp_clustering/code/eclust/gmm.py

@@ -1,163 +1,0 @@
-"""
-EM algorithm for GMMM
-
-We implement the Expectation Maximization (EM) algorithm for Gaussian
-Mixture Models (GMM). The algorithm is described as follows:
-
-1. Initialize \mu_k, \Sigma_k, and \pi_k. Evaluate log likelihood function.
-
-2. E Step: evaluate responsabilities using the current values of parameters.
-
-3. M Step: re-estimate \mu_k, \Sigma_k, and \pi_k using the updated
-   responsabilities.
-
-4. Evaluate log likelihood function and check for convergence. 
-   Repeat from step 2 if not converged.
-
-
-"""
-
-from __future__ import division
-
-import numpy as np
-import matplotlib.pyplot as plt
-from scipy.stats import multivariate_normal
-from matplotlib.patches import Ellipse
-
-import kmeanspp
-
-
-class GMM:
-    """Fit a GMM using EM algorithm."""
-
-    def __init__(self, n_clusters, tol=0.001):
-        self.K = n_clusters
-        self.tol = tol
-    
-    def gauss(self, n, k):
-        """Return multidimensional gaussian at point x_n for component k, 
-        i.e. N(x_n | \mu_k, \sigma_k).
-
-        """
-        return multivariate_normal.pdf(self.x[n], self.mu[k], self.sigma[k])
-
-    def log_likelihood(self):
-        """Compute log likelihood function."""
-        f = 0
-        for n in range(self.N):
-            f += np.array([self.pi[k]*self.gauss(n, k) 
-                                for k in range(self.K)]).sum()
-        return f
-
-    def xxT(self, n, k):
-        x = self.x[n] - self.mu[k]
-        return np.outer(x, x)
-
-    def gmm_em(self):
-        """Implementation of EM algorithm."""
-        converged = False
-        old_loglh = self.log_likelihood()
-
-        while not converged:
-            
-            # compute responsabilities, E-step
-            for n in range(self.N):
-                total = sum([self.pi[j]*self.gauss(n, j) 
-                                        for j in range(self.K)])
-                for k in range(self.K):
-                    self.gamma[n, k] = self.pi[k]*self.gauss(n, k)/total
-            
-            # compute parameters, M-step
-            for k in range(self.K):
-                Nk = self.gamma[:,k].sum()
[…]

### 2017-09-30 qcqp_clustering/code/old_tests/tests2.py

@@ -1,124 +1,0 @@
-"""
-Several tests and comparison for clustering.
-
-"""
-
-import numpy as np
-import matplotlib.pyplot as plt
-from sklearn.cluster import KMeans
-from sklearn.cluster import SpectralClustering
-from sklearn.mixture import GMM
-
-from eclust.kkmeans import KernelKMeans
-from eclust.metric import accuracy
-import eclust.data as data
-from eclust.withingraph import EClust
-from eclust.energy1d import two_clusters1D
-from eclust.energy import energy_kernel
-from eclust.objectives import kernel_score
-
-
-def kernel_energy(k, X, alpha=.5, cutoff=0, num_times=8):
-    best_score = 0
-    for i in range(num_times):
-        km = KernelKMeans(n_clusters=2, max_iter=300, kernel=energy_kernel,
-                      kernel_params={'alpha':alpha, 'cutoff':cutoff})
-        zh = km.fit_predict(X)
-        score = kernel_score(km.kernel_matrix_, zh)
-        if score > best_score:
-            best_score = score
-            best_z = zh
-    return best_z
-
-def spectral_energy(k, X, alpha=.5, cutoff=0, n_neighbors=12, num_times=8):
-    best_score = 0
-    for i in range(num_times):
-        sc = SpectralClustering(n_clusters=k, affinity=energy_kernel,
-                        assign_labels='kmeans', n_init=20, 
-                        kernel_params={'alpha':1, 'cutoff':0},
-                        n_neighbors=n_neighbors)
-        sc.fit(X)
-        zh = sc.labels_
-        score = kernel_score(sc.affinity_matrix_, zh)
-        if score > best_score:
-            best_z = zh
-            best_score = score
-    return best_z
-
-def kmeans(k, X):
-    km = KMeans(n_clusters=k)
-    zh = km.fit_predict(X)
-    return zh
-
-def gmm(k, X):
-    gmm = GMM(n_components=k)
-    gmm.fit(X)
-    zh = gmm.predict(X)
-    return zh
-
-def cluster_many(k, X, z, clustering_function, num_times):
-    accuracies = []
-    for i in range(num_times):
-        zh = clustering_function(k, X)
-        a = accuracy(z, zh)
-        accuracies.append(a)
-    return np.array(accuracies)
-
-def compare_algos(k, X, z, num_times=10):
[…]

### 2017-09-30 qcqp_clustering/code/old_tests/tests_nd.py

@@ -1,115 +1,0 @@
-"""
-Several tests and comparison for clustering.
-
-"""
-
-import numpy as np
-import matplotlib.pyplot as plt
-from sklearn.cluster import KMeans
-from sklearn.cluster import SpectralClustering
-from sklearn.mixture import GMM
-
-from eclust.kkmeans import KernelKMeans
-from eclust.metric import accuracy
-import eclust.data as data
-from eclust.withingraph import EClust
-from eclust.energy1d import two_clusters1D
-from eclust.energy import energy_kernel
-from eclust.objectives import kernel_score
-
-
-def kernel_energy(k, X, alpha=1, cutoff=0, num_times=10, truth=None):
-    best_score = 0
-    for i in range(num_times):
-        if truth is not None:
-            km = KernelKMeans(n_clusters=2, max_iter=300, kernel=energy_kernel,
-                      kernel_params={'alpha':alpha, 'cutoff':cutoff},
-                      labels=truth)
-        else:
-            km = KernelKMeans(n_clusters=2, max_iter=300, kernel=energy_kernel,
-                      kernel_params={'alpha':alpha, 'cutoff':cutoff})
-        zh = km.fit_predict(X)
-        score = kernel_score(km.kernel_matrix_, zh)
-        if score > best_score:
-            best_score = score
-            best_z = zh
-    return best_z
-
-def spectral_energy(k, X, alpha=1, cutoff=0, n_neighbors=10, num_times=10):
-    best_score = 0
-    for i in range(num_times):
-        sc = SpectralClustering(n_clusters=k, affinity=energy_kernel,
-                        assign_labels='kmeans', n_init=20, 
-                        kernel_params={'alpha':1, 'cutoff':0},
-                        n_neighbors=n_neighbors)
-        sc.fit(X)
-        zh = sc.labels_
-        score = kernel_score(sc.affinity_matrix_, zh)
-        if score > best_score:
-            best_z = zh
-            best_score = score
-    return best_z
-
-
-###############################################################################
-if __name__ == '__main__':
-    
-    m1 = np.array([0, 0, 0, 0])
-    s1 = np.array([
[…]

### 2017-09-30 qcqp_clustering/code/tests2.py

@@ -1,882 +1,0 @@
-"""
-Tests and comparison between Energy, K-means, and GMM.
-
-We run Energy, K-means, and GMM a few times, and show average accuracy 
-and variability for several settings.
-
-"""
-
-# Guilhere Franca <guifranca@gmail.com>
-# Johns Hopkins University, Neurodata
-
-from __future__ import division
-
-import numpy as np
-import scipy.stats
-import matplotlib as mpl
-import matplotlib.pyplot as plt
-import cPickle, gzip
-from sklearn.decomposition import PCA
-
-import eclust.data as data
-import eclust.kmeans as km
-import eclust.gmm as gm
-import eclust.qcqp as ke
-import eclust.kmeanspp as kpp
-import eclust.energy1d as e1d
-
-from eclust import metric 
-from eclust import objectives
-
-from pplotcust import *
-
-import sys
-
-
-###############################################################################
-# clustering functions
-
-def kernel_energy(k, X, kernel_matrix, z, run_times=5):
-    """Run few times and pick the best objective function value."""
-    G = kernel_matrix
-    best_score = -np.inf
-    for rt in range(run_times):
-        
-        z0 = kpp.kpp(k, X, ret='labels')
-        Z0 = ke.ztoZ(z0)
-        
-        zh = ke.kernel_kmeans(k, G, Z0, max_iter=300)
-        Zh = ke.ztoZ(zh)
-        score = ke.objective(Zh, G)
-        
-        if score > best_score:
-            best_score = score
-            best_z = zh
-
-    return metric.accuracy(z, best_z)
-
-def cost_energy(k, X, kernel_matrix, z, run_times=5):
-    """Run few times and pick the best objective function value."""
-    G = kernel_matrix
-    best_score = -np.inf
-    for rt in range(run_times):
-        
-        z0 = kpp.kpp(k, X, ret='labels')
-        Z0 = ke.ztoZ(z0)
-        
-        zh = ke.minw(k, G, Z0, max_iter=300)
-        Zh = ke.ztoZ(zh)
-        score = ke.objective(Zh, G)
-        if score > best_score:
-            best_score = score
-            best_z = zh
-    
-    return metric.accuracy(z, best_z)
-
-def kmeans(k, X, z, run_times=5):
-    """Run k-means couple times and pick the best answer."""
-    best_score = np.inf
-    for rt in range(run_times):
-        
[…]

### 2017-10-03 energy_clustering/draft/energy_clustering.tex
Finished 1st revision

@@
 \newcommand\om{\omega}
+\newcolumntype{g}{>{\columncolor{gray!20}}l}
 
@@
 Newtonian gravitational potential from classical mechanics, and it provides 
-a hypothesis test for equality of distributions. Energy statistics was 
-further generalized to data living metric spaces of strong negative type, 
-and more recently a connection with reproducing kernel Hilbert spaces (RKHS) 
-was proposed. Here we consider the problem of clustering data from an 
-energy statistics 
-theory perspective, called energy clustering for short, by providing a 
-precise mathematical formulation yielding a quadratically constrained 
-quadratic program (QCQP) in the associated RKHS, 
-which we show to be equivalent to kernel $k$-means optimization problem.
+a non-parametric hypothesis test for equality of distributions. 
+It was further generalized from Euclidean spaces to metric spaces of 
+strong negative type, and more recently, a connection with reproducing 
+kernel Hilbert spaces (RKHS) was established. 
+Here we consider the problem of clustering data from an 
+energy statistics theory perspective.
+We provide a precise mathematical formulation 
+yielding a quadratically constrained 
+quadratic program (QCQP) in the associated RKHS. We show that this QCQP
+is equivalent to kernel $k$-means optimization problem, once the kernel
+is fixed.
 Thus, our results imply a first principles derivation of kernel $k$-means 
 from energy statistics.
-Moreover, we also consider a weighted version of energy statistics 
-applied to clustering, making connection to graph partitioning problems.
+However, energy statistics fixes a family of standard kernels.
+Furthermore, we also consider a weighted version of energy statistics, 
+making connection to graph partitioning problems.
 To find local optimizers of such QCQP we consider an iterative algorithm based 
@@
 of the proposed method compared to kernel $k$-means, standard $k$-means, 
-and gaussian mixture models in a variety of settings.
+and Gaussian mixture models in a variety of settings.
 \end{abstract}
@@
 
+% @jovo: I corrected pretty much all your previous points, and also re-wrote 
+% a lot of other stuff throughout the paper. 
+% I deleted your previous comments which I believe were taken
+% care of, since most of them were simple. The ones which were not, I will
+% leave it, in the beggining of each section or in the middle.
+%
+% @jovo: About ToC, I agree and wouldn't mind remove it, however it is a 
+% tradition that most referees like to see. In this case I also included 
+% some important information about our work, and it helps to make a transition 
+% between introduction and the remaining of the paper. 
+% I would leave it, but if you prefer to remove it, let me know and I 
+% will adapt the paper.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Introduction}
-
-% @gui: algorithms don't make assumptions, theorems do.  also, algorithms cannot be parametric or non-parametric, that is a property only of the statistical model, or an estimator.
-
-% @gui: k-means is NOT a special case of GMM, GMM is a soft clustering (not fuzzy, in stats and machine learning we don't say "fuzzy"), where as k-means is a hard clustering.
-k-means is certainly closely related to GMM though.
-
-% @gui:  don't know what "clear resolution of cluster centers" adds to this?
-
-% @gui: algorithms cluster, doesn't matter how complicated the data are.  so, what do you actually mean here?  perform well empirically? what does "complicated" actually mean?
-
-% @gui: the words "this" and "it" are often vague and ambiguous, be specific/precise
-
-% @gui: not sure what you mean here, in kernel k-means, we don't "optimize" over any parameters typically.  sometimes we select hyperparameters of kernels via cross-validation, is that what you mean?
-
-% @gui: not sure what you mean "only one possible optimization problem".  we can reformulate it many different ways? also, below, "this" is ambiguous.
-
-% @gui: to me, we did not make explicit one of the most important points.
-in this paper, we make clear exactly when we expect energy clustering to perform better than k-means/GMM, and when we expect it to perform worse (empirically).
-
-% @gui: i find ToC paragraphs useless.
-
-% @jovo: corrected all of these points, and also re-wrote some other stuff.
-Please check.
-About ToC, I agree and wouldn't mind remove it, however it is a tradition that most referees like to see.
-In this case I included some important information about our work too, and it makes a transition between introduction and the remaining of the paper.
-If you want to remove it let me know and I will adapt.
 
@@
 in close analogy to Newton's gravitational potential in classical mechanics. 
-It provides a hypothesis test for equality of distributions which is achieved 
+It provides a non-parametric hypothesis test for equality of 
+distributions which is achieved 
 under minimum energy. When probability distributions are different the 
 statistical potential energy diverges as sample size increases, while tends 
-to a nondegenerate limit distribution, i.e. remains finite, when they are 
-equal. Energy statistics has been applied to several goodness-of-fit 
+to a nondegenerate limit distribution when probability
+distributions are equal. 
+Energy statistics has been applied to several goodness-of-fit 
 hypothesis tests, multi-sample tests of equality of distributions, 
@@
 The notion of distance covariance was further generalized from Euclidean 
-spaces to metric spaces of strong negative type \cite{Lyons}. Moreover, 
-the missing link between energy distance based tests and kernel based tests has 
+spaces to metric spaces of strong negative type \cite{Lyons}. Furthermore, 
+the missing link between energy distance based tests and kernel 
+based tests has 
 been recently resolved \cite{Sejdinovic2013}, where a unifying framework
@@
 distributions in reproducing kernel Hilbert spaces (RKHS), was established. 
-This equivalence provides the link between energy distance based measures and 
-RKHS based measures, commonly used in machine learning, and form the basis 
+This equivalence immediately relates energy statistics to
+kernel methods, often used in machine learning, and form the basis 
 of our approach.
 
-The clustering problem has such a long history in machine learning, making it
+Clustering has such a long history in machine learning, making it
 impossible to mention all important contributions in a short space. 
@@
 the cluster with closest center. The only statistical 
-information about each cluster comes from its mean, and it is thus sensitive 
+information about each cluster comes from its mean, making it sensitive 
 to outliers. Nevertheless, $k$-means works very well when data is 
-linearly separable in Euclidean space. Gaussian mixture models (GMM) is also 
-very commonly used for clustering, however, it makes strong assumptions about 
-the distribution of the data, as $k$-means which is closely related to GMM.
+linearly separable in Euclidean space. Gaussian mixture models (GMM) is 
+another very common approach, providing more flexibility than $k$-means, 
+however, it still makes strong assumptions about the distribution of 
+the data.
 
@@
 map data points to a RKHS, then clustering can be performed in the associated
-Hilbert space by using its inner product. Yet, the kernel choice remains 
+Hilbert space by using its inner product. However, the kernel choice remains 
 the biggest challenge since there is no principled theory to construct a kernel
@@
 in the feature space \cite{Girolami}. Furthermore, kernel $k$-means algorithm
-\cite{Dhillon2,Dhillon} is still based on Loyd's heuristic \cite{Lloyd} of
-using the mean of each cluster in the feature space. 
+\cite{Dhillon2,Dhillon} is still based on Loyd's heuristic \cite{Lloyd}
+of grouping points that are closer to a cluster center, now
+in feature space. 
 We refer the reader to \cite{Filippone} for a survey of clustering
@@
 in \cite{Kgroups}, the precise optimization problem behind this approach
-remains elusive, as well as the connection to kernel methods.
+remains elusive, as well as the connection with kernel methods.
 The main theoretical contribution of this paper is to fill this gap.
@@
 constrained quadratic program (QCQP) in the associated RKHS. Our results
-immediately establish the connection to kernel methods, by showing that this 
+immediately establish the connection with kernel methods, showing that this 
 QCQP is equivalent to kernel $k$-means optimization problem. The 
@@
 partitioning problems is well-known \cite{Dhillon,Dhillon2}. We demonstrate 
-how these connections arise from a weighted version of energy statistics.
-
-Our algorithmic contribution is to use Hartigan's method \cite{Hartigan} to 
+how these relations arise from a weighted version of energy statistics.
+
+Our main 
+algorithmic contribution is to use Hartigan's method \cite{Hartigan} to 
 find local solutions of the above mentioned QCQP, which is NP-hard in general.
-Hartigan's method was also used in \cite{Kgroups}, and its advantages
-over Lloyd's method was already being demonstrated \cite{Telgarsky,Slonin},
-but apparently it did not receive the desired attention. To the 
-best of our knowledge, Hartigan's method was not previously applied together 
-with kernel methods. We provide an algorithm based on Hartigan's method in 
-terms of the Gram matrix of the data computed from a kernel fixed by 
-energy statistics. We make clear the advantages of this proposal versus 
+Hartigan's method was also used in \cite{Kgroups}. More importantly, its 
+advantages
+over Lloyd's method was already demonstrated 
+in some simple settings
+\cite{Telgarsky,Slonin}, but apparently this method did not receive 
+the deserved attention. To the 
+best of our knowledge, Hartigan's method was not previously 
+employed together with kernel methods. 
+We provide a full kernel based Hartigan's algorithm for clustering,
+where the kernel is fixed by energy statistics. 
+We make clear the advantages of this proposal versus 
 Lloyd's method, which kernel $k$-means is based upon and will also be used 
-to locally solve our QCQP. We show that both algorithms  have the same
-time complexity.
+to solve our QCQP. We show that both algorithms  have the same
+time complexity, but Hartigan's method in kernel spaces offer several
+advantages. Furthermore, in the examples considered in this paper, it 
+also provides superior performance compared to a spectral clustering
+relaxation of the QCQP.
 
@@
 or \emph{energy clustering} for short, is more accurate and robust than 
-Lloyd's method. Moreover, in our experiments we put in evidence the
+Lloyd's method. Our experiments also put in evidence the
 flexibility of energy clustering, which provides a family of default kernels, 
@@
 very different distributions, contrary to $k$-means and GMM for instance.
-More specifically, the proposed algorithms for energy clustering perform 
-closely to $k$-means and GMM on normally distributed data. On the other
-hand, it performs considerably better than $k$-means and GMM on  data that 
+More specifically, our algorithms for energy clustering perform 
+closely to $k$-means and GMM on normally distributed data, but on the other
+hand, performs considerably better than $k$-means and GMM on  data that 
 is not normally distributed. It also performs better than $k$-means 
@@
 version of energy statistics, providing connections to graph
-partitioning problems and spectral clustering.
+partitioning problems.
 In Section~\ref{sec:twoclass}, we consider the  simple case of a two-class
@@
 Section~\ref{sec:numerics} contains some carefully designed numerical
-experiments showing that in several settings, energy clustering based
+experiments showing that, in a variety of settings, energy clustering based
 on Hartigan's algorithm
-outperms kernel $k$-means algorithm, standard $k$-means algorithm 
-and GMM with the expectation maximization algorithm. 
+outperform $k$-means, GMM, kernel $k$-means, and a version of
+spectral clustering.
 Our final conclusions are presented in Section~\ref{sec:conclusion}.
@@
 
-% @gui: i don't know what this means.  eqn 7 defines something for a population statistic, not a sample statistic.
-% @jovo: removed. I was thinking that you can always estimate the expected 
-% value by computing a sample mean. but I agree, it's confusing
-
 % @gui: i'm not sure the above is true/meaningful.  nonparametric means implicitly estimating an infinite number of parameters.  without any statistical theory, i'm not quite sure what we can say.
-% @jovo: I stoped saying non-parametric, however, Kolmogorov-Smirnov is also said to be non-parametric, so I thought that in this context it means that it does not assume any form of the distributions.
+% @jovo: I stoped saying non-parametric, however, Kolmogorov-Smirnov is also
+% said to be non-parametric. I think non-parametric tests also mean that
+% it does not assume any form of distributions, i.e. it is a distribution-free
+% statistical test.
 
@@
 allows us to compute quantities related to probability distributions over
-$\mathcal{X}$ in the Hilbert space $\mathcal{H}$.
+$\mathcal{X}$ in the associated Hilbert space $\mathcal{H}$.
 Even though the semimetric 
@@
 $\int f(x) d P(x) = \langle f, \kkk_P \rangle$ for all $f \in \Hk$,
-or alternatively $\kkk_P \equiv \int \kk( \, \cdot \,, x)  d P(x)$. 
+or alternatively, $\kkk_P \equiv \int \kk( \, \cdot \,, x)  d P(x)$. 
 We can now  introduce the 
@@
 If two different kernels generate the same $\rho$ they are
-equivalent kernels.
+said to be equivalent kernels.
 
@@
 \end{equation}
-Due to \eqref{eq:mmd} we can compute the energy distance using the inner 
+Due to \eqref{eq:mmd} we can compute the energy 
+distance $\mathcal{E}(P, Q)$ between two probability distributions
+using the inner 
 product of $\Hk$.
@@
 \sum_{x \in \C_i} 
-\sum_{y \in \C_j} \rho(x, y)
+\sum_{y \in \C_j} \rho(x, y) ,
 \end{equation}
@@
 we have that $S \to \infty$.
-This test does not make any assumptions
-about the distributions $P_j$.
+Note that this test does not make any assumptions
+about the distributions $P_j$, thus it is said to be non-parametric or
+distribution-free.
 
@@
 % @gui: consider only using equation numbers for equations that you later refer to? there are so many, it is easy to get lost.
-% @jovo: ok, will do that after we finish everything. There is a package that automatically remove numbers from unreferenced equations
+% @jovo: ok, will do that after we finish everything. There is a package that
+% automatically remove numbers from unreferenced equations. It is a common
+% practice however to number every equation in the paper,
+% specially for reviewing purposes. But later 
+% I will remove the numbers from the not cited ones if you don't like
 
 % @gui: where possible, i recommend using the "name" of the quantity rather than the number of the equation.  eg, here, write S and W, rather than (13) and (12).
-% @jovo: ok, I will do that for the main symbols, and refer to equations only when necessary
+% @jovo: I will do that for the main symbols, and refer to equations only
+% when necessary. But sometimes it is unnavoidable, specially
+% when referring to long 
+% expressions or derivation steps. It is way more clear and precise to say, 
+% replacing equation (1) into (2) we obtain ... than saying, 
+% replacing A into B we obtain ... and the reader has to look around
+% the whole paper to figure it out what  A, B are
 
@@
 % that this holds only for 0 < alpha < 2, with |x-y|^alpha, while
-% here I did for a general semimetric. The PhD thesis do not 
+% here I did for a general semimetric. Thus I don't really know if in general
+% this is equality is known or not, but it should because it's a very easy
+% derivation. The PhD thesis do not 
 % mention anywhere that minimizing W is equivalent to maximizing S.
-% Thus, although a trivial result, but important, I didn't see this 
-% written explicitly before.
-
-% @gui: "Let us" is used a bunch, though i think it never adds anything? i'd remove it everywhere.
-% @jovo: ok, remove it all along
+% Therefore, I don't want to sound like this is a big result and possibly
+% a referee can say  "well this is trivial and known".
+% Thus I just preferred to do my own derivation and not say anything.
+% What do you think??
 
@@
 % Since here A = Z^T G Z is a product of 3 matrices
-% we just put a parenthesis to avoid confusion, that's all
-
-% @gui: don't use equation numbers here.
-% @jovo: in some cases it is unavoidable. I tried to do that when the context
-% is obvious, by writing the symbols, but when we refer to long 
-% expressions or derivation steps it is way more clear so say, replace equation (1) into (2) to get ...
-than saying, replacing A into B to get ...
-and the reader has to glance through the entire paper to recall what A and B are.
+% it is common to just put a parenthesis to avoid confusion, that's all
 
 % @gui: this might be a better initialization than k-means++, which essentially initializes using a different kernel?
-% @jovo: yes it might be, and I agree, k-means++ is initializing in "Euclidean
-% space" not in the feature space where data might be better separable. We will do this from now on.
+% @jovo: This was a long issue this last week as you know. 
+% The spectral initialization is, often, very
+% close to a local optimum of E-clustering algorithm, and kernel k-means
+% as well. Now, since the set of local optima of Hartigan's is a subset of 
+% those of Lloyd's, if you initialize "too well", you might get stuck on 
+% a not so great local optimum of Lloyd's, which otherwise Hartigan's 
+% would be able to escape. Numerically, in most cases Hartigan's does
+% better when initializing with k-means++, or even at random.
 
 % @gui: i'm not sure about this.  who is to say the correct direction of causality? maybe energy clustering is a special case of kernel k-means?
-% @jovo: tried to not commit to a single view. We show that they are
+% @jovo: I tried to not commit to a single view. We show they are
 % equivalent, when opperating on the same kernel. However, kernel k-means per
@@
 % however, provides a theory valid on spaces of negative type, which guarantees
-% that a kernel constructed in this formalism will be positive-semidefinite.
-
-In this vein, kernel $k$-means is part of a
+% that a kernel constructed within this formalism will be 
+% positive-semidefinite. Thus, I still prefer to look in the first direction.
+
 This section contains the main theoretical results of this paper, where 
@@
 In other words, given a set of points coming from different probability
-distributions, $S$ should attain a maximum when each point is correctly
+distributions, the test statistic $S$ should attain a maximum when 
+each point is correctly
 classified as belonging to the cluster associated to its probability
@@
 energy statistics was already proposed in \cite{Kgroups}. However, it is
-important to note that this is equivalent to maximizing $S$
+important to note that this is equivalent to maximizing $S$,
 which is the test statistic for equality of distributions. In this current
-form, the relation with kernels and to other clustering methods is obscure.
+form, the relation with kernels and other clustering methods is obscure.
 In the following, we show what is the explicit optimization problem behind 
-in the corresponding RKHS, establishing the connection to kernel methods.
-
-We now formulate the analog of the optimization problem \eqref{eq:minimize} 
-in the corresponding RKHS. 
-Based on the relation between kernels and semimetrics, see 
-\eqref{eq:kernel_semimetric} and \eqref{eq:gen_kernel}, 
+\eqref{eq:minimize} in the corresponding RKHS, 
+establishing the connection with kernel methods.
+
+Based on the relation between kernels and semimetrics of negative
+type, 
 assume that the kernel $\kk: \mathcal{X} \times \mathcal{X} \to \mathbb{R}$ 
@@
 The resulting
-optimal objective function value is thus
+optimal objective function assumes the value 
 $\max \Tr \left( {Y^\star}^\top G \, Y^\star \right)  = 
@@
 i.e. $Y^\star \to \widetilde{Y}$ where 
-$\widetilde{Y}_{ij} \equiv Y^\star_{ij}/ \| Y^\star_{i\bullet} \|$. 
+$\widetilde{Y}_{ij} \equiv Y^\star_{ij} \big/ \| Y^\star_{i\bullet} \|$. 
 The vectors
@@
 set $\{ x_i \}_{i=1}^n$ onto an eigen-subspace of the kernel matrix $G$.
-One then group the points $\{ y_i \}_{i=1}^n$ into $k$ clusters 
-$\{ \tC_1,\dotsc,\tC_k \}$, using any clustering algorithm such as
-$k$-means, yielding the final solution
+After performing this embedding, one group the points 
+$\{ y_i \}_{i=1}^n$ into $k$ clusters 
+$\{ \tC_1,\dotsc,\tC_k \}$ using any clustering algorithm, such as
+$k$-means. This yields the final solution
 where the original point $x_i$ belongs
@@
 the matrix $D^{-1/2} G D^{-1/2}$ instead of $G$, and also 
-where $G$ is an affinity matrix with the 
-modification that its diagonal elements are set to zero $G_{ii}=0$, 
+where $G$ is an affinity matrix with the modification that
+its diagonal elements are set to zero $G_{ii}=0$, 
 and $D=\diag(d_{1},\dotsc,d_{n})$ is the degree matrix, 
-$d_{i} \equiv \sum_{j=1}^n G_{ij}$. 
-Here we consider a slightly different
-prescription which we call $\mathcal{E}$-spectral, using exactly the same 
-approach, but considering the
-eigen-decomposition of $G D^{-1}$ instead\footnote{
-We found empirically that our method is more stable and often more
-accurate than \cite{NgJordan}.}. Our method is described in 
-Algorithm~\ref{spectralalgo}, and will be used as an initialization to the
-iterative algorithms to solve the full optimization problem \eqref{eq:qcqp2}
+with $d_{i} \equiv \sum_{j=1}^n G_{ij}$. 
+We consider a slightly modification of this procedure,
+which we call \mbox{$\mathcal{E}^{S}$-clustering} 
+to emphasize that its a spectral
+method whose kernel comes from energy statistics, by 
+computing the 
+eigen-decomposition of $G D^{-1}$ instead.
+This  method is described in 
+Algorithm~\ref{spectralalgo} and will be used as a comparison
+to the other iterative algorithms 
 that will be proposed later.
@@
 has the top $k$ eigenvectors (multiplied by $D$) as columns
-\STATE normalize the rows $Y$, i.e. $\widetilde{Y}_{ij} \leftarrow Y_{ij} / 
+\STATE normalize the rows $Y$, i.e. 
+$\widetilde{Y}_{ij} \leftarrow Y_{ij} \Big/ 
 \left(\sum_{j=1}^n |Y_{ij}|^2\right)^{1/2}$
@@
 \label{spectralalgo}
-$\mathcal{E}$-spectral is an exact method to solve
+$\mathcal{E^S}$-clustering is an exact method to solve
 the relaxed problem \eqref{eq:relaxed}.
@@
 Note that the energy clustering problem \eqref{eq:qcqp2} 
-is valid for data living in an \emph{arbitrary} space of negative type where
+is valid for data living in an \emph{arbitrary} space of negative type, where
 a semimetric $\rho$, and thus the kernel $\kk$, are
-assumed to be known. For instance, standard energy statistics in
-Euclidean space fixes a family of choices, 
-$\rho(x,y) = \| x - y\|^\alpha$ and the corresponding
-kernel $K(x,y) = \tfrac{1}{2} \left( \| x \|^\alpha + \| y \|^\alpha - 
+assumed to be known. Standard energy statistics in
+Euclidean space fixes a family of choices 
+$\rho_\alpha(x,y) = \| x - y\|^\alpha$, with the corresponding
+kernel $K_\alpha(x,y) = \tfrac{1}{2} \left( \| x \|^\alpha + \| y \|^\alpha - 
 \| x-y\|^\alpha \right)$, where we
-set $x_0=0$ in \eqref{eq:kernel_semimetric}, for $0<\alpha\le 2$.
-For data living in a more general
-metric space $(\mathcal{X}, \rho)$ the corresponding semimetric $\rho$ fixes
-the kernel.
+set $x_0=0$ in \eqref{eq:kernel_semimetric} and  $0<\alpha\le 2$.
+The same would be valid for data living in a more general
+metric space $(\mathcal{X}, \rho)$, where the corresponding semimetric $\rho$ 
+fixes the kernel.
 In any case, the energy  clustering 
@@
 and is equivalent to choosing an appropriate kernel.
-However, if prior information is available to choose a suitable $\rho$,
-the energy clustering conviniently incorporate this measure in the 
+However, if prior information is available to choose an appropriate $\rho$,
+it can be conveniently incorporated in the 
 optimization problem \eqref{eq:qcqp2}.
@@
 that problem \eqref{eq:kernel_kmeans} 
-can be cast into a trace maximization problem in the same form as 
+can be cast into a trace maximization in the same form as 
 \eqref{eq:qcqp2}. The next result makes this explicit, showing that
@@
 energy statistics theory is valid for arbitrary semimetric spaces of
-negative type, and the corresponding kernel in the associated RKHS is
-guaranteed to be positive semi-definite.
+negative type, fixing the kernel function in the associated RKHS, which
+is guaranteed to be positive definite.
 
@@
 \eqref{eq:between}, with 
-$n_i \to s_i$ and $n \to s$ where $s = \sum_{j=1}^k s_j$,
-to obtain a weighted version of energy test statistic. With these
+$n_i \to s_i$ and $n \to s$, where $s = \sum_{j=1}^k s_j$,
+With these
 changes, Proposition~\ref{th:minimize} remains the unaltered, so the
@@
 where $w_i = w(x_i)$ and $\e \in \mathbb{R}^n$ is the all-ones
-vector. Now we can show the analogous of
+vector. We can now show the analogous of
 Proposition~\ref{th:qcqp2} to the case of problem \eqref{eq:minimize2}.
@@
 \label{th:qcqp3}
-The weighted version of energy statistics clustering given by
+The weighted energy clustering given by
 problem \eqref{eq:minimize2} is equivalent to
@@
 \end{equation}
-where $G$ is the Gram matrix \eqref{eq:kernel_matrix} and the other quantities
-are defined in 
-\eqref{eq:weighted_matrices}.
+where $G$ is the Gram matrix \eqref{eq:kernel_matrix}, 
+$\om = (w_1,\dotsc,w_n)^T$ contains the weights of each point, and
+$\mathcal{W} = \diag(\omega)$.
 \end{proposition}
@@
 % @jovo: it does not require initialization in the sense that we do not
-% assign points to clusters in an initial step, as k-means++ for instance, or
+% assign points to clusters as an initial step, as k-means++ for instance, or
 % random initialization. This algorithm is "exact" since nothing
@@
 i'm not sure the best way to visualize it, maybe play with a few options.
+% @jovo: I didn't do this yet and actually not sure how to do it.
+% Are you thinking about this: 
+% https://en.wikipedia.org/wiki/Kernel_density_estimation
+% Also, I'm not sure if the kernel from energy integrates to 1, etc. and is
+% this really worthy it?
+
+% @gui: why not use the same parameters as in fig 1a?  i feel like if we did, it would seem less like we are cherry picking?  
+% @jovo: yes, and GMM would do great! In 1D is very hard to find settings where
+% GMM and k-means perform poorly, so we have to cherry pick a little. In 1D you
+% basically do not have much choice, if data has minimal separation k-means and
+% GMM do well. The purpose is to
+% illustrate that there are settings where k-means/GMM do poorly and energy
+% does not. If I choose the same parameters as before all algorithms are doing
+% great, sometimes GMM can even do better than energy. All I did was to
+% translate one of the lognormals mu2=-1.5 and give it a big variance, so both
+% distributions overlap a lot, but the overlaping is essentially on the tail 
+% of one of them
 
@@
 \begin{minipage}{0.49\textwidth}
-\includegraphics[width=.8\textwidth]{hist_normal.pdf}\\[-.8em]
-\includegraphics[width=\textwidth]{gauss1d.pdf}\\[-1em]
-(a)
+\includegraphics[width=\textwidth]{1D_normal.pdf}\\[-1em](a)
 \end{minipage}
 \begin{minipage}{0.49\textwidth}
-\includegraphics[width=.8\textwidth]{hist_lognormal.pdf}\\[-.8em]
-\includegraphics[width=\textwidth]{loggauss1d.pdf}\\[-1em]
-(b)
+\includegraphics[width=\textwidth]{1D_lognormal.pdf}\\[-.8em](b)
 \end{minipage}
@@
 \label{fig:1d}
-Energy statistics clustering by Algorithm~\ref{algo1d}
-compared to $k$-means and GMM/EM.
-We have the same number of points in both clusters, and for each case
-we sample $100$ times from the distributions shown in the histograms. We 
-plot the average value of 
-\eqref{eq:accuracy} versus the total number of points 
-(error bars are standard error). The dashed line indicates the
-best possible classification accuracy computed from Bayes error.
-(a) Data coming from \eqref{eq:two_normal}, where the
+$\mathcal{E}^{1D}$-clustering versus $k$-means and GMM in one dimension.
+The two clusters have the same number of points, and we use
+$50$ Monte Carlo runs.
+We plot the average 
+accuracy \eqref{eq:accuracy} versus the total number of points, 
+where error bars are standard error. The dashed line indicates 
+Bayes accuracy. (a) Data distributed as \eqref{eq:two_normal}, where the
 optimal accuracy is $\approx 0.956$.
-(b) Data from \eqref{eq:two_lognormal}, where the optimal
+(b) Data from \eqref{eq:two_lognormal} where the optimal
 accuracy is $\approx 0.852$.
@@
 
-We now consider two simple experiments with equal number of points
-in each cluster. We plot the accuracy \eqref{eq:accuracy} versus the number
-of points in each cluster.
-The data is clustered using 
-the $\mathcal{E}^{1D}$-clustering algorithm,
-% @gui: i see why you call this algorithm 1, but can we call it something with semantic meaning? like 1D-\mathcal{E}-clustering? or maybe \mathcal{E}1 for short? same for the other variants.
-% @jovo: will call energy clustering, or $\mathcal{E}-clustering$ is name is to
-% big when I regenerate the plots with new figures
-GMM through EM algorithm, and $k$-means++ 
-algorithm. 
-% @gui: there are many k-means algorithms, be specific.
-% @jovo: there is only one k-means no? Anything else I would call kernel
-% k-means. I called $k$-means++ though
-We use the initialization procedure from 
-$k$-means++ \cite{Vassilvitskii} also for GMM.
-We run the algorithms $10$
-times choosing the result with  best objective function value. 
-% @gui: a few? be specific
-% @jovo: ok
+We now consider two simple experiments where we sample the same 
+number of points for each cluster. 
+We plot the average accuracy \eqref{eq:accuracy} versus the 
+total number of points, with error bars indicating standard error.
+The data is clustered using $\mathcal{E}^{1D}$-clustering algorithm,
+GMM through EM algorithm, and the standard $k$-means 
+algorithm. For $k$-means and GMM we use the implementations
+from the well-known \emph{scikit-learn} library in 
+Python \cite{scikit-learn}, where $k$-means is initialized through
+$k$-means++ procedure \cite{Vassilvitskii}, and GMM is initialized
+with the output of $k$-means. We choose to run both algorithms
+$5$ times with different initializations and the answer with best 
+objective function value is returned.
 Notice that $\mathcal{E}^{1D}$-clustering does not require any random
-initialization so we only
-% @gui: "no initiazation" is weird.  do any algs require an initialization?
-% @jovo: I guess so. In this algorithm we just compute something "exactly",
-% contrary to random initializing and iterating to optimize the solution.
-% There is no random initialization in this algo
-run it once.
-Moreover, for each case we sample $100$ times
-and show the average accuracy with error bars indicating
-the standard error.
+initialization so we only run it once.
+For each fixed number of points we use use $50$ Monte Carlo runs.
 In  Fig.~\ref{fig:1d}a 
-we have data sampled from two normal distributions with equal number of
-points in each cluster,
+we have the results for data sampled from two normal distributions, 
 \begin{equation}
@@
 \end{equation}
-For these distributions the optimal accuracy 
+In this case the optimal accuracy 
 obtained from Bayes classification error
-is $ \approx 0.956$ which is indicated by the dashed line in the plot.
-We see that the three methods
-perform closely. As expected, GMM has a slight advantage over the other
-methods since
-it corresponds to the true model of the data. 
-Energy statistics performs slightly better
+is $ \approx 0.956$, indicated by the dashed line in the plot.
+The three methods
+perform closely, with a slight advantage of GMM, as expected since
+it corresponds to the true model of the data, and
+$\mathcal{E}^{1D}$-clustering performs slightly better
 than $k$-means. On the other hand, in Fig.~\ref{fig:1d}b
@@
 \end{equation}
-% @gui: why not use the same parameters as in fig 1a?  i feel like if we did, it would seem less like we are cherry picking?  
-% @jovo: yes, and GMM would do great! In 1D is very hard to find setting where
-% GMM and k-means perform poorly, so we have to cherry pick. In 1D you
-% basically do not have much choice, if data has minimal separation k-means and
-% GMM does well.
-% The purpose is to
-% illustrate that there are settings where k-means/GMM do poorly and energy
-% does not. If I choose the same parameters as before all algorithms are doing
-% great, sometimes GMM can even do better than energy.
 The optimal classification accuracy from Bayes error is $\approx 0.852$.
-% @gui: computed numerically?
-% @jovo: yes!
-In this case,
-Algorithm~\ref{algo1d} provides a very accurate clustering while
+We can now see that $\mathcal{E}^{1D}$-clustering
+is still  very accurate, while
 GMM and $k$-means basically cluster at chance.
-Sometimes GMM/EM  was unable to estimate the parameters 
-thus giving zero accuracy.
 The two simple experiments of Fig.~\ref{fig:1d} illustrate
-how energy statistics clustering is nonparametric, being able
-to provide high quality clustering in settings where data comes
-from very different distributions.
+how energy statistics does not make assumptions about the distributions
+of the data, providing more flexibility and being able
+to accurately cluster data with very different distributions.
 
@@
 In this section we introduce an iterative algorithm to find a local
-maximizer of \eqref{eq:qcqp2}. Due to 
+maximizer of the optimization problem \eqref{eq:qcqp2}. Due to 
 Proposition~\ref{th:kernel_kmeans} we can also find an approximate
@@
 \bigg\{ Q = \sum_{j=1}^k \dfrac{Q_j}{n_j}  \bigg\},
-\qquad Q_j = \sum_{x,y\in\C_j} \kk(x,y),
+\qquad Q_j \equiv \sum_{x,y\in\C_j} \kk(x,y),
 \end{equation}
@@
 is weighted by the inverse
-of the number of its elements. For a data point $x_i$ we denote
+of the number of points in $\C_j$. For a data point $x_i$ we denote
 its own energy cost
@@
 
-%\subsection*{Kernel $\bm{k}$-Means Algorithm}
 \subsection*{Lloyd's Method for Energy Clustering}
@@
 The entire procedure is described in Algorithm~\ref{kmeans_algo}, which
-we call $\mathcal{E}^L$-clustering to emphasize
-that we are optimizing the energy function $W$ from energy statistics
+we name $\mathcal{E}^L$-clustering to emphasize
+that we are optimizing the withn energy function $W$ 
 based on Lloyd's method \cite{Lloyd}.
-% @gui: let's call this something more informative
-% @jovo: alright
 It can be shown that this algorithm converges provided $G$ is positive
@@
 % the standard formulation of kernel k-means.
-Actually, $\mathcal{E}^{L}$-clustering is preciselly
+Obviously, $\mathcal{E}^{L}$-clustering is precisely
 kernel $k$-means algorithm \cite{Dhillon2,Dhillon} 
@@
 where $\mu_\ell = \tfrac{1}{n_\ell} \sum_{x\in \C_\ell}x$ 
-is the mean (center) of cluster $\C_\ell$. 
+is the mean of cluster $\C_\ell$. 
 Therefore, $\min_\ell J^{(\ell)}(x_i) = 
@@
 which is the familiar Lloyd's
-heuristic approach.
+heuristic approach that kernel $k$-means is based upon.
 
@@
 \caption{\label{kmeans_algo}
-$\mathcal{E}^{L}$-clustering: Lloyd's method for energy clustering, which
-is preciselly kernel $k$-means algorithm. This procedure finds
-local solutions to the problem \eqref{eq:qcqp2}.
+$\mathcal{E}^{L}$-clustering is Lloyd's method for energy clustering, which
+is precisely~~~~~~~~ \\ 
+kernel $k$-means algorithm. This procedure finds
+local solutions to the problem \eqref{eq:qcqp2}.\hspace{\fill}
 }
@@
 \subsection*{Hartigan's Method for Energy Clustering}
+
+% @jovo: I also clarified this section, specially the derivation
+% of the algorithm. Actually there was a mistake in the previous formula,
+% which carried over to my code. Now it's fixed and better.
 
@@
 $\Delta Q^{j\to \ell} (x_i) \equiv Q^{(\ell)} - Q^{(j)}$, for $\ell\ne j$. 
-From \eqref{eq:maxQ}, by explicitly writting the costs related to these 
+From \eqref{eq:maxQ}, by explicitly writing the costs related to these 
 two cluster we obtain
@@
 maximum of \eqref{eq:maxQ} by
-moving $x_i$ to $\C_\ell$, otherwise we keep $x_i$ in $\C_j$. Based on
-this the proposed algorithm goes as follows.
+moving $x_i$ to $\C_\ell$, otherwise we keep $x_i$ in $\C_j$.
+
+We thus propose an algorithm that goes as follows.
 We start with an initial configuration for the label matrix $Z$, 
@@
 $\Delta Q^{j\to \ell}(x_i)$ for 
-$\ell=1,\dots,k$ with $\ell \ne j$ and $j$ denotes the index of its current
+$\ell=1,\dots,k$ with $\ell \ne j$, where $j$ denotes the index of its current
 partition, $x \in \C_j$. Hence, we choose
@@
             %for $\ell=1,2,\dots,k$ and $\ell \ne j$; 
-            using Eq. \eqref{eq:changeQ} \label{stepmove}
+            using \eqref{eq:changeQ} \label{stepmove}
         \IF{ $\Delta Q^{j \to j^\star}(x_i) > 0$ }
@@
 $\mathcal{E}^H$-clustering: Hartigan's method for energy clustering.
-This algorithm finds local solutions to  
-the optimization problem \eqref{eq:qcqp2}.
+This algorithm~~~~~~~~ \\ finds local solutions to  
+the optimization problem \eqref{eq:qcqp2}.\hspace{\fill}
 \hspace{\fill}
@@
 The cost of computing 
-\eqref{eq:costxij} is $\OO(n_j)$, thus the cost of step~$8$ in
-Algorithm~\ref{algo} is $\OO(k \max_j n_j)$ for $j=1,\dotsc,k$.
+%\eqref{eq:costxij} 
+$Q_\ell(x_i)$
+is $\OO(n_\ell)$, thus the cost of step~$6$ in
+Algorithm~\ref{algo} is $\OO(k \max_\ell n_\ell)$ for $\ell=1,\dotsc,k$.
 For the 
 entire dataset this gives a time complexity
-of $\OO(n k  \max_j n_j) =\OO(k n^2)$. Note that this is the same cost as
+of $\OO(n k  \max_\ell n_\ell) =\OO(k n^2)$. Note that this is the same cost as
 in $\mathcal{E}^L$-clustering, or kernel $k$-means algorithm. 
@@
 results about Hartigan's method.
-% @gui: "let us" again...
-% @jovo: ok
 
@@
 
-%Both conditions of the  above theorem are not satisfied 
-%by Lloyd's method, 
 Neither of these two conditions are guaranteed to be 
 satisfied  by Lloyd's method,
-% @gui: do you mean Lloyd's method satisfies neither, or does not satisfy both?  please clarify
-% @jovo: ok
 and consequently by $\mathcal{E}^L$-clustering 
@@
 Moreover, this improvement should be more prominent as
-dimension increases. Also, the improvement grows as $k$
-increases.
+dimension increases. Also, the improvement grows as the number of clusters
+$k$ increases.
 The empirical results of \cite{Telgarsky} show that 
 an implementation of Hartigan's method has comparable execution time 
-as an implementation of
+to an implementation of
 Lloyd's method,
@@
 $\mathcal{E}^L$- and $\mathcal{E}^H$-clustering
-have the same time complexity.
+have the same time complexity. To the best of our knowledge, Hartigan's
+method was not previously considered together with kernels, as we 
+are proposing
+in $\mathcal{E}^{H}$-clustering.
 
@@
 with any Bregman divergence. It was shown that the number of Hartigan's
-local minima is upper bounded by $\mathcal{O}(1/k)$ 
+local optima is upper bounded by $\mathcal{O}(1/k)$ 
 \cite[Proposition 5.1]{Slonin}. 
@@
 $\mathcal{E}^H$-clustering  outperforms
-of $\mathcal{E}^L$-clustering (kernel $k$-means) in several settings.
+of $\mathcal{E}^L$-clustering (kernel $k$-means) in several settings,
+specially in high dimensions.
 
@@
 
+% @gui: something is weird with the way you care computing accuracy.  GMM should not be significantly worse than chance.  i think you are probably counting the NaN's as *wrong*, rather than nothing?
+% now I'm using GMM from sklearn which eliminated previous weird things.
+% I think my method was too crude and the covariance estimate was too bad in
+% some cases
+
+% @gui: i'm not convinced you need alg 2 in these tables at all.
+if i were to add alg 2 to these tables, however, i would do the same thing i would for the figs, that is, for each simulation, compute the diff between performance on alg 2 and alg 3, and report that.
+% @jovo: yes I agree. I included new plots in the fig, but didn't do anything
+% for the tables cuz I don't think necessary
+
+% @jovo: added this paragraph
+The main goal of this section is threefold. First, we want to compare
+energy clustering, based on the kernels provided by the standard energy
+distance $\mathcal{E}_\alpha$, to $k$-means and GMM. Second, we want
+to compare $\mathcal{E}^H$-clustering, based on Hartigan's method, to
+$\mathcal{E}^L$-clustering, based on Lloyd's method. We also compare
+to $\mathcal{E}^S$-clustering
+which is an exact solution to the relaxed problem.
+Third, we want to illustrate the flexibility
+provided by energy clustering which do not make assumptions about the
+distribution of the data, being able to cluster accurately in different
+settings while keeping the same standard kernel.
+
 In the experiments below we fix the semimetric 
-according to the traditional energy distance \eqref{eq:energy}, and
+according to the traditional energy distance, $\alpha=1$, and
 the point $x_0=0$ is chosen in the corresponding kernel  
-\eqref{eq:kernel_semimetric}. We therefore use
+\eqref{eq:kernel_semimetric}. We thus have
 \begin{equation}
@@
 \end{equation}
-We will consider other kernels as well 
+We shall consider other kernels as well, 
 but \eqref{eq:standard_metric} will be the
@@
 
-Let us briefly mention that we compared $\mathcal{E}^H$-clustering,
+We briefly mention that we compared $\mathcal{E}^H$-clustering,
 as described in Algorithm~\ref{algo}, to
-the $1D\,\mathcal{E}$-clutering, described in Algorithm~\ref{algo1d}, 
+$\mathcal{E}^{1D}$-clutering, described in Algorithm~\ref{algo1d}, 
 for several univariate distributions.
@@
 However, we omit these results since we will analyse more interesting 
-scenarios in high dimensions and $k > 2$ number of clusters.
-
-The main goal of the 
-experiments  to follow
-is to compare Algorithm~\ref{algo} based on Hartigan's method to
-kernel $k$-means, as described in Algorithm~\ref{kmeans_algo}, which
-is based on Lloyd's method. 
-% @gui: i might call them \mathcal{E}-H & \mathcal{E}-L, or some such?
-% @jovo: it might be a good idea, let's decide when I re-do the plots
-From the discussion in the end of the previous section we can anticipate
-that Algorithm~\ref{algo} will be superior than Algorithm~\ref{kmeans_algo}.
-Another goal is to illustrate the nonparametric aspect of energy statistics.
-To this end we also compare
-Algorithm~\ref{algo} to standard $k$-means and GMM/EM since 
-these are reference
-clustering algorithms in practice.
-Moreover, for every algorithm, we always
-choose the initialization procedure from $k$-means++%
-\footnote{Notice that we just use the initialization
-procedure and not the full $k$-means++ algorithm.} \cite{Vassilvitskii}.
-Our measure of clustering quality will be the accuracy \eqref{eq:accuracy}
-based on the ground truth. Furthermore, in every experiment, we sample
-data many times and show the average value of the accuracy with error bars
-indicating the standard error. Whenever possible, 
-% @gui: bayes error is always possible to numerically estimate, no? maybe sometimes computationally intractable?
-% @jovo: true. I just didn't indicate Bayes error for Fig. 3 for instance cuz
-% will be very boring, will have to compute for every number of points, and
-% also we don't do that for MNIST and circles, thus the phrase
-we also indicate the optimal 
-accuracy computed from
-Bayes classification error.
-
-From the results of \cite{Telgarsky}, we expect the improvement of Hartigan's 
+scenarios in high dimensions.
+
+From the results of \cite{Telgarsky}, summarized in the end of the previous
+section, we expect the improvement of Hartigan's 
 over Lloyd's method to be more accentuated in high dimensions.
 Thus, we analyze
-how the algorithms degrade as the number of dimensions increase, while
-keeping the number of points in each cluster fixed. Consider
+how the algorithms degrade as the number of dimensions increase while
+keeping the number of points in each cluster fixed.
+Consider
 two clusters with multivariate normal distributions given by
@@
 Note that the Bayes error
-is fixed as $D$ increases, giving an optimal 
+is fixed as $D$ increases, yielding an optimal 
 accuracy of $\approx 0.86$.
@@
 sampling $100$ points for each cluster.
-% @gui: if you sample 100 points from each cluster, it breaks the GMM model, which assumes you sample n_i points from cluster i with probability \pi_i.
-i don't think it matters that much, but good to know :)
-% @jovo: humm ... true, didn't think about that. Do you think I should re-do
-% all experiments in this way? sampling from clusters with prob 1/2 for
-% instance, instead of exactly picking a fixed number of points for each?
 We apply each algorithm to the resulting dataset 
-and compute the average of the accuracy \eqref{eq:accuracy} (error bars
-are standard error).
-Algorithm~\ref{algo} and Algorithm~\ref{kmeans_algo} 
+and compute the accuracy \eqref{eq:accuracy}.
+$\mathcal{E}^H$- and $\mathcal{E}^L$-clustering
 both use the standard
-kernel \eqref{eq:standard_metric}.
-The results are shown in Fig.~\ref{fig:gauss}a.
-Note that GMM/EM is unable to estimate the covariance matrices
-when the number of dimensions exceeds the number of points in each cluster,
-%giving zero accuracy when dimensions are $\gtrsim 100$.
-i.e. when $D \gtrsim 100$.
-We see that Algorithm~\ref{algo} performs better than all the other ones,
-and in particular it outperforms kernel $k$-means, Algorithm~\ref{kmeans_algo},
-as the number
-of dimensions increase. 
-% @gui: when you say "kernel k-means" here, do you mean algorithm 2?  if so, please clarify, if not, please clarify :)
-% @jovo: will change again when we introduce algorithm's names
-
-%Note also that although GMM and $k$-means are
-%parametric they are consistent models to the data.
-Note that in this case GMM and $k$-means are consistent estimators.
-
-% @gui: gmm & k-means are not 'models". they are algorithms, or in this context, they are estimators.
-% @jovo: see if it's correct now. 
-% Don't know how to say this, what I wanted to say is
-% that GMM and k-means are algorithms that were built based on gaussian
-% mixtures models, and they should perform close to optimal when data
-% is normally distributed.
+kernel \eqref{eq:standard_metric}, as $\mathcal{E}^S$-clustering\footnote{We
+compared our implementation of spectral clustering provided in
+Algorithm~\ref{spectralalgo} with the spectral clustering routine provided
+by \emph{scikit-learn} library \cite{scikit-learn}, 
+which is based on \cite{Malik}, and in the few
+cases considered the algorithms had similar performance.}, 
+and we initialize the algorithms
+with the initialization procedure of $k$-means++ \cite{Vassilvitskii}.
+For $k$-means and GMM we use the implementation of \emph{scikit-learn} library
+\cite{scikit-learn}, where $k$-means is also initialized with $k$-means++,
+while GMM is initialized with the output of $k$-means. Every algorithm
+is ran $5$ times, picking the best objective function value.
+We plot the average accuracy versus $D$, with error bars indicating
+standard error.
+The results are shown in Fig.~\ref{fig:gauss}a, where 
+$\mathcal{E}^H$-clustering and $\mathcal{E}^S$-clustering have practically
+the same performance in this experiment, which is 
+much higher than $\mathcal{E}^L$-clustering (kernel
+$k$-means), $k$-means, and GMM. The improvement is noticeable 
+specially in high dimensions.
+Note that in this setting, $k$-means and GMM are consistent estimators,
+however, energy clustering degrades much less as dimension increases.
 
@@
 \begin{minipage}{0.49\textwidth}
-\includegraphics[width=1\textwidth]{gauss_dim.pdf}\\[-1.0em] (a)
+\includegraphics[width=1\textwidth]{normal_highdim_mean.pdf}\\[-1.0em] (a)
 \end{minipage}
 \begin{minipage}{0.49\textwidth}
-\includegraphics[width=1\textwidth]{gauss_cov_squareroot.pdf}\\[-1.0em] (b)
-\end{minipage}
-\begin{minipage}{0.49\textwidth}
-\includegraphics[width=1\textwidth]{gauss_cov_linear.pdf}\\[-1.0em] (c)
-\end{minipage}
-\begin{minipage}{0.49\textwidth}
-\includegraphics[width=1\textwidth]{gauss_cov_square.pdf}\\[-1.0em] (d)
+\includegraphics[width=1\textwidth]{normal_highdim_cov.pdf}\\[-1.0em] (b)
 \end{minipage}
@@
 \label{fig:gauss}
-Comparison of Algorithm~\ref{algo}, Algorithm~\ref{kmeans_algo},
-standard $k$-means and GMM/EM as the number of dimensions increase
-in Gaussian settings.
-% @gui: gaussian is always captilized
-% @gui: also, i don't understand this plot entirely, are you really starting with D=1?
-or 10?
-from the equations, it would seem that Bayes error changes when D< 10, but your Bayes error lines do not seem to change?
-% @jovo: I start with D = 10.
-% @gui: cep points out that in Fig 2a, GMM does very poorly even with d small, which does not make sense to us.  can you explain?
-% @jovo: you're right, didn't realize that. Will double check and maybe repeat
-% the experiment.
-We compute the average of \eqref{eq:accuracy} over $100$ samples with
-error bars being standard error. We have two clusters with $100$ points
-each. 
-(a) Data as in \eqref{eq:gauss1}, where the optimal accuracy
-from Bayes error is the dashed line equal to $\approx 0.86$.
-(b) Data from \eqref{eq:gauss2} with $q=1/2$ in \eqref{eq:cov}. 
-(c) Data from \eqref{eq:gauss2} with $q=1$ in \eqref{eq:cov}.
-(d) Data from \eqref{eq:gauss2} with $q=2$ in \eqref{eq:cov}.
-The optimal accuracy from Bayes error in (b--d) is $\approx 1$.
-% @gui: titles for each, or subcaptions, would be extremely useful.  in general, titles should provide the "setting" of the experiment, ie, the model, etc.
-% @jovo: ok will do when I re-generate the plots.
+Comparison of energy clustering algorithms to $k$-means and GMM in 
+high dimensional Gaussian settings. We plot the mean accuracy versus
+the number of dimensions, with error bars indicating standard 
+error from $100$ Monte
+Carlo runs.
+(a) Data distributed as in \eqref{eq:gauss1}, with Bayes
+accuracy $\approx 0.86$, over the range $D \in [10,200]$. 
+(b) Data distributed
+as in \eqref{eq:gauss2}, with Bayes accuracy $\approx 0.95$,
+over the range $D \in [10, 700]$.
 % @gui: also helpful here would be pairs plots, say, of the first 5 dimensions, showing the performance of our stuff (color points by Alg 3, maybe symbols denote "truth".
@@
 
-%Let us now consider the same experiment as before with data distributed as
-Consider now the following setting:
-% @gui: what do you mean "the same experiment as before"? clarify, or don't say that. there are obviously differences. 
-% @jovo: ok
+We now allow the diagonal covariance of the second Gaussian distribution to
+have different values by choosing
+$(\Sigma_2)_{ii} \stackrel{iid}{\sim} \textnormal{Uni}(1,5)$ for
+$i=1,\dotsc,10$, and $(\Sigma_2)_{ii} = 1$ for $i=11,\dotsc,D$. For instance,
+we consider the fixed setting
 \begin{equation}
 \label{eq:gauss2}
-x \in \C_i  \sim 
-\mathcal{N}(\mu_i,\Sigma_i) \quad (i=1,2),  \quad
-\mu_1 = (\underbrace{0,\dotsc,0}_{\times D})^\top , \quad
+\begin{split}
+x & \in \C_i  \sim 
+\mathcal{N}(\mu_i,\Sigma_i) \quad (i=1,2),  \\
+\mu_1 &= (\underbrace{0,\dotsc,0}_{\times D})^\top , \quad
 \mu_2 = (\underbrace{1,\dots,1}_{\times 10},
-\underbrace{0,\dots,0}_{\times (D-10)})^\top, 
-\end{equation}
-where
-\begin{equation}
-\label{eq:cov}
-(\Sigma_1)_{ij} = \begin{cases}
-i^{-q} \delta_{ij} & \mbox{if $i \le 10$} \\
-\delta_{ij} & \mbox{if $10 < i \le D$}
-\end{cases} \qquad
-(\Sigma_2)_{ij} = \begin{cases}
-i^{\,q} \delta_{ij} & \mbox{if $i \le 10$} \\
-\delta_{ij} & \mbox{if $10 < i \le D$}
-\end{cases} 
-\end{equation}
-and we choose $q \in \{ 1/2, 1, 2\}$. Above $\delta_{ij} = 1$ if $i=j$ and
-$\delta_{ij}=0$ if $i\ne j$ is the Kronecker delta. In these cases,
-% @gui: what is delta_ij??? can you explain the intuition for this setting, i don't get it, maybe because i don't know what delta_ij is.....
-% @jovo: delta_{ij} is the Kronecker delta, I included the definition
-the best possible accuracy from Bayes classification error is $\approx 1$.
-In Fig.~\ref{fig:gauss}b--d we have $q=1/2$, $q=1$, and $q=2$, respectivelly.
-Again, GMM/EM is unable to estimate the covariance matrices
-as dimensions get larger than $\gtrsim 100$, and it gives poor results even
-for number of dimensions much lower than this. Note that GMM requires
-a larger number of points to estimate the parameters accurately.
-Algorithm~\ref{algo} outperforms
-Algorithm~\ref{kmeans_algo}, and $k$-means degrades faster as 
-$q$ increases.
-
-To summarize, in the experiments of Fig.~\ref{fig:gauss}
-we see a better performance of Algorithm~\ref{algo} compared
-to the other ones, and in particular to kernel $k$-means,
-where we recall that both find local solutions
-to the same optimization problem \eqref{eq:qcqp2}.
-Algorithm~\ref{algo} is more robust as the number of dimensions
-increase.
-% @gui: is there any reason to initialize alg 3 with alg 2? is it even worth trying?
-% @jovo: based on the theorems I cited for Hartigan's this is actually worse,
-% since the set of local minima of Hartigan's is a subset of Lloyd's, thus if
-% you do that Hartigan's may not iterate. However, Hartigan's has possibly less
-% local minima than Lloyd's so if you start at random, or not on a local
-% minima it may converge to a better solution
+\underbrace{0,\dots,0}_{\times (D-10)})^\top, \quad
+\Sigma_1 = I_D, \quad
+\Sigma_2 = \left( \begin{array}{c|c}
+\widetilde{\Sigma}_{10} & 0 \\ \hline 
+0 & I_{D-10} \end{array}\right), \\
+\widetilde{\Sigma}_{10} &= \diag(1.367,  3.175,  3.247,  4.403,  1.249,
+1.969, 4.035,   4.237,  2.813,  3.637),
+\end{split}
+\end{equation}
+as we vary $D$. Note that there is nothing special about the numbers in
+$\widetilde{\Sigma}_{10}$, it is simply a random choice of sampling
+$\textnormal{Uni}(1,5)$ ten times, and one could pick any other choices
+as well.
+The Bayes error is kept fixed yielding
+an optimal accuracy $\approx 0.95$.
+In Fig.~\ref{fig:gauss}b we show the results of energy clustering,
+$k$-means, and GMM, where we do $100$ Monte Carlo runs, sampling
+$100$ points for each cluster. The results of $\mathcal{E}^H$-
+and $\mathcal{E}^S$-clustering are superior than the other algorithms, 
+specially
+in high dimensions.
 
@@
 \Sigma_2 = \left( 
-\begin{smallmatrix} 
-1/2 & 0 & 0 & 0\\
-0 & 1/2 & 0 & 0 \\
-0 & 0 & 1 & 0 \\
-0 & 0 & 0 & 1 
-\end{smallmatrix}\right), \quad
-n_1 = N - m, \quad  n_2 = N + m, \quad N=200.
+\begin{array}{c|c} 
+\tfrac{1}{2} I_2 & 0  \\ \hline
+0 & I_2 
+\end{array}\right), \quad
+n_1 = N - m, \quad  n_2 = N + m, \quad N=250.
 \end{split}
 \end{equation}
-We then increase $m$, 
-that is we make the clusters progressively more unbalanced,
-and plot the average of \eqref{eq:accuracy} 
-over
-$100$ samples for each $m$
-(error bars are standard error). The results are in Fig.~\ref{fig:unbalanced}.
+We then increase $m \in [0,240]$ making
+the clusters progressively more unbalanced.
+We plot the average accuracy over
+$100$ Monte Carlo runs for each $m$, with error bars indicating 
+standard error. 
+The results are shown in Fig.~\ref{fig:unbalanced}.
 For highly unbalanced clusters we see that GMM performs better than
-the other methods which have similar performance.
-%As expected, GMM
-%works better than the other algorithms since it is a fuzzy clustering.
-%The other methods based on hard assignments degrade similarly
-%and  more rapidly than GMM. This indicates that a soft version of
-%Algorithm~\ref{algo} should compensate for this effect.
-% @gui: in stats, we don't say "fuzzy", we say "soft".
-also, why do you think that when clusters are imbalanced that soft clustering is more important?
-i would say that soft clustering is more important when the cluster centers are close, i don't see how imbalanced matters.
-% @jovo: you are probably right, I'll remove this phrase. 
-% I thought that this should be the case because in this setting that's pretty
-% much the only difference between GMM and k-means
+the other methods, which have basically similar performance.
+Based on this experiment, an interesting problem would be to
+extend $\mathcal{E}^H$-clustering algorithm to account for 
+highly unbalanced clusters.
 
 \begin{figure}
-\includegraphics[width=0.5\textwidth]{gauss_pi.pdf}\vspace{-1em}
+\includegraphics[width=0.5\textwidth]{normal_unbalanced.pdf}\vspace{-1em}
 \caption{
 \label{fig:unbalanced}
-Comparison of Algorithm~\ref{algo}, Algorithm~\ref{kmeans_algo}, 
-$k$-means, and GMM/EM. The data is distributed as \eqref{eq:gauss3} where
-we make the clusters progressively more unbalanced.
+Comparison of energy clustering algorithms to $k$-means and GMM on
+unbalanced clusters. The data is distributed as \eqref{eq:gauss3}, where
+we vary $m \in [0, 240]$, and in each case we do $100$ Monte Carlo runs
+showing the average accuracy with standard error.
 }
@@
 Besides the standard kernel from energy statistics 
-\eqref{eq:standard_metric} 
-% @gui: please remove all "let us" from manuscript
-% @jovo: ok, hunting them
+\eqref{eq:standard_metric}, 
 consider the following two other semimetrics with their respective generating
@@
 \end{align}
-The kernel $K_{1/2}(x,y)$ corresponds to the energy distance
-\eqref{eq:energy2} with $\alpha=1/2$.
-We sample data from the following normal distribution in $D=20$:
+The kernel $K_{1/2}(x,y)$ corresponds to the standard energy distance
+with $\alpha=1/2$.
+We sample data from the following normal distribution in $20$ dimensions:
 \begin{equation}
@@
 \end{equation}
-We sample an equal number of points for each cluster, which is progressively
-increased. The optimal accuracy based on Bayes
+The optimal accuracy based on Bayes
 classification error is $\approx 0.90$. 
-Clustering results are shown in Fig.~\ref{fig:consist}a.
-Algorithm~\ref{algo} outperforms the other ones, and in 
-particular the kernel \eqref{eq:rhoe} provides better results.
-As the number of points get large enough GMM starts to approach
-optimal Bayes, as it should since it is a
-consistent model to the data. However, 
-Algorithm~\ref{algo} with kernel \eqref{eq:rhoe} approach optimal Bayes
-with a much smaller number of points. Moreover, Algorithm~\ref{algo} 
-outperforms
-Algorithm~\ref{kmeans_algo} for any of the kernel choices.
-% @gui: this is a separate point, that i would make a separate figure for if you want to make it.
-then, one fig would show E-H with 3 kernels vs.
-kmeans & gmm, and then another would show the difference between accuracy of E-H vs E-L for each of the 3 kernels.
-% @jovo: ok, point taken, will come together with other plots
-
-Now consider the same parameters as in \eqref{eq:20gauss} but with
-lognormal distributions, 
-\begin{equation}
-\label{eq:20loggauss}
-\log x \in \C_i \sim \mathcal{N}(\mu_i, \Sigma_i) \qquad (i=1,2).
-\end{equation}
-The same previous experiment is shown in 
-Fig.~\ref{fig:consist}b.
-Note that Algorithm~\ref{algo} still performs accurately, 
-while $k$-means works almost at chance,
-and GMM is not even able to estimate the parameters. 
-% @gui: something is weird with the way you care computing accuracy.  GMM should not be significantly worse than chance.  i think you are probably counting the NaN's as *wrong*, rather than nothing?
-% @jovo: yes, double checking that
-Again, the
-kernel \eqref{eq:rhoe}
-provides better results than \eqref{eq:standard_metric} or
-\eqref{eq:rhohalf}. 
-The experiments in Fig.~\ref{fig:consist}
-illustrate how energy statistics clustering is nonparametric.
+%Since previous results should be already convincing enough that
+%$\mathcal{E}^H$-clustering, based on Hartigan's method, is superior
+%to $\mathcal{E}^L$-clustering, which is based on Lloyd's method, in the
+%following 
+We sample an equal number of points for each cluster which is progressively
+increased, $n \in [5, 200]$. 
+The results over $100$ Monte Carlo trials are 
+shown in Fig.~\ref{fig:consist}a, where we compare
+$\mathcal{E}^H$-clustering, with different
+kernels, to $k$-means and GMM.
+Note that for small $n$ all methods
+are superior than GMM, which slowly catches up and tend to optimal Bayes,
+as expected since it is a consistent estimator in this case. 
+Note also that $\mathcal{E}^H$-clustering
+with kernel $\kk_e$ is as accurate as GMM for large number of points, however,
+it is superior for small number of points. Still for
+the same setting, in Fig.~\ref{fig:consist}b we
+we show the difference in accuracy provided by $\mathcal{E}^H$- minus
+$\mathcal{E}^L$- and $\mathcal{E}^S$-clustering, when using the
+kernel $\kk_e$.
+Note that $\mathcal{E}^H$ was
+always superior than the other methods.
 
@@
 \centering
-\includegraphics[width=1\textwidth]{gauss.pdf}\\[-1.0em]
-(a)
+\includegraphics[width=1\textwidth]{normal_kernels.pdf}\\[-1.0em](a)
 \end{minipage}
@@
 \centering
-\includegraphics[width=1\textwidth]{loggauss.pdf}\\[-1.0em]
-(b)
+\includegraphics[width=1\textwidth]{lognormal_kernels.pdf}\\[-1.0em](c)
+\end{minipage}
+\begin{minipage}{0.49\textwidth}
+\centering
+\includegraphics[width=1\textwidth]{normal_kernels_difference.pdf}\\[-1.0em](b)
+\end{minipage}
+\begin{minipage}{0.49\textwidth}
+\centering
+\includegraphics[width=1\textwidth]{lognormal_kernels_difference.pdf}\\[-1.0em]
+(d)
 \end{minipage}
@@
 \label{fig:consist}
-% @gui: too many lines, you demonstrated that E-H is better than E-L already, i don't think we need to show the E-L variants here.
-% @jovo: will fix that
-Algorithm~\ref{algo} and Algorithm~\ref{kmeans_algo} with kernels
-\eqref{eq:standard_metric}, \eqref{eq:rhohalf} and \eqref{eq:rhoe}, 
-$k$-means, and GMM. The optimal accuracy in both cases
-is $\approx 0.9$. We show the average of \eqref{eq:accuracy}
-over $100$ samples with standard error.
-(a) Data distributed as in 
-\eqref{eq:20gauss}. 
-(b) Data distributed as in \eqref{eq:20loggauss}.
+$\mathcal{E}^H$-clustering with different kernels versus
+$k$-means and GMM. In both settings Bayes accuracy
+is $\approx 0.9$. We show average accuracy (error bars are standard error)
+versus number of points for $100$ Monte Carlo runs.
+(a,b) Normal distribution \eqref{eq:20gauss}. 
+(c,d) Lognormal distribution~\eqref{eq:20loggauss}.
+The plots in (c) and (d) consider the difference in accuracy
+between $\mathcal{E}^H$- versus $\mathcal{E}^L$- and
+$\mathcal{E}^S$-clustering.
 }
 \end{figure}
+
+Now consider the same parameters as in \eqref{eq:20gauss} but with
+lognormal distributions, 
+\begin{equation}
+\label{eq:20loggauss}
+\log x \in \C_i \sim \mathcal{N}(\mu_i, \Sigma_i) \qquad (i=1,2).
+\end{equation}
+The same previous experiment is shown in 
+Fig.~\ref{fig:consist}c.
+Energy clustering still performs accurately, with any of the utilized
+kernels, providing better results than $k$-means and GMM on
+this non-normal data. The kernel $\kk_e$ still provides the best results
+for small number of points, but its performance is eventually achieved by
+$\kk_{1/2}$, indicating that $\alpha \approx 1/2$ in the standard
+energy distance should be more appropriate for skewed distributions.
+Still for the same setting, in Fig~.\ref{fig:consist}d
+we show the difference in accuracy provided by $\mathcal{E}^H$ versus
+$\mathcal{E}^L$ and $\mathcal{E}^S$, for the kernel $\kk_e$. Again,
+the accuracy provided by $\mathcal{E}^H$ is higher than the other methods,
+although not much higher than $\mathcal{E}^S$ in this example.
+The two experiments of Fig.~\ref{fig:consist}
+illustrate how energy statistics clustering does not make assumptions
+about the distribution of the data, performing well in different settings
+with the same kernel.
 
@@
 In Fig.~\ref{fig:other} we have examples of
-complex two dimensional datasets. The two parallel cigars of 
-Fig.~\ref{fig:other}a have $200$ points each. For the concentric circles
-of Fig.~\ref{fig:other}b and Fig.~\ref{fig:other}c we
-sample $400$ points for each class.
-We apply Algorithm~\ref{algo} and Algorithm~\ref{kmeans_algo}, 
-using the above kernels 
-\eqref{eq:rhoalpha}--\eqref{eq:rhogsigma}, as well as
-$k$-means and GMM. The results are shown
-in Table~\ref{table:other} where the respective choice of parameters for the
-kernels are indicated. For the data in Fig.~\ref{fig:other}a 
+complex two dimensional datasets. The two parallel cigars in (a)
+have $200$ points each. The concentric circles
+in (b) and (c) have $400$ points for each class.
+We apply $\mathcal{E}^H$-clustering  with different choices
+of kernels in \eqref{eq:rhoalpha}--\eqref{eq:rhogsigma}. We also consider
+the best kernel choice for each example for $\mathcal{E}^S$-clustering. 
+Moreover, we consider
+$k$-means and GMM. We perform $10$ Monte Carlo runs for each example
+and show the results in Table~\ref{table:other}, where the respective 
+choice of parameters for the kernels are indicated. For (a) we initialize
+all algorithms with $k$-means++, and for (b) and (c) we initialize at
+random.
+We observe that $\mathcal{E}^H$-clustering has superior performance
+in every example, and in particular better than $\mathcal{E}^S$-clustering.
+For the data in Fig.~\ref{fig:other}a 
 the semimetrics $\rho_1$ and $\rho_{1/2}$ are
-able to provide more accurate results compared to $k$-means. However,
-the gaussian kernel $\widetilde{\rho}_2$ gives very accurate
-results, similar to GMM,
-which is a consistent estimator for this data. For the data shown in
-Fig.~\ref{fig:other}b we see
-that the clustering quality is highly sensitive to the choice of kernel,
-% @gui: you mean performance is sensitive to the kernel?
-% @jovo: fixed
-and only \eqref{eq:rhoesigma}
-was able to cluster accurately. The same kernel choice to the case
-of Fig.~\ref{fig:other}c
-% @gui: Fig 5 a? 5c?
-% @jovo: ok
-still provides better results than
-the other kernels, but the results are less accurate 
-compared to the data in Fig.~\ref{fig:other}b.
-
-\newcolumntype{g}{>{\columncolor{gray!20}}l}
+still able to provide accurate results, however, for the
+examples in (b) and (c) the kernel choice is more sensitive.
+
 \begin{table}[h]
@@
 \midrule[0.5pt]
-\multirow{4}{*}{\emph{Algorithm \ref{algo}}~~~~} 
-& $\rho_{1}$ & $0.766\pm0.066$
-& $\rho_{1}$ & $0.522\pm0.006$
-& $\rho_{1}$ & $0.437\pm0.030$ \\
-& $\rho_{1/2}$ & $0.859\pm0.062$
-& $\rho_{1/2}$ & $0.524\pm0.007$
-& $\rho_{1/2}$ & $0.547\pm0.026$ \\
-& $\widetilde{\rho}_{2}$ & $0.971\pm0.015$
-& $\widetilde{\rho}_{1}$ & $\bm{0.9999\pm0.0001}$
-& $\widetilde{\rho}_{2}$ & $\bm{0.677\pm0.003}$ \\
-& $\widehat{\rho}_{2}$  & $\bm{0.998\pm0.001}$
-& $\widehat{\rho}_{1}$  & $0.597\pm0.052$
-& $\widehat{\rho}_{2}$ & $0.645\pm0.012$ \\
+\multirow{4}{*}{\emph{$\mathcal{E}^H$-clustering~~~~}}
+& $\rho_{1}$ & $0.705\pm 0.065$
+& $\rho_{1}$ & $0.521\pm 0.005$
+& $\rho_{1}$ & $0.366\pm 0.005$ \\
+& $\rho_{1/2}$ & $0.952\pm 0.048$
+& $\rho_{1/2}$ & $0.522\pm 0.004$
+& $\rho_{1/2}$ & $0.367\pm 0.005$ \\
+& $\widetilde{\rho}_{2}$ & $\bm{0.9987\pm 0.0008}$
+& $\widetilde{\rho}_{1}$ & $0.778\pm 0.075$
+& $\widetilde{\rho}_{2}$ & $0.374\pm 0.009$ \\
+& $\widehat{\rho}_{2}$  & $0.956\pm 0.020$
+& $\widehat{\rho}_{1}$  & $\bm{1.0\pm 0.0}$
+& $\widehat{\rho}_{2}$ & $\bm{0.572\pm 0.041}$ \\
 \arrayrulecolor{gray!80}\midrule[0.5pt]
-\multirow{4}{*}{\emph{Algorithm \ref{kmeans_algo}}~~~~} 
-& $\rho_{1}$ & $0.758\pm 0.069$
-& $\rho_{1}$ & $0.516\pm 0.002$
-& $\rho_{1}$  & $0.452\pm 0.030$ \\
-& $\rho_{1/2}$ & $0.901\pm 0.060$
-& $\rho_{1/2}$ & $0.524\pm 0.007$
-& $\rho_{1/2}$ & $0.570\pm 0.016$ \\
-& $\widetilde{\rho}_{2}$ & $0.971\pm 0.015$
-& $\widetilde{\rho}_{1}$ & $\bm{0.9999\pm 0.0001}$
-& $\widetilde{\rho}_{2}$ & $\bm{0.673\pm 0.002}$ \\
-& $\widehat{\rho}_{2}$ & $\bm{0.998\pm 0.001}$
-& $\widehat{\rho}_{1}$ & $0.528\pm 0.008$
-& $\widehat{\rho}_{2}$ & $0.640\pm 0.013$ \\
+\emph{$\mathcal{E}$-spectral~~~~}
+& $\widetilde{\rho}_{2}$ & $0.557\pm 0.014$ 
+& $\widehat{\rho}_{1}$ & $0.732\pm 0.002$ 
+& $\widehat{\rho}_{2}$ & $0.370\pm 0.005$  \\
 \arrayrulecolor{gray!80}\midrule[0.5pt]
 \emph{$k$-means}~~~~ 
-& & $0.599\pm0.046$
-& & $0.521\pm0.005$
-& & $0.360\pm0.004$ \\
+& & $0.550\pm 0.011$
+& & $0.522\pm 0.004$
+& & $0.362\pm 0.004$ \\
 \emph{GMM}~~~~
-& & $0.9995\pm0.0003$
-& & $0.598\pm0.018$
-& & $0.479\pm0.021$ \\
+& & $0.903\pm 0.064$
+& & $0.595\pm 0.011$
+& & $0.520\pm 0.034$ \\
 \arrayrulecolor{black}\bottomrule[1pt]
@@
 \caption{\label{table:other}
-Clustering the data shown in Fig.~\ref{fig:other} with Algorithm~\ref{algo}
-and Algorithm~\ref{kmeans_algo}, with kernels 
-\eqref{eq:rhoalpha}--\eqref{eq:rhogsigma}, as well as $k$-means and GMM. 
-We sample $10$ times and show the average
-accuracy \eqref{eq:accuracy} with standard error.
+Clustering data from Fig.~\ref{fig:other}.
+We show average
+accuracy over $10$ Monte Carlo runs.
 }
@@
 
-Next we consider 
-the well-known MNIST handwritten digit dataset 
+\begin{table}[h]
+\renewcommand*{\arraystretch}{0.75}
+\begin{tabular}{@{}r@{}l@{}|@{}l@{}|@{}l@{}|@{}l@{}|@{}l@{}}
+\toprule[1pt]
+\emph{Class Subset}~~~~ & & 
+$~\{0,1,2\}$ &
+$~\{0,1,\dotsc,4\}~$ &
+$~\{0,1,\dotsc,6\}~$ &
+$~\{0,1,\dotsc,8\}~$ \\
+\midrule[0.5pt]
+\multirow{3}{*}{\emph{$\mathcal{E}^H$-clustering~~~~}}
+& $\rho_{1}$\hspace{1em} 
+&$~0.937\pm 0.006~$
+&$~0.873\pm 0.025~$
+&$~0.731\pm 0.016~$
+&$~0.687\pm 0.016~$
+\\
+& $\rho_{1/2}$ 
+&$~0.939\pm 0.006~$
+&$~0.874\pm 0.027~$
+&$~0.722\pm 0.017~$
+&$~0.647\pm 0.017~$
+\\
+& $\widetilde{\rho}_{\sigma}$ 
+&$~0.939\pm 0.006~$
+&$~0.847\pm 0.031~$
+&$~0.695\pm 0.023~$
+&$~0.657\pm 0.014~$
+\\
+& $\widehat{\rho}_{\sigma}$ 
+&$~0.933\pm 0.005~$
+&$~0.891\pm 0.009~$
+&$~0.759\pm 0.011~$
+&$~0.704\pm 0.011~$
+\\
+\arrayrulecolor{gray!80}\midrule[0.5pt]
+\emph{$\mathcal{E}$-spectral~~~~} 
+& $\widehat{\rho}_{\sigma}$ 
+&$~0.823\pm 0.015~$
+&$~0.769\pm 0.012~$
+&$~0.678\pm 0.014~$
+&$~0.649\pm 0.018~$
+\\
+\midrule[0.5pt]
+\rowcolor{gray!20}
+& {$\sigma$}
+& $~10.34~$
+& $~10.41~$
+& $~10.41~$
+& $~10.37~$
+\\
+\arrayrulecolor{gray!80}\midrule[0.5pt]
+\emph{$k$-means}~~~~ &
+&$~0.927\pm 0.004~$
+&$~0.878\pm 0.010~$
+&$~0.744\pm 0.008~$
+&$~0.695\pm 0.012~$
+\\
+\arrayrulecolor{gray!80}\midrule[0.5pt]
+\emph{GMM}~~~~ &
+&$~0.952\pm 0.005~$
+&$~0.839\pm 0.015~$
+&$~0.694\pm 0.010~$
+&$~0.621\pm 0.009~$
+\\
+\arrayrulecolor{black}\bottomrule[1pt]
+\end{tabular}
+\caption{\label{table:mnist}
+Clustering MNIST data from Fig.~\ref{fig:other}d.
+For each subset of digits we perform $10$ Monte Carlo runs, sampling 
+$100$ points for  each class.
+}
+\end{table}
+
+Next, we consider 
+the infamous MNIST handwritten digits
 as illustrated in Fig.~\ref{fig:other}d.
@@
 Besides the kernel \eqref{eq:rhoalpha}, we
-consider the gaussian kernel \eqref{eq:rhogsigma} with 
+consider the Gaussian kernel \eqref{eq:rhogsigma} with 
 \begin{equation}
@@
 \end{equation}
-which is computed from a 
-%sample  
-given dataset $\{ x_i \}_{i=1}^n$.
-% @gui: it is cheating if this sample is part of the data, this should be a sample of "training data", not used for subsequent testing. please clarify
-% @jovo: All I'm doing is, given a data set
-% I compute \sigma as above. Then I use this in the kernel and perform
-% clustering. How can this be cheating if I'm not using any labels, or any
-% other information besides the data I'm clustering?
-We consider subsets of the  classes $\{0,1,\dotsc,9 \}$, 
-where we sample $100$ points 
-for each class. We perform clustering through 
-Algorithm~\ref{algo},
-Algorithm~\ref{kmeans_algo}, and $k$-means.
-The results are shown in Table~\ref{table:mnist} where the kernel
-and its parameter for each case is indicated.
-Algorithm~\ref{algo} performed slightly better than $k$-means, except
-for the last column where all the methods are comparable. 
-Unsupervised clustering on MNIST dataset without any feature extraction
-or dimensionality reduction is not an easy task. For instance,
+computed from a separate
+training set.
+We consider subsets of $\{0,1,\dotsc,9 \}$, 
+sampling $100$ points 
+for each class. 
+The results are shown in Table~\ref{table:mnist}, where kernels
+and parameters are indicated.
+$\mathcal{E}^H$-clustering performs slightly better than $k$-means
+and GMM, however the difference is not considerable.
+Unsupervised clustering on MNIST without any feature extraction
+is not trivial. For instance,
 the same experiment was performed in \cite{Sapiro} where a low-rank
 transformation is learned then subsequently used in subspace clustering,
-providing very accurate results. One could explore analogous methods
+providing very accurate results. It would be interesting to 
+explore analogous methods
 for learning a better representation of the data and subsequently apply
-Algorithm~\ref{algo} for clustering.
-
-\newcolumntype{g}{>{\columncolor{gray!20}}l}
-\begin{table}[h]
-\renewcommand*{\arraystretch}{0.75}
-\begin{tabular}{@{}rl|l|l|l|l@{}}
-\toprule[1pt]
-\emph{Class Subset}~~~~ & & $\{0,1,2\}$ &
-$\{0,1,\dotsc,4\}$ &
-$\{0,1,\dotsc,6\}$ &
-$\{0,1,\dotsc,8\}$ \\
-\midrule[0.5pt]
-\multirow{3}{*}{\emph{Algorithm \ref{algo}}~~~~} 
-& $\rho_{1}$\hspace{1em} 
-&$0.907\pm 0.007$
-&$0.866\pm 0.006$
-&$0.715\pm 0.013$
-&$0.616\pm 0.019$
-\\
-& $\rho_{1/2}$ 
-&$\bm{0.918\pm 0.006}$
-&$0.849\pm 0.025$
-&$0.711\pm 0.010$
-&$0.642\pm 0.009$
-\\
-& $\widetilde{\rho}_{\sigma}$ 
-&$0.900\pm 0.007$
-&$\bm{0.871\pm 0.005}$
-&$\bm{0.719\pm 0.010}$
-&$0.630\pm 0.016$
-\\
-\arrayrulecolor{gray!80}\midrule[0.5pt]
-\multirow{3}{*}{\emph{Algorithm \ref{kmeans_algo}}~~~~} 
-& $\rho_{1}$ 
-&$0.914\pm 0.006$
-&$0.845\pm 0.023$
-&$0.664\pm 0.022$
-&$0.614\pm 0.014$
-\\
-& $\rho_{1/2}$ 
-&$0.895\pm 0.011$
-&$0.822 \pm 0.026$
-&$0.669\pm 0.021$
-&$0.591\pm 0.019$
-\\
-& $\widetilde{\rho}_{\sigma}$ 
-&$0.896\pm 0.007$
-&$0.869\pm 0.006$
-&$0.705\pm 0.016$
-&$\bm{0.646\pm 0.020}$
-\\
-\arrayrulecolor{gray!80}\midrule[0.5pt]
-\emph{$k$-means}~~~~ &
-&$0.871\pm 0.015$
-&$0.840\pm 0.022$
-&$0.707\pm 0.012$
-&$0.634\pm 0.011$
-\\
-\arrayrulecolor{black}\bottomrule[1pt]
-\end{tabular}
-% @gui: i'm not convinced you need alg 2 in these tables at all.
-if i were to add alg 2 to these tables, however, i would do the same thing i would for the figs, that is, for each simulation, compute the diff between performance on alg 2 and alg 3, and report that.
-% @jovo: will check how the difference looks and change this
-\caption{\label{table:mnist}
-Clustering the data shown in Fig.~\ref{fig:other}d with Algorithm~\ref{algo},
-Algorithm~\ref{kmeans_algo}, and $k$-means.
-We use the kernel \eqref{eq:rhoalpha} with $\alpha\in\{1,2\}$ and
-the gaussian kernel 
-\eqref{eq:rhogsigma} with $\sigma$ given by \eqref{eq:sigma}. 
-For each subset of digits we sample $10$ times and show the average
-accuracy \eqref{eq:accuracy} with standard error. We sample $100$ points for 
-each class.
-}
-\end{table}
-
-%Energy standard  : 0.991 0.00221108319357
-%Energy half      : 0.993 0.000816496580928
-%Energy rbf       : 0.99 0.00235702260396
-%Kernel standard  : 0.9915 0.00183333333333
-%Kernel half      : 0.992 0.00110554159679
-%Kernel rbf       : 0.99 0.00235702260396
-%k-means          : 0.9865 0.00258736244938
-%
-%
-%%%%%%%%%%%%%%%%%%
-%Energy standard  : 0.907333333333 0.00745024649522
-%Energy half      : 0.918 0.00638671460587
-%Energy rbf       : 0.899666666667 0.00708763485947
-%Kernel standard  : 0.913666666667 0.00648740470089
-%Kernel half      : 0.895333333333 0.0113442213494
-%Kernel rbf       : 0.895666666667 0.00723588451341
-%k-means          : 0.870666666667 0.0152412695107
-%
-%
-%
-%Energy standard  : 0.82975 0.0289550657629
-%Energy half      : 0.857 0.0138232895265
-%Energy rbf       : 0.85475 0.0164335784295
-%Kernel standard  : 0.84875 0.016386181509
-%Kernel half      : 0.79025 0.036198622595
-%Kernel rbf       : 0.8545 0.0145334862377
-%k-means          : 0.83425 0.01634289686
-%
-%
-%%%%%%%%%%%%%%%%%%%%%
-%Energy standard  : 0.8662 0.00636972003571
-%Energy half      : 0.8494 0.0250138628231
-%Energy rbf       : 0.8714 0.00536697721669
-%Kernel standard  : 0.845 0.0234932614452
-%Kernel half      : 0.8224 0.0261921107715
-%Kernel rbf       : 0.8688 0.00640971484892
-%k-means          : 0.8398 0.0224151734323
-%
-%
-%Energy standard  : 0.703333333333 0.0166295883857
-%Energy half      : 0.709833333333 0.0133690493859
-%Energy rbf       : 0.702 0.0158149015524
-%Kernel standard  : 0.688833333333 0.0195426883173
-%Kernel half      : 0.708833333333 0.0222403398366
-%Kernel rbf       : 0.7175 0.0153744416741
-%k-means          : 0.693333333333 0.0132590523468
-%
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%Energy standard  : 0.714714285714 0.0132292707827
-%Energy half      : 0.711142857143 0.00970728038727
-%Energy rbf       : 0.719142857143 0.010446277641
-%Kernel standard  : 0.663857142857 0.0220872830507
-%Kernel half      : 0.668571428571 0.0209923860509
-%Kernel rbf       : 0.705 0.0157981169405
-%k-means          : 0.707428571429 0.0115799498313
-%
-%Energy standard  : 0.671625 0.0167819450766
-%Energy half      : 0.652875 0.0162233221183
-%Energy rbf       : 0.686375 0.0142173724913
-%Kernel standard  : 0.674875 0.012649728258
-%Kernel half      : 0.654375 0.0232537706023
-%Kernel rbf       : 0.706 0.0188513925215
-%k-means          : 0.6705 0.0121037184369
-%
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%Energy standard  : 0.615555555556 0.0190076152525
-%Energy half      : 0.641888888889 0.00939463992175
-%Energy rbf       : 0.630333333333 0.0158222725065
-%Kernel standard  : 0.614444444444 0.0136113000743
-%Kernel half      : 0.590777777778 0.0188685802226
-%Kernel rbf       : 0.645888888889 0.0202454418315
-%k-means          : 0.634222222222 0.0111083947297
-%
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%Energy standard  : 0.5369 0.0136100044902
-%Energy half      : 0.5395 0.0138782403624
-%Energy rbf       : 0.5276 0.00942125257065
-%Kernel standard  : 0.518 0.0162446572413
-%Kernel half      : 0.5193 0.0146674620996
-%Kernel rbf       : 0.5289 0.0144001928999
-%k-means          : 0.5066 0.0120103658932
+$\mathcal{E}^H$-clustering.
 
@@
 We considered clustering from the perspective of energy
-statistics theory, coined energy clustering for short, 
-which provides a nonparametric test for 
+statistics, which provides a non-parametric test for 
 equality of distributions.
-We showed that the mathematical formulation of energy clustering 
-reduces to a quadratically
-constrained quadratic program, 
-as described in Proposition~\ref{th:qcqp2}.
-Moreover, the energy clustering optimization problem is equivalent
-to kernel $k$-means optimization problem, once the kernel is fixed; see
+The mathematical formulation of energy clustering 
+reduces to a QCQP in the associated RKHS, as demonstrated in 
+Proposition~\ref{th:qcqp2}.
+We showed that the optimization problem
+is equivalent
+to kernel $k$-means, once the kernel is fixed; see
 Proposition~\ref{th:kernel_kmeans}. Energy statistics, however, fixes
-a family of standard kernels consistent with the energy 
-distance \eqref{eq:energy2}.
-More general kernels 
-%related to the energy distance \eqref{eq:energy3},
-defined on semimetric spaces of negative type can also be obtained.
-%Our results imply that kernel $k$-means optimization problem
-%can be interpreted as a consequence of energy clustering.
-% @gui: "this"? what is "this:? clarify
-% @jovo: removed, i think it's enough the way it is already
-%thus placing it this 
-%method into a principled statistical basis.
-We also considered a weighted version of energy statistics whose 
-clustering formulation establishes connections with spectral
-clustering and graph partitioning problems.
-We proposed the $\mathcal{E}^H$-clustering algorithm based on 
-Hartigan's method, and compared with $\mathcal{E}^L$-clustering algorithm
+a family of standard kernels in Euclidean space, and
+more general kernels 
+on semimetric spaces of negative type can also be obtained.
+We also considered a weighted version of energy statistics, whose 
+clustering formulation establishes connections with 
+graph partitioning.
+We proposed the iterative $\mathcal{E}^H$-clustering algorithm based on 
+Hartigan's method, which was compared to kernel $k$-means algorithm
 based on Lloyd's heuristic.
-%, which is the same as kernel $k$-means algorithm.
-Both algorithms have the same time complexity, however, the numerical 
-results provide compelling evidence that $\mathcal{E}^H$-clustering
-is more robust with
-a superior clustering performance. The fact that $\mathcal{E}^H$-clustering
-has more desirable properties compared to $\mathcal{E}^L$-clustering
-is also supported by theoretical results, 
-as described in the end of Section~\ref{sec:algo}.
+Both have the same time complexity, however, numerical and theoretical
+results provide compelling evidence that $\mathcal{E}^H$
+is more robust with a superior performance, specially in high
+dimensions. 
+Furthermore, energy clustering, with standard kernels from energy
+statistics, outperformed $k$-means and GMM
+on several settings, illustrating the flexibility
+of the proposed method which is distribution-free. In many
+settings, the iterative
+$\mathcal{E}^H$-clustering also surpassed $\mathcal{E}^S$-clustering, which
+is a spectral relaxation of the QCQP.
 
@@
 of room to make $\mathcal{E}^H$-clustering algorithm more scalable.
-%For instance, an approach to make kernel $k$-means scalable
-%was recently proposed \cite{Mahoney}, and a modified technique for
-%rank reduction in Nystr\"om method was also recently considered 
-%\cite{Becker}. It would be interesting to apply these and related
-%techniques to the case of $\mathcal{E}^H$-clustering algorithm.
-%Another fruitful avenue of exploration would be to find 
-%better methods to tackle the optimization problem \eqref{eq:qcqp2}.
 
@@
 % Besides a summary I don't know what else to say
-
-\subsection*{Acknowledgements}
+% I modified above but check if still not good enough
+
+\begin{acknowledgements}
+\vspace{-1em}
 We would like to thank Carey Priebe for discussions.
 This work was supported by NIH TRA  grant.
+\end{acknowledgements}
 

### 2017-10-03 energy_clustering/code/experiment2.py

@@ -14,16 +14,13 @@ import plot
 
 def gauss_dimensions_mean(dimensions=range(2,100,20), num_points=[100, 100],
                           num_experiments=100, d=None):
-    """Here we keep signal in one dimension and increase the ambient dimension.
-    The covariances are kept fixed.
-    
-    """
+    # data distribution
     k = 2
     delta = 0.7
     if not d:
         d = dimensions[0]
     n1, n2 = num_points
-    table = np.zeros((num_experiments*len(dimensions), 5))
+    table = np.zeros((num_experiments*len(dimensions), 6))
     count = 0
 
     for D in dimensions:
@@ -39,27 +36,32 @@ def gauss_dimensions_mean(dimensions=range(2,100,20), num_points=[100, 100],
             rho = lambda x, y: np.linalg.norm(x-y)
             G = eclust.kernel_matrix(X, rho)
 
+            # can change the number of times we execute each experiment
+            # and initialization method as well
             table[count, 0] = D
             table[count, 1] = run_clustering.energy_hartigan(k, X, G, z, 
-                                    init="spectral", run_times=1)
+                                                init="k-means++", run_times=5)
             table[count, 2] = run_clustering.energy_lloyd(k, X, G, z, 
-                                    init="spectral", run_times=1)
-            table[count, 3] = run_clustering.kmeans(k, X, z)
-            table[count, 4] = run_clustering.gmm(k, X, z)
-
+                                                init="k-means++", run_times=5)
+            table[count, 3] = run_clustering.energy_spectral(k, X, G, z, 
+                                                init="k-means++", run_times=5)
+            table[count, 4] = run_clustering.kmeans(k, X, z, 
+                                                init="k-means++", run_times=5)
+            table[count, 5] = run_clustering.gmm(k, X, z, 
+                                                init="kmeans", run_times=5)
+            
             count += 1
 
     return table
 
 def gauss_dimensions_cov(dimensions=range(2,100,20), num_points=[100, 100],
-                         num_experiments=100, d=None):
+                         num_experiments=100, d=10):
     """High dimensions but with nontrivial covariance."""
     k = 2
-    q = 0.5
     if not d:
         d = dimensions[0]
     n1, n2 = num_points
[…]

### 2017-10-03 energy_clustering/code/experiment3.py

@@ -17,8 +17,8 @@ def gauss_dimensions_pi(num_points=range(0, 180, 10), num_experiments=10):
     k = 2
     D = 4
     d = 2
-    N = 210
-    table = np.zeros((num_experiments*len(num_points), 5))
+    N = 250
+    table = np.zeros((num_experiments*len(num_points), 6))
     count = 0
 
     for p in num_points:
@@ -34,15 +34,19 @@ def gauss_dimensions_pi(num_points=range(0, 180, 10), num_experiments=10):
 
             X, z = data.multivariate_normal([m1, m2], [s1, s2], [n1, n2])
             rho = lambda x, y: np.linalg.norm(x-y)
-            G = ke.kernel_matrix(X, rho)
+            G = eclust.kernel_matrix(X, rho)
 
             table[count, 0] = p
-            table[count, 1] = eclust.energy_hartigan(k, X, G, z, 
-                                    init="spectral", run_times=1)
-            table[count, 2] = eclust.energy_lloyd(k, X, G, z, 
-                                    init="spectral", run_times=1)
-            table[count, 3] = eclust.kmeans(k, X, z)
-            table[count, 4] = eclust.gmm(k, X, z)
+            table[count, 1] = run_clustering.energy_hartigan(k, X, G, z,
+                                                init="k-means++", run_times=5)
+            table[count, 2] = run_clustering.energy_lloyd(k, X, G, z,
+                                                init="k-means++", run_times=5)
+            table[count, 3] = run_clustering.energy_spectral(k, X, G, z,
+                                                init="k-means++", run_times=5)
+            table[count, 4] = run_clustering.kmeans(k, X, z,
+                                                init="k-means++", run_times=5)
+            table[count, 5] = run_clustering.gmm(k, X, z,
+                                                init="kmeans", run_times=5)
 
             count += 1
 
@@ -61,10 +65,12 @@ def make_plot(*data_files):
     p.xlabel = 'number of unbalanced points'
     p.legends = [r'$\mathcal{E}^{H}$-clustering', 
                  r'$\mathcal{E}^{L}$-clustering', 
+                 r'$\mathcal{E}$-spectral', 
                  r'$k$-means', 
-                 'GMM']
-    p.colors = ['b', 'r', 'g', 'm']
-    p.symbols = ['o', 's', '^', 'v']
+                 r'GMM']
+    p.colors = ['b', 'r', 'g', 'm', 'c']
[…]

### 2017-10-03 energy_clustering/code/experiment5.py

@@ -1,0 +1,177 @@
+"""Other experiments such as cigars, circles and MNIST."""
+
+# Guilhere Franca <guifranca@gmail.com>
+# Johns Hopkins University, Neurodata
+
+from __future__ import division
+
+import numpy as np
+import multiprocessing as mp
+import cPickle, gzip
+import scipy.stats
+
+import energy.data as data
+import energy.eclust as eclust
+import run_clustering
+import energy.gmm
+import energy.metric
+
+
+# some kernels/semmimetric function
+rho_standard = lambda x, y: np.power(np.linalg.norm(x-y), 1)
+rho_half = lambda x, y: np.power(np.linalg.norm(x-y), 1/2)
+rho_gauss = lambda x, y: 2-2*np.exp(-np.linalg.norm(x-y)**2/4)
+rho_exp = lambda x, y, sigma: 2-2*np.exp(-np.linalg.norm(x-y)/(2*sigma))
+rho_rbf = lambda x, y, sigma: 2-2*np.exp(-np.linalg.norm(x-y)**2/(2*sigma**2))
+
+def cigars_circles(num_experiments=10, run_times=5, kind='cigars'):
+    table = []
+    for i in range(num_experiments):
+        this_experiment = []
+        
+        if kind == 'cigars':
+            m1 = [0,0]
+            m2 = [6.5,0]
+            s1 = np.array([[1,0],[0,20]])
+            s2 = np.array([[1,0],[0,20]])
+            X, z = data.multivariate_normal([m1, m2], [s1, s2], [200, 200])
+            k = 2
+            init = 'k-means++'
+        elif kind == '2circles':
+            X, z = data.circles([1, 3], [0.2, 0.2], [400, 400])
+            k = 2
+            init = 'random'
+        elif kind == '3circles':
+            X, z = data.circles([1, 3, 5], [0.2, 0.2, 0.2], [400, 400, 400])
+            init = 'spectral'
+            k = 3
+        else:
+            raise ValueError("Don't know which example to sample.")
+
+        G = eclust.kernel_matrix(X, rho_standard)
+        G_half = eclust.kernel_matrix(X, rho_half)
+        G_exp = eclust.kernel_matrix(X, lambda x,y: rho_exp(x, y, 2))
+        G_rbf = eclust.kernel_matrix(X, lambda x,y: rho_rbf(x, y, 2))
+        #G_exp = eclust.kernel_matrix(X, lambda x,y: rho_exp(x, y, 1))
[…]

### 2017-10-03 energy_clustering/code/run_clustering.py

@@ -15,20 +15,21 @@ import energy.eclust as eclust
 import energy.energy1d as energy1d
 
 def initialize(init, k, G, X):
-        if init == "spectral":
-            z0 = initialization.topeigen(k, G)
-            Z0 = eclust.ztoZ(z0)
-        elif init == "kmeans++":
-            z0 = initialization.kmeanspp(k, X)
-            Z0 = eclust.ztoZ(z0)
-        elif init == "random":
-            z0 = np.random.randint(0, k, len(X))
-            Z0 = eclust.ztoZ(z0)
-        else:
-            raise ValueError("No initialization method provided")
-        return Z0
+    """Initializaton function to use in energy Lloyd and Hartigan."""
+    if init == "spectral":
+        z0 = initialization.topeigen(k, G)
+        Z0 = eclust.ztoZ(z0)
+    elif init == "k-means++":
+        z0 = initialization.kmeanspp(k, X)
+        Z0 = eclust.ztoZ(z0)
+    elif init == "random":
+        z0 = np.random.randint(0, k, len(X))
+        Z0 = eclust.ztoZ(z0)
+    else:
+        raise ValueError("No initialization method provided")
+    return Z0
 
-def energy_lloyd(k, X, G, z, run_times=1, init="spectral"):
+def energy_lloyd(k, X, G, z, run_times=10, init="spectral"):
     """Run few times and pick the best objective function value."""
     best_score = -np.inf
     for rt in range(run_times):
@@ -43,12 +44,29 @@ def energy_lloyd(k, X, G, z, run_times=1, init="spectral"):
 
     return metric.accuracy(z, best_z)
 
-def energy_hartigan(k, X, G, z, run_times=1, init="spectral"):
+def energy_hartigan(k, X, G, z, run_times=10, init="spectral"):
     """Run few times and pick the best objective function value."""
     best_score = -np.inf
     for rt in range(run_times):
         Z0 = initialize(init, k, G, X)
         zh = eclust.energy_hartigan(k, G, Z0, max_iter=300)
+        Zh = eclust.ztoZ(zh)
+        score = eclust.objective(Zh, G)
+        
+        if score > best_score:
+            best_score = score
+            best_z = zh
+
+    return metric.accuracy(z, best_z)
+
[…]

### 2017-10-07 energy_clustering/draft/energy_clustering.tex

@@
 \label{spectralalgo}
-$\mathcal{E^S}$-clustering is an exact method to solve
+$\mathcal{E^S}$-clustering is an exact spectral method to solve
 the relaxed problem \eqref{eq:relaxed}.
@@
 \caption{\label{algo}
-$\mathcal{E}^H$-clustering: Hartigan's method for energy clustering.
-This algorithm~~~~~~~~ \\ finds local solutions to  
+$\mathcal{E}^H$-clustering is Hartigan's method for energy clustering.
+This algorithm~~~~~~ \\ finds local solutions to  
 the optimization problem \eqref{eq:qcqp2}.\hspace{\fill}
@@
 We apply each algorithm to the resulting dataset 
-and compute the accuracy \eqref{eq:accuracy}.
+and compute the accuracy \eqref{eq:accuracy}. The algorithms
 $\mathcal{E}^H$- and $\mathcal{E}^L$-clustering
 both use the standard
-kernel \eqref{eq:standard_metric}, as $\mathcal{E}^S$-clustering\footnote{We
+kernel \eqref{eq:standard_metric}, as well as the spectral
+$\mathcal{E}^S$-clustering method\footnote{We
 compared our implementation of spectral clustering provided in
@@
 by \emph{scikit-learn} library \cite{scikit-learn}, 
-which is based on \cite{Malik}, and in the few
+which is based on \cite{Malik}. In the few
 cases considered the algorithms had similar performance.}, 
 and we initialize the algorithms
-with the initialization procedure of $k$-means++ \cite{Vassilvitskii}.
+with the initialization of $k$-means++ \cite{Vassilvitskii}.
 For $k$-means and GMM we use the implementation of \emph{scikit-learn} library
@@
 while GMM is initialized with the output of $k$-means. Every algorithm
-is ran $5$ times, picking the best objective function value.
+is ran $5$ times, returning the answer with best objective function value.
 We plot the average accuracy versus $D$, with error bars indicating
 standard error.
-The results are shown in Fig.~\ref{fig:gauss}a, where 
-$\mathcal{E}^H$-clustering and $\mathcal{E}^S$-clustering have practically
-the same performance in this experiment, which is 
+The results are shown in Fig.~\ref{fig:gauss}a. We can see that
+$\mathcal{E}^H$- and $\mathcal{E}^S$-clustering have practically
+the same performance, which is 
 much higher than $\mathcal{E}^L$-clustering (kernel
-$k$-means), $k$-means, and GMM. The improvement is noticeable 
-specially in high dimensions.
-Note that in this setting, $k$-means and GMM are consistent estimators,
+$k$-means). Moreover, $\mathcal{E}^H$-clustering outperforms 
+$k$-means and GMM, where the improvement is noticeable specially in 
+high dimensions.
+Note that in this setting $k$-means and GMM are consistent estimators,
 however, energy clustering degrades much less as dimension increases.
@@
 In Fig.~\ref{fig:gauss}b we show the results of energy clustering,
-$k$-means, and GMM, where we do $100$ Monte Carlo runs, sampling
-$100$ points for each cluster. The results of $\mathcal{E}^H$-
-and $\mathcal{E}^S$-clustering are superior than the other algorithms, 
-specially
-in high dimensions.
+$k$-means, and GMM.
+Note that GMM performs better in low dimensions, 
+since it is a consistent estimator.
+However, $\mathcal{E}^H$ and $\mathcal{E}^S$ degrade much less as dimensions
+increase, performing better than $k$-means and GMM in high dimensions.
 
@@
 classification error is $\approx 0.90$. 
-%Since previous results should be already convincing enough that
-%$\mathcal{E}^H$-clustering, based on Hartigan's method, is superior
-%to $\mathcal{E}^L$-clustering, which is based on Lloyd's method, in the
-%following 
 We sample an equal number of points for each cluster which is progressively
@@
 the same setting, in Fig.~\ref{fig:consist}b we
-we show the difference in accuracy provided by $\mathcal{E}^H$- minus
-$\mathcal{E}^L$- and $\mathcal{E}^S$-clustering, when using the
+we show the difference in accuracy provided by $\mathcal{E}^H$ minus
+$\mathcal{E}^L$ and $\mathcal{E}^S$, when using the
 kernel $\kk_e$.
 Note that $\mathcal{E}^H$ was
-always superior than the other methods.
+always superior than the other 
+methods (there would be points negative values on the $y$-axis otherwise).
 
@@
 is $\approx 0.9$. We show average accuracy (error bars are standard error)
-versus number of points for $100$ Monte Carlo runs.
+versus number of points for $100$ Monte Carlo trials.
 (a,b) Normal distribution \eqref{eq:20gauss}. 
@@
 The plots in (c) and (d) consider the difference in accuracy
-between $\mathcal{E}^H$- versus $\mathcal{E}^L$- and
-$\mathcal{E}^S$-clustering.
+between $\mathcal{E}^H$ versus $\mathcal{E}^L$ and
+$\mathcal{E}^S$-clustering, with semimetric $\rho_e$.
 }
@@
 energy distance should be more appropriate for skewed distributions.
-Still for the same setting, in Fig~.\ref{fig:consist}d
+Still for the same setting, in Fig.~\ref{fig:consist}d
 we show the difference in accuracy provided by $\mathcal{E}^H$ versus
@@
 \end{align}
-In Fig.~\ref{fig:other} we have examples of
+In Fig.~\ref{fig:other}a--c we have 
 complex two dimensional datasets. The two parallel cigars in (a)
@@
 of kernels in \eqref{eq:rhoalpha}--\eqref{eq:rhogsigma}. We also consider
-the best kernel choice for each example for $\mathcal{E}^S$-clustering. 
+the best kernel choice for each example for the spectral
+$\mathcal{E}^S$-clustering. 
 Moreover, we consider
-$k$-means and GMM. We perform $10$ Monte Carlo runs for each example
-and show the results in Table~\ref{table:other}, where the respective 
-choice of parameters for the kernels are indicated. For (a) we initialize
+$k$-means and GMM. We perform $10$ Monte Carlo runs for each example.
+The results are in Table~\ref{table:other}.
+%, where the respective 
+%choice of parameters for the kernels are indicated. 
+For (a) we initialize
 all algorithms with $k$-means++, and for (b) and (c) we initialize at
 random.
-We observe that $\mathcal{E}^H$-clustering has superior performance
-in every example, and in particular better than $\mathcal{E}^S$-clustering.
+$\mathcal{E}^H$ has superior performance
+in every example, and in particular better than the
+spectral $\mathcal{E}^S$.
 For the data in Fig.~\ref{fig:other}a 
@@
 
-\begin{table}[h]
+\begin{table}[t]
 \renewcommand*{\arraystretch}{0.75}
@@
 & $\rho_{1}$ & $0.521\pm 0.005$
-& $\rho_{1}$ & $0.366\pm 0.005$ \\
+& $\rho_{1}$ & $0.393\pm 0.020$ \\
 & $\rho_{1/2}$ & $0.952\pm 0.048$
 & $\rho_{1/2}$ & $0.522\pm 0.004$
-& $\rho_{1/2}$ & $0.367\pm 0.005$ \\
+& $\rho_{1/2}$ & $0.486\pm 0.040$ \\
 & $\widetilde{\rho}_{2}$ & $\bm{0.9987\pm 0.0008}$
 & $\widetilde{\rho}_{1}$ & $0.778\pm 0.075$
-& $\widetilde{\rho}_{2}$ & $0.374\pm 0.009$ \\
+& $\widetilde{\rho}_{2}$ & $0.666\pm 0.007$ \\
 & $\widehat{\rho}_{2}$  & $0.956\pm 0.020$
 & $\widehat{\rho}_{1}$  & $\bm{1.0\pm 0.0}$
-& $\widehat{\rho}_{2}$ & $\bm{0.572\pm 0.041}$ \\
+& $\widehat{\rho}_{2}$ & $\bm{0.676\pm 0.002}$ \\
 \arrayrulecolor{gray!80}\midrule[0.5pt]
@@
 & $\widehat{\rho}_{1}$ & $0.732\pm 0.002$ 
-& $\widehat{\rho}_{2}$ & $0.370\pm 0.005$  \\
+& $\widehat{\rho}_{2}$ & $0.364\pm 0.004$  \\
 \arrayrulecolor{gray!80}\midrule[0.5pt]
@@
 & & $0.522\pm 0.004$
-& & $0.362\pm 0.004$ \\
+& & $0.368\pm 0.005$ \\
 \emph{GMM}~~~~
@@
 & & $0.595\pm 0.011$
-& & $0.520\pm 0.034$ \\
+& & $0.465\pm 0.030$ \\
 \arrayrulecolor{black}\bottomrule[1pt]
@@
 Both have the same time complexity, however, numerical and theoretical
-results provide compelling evidence that $\mathcal{E}^H$
+results provide compelling evidence that $\mathcal{E}^H$-clustering
 is more robust with a superior performance, specially in high

### 2017-10-08 energy_clustering/draft/energy_clustering.tex
more comments

@@
 
-% @gui: have other people shown a relationship between kernel k-means and 
-% say, RKHS, or something else? 
-% @jovo: any kernel method is defined on a RKHS. 
-% For instance ref [11] is the first one to consider kernel k-means in a 
-% RKHS. The problem is how do you formulate a 
-% statistical problem in a RKHS, or how do you know if there is a kernel
-% related to it. In the case of energy statistics this was done by [6], and
-% is a nontrivial result. It's based on this work that we have the theory 
-% presented in this paper
+% @gui: you didn't update the abstract?
 
@@
 
-% @jovo: I corrected pretty much all your previous points, and also re-wrote 
-% a lot of other stuff throughout the paper. 
-% I deleted your previous comments which I believe were taken
-% care of, since most of them were simple. The ones which were not, I will
-% leave it, in the beggining of each section or in the middle.
-%
 % @jovo: About ToC, I agree and wouldn't mind remove it, however it is a 
@@
 % will adapt the paper.
+
+% @gui: my experience is that most reviewers absolutely do *NOT* want it, rather, we skip it, and journals often remove it prior to publication.
 
@@
 % @gui: i'm not sure the above is true/meaningful.  nonparametric means implicitly estimating an infinite number of parameters.  without any statistical theory, i'm not quite sure what we can say.
-% @jovo: I stoped saying non-parametric, however, Kolmogorov-Smirnov is also
+% @jovo: I stopped saying non-parametric, however, Kolmogorov-Smirnov is also
 % said to be non-parametric. I think non-parametric tests also mean that
@@
 % statistical test.
+% @gui: KS test is nonparametric because it can reject for any distribution, we cannot cluster for any distribution, so we are not.
 
@@
 % I will remove the numbers from the not cited ones if you don't like
-
-% @gui: where possible, i recommend using the "name" of the quantity rather than the number of the equation.  eg, here, write S and W, rather than (13) and (12).
-% @jovo: I will do that for the main symbols, and refer to equations only
-% when necessary. But sometimes it is unnavoidable, specially
-% when referring to long 
-% expressions or derivation steps. It is way more clear and precise to say, 
-% replacing equation (1) into (2) we obtain ... than saying, 
-% replacing A into B we obtain ... and the reader has to look around
-% the whole paper to figure it out what  A, B are
+% @gui: maybe common in math, but certainly not in stats / data science. 
+% in any case, what is the package that does that?
 
@@
 % What do you think??
-
-% @gui: i don't understand the notaiton (.)_jj. is a sum missing?
-% @jovo: this is standard.  A_{ij} denotes the 
-% $(i,j)$th element of matrix A as you know. 
-% Since here A = Z^T G Z is a product of 3 matrices
-% it is common to just put a parenthesis to avoid confusion, that's all
-
-% @gui: this might be a better initialization than k-means++, which essentially initializes using a different kernel?
-% @jovo: This was a long issue this last week as you know. 
-% The spectral initialization is, often, very
-% close to a local optimum of E-clustering algorithm, and kernel k-means
-% as well. Now, since the set of local optima of Hartigan's is a subset of 
-% those of Lloyd's, if you initialize "too well", you might get stuck on 
-% a not so great local optimum of Lloyd's, which otherwise Hartigan's 
-% would be able to escape. Numerically, in most cases Hartigan's does
-% better when initializing with k-means++, or even at random.
+% @gui: i'd state it as a lemma then.
 
@@
 % positive-semidefinite. Thus, I still prefer to look in the first direction.
+% @gui: ok, did you update the text? i don't recall where i made this comment and don't want to search for it again. can you point me to the place that you say this?
 
@@
 % D^{-1} Z^T G Z which is the definition of the trace
+% @gui: obvious to you \neq obvious.  i understood, i want you to explain it.
 Now we can
@@
 on graphs.
-% @gui: can you provide some guidance here, in particular, what do the weights "mean" or "correspond to"? right now, it feels like just math.
-% @jovo: ok, added the phrase below
-% let me know if not enough. In practice the weights can be anything attributed
-% to the nodes of the graph
 Above, the weight $w_i = w(i)$ of node $i\in \mathcal{V}$ can be, for instance,
@@
 % is random in it, and multiple runs will always give the same answer
+% @gui: i know what you mean, my point is that it is unclear.
+state in the manuscript what you mean: "it is a deterministic algorithm." or state something else, that it does not require an initialization i don't feel is correct or informative.
 
@@
 % this really worthy it?
+% @gui: yes, it is really worth it, that is why i suggested it. it is trivial to do.  currently, the reader has no idea *why* the different algs get different answers.
 
@@
 % of one of them
+% @gui: ok, well why not use the same params from fig 1b in fig 1a then?
 
@@
 accuracy is $\approx 0.852$.
+% @gui: if you insist on using the equations, at least also mention something semantic about them, ie, Data distributed Gaussian (as in 46), or Data distributed log Gaussian (as in 47).
+figures & captions should stand on their own without requiring referring to the paper.
 }
+% @gui: why did you take away the plot of the distribution? those are incredibly helpful.  
 \end{figure}
@@
 kernel $k$-means algorithm \cite{Dhillon2,Dhillon} 
-but written in a more concise and clear manner. Indeed, recalling that
+but written more concisely and with the kernel induced by energy statistics. Indeed, recalling that
 $\kk(x,y)=\langle \varphi(x), \varphi(y) \rangle$ where $\varphi: \mathcal{X}
@@
 is precisely~~~~~~~~ \\ 
-kernel $k$-means algorithm. This procedure finds
+kernel $k$-means algorithm, with the kernel induced by energy statistics. This procedure finds
 local solutions to the problem \eqref{eq:qcqp2}.\hspace{\fill}
+% @gui: spacing is weird.
 }
@@
 The entire procedure is explicitly described in Algorithm~\ref{algo}, which we
-coin $\mathcal{E}^H$-clustering to emphasize that it is based on
+denote $\mathcal{E}^H$-clustering to emphasize that it is based on
 Hartigan's method.
@@
 the optimization problem \eqref{eq:qcqp2}.\hspace{\fill}
+% @gui: clarify precisely which steps are different from alg 3 in caption.
 \hspace{\fill}
@@
 $k$ increases.
-The empirical results of \cite{Telgarsky} show that 
+The empirical results of \citet{Telgarsky} show that 
 an implementation of Hartigan's method has comparable execution time 
@@
 Lloyd's method,
-but no explicit complexity was provided. In our case, we showed that both
+but no explicit complexity was provided. We show that both
 $\mathcal{E}^L$- and $\mathcal{E}^H$-clustering
@@
 
-In \cite{Slonin} Hartigan's method was applied to $k$-means problem
+% @gui: if you refer to a manuscript, it is not usually acceptable to just write the number. for example, sometimes the numbers are superscripts.  citet should work.
+In \citet{Slonin}, Hartigan's method was applied to $k$-means problem
 with any Bregman divergence. It was shown that the number of Hartigan's
@@
 which is based on \cite{Malik}. In the few
-cases considered the algorithms had similar performance.}, 
+cases considered the algorithms had similar performance.
+% @gui: dude, this is a crazy sentence to write. use sklearn, or have a *very* good reason not to.  
+% currently, you utterly fail to justify the difference.
+}, 
 and we initialize the algorithms
 with the initialization of $k$-means++ \cite{Vassilvitskii}.
+% @gui: i thought you were goign to initialize E-H & E-L with E-S?
 For $k$-means and GMM we use the implementation of \emph{scikit-learn} library
@@
 however, energy clustering degrades much less as dimension increases.
+% @gui: k-means is only consistent if the prob of each class is equal.
+but you did not actually define this parameter, because you sampled 100 from each.
+so, you gotta define the parameter at least if you want to say anythng about consistency.
+also, what do you mean k-means is a consistent estimator?
+estimator of what?
+clarify.
 
@@
 as well.
+% @gui: if there is nothing special, why not just use 2*ones?  this seems suspicious.  also, it is the result of sampling 1 time, not 10 times.
 The Bayes error is kept fixed yielding
 an optimal accuracy $\approx 0.95$.
+% @gui: bayes error changes for each random sample, unless you are integrating over all possible samples?  it is not fixed.
 In Fig.~\ref{fig:gauss}b we show the results of energy clustering,
@@
 increase, performing better than $k$-means and GMM in high dimensions.
-
+% @gui: you gotta explain why GMM doesn't utterly fail when D>n!
+
+ % @gui: maybe put this below section in the discussion?
 Consider the effect of having 
@@
 with the same kernel. 
+% @gui: algs don't make assumptions.
 
@@
 \midrule[0.5pt]
-\multirow{4}{*}{\emph{$\mathcal{E}^H$-clustering~~~~}}
+\multirow{4}{*}{\emph{$\mathcal{E}^H$~~~~}}
 & $\rho_{1}$ & $0.705\pm 0.065$
@@
 \arrayrulecolor{gray!80}\midrule[0.5pt]
-\emph{$\mathcal{E}$-spectral~~~~}
+\emph{$\mathcal{E}^S$~~~~}
 & $\widetilde{\rho}_{2}$ & $0.557\pm 0.014$ 
@@
 for each class. 
+% @gui: i don't understand these results. {0,1,2} we do worse.  why don't we show {0,1,...,9}?
+% also, why don't you use a better metric/similarity designed for images? like using pairwise alignment followed by rho?
 The results are shown in Table~\ref{table:mnist}, where kernels
@@
 % I modified above but check if still not good enough
+% @gui: better, but still important to have another paragraph between those connecting it to the existing literature.  compare this work with 3-4 other works explicitly stating the difference.
 

### 2017-10-08 energy_clustering/draft/energy_clustering.tex
more comments

@@
 
-% @jovo: About ToC, I agree and wouldn't mind remove it, however it is a 
-% tradition that most referees like to see. In this case I also included 
-% some important information about our work, and it helps to make a transition 
-% between introduction and the remaining of the paper. 
-% I would leave it, but if you prefer to remove it, let me know and I 
-% will adapt the paper.
-
-% @gui: my experience is that most reviewers absolutely do *NOT* want it, rather, we skip it, and journals often remove it prior to publication.
+% @gui: my experience is that most reviewers absolutely do *NOT* want a ToC, rather, we skip it, and journals often remove it prior to publication.
 
@@
 % the standard formulation of kernel k-means.
-Obviously, $\mathcal{E}^{L}$-clustering is precisely
+$\mathcal{E}^{L}$-clustering is precisely
 kernel $k$-means algorithm \cite{Dhillon2,Dhillon} 
@@
 % @gui: i thought you were goign to initialize E-H & E-L with E-S?
+% @gui: you gotta clearly define E-S, with a pseudocode block, just like the others.
+in particular, which matrix are you doing an eigendecomposition on, and how you are choosing the number of dimensions to embed into.
 For $k$-means and GMM we use the implementation of \emph{scikit-learn} library
@@
 % @jovo: didn't understand this
+% @gui: https://www.google.com/search?q=pairs+plots&rlz=1C5CHFA_enUS731US731&source=lnms&tbm=isch&sa=X&ved=0ahUKEwjwo9u5w-DWAhXMKyYKHYkhAKMQ_AUICigB&biw=1440&bih=776
+% to show how well separated the classes are.
 }
@@
 
+% @gui: for eq 59, E-H & E-S yield the same performance.  but for eq 63, E-H is much better than E-S.  do you have any idea why that is?
 We now allow the diagonal covariance of the second Gaussian distribution to

### 2017-10-12 energy_clustering/aistats18/energy_clustering.tex
aistats18

\documentclass[twoside]{article}
%\usepackage{aistats2018}
 % If your paper is accepted, change the options for the package
% aistats2018 as follows:
%
\usepackage[accepted]{aistats2018}
%
% This option will print headings for the title of your paper and
% headings for the authors names, plus a copyright note at the end of
% the first column of the first page.

\usepackage[compress,round]{natbib}
\usepackage{amsmath,amssymb,amsfonts,amsthm,amscd,bm,bbm}
\usepackage{mathtools}
\mathtoolsset{showonlyrefs}
\usepackage[inline]{enumitem}
\usepackage[inline]{enumitem}
\usepackage[pdftex]{graphicx}
\graphicspath{{./figs/}}
\usepackage{algorithmic}
\usepackage{algorithm}
\usepackage{multirow}
\usepackage{colortbl}
\usepackage{array,booktabs}
\usepackage{xcolor}
\usepackage{makecell}
\usepackage{pifont}
\newcommand{\cmark}{\ding{51}}
\newcommand{\xmark}{\ding{55}}

\newtheorem{theorem}{Theorem}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{assumption}[theorem]{Assumption}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{example}{Example}

\DeclareMathOperator{\aff}{aff}
\DeclareMathOperator{\st}{s.t.}
\DeclareMathOperator{\affnot}{aff_0}
\DeclareMathOperator{\conv}{conv}
\DeclareMathOperator{\relint}{relint}
\DeclareMathOperator{\vol}{vol}
\DeclareMathOperator{\range}{range}
\DeclareMathOperator{\image}{im}
\DeclareMathOperator{\nullspace}{null}
\DeclareMathOperator{\area}{area}
\DeclareMathOperator{\vspan}{span}
\DeclareMathOperator{\id}{Id}
\DeclareMathOperator{\cond}{cond}
\DeclareMathOperator{\prox}{prox}
\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator*{\minimize}{minimize}
\DeclareMathOperator{\diag}{diag}
\DeclareMathOperator{\Tr}{Tr}

\newcommand\Energy{\mathcal{E}}
\newcommand\EnergyH{\mathcal{E}^{H}}
\newcommand\EnergyL{\mathcal{E}^{L}}
\newcommand\EnergyOne{\mathcal{E}^{1D}}
\newcommand\E{\mathbb{E}}
\newcommand\kk{K}
\newcommand\kkk{h}
\newcommand\Hk{{\mathcal{H}}_{\kk}}
\newcommand\HH{\mathcal{H}}
\newcommand\C{{\mathcal{C}}}
\newcommand\tC{{\widetilde{\C}}}
\newcommand\OO{{\mathcal{O}}}
\newcommand\Zt{Y}
\newcommand{\Ind}[1]{\mathbbm{1}_{#1}}
\newcommand\e{e}
\newcommand\om{\omega}
\newcolumntype{g}{>{\columncolor{gray!20}}l}

\begin{document}

% If your paper is accepted and the title of your paper is very long,
% the style will print as headings an error message. Use the following
% command to supply a shorter title of your paper so that it can be
% used as headings.
%
%\runningtitle{I use this title instead because the last one was very long}

% If your paper is accepted and the number of authors is large, the
% style will print as headings an error message. Use the following
% command to supply a shorter version of the authors names so that
% they can be used as headings (for example, use only the surnames)
%
%\runningauthor{Surname 1, Surname 2, Surname 3, ...., Surname n}

\twocolumn[

\aistatstitle{Clustering via Generalized Energy Statistics}

\aistatsauthor{Guilherme Fran\c{c}a \And  Joshua T. Vogelstein}

\aistatsaddress{ Johns Hopkins University } ]

\begin{abstract}
Energy statistics introduces the notion of potential
energy between probability distributions, in  
close analogy to Newton's gravitational
potential in physics. We propose a principled approach to the
clustering problem based on energy statistics theory.
Our mathematical formulation establishes connection to kernel methods,
leading to a quadratically constrained quadratic program (QCQP).
To obtain local solutions of such NP-hard QCQP 
we introduce an iterative
algorithm based
on Hartigan's method, which has the same computational cost but offers
several advantages compared to  
kernel $k$-means, based on Lloyd's heuristic.
We provide carefully designed numerical experiments showing
that our method is more flexible and outperforms
kernel $k$-means, spectral clustering,
standard $k$-means and Gaussian mixture models in a variety of settings,
specially in high dimensions. 
\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}

Energy statistics \citep{Szkely2013}
provides a hypothesis test for equality of 
distributions which is achieved 
under minimum statistical potential energy. 
When probability distributions are different the 
statistical potential energy diverges as sample size increases, while tends 
to a nondegenerate limit distribution when probability
distributions are equal. 
The test statistic has compact representation
in terms of expectations of pairwise distances, providing
straightforward empirical estimates. Energy statistics
has been applied to several goodness-of-fit 
hypothesis tests, multi-sample tests of equality of distributions, 
analysis of variance \citep{RizzoVariance}, nonlinear dependence tests through
distance covariance and distance correlation, which generalizes the Pearson
correlation coefficient, and hierarchical clustering \citep{RizzoClustering} 
by extending Ward's method of minimum variance. Moreover, an application of 
energy statistics to clustering, in Euclidean spaces, was 
proposed \citep{Kgroups}, which in part motivated this paper. 
We refer to \citep{Szkely2013} for an overview.

Distance covariance was further generalized from Euclidean 
to metric spaces of strong negative type \citep{Lyons}. Furthermore, 
the missing link between energy distance based tests and kernel 
based tests has 
been recently resolved \citep{Sejdinovic2013}, where a unifying framework
establishing an equivalence between generalized energy distances to maximum
mean discrepancies (MMD), which are distances between embeddings of 
distributions in reproducing kernel Hilbert spaces (RKHS), was established. 
This equivalence immediately relates energy statistics to
kernel methods often used in machine learning, and form the basis 
of our approach.

Clustering has such a long history in machine learning, making it
impossible to mention all important contributions in a short space. 
Perhaps, the most used method is $k$-means \citep{Lloyd,MacQueen,Forgy}, which
is based on Lloyd's heuristic \citep{Lloyd} of assigning a data point to
the cluster with closest center. The only statistical 
information about each cluster comes from its mean, making it sensitive 
to outliers. Nevertheless, $k$-means works very well when data is 
linearly separable in Euclidean space. Gaussian mixture models (GMM) is 
another very common approach, providing more flexibility than $k$-means, 
however, it still makes strong assumptions about the distribution of 
the data.

To account for nonlinearities, kernel methods were introduced 
\citep{Smola,Girolami}. A mercer kernel \citep{Mercer} is used to implicitly
map data points to a RKHS, then clustering can be performed in the associated
Hilbert space by exploiting its inner product. However, the kernel choice remains 
the biggest challenge since there is no principled theory to construct a kernel
for a given dataset, and usually kernels introduce hyperparameters that 
need to be carefully chosen.
The well-known kernel $k$-means optimization problem is nothing but $k$-means 
in the feature space \citep{Girolami}. Furthermore, kernel $k$-means algorithm
\citep{Dhillon2,Dhillon} is still based on Loyd's heuristic \citep{Lloyd}
of grouping points that are closer to a cluster center, now
in feature space. 
We refer the reader to \citep{Filippone} for a survey of clustering
methods.

Although clustering from energy statistics in Euclidean spaces was considered
in \citep{Kgroups}, the precise optimization problem behind this approach
remains elusive, as well as the connection with kernel methods.
The main theoretical contribution of this paper is to fill this gap.
Since the statistical potential energy is minimum when
distributions are equal, the principle behind clustering is to maximize 
the statistical energy,  enforcing probability distributions associated to 
each cluster to be different from one another. We provide a precise 
mathematical formulation to this statement, leading to a quadratically 
constrained quadratic program (QCQP) in the associated RKHS. Our results
immediately establish the connection with kernel methods, showing that this 
QCQP is equivalent to kernel $k$-means optimization problem.

Our main 
algorithmic contribution is to use Hartigan's method \citep{Hartigan} to 
find local solutions of the above mentioned QCQP, which is NP-hard in general.
Hartigan's method was also used in \citep{Kgroups}, without any
connection to kernels. More importantly, its 
advantages
over Lloyd's method was already demonstrated 
in some simple settings
\citep{Telgarsky,Slonin}, but apparently this method did not receive 
the deserved attention. To the 
best of our knowledge, Hartigan's method was not previously 
employed together with kernel methods. 
We provide a full kernel based Hartigan's algorithm for clustering,
where the kernel is fixed by energy statistics. 
We make clear the advantages of this proposal versus 
Lloyd's method, which kernel $k$-means is based upon and will also be used 
to solve our QCQP. We show that both algorithms  have the same
time complexity, but Hartigan's method in kernel spaces offer several
advantages. Furthermore, in the examples considered in this paper, it 
also provides superior performance compared to a spectral clustering.
Our numerical results provide compelling evidence that 
Hartigan's method applied to energy statistics based clustering
is more accurate and robust than 
kernel $k$-means. Moreover, we illustrate the
flexibility of energy clustering,  
showing that it is able to perform accurately on data coming from 
different distributions, contrary to $k$-means and GMM for instance.
The proposed algorithm performs 
closely to $k$-means and GMM on normally distributed data, however,
it is significantly better on other settings. 
Its superiority in high dimensions is striking, being more accurate 
than $k$-means and GMM even on Gaussian settings.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Background}
\label{sec:background}

In this section we introduce the main concepts from (generalized) energy
statistics \citep{Szkely2013,Lyons} and its relation to 
RKHS \citep{Sejdinovic2013}
which form the basis of our work.

Consider random vectors $X_i$ living in an arbitrary space
$\mathcal{X}$ of \emph{negative type}, which means that $\mathcal{X}$
is endowed with a \emph{semimetric} 
$\rho: \mathcal{X}\times\mathcal{X} \to \mathbb{R}$ satisfying
$\sum_{i,j=1}^n c_i c_j \rho(X_i, X_j) \le 0$,
where $c_i \in \mathbb{R}$ and
$\sum_{i=1}^n c_i = 0$. 
Let $X,X' \stackrel{iid}{\sim} P$ and 
$Y,Y' \stackrel{iid}{\sim} Q$, where $P$ and $Q$ are cumulative
distribution functions with finite first moments, and 
$X,X',Y,Y' \in \mathcal{X}$. 
The \emph{generalized energy distance} between $P$ and $Q$ is
given by 
\begin{equation}
\label{eq:energy3}
\Energy(P, Q) \equiv 2 \E \rho(X,Y) - \E \rho(X, X') - \E \rho(Y,Y').
\end{equation} 
This quantity is nonnegative, $\Energy(P,Q) \ge 0$, and 
$\Energy^{1/2}$ is
a metric on the space of distributions. Energy distance provides
a characterization of equality of distributions.

For instance, the \emph{standard
energy distance} \citep{Szkely2013} in Euclidean spaces uses
the semimetric
\begin{equation}
\label{eq:rho_standard}
\rho_\alpha(X,Y) = \| X - Y\|^\alpha,
\end{equation} 
where $0< \alpha \le 2$ and $\| \cdot \|$ is the
Euclidean norm in $\mathcal{X}=\mathbb{R}^D$. In this case $\Energy(P,Q)$
is rotationally invariant. For $0<\alpha<2$ we have
$\Energy(P,Q) = 0$ if and only if $P=Q$. However, for $\alpha=2$
we get $\Energy(P,Q) = 2\| \E(X) - \E(Y) \|^2$, thus
$\Energy(P,Q)=0$ does 
not imply equality of distributions but only equality
of the means.

Consider data $\mathbb{X} = \{ x_1,\dotsc, x_n \}$ 
sampled from $k$ unknown distributions $\{ P_j \}_{j=1}^k$,
with points lying on a space of negative type,
$x_i \in \mathcal{X}$.
Let $\mathbb{X} = \bigcup_{j=1}^k \C_j$ be a disjoint
partition, $\C_i \cap \C_j = \emptyset$.
Each expectation in the generalized energy distance
can be empirically estimated with the aid of the
function
\begin{equation}
\label{eq:g_def}
g (\C_i, \C_j) \equiv 
\dfrac{1}{n_i n_j}
\sum_{x \in \C_i} 
\sum_{y \in \C_j} \rho(x, y) ,
\end{equation}
where $n_i = |\C_i|$ is the number of elements in partition
$\C_i$. 
Define the \emph{within energy dispersion} as
\begin{equation}
\label{eq:within}
W \equiv
\sum_{j=1}^{k} \dfrac{n_j}{2} g(\C_j, \C_j),
\end{equation}
and the \emph{between-sample energy statistic} as
\begin{equation}
\label{eq:between}
S \equiv \hspace{-1em}
\sum_{1 \le  i < j \le k } \hspace{-.5em} \dfrac{n_i n_{j}}{2 n} \left[
2 g(\C_i, \C_j) - 
g(\C_i, \C_i) - 
g(\C_j, \C_j)
\right],
\end{equation}
where $n = \sum_{j=1}^k n_j$.
A given point $x_i$ belongs to partition $\C_j$
if and only if $x_i \sim P_j$. 
The quantity $S$ is
a test statistic for equality of distributions
\citep{Szkely2013}.
When the sample size is large enough, $n\to \infty$,
under the null hypothesis $H_0: P_1=P_2=\dotsm=P_k$ we have that
$S\to 0$, 
and under
the alternative hypothesis $H_1: P_\ell \ne P_j$ for at least two $\ell\ne j$, 
we have that $S \to \infty$.

Let $\HH_\kk$ be a Hilbert space of real-valued functions
over $\mathcal{X}$ with an associated kernel
$\kk : \mathcal{X} \times \mathcal{X} \to 
\mathbb{R}$, which is a symmetric and positive definite function, i.e. 
$\kk(x_i,x_j) = \kk(x_j,x_i)$ and 
$\sum_{i,j=1}^n c_i c_j \kk(x_i, x_j) \ge 0$, or equivalently,
if $G$ is the Gram matrix with entries $G_{ij} = \kk(x_i,x_j)$
then $G = G^\top$ and $v^\top G v \ge 0$ for any $v \in \mathbb{R}^n$.
For every $x \in \mathcal{X}$ there exists 
$h_x \equiv \kk(\cdot,x) \in \Hk$ such that  $\langle h_x, f \rangle = f(x)$
for any function $f \in \Hk$. Thus, 
$\langle h_x, h_y \rangle = \kk(x,y)$.
Conversely \citep{Aronszajn},  
for every symmetric
positive definite function $\kk: \mathcal{X}\times \mathcal{X} \to
\mathbb{R}$ there is a Hilbert space $\Hk$ with reproducing
kernel $\kk$, with a 
\emph{feature map} $\varphi: x \mapsto \kkk_x \in \Hk$ such
that $\langle \varphi(x), \varphi(y) \rangle = \kk(x, y)$.

Define the embedding of a probability measure
$P \mapsto \kkk_P \in \Hk$ through
$\kkk_P \equiv \int \kk( \, \cdot \,, x)  d P(x)$. 
The distance between two probability measures, 
called maximum mean discrepancy (MMD), is thus given by
\begin{equation}
\label{eq:mmd}
\gamma_\kk(P,Q) \equiv \| \kkk_P - \kkk_Q \|_{\Hk},
\end{equation}
which can also be written as \citep{Gretton2012}
\begin{equation}\label{eq:mmd2}
\gamma_\kk^2(P,Q) = \E \kk(X,X') + \E \kk(Y,Y') - 2 \E \kk(X, Y)
\end{equation}
where $X,X' \stackrel{iid}{\sim} P$ and $Y,Y'\stackrel{iid}{\sim} Q$.
The equality between \eqref{eq:mmd} and \eqref{eq:mmd2}
gives $\langle \kkk_P, \kkk_Q \rangle_{\Hk} = \E \, \kk(X, Y)$,
thus in practice we can estimate the inner product between  
embedded distributions by averaging the kernel function over sampled data.

The following important result shows that semimetrics of negative
type and symmetric positive definite kernels are closely related
\citep{Berg1984}. Let $\rho: \mathcal{X} \times \mathcal{X} \to \mathbb{R}$
and $x_0 \in \mathcal{X}$ an arbitrary but fixed point.
Define
\begin{equation}
\label{eq:kernel_semimetric}
\kk(x,y) \equiv 
\tfrac{1}{2} \left[  \rho(x,x_0) + \rho(y,x_0) - \rho(x,y)\right].
\end{equation}
Then, it can be shown that 
$\kk$ is positive definite if and only if $\rho$ is a semimetric
of negative type.
We have a family of kernels, one for each choice of $x_0$. Conversely,
if $\rho$ is a semimetric of negative type and $\kk$ is a kernel in this
family, then 
\begin{equation}
\label{eq:gen_kernel}
\begin{split}
\rho(x,y) &= \kk(x,x) + \kk(y,y) -2\kk(x,y) \\
&=  \| \kkk_x - \kkk_y \|^2_{\Hk}
\end{split}
\end{equation}
and the canonical feature map 
$\varphi: x \mapsto \kkk_x$ is injective \citep{Sejdinovic2013}.
When these conditions are satisfied we say that the kernel $\kk$ 
generates the semimetric $\rho$. 
If two different kernels generate the same $\rho$ they are
said to be equivalent kernels.
Now we can state the equivalence between energy distance and
inner products on RKHS, which is one of the main results of
\citep{Sejdinovic2013}. If $\rho$ is a semimetric
of negative type and $\kk$ a kernel that generates $\rho$, then
replacing \eqref{eq:gen_kernel} into
\eqref{eq:energy3}, and using \eqref{eq:mmd2}, yields
\begin{equation}
\label{eq:Erho}
\begin{split}
\tfrac{1}{2}\Energy(P, Q) &= 
\E \, \kk(X, X') + \E \, \kk(Y, Y') - 2\E \, \kk(X, Y) \\ 
&= \gamma_\kk^2(P,Q) ,
\end{split}
\end{equation}
so we can compute the energy distance using the inner product
of $\Hk$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Clustering via Energy Statistics}
\label{sec:clustering_theory}

This section contains our main theoretical results, where 
we formulate an optimization problem for clustering 
based on energy statistics in the RKHS introduced in the previous section.
The proofs are contained in supplementary material.

Due to the energy test statistic for equality of distributions,
the obvious
criterion for clustering data is to 
maximize $S$ which makes 
each cluster as different
as possible from the other ones.
In other words, given a set of points coming from different probability
distributions, the test statistic $S$ should attain a maximum when 
each point is correctly
classified as belonging to the cluster associated to its probability
distribution.
The following 
straightforward result
shows that maximizing $S$ is, however, equivalent to minimizing
$W$ which has a more convenient form.

\begin{lemma}
\label{th:minimize}
Let $\mathbb{X} = \{x_1,\dotsc,x_n\}$ where each data point
$x_i$ lives in a space $\mathcal{X}$ of negative type. 
For a fixed integer $k$,
the partition
$\mathbb{X} = \bigcup_{j=1}^k \C_j$, where $\C_i \cap C_j = \emptyset$ for
all $i\ne j$, maximizes the between-sample statistic $S$, defined
in equation \eqref{eq:between}, if and only if
\begin{equation}
\label{eq:minimize}
\min_{\C_1,\dotsc,C_k  } W(
\C_1, \dotsc, \C_k),
\end{equation}
where the within energy dispersion $W$ is defined by \eqref{eq:within}.
\end{lemma}

In the Euclidean case, the optimization problem \eqref{eq:minimize} based on
energy statistics was already proposed in \citep{Kgroups}. However, it is
important to note that this is equivalent to maximizing $S$,
which is the test statistic for equality of distributions. In this current
form, the relation with kernels and other clustering methods is obscure.
In the following, we show what is the explicit optimization problem behind 
\eqref{eq:minimize} in the corresponding RKHS, 
establishing the connection with kernel methods.

Assume that the kernel $\kk: \mathcal{X} \times \mathcal{X} \to \mathbb{R}$ 
generates $\rho$.  Define  the Gram matrix $G \in \mathbb{R}^{n\times n}$ as
\begin{equation}
\label{eq:kernel_matrix}
G_{ij} \equiv \kk(x_i,x_j).
\end{equation}
Let $Z \in \{ 0,1 \}^{n\times k}$ be the label matrix, 
with only one nonvanishing entry per row, 
indicating to which cluster (column)
each point (row) belongs to. This matrix satisfies
$Z^\top Z = D$, where  
$D = \diag( n_1,\dotsc, n_k )$  contains
the number of points in each cluster. We also introduce the rescaled
matrix  $Y \equiv Z D^{-1/2}$. In component form they are given by
\begin{equation}
\label{eq:label_matrix}
Z_{ij} \equiv \begin{cases}
1 & \mbox{if $x_i \in \C_j$ } \\
0 & \mbox{otherwise}
\end{cases} \
\Zt_{ij} \equiv \begin{cases}
\tfrac{1}{\sqrt{n_j}} & \mbox{if $x_i \in \C_j$ } \\
0 & \mbox{otherwise}
\end{cases}
\end{equation}
Throughout the paper, we use the notation $M_{i\bullet}$ to denote
the $i$th row of a matrix $M$, and $M_{\bullet j}$ denotes its $j$th column.
Our next result shows that the optimization problem \eqref{eq:minimize}
is NP-hard since
it is a quadratically constrained quadratic program (QCQP) in the RKHS.

\begin{proposition} 
\label{th:qcqp2}
The optimization problem \eqref{eq:minimize} is equivalent to
\begin{equation}
\label{eq:qcqp2}
\begin{split}
&\max_{\Zt} \Tr \left( \Zt^\top G \, \Zt \right) \\
&\mbox{s.t. $\Zt \ge 0$, $\Zt^\top \Zt = I$, 
$\Zt \Zt^\top \e = \e$},
\end{split}
\end{equation}
where $\e = (1,\dots,1)^\top \in \mathbb{R}^n$ is the all-ones vector,
and $G$ is the Gram matrix \eqref{eq:kernel_matrix}.
\end{proposition}

Therefore, to group data
$\mathbb{X} = \{ x_1,\dotsc,x_n \}$
into  $k$ clusters we first compute the Gram matrix
$G$ and then 
solve the optimization problem \eqref{eq:qcqp2} for $\Zt \in
\mathbb{R}^{n\times k}$. The $i$th row
of $\Zt$ will contain a single nonzero element in some $j$th column,
indicating that $x_i \in \C_j$. 
This optimization problem is nonconvex, and also NP-hard,
thus a direct approach 
is computational prohibitive even for small datasets.
However, one can find approximate solutions by relaxing some 
of the constraints.
For instance, the relaxed problem
$\max_{Y} \Tr \left( Y^\top G \, Y \right)$ s.t. $Y^\top Y = I$,
has a well-known closed form solution $Y^\star = U R$, where the
columns of $U \in \mathbb{R}^{n\times k}$ 
contain the top $k$ eigenvectors of $G$ corresponding
to the $k$ largest eigenvalues $\lambda_1\ge \lambda_2\ge\dotsc\ge\lambda_k$, 
and
$R \in \mathbb{R}^{k\times k}$ is an arbitrary orthogonal matrix. 
\emph{Spectral clustering} is based on (variants of) this approach 
and will be compared to the iterative method that will be proposed 
in the following.

Note that the energy clustering problem \eqref{eq:qcqp2} 
is valid for data living in an \emph{arbitrary} space of negative type, where
a semimetric $\rho$, and thus the kernel $\kk$, are
assumed to be known. Standard energy statistics in
Euclidean spaces fixes a family of choices through 
\eqref{eq:rho_standard}.
The same would be valid for data living in more general
spaces $(\mathcal{X}, \rho)$.
In any case, energy  clustering 
is model-free, 
contrary to $k$-means and GMM, for example.
In practice, however,
the clustering quality strongly depend on the choice of a suitable
$\rho$ which measures the similarity between data points.
If prior information is available to choose $\rho$
it can be conveniently incorporated in the 
optimization problem \eqref{eq:qcqp2}.

\paragraph{Relation to Kernel $\bm{k}$-Means.}
One may wonder how energy clustering 
relates to the well-known kernel $k$-means problem%
\footnote{When we refer to kernel $k$-means problem we mean specifically 
the optimization problem \eqref{eq:kernel_kmeans}, which should not be 
confused with kernel $k$-means algorithm that is just one possible recipe 
to solve \eqref{eq:kernel_kmeans}.}, 
extensively used in machine learning.
For a positive semidefinite Gram matrix $G$, as defined in
\eqref{eq:kernel_matrix},
there exists a feature map
$\varphi: \mathcal{X} \to \HH_\kk$ such that
$G_{ij} = \langle \varphi(x_i), \varphi(x_j) \rangle$. The kernel $k$-means optimization
problem,
in feature space,
is defined by
\begin{equation}
\label{eq:kernel_kmeans}
\min_{\C_1,\dotsc,\C_k}\bigg\{ 
\sum_{j=1}^k
\sum_{x \in \C_j} \| \varphi(x) - \varphi(\mu_j) \|^2
\bigg\}
\end{equation}
where $\mu_j = \tfrac{1}{n_j} \sum_{x \in \C_j} x$ is the  mean of cluster
$\C_j$ in the ambient space $\mathcal{X}$. 
Notice that the above objective function
is strongly tied to the idea of minimizing distances between points
and cluster centers, which arises from $k$-means objective function based
on Lloyd's heuristic \citep{Lloyd}.
It is known \citep{Dhillon2,Dhillon}
that kernel $k$-means problem 
can be cast into a trace maximization in the same form as 
\eqref{eq:qcqp2}. The next result makes this explicit.

\begin{proposition}
\label{th:kernel_kmeans}
For a fixed kernel,
the energy clustering optimization problem
\eqref{eq:minimize} 
is equivalent to the kernel $k$-means optimization problem
\eqref{eq:kernel_kmeans}, and both are equivalent to \eqref{eq:qcqp2}.
\end{proposition}

The above result shows that 
kernel $k$-means problem is equivalent to the clustering problem
formulated in the energy statistics framework, when operating on the same
kernel. Notice, however, that 
energy statistics theory is valid for arbitrary semimetric spaces of
negative type, fixing the kernel function in the associated RKHS, which
is guaranteed to be positive definite.
As shown by \citep{Dhillon2,Dhillon}, kernel $k$-means, spectral clustering,
and graph partitioning problems such as ratio association, ratio cut, and
normalized cut are all equivalent to a QCQP of the form \eqref{eq:qcqp2}.
Thus one can view all these problems arising from energy statistics as well.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Hartigan's for Energy Clustering}
\label{sec:algo}

We introduce an iterative algorithm based
on Hartigan's method \citep{Hartigan} to find a local
maximizer of the optimization problem \eqref{eq:qcqp2}. Due to 
Proposition~\ref{th:kernel_kmeans} we can use 
kernel $k$-means algorithm \citep{Dhillon2,Dhillon}, which
will be compared with the proposed algorithm.

We can write the optimization problem \eqref{eq:qcqp2} as
\begin{equation}
\label{eq:maxQ}
\max_{\{ \C_1,\dotsc,\C_k \}} 
\bigg\{ Q = \sum_{j=1}^k \dfrac{Q_j}{n_j}  \bigg\},
\quad Q_j \equiv \sum_{x,y\in\C_j} \kk(x,y),
\end{equation}
where $Q_j$ represents an internal energy cost of cluster $\C_j$, and
$Q$ is the total energy cost where each $Q_j$ 
is weighted by the inverse
of the number of points in $\C_j$. For a data point $x_i$ we denote
its own energy cost
with the entire cluster $\C_\ell$ by
\begin{equation}
\label{eq:costxij}
Q_\ell(x_i) \equiv \sum_{y\in\C_\ell} \kk(x_i, y) = 
G_{i \bullet} \cdot Z_{\bullet \ell},
\end{equation}
where we recall that $G_{i\bullet}$ ($G_{\bullet i}$) denotes
the $i$th row (column) of matrix $G$.

For a given configuration, we consider the maximum change
in the total cost function $Q$ when moving each data point to
another cluster. More specifically, 
suppose $x_i$
is currently assigned to  cluster $\C_j$, yielding
a total cost function denoted by $Q^{(j)}$.
Moving $x_i$ to cluster $\C_\ell$ yields another total cost function
denoted by $Q^{(\ell)}$. We are interested in computing the maximum 
cost change
$\Delta Q^{j\to \ell} (x_i) \equiv Q^{(\ell)} - Q^{(j)}$, for $\ell\ne j$. 
From \eqref{eq:maxQ}, by explicitly writing the costs related to these 
two cluster we obtain
\begin{equation}
\Delta Q^{j\to \ell} (x_i) = \dfrac{Q_\ell^{+}}{n_\ell+1} + 
\dfrac{Q_j^-}{n_j-1} - \dfrac{Q_j}{n_j} - \dfrac{Q_\ell}{n_\ell}
\end{equation}
where $Q^{+}_\ell$ denote the cost of the new $\ell$th cluster
with the point $x_i$ added to it, and $Q^-_j$ is the cost of new 
$j$th cluster with $x_i$ removed from it. Observe that 
$Q_\ell^{+} = Q_\ell + 2 Q_\ell(x_i) + G_{ii}$ and
$Q_j^{-} = Q_j - 2 Q_j(x_i) + G_{ii}$, hence
\begin{equation}
\label{eq:changeQ}
\begin{split}
\Delta Q^{j \to \ell}(x_i)  = 
\dfrac{1}{n_j - 1} & \left[ \dfrac{Q_j}{n_j} - 2 Q_j(x_i) + G_{ii} \right] \\
- \dfrac{1}{n_\ell + 1}&\left[ \dfrac{Q_\ell}{n_\ell} - 2 Q_\ell(x_i) 
- G_{ii} \right].
\end{split}
\end{equation}
Therefore, if $\Delta Q^{j\to \ell}(x_i) > 0$ we get closer to a 
maximum of \eqref{eq:maxQ} by
moving $x_i$ to $\C_\ell$, otherwise we should keep $x_i$ in $\C_j$.

We thus propose the following algorithm.
Start with an initial label matrix $Z=Z_0$, 
then for each point $x_i$, assuming it currently belongs to cluster
$\C_j$, compute the cost of moving it to $\C_\ell$, i.e.
$\Delta Q^{j\to \ell}(x_i)$ for 
$\ell=1,\dots,k$ with $\ell \ne j$. Then, choose
\begin{equation}
j^\star = \argmax_{\ell=1,\dotsc,k \, | \, \ell\ne j} 
\Delta^{j \to \ell}(x_i).
\end{equation}
If $\Delta Q^{j \to j^\star}(x_i) > 0$ move $x_i$ to $\C_{j^\star}$, 
otherwise keep $x_i$ in its original cluster $\C_j$. 
Repeat this process until no new assignments are made.
The entire procedure is explicitly described in Algorithm~\ref{algo}, 
which we call $\mathcal{E}^H$-clustering to emphasize that it is based on
Hartigan's method.
Note that the objective function is
monotonically increasing at each iteration, consequently the algorithm
converges in a finite number of steps.

\begin{algorithm}
\begin{algorithmic}[1]
    \INPUT number of clusters $k$, Gram matrix $G$, 
                initial label matrix $Z \leftarrow Z_0$
    \OUTPUT label matrix $Z$
  \STATE $q \leftarrow (Q_1, \dotsc, Q_k)^\top$ 
            have the energy costs of each cluster, defined in \eqref{eq:maxQ}
  \STATE $n \leftarrow (n_1,\dotsc,n_k)^\top$ have the number of points 
        in each cluster%, obtained from $D=Z^\top Z$
  \REPEAT
    \FOR{ $i=1,\dotsc,n$}
        \STATE let $j$ be such that $x_i \in \C_j$
        \STATE $j^\star \leftarrow \argmax_{\ell=1,\dotsc,k \, | \, \ell\ne j} 
                \Delta Q^{j\to \ell}(x_i)$
        \IF{ $\Delta Q^{j \to j^\star}(x_i) > 0$ }
            \STATE move $x_i$ to $\C_{j^\star}$: $Z_{ij} \leftarrow 0$ and 
            $Z_{ij^\star} \leftarrow 1$
            \STATE update $n$: $n_j \leftarrow n_j - 1$ and
                    $n_{j^\star} \leftarrow n_{j^\star} + 1$
            \STATE update $q$: $q_j \leftarrow q_j - 2Q_j(x_i) + G_{ii}$ and \\
             \hspace{4.5em}$q_{j^\star} \leftarrow q_{j^\star} + 
                            2Q_{j^\star}(x_i)+ G_{ii}$
    %    \ELSE
    %        \STATE Do nothing;
        \ENDIF
    \ENDFOR
  \UNTIL{convergence}
\end{algorithmic}
\caption{\label{algo}
$\mathcal{E}^H$-clustering is Hartigan's method to find local solutions to the optimization problem \eqref{eq:qcqp2}.
}
\end{algorithm}

%Computing  $G$ requires $\OO( D n^2)$ operations, where 
%$D$ is the dimension of $\mathcal{X}$ and $n$ is the data size, however,
%the algorithm assumes $G$ is given. 
%There are efficient methods to compute $G$, specially if it is sparse, 
%but we will not consider this further.
The worst time complexity of $\mathcal{E}^H$-clustering is
$\OO(n k  \max_\ell n_\ell) =\OO(k n^2)$. 
This is the same cost as kernel $k$-means algorithm
\citep{Dhillon2,Dhillon}. 
If $G$ is sparse
this can be further reduced to $\OO(k n')$ where $n'$ is the number of 
nonzero entries.

There are known results about Hartigan's method,
indicating its advantages over Lloyd's method.

\begin{theorem}[\cite{Telgarsky}]
If $n > k$ the resulting partition obtained from Hartigan's method has 
\begin{enumerate*}
\item \label{noempty} no empty clusters, and
\item \label{diffmean} distinct means.
\end{enumerate*}
\end{theorem}

Neither of these two conditions are guaranteed to hold 
for Lloyd's method,
and consequently for (kernel) $k$-means algorithm.

\begin{theorem}[\cite{Telgarsky}]
The set of local optima of Hartigan's method is a (possibly strict) subset
of local optima of Lloyd's method.
\end{theorem}

This means that Hartigan's method can potentially 
escape local optima of Lloyd's method.
Kernel $k$-means cannot
improve on a local optima of $\mathcal{E}^H$-clustering, but on the other hand,
$\mathcal{E}^H$-clustering might improve on a local optima of 
kernel $k$-means. 
Lloyd's method forms Voronoi partitions,
while Hartigan's method groups data
in regions called circlonoi cells.
The circlonoi cells are within
a smaller volume of a Voronoi cell, and this excess volume grows
exponentially with the dimension of $\mathcal{X}$ 
\citep[Theorems 2.4 and 3.1]{Telgarsky}. 
Points in this excess volume
force Hartigan's to iterate, contrary
to Lloyd's. 
Moreover, this improvement should be more prominent as
dimension increases. Also, the improvement grows as the number of clusters
$k$ increases.
The empirical results of \citep{Telgarsky} show that 
an implementation of Hartigan's method has comparable execution time 
to an implementation of
Lloyd's method,
but no explicit complexity was provided. In our case both
$\mathcal{E}^H$-clustering and kernel $k$-means
have the same time complexity. To the best of our knowledge, Hartigan's
method was not previously considered together with kernels, as we 
are proposing here.

In \citep{Slonin}, Hartigan's method was applied to $k$-means problem
with any Bregman divergence. It was shown that the number of Hartigan's
local optima is upper bounded by $\mathcal{O}(1/k)$. 
In addition, examples were provided where
\emph{any} initial partition correspond to a Lloyd's local optima, while  
the number of Hartigan's local optima  is small and 
correspond to true partitions of the data. Empirically, the number of
Hartigan's local optima was considerably smaller.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Numerical Experiments}
\label{sec:numerics}

The main goal of this section is threefold. First, to compare
$\mathcal{E}$-clustering in Euclidean space 
to $k$-means and GMM. 
Second, to compare $\mathcal{E}^H$-clustering  to
kernel $k$-means and also to spectral clustering, when they all operate
on the same kernel.
Third, to show the flexibility
of energy clustering, able to perform 
accurately in different settings using the same kernel.

The following experimental setup holds unless specified
otherwise. 
We consider
$\mathcal{E}$-clustering, kernel $k$-means and spectral clustering
with semimetrics
\begin{equation}
\begin{split}
\rho_{\alpha}(x,y) &= \| x-y \|^{\alpha}, \\
%\kk_{\alpha}(x,y) &= \tfrac{1}{2} \left( \| x \|^{\alpha} + \| y \|^{\alpha} 
%- \| x-y \|^{\alpha} \right) \\
%
\widetilde{\rho}_{\sigma}(x,y) &= 2 - 2 e^{-\tfrac{\|x-y\|}{2 \sigma}},  \\
%\widetilde{\kk}_{\sigma}(x,y) &= e^{-\tfrac{\|x-y\|}{2\sigma}} \\
%
\widehat{\rho}_{\sigma}(x,y) &= 2 - 2 e^{-\tfrac{\|x-y\|^2}{2 \sigma^2}}  
%\widehat{\kk}_{\sigma}(x,y) &= e^{-\tfrac{\|x-y\|^2}{2\sigma^2}}  
\end{split}
\end{equation}
These are generated by the kernel \eqref{eq:kernel_semimetric}
where we fix $x_0=0$.
The standard $\rho_1$, from energy statistics in Euclidean
spaces, will always be present as a reference and it
is implied unless explicitly specified.
For $k$-means, GMM and spectral clustering we use the robust 
implementations of \emph{scikit-learn} library \citep{scikit-learn}, where  
$k$-means is initialized with 
$k$-means++ \citep{Vassilvitskii}, 
and GMM with the output of $k$-means, making it more robust and not
breaking in high dimensions. 
We implemented kernel $k$-means
as described in \citep{Dhillon2,Dhillon}, and $\mathcal{E}^H$-clustering
as described in Algorithm~\ref{algo}. Both
will also be initialized with $k$-means++.
We run the algorithms $5$ times with different initializations, picking
the result with best objective. 
We evaluate clustering quality by
the \emph{accuracy} based on the true labels. For each setting we show 
the average accuracy over $100$ Monte
Carlo trials, with error bars indicating the standard error.

\begin{figure}
\includegraphics[scale=0.41]{normal_highdim_mean.pdf}
\includegraphics[scale=0.41]{normal_highdim_cov.pdf}
\put(-180,-7){(a)}
\put(-60,-7){(b)}
\caption{
\label{fig:gauss}
High dimensional Gaussian mixture.
Comparison of $\mathcal{E}^H$-clustering, kernel $k$-means, spectral
clustering, $k$-means and GMM (last three from scikit-learn).
(a) Parameters as in \eqref{eq:gauss1}. 
(b) Parameters as in \eqref{eq:gauss2}. Dashed line is Bayes accuracy.
We plot mean accuracy versus number of dimensions $D$ over $100$ Monte
Carlo trials, error bars are standard error.
}
\end{figure}

First we analyze
how the algorithms degrade as the number of dimensions increase while
keeping Bayes error fixed.
Consider data from a Gaussian mixture
$x  \stackrel{iid}{\sim} 
\tfrac{1}{2}\left[\mathcal{N}(\mu_1,\Sigma_1) +
\mathcal{N}(\mu_2,\Sigma_2)\right]$ in $\mathbb{R}^D$ with
$\Sigma_1=\Sigma_2 = I_D$,   
\begin{equation}
\label{eq:gauss1}
\mu_1 = (\underbrace{0,\dotsc,0}_{\times D})^\top, \
\mu_2 = 0.7 (\underbrace{1,\dots,1}_{\times 10},
\underbrace{0,\dots,0}_{\times (D-10)})^\top.
\end{equation}
The Bayes error gives an optimal accuracy
$\approx 0.86$, which is fixed as $D$ increases.
We sample $200$ points for each trial.
The results are shown in Fig.~\ref{fig:gauss}a, where
$\mathcal{E}^H$- and spectral-clustering have close
performance,  
much better than kernel $k$-means, and also 
better than 
$k$-means and GMM. The improvement is noticeable in 
higher dimensions.
Still for a two-class Gaussian mixture we now choose
different numbers for the  diagonal covariance $\Sigma_2$.
We have $\Sigma_1=I_D$, $\mu_1=(0,\dotsc,0)^\top \in \mathbb{R}^D$,
$\mu_2=(1,\dotsc,1,0,\dotsc,0)^T \in \mathbb{R}^D$ 
with signal in the first $10$ dimensions, and
\begin{equation}
\label{eq:gauss2}
\begin{split}
\Sigma_2 &= \left( \begin{array}{c|c}
\widetilde{\Sigma}_{10} & 0 \\ \hline 
0 & I_{D-10} \end{array}\right), \\
\widetilde{\Sigma}_{10} &= \diag(1.367,  3.175,  3.247,  4.403,  1.249,\\
&\hspace{3.6em}1.969, 4.035,   4.237,  2.813,  3.637).
\end{split}
\end{equation}
We simply chose $10$ number uniformly at random on the interval
$[1,5]$, and any other choice would give analogous results.
The Bayes  accuracy is fixed at $\approx 0.95$.
Fig.~\ref{fig:gauss}b we show the results. 
GMM performs better in low dimensions, 
but it quickly degenerates
as $D$ increases, as (kernel) $k$-means, while  
$\mathcal{E}^H$- and spectral-clustering remains much more stable.

\begin{figure}[t]
\includegraphics[scale=.41]{normal_kernels.pdf}
\includegraphics[scale=.41]{lognormal_kernels.pdf}\\
\includegraphics[scale=.41]{normal_kernels_difference.pdf}
\includegraphics[scale=.41]{lognormal_kernels_difference.pdf}
\put(-180,117){(a)}
\put(-60,117){(c)}
\put(-180,-10){(b)}
\put(-60,-10){(d)}
\caption{
\label{fig:consist}
(a,b) Normal and (c,d) lognormal distributions in $\mathbb{R}^{20}$ with
parameters \eqref{eq:20gauss}. We use different kernels
for $\mathcal{E}^H$-clustering, compared to 
$k$-means and GMM. Bayes accuracy
is $\approx 0.9$. We plot average accuracy (error bars are standard error)
versus number of points for $100$ Monte Carlo trials.
The plots in (c) and (d) consider the difference in accuracy
between $\mathcal{E}^H$ versus kernel $k$-means and
spectral clustering, with $\widetilde{\rho}_1$.
}
\end{figure}

Now we sample $n$ points from the Gaussian mixture
$x \stackrel{iid}{\sim} \tfrac{1}{2}\left[ \mathcal{N}(\mu_1,\Sigma_1)+
\mathcal{N}(\mu_2, \Sigma_2) \right]$ in $\mathbb{R}^{20}$ with 
$\Sigma_1=\tfrac{1}{2} I_{20}$, $\Sigma_2 = I_{20}$, 
\begin{equation}
\label{eq:20gauss}
\mu_1 = (\underbrace{0,\dotsc,0}_{\times 20})^\top ,\quad
\mu_2 = \tfrac{1}{2} 
(\underbrace{1,\dotsc,1}_{5},\underbrace{0,\dotsc,0}_{15})^\top.
\end{equation}
Bayes accuracy is $\approx 0.90$. 
We increase the sample size in the range $n \in [10, 400]$. 
The results are 
shown in Fig.~\ref{fig:consist}a, where we compare
$\mathcal{E}^H$-clustering with different
kernels, indicated in the legend, to $k$-means and GMM.
%For small $n$ all methods
%are superior than GMM, which eventually catches up and tend to optimal Bayes,
%as expected since it is consistent to this setting. 
Note that $\mathcal{E}^H$-clustering
with $\widetilde{\rho}_1$ is as accurate as GMM for large number of points, 
however, it is superior for small number of points. Still for
the same setting, in Fig.~\ref{fig:consist}b we
we show the difference in accuracy provided by $\mathcal{E}^H$ minus
kernel $k$-means and $\mathcal{E}^H$ minus spectral clustering, when using the
semimetric $\widetilde{\rho}_1$.
Note that $\mathcal{E}^H$ was
always superior (there would be points with negative values on 
the $y$-axis otherwise).
Now consider the same parameters as in \eqref{eq:20gauss} but with
lognormal mixtures in $\mathbb{R}^D$,
$\log x \stackrel{iid}{\sim} \tfrac{1}{2}\left[ \mathcal{N}(\mu_1,\Sigma_1)+
\mathcal{N}(\mu_2, \Sigma_2) \right]$. 
The same experiments are shown in 
Fig.~\ref{fig:consist}c and Fig.~\ref{fig:consist}d.
Note that $\mathcal{E}^H$-clustering still performs accurately, 
with any of those
kernels, providing better results than $k$-means and GMM. 
%The semimetric $\widetilde{\rho}_1$ still provides the 
%best results for small number of points, but its performance is eventually 
%achieved by $\rho_{1/2}$, indicating that $\alpha \approx 1/2$ in the standard
%energy distance should be more appropriate for skewed distributions.
These two experiments
illustrate how energy clustering is flexible,
since it is accurate on data from very different distributions
with the same kernel.

\begin{figure}[t!]
\includegraphics[scale=0.21]{complex_data.pdf}
\put(-210,-10){(a)}
\put(-150,-10){(b)}
\put(-90,-10){(c)}
\put(-35,-10){(d)}
\caption{\label{fig:other}
(a) Parallel cigars, $200$ points each. (b,c) Two and three  
concentric circles with Gaussian noise, $400$ points each.   
(d) MNIST handwritten digits.
Clustering results are in Tables~\ref{table:other} and
\ref{table:mnist}.
}
\end{figure}

\begin{table*}
\caption{\label{table:other}
Clustering data from Fig.~\ref{fig:other}a--c.
}
\begin{center}
%\footnotesize{
\begin{tabular}{@{}r|  l l  l l  l l@{}}
\toprule[1pt]
 & & \emph{Fig.~\ref{fig:other}a}
 & & \emph{Fig.~\ref{fig:other}b}
 & & \emph{Fig.~\ref{fig:other}c} \\
\midrule[0.5pt]
\multirow{4}{*}{\emph{$\mathcal{E}^H$-clustering~~~~}}
& $\rho_{1}$ & $0.705\pm 0.065$
& $\rho_{1}$ & $0.521\pm 0.005$
& $\rho_{1}$ & $0.393\pm 0.020$ \\
& $\rho_{1/2}$ & $0.952\pm 0.048$
& $\rho_{1/2}$ & $0.522\pm 0.004$
& $\rho_{1/2}$ & $0.486\pm 0.040$ \\
& $\widetilde{\rho}_{2}$ & $\bm{0.9987\pm 0.0008}$
& $\widetilde{\rho}_{1}$ & $0.778\pm 0.075$
& $\widetilde{\rho}_{2}$ & $0.666\pm 0.007$ \\
& $\widehat{\rho}_{2}$  & $0.956\pm 0.020$
& $\widehat{\rho}_{1}$  & $\bm{1.0\pm 0.0}$
& $\widehat{\rho}_{2}$ & $\bm{0.676\pm 0.002}$ \\
\midrule[0.5pt]
\emph{spectral-clustering~~~~}
& $\widetilde{\rho}_{2}$ & $0.557\pm 0.014$ 
& $\widehat{\rho}_{1}$ & $0.732\pm 0.002$ 
& $\widehat{\rho}_{2}$ & $0.364\pm 0.004$  \\
\emph{$k$-means}~~~~ 
& \xmark & $0.550\pm 0.011$
& \xmark & $0.522\pm 0.004$
& \xmark & $0.368\pm 0.005$ \\
\emph{GMM}~~~~
& \xmark & $0.903\pm 0.064$
& \xmark & $0.595\pm 0.011$
& \xmark & $0.465\pm 0.030$ \\
\bottomrule[1pt]
\end{tabular}
%}
\end{center}
\end{table*}

\begin{table*}
\caption{\label{table:mnist}
Clustering MNIST data from Fig.~\ref{fig:other}d.
}
\begin{center}
%\footnotesize{
\begin{tabular}{@{}r@{}l@{}|@{}l@{}|@{}l@{}|@{}l@{}|@{}l@{}}
\toprule[1pt]
\emph{Class Subset}~~~~ & & 
$~\{0,1,2\}$ &
$~\{0,1,\dotsc,4\}~$ &
$~\{0,1,\dotsc,6\}~$ &
$~\{0,1,\dotsc,8\}~$ \\
parameter~~~~ & {$\sigma$}
& $~10.34~$
& $~10.41~$
& $~10.41~$
& $~10.37~$ \\
\midrule[0.5pt]
\multirow{3}{*}{\emph{$\mathcal{E}^H$-clustering~~~~}}
& $\rho_{1}$\hspace{1em} 
&$~0.937\pm 0.006~$
&$~0.873\pm 0.025~$
&$~0.731\pm 0.016~$
&$~0.687\pm 0.016~$
\\
& $\rho_{1/2}$ 
&$~0.939\pm 0.006~$
&$~0.874\pm 0.027~$
&$~0.722\pm 0.017~$
&$~0.647\pm 0.017~$
\\
& $\widetilde{\rho}_{\sigma}$ 
&$~0.939\pm 0.006~$
&$~0.847\pm 0.031~$
&$~0.695\pm 0.023~$
&$~0.657\pm 0.014~$
\\
& $\widehat{\rho}_{\sigma}$ 
&$~0.933\pm 0.005~$
&$~\bm{0.891\pm 0.009}~$
&$~\bm{0.759\pm 0.011}~$
&$~\bm{0.704\pm 0.011}~$
\\
\midrule[.5pt]
\emph{spectral-clustering~~~~} 
& $\widehat{\rho}_{\sigma}$ 
&$~0.823\pm 0.015~$
&$~0.769\pm 0.012~$
&$~0.678\pm 0.014~$
&$~0.649\pm 0.018~$
\\
\emph{$k$-means}~~~~ & \xmark
&$~0.927\pm 0.004~$
&$~0.878\pm 0.010~$
&$~0.744\pm 0.008~$
&$~0.695\pm 0.012~$
\\
\emph{GMM}~~~~ & \xmark
&$~\bm{0.952\pm 0.005}~$
&$~0.839\pm 0.015~$
&$~0.694\pm 0.010~$
&$~0.621\pm 0.009~$
\\
\bottomrule[1pt]
\end{tabular}
%}
\end{center}
\end{table*}

In Fig.~\ref{fig:other}a--c we have 
complex two dimensional datasets. 
We apply $\mathcal{E}^H$-clustering  with different kernel choices.
We also consider the best kernel choice for spectral
clustering, besides 
$k$-means and GMM. Here we perform only $10$ Monte Carlo runs.
In (a) we initialize
all algorithms with $k$-means++, and in (b) and (c) we initialize at
random.
The results are in Table~\ref{table:other}.
$\mathcal{E}^H$-clustering has superior performance
in every example, and in particular better than the
spectral clustering.
For the data in Fig.~\ref{fig:other}a 
the semimetrics $\rho_1$ and $\rho_{1/2}$ still provide
accurate results, however, for the
examples in (b) and (c) the kernel choice is more sensitive.

Next, we consider 
the infamous MNIST handwritten digits
as illustrated in Fig.~\ref{fig:other}d.
Each data point is an $8$-bit gray scale
image forming a $784$-dimensional vector 
corresponding to the digits $\{0,1,\dotsc,9 \}$.
We compute the parameter 
\begin{equation}
\label{eq:sigma}
\sigma^2 = \dfrac{1}{n^2} \sum_{i,j=1}^n \| x_i - x_j \|^2 
\end{equation}
from a separate
training set, to be used in the kernels.
We consider subsets of $\{0,1,\dotsc,9 \}$, 
sampling $100$ points 
for each class over $10$ Monte Carlo trials.
The results are shown in Table~\ref{table:mnist}.
Unsupervised clustering on MNIST without any feature extraction
is not trivial. This 
same experiment was performed in \citep{Sapiro} where a low-rank
transformation is learned then subsequently used in subspace clustering. 
It would be interesting to explore methods
for learning a better representation of the data and subsequently apply
$\mathcal{E}^H$-clustering.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}
\label{sec:conclusion}

We considered clustering from the perspective of generalized energy
statistics, valid for arbitrary spaces of negative type.
Our mathematical formulation of energy clustering 
reduces to a QCQP in the associated RKHS, as demonstrated in 
Proposition~\ref{th:qcqp2}.
We showed that the optimization problem
is equivalent
to kernel $k$-means, once the kernel is fixed; see
Proposition~\ref{th:kernel_kmeans}. Energy statistics, however, fixes
a family of standard kernels in Euclidean space, and
more general kernels 
on spaces of negative type can also be obtained.
We proposed the iterative $\mathcal{E}^H$-clustering algorithm based on 
Hartigan's method, which was compared to kernel $k$-means algorithm
based on Lloyd's heuristic.
Both have the same time complexity, however, numerical and theoretical
results provide compelling evidence that $\mathcal{E}^H$-clustering
is more robust with a superior performance, specially in high
dimensions. 
Furthermore, energy clustering, with standard kernels from energy
statistics, outperformed $k$-means and GMM
on several settings, even on Gaussian ones, illustrating the flexibility
of the proposed method. In some
examples the iterative
$\mathcal{E}^H$-clustering also surpassed spectral clustering, and in
others performed similarly but never worse.

A limitation of $\mathcal{E}^H$-clustering  is that it cannot
handle accurately highly unbalanced clusters. 
An interesting problem would be to
extend $\mathcal{E}^H$-clustering to these cases.
Moreover, it would also be interesting to formally 
demonstrate cases where energy clustering is a 
consistent. A soft version of energy clustering is also an
interesting extension.
Finally, kernel methods can benefit from sparsity and
fixed-rank approximations of the Gram matrix, and there is plenty
of room to make $\mathcal{E}^H$-clustering algorithm more scalable.

\subsection*{Acknowledgements}
We would like to thank Carey Priebe for discussions.
This work was supported by NIH TRA grant.

%\bibliographystyle{unsrt}
%\bibliographystyle{plainnat}
\bibliographystyle{abbrvnat}
\bibliography{biblio.bib}

\clearpage

\appendix

\section{Supplementary Material}

Here we colect the proofs of our main results.

\begin{proof}[Proof of Lemma~\ref{th:minimize}]
From \eqref{eq:within} and \eqref{eq:between}
we have
\begin{equation}
\begin{split}
S + W &= 
\dfrac{1}{2n} \sum_{\substack{i,j=1 \\ i\ne j}}^k n_i n_j g(\C_i, \C_j)
\\&+ \dfrac{1}{2n} \sum_{i=1}^{k} 
\bigg[ n - 
\sum_{j\ne i = 1}^k n_j \bigg] 
n_i g(\C_i, \C_i) \\
& = \dfrac{1}{2n} \sum_{i,j=1}^k n_i n_j g(\C_i, \C_j) \\
&= \dfrac{1}{2n} \sum_{x \in \mathbb{X}} \sum_{y \in \mathbb{X}} \rho(x,y)\\
&= \dfrac{n}{2} g(\mathbb{X}, \mathbb{X}).
\end{split}
\end{equation}
Note that the right hand side of this equation 
only depends on the pooled data, so it is a constant
independent of the choice of partition. Therefore, maximizing
$S$ over the choice of partition is equivalent to minimizing $W$.
\end{proof}

\begin{proof}[Proof of Proposition~\ref{th:qcqp2}]
From 
\eqref{eq:gen_kernel},
\eqref{eq:g_def}, and
\eqref{eq:within}
we have
\begin{equation}
\label{eq:W2}
\begin{split}
W
&= \dfrac{1}{2} \sum_{j=1}^k \dfrac{1}{n_j} \sum_{x,y \in \C_j} \rho(x,y) \\
&= \sum_{j=1}^k \sum_{x \in \C_j}  \bigg(
\kk(x,x) - \dfrac{1}{n_j} \sum_{y \in \C_j} \kk(x,y) \bigg).
\end{split}
\end{equation}
Note that the first term is global so it does not contribute to the 
optimization problem.
Therefore, minimizing \eqref{eq:W2} is equivalent to
\begin{equation}
\label{eq:max_prob}
\max_{ \C_1,\dotsc,\C_k } 
\sum_{j=1}^k \dfrac{1}{n_j} \sum_{x,y\in C_j} \kk(x,y) .
\end{equation}
But 
\begin{equation}
\sum_{x, y \in \C_j} \kk(x, y) =
\sum_{p=1}^{n} \sum_{q=1}^{n} Z_{pj} Z_{qj} G_{pq} = 
(Z^\top G \, Z)_{jj},
\end{equation}
where we used the definitions \eqref{eq:kernel_matrix} and
\eqref{eq:label_matrix}. 
Notice that $n_j^{-1} = D^{-1}_{jj}$, where the diagonal matrix $D = 
\diag(n_1,\dotsc,n_k)$ contains the number of points in each cluster, 
thus the objective function in 
\eqref{eq:max_prob} is equal to $\sum_{j=1}^k D^{-1}_{jj} 
\left( Z^\top G Z \right)_{jj} = \Tr \left( D^{-1} Z^\top G Z \right)$. 
Now we can
use the cyclic property
of the trace, and by the  definition of the matrix $Z$
in \eqref{eq:label_matrix}, we obtain the following integer
programing problem:
\begin{equation}\label{eq:qcqp}
\begin{split}
&\max_{Z} \Tr\Big( \big( Z D^{-1/2}\big)^\top G 
\big( ZD^{-1/2} \big) 
\Big) \\
&\mbox{s.t. $Z_{ij} \in \{0,1\}$, $\sum_{j=1}^k Z_{ij} = 1$, 
$\sum_{i=1}^n Z_{ij} = n_j$.}
\end{split}
\end{equation}

Now we write this in terms of the matrix $Y = Z D^{-1/2}$.
The objective function immediately becomes
$\Tr\left( Y^\top G \, Y\right)$. Notice that the above constraints
imply that $Z^T Z = D$, which in turn gives
$D^{-1/2} Y^T Y D^{-1/2} = D$, or $Y^\top Y = I$. 
Also, every entry of $Y$ is positive by definition,
$Y \ge 0$. Now it only remains to show the last 
constraint in \eqref{eq:qcqp2}, which comes from the last
constraint in \eqref{eq:qcqp}. In matrix form this reads
$Z^T \e = D \e$. Replacing $Z=YD^{1/2}$ we have
$Y^\top \e = D^{1/2} \e$. Multiplying this last equation
on the left by $Y$, and noticing
that $Y D^{1/2} \e = Z \e = \e$, we finally obtain
$Y Y^\top \e = \e$. Therefore, the optimization 
problem \eqref{eq:qcqp} is equivalent
to \eqref{eq:qcqp2} .
\end{proof}

\begin{proof}[Proof of Proposition~\ref{th:kernel_kmeans}]
Notice that 
\begin{equation}
\begin{split}
\| \varphi(x) - \varphi(\mu_j) \|^2 &= \langle 
\varphi(x),  \varphi(x) \rangle
- 2 \langle \varphi(x), \varphi(\mu_j)\rangle \\
& \qquad + \langle \varphi(\mu_j), \varphi(\mu_j) \rangle,
\end{split}
\end{equation}
therefore, kernel $k$-means objective function becomes
\begin{equation}
\label{eq:J}
\begin{split}
\sum_{j=1}^k \sum_{x\in\C_j} \bigg(
\kk(x,x) - 
\dfrac{2}{n_j} \sum_{y\in \C_j} \kk(x,y) \\
+ \dfrac{1}{n_j^2}
\sum_{y,z \in \C_j} \kk(y,z) \bigg).
\end{split}
\end{equation}
The first term is global so it does not contribute to the optimization
problem. Notice that the third term gives
$\sum_{x\in\C_j} \tfrac{1}{n_j^2} \sum_{y,z\in\C_j} \kk(y,z) =
\tfrac{1}{n_j}\sum_{y,z\in\C_j} \kk(y,z)$, which is the same as
the second term. Thus, problem
\eqref{eq:kernel_kmeans} is equivalent to
\begin{equation}
\max_{\C_1,\dotsc,\C_k}
\sum_{j=1}^k \dfrac{1}{n_j} \sum_{x,y \in\C_j} \kk(x,y) 
\end{equation}
which is exactly the same as 
\eqref{eq:max_prob} from the energy statistics formulation. Therefore,
once the kernel $\kk$ is fixed, the function 
$W$ given by \eqref{eq:within} is the same
as $J$ in \eqref{eq:kernel_kmeans}.
The remaining of the proof proceeds as 
already shown in the proof of Proposition~\ref{th:qcqp2}, leading to
the optimization problem \eqref{eq:qcqp2}.
\end{proof}

\end{document}

### 2017-10-12 energy_clustering/draft/energy_clustering.tex
aistats18

@@
 Newtonian gravitational potential from classical mechanics, and it provides 
-a non-parametric hypothesis test for equality of distributions. 
+a 
+hypothesis test for equality of distributions. 
 It was further generalized from Euclidean spaces to metric spaces of 
@@
 % @gui: you didn't update the abstract?
+% @jovo: I did, removed some stuff that you didn't like before, now
+% removing the non-parametric word. You think it's not good enough?
 
 \maketitle
-
-% @gui: my experience is that most reviewers absolutely do *NOT* want a ToC, rather, we skip it, and journals often remove it prior to publication.
 
@@
 
-Our work is organized as follows. In section~\ref{sec:background}, we introduce
-the necessary background on energy statistics and RKHS.
-Section~\ref{sec:clustering_theory} contains the main theoretical 
-results of this paper,
-where we consider a clustering theory based on energy statistics leading
-to a QCQP, which is NP-hard in general. We also show the equivalence to 
-kernel $k$-means optimization problem.
-In Section~\ref{sec:weighted}, we generalize these results to a weighted
-version of energy statistics, providing connections to graph
-partitioning problems.
-In Section~\ref{sec:twoclass}, we consider the  simple case of a two-class
-problem in one dimension,
-where we propose an algorithm which requires no random initialization.
-In section~\ref{sec:algo}, we consider Lloyd's and Hartigan's methods
-and propose iterative algorithms to solve the QCQP.
-Section~\ref{sec:numerics} contains some carefully designed numerical
-experiments showing that, in a variety of settings, energy clustering based
-on Hartigan's algorithm
-outperform $k$-means, GMM, kernel $k$-means, and a version of
-spectral clustering.
-Our final conclusions are presented in Section~\ref{sec:conclusion}.
+%Our work is organized as follows. In section~\ref{sec:background}, we introduce
+%the necessary background on energy statistics and RKHS.
+%Section~\ref{sec:clustering_theory} contains the main theoretical 
+%results of this paper,
+%where we consider a clustering theory based on energy statistics leading
+%to a QCQP, which is NP-hard in general. We also show the equivalence to 
+%kernel $k$-means optimization problem.
+%In Section~\ref{sec:weighted}, we generalize these results to a weighted
+%version of energy statistics, providing connections to graph
+%partitioning problems.
+%In Section~\ref{sec:twoclass}, we consider the  simple case of a two-class
+%problem in one dimension,
+%where we propose an algorithm which requires no random initialization.
+%In section~\ref{sec:algo}, we consider Lloyd's and Hartigan's methods
+%and propose iterative algorithms to solve the QCQP.
+%Section~\ref{sec:numerics} contains some carefully designed numerical
+%experiments showing that, in a variety of settings, energy clustering based
+%on Hartigan's algorithm
+%outperform $k$-means, GMM, kernel $k$-means, and a version of
+%spectral clustering.
+%Our final conclusions are presented in Section~\ref{sec:conclusion}.
 
@@
 \label{sec:background}
-
-% @gui: i'm not sure the above is true/meaningful.  nonparametric means implicitly estimating an infinite number of parameters.  without any statistical theory, i'm not quite sure what we can say.
-% @jovo: I stopped saying non-parametric, however, Kolmogorov-Smirnov is also
-% said to be non-parametric. I think non-parametric tests also mean that
-% it does not assume any form of distributions, i.e. it is a distribution-free
-% statistical test.
-% @gui: KS test is nonparametric because it can reject for any distribution, we cannot cluster for any distribution, so we are not.
 
@@
 
-% @gui: consider only using equation numbers for equations that you later refer to? there are so many, it is easy to get lost.
-% @jovo: ok, will do that after we finish everything. There is a package that
-% automatically remove numbers from unreferenced equations. It is a common
-% practice however to number every equation in the paper,
-% specially for reviewing purposes. But later 
-% I will remove the numbers from the not cited ones if you don't like
 % @gui: maybe common in math, but certainly not in stats / data science. 
 % in any case, what is the package that does that?
-
-% @gui: do other people know this result? (about min W = max X) if so, cite them, if not, state that.
-% @jovo: eq (15), i.e. S + W = T is well-known. Gabor and Rizzo wrote that
-% in their paper, however, they didn't write a proof. They should have
-% it in some other paper.  However, they mention 
-% that this holds only for 0 < alpha < 2, with |x-y|^alpha, while
-% here I did for a general semimetric. Thus I don't really know if in general
-% this is equality is known or not, but it should because it's a very easy
-% derivation. The PhD thesis do not 
-% mention anywhere that minimizing W is equivalent to maximizing S.
-% Therefore, I don't want to sound like this is a big result and possibly
-% a referee can say  "well this is trivial and known".
-% Thus I just preferred to do my own derivation and not say anything.
-% What do you think??
-% @gui: i'd state it as a lemma then.
+% @jovo: \usepackage{mathtools} \mathtoolsset{showonlyrefs}
 
 % @gui: i'm not sure about this.  who is to say the correct direction of causality? maybe energy clustering is a special case of kernel k-means?
-% @jovo: I tried to not commit to a single view. We show they are
-% equivalent, when opperating on the same kernel. However, kernel k-means per
-% se does not consider anything about the kernel choice. Energy statistics,
-% however, provides a theory valid on spaces of negative type, which guarantees
-% that a kernel constructed within this formalism will be 
-% positive-semidefinite. Thus, I still prefer to look in the first direction.
-% @gui: ok, did you update the text? i don't recall where i made this comment and don't want to search for it again. can you point me to the place that you say this?
+% @jovo: I'll put a comment below, I simply removed the previous phrase,
+% didn't say anything instead.
 
@@
 
-\begin{proposition}
+\begin{lemma}
 \label{th:minimize}
@@
 where the within energy dispersion $W$ is defined by \eqref{eq:within}.
-\end{proposition}
+\end{lemma}
 \begin{proof}
@@
 where we used the definitions \eqref{eq:kernel_matrix} and
-\eqref{eq:label_matrix}. Thus, the objective function in 
-\eqref{eq:max_prob} is equal to
-$\Tr \left( D^{-1} Z^\top G Z \right)$. 
-% @gui: the above step is not obvious
-% @jovo: It is obvious. 
-% From (21) we have \sum_j 1/n_j (Z^T G Z)_jj. But 1/n_j is just
-% D^{-1}_jj (the diagonal matrix). The sum over j goes over the diagonals of
-% D^{-1} Z^T G Z which is the definition of the trace
-% @gui: obvious to you \neq obvious.  i understood, i want you to explain it.
+\eqref{eq:label_matrix}. 
+Notice that $n_j^{-1} = D^{-1}_{jj}$, where the diagonal matrix $D = 
+\diag(n_1,\dotsc,n_k)$ contains the number of points in each cluster, 
+thus the objective function in 
+\eqref{eq:max_prob} is equal to $\sum_{j=1}^k D^{-1}_{jj} 
+\left( Z^\top G Z \right)_{jj} = \Tr \left( D^{-1} Z^\top G Z \right)$. 
 Now we can
@@
 $\Tr\left( Y^\top G \, Y\right)$. Notice that the above constraints
-imply that $Z^T Z = D$, where $D=\diag(n_1,\dotsc,n_k)$, which in turn gives
+imply that $Z^T Z = D$, which in turn gives
 $D^{-1/2} Y^T Y D^{-1/2} = D$, or $Y^\top Y = I$. 
@@
 
+% @jovo: the previous paragraph saying that kernel k-means is a consequence
+% of energy statistics WAS here, I removed and now don't say anything.
 The above result shows that 
@@
 
-% @gui: "does not require any initiializaton" i think you mean is guaranteed to find the optimal solution in O(n^2) time.
-% @jovo: it does not require initialization in the sense that we do not
-% assign points to clusters as an initial step, as k-means++ for instance, or
-% random initialization. This algorithm is "exact" since nothing
-% is random in it, and multiple runs will always give the same answer
-% @gui: i know what you mean, my point is that it is unclear.
-state in the manuscript what you mean: "it is a deterministic algorithm." or state something else, that it does not require an initialization i don't feel is correct or informative.
-
 % @gui: to provide intuition to the reader, let's show the kernel density estimate for each of the 3 algorithms for both classes.
@@
 $\mathcal{E}^{1D}$-clustering.
-Note that this algorithm is exact in the sense that it does not involve
-any random initialization,
+Note that this algorithm is deterministic,
 however,
@@
 where error bars are standard error. The dashed line indicates 
-Bayes accuracy. (a) Data distributed as \eqref{eq:two_normal}, where the
+Bayes accuracy. (a) Data normally distributed as in 
+\eqref{eq:two_normal}, where the
 optimal accuracy is $\approx 0.956$.
-(b) Data from \eqref{eq:two_lognormal} where the optimal
-accuracy is $\approx 0.852$.
+(b) Data log-normally distributed as in \eqref{eq:two_lognormal} where 
+the optimal accuracy is $\approx 0.852$.
+}
 % @gui: if you insist on using the equations, at least also mention something semantic about them, ie, Data distributed Gaussian (as in 46), or Data distributed log Gaussian (as in 47).
 figures & captions should stand on their own without requiring referring to the paper.
-}
 % @gui: why did you take away the plot of the distribution? those are incredibly helpful.
+% @jovo: ok will restore them.
 \end{figure}
@@
 
-% @jovo: check this paragraph. I included it to answer your question
-% about having K(x,x) and \ell=1,...,k without exluding \ell=j. Essentially,
-% the term K(x,x) will be hidden inside the mean of the cluster to which
-% x is currently assigned to. This also clarifies the connection to 
-% the standard formulation of kernel k-means.
 $\mathcal{E}^{L}$-clustering is precisely
 kernel $k$-means algorithm \cite{Dhillon2,Dhillon} 
-but written more concisely and with the kernel induced by energy statistics. Indeed, recalling that
+but written more concisely and with the kernel induced by energy statistics. 
+Indeed, recalling that
 $\kk(x,y)=\langle \varphi(x), \varphi(y) \rangle$ where $\varphi: \mathcal{X}
@@
 $\mathcal{E}^{L}$-clustering is Lloyd's method for energy clustering, which
-is precisely~~~~~~~~ \\ 
-kernel $k$-means algorithm, with the kernel induced by energy statistics. This procedure finds
+is precisely~~~~~~~~~ \\
+kernel $k$-means algorithm, with the kernel induced by energy statistics. 
+This procedure~~~~~~~ \\
+finds
 local solutions to the problem \eqref{eq:qcqp2}.\hspace{\fill}
-% @gui: spacing is weird.
 }
@@
 \subsection*{Hartigan's Method for Energy Clustering}
-
-% @jovo: I also clarified this section, specially the derivation
-% of the algorithm. Actually there was a mistake in the previous formula,
-% which carried over to my code. Now it's fixed and better.
 
@@
 moving $x_i$ to $\C_\ell$, otherwise we keep $x_i$ in $\C_j$.
-
-We thus propose an algorithm that goes as follows.
-We start with an initial configuration for the label matrix $Z$, 
-then for each
-point $x_i$ 
-we compute the cost of moving it to another cluster $\C_\ell$, i.e.
-$\Delta Q^{j\to \ell}(x_i)$ for 
-$\ell=1,\dots,k$ with $\ell \ne j$, where $j$ denotes the index of its current
-partition, $x \in \C_j$. Hence, we choose
-\begin{equation}
-j^\star = \argmax_{\ell=1,\dotsc,k \, | \, \ell\ne j} 
-\Delta^{j \to \ell}(x_i).
-\end{equation}
-If $\Delta Q^{j \to j^\star}(x_i) > 0$ 
-we move $x_i$ to cluster $\C_{j^\star}$, otherwise 
-we keep $x_i$ in its original cluster $\C_j$. 
-This process is repeated
-until no points are assigned to new clusters. 
-The entire procedure is explicitly described in Algorithm~\ref{algo}, which we
-denote $\mathcal{E}^H$-clustering to emphasize that it is based on
-Hartigan's method.
-This method automatically ensures that the objective function is
-monotonically increasing at each iteration, and consequently the algorithm
-converges in a finite number of steps.
 
@@
 This algorithm~~~~~~ \\ finds local solutions to  
-the optimization problem \eqref{eq:qcqp2}.\hspace{\fill}
-% @gui: clarify precisely which steps are different from alg 3 in caption.
-\hspace{\fill}
+the optimization problem \eqref{eq:qcqp2}.
+The steps $6$ and $10$ are different~~~~~~~~ \\
+than $\mathcal{E}^L$-clustering
+described in Algorithm~\ref{kmeans_algo}. \hspace{\fill}
 }
@@
 \end{figure}
+
+We thus propose an algorithm that goes as follows.
+We start with an initial configuration for the label matrix $Z$, 
+then for each
+point $x_i$ 
+we compute the cost of moving it to another cluster $\C_\ell$, i.e.
+$\Delta Q^{j\to \ell}(x_i)$ for 
+$\ell=1,\dots,k$ with $\ell \ne j$, where $j$ denotes the index of its current
+partition, $x \in \C_j$. Hence, we choose
+\begin{equation}
+j^\star = \argmax_{\ell=1,\dotsc,k \, | \, \ell\ne j} 
+\Delta^{j \to \ell}(x_i).
+\end{equation}
+If $\Delta Q^{j \to j^\star}(x_i) > 0$ 
+we move $x_i$ to cluster $\C_{j^\star}$, otherwise 
+we keep $x_i$ in its original cluster $\C_j$. 
+This process is repeated
+until no points are assigned to new clusters. 
+The entire procedure is explicitly described in Algorithm~\ref{algo}, which we
+denote $\mathcal{E}^H$-clustering to emphasize that it is based on
+Hartigan's method.
+This method automatically ensures that the objective function is
+monotonically increasing at each iteration, and consequently the algorithm
+converges in a finite number of steps.
 
@@
 The cost of computing 
-%\eqref{eq:costxij} 
 $Q_\ell(x_i)$
@@
 
-% @gui: something is weird with the way you care computing accuracy.  GMM should not be significantly worse than chance.  i think you are probably counting the NaN's as *wrong*, rather than nothing?
-% now I'm using GMM from sklearn which eliminated previous weird things.
-% I think my method was too crude and the covariance estimate was too bad in
-% some cases
-
-% @gui: i'm not convinced you need alg 2 in these tables at all.
-if i were to add alg 2 to these tables, however, i would do the same thing i would for the figs, that is, for each simulation, compute the diff between performance on alg 2 and alg 3, and report that.
-% @jovo: yes I agree. I included new plots in the fig, but didn't do anything
-% for the tables cuz I don't think necessary
-
-% @jovo: added this paragraph
 The main goal of this section is threefold. First, we want to compare
@@
 kernel \eqref{eq:standard_metric}, as well as the spectral
-$\mathcal{E}^S$-clustering method\footnote{We
-compared our implementation of spectral clustering provided in
-Algorithm~\ref{spectralalgo} with the spectral clustering routine provided
-by \emph{scikit-learn} library \cite{scikit-learn}, 
-which is based on \cite{Malik}. In the few
-cases considered the algorithms had similar performance.
+$\mathcal{E}^S$-clustering method,
 % @gui: dude, this is a crazy sentence to write. use sklearn, or have a *very* good reason not to.  
 % currently, you utterly fail to justify the difference.
-}, 
+% @jovo: Removed the sentence. I have a very good reason to use E^S clustering
+% instead of the spectral clustering from sklearn library. The reason is that
+% E^S clustering is fundamentally different, being an exact solution to
+% a relaxed problem, while 
+% spectral clustering 
+% from sklearn library is, apparently, Shi-Malik's method which is
+% not directly related to the original problem.
+% We have two options. (1) We keep these plots using E^S clustering proposed 
+% before. (2) We remove E^S clustering 
+% from the paper altogether, and if you want to compare with spectral 
+% clustering at all we use sklearn library. However this will be totally 
+% disconnected from the theory. 
 and we initialize the algorithms
@@
 in particular, which matrix are you doing an eigendecomposition on, and how you are choosing the number of dimensions to embed into.
+% @jovo: what about Algorithm 1? All your points are clearly defined there.
+% I don't think it's a good idea to initialize  E-H and E-L with E-S 
+% because of Theorem 6. If you do this initialization you will not see much
+% difference between E^L and E^H, missing the opportunity to illustrate a good
+% selling point of this paper. What we can do, is to do a separate experiment
+% comparing different initializations and see how the algorithms perform.
 For $k$-means and GMM we use the implementation of \emph{scikit-learn} library
@@
 high dimensions.
-Note that in this setting $k$-means and GMM are consistent estimators,
-however, energy clustering degrades much less as dimension increases.
+%Note that in this setting $k$-means and GMM are consistent estimators,
+%however, energy clustering degrades much less as dimension increases.
+Note that while $k$-means and GMM are appropriate to this type of 
+dataset, energy clustering degrades much less as dimension increases.
 % @gui: k-means is only consistent if the prob of each class is equal.
@@
 clarify.
-
+% @jovo: changed the phrase above. I could re-do all experiments again
+% fliping a 1/2 coin for each distribution, but I don't think it's worthy since
+% nothing will change and will just take extra time. So I just rephrased above.
 \begin{figure}
@@
 % @gui: for eq 59, E-H & E-S yield the same performance.  but for eq 63, E-H is much better than E-S.  do you have any idea why that is?
+% @jovo: the difference is not increadibly large, but it is a litlle. I don't
+% know why. And in Fig.3b this difference is very small, also don't know why
 We now allow the diagonal covariance of the second Gaussian distribution to
@@
 % @gui: if there is nothing special, why not just use 2*ones?  this seems suspicious.  also, it is the result of sampling 1 time, not 10 times.
+% @jovo: This is actually less suspicious since  I'm not even choosing the 
+% numbers, just picking them at random. The point is to make the diagonal 
+% entries different, so 2*ones does not help. Also, I don't want
+% to keep trying many experiments everytime, it's super time consuming.
+% This setting works to illustrate what we want ... a setting where k-means
+% does poorly than GMM, and energy still beats GMM in high D.
 The Bayes error is kept fixed yielding
@@
 % @gui: bayes error changes for each random sample, unless you are integrating over all possible samples?  it is not fixed.
+% @jovo: I said explicitly "the fixed setting". 
+% It is exactly as in eq. (59) with those numbers 
+% fixed everytime, this is why I wrote the numbers explicitly. 
+% I'm not generating new numbers everytime I sample.
 In Fig.~\ref{fig:gauss}b we show the results of energy clustering,
@@
 % @gui: you gotta explain why GMM doesn't utterly fail when D>n!
-
- % @gui: maybe put this below section in the discussion?
-Consider the effect of having 
-unbalanced clusters according to
-\begin{equation}
-\label{eq:gauss3}
-\begin{split}
-x &\in \C_i \sim  
-\dfrac{n_i}{N} \mathcal{N}(\mu_i,\Sigma_i) \quad (i=1,2), \quad 
-\mu_1 = (0,0,0,0)^\top , \quad
-\mu_2 = 1.5\times (1,1,0,0)^\top, \\
-\Sigma_1 &= I_4, \quad
-\Sigma_2 = \left( 
-\begin{array}{c|c} 
-\tfrac{1}{2} I_2 & 0  \\ \hline
-0 & I_2 
-\end{array}\right), \quad
-n_1 = N - m, \quad  n_2 = N + m, \quad N=250.
-\end{split}
-\end{equation}
-We then increase $m \in [0,240]$ making
-the clusters progressively more unbalanced.
-We plot the average accuracy over
-$100$ Monte Carlo runs for each $m$, with error bars indicating 
-standard error. 
-The results are shown in Fig.~\ref{fig:unbalanced}.
-For highly unbalanced clusters we see that GMM performs better than
-the other methods, which have basically similar performance.
-Based on this experiment, an interesting problem would be to
-extend $\mathcal{E}^H$-clustering algorithm to account for 
-highly unbalanced clusters.
-
-\begin{figure}
-\includegraphics[width=0.5\textwidth]{normal_unbalanced.pdf}\vspace{-1em}
-\caption{
-\label{fig:unbalanced}
-Comparison of energy clustering algorithms to $k$-means and GMM on
-unbalanced clusters. The data is distributed as \eqref{eq:gauss3}, where
-we vary $m \in [0, 240]$, and in each case we do $100$ Monte Carlo runs
-showing the average accuracy with standard error.
-}
-\end{figure}
+% @jovo: check below
+Notice that a naive implementation of GMM should not be able to 
+estimate the covariances when $D \gtrsim 100$, however, scikit-learn library
+uses $k$-means output as initialization, and a more refined estimation
+of the covariances, therefore the
+output of GMM in this implementation is at least as good as $k$-means, making
+the algorithm more robust in high dimensions.
 
@@
 The two experiments of Fig.~\ref{fig:consist}
-illustrate how energy statistics clustering does not make assumptions
-about the distribution of the data, performing well in different settings
+illustrate how energy clustering is more flexible,
+performing well in different settings
 with the same kernel. 
 % @gui: algs don't make assumptions.
+% @jovo: see if above is ok, just said flexible instead
 
@@
 The results are in Table~\ref{table:other}.
-%, where the respective 
-%choice of parameters for the kernels are indicated. 
 For (a) we initialize
@@
 % also, why don't you use a better metric/similarity designed for images? like using pairwise alignment followed by rho?
+% @jovo: We can't claim we do worse on {0,1,2}, but also don't do better.
+% I did the same experiment as Sapiro, and didn't fit in the table this is
+% why I didn't include {0,...,9}, but I can do that.
+% To treat the images better I would need to do some pre-processing, which
+% can be time consuming, and you know how much I love manipulating images.
+% I don't have any robust code for this, and anything I try to
+% implement will take more than a week to do. I will check if I find something
+% that works out of the box in python though.
 The results are shown in Table~\ref{table:mnist}, where kernels
@@
 We considered clustering from the perspective of energy
-statistics, which provides a non-parametric test for 
-equality of distributions.
+statistics theory.
 The mathematical formulation of energy clustering 
@@
 
-It would be interesting to formally 
+A limitation of the proposed methods for energy clustering is that it cannot
+handle accurately highly unbalanced clusters. As an illustration
+consider the following setting:
+\begin{equation}
+\label{eq:gauss3}
+\begin{split}
+x &\in \C_i \sim  
+\dfrac{n_i}{N} \mathcal{N}(\mu_i,\Sigma_i) \quad (i=1,2), \quad 
+\mu_1 = (0,0,0,0)^\top , \quad
+\mu_2 = 1.5\times (1,1,0,0)^\top, \\
+\Sigma_1 &= I_4, \quad
+\Sigma_2 = \left( 
+\begin{array}{c|c} 
+\tfrac{1}{2} I_2 & 0  \\ \hline
+0 & I_2 
+\end{array}\right), \quad
+n_1 = N - m, \quad  n_2 = N + m, \quad N=250.
+\end{split}
+\end{equation}
+We then increase $m \in [0,240]$ making
+the clusters progressively more unbalanced.
+We plot the average accuracy over
+$100$ Monte Carlo runs for each $m$, with error bars indicating 
+standard error. 
+The results are shown in Fig.~\ref{fig:unbalanced}.
+For highly unbalanced clusters we see that GMM performs better than
+the other methods, which have basically similar performance.
+Based on this experiment, an interesting problem would be to
+extend $\mathcal{E}^H$-clustering algorithm to account for 
+highly unbalanced clusters.
+
+\begin{figure}
+\includegraphics[width=0.5\textwidth]{normal_unbalanced.pdf}\vspace{-1em}
+\caption{
+\label{fig:unbalanced}
+Comparison of energy clustering algorithms to $k$-means and GMM on
+unbalanced clusters. The data is normally 
+distributed as \eqref{eq:gauss3}, where
+we vary $m \in [0, 240]$, and in each case we do $100$ Monte Carlo runs
+showing the average accuracy with standard error.
+}
+\end{figure}
+
+Moreover, it would also be interesting to formally 
 demonstrate cases where energy clustering is a 
@@
 
-% @gui: your discussion is basically just a summary, which i don't find very helpful to the reader.
-if you want to summarize results, 1 paragraph is sufficient.
-another paragraph connects it to previous work not already mentioned, and a third paragraph on additional possible extensions.
-as you know, one thing i'd like to do next is to theoretically prove when Energy Clustering asymptoticaly gets the right answer.
-you also mentioned soft clustering, which is another potential extension, or incorporating this into subspace clustering.
-% @jovo: not clear how to make the conclusion?
-% how do you think I should modify? A summary is important, because it
-% condenses the results of the paper in one place in a non-technical manner.
-% Besides a summary I don't know what else to say
-% I modified above but check if still not good enough
 % @gui: better, but still important to have another paragraph between those connecting it to the existing literature.  compare this work with 3-4 other works explicitly stating the difference.
@@
 \begin{acknowledgements}
-\vspace{-1em}
 We would like to thank Carey Priebe for discussions.

### 2017-10-12 energy_clustering/code/customize_plots.py

@@ -6,22 +6,22 @@ import numpy as np
 pylab.rc('lines', linewidth=.5, antialiased=True, markersize=4,
             markeredgewidth=0.00)
 pylab.rc('font', family='computer modern roman', style='normal',
-         weight='normal', serif='computer modern sans serif', size=10)
+         weight='normal', serif='computer modern sans serif', size=14)
 pylab.rc('text', usetex=True)
 pylab.rc('text.latex', preamble=[
         '\usepackage{amsmath,amsfonts,amssymb,relsize,cancel}'])
-pylab.rc('axes', linewidth=0.5, labelsize=14)
-pylab.rc('xtick', labelsize=9)
-pylab.rc('ytick', labelsize=9)
+pylab.rc('axes', linewidth=0.5, labelsize=20)
+pylab.rc('xtick', labelsize=12)
+pylab.rc('ytick', labelsize=12)
 #pylab.rc('legend', numpoints=1, fontsize=10, handlelength=0.5)
-pylab.rc('legend', numpoints=1, fontsize=10)
+pylab.rc('legend', numpoints=1, fontsize=16)
 fig_width_pt = 455.0 / 1.5 # take this from LaTeX \textwidth in points
 #fig_width_pt = 430.0 / 3 # take this from LaTeX \textwidth in points
 inches_per_pt = 1.0/72.27
 golden_mean = (np.sqrt(5.0)-1.0)/2.0
 fig_width = fig_width_pt*inches_per_pt
-#fig_height = fig_width
-fig_height = fig_width*golden_mean
+fig_height = fig_width
+#fig_height = fig_width*golden_mean
 #fig_height = fig_width
 pylab.rc('figure', figsize=(fig_width, fig_height))
 

### 2017-10-12 energy_clustering/code/energy/bayes.py

@@ -18,12 +18,14 @@ import data
 
 
 def bayes_univariate_normal(mu1, mu2, sigma1, sigma2, 
-                            num_times=10, num_points=1000):
+                            num_times=10, num_points=1000, p=0.5):
     """Estimate Bayes' error for two-class normal distributions in 1D."""
     error = np.zeros(num_times)
     for i in range(num_times):
-        x1 = np.random.normal(mu1, sigma1, num_points)
-        x2 = np.random.normal(mu2, sigma2, num_points)
+        q = 1-p
+        n1, n2 = np.random.multinomial(num_points, [p,q])
+        x1 = np.random.normal(mu1, sigma1, n1)
+        x2 = np.random.normal(mu2, sigma2, n2)
     
         error1 = sum([1 for x in x1 
             if stats.norm.pdf(x,mu1,sigma1)-stats.norm.pdf(x,mu2,sigma2)<=0
@@ -31,10 +33,10 @@ def bayes_univariate_normal(mu1, mu2, sigma1, sigma2,
         error2 = sum([1 for x in x2 
             if stats.norm.pdf(x,mu1,sigma1)-stats.norm.pdf(x,mu2,sigma2)>=0
         ])
-        error[i] = error1+error2
-    return 1-error.mean()/(2*num_points)
+        error[i] = p*error1/n1+q*error2/n2
+    return 1-error.mean()
 
-def integrate_univariate_normal(mu1, mu2, sigma1, sigma2):
+def integrate_univariate_normal(mu1, mu2, sigma1, sigma2, p=0.5):
     """Same as above but through numerical integration."""
     
     g1 = partial(stats.norm.pdf, loc=mu1, scale=sigma1)
@@ -58,9 +60,9 @@ def integrate_univariate_normal(mu1, mu2, sigma1, sigma2):
     #    {'type': 'ineq', 'fun': lambda x: -(x-2)},
     #)
     #x0 = minimize(lambda x: np.abs(g1(x)-g2(x)), x0=1, constraints=cons).x[0]
-    
-    e1 = integrate.quad(lambda x: 0.5*g1(x), x0, np.inf)[0] 
-    e2 = integrate.quad(lambda x: 0.5*g2(x), -np.inf, x0)[0]
+    q = 1-0.5
+    e1 = integrate.quad(lambda x: p*g1(x), x0, np.inf)[0] 
+    e2 = integrate.quad(lambda x: q*g2(x), -np.inf, x0)[0]
     accuracy = 1 - (e1+e2)
     return accuracy
 
[…]

### 2017-10-12 energy_clustering/code/plot.py

@@ -69,11 +69,11 @@ class ErrorBar:
                 table[np.where(self.col0==x)[0],i]) for x in self.xs])
             if self.bayes:
                 ax.plot(self.xs2, self.bayes, '--', color='black', 
-                        linewidth=1, zorder=0)
+                        linewidth=1.5, zorder=0)
             ax.errorbar(self.xs2, mean_, yerr=stderr_, 
-                        linestyle=t, marker=m, color=c, markersize=4, 
-                        elinewidth=.5,  capthick=0.4, label=l, 
-                        linewidth=1, barsabove=False)
+                        linestyle=t, marker=m, color=c, markersize=5, 
+                        elinewidth=.5,  capthick=0.5, label=l, 
+                        linewidth=1.5, barsabove=False)
         
         ax.set_xlabel(self.xlabel)
         ax.set_ylabel(self.ylabel)

### 2017-10-13 energy_clustering/aistats18/energy_clustering.tex
energy to aistats

@@
 \documentclass[twoside]{article}
-%\usepackage{aistats2018}
+\usepackage{aistats2018}
  % If your paper is accepted, change the options for the package
@@
 %
-\usepackage[accepted]{aistats2018}
+%\usepackage[accepted]{aistats2018}
 %
@@
 
-\aistatsauthor{Guilherme Fran\c{c}a \And  Joshua T. Vogelstein}
-
-\aistatsaddress{ Johns Hopkins University } ]
+\aistatsauthor{ Author 1 \And Author 2 \And  Author 3 }
+
+\aistatsaddress{ Institution 1 \And  Institution 2 \And Institution 3 }
+
+%\aistatsauthor{Guilherme Fran\c{c}a \And  Joshua T. Vogelstein}
+
+%\aistatsaddress{ Johns Hopkins University }
+
+]
 
@@
 potential in physics. We propose a principled approach to the
-clustering problem based on energy statistics theory.
+\mbox{clustering} problem based on energy statistics theory.
 Our mathematical formulation establishes connection to kernel methods,
-leading to a quadratically constrained quadratic program (QCQP).
-To obtain local solutions of such NP-hard QCQP 
+leading to a quadratically constrained quadratic program.
+To obtain local solutions of such an optimization problem, which is NP-hard, 
 we introduce an iterative
@@
 on Hartigan's method, which has the same computational cost but offers
-several advantages compared to  
-kernel $k$-means, based on Lloyd's heuristic.
+several advantages compared to the 
+well-known kernel $k$-means, based on Lloyd's heuristic.
 We provide carefully designed numerical experiments showing
@@
 standard $k$-means and Gaussian mixture models in a variety of settings,
-specially in high dimensions. 
+specially in high \mbox{dimensions}. 
 \end{abstract}
@@
 provides a hypothesis test for equality of 
-distributions which is achieved 
-under minimum statistical potential energy. 
+distributions, which is achieved 
+under the minimum of the statistical potential energy. 
 When probability distributions are different the 
@@
 by extending Ward's method of minimum variance. Moreover, an application of 
-energy statistics to clustering, in Euclidean spaces, was 
-proposed \citep{Kgroups}, which in part motivated this paper. 
+energy statistics to clustering in Euclidean spaces was 
+considered \citep{Kgroups}.  
 We refer to \citep{Szkely2013} for an overview.
@@
 constrained quadratic program (QCQP) in the associated RKHS. Our results
-immediately establish the connection with kernel methods, showing that this 
+immediately establish the connection between energy statistics based
+clustering and kernel methods. Moreover, we show that that this 
 QCQP is equivalent to kernel $k$-means optimization problem.
@@
 find local solutions of the above mentioned QCQP, which is NP-hard in general.
-Hartigan's method was also used in \citep{Kgroups}, without any
-connection to kernels. More importantly, its 
-advantages
+Hartigan's method was also used in \citep{Kgroups}, but without any
+connection to kernels. More importantly, the 
+advantages of Hartigan's
 over Lloyd's method was already demonstrated 
@@
 employed together with kernel methods. 
-We provide a full kernel based Hartigan's algorithm for clustering,
+We provide a fully kernel based Hartigan's algorithm for clustering,
 where the kernel is fixed by energy statistics. 
@@
 to solve our QCQP. We show that both algorithms  have the same
-time complexity, but Hartigan's method in kernel spaces offer several
-advantages. Furthermore, in the examples considered in this paper, it 
-also provides superior performance compared to a spectral clustering.
+time complexity, however, Hartigan's method in kernel spaces is superior. 
+Furthermore, in the examples considered in this paper, it 
+also provides equal or better performance than spectral clustering,
+which in fact solves a relaxed version of our QCQP.
 Our numerical results provide compelling evidence that 
@@
 showing that it is able to perform accurately on data coming from 
-different distributions, contrary to $k$-means and GMM for instance.
-The proposed algorithm performs 
+very different distributions, contrary to $k$-means and GMM for instance.
+More specifically, the proposed method performs 
 closely to $k$-means and GMM on normally distributed data, however,
@@
 a characterization of equality of distributions.
-
 For instance, the \emph{standard
@@
 \label{eq:rho_standard}
-\rho_\alpha(X,Y) = \| X - Y\|^\alpha,
+\rho_\alpha(X,Y) = \| X - Y\|^\alpha
 \end{equation} 
@@
 
-Consider data $\mathbb{X} = \{ x_1,\dotsc, x_n \}$ 
-sampled from $k$ unknown distributions $\{ P_j \}_{j=1}^k$,
+Consider a sample $\mathbb{X} = \{ x_1,\dotsc, x_n \}$ 
+from $k$ unknown distributions $\{ P_j \}_{j=1}^k$,
 with points lying on a space of negative type,
@@
 Let $\mathbb{X} = \bigcup_{j=1}^k \C_j$ be a disjoint
-partition, $\C_i \cap \C_j = \emptyset$.
+partition, i.e. $\C_i \cap \C_j = \emptyset$.
 Each expectation in the generalized energy distance
@@
 \end{equation}
-where $n_i = |\C_i|$ is the number of elements in partition
+where $n_i = |\C_i|$ denotes the number of elements in partition
 $\C_i$. 
@@
 type and symmetric positive definite kernels are closely related
-\citep{Berg1984}. Let $\rho: \mathcal{X} \times \mathcal{X} \to \mathbb{R}$
-and $x_0 \in \mathcal{X}$ an arbitrary but fixed point.
+\citep{Berg1984}. Let $\rho: \mathcal{X} \times \mathcal{X} \to \mathbb{R}$,
+and $x_0 \in \mathcal{X}$ be an arbitrary but fixed point.
 Define
@@
 $\varphi: x \mapsto \kkk_x$ is injective \citep{Sejdinovic2013}.
-When these conditions are satisfied we say that the kernel $\kk$ 
+When these conditions are met we say that the kernel $\kk$ 
 generates the semimetric $\rho$. 
@@
 \E \, \kk(X, X') + \E \, \kk(Y, Y') - 2\E \, \kk(X, Y) \\ 
-&= \gamma_\kk^2(P,Q) ,
+&= \gamma_\kk^2(P,Q) .
 \end{split}
 \end{equation}
-so we can compute the energy distance using the inner product
+Therefore, we can compute the energy distance using the inner product
 of $\Hk$.
@@
 
-This section contains our main theoretical results, where 
+This section contains our main theoretical results where 
 we formulate an optimization problem for clustering 
-based on energy statistics in the RKHS introduced in the previous section.
+based on energy statistics in the RKHS.
 The proofs are contained in supplementary material.
@@
 all $i\ne j$, maximizes the between-sample statistic $S$, defined
-in equation \eqref{eq:between}, if and only if
+in \eqref{eq:between}, if and only if
 \begin{equation}
@@
 form, the relation with kernels and other clustering methods is obscure.
-In the following, we show what is the explicit optimization problem behind 
+In the following we demonstrate what is the explicit optimization problem behind 
 \eqref{eq:minimize} in the corresponding RKHS, 
@@
 \end{equation}
-Throughout the paper, we use the notation $M_{i\bullet}$ to denote
-the $i$th row of a matrix $M$, and $M_{\bullet j}$ denotes its $j$th column.
+%Throughout the paper, we use the notation $M_{i\bullet}$ to denote
+%the $i$th row of a matrix $M$, and $M_{\bullet j}$ denotes its $j$th column.
 Our next result shows that the optimization problem \eqref{eq:minimize}
 is NP-hard since
-it is a quadratically constrained quadratic program (QCQP) in the RKHS.
+it is a quadratically constrained quadratic program (QCQP) in the 
+associated RKHS.
 
@@
 indicating that $x_i \in \C_j$. 
-This optimization problem is nonconvex, and also NP-hard,
+This optimization problem is nonconvex, and NP-hard,
 thus a direct approach 
@@
 and will be compared to the iterative method that will be proposed 
-in the following.
-
-Note that the energy clustering problem \eqref{eq:qcqp2} 
+in the next section.
+
+Note that the energy clustering optimization problem \eqref{eq:qcqp2} 
 is valid for data living in an \emph{arbitrary} space of negative type, where
@@
 In practice, however,
-the clustering quality strongly depend on the choice of a suitable
+the clustering quality strongly depends on the choice of a suitable
 $\rho$ which measures the similarity between data points.
-If prior information is available to choose $\rho$
-it can be conveniently incorporated in the 
-optimization problem \eqref{eq:qcqp2}.
+If prior information is available to make this choice
+it can be immediately incorporated in the 
+optimization problem \eqref{eq:qcqp2} through the kernel matrix $G$.
 
@@
 confused with kernel $k$-means algorithm that is just one possible recipe 
-to solve \eqref{eq:kernel_kmeans}.}, 
+to solve \eqref{eq:kernel_kmeans}. The distinction should be clear from 
+the context.}, 
 extensively used in machine learning.
@@
 $\varphi: \mathcal{X} \to \HH_\kk$ such that
-$G_{ij} = \langle \varphi(x_i), \varphi(x_j) \rangle$. The kernel $k$-means optimization
+$G_{ij} = \langle \varphi(x_i), \varphi(x_j) \rangle$. 
+The kernel $k$-means optimization
 problem,
@@
 \eqref{eq:minimize} 
-is equivalent to the kernel $k$-means optimization problem
+is equivalent to kernel $k$-means optimization problem
 \eqref{eq:kernel_kmeans}, and both are equivalent to \eqref{eq:qcqp2}.
@@
 
-We introduce an iterative algorithm based
+We now introduce an iterative algorithm based
 on Hartigan's method \citep{Hartigan} to find a local
 maximizer of the optimization problem \eqref{eq:qcqp2}. Due to 
-Proposition~\ref{th:kernel_kmeans} we can use 
-kernel $k$-means algorithm \citep{Dhillon2,Dhillon}, which
-will be compared with the proposed algorithm.
-
-We can write the optimization problem \eqref{eq:qcqp2} as
+Proposition~\ref{th:kernel_kmeans} we can also use 
+kernel $k$-means algorithm \citep{Dhillon2,Dhillon} to solve
+this optimization problem, and it
+will be compared with the proposed algorithm in our numerical experiments.
+
+We can rewrite the optimization problem \eqref{eq:qcqp2} as
 \begin{equation}
@@
 \end{equation}
-where we recall that $G_{i\bullet}$ ($G_{\bullet i}$) denotes
-the $i$th row (column) of matrix $G$.
+where $G_{i\bullet}$ ($G_{\bullet i}$) denotes
+the $i$th row (column) of the matrix $G$.
 
@@
 $\Delta Q^{j\to \ell}(x_i)$ for 
-$\ell=1,\dots,k$ with $\ell \ne j$. Then, choose
+$\ell=1,\dots,k$ with $\ell \ne j$. Choose
 \begin{equation}
@@
 Repeat this process until no new assignments are made.
-The entire procedure is explicitly described in Algorithm~\ref{algo}, 
-which we call $\mathcal{E}^H$-clustering to emphasize that it is based on
+This procedure is explicitly described in Algorithm~\ref{algo}, 
+which we name $\mathcal{E}^H$-clustering to emphasize that it is based on
 Hartigan's method.
 Note that the objective function is
-monotonically increasing at each iteration, consequently the algorithm
+monotonically increasing at each iteration, consequently, the algorithm
 converges in a finite number of steps.
@@
                             2Q_{j^\star}(x_i)+ G_{ii}$
-    %    \ELSE
-    %        \STATE Do nothing;
         \ENDIF
@@
 \caption{\label{algo}
-$\mathcal{E}^H$-clustering is Hartigan's method to find local solutions to the optimization problem \eqref{eq:qcqp2}.
+$\mathcal{E}^H$-clustering is Hartigan's method to find local solutions 
+to the optimization problem \eqref{eq:qcqp2}.
 }
@@
 
-%Computing  $G$ requires $\OO( D n^2)$ operations, where 
-%$D$ is the dimension of $\mathcal{X}$ and $n$ is the data size, however,
-%the algorithm assumes $G$ is given. 
-%There are efficient methods to compute $G$, specially if it is sparse, 
-%but we will not consider this further.
-The worst time complexity of $\mathcal{E}^H$-clustering is
-$\OO(n k  \max_\ell n_\ell) =\OO(k n^2)$. 
-This is the same cost as kernel $k$-means algorithm
+The worst time complexity of the above algorithm is
+$\OO(n k  \max_\ell n_\ell) =\OO(k n^2)$, which
+is the same cost as kernel $k$-means algorithm
 \citep{Dhillon2,Dhillon}. 
@@
 
-There are known results about Hartigan's method,
+There are known results about Hartigan's method
 indicating its advantages over Lloyd's method.
@@
 
-This means that Hartigan's method can potentially 
+This means that Hartigan's can potentially 
 escape local optima of Lloyd's method.
@@
 while Hartigan's method groups data
-in regions called circlonoi cells.
+into regions called circlonoi cells.
 The circlonoi cells are within
@@
 The main goal of this section is threefold. First, to compare
-$\mathcal{E}$-clustering in Euclidean space 
+$\mathcal{E}^H$-clustering in Euclidean space 
 to $k$-means and GMM. 
@@
 
-The following experimental setup holds unless specified
+\textbf{Experimental Setup.} The following setup holds unless specified
 otherwise. 
 We consider
-$\mathcal{E}$-clustering, kernel $k$-means and spectral clustering
-with semimetrics
+$\mathcal{E}^H$-clustering, kernel $k$-means and spectral clustering
+with \mbox{generating} kernels to the following semimetrics:
 \begin{equation}
@@
 %
-\widehat{\rho}_{\sigma}(x,y) &= 2 - 2 e^{-\tfrac{\|x-y\|^2}{2 \sigma^2}}  
+\widehat{\rho}_{\sigma}(x,y) &= 2 - 2 e^{-\tfrac{\|x-y\|^2}{2 \sigma^2}}.  
 %\widehat{\kk}_{\sigma}(x,y) &= e^{-\tfrac{\|x-y\|^2}{2\sigma^2}}  
@@
 \end{equation}
-These are generated by the kernel \eqref{eq:kernel_semimetric}
-where we fix $x_0=0$.
-The standard $\rho_1$, from energy statistics in Euclidean
-spaces, will always be present as a reference and it
-is implied unless explicitly specified.
+These are generated by formula \eqref{eq:kernel_semimetric},
+and we fix $x_0=0$.
+The standard $\rho_1$ will always be present in the experiments 
+as a reference, being the implied choice unless explicitly mentioned.
 For $k$-means, GMM and spectral clustering we use the robust 
@@
 $k$-means++ \citep{Vassilvitskii}, 
-and GMM with the output of $k$-means, making it more robust and not
-breaking in high dimensions. 
+and GMM with the output of $k$-means, making it more robust and preventing
+it from breaking in high dimensions. 
+Spectral clustering is based on \citep{Malik}. 
 We implemented kernel $k$-means
@@
 the average accuracy over $100$ Monte
-Carlo trials, with error bars indicating the standard error.
+Carlo trials, with error bars indicating standard error.
+
+\begin{figure}
+\includegraphics[scale=0.381]{hist_normal.pdf}
+\includegraphics[scale=0.381]{hist_lognormal.pdf}\\[-0.6em]
+\includegraphics[scale=0.41]{1D_normal.pdf}
+\includegraphics[scale=0.41]{1D_lognormal.pdf}
+\put(-180,114){(a)}
+\put(-60,114){(c)}
+\put(-180,-8){(b)}
+\put(-60,-8){(d)}
+\caption{
+\label{fig:1D}
+Clustering results for one dimensional data. (a) and (c) show
+the probability densities of \eqref{eq:uni_normal} and
+\eqref{eq:uni_lognormal}, respectively. The corresponding 
+clustering accuracy versus sample size 
+are respectively shown in (b) and (d).
+Bayes accuracy in both cases are $\approx 0.88$.
+}
+\end{figure}
+
+The first two experiments involve univariate
+normal and lognormal mixtures:
+\begin{align}
+x &\stackrel{iid}{\sim} \tfrac{1}{2}
+\mathcal{N}(\mu_1,\sigma_1^2) + 
+\tfrac{1}{2} \mathcal{N}(\mu_2,\sigma_2^2) \label{eq:uni_normal} \\
+x & \stackrel{iid}{\sim} \tfrac{1}{2}
+\exp\left\{ \mathcal{N}(\mu_1,\sigma_1^2) \right\} + 
+\tfrac{1}{2} \exp \left\{ \mathcal{N}(\mu_2,\sigma_2^2)\right\} 
+\label{eq:uni_lognormal} \\
+\mu_1 &= 1.5, \ \sigma_1=0.3, \ \mu_2=0, \ \sigma_2=1.5.
+\end{align}
+We increase the sample size $n$ and show clustering accuracy 
+versus $n$ in Fig.~\ref{fig:1D}. $\mathcal{E}^H$-clustering 
+performs better than $k$-means for normally distributed data, 
+and worse than GMM 
+as expected. 
+However, for lognormal distributions $k$-means and GMM are only slightly
+better than chance while $\mathcal{E}^H$-clustering is still very
+accurate.
 
@@
 \label{fig:gauss}
-High dimensional Gaussian mixture.
-Comparison of $\mathcal{E}^H$-clustering, kernel $k$-means, spectral
-clustering, $k$-means and GMM (last three from scikit-learn).
-(a) Parameters as in \eqref{eq:gauss1}. 
-(b) Parameters as in \eqref{eq:gauss2}. Dashed line is Bayes accuracy.
-We plot mean accuracy versus number of dimensions $D$ over $100$ Monte
-Carlo trials, error bars are standard error.
+High dimensional Gaussian mixtures.
+(a) Parameters as in \eqref{eq:gauss1}, and  
+(b) as in \eqref{eq:gauss2}. Dashed line is Bayes accuracy.
+We plot accuracy versus $D$.
 }
@@
 
-First we analyze
-how the algorithms degrade as the number of dimensions increase while
-keeping Bayes error fixed.
+Next we analyze
+how the algorithms degrade as the number of dimensions increase.
 Consider data from a Gaussian mixture
 $x  \stackrel{iid}{\sim} 
-\tfrac{1}{2}\left[\mathcal{N}(\mu_1,\Sigma_1) +
-\mathcal{N}(\mu_2,\Sigma_2)\right]$ in $\mathbb{R}^D$ with
-$\Sigma_1=\Sigma_2 = I_D$,   
+\tfrac{1}{2} \mathcal{N}(\mu_1,\Sigma_1) +
+\tfrac{1}{2} \mathcal{N}(\mu_2,\Sigma_2)$ in $\mathbb{R}^D$ with
+$\Sigma_1=\Sigma_2 = I_D$,  and
 \begin{equation}
@@
 \end{equation}
-The Bayes error gives an optimal accuracy
-$\approx 0.86$, which is fixed as $D$ increases.
-We sample $200$ points for each trial.
-The results are shown in Fig.~\ref{fig:gauss}a, where
+Bayes error is fixed as $D$ increases giving an optimal accuracy
+$\approx 0.86$.
+We sample $200$ points on each trial.
+As shown in Fig.~\ref{fig:gauss}a, 
 $\mathcal{E}^H$- and spectral-clustering have close
 performance,  
-much better than kernel $k$-means, and also 
-better than 
+much better than kernel $k$-means, 
 $k$-means and GMM. The improvement is noticeable in 
@@
 We simply chose $10$ number uniformly at random on the interval
-$[1,5]$, and any other choice would give analogous results.
-The Bayes  accuracy is fixed at $\approx 0.95$.
-Fig.~\ref{fig:gauss}b we show the results. 
-GMM performs better in low dimensions, 
+$[1,5]$, and other choice would give analogous results.
+Bayes  accuracy is fixed at $\approx 0.95$.
+From Fig.~\ref{fig:gauss}b we see that 
+GMM is better in low dimensions, 
 but it quickly degenerates
-as $D$ increases, as (kernel) $k$-means, while  
+as $D$ increases, as kernel $k$-means and $k$-means, while  
 $\mathcal{E}^H$- and spectral-clustering remains much more stable.
@@
 \label{fig:consist}
-(a,b) Normal and (c,d) lognormal distributions in $\mathbb{R}^{20}$ with
+(a,b) Normal and (c,d) lognormal mixtures with
 parameters \eqref{eq:20gauss}. We use different kernels
-for $\mathcal{E}^H$-clustering, compared to 
-$k$-means and GMM. Bayes accuracy
-is $\approx 0.9$. We plot average accuracy (error bars are standard error)
-versus number of points for $100$ Monte Carlo trials.
-The plots in (c) and (d) consider the difference in accuracy
+for $\mathcal{E}^H$-clustering. Bayes accuracy
+is $\approx 0.9$.
+In (c) and (d) we show the difference in accuracy
 between $\mathcal{E}^H$ versus kernel $k$-means and
@@
 
-Now we sample $n$ points from the Gaussian mixture
-$x \stackrel{iid}{\sim} \tfrac{1}{2}\left[ \mathcal{N}(\mu_1,\Sigma_1)+
-\mathcal{N}(\mu_2, \Sigma_2) \right]$ in $\mathbb{R}^{20}$ with 
-$\Sigma_1=\tfrac{1}{2} I_{20}$, $\Sigma_2 = I_{20}$, 
+Consider
+$x \stackrel{iid}{\sim} \tfrac{1}{2} \mathcal{N}(\mu_1,\Sigma_1)+
+\tfrac{1}{2} \mathcal{N}(\mu_2, \Sigma_2)$ in $\mathbb{R}^{20}$ with 
+$\Sigma_1=\tfrac{1}{2} I_{20}$, $\Sigma_2 = I_{20}$, and
 \begin{equation}
@@
 Bayes accuracy is $\approx 0.90$. 
-We increase the sample size in the range $n \in [10, 400]$. 
-The results are 
-shown in Fig.~\ref{fig:consist}a, where we compare
+We increase the sample size $n \in [10, 400]$ and show
+the accuracy versus $n$ in Fig.~\ref{fig:consist}a, where we compare
 $\mathcal{E}^H$-clustering with different
 kernels, indicated in the legend, to $k$-means and GMM.
-%For small $n$ all methods
-%are superior than GMM, which eventually catches up and tend to optimal Bayes,
-%as expected since it is consistent to this setting. 
 Note that $\mathcal{E}^H$-clustering
-with $\widetilde{\rho}_1$ is as accurate as GMM for large number of points, 
-however, it is superior for small number of points. Still for
+with $\widetilde{\rho}_1$ is as accurate as GMM for large $n$ but
+superior for small $n$. For
 the same setting, in Fig.~\ref{fig:consist}b we
 we show the difference in accuracy provided by $\mathcal{E}^H$ minus
-kernel $k$-means and $\mathcal{E}^H$ minus spectral clustering, when using the
-semimetric $\widetilde{\rho}_1$.
-Note that $\mathcal{E}^H$ was
-always superior (there would be points with negative values on 
-the $y$-axis otherwise).
-Now consider the same parameters as in \eqref{eq:20gauss} but with
-lognormal mixtures in $\mathbb{R}^D$,
-$\log x \stackrel{iid}{\sim} \tfrac{1}{2}\left[ \mathcal{N}(\mu_1,\Sigma_1)+
-\mathcal{N}(\mu_2, \Sigma_2) \right]$. 
+kernel $k$-means and spectral clustering when using $\widetilde{\rho}_1$.
+$\mathcal{E}^H$-clustering was
+always superior, otherwise there would be points with negative $y$-values.
+Consider the same parameters as in \eqref{eq:20gauss} but now with
+lognormal mixtures in $\mathbb{R}^D$.
 The same experiments are shown in 
-Fig.~\ref{fig:consist}c and Fig.~\ref{fig:consist}d.
-Note that $\mathcal{E}^H$-clustering still performs accurately, 
+Fig.~\ref{fig:consist}c--d, where
+$\mathcal{E}^H$-clustering still performs accurately 
 with any of those
-kernels, providing better results than $k$-means and GMM. 
-%The semimetric $\widetilde{\rho}_1$ still provides the 
-%best results for small number of points, but its performance is eventually 
-%achieved by $\rho_{1/2}$, indicating that $\alpha \approx 1/2$ in the standard
-%energy distance should be more appropriate for skewed distributions.
-These two experiments
-illustrate how energy clustering is flexible,
-since it is accurate on data from very different distributions
-with the same kernel.
-
-\begin{figure}[t!]
+kernels, contrary to $k$-means and GMM.
+
+\begin{figure}[b!]
 \includegraphics[scale=0.21]{complex_data.pdf}
@@
 \begin{center}
-%\footnotesize{
-\begin{tabular}{@{}r|  l l  l l  l l@{}}
+\footnotesize{
+\begin{tabular}{@{}r  l l  l l  l l@{}}
 \toprule[1pt]
@@
 \end{tabular}
-%}
+}
 \end{center}
@@
 \begin{center}
-%\footnotesize{
-\begin{tabular}{@{}r@{}l@{}|@{}l@{}|@{}l@{}|@{}l@{}|@{}l@{}}
+\footnotesize{
+\begin{tabular}{@{}r@{}l@{}@{}l@{}@{}l@{}@{}l@{}l@{}}
 \toprule[1pt]
 \emph{Class Subset}~~~~ & & 
-$~\{0,1,2\}$ &
+%$~\{0,1,2\}$ &
 $~\{0,1,\dotsc,4\}~$ &
 $~\{0,1,\dotsc,6\}~$ &
-$~\{0,1,\dotsc,8\}~$ \\
-parameter~~~~ & {$\sigma$}
-& $~10.34~$
+$~\{0,1,\dotsc,8\}~$ &
+$~\{0,1,\dotsc,9\}~$ \\
+\emph{parameter}~~~~ & {$\sigma$}
+%& $~10.34~$
 & $~10.41~$
 & $~10.41~$
-& $~10.37~$ \\
+& $~10.37~$
+& $~10.19~$ \\
 \midrule[0.5pt]
@@
 & $\rho_{1}$\hspace{1em} 
-&$~0.937\pm 0.006~$
+%&$~0.937\pm 0.006~$
 &$~0.873\pm 0.025~$
@@
 &$~0.687\pm 0.016~$
+&$~0.581\pm 0.011~$
 \\
 & $\rho_{1/2}$ 
-&$~0.939\pm 0.006~$
+%&$~0.939\pm 0.006~$
 &$~0.874\pm 0.027~$
@@
 &$~0.647\pm 0.017~$
+&$~\bm{0.600\pm 0.009}~$
 \\
 & $\widetilde{\rho}_{\sigma}$ 
-&$~0.939\pm 0.006~$
+%&$~0.939\pm 0.006~$
 &$~0.847\pm 0.031~$
@@
 &$~0.657\pm 0.014~$
+&$~0.584\pm 0.013~$
 \\
 & $\widehat{\rho}_{\sigma}$ 
-&$~0.933\pm 0.005~$
+%&$~0.933\pm 0.005~$
 &$~\bm{0.891\pm 0.009}~$
@@
 &$~\bm{0.704\pm 0.011}~$
+&$~\bm{0.591\pm 0.012}~$
 \\
@@
 & $\widehat{\rho}_{\sigma}$ 
-&$~0.823\pm 0.015~$
+%&$~0.823\pm 0.015~$
 &$~0.769\pm 0.012~$
@@
 &$~0.649\pm 0.018~$
+&$~0.565\pm 0.009~$
 \\
 \emph{$k$-means}~~~~ & \xmark
-&$~0.927\pm 0.004~$
+%&$~0.927\pm 0.004~$
 &$~0.878\pm 0.010~$
@@
 &$~0.695\pm 0.012~$
+&$~0.557\pm 0.012~$
 \\
 \emph{GMM}~~~~ & \xmark
-&$~\bm{0.952\pm 0.005}~$
+%&$~\bm{0.952\pm 0.005}~$
 &$~0.839\pm 0.015~$
@@
 &$~0.621\pm 0.009~$
+&$~0.540\pm 0.009~$
 \\
@@
 In Fig.~\ref{fig:other}a--c we have 
-complex two dimensional datasets. 
-We apply $\mathcal{E}^H$-clustering  with different kernel choices.
-We also consider the best kernel choice for spectral
-clustering, besides 
+complex two dimensional data. 
+We apply $\mathcal{E}^H$  with different kernel choices,
+and also spectral clustering with the best kernel choice, besides 
 $k$-means and GMM. Here we perform only $10$ Monte Carlo runs.
@@
 The results are in Table~\ref{table:other}.
-$\mathcal{E}^H$-clustering has superior performance
-in every example, and in particular better than the
+$\mathcal{E}^H$ has superior performance
+in every example, in particular better than the
 spectral clustering.
-For the data in Fig.~\ref{fig:other}a 
-the semimetrics $\rho_1$ and $\rho_{1/2}$ still provide
-accurate results, however, for the
-examples in (b) and (c) the kernel choice is more sensitive.
-
-Next, we consider 
+In (a) the semimetrics $\rho_1$ and $\rho_{1/2}$ still provide
+accurate results, however, for
+ (b) and (c) the kernel choice is more sensitive.
+
+Consider 
 the infamous MNIST handwritten digits
-as illustrated in Fig.~\ref{fig:other}d.
+from Fig.~\ref{fig:other}d.
 Each data point is an $8$-bit gray scale
@@
 We compute the parameter 
-\begin{equation}
-\label{eq:sigma}
-\sigma^2 = \dfrac{1}{n^2} \sum_{i,j=1}^n \| x_i - x_j \|^2 
-\end{equation}
+%\begin{equation}
+%\label{eq:sigma}
+$
+\sigma^2 = \tfrac{1}{n^2} \sum_{i,j=1}^n \| x_i - x_j \|^2 
+$
+%\end{equation}
 from a separate
 training set, to be used in the kernels.
-We consider subsets of $\{0,1,\dotsc,9 \}$, 
-sampling $100$ points 
-for each class over $10$ Monte Carlo trials.
+We sample subsets of $\{0,1,\dotsc,9 \}$ with $100$ points 
+for each class, over $10$ Monte Carlo trials.
 The results are shown in Table~\ref{table:mnist}.
-Unsupervised clustering on MNIST without any feature extraction
-is not trivial. This 
-same experiment was performed in \citep{Sapiro} where a low-rank
+Unsupervised clustering on MNIST without feature extraction
+is not trivial. Similar 
+experiment was performed in \citep{Sapiro} where a low-rank
 transformation is learned then subsequently used in subspace clustering. 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Discussion}
+\section{Final Remarks}
 \label{sec:conclusion}
 
-We considered clustering from the perspective of generalized energy
+We proposed clustering from the perspective of generalized energy
 statistics, valid for arbitrary spaces of negative type.
 Our mathematical formulation of energy clustering 
-reduces to a QCQP in the associated RKHS, as demonstrated in 
-Proposition~\ref{th:qcqp2}.
-We showed that the optimization problem
-is equivalent
-to kernel $k$-means, once the kernel is fixed; see
+reduces to a QCQP in the associated kernel Hilbert space; 
+see  Proposition~\ref{th:qcqp2}.
+We demonstrated that such QCQP
+is equivalent to kernel $k$-means optimization problem, 
+once the kernel is fixed; see
 Proposition~\ref{th:kernel_kmeans}. Energy statistics, however, fixes
@@
 on spaces of negative type can also be obtained.
-We proposed the iterative $\mathcal{E}^H$-clustering algorithm based on 
-Hartigan's method, which was compared to kernel $k$-means algorithm
-based on Lloyd's heuristic.
+
+We proposed the iterative $\mathcal{E}^H$-clustering which can be seen
+as a kernelized version of Hartigan's method; see Algorithm~\ref{algo}. 
+It was compared to kernel $k$-means algorithm
+which is based on Lloyd's heuristic.
 Both have the same time complexity, however, numerical and theoretical
@@
 dimensions. 
-Furthermore, energy clustering, with standard kernels from energy
-statistics, outperformed $k$-means and GMM
-on several settings, even on Gaussian ones, illustrating the flexibility
-of the proposed method. In some
-examples the iterative
+Furthermore, $\mathcal{E}^H$-clustering with standard kernels from energy
+statistics outperformed $k$-means and GMM
+on several settings, even on Gaussian ones.
+In some cases
 $\mathcal{E}^H$-clustering also surpassed spectral clustering, and in
-others performed similarly but never worse.
-
-A limitation of $\mathcal{E}^H$-clustering  is that it cannot
-handle accurately highly unbalanced clusters. 
-An interesting problem would be to
-extend $\mathcal{E}^H$-clustering to these cases.
-Moreover, it would also be interesting to formally 
-demonstrate cases where energy clustering is a 
-consistent. A soft version of energy clustering is also an
-interesting extension.
+others performed similarly but never worse. Moreover, computing eigenvectors
+of large matrices can be unfeasible and iterative methods are preferred.
 Finally, kernel methods can benefit from sparsity and
-fixed-rank approximations of the Gram matrix, and there is plenty
-of room to make $\mathcal{E}^H$-clustering algorithm more scalable.
+fixed-rank approximations of the Gram matrix and there is plenty
+of room to make $\mathcal{E}^H$-clustering more scalable.
 
 \subsection*{Acknowledgements}
-We would like to thank Carey Priebe for discussions.
-This work was supported by NIH TRA grant.
+We would like to thank XXX
+%Carey Priebe 
+for discussions.
+This work was supported by 
+%NIH TRA 
+XXX
+grant.
 
@@
 
-Here we colect the proofs of our main results.
+Here we collect the proofs of our main results.
 

### 2017-10-13 energy_clustering/code/energy/bayes.py

@@ -67,7 +67,7 @@ def integrate_univariate_normal(mu1, mu2, sigma1, sigma2, p=0.5):
     return accuracy
 
 def bayes_univariate_lognormal(mu1, mu2, sigma1, sigma2, 
-                            num_times=10, num_points=1000):
+                            num_times=10, num_points=1000, p=0.5):
     """Estimate Bayes' error for two-class normal distributions in 1D."""
 
     def lognorm(x, mu, sigma):
@@ -75,15 +75,17 @@ def bayes_univariate_lognormal(mu1, mu2, sigma1, sigma2,
    
     error = np.zeros(num_times)
     for i in range(num_times):
-        x1 = np.random.lognormal(mu1, sigma1, num_points)
-        x2 = np.random.lognormal(mu2, sigma2, num_points)
+        q = 1-p
+        n1, n2 = np.random.multinomial(num_points, [p,q])
+        x1 = np.random.lognormal(mu1, sigma1, n1)
+        x2 = np.random.lognormal(mu2, sigma2, n2)
         
         error1 = sum([1 for x in x1 
                       if lognorm(x,mu1,sigma1)-lognorm(x,mu2,sigma2)<=0])
         error2 = sum([1 for x in x2 
                       if lognorm(x,mu1,sigma1)-lognorm(x,mu2,sigma2)>=0])
-        error[i] = error1+error2
-    return 1-error.mean()/(2*num_points)
+        error[i] = p*error1/n1+q*error2/n2
+    return 1-error.mean()
 
 def integrate_univariate_lognormal(mu1, mu2, sigma1, sigma2):
     """Same as above but through numerical integration."""
@@ -156,20 +158,20 @@ def bayes_multivariate_lognormal(mu1, mu2, sigma1, sigma2,
     
 #############################################################################
 if __name__ == '__main__':
-    
-    #print bayes_univariate_normal(0, 5, 1, 2,num_times=30,
-    #                                num_points=10000,p=0.9)
-    #print integrate_univariate_normal(0, 5, 1, 2, p=0.9)
-    
-    #print bayes_univariate_lognormal(0, -1.5, 0.3, 1.5, 
-    #                                 num_times=10, num_points=10000)
-    #print integrate_univariate_lognormal(0, -1.5, 0.3, 1.5)
-    
-    D = 25
-    m1 = np.zeros(D)
[…]

### 2017-10-13 energy_clustering/code/energy/data.py

@@ -225,36 +225,37 @@ def histogram_gauss_loggauss(xlim=None, ylim=None, fname='plot.pdf'):
         return 1/(np.sqrt(2*np.pi*sigma**2)*x)*np.exp(
                     -(np.log(x)-mu)**2/(2*sigma**2))
 
-    colors = iter(plt.cm.brg(np.linspace(0,1,6)))
+    #colors = iter(plt.cm.brg(np.linspace(0,1,6)))
+    colors = iter(['b', 'r'])
 
     fig = plt.figure()
     ax = fig.add_subplot(111)
     
-    xs = np.arange(-5, 15, 0.001)
-    #xs = np.arange(0.0001, 5, 0.001)
-    y1s = np.array([gauss_func(x, 0, 1) for x in xs])
-    y2s = np.array([gauss_func(x, 5, 2) for x in xs])
-    #y1s = np.array([loggauss_func(x, 0, 0.3) for x in xs])
-    #y2s = np.array([loggauss_func(x, -1.5, 1.5) for x in xs])
+    #xs = np.arange(-4, 4, 0.001)
+    xs = np.arange(0.000001, 10, 0.001)
+    #y1s = np.array([gauss_func(x, 1.5, 0.3) for x in xs])
+    #y2s = np.array([gauss_func(x, 0, 1.5) for x in xs])
+    y1s = np.array([loggauss_func(x, 1.5, 0.3) for x in xs])
+    y2s = np.array([loggauss_func(x, 0, 1.5) for x in xs])
     c = next(colors)
     ax.plot(xs, y1s, '-', color=c, linewidth=1)
-    ax.fill_between(xs, 0, y1s, color=c, alpha=0.5)
+    ax.fill_between(xs, 0, y1s, color=c, alpha=0.3)
     c = next(colors)
     ax.plot(xs, y2s, '-', color=c, linewidth=1)
-    ax.fill_between(xs, 0, y2s, color=c, alpha=0.5)
+    ax.fill_between(xs, 0, y2s, color=c, alpha=0.3)
 
-    ax.set_xlabel(r'$x$')
+    #ax.set_xlabel(r'$x$')
     if xlim:
         ax.set_xlim(xlim)
     if ylim:
         ax.set_ylim(ylim)
-    plt.tick_params(top='off', bottom='on', left='off', right='off',
-            labelleft='off', labelbottom='on')
-    for i, spine in enumerate(plt.gca().spines.values()):
-        if i !=2:
[…]

### 2017-10-13 energy_clustering/code/experiment1_2.py

@@ -1,0 +1,125 @@
+"""First experiment. 1D data."""
+
+# Guilhere Franca <guifranca@gmail.com>
+# Johns Hopkins University, Neurodata
+
+import numpy as np
+import multiprocessing as mp
+
+import energy.data as data
+import energy.eclust as eclust
+import energy.initialization as initialization
+import run_clustering
+import plot
+
+def sample_normal(m1, m2, s1, s2, n1, n2):
+    return data.univariate_normal([m1, m2], [s1, s2], [n1, n2])
+
+def sample_lognormal(m1, m2, s1, s2, n1, n2):
+    return data.univariate_lognormal([m1, m2], [s1, s2], [n1, n2])
+
+
+class Clustering1D:
+
+    def __init__(self):
+        # parameters for distributions
+        self.m1 = 1.5
+        self.s1 = 0.3
+        self.m2 = 0
+        self.s2 = 1.5 
+        self.distr = 'normal' # should be normal or lognormal
+        self.numpoints = range(50, 100, 20)
+        self.experiments = 100 # number of experiments
+
+    def get_sample(self, n):
+        n1, n2 = np.random.multinomial(n, [0.5, 0.5])
+        if self.distr == 'normal':
+            X, z = sample_normal(self.m1, self.m2, self.s1, self.s2, n1, n2)
+        elif self.distr == 'lognormal':
+            X, z = sample_lognormal(self.m1, self.m2, self.s1, self.s2, n1, n2)
+        else:
+            raise ValueError("Clustering 1D, unknown distribution to sample.")
+        Y = np.array([[x] for x in X])
+        return Y, z
+
+    def run(self):
+        k = 2
+        ncols = 3 + 1 # number of clustering methods + 1
+        table = np.zeros((self.experiments*len(self.numpoints), ncols))
+        count = 0
+        init = 'k-means++'
+        for i, n in enumerate(self.numpoints):
+            for ne in range(self.experiments):
+                Y, z = self.get_sample(n)
+                G = eclust.kernel_matrix(Y, lambda x, y: np.linalg.norm(x-y))
+                table[count, 0] = n
+                table[count, 1] = run_clustering.energy_hartigan(k, Y, G, z,
+                                                    init=init, run_times=5)
+                table[count, 2] = run_clustering.kmeans(k, Y, z, init=init,
+                                                    run_times=5)
+                table[count, 3] = run_clustering.gmm(k, Y, z, init='kmeans',
[…]

### 2017-10-13 energy_clustering/code/experiment3_2.py

@@ -1,0 +1,106 @@
+"""Third experiment. Normal distributions for unbalanced clusters.
+Using mixtures now.
+
+"""
+
+# Guilhere Franca <guifranca@gmail.com>
+# Johns Hopkins University, Neurodata
+
+import numpy as np
+import multiprocessing as mp
+
+import energy.data as data
+import energy.eclust as eclust
+import energy.initialization as initialization
+import run_clustering
+import plot
+
+def gauss_dimensions_pi(num_points=range(0, 180, 10), num_experiments=10):
+    """Test unbalanced clusters."""
+    k = 2
+    D = 4
+    d = 2
+    N = 250
+    table = np.zeros((num_experiments*len(num_points), 6))
+    count = 0
+
+    for p in num_points:
+        for i in range(num_experiments):
+
+            # generate data
+            m1 = np.zeros(D)
+            s1 = np.eye(D)
+            m2 = np.concatenate((1.5*np.ones(d), np.zeros(D-d)))
+            s2 = np.diag(np.concatenate((.5*np.ones(d), np.ones(D-d))))
+            pi1 = (N-p)/N
+            pi2 = (N+p)/N
+            print pi1, pi2
+            n1, n2 = np.random.multinomial(N, [pi1,pi2])
+            X, z = data.multivariate_normal([m1, m2], [s1, s2], [n1, n2])
+            rho = lambda x, y: np.linalg.norm(x-y)
+            G = eclust.kernel_matrix(X, rho)
+
+            table[count, 0] = p
+            table[count, 1] = run_clustering.energy_hartigan(k, X, G, z,
+                                                init="k-means++", run_times=5)
+            table[count, 2] = run_clustering.energy_lloyd(k, X, G, z,
+                                                init="k-means++", run_times=5)
+            table[count, 3] = run_clustering.spectral(k, X, G, z,
+                                                run_times=5)
+            table[count, 4] = run_clustering.kmeans(k, X, z,
+                                                init="k-means++", run_times=5)
+            table[count, 5] = run_clustering.gmm(k, X, z,
+                                                init="kmeans", run_times=5)
+
+            count += 1
+
+    return table
+
+def make_plot(*data_files):
+    table = []
+    for f in data_files:
+        t = np.loadtxt(f, delimiter=',')
+        t[:,0] = np.array([int(D) for D in t[:,0]])
+        table.append(t)
[…]

### 2017-10-24 energy_clustering/draft_v2/energy_clustering.tex
draft v2

--- 2017-10-12 energy_clustering/draft/energy_clustering.tex
@@
 %\usepackage{cite}
+\usepackage{mathtools}
+\mathtoolsset{showonlyrefs}
 \usepackage[inline]{enumitem}
@@
 \usepackage{makecell}
+\usepackage{pifont}
+\newcommand{\cmark}{\ding{51}}
+\newcommand{\xmark}{\ding{55}}
 
@@
 kernel Hilbert spaces (RKHS) was established. 
-Here we consider the problem of clustering data from an 
-energy statistics theory perspective.
-We provide a precise mathematical formulation 
+Here we consider the clustering problem from an 
+energy statistics theory perspective, 
+providing a precise mathematical formulation 
 yielding a quadratically constrained 
-quadratic program (QCQP) in the associated RKHS. We show that this QCQP
-is equivalent to kernel $k$-means optimization problem, once the kernel
+quadratic program (QCQP) in the associated RKHS, thus establishing the
+connection with kernel methods. We show that this QCQP
+is equivalent to kernel $k$-means optimization problem once the kernel
 is fixed.
-Thus, our results imply a first principles derivation of kernel $k$-means 
+These results imply a first principles derivation of kernel $k$-means 
 from energy statistics.
@@
 making connection to graph partitioning problems.
-To find local optimizers of such QCQP we consider an iterative algorithm based 
+To find local optimizers of such QCQP we propose an iterative algorithm based 
 on Hartigan's method, which in this case has the same computational cost 
-as kernel $k$-means algorithm, based on Lloyd's method, but usually 
+as kernel $k$-means algorithm, based on Lloyd's heuristic, but usually 
 with better clustering quality. 
 We provide carefully designed numerical experiments showing the superiority 
-of the proposed method compared to kernel $k$-means, standard $k$-means, 
-and Gaussian mixture models in a variety of settings.
+of the proposed method compared to kernel $k$-means, spectral clustering,
+standard $k$-means, and Gaussian mixture models in a variety of settings.
 \end{abstract}
-
-% @gui: you didn't update the abstract?
-% @jovo: I did, removed some stuff that you didn't like before, now
-% removing the non-parametric word. You think it's not good enough?
 
@@
 in close analogy to Newton's gravitational potential in classical mechanics.
-It provides a non-parametric hypothesis test for equality of 
+It provides a model-free hypothesis test for equality of 
 distributions which is achieved 
@@
 correlation coefficient, and hierarchical clustering \cite{RizzoClustering} 
-by extending Ward's method of minimum variance. Moreover, an application of 
-energy statistics to clustering was already proposed \cite{Kgroups}, 
-which in part motivated this paper. We refer the reader to \cite{Szkely2013}, 
+by extending Ward's method of minimum variance. Moreover, 
+in Euclidean spaces, an application of 
+energy statistics to clustering
+was already proposed \cite{Kgroups}. 
+We refer the reader to \cite{Szkely2013}, 
 and references therein, for an overview of energy statistics theory and 
@@
 based tests has 
-been recently resolved \cite{Sejdinovic2013}, where a unifying framework
+been recently resolved \cite{Sejdinovic2013}, 
 establishing an equivalence between generalized energy distances to maximum
 mean discrepancies (MMD), which are distances between embeddings of 
-distributions in reproducing kernel Hilbert spaces (RKHS), was established. 
+distributions in reproducing kernel Hilbert spaces (RKHS).
 This equivalence immediately relates energy statistics to
-kernel methods, often used in machine learning, and form the basis 
+kernel methods often used in machine learning, and form the basis 
 of our approach.
@@
 for a given dataset, and usually a kernel introduces hyperparameters that 
-need to be carefully chosen.
-The well-known kernel $k$-means optimization problem is nothing but $k$-means 
-in the feature space \cite{Girolami}. Furthermore, kernel $k$-means algorithm
+need to be carefully chosen. A well-known kernel based clustering method
+is kernel $k$-means, which is preciselly $k$-means 
+formulated in the feature space \cite{Girolami}. 
+Furthermore, kernel $k$-means algorithm
 \cite{Dhillon2,Dhillon} is still based on Loyd's heuristic \cite{Lloyd}
-of grouping points that are closer to a cluster center, now
-in feature space. 
+of grouping points that are closer to a cluster center.
 We refer the reader to \cite{Filippone} for a survey of clustering
@@
 
-Although clustering from energy statistics was considered
-in \cite{Kgroups}, the precise optimization problem behind this approach
+Although clustering from energy statistics, in Euclidean spaces,
+was considered in \cite{Kgroups}, 
+the precise optimization problem behind this approach
 remains elusive, as well as the connection with kernel methods.
@@
 mathematical formulation to this statement, leading to a quadratically 
-constrained quadratic program (QCQP) in the associated RKHS. Our results
-immediately establish the connection with kernel methods, showing that this 
-QCQP is equivalent to kernel $k$-means optimization problem. The 
-equivalence between kernel $k$-means, spectral clustering, and graph 
-partitioning problems is well-known \cite{Dhillon,Dhillon2}. We demonstrate 
+constrained quadratic program (QCQP) in the associated RKHS. This immediatelly
+establishes the connection between energy statistics based clustering,
+or \emph{energy clustering} for short, with kernel methods. Moreover,
+our formulation holds for general semimetric spaces of negative type.
+We also show that such QCQP is equivalent to kernel $k$-means 
+optimization problem, however, the kernel is fixed by 
+energy statistics.
+The equivalence between kernel $k$-means, spectral clustering, and graph 
+partitioning problems is well-known \cite{Dhillon,Dhillon2}. We further
+demonstrate 
 how these relations arise from a weighted version of energy statistics.
@@
 find local solutions of the above mentioned QCQP, which is NP-hard in general.
-Hartigan's method was also used in \cite{Kgroups}. More importantly, its 
+Hartigan's method was also used in \cite{Kgroups}, but without
+any connection to kernels. More importantly, the
 advantages
-over Lloyd's method was already demonstrated 
+of Hartigan's over Lloyd's method was already demonstrated 
 in some simple settings
@@
 advantages. Furthermore, in the examples considered in this paper, it 
-also provides superior performance compared to a spectral clustering
-relaxation of the QCQP.
+also provides superior performance compared spectral clustering, which
+is more expensive and in fact solves a relaxed version of our QCQP.
 
 Our numerical results provide compelling evidence that 
-Hartigan's method applied to energy statistics based clustering, 
-or \emph{energy clustering} for short, is more accurate and robust than 
-Lloyd's method. Our experiments also put in evidence the
-flexibility of energy clustering, which provides a family of default kernels, 
+Hartigan's method applied to energy clustering 
+is more accurate and robust than kernel $k$-means algorithm.
+Furthermore, our experiments illustrate 
+the flexibility of energy clustering, 
 showing that it is able to perform accurately on data coming from 
 very different distributions, contrary to $k$-means and GMM for instance.
-More specifically, our algorithms for energy clustering perform 
-closely to $k$-means and GMM on normally distributed data, but on the other
-hand, performs considerably better than $k$-means and GMM on  data that 
-is not normally distributed. It also performs better than $k$-means 
-and GMM in high dimensions, even on Gaussian settings. However, it performs 
-worse than GMM, and more closely to $k$-means, for highly unbalanced clusters
-with normal distributions.
-
-%Our work is organized as follows. In section~\ref{sec:background}, we introduce
-%the necessary background on energy statistics and RKHS.
-%Section~\ref{sec:clustering_theory} contains the main theoretical 
-%results of this paper,
-%where we consider a clustering theory based on energy statistics leading
-%to a QCQP, which is NP-hard in general. We also show the equivalence to 
-%kernel $k$-means optimization problem.
-%In Section~\ref{sec:weighted}, we generalize these results to a weighted
-%version of energy statistics, providing connections to graph
-%partitioning problems.
-%In Section~\ref{sec:twoclass}, we consider the  simple case of a two-class
-%problem in one dimension,
-%where we propose an algorithm which requires no random initialization.
-%In section~\ref{sec:algo}, we consider Lloyd's and Hartigan's methods
-%and propose iterative algorithms to solve the QCQP.
-%Section~\ref{sec:numerics} contains some carefully designed numerical
-%experiments showing that, in a variety of settings, energy clustering based
-%on Hartigan's algorithm
-%outperform $k$-means, GMM, kernel $k$-means, and a version of
-%spectral clustering.
-%Our final conclusions are presented in Section~\ref{sec:conclusion}.
+More specifically, the proposed method performs 
+closely to $k$-means and GMM on normally distributed data, however,
+it is significantly better on data that 
+is not normally distributed. 
+Its superiority in high dimensions
+is striking, being more accurate than $k$-means and GMM 
+even on Gaussian settings.
 
@@
 $\rho^{1/2}$ does since it can be shown to be a proper metric. 
+Our energy clustering formulation, proposed in the next section,
+will be based on the generalized
+energy distance \eqref{eq:energy3}.
 
@@
 
-Now we can state the equivalence between energy distance $\Energy$ and
+Now we can state the equivalence between the generalized 
+energy distance \eqref{eq:energy3} and
 inner products on RKHS, which is one of the main results of
@@
 
-Finally, let us recall the main formulas from energy statistics
+Finally, let us recall the main formulas from generalized energy statistics
 for the test statistic of equality of distributions \cite{Szkely2013}. 
@@
 
-% @gui: maybe common in math, but certainly not in stats / data science. 
-% in any case, what is the package that does that?
-% @jovo: \usepackage{mathtools} \mathtoolsset{showonlyrefs}
-
-% @gui: i'm not sure about this.  who is to say the correct direction of causality? maybe energy clustering is a special case of kernel k-means?
-% @jovo: I'll put a comment below, I simply removed the previous phrase,
-% didn't say anything instead.
-
 This section contains the main theoretical results of this paper, where 
@@
 
-Due to the energy test statistic for equality of distributions,
+Due to the previous test statistic for equality of distributions,
 the obvious
@@
 \label{eq:relaxed}
-\max_{Y} \Tr \left( Y^\top G \, Y \right) \quad \mbox{s.t. $Y^\top Y = I$}
+\max_{Y} \Tr \left( Y^\top G \, Y \right) \qquad \mbox{s.t. $Y^\top Y = I$}
 \end{equation}
@@
 \sum_{i=1}^k \lambda_i$. 
-
 Spectral clustering is based on the above approach, where
-one further normalize the rows of $Y^\star$, 
-i.e. $Y^\star \to \widetilde{Y}$ where 
-$\widetilde{Y}_{ij} \equiv Y^\star_{ij} \big/ \| Y^\star_{i\bullet} \|$. 
-The vectors
-$y_i \equiv \widetilde{Y}_{i\bullet} \in \mathbb{R}^k$, which are the
-rows of matrix $\widetilde{Y}$, lie on the surface of the $k$-dimensional 
-unit sphere.
-The set $\{ y_i \}_{i=1}^n$ is a spectral embedding of the original
-set $\{ x_i \}_{i=1}^n$ onto an eigen-subspace of the kernel matrix $G$.
-After performing this embedding, one group the points 
-$\{ y_i \}_{i=1}^n$ into $k$ clusters 
-$\{ \tC_1,\dotsc,\tC_k \}$ using any clustering algorithm, such as
-$k$-means. This yields the final solution
-where the original point $x_i$ belongs
-to cluster $\C_j$ if and only if $y_i$ belongs to $\tC_j$.
-A procedure on these lines was proposed in the seminal paper
-\cite{NgJordan}, which considers the eigen-decomposition of
-the matrix $D^{-1/2} G D^{-1/2}$ instead of $G$, and also 
-where $G$ is an affinity matrix with the modification that
-its diagonal elements are set to zero $G_{ii}=0$, 
-and $D=\diag(d_{1},\dotsc,d_{n})$ is the degree matrix, 
-with $d_{i} \equiv \sum_{j=1}^n G_{ij}$. 
-We consider a slightly modification of this procedure,
-which we call \mbox{$\mathcal{E}^{S}$-clustering} 
-to emphasize that its a spectral
-method whose kernel comes from energy statistics, by 
-computing the 
-eigen-decomposition of $G D^{-1}$ instead.
-This  method is described in 
-Algorithm~\ref{spectralalgo} and will be used as a comparison
-to the other iterative algorithms 
-that will be proposed later.
-
-\begin{figure}
-\begin{flushleft}
-\begin{algorithm}[H]
-\vspace{.5em}
-\begin{algorithmic}[1]
-\INPUT Gram matrix $G\in\mathbb{R}^{n\times n}$, and number of clusters $k$
-\OUTPUT label matrix $Z \in \{ 0, 1\}^{n \times k}$
-\STATE set the diagonal of $G$ to zero: $G_{ii} \leftarrow 0$ 
-        for $i=1,\dotsc,n$
-\STATE $D\leftarrow\diag(d_1,\dotsc,d_n)$ 
-        where $d_i \leftarrow \sum_{j=1}^n G_{ij}$
-\STATE solve the generalized eigenvalue problem $G v_i = \lambda_i D v_i$,
-        where $i=1,\dotsc,n$ with eigenvalues, and corresponding eigenvectors,
-        obeying the ordering
-        $\lambda_1 \ge \lambda_2 \ge \dotsc \lambda_n$
-\STATE $Y \leftarrow (D v_1| \dotsm | D v_k) 
-        \in \mathbb{R}^{n\times k}$ 
-has the top $k$ eigenvectors (multiplied by $D$) as columns
-\STATE normalize the rows $Y$, i.e. 
-$\widetilde{Y}_{ij} \leftarrow Y_{ij} \Big/ 
-\left(\sum_{j=1}^n |Y_{ij}|^2\right)^{1/2}$
-\STATE apply $k$-means on $\{ y_i \}_{i=1}^n$, 
-where $y_i = \widetilde{Y}_{i\bullet}$
-is the $i$th row of $\widetilde{Y}$
-\STATE $Z_{ij} \leftarrow 1$ if $y_i$ is in cluster $j$, 
-        and $Z_{ij} \leftarrow 0$ otherwise
-\end{algorithmic}
-\caption{
-\label{spectralalgo}
-$\mathcal{E^S}$-clustering is an exact spectral method to solve
-the relaxed problem \eqref{eq:relaxed}.
-\hspace{\fill}
-}
-\end{algorithm}
-\end{flushleft}
-\end{figure}
-
-Note that the energy clustering problem \eqref{eq:qcqp2} 
+one further normalize the rows of $Y^\star$, then cluster
+the resulting rows as data points.
+A procedure on these lines was proposed in the seminal papers
+\cite{Malik,NgJordan}.
+
+Note that the optimization problem \eqref{eq:qcqp2} based on energy
+statistics
 is valid for data living in an \emph{arbitrary} space of negative type, where
 a semimetric $\rho$, and thus the kernel $\kk$, are
-assumed to be known. Standard energy statistics in
-Euclidean space fixes a family of choices 
-$\rho_\alpha(x,y) = \| x - y\|^\alpha$, with the corresponding
-kernel $K_\alpha(x,y) = \tfrac{1}{2} \left( \| x \|^\alpha + \| y \|^\alpha - 
-\| x-y\|^\alpha \right)$, where we
-set $x_0=0$ in \eqref{eq:kernel_semimetric} and  $0<\alpha\le 2$.
+assumed to be known. The standard energy distance \eqref{eq:energy2}
+fixes a family of choices in Euclidean spaces given by
+\begin{equation}
+\rho_\alpha(x,y) = \| x - y\|^\alpha, \qquad 
+K_\alpha(x,y) = \tfrac{1}{2} \left( \| x \|^\alpha + \| y \|^\alpha - 
+\| x-y\|^\alpha \right) ,
+\end{equation}
+for $0<\alpha\le 2$ and 
+we fix $x_0=0$ in \eqref{eq:kernel_semimetric}.
 The same would be valid for data living in a more general
-metric space $(\mathcal{X}, \rho)$, where the corresponding semimetric $\rho$ 
+semimetric space $(\mathcal{X}, \rho)$ where $\rho$ 
 fixes the kernel.
-In any case, the energy  clustering 
-formulation does not make assumptions
-about  the distribution of the data, 
-contrary to $k$-means and GMM, for example.
-In practice, however,
-the clustering quality strongly depend on the choice of a suitable
-$\rho$, which measures the similarity between data points,
-and is equivalent to choosing an appropriate kernel.
-However, if prior information is available to choose an appropriate $\rho$,
-it can be conveniently incorporated in the 
+In practice, the clustering quality strongly depend on the choice 
+of a suitable $\rho$. 
+Nevertheless, if prior information is available to make this choice, it
+can be immediately incorporated in the 
 optimization problem \eqref{eq:qcqp2}.
@@
 confused with kernel $k$-means algorithm that is just one possible recipe 
-to solve \eqref{eq:kernel_kmeans}.} 
+to solve \eqref{eq:kernel_kmeans}. The distinction should also be clear
+from the context.} 
 which is extensively used in machine learning.
@@
 $\varphi: \mathcal{X} \to \HH_\kk$ such that
-$\kk(x,y) = \varphi(x)^\top \varphi(y)$. The kernel $k$-means optimization
+$\kk(x,y) = \langle \varphi(x), \varphi(y) \rangle$. 
+Kernel $k$-means optimization
 problem,
@@
 \begin{proof}
-Notice that $\| \varphi(x) - \varphi(\mu_j) \|^2 = \varphi(x)^\top \varphi(x)
-- 2 \varphi(x)^\top \varphi(\mu_j) + \varphi(\mu_j)^\top \varphi(\mu_j)$,
+Notice that $\| \varphi(x) - \varphi(\mu_j) \|^2 = 
+\langle \varphi(x), \varphi(x) \rangle
+- 2 \langle \varphi(x), \varphi(\mu_j) \rangle +
+\langle \varphi(\mu_j), \varphi(\mu_j)\rangle$,
 therefore
@@
 
-% @jovo: the previous paragraph saying that kernel k-means is a consequence
-% of energy statistics WAS here, I removed and now don't say anything.
 The above result shows that 
-kernel $k$-means problem is equivalent to the clustering problem
+kernel $k$-means optimization problem is equivalent to the clustering problem
 formulated in the energy statistics framework, when operating on the same
 kernel. Notice, however, that 
-energy statistics theory is valid for arbitrary semimetric spaces of
+energy statistics is valid for arbitrary semimetric spaces of
 negative type, fixing the kernel function in the associated RKHS, which
-is guaranteed to be positive definite.
-
-As shown in \cite{Dhillon2,Dhillon}, kernel $k$-means, spectral clustering,
+is guaranteed to be positive definite. On the other hand, 
+kernel $k$-means \eqref{eq:kernel_kmeans} by itself is just an heuristic 
+approach that does not make any explicit mention to the kernel. 
+Based on Proposition~\ref{th:kernel_kmeans}
+one may view kernel $k$-means as being derived from the energy statistics
+framework.
+
+Kernel $k$-means, spectral clustering,
 and graph partitioning problems such as ratio association, ratio cut, and
-normalized cut are all equivalent to a QCQP of the form \eqref{eq:qcqp2}. Thus
-one can use kernel $k$-means algorithm to solve these problems as well.
+normalized cut are all equivalent to a QCQP of the form \eqref{eq:qcqp2}
+\cite{Dhillon2,Dhillon}.
+One can thus use kernel $k$-means algorithm to solve these problems as well.
 This correspondence involves a weighted version of problem 
 \eqref{eq:qcqp2}, that
-we demonstrate in the following from the perspective of energy statistics.
+will be demonstrated in the following from the perspective 
+of energy statistics.
 
@@
 
-We generalize the formulas from energy statistics to incorporate
+We now generalize energy statistics to incorporate
 weights associated to each data point.
 Let $w(x)$ be a weight function associated to point $x \in \mathcal{X}$.
-We can generalize \eqref{eq:g_def} as follows:
+Define
 \begin{equation}
@@
 \end{equation}
-Now we replace this function in the formulas \eqref{eq:within} and
+Replace this function in the formulas \eqref{eq:within} and
 \eqref{eq:between}, with 
-$n_i \to s_i$ and $n \to s$, where $s = \sum_{j=1}^k s_j$,
+$n_i \to s_i$ and $n \to s$, where $s = \sum_{j=1}^k s_j$.
 With these
-changes, Proposition~\ref{th:minimize} remains the unaltered, so the
+changes Proposition~\ref{th:minimize} remains the unaltered, so the
 clustering problem becomes
@@
 where now $g$ is given by \eqref{eq:g_def2}. 
-Let us define the following matrices and vector:
+Define the following matrices and vector:
 \begin{equation}
@@
 where $w_i = w(x_i)$ and $\e \in \mathbb{R}^n$ is the all-ones
-vector. We can now show the analogous of
-Proposition~\ref{th:qcqp2} to the case of problem \eqref{eq:minimize2}.
+vector.The analogous of
+Proposition~\ref{th:qcqp2} is as follows.
 
@@
 \end{equation}
-therefore $H^\top H = I$. This is a constraint on the rows of $H$.
+where $\delta_{ij}=1$ if $i=j$ and $\delta_{ij}=0$ if $i\ne j$ is the
+Kronecker delta. Therefore, $H^\top H = I$. 
+This is a constraint on the rows of $H$.
 To obtain a condition on its columns
@@
 where $\mathcal{V}$ is the set of vertices, $\mathcal{E}$ the set of edges,
-and $\mathcal{A}$ is an affinity matrix of the graph 
-that measures the 
+and $\mathcal{A}$ is an affinity matrix of the graph, 
+which measures the 
 similarities between pairs of nodes. Thus, $\mathcal{A}_{ij} \ne 0$
@@
 \eqref{eq:weighted_matrices}. Therefore, the ratio association problem
-can be written in the form \eqref{eq:qcqp3} as
+can be written in the form \eqref{eq:qcqp3}, i.e.
 \begin{equation}
@@
 
-\begin{figure}
-\begin{minipage}{0.49\textwidth}
-\includegraphics[width=\textwidth]{1D_normal.pdf}\\[-1em](a)
-\end{minipage}
-\begin{minipage}{0.49\textwidth}
-\includegraphics[width=\textwidth]{1D_lognormal.pdf}\\[-.8em](b)
-\end{minipage}
-\caption{
-\label{fig:1d}
-$\mathcal{E}^{1D}$-clustering versus $k$-means and GMM in one dimension.
-The two clusters have the same number of points, and we use
-$50$ Monte Carlo runs.
-We plot the average 
-accuracy \eqref{eq:accuracy} versus the total number of points, 
-where error bars are standard error. The dashed line indicates 
-Bayes accuracy. (a) Data normally distributed as in 
-\eqref{eq:two_normal}, where the
-optimal accuracy is $\approx 0.956$.
-(b) Data log-normally distributed as in \eqref{eq:two_lognormal} where 
-the optimal accuracy is $\approx 0.852$.
-}
-% @gui: if you insist on using the equations, at least also mention something semantic about them, ie, Data distributed Gaussian (as in 46), or Data distributed log Gaussian (as in 47).
-figures & captions should stand on their own without requiring referring to the paper.
-% @gui: why did you take away the plot of the distribution? those are incredibly helpful.
-% @jovo: ok will restore them.
-\end{figure}
-
-We now consider two simple experiments where we sample the same 
-number of points for each cluster. 
-We plot the average accuracy \eqref{eq:accuracy} versus the 
-total number of points, with error bars indicating standard error.
+We now consider two simple experiments where we sample $n$ points
+from a two-class mixture.
+We plot the average accuracy \eqref{eq:accuracy} versus $n$, 
+with error bars indicating standard error.
 The data is clustered using $\mathcal{E}^{1D}$-clustering algorithm,
-GMM through EM algorithm, and the standard $k$-means 
-algorithm. For $k$-means and GMM we use the implementations
+GMM and $k$-means. 
+For GMM and $k$-means we use the implementations
 from the well-known \emph{scikit-learn} library in 
@@
 $k$-means++ procedure \cite{Vassilvitskii}, and GMM is initialized
-with the output of $k$-means. We choose to run both algorithms
-$5$ times with different initializations and the answer with best 
-objective function value is returned.
-Notice that $\mathcal{E}^{1D}$-clustering does not require any random
+with the output of $k$-means. We run both algorithms
+$5$ times with different initializations and pick the answer with best 
+objective function value.
+Notice that $\mathcal{E}^{1D}$-clustering does not require random
 initialization so we only run it once.
-For each fixed number of points we use use $50$ Monte Carlo runs.
+For each $n$ we use use $100$ Monte Carlo runs.
 In  Fig.~\ref{fig:1d}a 
-we have the results for data sampled from two normal distributions, 
+we have the results for data sampled from the Gaussian mixture
 \begin{equation}
 \label{eq:two_normal}
-x \sim \mathcal{N}\big(\mu_i,\sigma_i^2\big) 
-\quad 
-\mbox{with $\mu_1 = 0$, $\sigma_1=1$ and
-$\mu_2 = 5$, $\sigma_2 = 2$.}
+x \stackrel{iid}{\sim} 
+\tfrac{1}{2}\mathcal{N}\big(\mu_1,\sigma_1^2\big) 
++\tfrac{1}{2}\mathcal{N}\big(\mu_2,\sigma_2^2\big), 
+\qquad 
+\mu_1 = 1.5, 
+\sigma_1=0.3, 
+\mu_2 = 0, 
+\sigma_2 = 1.5.
 \end{equation}
@@
 obtained from Bayes classification error
-is $ \approx 0.956$, indicated by the dashed line in the plot.
+is $ \approx 0.88$, indicated by the dashed line in the plot.
 The three methods
 perform closely, with a slight advantage of GMM, as expected since
-it corresponds to the true model of the data, and
+it is a consistent model to the data, and 
 $\mathcal{E}^{1D}$-clustering performs slightly better
 than $k$-means. On the other hand, in Fig.~\ref{fig:1d}b
-we consider two clusters with lognormal distributions,
+we consider a mixture of lognormal distributions,
 \begin{equation}
 \label{eq:two_lognormal}
-\log x \sim \mathcal{N}\big( \mu_i,\sigma_i^2\big) \quad
-\mbox{with
-$\mu_1 = 0$,
-$\sigma_1 = 0.3$ and
-$\mu_2 = -1.5$,
-$\sigma_2 = 1.5$.}
-\end{equation}
-The optimal classification accuracy from Bayes error is $\approx 0.852$.
+x \stackrel{iid}{\sim} 
+\tfrac{1}{2} \exp\left\{\mathcal{N}\big( \mu_1,\sigma_1^2\big)\right\} 
++\tfrac{1}{2} \exp\left\{\mathcal{N}\big( \mu_2,\sigma_2^2\big)\right\}, 
+\qquad 
+\mu_1 = 1.5, 
+\sigma_1=0.3, 
+\mu_2 = 0, 
+\sigma_2 = 1.5.
+\end{equation}
+The optimal Bayes accuracy is again $\approx 0.88$.
 We can now see that $\mathcal{E}^{1D}$-clustering
@@
 The two simple experiments of Fig.~\ref{fig:1d} illustrate
-how energy statistics does not make assumptions about the distributions
-of the data, providing more flexibility and being able
-to accurately cluster data with very different distributions.
+how energy clustering is more flexible compared to $k$-means and GMM.
+
+\begin{figure}
+\begin{minipage}{0.49\textwidth}
+~~~~~\includegraphics[width=.95\textwidth]{hist_normal.pdf}\\[-1em]
+\includegraphics[width=\textwidth]{1D_normal.pdf}\\[-1em](a)
+\end{minipage}
+\begin{minipage}{0.49\textwidth}
+~~~~~\includegraphics[width=.95\textwidth]{hist_lognormal.pdf}\\[-1em]
+\includegraphics[width=\textwidth]{1D_lognormal.pdf}\\[-1em](b)
+\end{minipage}
+\caption{
+\label{fig:1d}
+$\mathcal{E}^{1D}$-clustering versus $k$-means and GMM.
+We plot the average 
+accuracy \eqref{eq:accuracy}, over $100$ Monte Carlo trials, 
+versus the number of sampled points. 
+Error bars are standard error. The dashed line indicates 
+Bayes accuracy ($\approx 0.88$ in both cases). (a) Clustering results for data normally distributed as in 
+\eqref{eq:two_normal}, where we show the histograms of each component
+above the experimental plot.
+(b) Data lognormally distributed as in \eqref{eq:two_lognormal}
+with histograms of the components in the mixture.
+}
+\end{figure}
 
@@
 solution by the well-known kernel $k$-means algorithm based
-on Lloyd's heuristic (see \cite{Dhillon2,Dhillon}), which 
+on Lloyd's heuristic \cite{Dhillon2,Dhillon}, which 
 for convenience will also be restated in the present context.
@@
 Consider the optimization problem 
-written in the form \eqref{eq:max_prob} as follows:
+\eqref{eq:max_prob} 
+written as
 \begin{equation}
@@
 
-To optimize the kernel $k$-means objective function
+To optimize kernel $k$-means objective function
 \eqref{eq:J} we remove the global term and define the function
@@
 we name $\mathcal{E}^L$-clustering to emphasize
-that we are optimizing the withn energy function $W$ 
+that we are optimizing the within energy function $W$ 
 based on Lloyd's method \cite{Lloyd}.
@@
 finds
-local solutions to the problem \eqref{eq:qcqp2}.\hspace{\fill}
+local solutions to the optimization problem \eqref{eq:qcqp2}.\hspace{\fill}
 }
@@
 denoted by $Q^{(\ell)}$. We are interested in computing the maximum 
-cost change
+change
 $\Delta Q^{j\to \ell} (x_i) \equiv Q^{(\ell)} - Q^{(j)}$, for $\ell\ne j$. 
@@
 moving $x_i$ to $\C_\ell$, otherwise we keep $x_i$ in $\C_j$.
+
+We thus propose the following algorithm.
+We start with an initial configuration for the label matrix $Z$, 
+then for each
+point $x_i$ 
+we compute the cost of moving it to another cluster $\C_\ell$, i.e.
+$\Delta Q^{j\to \ell}(x_i)$ for 
+$\ell=1,\dots,k$ with $\ell \ne j$, where $j$ denotes the index of its current
+partition, $x \in \C_j$. Hence, we choose
+\begin{equation}
+j^\star = \argmax_{\ell=1,\dotsc,k \, | \, \ell\ne j} 
+\Delta^{j \to \ell}(x_i).
+\end{equation}
+If $\Delta Q^{j \to j^\star}(x_i) > 0$ 
+we move $x_i$ to cluster $\C_{j^\star}$, otherwise 
+we keep $x_i$ in its original cluster $\C_j$. 
+This process is repeated
+until no points are assigned to new clusters. 
+The entire procedure is explicitly described in Algorithm~\ref{algo}, which we
+denote $\mathcal{E}^H$-clustering to emphasize that it is based on
+Hartigan's method.
+This method automatically ensures that the objective function is
+monotonically increasing at each iteration, and consequently the algorithm
+converges in a finite number of steps.
 
@@
 
-We thus propose an algorithm that goes as follows.
-We start with an initial configuration for the label matrix $Z$, 
-then for each
-point $x_i$ 
-we compute the cost of moving it to another cluster $\C_\ell$, i.e.
-$\Delta Q^{j\to \ell}(x_i)$ for 
-$\ell=1,\dots,k$ with $\ell \ne j$, where $j$ denotes the index of its current
-partition, $x \in \C_j$. Hence, we choose
-\begin{equation}
-j^\star = \argmax_{\ell=1,\dotsc,k \, | \, \ell\ne j} 
-\Delta^{j \to \ell}(x_i).
-\end{equation}
-If $\Delta Q^{j \to j^\star}(x_i) > 0$ 
-we move $x_i$ to cluster $\C_{j^\star}$, otherwise 
-we keep $x_i$ in its original cluster $\C_j$. 
-This process is repeated
-until no points are assigned to new clusters. 
-The entire procedure is explicitly described in Algorithm~\ref{algo}, which we
-denote $\mathcal{E}^H$-clustering to emphasize that it is based on
-Hartigan's method.
-This method automatically ensures that the objective function is
-monotonically increasing at each iteration, and consequently the algorithm
-converges in a finite number of steps.
-
 The complexity analysis of $\mathcal{E}^H$-clustering is the following.
@@
 improve on a local optima of $\mathcal{E}^H$-clustering. On the other hand,
-$\mathcal{E}^H$-clustering might improve on a local optima of 
-$\mathcal{E}^L$-clustering. Lloyd's method forms Voronoi partitions,
+$\mathcal{E}^H$ might improve on a local optima of 
+$\mathcal{E}^L$. Lloyd's method forms Voronoi partitions,
 while Hartigan's method groups data
@@
 $k$ increases.
-The empirical results of \citet{Telgarsky} show that 
+The empirical results of \cite{Telgarsky} show that 
 an implementation of Hartigan's method has comparable execution time 
@@
 are proposing
-in $\mathcal{E}^{H}$-clustering.
-
-% @gui: if you refer to a manuscript, it is not usually acceptable to just write the number. for example, sometimes the numbers are superscripts.  citet should work.
-In \citet{Slonin}, Hartigan's method was applied to $k$-means problem
+in $\mathcal{E}^{H}$-clustering algorithm.
+
+In \cite{Slonin}, Hartigan's method was applied to $k$-means problem
 with any Bregman divergence. It was shown that the number of Hartigan's
@@
 be supported by our numerical experiments in the next section where 
-$\mathcal{E}^H$-clustering  outperforms
-of $\mathcal{E}^L$-clustering (kernel $k$-means) in several settings,
+$\mathcal{E}^H$  outperforms
+of $\mathcal{E}^L$ (kernel $k$-means) in several settings,
 specially in high dimensions.
@@
 The main goal of this section is threefold. First, we want to compare
-energy clustering, based on the kernels provided by the standard energy
-distance $\mathcal{E}_\alpha$, to $k$-means and GMM. Second, we want
+$\mathcal{E}^H$-clustering in Euclidean space to 
+$k$-means and GMM. Second, we want
 to compare $\mathcal{E}^H$-clustering, based on Hartigan's method, to
-$\mathcal{E}^L$-clustering, based on Lloyd's method. We also compare
-to $\mathcal{E}^S$-clustering
-which is an exact solution to the relaxed problem.
+$\mathcal{E}^L$-clustering or kernel $k$-means, based on Lloyd's method,
+and also to 
+spectral clustering, when they all operate on the same kernel.
 Third, we want to illustrate the flexibility
-provided by energy clustering which do not make assumptions about the
-distribution of the data, being able to cluster accurately in different
-settings while keeping the same standard kernel.
-
-In the experiments below we fix the semimetric 
-according to the traditional energy distance, $\alpha=1$, and
-the point $x_0=0$ is chosen in the corresponding kernel  
-\eqref{eq:kernel_semimetric}. We thus have
-\begin{equation}
-\label{eq:standard_metric}
-\rho(x,y) = \| x-y\| \qquad \mbox{and} \qquad \kk(x,y) = 
-\tfrac{1}{2}\left( \| x \| + \| y \| - \| x-y \| \right).
-\end{equation}
-We shall consider other kernels as well, 
-but \eqref{eq:standard_metric} will be the
-standard kernel for energy statistics and will always be present in the
-experiments as a reference.
+provided by energy clustering, which is able to cluster accurately in 
+different settings while keeping the same kernel.
+
+The following experimental setup holds unless specified
+otherwise. 
+We consider
+$\mathcal{E}^H$-clustering, $\mathcal{E}^L$-clustering 
+and spectral clustering with the following semimetrics and corresponding
+generating kernels:
+\begin{align}
+\rho_{\alpha}(x,y) &= \| x-y \|^{\alpha}, 
+&\kk_{\alpha}(x,y) &= \tfrac{1}{2} \left( \| x \|^{\alpha} + \| y \|^{\alpha} 
+- \| x-y \|^{\alpha} \right), \label{eq:rho_alpha} \\
+%
+\widetilde{\rho}_{\sigma}(x,y) &= 2 - 2 e^{-\tfrac{\|x-y\|}{2 \sigma}},  
+&\widetilde{\kk}_{\sigma}(x,y) &= e^{-\tfrac{\|x-y\|}{2\sigma}}, 
+\label{eq:rho_tilde}\\
+%
+\widehat{\rho}_{\sigma}(x,y) &= 2 - 2 e^{-\tfrac{\|x-y\|^2}{2 \sigma^2}}, 
+&\widehat{\kk}_{\sigma}(x,y) &= e^{-\tfrac{\|x-y\|^2}{2\sigma^2}}  .
+\label{eq:rho_hat}
+\end{align}
+The relation between kernel and semimetric is given by 
+formula \eqref{eq:kernel_semimetric} where we fix $x_0=0$.
+The standard $\rho_1$, from the original energy distance \eqref{eq:energy}, 
+will always be present in the experiments 
+as a reference, being the implied choice unless explicitly mentioned.
+For $k$-means, GMM and spectral clustering we use the robust 
+implementations of \emph{scikit-learn} library \citep{scikit-learn}, where  
+$k$-means is initialized with 
+$k$-means++ \citep{Vassilvitskii}, 
+and GMM with the output of $k$-means, making it more robust and preventing
+it from breaking in high dimensions. 
+Spectral clustering implementation is based on \citep{Malik}. 
+We implemented $\mathcal{E}^L$-clustering as described
+in Algorithm~\ref{kmeans_algo}, and $\mathcal{E}^H$-clustering
+as described in Algorithm~\ref{algo}. Both
+will also be initialized with $k$-means++.
+We run the algorithms $5$ times with different initializations, picking
+the result with best objective function value. 
+We evaluate clustering quality by
+the accuracy \eqref{eq:accuracy} 
+based on the true labels. For each setting we show 
+the average accuracy over $100$ Monte
+Carlo trials, with error bars indicating standard error.
 
@@
 for several univariate distributions.
-Both perform almost indistinguishable regarding the clustering quality.
+Both perform very closely.
 However, we omit these results since we will analyse more interesting 
@@
 keeping the number of points in each cluster fixed.
-Consider
-two clusters with multivariate normal distributions given by
+Consider data from the Gaussian mixture
 \begin{equation}
@@
 \begin{split}
-x &\in \C_i  \sim 
-\mathcal{N}(\mu_i,\Sigma_i) \qquad (i=1,2),  \\
-\mu_1 &= (\underbrace{0,\dotsc,0}_{\times D})^\top , \quad
-\mu_2 = 0.7 \times (\underbrace{1,\dots,1}_{\times 10},
-\underbrace{0,\dots,0}_{\times (D-10)})^\top, \quad
-\Sigma_1 = \Sigma_2 = I_D.
+x  &\stackrel{iid}{\sim} 
+\tfrac{1}{2} \mathcal{N}(\mu_1,\Sigma_1) +
+\tfrac{1}{2} \mathcal{N}(\mu_2,\Sigma_2), \\
+\Sigma_1&=\Sigma_2 = I_D, \quad
+\mu_1 = (\underbrace{0,\dotsc,0}_{\times D})^\top, \quad
+\mu_2 = 0.7 (\underbrace{1,\dots,1}_{\times 10},
+\underbrace{0,\dots,0}_{\times (D-10)})^\top.
 \end{split}
@@
 accuracy of $\approx 0.86$.
-For each $D$ we generate $100$ Monte Carlo runs, 
-sampling $100$ points for each cluster.
-We apply each algorithm to the resulting dataset 
-and compute the accuracy \eqref{eq:accuracy}. The algorithms
-$\mathcal{E}^H$- and $\mathcal{E}^L$-clustering
-both use the standard
-kernel \eqref{eq:standard_metric}, as well as the spectral
-$\mathcal{E}^S$-clustering method,
-% @gui: dude, this is a crazy sentence to write. use sklearn, or have a *very* good reason not to.  
-% currently, you utterly fail to justify the difference.
-% @jovo: Removed the sentence. I have a very good reason to use E^S clustering
-% instead of the spectral clustering from sklearn library. The reason is that
-% E^S clustering is fundamentally different, being an exact solution to
-% a relaxed problem, while 
-% spectral clustering 
-% from sklearn library is, apparently, Shi-Malik's method which is
-% not directly related to the original problem.
-% We have two options. (1) We keep these plots using E^S clustering proposed 
-% before. (2) We remove E^S clustering 
-% from the paper altogether, and if you want to compare with spectral 
-% clustering at all we use sklearn library. However this will be totally 
-% disconnected from the theory. 
-and we initialize the algorithms
-with the initialization of $k$-means++ \cite{Vassilvitskii}.
-% @gui: i thought you were goign to initialize E-H & E-L with E-S?
-% @gui: you gotta clearly define E-S, with a pseudocode block, just like the others.
-in particular, which matrix are you doing an eigendecomposition on, and how you are choosing the number of dimensions to embed into.
-% @jovo: what about Algorithm 1? All your points are clearly defined there.
-% I don't think it's a good idea to initialize  E-H and E-L with E-S 
-% because of Theorem 6. If you do this initialization you will not see much
-% difference between E^L and E^H, missing the opportunity to illustrate a good
-% selling point of this paper. What we can do, is to do a separate experiment
-% comparing different initializations and see how the algorithms perform.
-For $k$-means and GMM we use the implementation of \emph{scikit-learn} library
-\cite{scikit-learn}, where $k$-means is also initialized with $k$-means++,
-while GMM is initialized with the output of $k$-means. Every algorithm
-is ran $5$ times, returning the answer with best objective function value.
-We plot the average accuracy versus $D$, with error bars indicating
-standard error.
+We sample $200$ points on each trial.
 The results are shown in Fig.~\ref{fig:gauss}a. We can see that
-$\mathcal{E}^H$- and $\mathcal{E}^S$-clustering have practically
+$\mathcal{E}^H$ and spectral clustering have practically
 the same performance, which is 
-much higher than $\mathcal{E}^L$-clustering (kernel
-$k$-means). Moreover, $\mathcal{E}^H$-clustering outperforms 
+higher than $\mathcal{E}^L$-clustering (kernel
+$k$-means). Moreover, $\mathcal{E}^H$ outperforms 
 $k$-means and GMM, where the improvement is noticeable specially in 
 high dimensions.
-%Note that in this setting $k$-means and GMM are consistent estimators,
-%however, energy clustering degrades much less as dimension increases.
-Note that while $k$-means and GMM are appropriate to this type of 
-dataset, energy clustering degrades much less as dimension increases.
-% @gui: k-means is only consistent if the prob of each class is equal.
-but you did not actually define this parameter, because you sampled 100 from each.
-so, you gotta define the parameter at least if you want to say anythng about consistency.
-also, what do you mean k-means is a consistent estimator?
-estimator of what?
-clarify.
-% @jovo: changed the phrase above. I could re-do all experiments again
-% fliping a 1/2 coin for each distribution, but I don't think it's worthy since
-% nothing will change and will just take extra time. So I just rephrased above.
+Note that in this setting $k$-means and GMM are consistent models to the data,
+however, energy clustering degrades much less as dimension increases.
+
 \begin{figure}
@@
 \label{fig:gauss}
-Comparison of energy clustering algorithms to $k$-means and GMM in 
+Comparison of $\mathcal{E}^H$-clustering, $\mathcal{E}^L$-clustering (kernel
+$k$-means), spectral clustering, $k$-means and GMM in
 high dimensional Gaussian settings. We plot the mean accuracy versus
@@
 Carlo runs.
-(a) Data distributed as in \eqref{eq:gauss1}, with Bayes
+(a) Data normally distributed as in \eqref{eq:gauss1}, with Bayes
 accuracy $\approx 0.86$, over the range $D \in [10,200]$. 
-(b) Data distributed
+(b) Data normally distributed
 as in \eqref{eq:gauss2}, with Bayes accuracy $\approx 0.95$,
@@
 i have no intuition how difficult these are.
-% @jovo: didn't understand this
 % @gui: https://www.google.com/search?q=pairs+plots&rlz=1C5CHFA_enUS731US731&source=lnms&tbm=isch&sa=X&ved=0ahUKEwjwo9u5w-DWAhXMKyYKHYkhAKMQ_AUICigB&biw=1440&bih=776
@@
 
-% @gui: for eq 59, E-H & E-S yield the same performance.  but for eq 63, E-H is much better than E-S.  do you have any idea why that is?
-% @jovo: the difference is not increadibly large, but it is a litlle. I don't
-% know why. And in Fig.3b this difference is very small, also don't know why
-We now allow the diagonal covariance of the second Gaussian distribution to
+Still for a two-class Gaussian mixture, 
+we now allow the diagonal entries of one of the covariances to
 have different values by choosing
-$(\Sigma_2)_{ii} \stackrel{iid}{\sim} \textnormal{Uni}(1,5)$ for
-$i=1,\dotsc,10$, and $(\Sigma_2)_{ii} = 1$ for $i=11,\dotsc,D$. For instance,
-we consider the fixed setting
 \begin{equation}
@@
 \begin{split}
-x & \in \C_i  \sim 
-\mathcal{N}(\mu_i,\Sigma_i) \quad (i=1,2),  \\
+x & \stackrel{iid}{\sim} 
+\tfrac{1}{2}\mathcal{N}(\mu_1,\Sigma_1)+
+\tfrac{1}{2}\mathcal{N}(\mu_2,\Sigma_2),  \\
 \mu_1 &= (\underbrace{0,\dotsc,0}_{\times D})^\top , \quad
@@
 \end{equation}
-as we vary $D$. Note that there is nothing special about the numbers in
-$\widetilde{\Sigma}_{10}$, it is simply a random choice of sampling
-$\textnormal{Uni}(1,5)$ ten times, and one could pick any other choices
-as well.
-% @gui: if there is nothing special, why not just use 2*ones?  this seems suspicious.  also, it is the result of sampling 1 time, not 10 times.
-% @jovo: This is actually less suspicious since  I'm not even choosing the 
-% numbers, just picking them at random. The point is to make the diagonal 
-% entries different, so 2*ones does not help. Also, I don't want
-% to keep trying many experiments everytime, it's super time consuming.
-% This setting works to illustrate what we want ... a setting where k-means
-% does poorly than GMM, and energy still beats GMM in high D.
-The Bayes error is kept fixed yielding
+We simply chose a fixed set of $10$ numbers uniformly at random on 
+the interval $[1,5]$ for
+the diagonal of $\widetilde{\Sigma}_{10}$, and any other choice would give 
+analogous results. We sample a total of $200$ points from \eqref{eq:gauss2}
+on each trial.
+The Bayes error is kept fixed when increasing $D$ yielding
 an optimal accuracy $\approx 0.95$.
-% @gui: bayes error changes for each random sample, unless you are integrating over all possible samples?  it is not fixed.
-% @jovo: I said explicitly "the fixed setting". 
-% It is exactly as in eq. (59) with those numbers 
-% fixed everytime, this is why I wrote the numbers explicitly. 
-% I'm not generating new numbers everytime I sample.
-In Fig.~\ref{fig:gauss}b we show the results of energy clustering,
-$k$-means, and GMM.
-Note that GMM performs better in low dimensions, 
-since it is a consistent estimator.
-However, $\mathcal{E}^H$ and $\mathcal{E}^S$ degrade much less as dimensions
-increase, performing better than $k$-means and GMM in high dimensions.
-% @gui: you gotta explain why GMM doesn't utterly fail when D>n!
-% @jovo: check below
+In Fig.~\ref{fig:gauss}b we see that GMM performs better in low dimensions, 
+but it quickly degenerates as $D$ increases. 
+The same is true for $k$-means and $\mathcal{E}^L$-clustering.
+However, $\mathcal{E}^H$ and spectral clustering remains much
+more stable in high dimensions.
 Notice that a naive implementation of GMM should not be able to 
 estimate the covariances when $D \gtrsim 100$, however, scikit-learn library
-uses $k$-means output as initialization, and a more refined estimation
-of the covariances, therefore the
-output of GMM in this implementation is at least as good as $k$-means, making
-the algorithm more robust in high dimensions.
-
-Besides the standard kernel from energy statistics 
-\eqref{eq:standard_metric}, 
-consider the following two other semimetrics with their respective generating
-kernels:
-\begin{align}
-\rho_{1/2}(x,y) &= \| x-y \|^{1/2} & 
- \kk_{1/2}(x,y) &= \tfrac{1}{2} \left( 
-\| x \|^{1/2} + \| y \|^{1/2} 
-- \| x-y \|^{1/2} \right), \label{eq:rhohalf}\\
-\rho_{e}(x,y) &= 
-2 - 2 e^{-\tfrac{1}{2}\| x- y\|} &
- \kk_{e}(x,y) &= e^{-\tfrac{1}{2}\| x-y\|}.
-\label{eq:rhoe}
-\end{align}
-The kernel $K_{1/2}(x,y)$ corresponds to the standard energy distance
-with $\alpha=1/2$.
-We sample data from the following normal distribution in $20$ dimensions:
+uses $k$-means output as initialization, therefore the
+output of GMM in this implementation is at least as good as $k$-means and
+the algorithm is more robust in high dimensions.
+
+Consider sampling data from the following Gaussian mixture in 
+$\mathbb{R}^{20}$:
 \begin{equation}
@@
 \begin{split}
-x &\in \C_i \sim \mathcal{N}(\mu_i,\Sigma_i) \qquad (i=1,2), \\
+x &\stackrel{iid}{\sim} \tfrac{1}{2} \mathcal{N}(\mu_1,\Sigma_1)+
+\tfrac{1}{2} \mathcal{N}(\mu_2,\Sigma_2) , \\
 \mu_1 &= (\underbrace{0,\dotsc,0}_{\times 20})^\top ,\quad
@@
 classification error is $\approx 0.90$. 
-We sample an equal number of points for each cluster which is progressively
-increased, $n \in [5, 200]$. 
-The results over $100$ Monte Carlo trials are 
-shown in Fig.~\ref{fig:consist}a, where we compare
-$\mathcal{E}^H$-clustering, with different
-kernels, to $k$-means and GMM.
+We increase the sample size $n \in [10,400]$ and show the accuracy versus
+$n$ for the different kernels 
+\eqref{eq:rho_alpha} and \eqref{eq:rho_tilde}
+within $\mathcal{E}^H$-clustering
+algorithm, which are compared to $k$-means and GMM. The results are
+in Fig.~\ref{fig:consist}a.
 Note that for small $n$ all methods
 are superior than GMM, which slowly catches up and tend to optimal Bayes,
-as expected since it is a consistent estimator in this case. 
+as expected since it is a consistent model to the data. 
 Note also that $\mathcal{E}^H$-clustering
-with kernel $\kk_e$ is as accurate as GMM for large number of points, however,
+with kernel $\widetilde{\kk}_{1}$ is as accurate as GMM for 
+large number of points, however,
 it is superior for small number of points. Still for
 the same setting, in Fig.~\ref{fig:consist}b we
-we show the difference in accuracy provided by $\mathcal{E}^H$ minus
-$\mathcal{E}^L$ and $\mathcal{E}^S$, when using the
-kernel $\kk_e$.
+show the difference in accuracy provided by $\mathcal{E}^H$ minus
+$\mathcal{E}^L$ and $\mathcal{E}^H$ minus spectral clustering, when using the
+kernel $\widetilde{\kk}_{1}$.
 Note that $\mathcal{E}^H$ was
-always superior than the other 
-methods (there would be points negative values on the $y$-axis otherwise).
+always superior than kernel $k$-means and spectral clustering,  
+otherwise there would be points negative values on the $y$-axis.
 
@@
 \label{fig:consist}
-$\mathcal{E}^H$-clustering with different kernels versus
+$\mathcal{E}^H$-clustering with kernels \eqref{eq:rho_alpha} and 
+\eqref{eq:rho_tilde} versus
 $k$-means and GMM. In both settings Bayes accuracy
@@
 versus number of points for $100$ Monte Carlo trials.
-(a,b) Normal distribution \eqref{eq:20gauss}. 
-(c,d) Lognormal distribution~\eqref{eq:20loggauss}.
+(a,b) Gaussian mixture \eqref{eq:20gauss}. 
+(c,d) Lognormal mixture \eqref{eq:20loggauss}.
 The plots in (c) and (d) consider the difference in accuracy
-between $\mathcal{E}^H$ versus $\mathcal{E}^L$ and
-$\mathcal{E}^S$-clustering, with semimetric $\rho_e$.
+between $\mathcal{E}^H$ versus $\mathcal{E}^L$ (kernel $k$-means) and
+spectral clustering, with the kernel $\widetilde{K}_1$.
 }
@@
 
-Now consider the same parameters as in \eqref{eq:20gauss} but with
-lognormal distributions, 
+Consider the same experiment but now with a
+lognormal mixture, 
 \begin{equation}
 \label{eq:20loggauss}
-\log x \in \C_i \sim \mathcal{N}(\mu_i, \Sigma_i) \qquad (i=1,2).
-\end{equation}
-The same previous experiment is shown in 
-Fig.~\ref{fig:consist}c.
+\begin{split}
+x &\stackrel{iid}{\sim} \tfrac{1}{2} 
+\exp\left\{\mathcal{N}(\mu_1,\Sigma_1)\right\}+
+\tfrac{1}{2} \exp\left\{\mathcal{N}(\mu_2,\Sigma_2)\right\} , \\
+\mu_1 &= (\underbrace{0,\dotsc,0}_{\times 20})^\top ,\quad
+\mu_2 = \tfrac{1}{2} 
+(\underbrace{1,\dotsc,1}_{5},\underbrace{0,\dotsc,0}_{15})^\top, \quad
+\Sigma_1 = \tfrac{1}{2} I_{20},  \quad
+\Sigma_2 = I_{20}.
+\end{split}
+\end{equation}
+The results are in Fig.~\ref{fig:consist}c.
 Energy clustering still performs accurately, with any of the utilized
 kernels, providing better results than $k$-means and GMM on
-this non-normal data. The kernel $\kk_e$ still provides the best results
-for small number of points, but its performance is eventually achieved by
+this non-normal data. The kernel $\widetilde{\kk}_1$ still provides 
+the best results for small number of points, 
+but its performance is eventually achieved by
 $\kk_{1/2}$, indicating that $\alpha \approx 1/2$ in the standard
 energy distance should be more appropriate for skewed distributions.
-Still for the same setting, in Fig.~\ref{fig:consist}d
-we show the difference in accuracy provided by $\mathcal{E}^H$ versus
-$\mathcal{E}^L$ and $\mathcal{E}^S$, for the kernel $\kk_e$. Again,
+In Fig.~\ref{fig:consist}d we show the difference between 
+$\mathcal{E}^H$-clustering to kernel $k$-means and spectral clustering,
+with the kernel $\widetilde{\kk}_1$. Again,
 the accuracy provided by $\mathcal{E}^H$ is higher than the other methods,
-although not much higher than $\mathcal{E}^S$ in this example.
+although not much higher than spectral clustering in this example.
 The two experiments of Fig.~\ref{fig:consist}
@@
 performing well in different settings
-with the same kernel. 
-% @gui: algs don't make assumptions.
-% @jovo: see if above is ok, just said flexible instead
+with the same kernel, contrary to $k$-means and GMM.
 
@@
 
-Consider the following choices of semimetric and 
-corresponding generating kernel:
-\begin{align}
-\rho_{\alpha}(x,y) &= \| x - y\|^\alpha & 
-\kk_{\alpha}(x,y) &= \tfrac{1}{2}\left(
-\| x \|^\alpha +
-\| y \|^\alpha -
-\| x-y \|^\alpha \right) ,
-\label{eq:rhoalpha} \\
-%
-\widetilde{\rho}_{\sigma}(x,y) &= 2 - 2 e^{-\tfrac{\|x-y\|}{2 \sigma}} &
-\widetilde{\kk}_{\sigma}(x,y) &= e^{-\tfrac{\|x-y\|}{2\sigma}} ,
-\label{eq:rhoesigma} \\
-%
-\widehat{\rho}_{\sigma}(x,y) &= 2 - 2 e^{-\tfrac{\|x-y\|^2}{2 \sigma^2}} &
-\widehat{\kk}_{\sigma}(x,y) &= e^{-\tfrac{\|x-y\|^2}{2\sigma^2}} .
-\label{eq:rhogsigma} 
-\end{align}
 In Fig.~\ref{fig:other}a--c we have 
@@
 in (b) and (c) have $400$ points for each class.
-We apply $\mathcal{E}^H$-clustering  with different choices
-of kernels in \eqref{eq:rhoalpha}--\eqref{eq:rhogsigma}. We also consider
-the best kernel choice for each example for the spectral
-$\mathcal{E}^S$-clustering. 
+We apply $\mathcal{E}^H$-clustering  with the 
+kernels \eqref{eq:rho_alpha}, \eqref{eq:rho_tilde} and \eqref{eq:rho_hat}. 
+We also consider the best kernel choice for each example for spectral
+clustering. 
 Moreover, we consider
@@
 in every example, and in particular better than the
-spectral $\mathcal{E}^S$.
-For the data in Fig.~\ref{fig:other}a 
-the semimetrics $\rho_1$ and $\rho_{1/2}$ are
-still able to provide accurate results, however, for the
-examples in (b) and (c) the kernel choice is more sensitive.
-
-\begin{table}[t]
-\renewcommand*{\arraystretch}{0.75}
-\begin{tabular}{@{}r  l g  l g  l g@{}}
+spectral clustering.
+In (a) 
+the standard kernel from energy statistics in Euclidean
+space, $\kk_1$ and $\kk_{1/2}$, are
+able to provide accurate results, however, for the
+examples in (b) and (c) the kernel choice is more sensitive, where
+$\widehat{\kk}_1$ and $\widehat{\kk}_2$ provide a significant improvement.
+
+\begin{table*}
+\caption{\label{table:other}
+Clustering data from Fig.~\ref{fig:other}a--c.
+}
+\begin{center}
+\footnotesize{
+\begin{tabular}{@{}r  l l  l l  l l@{}}
 \toprule[1pt]
@@
 \midrule[0.5pt]
-\multirow{4}{*}{\emph{$\mathcal{E}^H$~~~~}}
+\multirow{4}{*}{\emph{$\mathcal{E}^H$-clustering~~~~}}
 & $\rho_{1}$ & $0.705\pm 0.065$
@@
 & $\widehat{\rho}_{2}$ & $\bm{0.676\pm 0.002}$ \\
-\arrayrulecolor{gray!80}\midrule[0.5pt]
-\emph{$\mathcal{E}^S$~~~~}
+\midrule[0.5pt]
+\emph{spectral-clustering~~~~}
 & $\widetilde{\rho}_{2}$ & $0.557\pm 0.014$ 
@@
 & $\widehat{\rho}_{2}$ & $0.364\pm 0.004$  \\
-\arrayrulecolor{gray!80}\midrule[0.5pt]
 \emph{$k$-means}~~~~ 
-& & $0.550\pm 0.011$
-& & $0.522\pm 0.004$
-& & $0.368\pm 0.005$ \\
+& \xmark & $0.550\pm 0.011$
+& \xmark & $0.522\pm 0.004$
+& \xmark & $0.368\pm 0.005$ \\
 \emph{GMM}~~~~
-& & $0.903\pm 0.064$
-& & $0.595\pm 0.011$
-& & $0.465\pm 0.030$ \\
-\arrayrulecolor{black}\bottomrule[1pt]
+& \xmark & $0.903\pm 0.064$
+& \xmark & $0.595\pm 0.011$
+& \xmark & $0.465\pm 0.030$ \\
+\bottomrule[1pt]
 \end{tabular}
-\caption{\label{table:other}
-Clustering data from Fig.~\ref{fig:other}.
-We show average
-accuracy over $10$ Monte Carlo runs.
 }
-\end{table}
-
-\begin{table}[t]
-\renewcommand*{\arraystretch}{0.75}
-\begin{tabular}{@{}r@{}l@{}|@{}l@{}|@{}l@{}|@{}l@{}|@{}l@{}}
+\end{center}
+\end{table*}
+
+\begin{table*}
+\caption{\label{table:mnist}
+Clustering MNIST data from Fig.~\ref{fig:other}d.
+}
+\begin{center}
+\footnotesize{
+\begin{tabular}{@{}r@{}l@{}@{}l@{}@{}l@{}@{}l@{}l@{}}
 \toprule[1pt]
 \emph{Class Subset}~~~~ & & 
-$~\{0,1,2\}$ &
+%$~\{0,1,2\}$ &
 $~\{0,1,\dotsc,4\}~$ &
 $~\{0,1,\dotsc,6\}~$ &
-$~\{0,1,\dotsc,8\}~$ \\
+$~\{0,1,\dotsc,8\}~$ &
+$~\{0,1,\dotsc,9\}~$ \\
+\emph{parameter}~~~~ & {$\sigma$}
+%& $~10.34~$
+& $~10.41~$
+& $~10.41~$
+& $~10.37~$
+& $~10.19~$ \\
 \midrule[0.5pt]
@@
 &$~0.657\pm 0.014~$
+&$~0.584\pm 0.013~$
 \\
 & $\widehat{\rho}_{\sigma}$ 
-&$~0.933\pm 0.005~$
-&$~0.891\pm 0.009~$
-&$~0.759\pm 0.011~$
-&$~0.704\pm 0.011~$
+%&$~0.933\pm 0.005~$
+&$~\bm{0.891\pm 0.009}~$
+&$~\bm{0.759\pm 0.011}~$
+&$~\bm{0.704\pm 0.011}~$
+&$~\bm{0.591\pm 0.012}~$
 \\
-\arrayrulecolor{gray!80}\midrule[0.5pt]
-\emph{$\mathcal{E}$-spectral~~~~} 
+\midrule[.5pt]
+\emph{spectral-clustering~~~~} 
 & $\widehat{\rho}_{\sigma}$ 
-&$~0.823\pm 0.015~$
+%&$~0.823\pm 0.015~$
 &$~0.769\pm 0.012~$
@@
 &$~0.649\pm 0.018~$
+&$~0.565\pm 0.009~$
 \\
-\midrule[0.5pt]
-\rowcolor{gray!20}
-& {$\sigma$}
-& $~10.34~$
-& $~10.41~$
-& $~10.41~$
-& $~10.37~$
-\\
-\arrayrulecolor{gray!80}\midrule[0.5pt]
-\emph{$k$-means}~~~~ &
-&$~0.927\pm 0.004~$
+\emph{$k$-means}~~~~ & \xmark
+%&$~0.927\pm 0.004~$
 &$~0.878\pm 0.010~$
@@
 &$~0.695\pm 0.012~$
+&$~0.557\pm 0.012~$
 \\
-\arrayrulecolor{gray!80}\midrule[0.5pt]
-\emph{GMM}~~~~ &
-&$~0.952\pm 0.005~$
+\emph{GMM}~~~~ & \xmark
+%&$~\bm{0.952\pm 0.005}~$
 &$~0.839\pm 0.015~$
@@
 &$~0.621\pm 0.009~$
+&$~0.540\pm 0.009~$
 \\
-\arrayrulecolor{black}\bottomrule[1pt]
+\bottomrule[1pt]
 \end{tabular}
-\caption{\label{table:mnist}
-Clustering MNIST data from Fig.~\ref{fig:other}d.
-For each subset of digits we perform $10$ Monte Carlo runs, sampling 
-$100$ points for  each class.
 }
-\end{table}
+\end{center}
+\end{table*}
 
@@
 corresponding to the digits $\{0,1,\dotsc,9 \}$.
-Besides the kernel \eqref{eq:rhoalpha}, we
-consider the Gaussian kernel \eqref{eq:rhogsigma} with 
+We compute the parameter
 \begin{equation}
@@
 \end{equation}
-computed from a separate
-training set.
+from a separate
+training set, to be used in the kernels \eqref{eq:rho_tilde} and 
+\eqref{eq:rho_hat}.
 We consider subsets of $\{0,1,\dotsc,9 \}$, 
@@
 for each class. 
-% @gui: i don't understand these results. {0,1,2} we do worse.  why don't we show {0,1,...,9}?
-% also, why don't you use a better metric/similarity designed for images? like using pairwise alignment followed by rho?
-% @jovo: We can't claim we do worse on {0,1,2}, but also don't do better.
-% I did the same experiment as Sapiro, and didn't fit in the table this is
-% why I didn't include {0,...,9}, but I can do that.
-% To treat the images better I would need to do some pre-processing, which
-% can be time consuming, and you know how much I love manipulating images.
-% I don't have any robust code for this, and anything I try to
-% implement will take more than a week to do. I will check if I find something
-% that works out of the box in python though.
 The results are shown in Table~\ref{table:mnist}, where kernels
@@
 
-We considered clustering from the perspective of energy
-statistics theory.
-The mathematical formulation of energy clustering 
+We proposed clustering from the perspective of generalized energy
+statistics, valid for arbitrary spaces of negative type.
+Our mathematical formulation of energy clustering 
 reduces to a QCQP in the associated RKHS, as demonstrated in 
@@
 more general kernels 
-on semimetric spaces of negative type can also be obtained.
+on spaces of negative type can also be obtained.
 We also considered a weighted version of energy statistics, whose 
@@
 on several settings, illustrating the flexibility
-of the proposed method which is distribution-free. In many
+of the proposed method which is model-free. In many
 settings, the iterative
-$\mathcal{E}^H$-clustering also surpassed $\mathcal{E}^S$-clustering, which
-is a spectral relaxation of the QCQP.
+$\mathcal{E}^H$-clustering also surpassed spectral clustering, which
+is solves a relaxation of the original QCQP, and in other settings performed
+closely but never worse. Note that spectral clustering is more expensive
+than our iterative method, going up to $\mathcal{O}(n^3)$, and finding
+eigenvectors of very large matrices is problematic.
 
 A limitation of the proposed methods for energy clustering is that it cannot
-handle accurately highly unbalanced clusters. As an illustration
-consider the following setting:
+handle accurately highly unbalanced clusters. As an illustration,
+consider the following Gaussian mixture:
 \begin{equation}
@@
 \begin{split}
-x &\in \C_i \sim  
-\dfrac{n_i}{N} \mathcal{N}(\mu_i,\Sigma_i) \quad (i=1,2), \quad 
+x &\stackrel{iid} \sim  
+\dfrac{n_1}{2N} \mathcal{N}(\mu_1,\Sigma_1)+
+\dfrac{n_1}{2N} \mathcal{N}(\mu_1,\Sigma_1), \quad 
 \mu_1 = (0,0,0,0)^\top , \quad
@@
 \end{array}\right), \quad
-n_1 = N - m, \quad  n_2 = N + m, \quad N=250.
+n_1 = N - m, \quad  n_2 = N + m, \quad N=300.
 \end{split}
@@
 
-Moreover, it would also be interesting to formally 
+Moreover, it would be interesting to formally 
 demonstrate cases where energy clustering is a 
-consistent estimator. A soft version of energy clustering is also an
+consistent in the large $n$ limit. 
+A soft version of energy clustering is also an
 interesting extension.
@@
 We would like to thank Carey Priebe for discussions.
-This work was supported by NIH TRA grant.
+We would like to acknowledge the support of the Transformative
+Research Award (NIH \#R01NS092474) and  the Defense Advanced Research Projects
+Agency’s (DARPA) SIMPLEX program through SPAWAR contract N66001-15-C-4041.
 \end{acknowledgements}
@@
 \bibliographystyle{unsrt}
+%\bibliographystyle{abbrvnat}
 \bibliography{biblio.bib}

### 2017-10-24 energy_clustering/code/experiment2_2.py

@@ -115,11 +115,11 @@ def make_plot(*data_files):
 
     ## customize plot below ##
     p = plot.ErrorBar()
-    #p.xlabel = r'$\#$ dimensions'
-    p.xlabel = ''
-    p.ylabel = ''
+    p.xlabel = r'$\#$ dimensions'
+    p.ylabel = 'accuracy'
     p.legends = [r'$\mathcal{E}^{H}$', 
-                 r'kernel', 
+                 #r'kernel', 
+                 r'$\mathcal{E}^L$', 
                  r'spectral', 
                  r'$k$-means', 
                  r'GMM']
@@ -128,7 +128,7 @@ def make_plot(*data_files):
     p.symbols = ['o', 's', 'D', '^', 'v']
     p.lines = ['-', '-', '-', '-', '-']
     #p.output = './experiments_figs2/normal_highdim_mean.pdf'
-    p.output = './experiments_figs2/normal_highdim_cov.pdf'
+    p.output = './experiments_figs/normal_highdim_cov.pdf'
     #p.bayes = 0.86
     p.bayes = 0.9537075
     #p.xlim = [10, 200]

### 2017-10-24 energy_clustering/code/experiment3_2.py

@@ -6,6 +6,8 @@ Using mixtures now.
 # Guilhere Franca <guifranca@gmail.com>
 # Johns Hopkins University, Neurodata
 
+from __future__ import division
+
 import numpy as np
 import multiprocessing as mp
 
@@ -20,7 +22,7 @@ def gauss_dimensions_pi(num_points=range(0, 180, 10), num_experiments=10):
     k = 2
     D = 4
     d = 2
-    N = 250
+    N = 300
     table = np.zeros((num_experiments*len(num_points), 6))
     count = 0
 
@@ -32,9 +34,8 @@ def gauss_dimensions_pi(num_points=range(0, 180, 10), num_experiments=10):
             s1 = np.eye(D)
             m2 = np.concatenate((1.5*np.ones(d), np.zeros(D-d)))
             s2 = np.diag(np.concatenate((.5*np.ones(d), np.ones(D-d))))
-            pi1 = (N-p)/N
-            pi2 = (N+p)/N
-            print pi1, pi2
+            pi1 = (N-p)/N/2.
+            pi2 = (N+p)/N/2.
             n1, n2 = np.random.multinomial(N, [pi1,pi2])
             X, z = data.multivariate_normal([m1, m2], [s1, s2], [n1, n2])
             rho = lambda x, y: np.linalg.norm(x-y)
@@ -66,10 +67,11 @@ def make_plot(*data_files):
 
     ## customize plot below ##
     p = plot.ErrorBar()
-    p.xlabel = 'number of unbalanced points'
+    p.xlabel = r'$\#$ unbalanced points'
+    p.ylabel = r'accuracy'
     p.legends = [r'$\mathcal{E}^{H}$', 
                  r'$\mathcal{E}^{L}$', 
-                 r'$\mathcal{E}^{S}$', 
+                 r'spectral', 
                  r'$k$-means', 
                  r'GMM']
     p.colors = ['b', 'r', 'g', 'm', 'c']
@@ -102,5 +104,5 @@ def worker(dimensions, fname):
 ###############################################################################
 if __name__ == '__main__':
     fname = './experiments_data2/experiment_unbalanced_%i.csv'
-    gen_data(fname)
-    #make_plot(fname%0, fname%1, fname%2, fname%3)
+    #gen_data(fname)
+    make_plot(fname%0, fname%1, fname%2, fname%3)

### 2017-10-26 energy_clustering/draft_v2/energy_clustering.tex
draft v2

@@
 need to be carefully chosen. A well-known kernel based clustering method
-is kernel $k$-means, which is preciselly $k$-means 
+is kernel $k$-means, which is precisely $k$-means 
 formulated in the feature space \cite{Girolami}. 
@@
 mathematical formulation to this statement, leading to a quadratically 
-constrained quadratic program (QCQP) in the associated RKHS. This immediatelly
+constrained quadratic program (QCQP) in the associated RKHS. This immediately
 establishes the connection between energy statistics based clustering,
@@
 
-% @gui: to provide intuition to the reader, let's show the kernel density estimate for each of the 3 algorithms for both classes.
-i'm not sure the best way to visualize it, maybe play with a few options.
-% @jovo: I didn't do this yet and actually not sure how to do it.
-% Are you thinking about this: 
-% https://en.wikipedia.org/wiki/Kernel_density_estimation
-% Also, I'm not sure if the kernel from energy integrates to 1, etc. and is
-% this really worthy it?
-% @gui: yes, it is really worth it, that is why i suggested it. it is trivial to do.  currently, the reader has no idea *why* the different algs get different answers.
-
-% @gui: why not use the same parameters as in fig 1a?  i feel like if we did, it would seem less like we are cherry picking?  
-% @jovo: yes, and GMM would do great! In 1D is very hard to find settings where
-% GMM and k-means perform poorly, so we have to cherry pick a little. In 1D you
-% basically do not have much choice, if data has minimal separation k-means and
-% GMM do well. The purpose is to
-% illustrate that there are settings where k-means/GMM do poorly and energy
-% does not. If I choose the same parameters as before all algorithms are doing
-% great, sometimes GMM can even do better than energy. All I did was to
-% translate one of the lognormals mu2=-1.5 and give it a big variance, so both
-% distributions overlap a lot, but the overlaping is essentially on the tail 
-% of one of them
-% @gui: ok, well why not use the same params from fig 1b in fig 1a then?
-
 Before stating a general algorithm to solve the optimization problem
@@
 $\mathcal{E}^{1D}$-clustering performs slightly better
-than $k$-means. On the other hand, in Fig.~\ref{fig:1d}b
+than $k$-means. 
+In Fig.~\ref{fig:1d}c we show a density estimation from clustering
+$1000$ points from this mixture using the three algorithms.
+Notice that all of them are able to distinguish the two classes.
+On the other hand, in Fig.~\ref{fig:1d}b
 we consider a mixture of lognormal distributions,
@@
 GMM and $k$-means basically cluster at chance.
-The two simple experiments of Fig.~\ref{fig:1d} illustrate
+Density estimation after clustering $1000$ points this
+mixture using the three algorithms 
+are
+is shown in Fig.~\ref{fig:1d}d. Note that only $\mathcal{E}^{1D}$-clustering
+was able to distinguish the two classes. $k$-means and GMM put most of the
+points in a single cluster, and points on the tail of the second component
+of \eqref{eq:two_lognormal} in the other cluster.
+The experiments of Fig.~\ref{fig:1d} illustrate
 how energy clustering is more flexible compared to $k$-means and GMM.
@@
 \begin{minipage}{0.49\textwidth}
-~~~~~\includegraphics[width=.95\textwidth]{hist_normal.pdf}\\[-1em]
-\includegraphics[width=\textwidth]{1D_normal.pdf}\\[-1em](a)
+\includegraphics[width=\textwidth]{1D_normal.pdf}\\[-1em](a)\\
+~~~~~\includegraphics[width=.9\textwidth]{normal_density.pdf}\\[-1.5em](c)
 \end{minipage}
 \begin{minipage}{0.49\textwidth}
-~~~~~\includegraphics[width=.95\textwidth]{hist_lognormal.pdf}\\[-1em]
-\includegraphics[width=\textwidth]{1D_lognormal.pdf}\\[-1em](b)
+\includegraphics[width=\textwidth]{1D_lognormal.pdf}\\[-1em](b)\\
+~~~~~\includegraphics[width=.9\textwidth]{lognormal_density.pdf}\\[-1.5em](d)
 \end{minipage}
@@
 $\mathcal{E}^{1D}$-clustering versus $k$-means and GMM.
-We plot the average 
-accuracy \eqref{eq:accuracy}, over $100$ Monte Carlo trials, 
+(a,b) We plot the mean accuracy
+\eqref{eq:accuracy} over $100$ Monte Carlo trials, 
 versus the number of sampled points. 
@@
 Bayes accuracy ($\approx 0.88$ in both cases). (a) Clustering results for data normally distributed as in 
-\eqref{eq:two_normal}, where we show the histograms of each component
-above the experimental plot.
-(b) Data lognormally distributed as in \eqref{eq:two_lognormal}
-with histograms of the components in the mixture.
+\eqref{eq:two_normal}.
+(b) Data lognormally distributed as in \eqref{eq:two_lognormal}.
+(c) Density estimation of each component in the mixture \eqref{eq:two_normal} 
+after clustering $1000$ sampled points using the three algorithms, compared
+to the ground truth. 
+(d) The same but for lognormal data \eqref{eq:two_lognormal}.
 }
@@
 as described in Algorithm~\ref{algo}, to
-$\mathcal{E}^{1D}$-clutering, described in Algorithm~\ref{algo1d}, 
+$\mathcal{E}^{1D}$-clustering, described in Algorithm~\ref{algo1d}, 
 for several univariate distributions.
@@
 scenarios in high dimensions.
+
+\begin{figure}
+\begin{minipage}{0.49\textwidth}
+\includegraphics[width=1\textwidth]{pairsplot1.pdf}\\[-1.0em] (a)
+\end{minipage}
+\begin{minipage}{0.49\textwidth}
+\includegraphics[width=1\textwidth]{pairsplot2.pdf}\\[-1.0em] (a)
+\end{minipage}
+\caption{
+\label{fig:pairsplot}
+Pair plots for the first $5$ dimensions. (a) Data normally distributed
+as in \eqref{eq:gauss1}. (b) Data normally distributed as in 
+\eqref{eq:gauss2}. We sample $200$ points for both cases.
+We can see that there is a considerable overlap between the clusters.
+}
+\end{figure}
 
@@
 \end{equation}
+To get some intuition about how separated data points from each class
+are, we show scatter plots between the first $5$ dimensions in 
+Fig.~\ref{fig:pairsplot}a.
 Note that the Bayes error
@@
 over the range $D \in [10, 700]$.
-% @gui: also helpful here would be pairs plots, say, of the first 5 dimensions, showing the performance of our stuff (color points by Alg 3, maybe symbols denote "truth".
-i have no intuition how difficult these are.
-% @gui: https://www.google.com/search?q=pairs+plots&rlz=1C5CHFA_enUS731US731&source=lnms&tbm=isch&sa=X&ved=0ahUKEwjwo9u5w-DWAhXMKyYKHYkhAKMQ_AUICigB&biw=1440&bih=776
-% to show how well separated the classes are.
 }
@@
 \widetilde{\Sigma}_{10} &= \diag(1.367,  3.175,  3.247,  4.403,  1.249,
-1.969, 4.035,   4.237,  2.813,  3.637),
+1.969, 4.035,   4.237,  2.813,  3.637).
 \end{split}
@@
 the diagonal of $\widetilde{\Sigma}_{10}$, and any other choice would give 
-analogous results. We sample a total of $200$ points from \eqref{eq:gauss2}
+analogous results. 
+We show pair plots of this data in Fig.~\ref{fig:pairsplot}b.
+We sample a total of $200$ points from \eqref{eq:gauss2}
 on each trial.

### 2018-02-09 energy_clustering/icml18/energy_clustering.tex
energy icml18

--- 2017-10-13 energy_clustering/aistats18/energy_clustering.tex
@@
-\documentclass[twoside]{article}
-\usepackage{aistats2018}
- % If your paper is accepted, change the options for the package
-% aistats2018 as follows:
-%
-%\usepackage[accepted]{aistats2018}
-%
-% This option will print headings for the title of your paper and
-% headings for the authors names, plus a copyright note at the end of
-% the first column of the first page.
-
-\usepackage[compress,round]{natbib}
+%%%%%%%% ICML 2018 EXAMPLE LATEX SUBMISSION FILE %%%%%%%%%%%%%%%%%
+
+\documentclass{article}
+
+% Recommended, but optional, packages for figures and better typesetting:
+\usepackage{microtype}
+\usepackage{graphicx}
+\graphicspath{{./figs/}}
+\usepackage{subfigure}
+\usepackage{booktabs} % for professional tables
+
 \usepackage{amsmath,amssymb,amsfonts,amsthm,amscd,bm,bbm}
 \usepackage{mathtools}
-\mathtoolsset{showonlyrefs}
+
+% hyperref makes hyperlinks in the resulting PDF.
+% If your build breaks (sometimes temporarily if a hyperlink spans a page)
+% please comment out the following usepackage line and replace
+% \usepackage{icml2018} with \usepackage[nohyperref]{icml2018} above.
+\usepackage{hyperref}
 \usepackage[inline]{enumitem}
-\usepackage[inline]{enumitem}
-\usepackage[pdftex]{graphicx}
-\graphicspath{{./figs/}}
-\usepackage{algorithmic}
-\usepackage{algorithm}
 \usepackage{multirow}
@@
 \usepackage{pifont}
-\newcommand{\cmark}{\ding{51}}
-\newcommand{\xmark}{\ding{55}}
+
+% Attempt to make hyperref and algorithmic work together better:
+\newcommand{\theHalgorithm}{\arabic{algorithm}}
+
+% Use the following line for the initial blind version submitted for review:
+%\usepackage{icml2018}
+
+% If accepted, instead use the following line for the camera-ready submission:
+\usepackage[accepted]{icml2018}
+
+% The \icmltitle you define below is probably too long as a header.
+% Therefore, a short form for the running title is supplied here:
+\icmltitlerunning{Clustering via Generalized Energy Statistics}
 
@@
 
+\newcommand{\cmark}{\ding{51}}
+\newcommand{\xmark}{\ding{55}}
+
 \newcommand\Energy{\mathcal{E}}
@@
 
-% If your paper is accepted and the title of your paper is very long,
-% the style will print as headings an error message. Use the following
-% command to supply a shorter title of your paper so that it can be
-% used as headings.
-%
-%\runningtitle{I use this title instead because the last one was very long}
-
-% If your paper is accepted and the number of authors is large, the
-% style will print as headings an error message. Use the following
-% command to supply a shorter version of the authors names so that
-% they can be used as headings (for example, use only the surnames)
-%
-%\runningauthor{Surname 1, Surname 2, Surname 3, ...., Surname n}
-
 \twocolumn[
-
-\aistatstitle{Clustering via Generalized Energy Statistics}
-
-\aistatsauthor{ Author 1 \And Author 2 \And  Author 3 }
-
-\aistatsaddress{ Institution 1 \And  Institution 2 \And Institution 3 }
-
-%\aistatsauthor{Guilherme Fran\c{c}a \And  Joshua T. Vogelstein}
-
-%\aistatsaddress{ Johns Hopkins University }
-
+\icmltitle{Clustering via Generalized Energy Statistics}
+
+% It is OKAY to include author information, even for blind
+% submissions: the style file will automatically remove it for you
+% unless you've provided the [accepted] option to the icml2018
+% package.
+
+% List of affiliations: The first argument should be a (short)
+% identifier you will use later to specify author affiliations
+% Academic affiliations should list Department, University, City, Region,
+% Country
+% Industry affiliations should list Company, City, Region, Country
+
+% You can specify symbols, otherwise they are numbered in order.
+% Ideally, you should not use this facility. Affiliations will be numbered
+% in order of appearance and this is the preferred way.
+\icmlsetsymbol{equal}{*}
+
+\begin{icmlauthorlist}
+\icmlauthor{Guilherme Fran\c ca}{to}
+\icmlauthor{Joshua Vogelstein}{to}
+\end{icmlauthorlist}
+
+\icmlaffiliation{to}{Johns Hopkins University}
+
+\icmlcorrespondingauthor{Guilherme Fran\c ca}{guifranca@jhu.edu}
+\icmlcorrespondingauthor{Joshua Vogelstein}{jovo@jhu.edu}
+
+% You may provide any keywords that you
+% find helpful for describing your paper; these are used to populate
+% the "keywords" metadata in the PDF but will not be shown in the document
+\icmlkeywords{Clustering, Energy Statistics}
+
+\vskip 0.3in
 ]
+
+% this must go after the closing bracket ] following \twocolumn[ ...
+
+% This command actually creates the footnote in the first column
+% listing the affiliations and the copyright notice.
+% The command takes one argument, which is text to display at the start of the
+% footnote.
+% The \icmlEqualContribution command is standard text for equal contribution.
+% Remove it (just {}) if you do not need this facility.
+
+\printAffiliationsAndNotice{}  % leave blank if no need to mention equal contribution
+%\printAffiliationsAndNotice{\icmlEqualContribution} % otherwise use the
+%standard text.
 
@@
 close analogy to Newton's gravitational
-potential in physics. We propose a principled approach to the
-\mbox{clustering} problem based on energy statistics theory.
+potential in physics. In this paper, we propose a principled approach to 
+clustering based on energy statistics theory.
 Our mathematical formulation establishes connection to kernel methods,
-leading to a quadratically constrained quadratic program.
-To obtain local solutions of such an optimization problem, which is NP-hard, 
-we introduce an iterative
-algorithm based
-on Hartigan's method, which has the same computational cost but offers
-several advantages compared to the 
-well-known kernel $k$-means, based on Lloyd's heuristic.
-We provide carefully designed numerical experiments showing
-that our method is more flexible and outperforms
+leading to a quadratically constrained quadratic program in the associated
+feature space.
+To obtain local solutions of such an NP-hard optimization problem,
+we introduce an iterative algorithm based
+on Hartigan's method. This algorithm has the same computational 
+cost as kernel $k$-means but offers
+several advantages. 
+We provide carefully designed numerical experiments illustrating
+that the proposed method is more flexible and outperforms
 kernel $k$-means, spectral clustering,
 standard $k$-means and Gaussian mixture models in a variety of settings,
-specially in high \mbox{dimensions}. 
+specially in high dimensions. We also employ the method 
+in a real data application describing shapes of neural synapses.
 \end{abstract}
@@
 distributions, which is achieved 
-under the minimum of the statistical potential energy. 
-When probability distributions are different the 
-statistical potential energy diverges as sample size increases, while tends 
-to a nondegenerate limit distribution when probability
+under the minimum of a ``statistical energy''. 
+When probability distributions are different, this
+statistical energy diverges as sample size increases. On the other hand,
+it tends to a nondegenerate limit distribution when probability
 distributions are equal.
-The test statistic has compact representation
+The test statistic has a compact representation
 in terms of expectations of pairwise distances, providing
@@
 
-Distance covariance was further generalized from Euclidean 
-to metric spaces of strong negative type \citep{Lyons}. Furthermore, 
+\citet{Lyons} generalized
+distance covariance from Euclidean 
+to arbitrary metric spaces of strong negative type. Furthermore, 
 the missing link between energy distance based tests and kernel 
 based tests has 
-been recently resolved \citep{Sejdinovic2013}, where a unifying framework
+been recently resolved by \citet{Sejdinovic2013}, where a unifying framework
 establishing an equivalence between generalized energy distances to maximum
@@
 kernel methods often used in machine learning, and form the basis 
-of our approach.
+of our approach in this paper.
 
@@
 map data points to a RKHS, then clustering can be performed in the associated
-Hilbert space by exploiting its inner product. However, the kernel choice remains 
+Hilbert space by exploiting its inner product. However, 
+the kernel choice remains 
 the biggest challenge since there is no principled theory to construct a kernel
@@
 in the feature space \citep{Girolami}. Furthermore, kernel $k$-means algorithm
-\citep{Dhillon2,Dhillon} is still based on Loyd's heuristic \citep{Lloyd}
+\citep{Dhillon2,Dhillon} is still based on Loyd's heuristic 
 of grouping points that are closer to a cluster center, now
-in feature space. 
-We refer the reader to \citep{Filippone} for a survey of clustering
+in the feature space. 
+We refer the reader to \citet{Filippone} for a survey of clustering
 methods.
@@
 Although clustering from energy statistics in Euclidean spaces was considered
-in \citep{Kgroups}, the precise optimization problem behind this approach
+in \citet{Kgroups}, the precise optimization problem behind this approach
 remains elusive, as well as the connection with kernel methods.
@@
 find local solutions of the above mentioned QCQP, which is NP-hard in general.
-Hartigan's method was also used in \citep{Kgroups}, but without any
+Hartigan's method was also used by \citet{Kgroups}, but without any
 connection to kernels. More importantly, the 
@@
 over Lloyd's method was already demonstrated 
-in some simple settings
-\citep{Telgarsky,Slonin}, but apparently this method did not receive 
-the deserved attention. To the 
+in some simple settings by
+\citet{Telgarsky,Slonin}, but apparently this method did not receive 
+the deserved attention by the statistics and machine learning communities. 
+To the 
 best of our knowledge, Hartigan's method was not previously 
 employed together with kernel methods. 
-We provide a fully kernel based Hartigan's algorithm for clustering,
+Here we provide a fully kernel based Hartigan's algorithm for clustering,
 where the kernel is fixed by energy statistics. 
@@
 showing that it is able to perform accurately on data coming from 
-very different distributions, contrary to $k$-means and GMM for instance.
+very different distributions, contrary to $k$-means and GMM, for instance.
 More specifically, the proposed method performs 
@@
 than $k$-means and GMM even on Gaussian settings.
+We illustrate the applicability of the proposed method on a real dataset
+obtained by neuroscientists experts, which describes 
+shapes of neural synapses.
 
@@
 statistics \citep{Szkely2013,Lyons} and its relation to 
-RKHS \citep{Sejdinovic2013}
-which form the basis of our work.
-
-Consider random vectors $X_i$ living in an arbitrary space
+RKHS \citep{Sejdinovic2013}.
+
+Consider random vectors $X_i$ ($i=1,\dotsc,n$) living in an arbitrary space
 $\mathcal{X}$ of \emph{negative type}, which means that $\mathcal{X}$
@@
 For instance, the \emph{standard
-energy distance} \citep{Szkely2013} in Euclidean spaces uses
+energy distance} 
+in Euclidean spaces 
+introduced by \citet{Szkely2013} 
+uses
 the semimetric
@@
 where $0< \alpha \le 2$ and $\| \cdot \|$ is the
-Euclidean norm in $\mathcal{X}=\mathbb{R}^D$. In this case $\Energy(P,Q)$
+Euclidean norm in $\mathcal{X}=\mathbb{R}^D$. In this case, $\Energy(P,Q)$
 is rotationally invariant. For $0<\alpha<2$ we have
@@
 from $k$ unknown distributions $\{ P_j \}_{j=1}^k$,
-with points lying on a space of negative type,
-$x_i \in \mathcal{X}$.
+where $x_i \in \mathcal{X}$.
 Let $\mathbb{X} = \bigcup_{j=1}^k \C_j$ be a disjoint
@@
 \end{equation}
-where $n_i = |\C_i|$ denotes the number of elements in partition
+where $n_i = |\C_i|$ denotes the number of points in partition
 $\C_i$. 
@@
 if and only if $x_i \sim P_j$. 
-The quantity $S$ is
+The quantity $S$ defined above is
 a test statistic for equality of distributions
@@
 $\kk : \mathcal{X} \times \mathcal{X} \to 
-\mathbb{R}$, which is a symmetric and positive definite function, i.e. 
-$\kk(x_i,x_j) = \kk(x_j,x_i)$ and 
-$\sum_{i,j=1}^n c_i c_j \kk(x_i, x_j) \ge 0$, or equivalently,
-if $G$ is the Gram matrix with entries $G_{ij} = \kk(x_i,x_j)$
-then $G = G^\top$ and $v^\top G v \ge 0$ for any $v \in \mathbb{R}^n$.
+\mathbb{R}$, which is a symmetric and positive definite function.
+%, i.e. 
+%$\kk(x_i,x_j) = \kk(x_j,x_i)$ and 
+%$\sum_{i,j=1}^n c_i c_j \kk(x_i, x_j) \ge 0$, or equivalently,
+%if $G$ is the Gram matrix with entries $G_{ij} = \kk(x_i,x_j)$
+%then $G = G^\top$ and $v^\top G v \ge 0$ for any $v \in \mathbb{R}^n$.
 For every $x \in \mathcal{X}$ there exists 
@@
 $\langle h_x, h_y \rangle = \kk(x,y)$.
-Conversely \citep{Aronszajn},  
+Conversely, 
 for every symmetric
@@
 \emph{feature map} $\varphi: x \mapsto \kkk_x \in \Hk$ such
-that $\langle \varphi(x), \varphi(y) \rangle = \kk(x, y)$.
+that $\langle \varphi(x), \varphi(y) \rangle = \kk(x, y)$
+\citep{Aronszajn}.
 
@@
 The equality between \eqref{eq:mmd} and \eqref{eq:mmd2}
-gives $\langle \kkk_P, \kkk_Q \rangle_{\Hk} = \E \, \kk(X, Y)$,
-thus in practice we can estimate the inner product between  
-embedded distributions by averaging the kernel function over sampled data.
-
-The following important result shows that semimetrics of negative
-type and symmetric positive definite kernels are closely related
-\citep{Berg1984}. Let $\rho: \mathcal{X} \times \mathcal{X} \to \mathbb{R}$,
+gives $\langle \kkk_P, \kkk_Q \rangle_{\Hk} = \E \, \kk(X, Y)$.
+%thus in practice we can estimate the inner product between  
+%embedded distributions by averaging the kernel function over sampled data.
+
+The following important result of 
+\citet{Berg1984}
+shows that semimetrics of negative
+type and symmetric positive definite kernels are closely related. 
+Let $\rho: \mathcal{X} \times \mathcal{X} \to \mathbb{R}$,
 and $x_0 \in \mathcal{X}$ be an arbitrary but fixed point.
@@
 of negative type.
-We have a family of kernels, one for each choice of $x_0$. Conversely,
+We thus have a family of kernels, one for each choice of $x_0$. Conversely,
+as shown by \citet{Sejdinovic2013},
 if $\rho$ is a semimetric of negative type and $\kk$ is a kernel in this
@@
 and the canonical feature map 
-$\varphi: x \mapsto \kkk_x$ is injective \citep{Sejdinovic2013}.
-When these conditions are met we say that the kernel $\kk$ 
+$\varphi: x \mapsto \kkk_x$ is injective. 
+When the above conditions are met we say that the kernel $\kk$ 
 generates the semimetric $\rho$. 
@@
 inner products on RKHS, which is one of the main results of
-\citep{Sejdinovic2013}. If $\rho$ is a semimetric
+\citet{Sejdinovic2013}. If $\rho$ is a semimetric
 of negative type and $\kk$ a kernel that generates $\rho$, then
@@
 \eqref{eq:energy3}, and using \eqref{eq:mmd2}, yields
-\begin{equation}
+\begin{align}
+\tfrac{1}{2}\Energy(P, Q) &= 
+\E \, \kk(X, X') + \E \, \kk(Y, Y') - 2\E \, \kk(X, Y) \nonumber \\ 
+&= \gamma_\kk^2(P,Q) .
 \label{eq:Erho}
-\begin{split}
-\tfrac{1}{2}\Energy(P, Q) &= 
-\E \, \kk(X, X') + \E \, \kk(Y, Y') - 2\E \, \kk(X, Y) \\ 
-&= \gamma_\kk^2(P,Q) .
-\end{split}
-\end{equation}
+\end{align}
 Therefore, we can compute the energy distance using the inner product
-of $\Hk$.
+of the associated Hilbert space $\Hk$.
 
@@
 
-This section contains our main theoretical results where 
+This section contains our main theoretical contributions, where 
 we formulate an optimization problem for clustering 
-based on energy statistics in the RKHS.
+based on energy statistics in the associated RKHS.
 The proofs are contained in supplementary material.
 
-Due to the energy test statistic for equality of distributions,
-the obvious
+Due to the energy test statistic for equality of distributions previously
+discussed, the obvious
 criterion for clustering data is to 
-maximize $S$ which makes 
+maximize $S$, defined in \eqref{eq:between}, which makes 
 each cluster as different
@@
 shows that maximizing $S$ is, however, equivalent to minimizing
-$W$ which has a more convenient form.
+$W$, which has a more convenient form.
 
@@
 \label{th:minimize}
-Let $\mathbb{X} = \{x_1,\dotsc,x_n\}$ where each data point
+Let $\mathbb{X} = \{x_1,\dotsc,x_n\}$, where each data point
 $x_i$ lives in a space $\mathcal{X}$ of negative type. 
@@
 the partition
-$\mathbb{X} = \bigcup_{j=1}^k \C_j$, where $\C_i \cap C_j = \emptyset$ for
+$\mathbb{X} = \bigcup_{j=1}^k \C^\star_j$, where 
+$\C^\star_i \cap C^\star_j = \emptyset$ for
 all $i\ne j$, maximizes the between-sample statistic $S$, defined
@@
 \label{eq:minimize}
-\min_{\C_1,\dotsc,C_k  } W(
+\{C_1^\star, \dotsc,\C_k^\star\} = \argmin_{\{\C_1,\dotsc,C_k\}  } W(
 \C_1, \dotsc, \C_k),
@@
 
-In the Euclidean case, the optimization problem \eqref{eq:minimize} based on
-energy statistics was already proposed in \citep{Kgroups}. However, it is
+In the particular Euclidean case, 
+the optimization problem \eqref{eq:minimize} based on
+energy statistics was already proposed in \citet{Kgroups}. However, it is
 important to note that this is equivalent to maximizing $S$,
-which is the test statistic for equality of distributions. In this current
-form, the relation with kernels and other clustering methods is obscure.
-In the following we demonstrate what is the explicit optimization problem behind 
-\eqref{eq:minimize} in the corresponding RKHS, 
-establishing the connection with kernel methods.
+which is the test statistic for equality of distributions.
+
+The clustering
+problem as stated in the
+form \eqref{eq:minimize} makes
+the relation with kernels and other clustering methods obscure.
+In the following, we demonstrate what is the explicit optimization 
+problem behind 
+\eqref{eq:minimize} in the associated RKHS, directly 
+establishing the connection with kernel methods commonly used in machine
+learning.
 
 Assume that the kernel $\kk: \mathcal{X} \times \mathcal{X} \to \mathbb{R}$ 
-generates $\rho$.  Define  the Gram matrix $G \in \mathbb{R}^{n\times n}$ as
+generates $\rho$.  Define  the Gram matrix $G \in \mathbb{R}^{n\times n}$ 
+with components
 \begin{equation}
@@
 0 & \mbox{otherwise}
-\end{cases} \
+\end{cases} , \quad
 \Zt_{ij} \equiv \begin{cases}
@@
 0 & \mbox{otherwise}
-\end{cases}
+\end{cases} .
 \end{equation}
@@
 \label{th:qcqp2}
-The optimization problem \eqref{eq:minimize} is equivalent to
+Problem \eqref{eq:minimize} is equivalent to
 \begin{equation}
@@
 
-Therefore, to group data
+Thus, to group data
 $\mathbb{X} = \{ x_1,\dotsc,x_n \}$
@@
 indicating that $x_i \in \C_j$.
-This optimization problem is nonconvex, and NP-hard,
-thus a direct approach 
-is computational prohibitive even for small datasets.
+
+Besides NP-hard, the optimization problem \eqref{eq:qcqp2} is nonconvex, and
+a direct computational approach is prohibitive even for small datasets.
 However, one can find approximate solutions by relaxing some 
@@
 
-Note that the energy clustering optimization problem \eqref{eq:qcqp2} 
+Note also that the optimization problem \eqref{eq:qcqp2} 
 is valid for data living in an \emph{arbitrary} space of negative type, where
@@
 \eqref{eq:rho_standard}.
-The same would be valid for data living in more general
+The same is valid for data living in more general
 spaces $(\mathcal{X}, \rho)$.
@@
 $\rho$ which measures the similarity between data points.
-If prior information is available to make this choice
-it can be immediately incorporated in the 
-optimization problem \eqref{eq:qcqp2} through the kernel matrix $G$.
-
-\paragraph{Relation to Kernel $\bm{k}$-Means.}
+If prior information is available to make this choice, then
+it can be immediately incorporated  
+into \eqref{eq:qcqp2} through the kernel matrix $G$.
+
+\subsection{Relation to Kernel $\bm{k}$-Means Optimization Problem}
 One may wonder how energy clustering 
-relates to the well-known kernel $k$-means problem%
-\footnote{When we refer to kernel $k$-means problem we mean specifically 
+relates to the well-known kernel $k$-means optimization problem%
+\footnote{By this we mean specifically 
 the optimization problem \eqref{eq:kernel_kmeans}, which should not be 
-confused with kernel $k$-means algorithm that is just one possible recipe 
-to solve \eqref{eq:kernel_kmeans}. The distinction should be clear from 
+confused with kernel $k$-means \emph{algorithm} that is 
+just one possible recipe 
+to solve \eqref{eq:kernel_kmeans}. The distinction between kernel $k$-means
+problem and kernel $k$-means algorithm should be clear from 
 the context.}, 
@@
 The kernel $k$-means optimization
-problem,
-in feature space,
+problem
 is defined by
@@
 \sum_{x \in \C_j} \| \varphi(x) - \varphi(\mu_j) \|^2
-\bigg\}
+\bigg\} ,
 \end{equation}
@@
 on Lloyd's heuristic \citep{Lloyd}.
-It is known \citep{Dhillon2,Dhillon}
+It was shown by \citet{Dhillon2,Dhillon}
 that kernel $k$-means problem 
@@
 is guaranteed to be positive definite.
-As shown by \citep{Dhillon2,Dhillon}, kernel $k$-means, spectral clustering,
+As shown by \citet{Dhillon2,Dhillon}, kernel $k$-means, spectral clustering,
 and graph partitioning problems such as ratio association, ratio cut, and
 normalized cut are all equivalent to a QCQP of the form \eqref{eq:qcqp2}.
-Thus one can view all these problems arising from energy statistics as well.
+Therefore, one can view all of 
+these problems as arising from energy statistics as well.
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Hartigan's for Energy Clustering}
+\section{Hartigan's Method for Energy Clustering}
 \label{sec:algo}
@@
 We now introduce an iterative algorithm based
-on Hartigan's method \citep{Hartigan} to find a local
+on \citet{Hartigan} to find a local
 maximizer of the optimization problem \eqref{eq:qcqp2}. Due to 
@@
 \end{equation}
-where $Q_j$ represents an internal energy cost of cluster $\C_j$, and
-$Q$ is the total energy cost where each $Q_j$ 
+where $Q_j$ represents an internal cost of cluster $\C_j$, and
+$Q$ is the total cost where each $Q_j$ 
 is weighted by the inverse
 of the number of points in $\C_j$. For a data point $x_i$ we denote
-its own energy cost
+its own cost
 with the entire cluster $\C_\ell$ by
@@
 Q_\ell(x_i) \equiv \sum_{y\in\C_\ell} \kk(x_i, y) = 
-G_{i \bullet} \cdot Z_{\bullet \ell},
-\end{equation}
-where $G_{i\bullet}$ ($G_{\bullet i}$) denotes
+G_{i \bullet} \cdot Z_{\bullet \ell}.
+\end{equation}
+Above, $G_{i\bullet}$ ($G_{\bullet i}$) denotes
 the $i$th row (column) of the matrix $G$.
-
 For a given configuration, we consider the maximum change
@@
 denoted by $Q^{(\ell)}$. We are interested in computing the maximum 
-cost change
 $\Delta Q^{j\to \ell} (x_i) \equiv Q^{(\ell)} - Q^{(j)}$, for $\ell\ne j$. 
@@
 \Delta Q^{j\to \ell} (x_i) = \dfrac{Q_\ell^{+}}{n_\ell+1} + 
-\dfrac{Q_j^-}{n_j-1} - \dfrac{Q_j}{n_j} - \dfrac{Q_\ell}{n_\ell}
+\dfrac{Q_j^-}{n_j-1} - \dfrac{Q_j}{n_j} - \dfrac{Q_\ell}{n_\ell} ,
 \end{equation}
@@
 $Q_j^{-} = Q_j - 2 Q_j(x_i) + G_{ii}$, hence
-\begin{equation}
+\begin{multline}
 \label{eq:changeQ}
-\begin{split}
-\Delta Q^{j \to \ell}(x_i)  = 
-\dfrac{1}{n_j - 1} & \left[ \dfrac{Q_j}{n_j} - 2 Q_j(x_i) + G_{ii} \right] \\
-- \dfrac{1}{n_\ell + 1}&\left[ \dfrac{Q_\ell}{n_\ell} - 2 Q_\ell(x_i) 
+\Delta Q^{j \to \ell}(x_i)  i = 
+\dfrac{1}{n_j - 1} \left[ \dfrac{Q_j}{n_j} - 2 Q_j(x_i) + G_{ii} \right] \\
+- \dfrac{1}{n_\ell + 1}\left[ \dfrac{Q_\ell}{n_\ell} - 2 Q_\ell(x_i) 
 - G_{ii} \right].
-\end{split}
-\end{equation}
+\end{multline}
 Therefore, if $\Delta Q^{j\to \ell}(x_i) > 0$ we get closer to a 
@@
 moving $x_i$ to $\C_\ell$, otherwise we should keep $x_i$ in $\C_j$. 
+A similar analysis leading to \eqref{eq:changeQ} in the particular
+case of Euclidean distances was considered in \citet{Kgroups}.
 
@@
 
-\begin{theorem}[\cite{Telgarsky}]
+\begin{theorem}[\citet{Telgarsky}]
 If $n > k$ the resulting partition obtained from Hartigan's method has
-\begin{enumerate*}
-\item \label{noempty} no empty clusters, and
-\item \label{diffmean} distinct means.
-\end{enumerate*}
+no empty clusters, and distinct means.
+%\begin{enumerate*}
+%\item \label{noempty} no empty clusters;
+%\item \label{diffmean} distinct means.
+%\end{enumerate*}
 \end{theorem}
@@
 
-\begin{theorem}[\cite{Telgarsky}]
+\begin{theorem}[\citet{Telgarsky}]
 The set of local optima of Hartigan's method is a (possibly strict) subset
@@
 escape local optima of Lloyd's method.
-Kernel $k$-means cannot
+Therefore, kernel $k$-means cannot
 improve on a local optima of $\mathcal{E}^H$-clustering, but on the other hand,
@@
 Points in this excess volume
-force Hartigan's to iterate, contrary
-to Lloyd's. 
+force Hartigan's method to iterate, contrary
+to Lloyd's method. 
 Moreover, this improvement should be more prominent as
@@
 $k$ increases.
-The empirical results of \citep{Telgarsky} show that 
+The empirical results of \citet{Telgarsky} show that 
 an implementation of Hartigan's method has comparable execution time 
@@
 Lloyd's method,
-but no explicit complexity was provided. In our case both
+but no explicit complexity was provided. In our case, both
 $\mathcal{E}^H$-clustering and kernel $k$-means
@@
 
-In \citep{Slonin}, Hartigan's method was applied to $k$-means problem
+In \citet{Slonin}, Hartigan's method was applied to $k$-means problem
 with any Bregman divergence. It was shown that the number of Hartigan's
@@
 In addition, examples were provided where
-\emph{any} initial partition correspond to a Lloyd's local optima, while  
-the number of Hartigan's local optima  is small and 
+\emph{any} initial partition correspond to a local optima of Lloyd's method,
+while  
+the number of local optima of Hartigan's method is small and 
 correspond to true partitions of the data. Empirically, the number of
-Hartigan's local optima was considerably smaller.
+local optima of Hartigan's method was considerably smaller than that of Lloyd's
+method.
 
@@
 
-The main goal of this section is threefold. First, to compare
+\subsection{Synthetic Experiments}
+
+The goal of this section is threefold. First, to compare
 $\mathcal{E}^H$-clustering in Euclidean space 
@@
 Third, to show the flexibility
-of energy clustering, able to perform 
+of energy clustering since it is able to perform 
 accurately in different settings using the same kernel.
 
-\textbf{Experimental Setup.} The following setup holds unless specified
+\emph{Experimental Setup.} 
+The following setup holds unless specified
 otherwise. 
@@
 with \mbox{generating} kernels to the following semimetrics:
-\begin{equation}
-\begin{split}
+\begin{subequations}
+\label{eq:test_metrics}
+\begin{align}
 \rho_{\alpha}(x,y) &= \| x-y \|^{\alpha}, \\
-%\kk_{\alpha}(x,y) &= \tfrac{1}{2} \left( \| x \|^{\alpha} + \| y \|^{\alpha} 
-%- \| x-y \|^{\alpha} \right) \\
-%
 \widetilde{\rho}_{\sigma}(x,y) &= 2 - 2 e^{-\tfrac{\|x-y\|}{2 \sigma}},  \\
-%\widetilde{\kk}_{\sigma}(x,y) &= e^{-\tfrac{\|x-y\|}{2\sigma}} \\
-%
 \widehat{\rho}_{\sigma}(x,y) &= 2 - 2 e^{-\tfrac{\|x-y\|^2}{2 \sigma^2}}.  
-%\widehat{\kk}_{\sigma}(x,y) &= e^{-\tfrac{\|x-y\|^2}{2\sigma^2}}  
-\end{split}
-\end{equation}
-These are generated by formula \eqref{eq:kernel_semimetric},
+\end{align}
+\end{subequations}
+These are generated by formula \eqref{eq:kernel_semimetric}
 and we fix $x_0=0$.
@@
 it from breaking in high dimensions. 
-Spectral clustering is based on \citep{Malik}. 
+Spectral clustering is based on \citet{Malik}. 
 We implemented kernel $k$-means
-as described in \citep{Dhillon2,Dhillon}, and $\mathcal{E}^H$-clustering
+as described in \citet{Dhillon2,Dhillon}, and $\mathcal{E}^H$-clustering
 as described in Algorithm~\ref{algo}. Both
 will also be initialized with $k$-means++.
-We run the algorithms $5$ times with different initializations, picking
-the result with best objective. 
+We run each algorithm $5$ times with different initializations, picking
+the result with best objective function value. 
 We evaluate clustering quality by
-the \emph{accuracy} based on the true labels. For each setting we show 
+the \emph{accuracy}, which is the fraction of correctly clustered points
+based on the ground truth. More precisely, let $Z$ be the true label
+matrix and $\hat{Z}$ the estimated label matrix produced by 
+an algorithm. The accuracy is then defined as
+\begin{equation}
+\label{eq:accuracy}
+\textnormal{accuracy}(\hat{Z}) \equiv \max_\pi \dfrac{1}{n} \sum_{i=1}^n
+\sum_{j=1}^k \hat{Z}_{i \pi(j)} Z_{ij},
+\end{equation}
+where $\pi$ is a permutation of the $k$ cluster groups. The accuracy
+is always between $[0,1]$, where $1$ correspond to all points correctly
+clustered, and $0$ to all points wrongly clustered.
+For each setting we show 
 the average accuracy over $100$ Monte
-Carlo trials, with error bars indicating standard error.
+Carlo trials, with error bars being standard error.
 
 \begin{figure}
-\includegraphics[scale=0.381]{hist_normal.pdf}
-\includegraphics[scale=0.381]{hist_lognormal.pdf}\\[-0.6em]
-\includegraphics[scale=0.41]{1D_normal.pdf}
-\includegraphics[scale=0.41]{1D_lognormal.pdf}
-\put(-180,114){(a)}
-\put(-60,114){(c)}
-\put(-180,-8){(b)}
-\put(-60,-8){(d)}
+\includegraphics[width=0.23\textwidth]{hist_normal.pdf}
+\includegraphics[width=0.23\textwidth]{hist_lognormal.pdf}\\[-0.6em]
+\includegraphics[width=0.23\textwidth]{1D_normal.pdf}
+\includegraphics[width=0.23\textwidth]{1D_lognormal.pdf}
 \caption{
 \label{fig:1D}
-Clustering results for one dimensional data. (a) and (c) show
-the probability densities of \eqref{eq:uni_normal} and
-\eqref{eq:uni_lognormal}, respectively. The corresponding 
-clustering accuracy versus sample size 
-are respectively shown in (b) and (d).
-Bayes accuracy in both cases are $\approx 0.88$.
+Clustering results for one dimensional data
+\eqref{eq:uni_normal}--\eqref{eq:uni_params}. \emph{Top Left:} Probability
+density of \eqref{eq:uni_normal}. \emph{Top Right:} Probability
+density of \eqref{eq:uni_lognormal}.
+\emph{Bottom Left:} Clustering data from \eqref{eq:uni_normal}.
+\emph{Bottom Right:} Clustering data from \eqref{eq:uni_lognormal}.
+The dashed lines show the optimal Bayes accuracy, which in both cases 
+are $\approx 0.88$.
 }
@@
 \mu_1 &= 1.5, \ \sigma_1=0.3, \ \mu_2=0, \ \sigma_2=1.5.
+\label{eq:uni_params}
 \end{align}
 We increase the sample size $n$ and show clustering accuracy 
+\eqref{eq:accuracy}
 versus $n$ in Fig.~\ref{fig:1D}. $\mathcal{E}^H$-clustering 
 performs better than $k$-means for normally distributed data, 
-and worse than GMM 
+and worse than GMM,
 as expected. 
 However, for lognormal distributions $k$-means and GMM are only slightly
-better than chance while $\mathcal{E}^H$-clustering is still very
-accurate.
+better than chance while $\mathcal{E}^H$-clustering is still 
+accurate. The reason is that $k$-means and GMM models are inconsistent
+with this lognormal data. Energy clustering on the other hand
+is model-free.
 
 \begin{figure}
-\includegraphics[scale=0.41]{normal_highdim_mean.pdf}
-\includegraphics[scale=0.41]{normal_highdim_cov.pdf}
-\put(-180,-7){(a)}
-\put(-60,-7){(b)}
+\includegraphics[width=0.23\textwidth]{normal_highdim_mean.pdf}
+\includegraphics[width=0.23\textwidth]{normal_highdim_cov.pdf}
 \caption{
@@
 High dimensional Gaussian mixtures.
-(a) Parameters as in \eqref{eq:gauss1}, and  
-(b) as in \eqref{eq:gauss2}. Dashed line is Bayes accuracy.
-We plot accuracy versus $D$.
+\emph{Left:} Parameters as in \eqref{eq:gauss1}.
+\emph{Right:} Parameters  as in \eqref{eq:gauss2}. 
+The dashed lines are Bayes accuracy.
 }
@@
 \tfrac{1}{2} \mathcal{N}(\mu_1,\Sigma_1) +
-\tfrac{1}{2} \mathcal{N}(\mu_2,\Sigma_2)$ in $\mathbb{R}^D$ with
+\tfrac{1}{2} \mathcal{N}(\mu_2,\Sigma_2)$ in $\mathbb{R}^D$, with
 $\Sigma_1=\Sigma_2 = I_D$,  and
@@
 Bayes error is fixed as $D$ increases giving an optimal accuracy
-$\approx 0.86$.
+of $\approx 0.86$.
 We sample $200$ points on each trial.
-As shown in Fig.~\ref{fig:gauss}a, 
-$\mathcal{E}^H$- and spectral-clustering have close
+As shown in Fig.~\ref{fig:gauss} (left), 
+$\mathcal{E}^H$- and spectral clustering have close
 performance,  
@@
 \end{equation}
-We simply chose $10$ number uniformly at random on the interval
-$[1,5]$, and other choice would give analogous results.
+We simply chose $10$ numbers uniformly at random between
+$[1,5]$ and other choice would give analogous results.
 Bayes  accuracy is fixed at $\approx 0.95$.
-From Fig.~\ref{fig:gauss}b we see that 
+From Fig.~\ref{fig:gauss} (right) we see that 
 GMM is better in low dimensions, 
@@
 as $D$ increases, as kernel $k$-means and $k$-means, while  
-$\mathcal{E}^H$- and spectral-clustering remains much more stable.
+$\mathcal{E}^H$- and spectral clustering remains much more stable.
 
 \begin{figure}[t]
-\includegraphics[scale=.41]{normal_kernels.pdf}
-\includegraphics[scale=.41]{lognormal_kernels.pdf}\\
-\includegraphics[scale=.41]{normal_kernels_difference.pdf}
-\includegraphics[scale=.41]{lognormal_kernels_difference.pdf}
-\put(-180,117){(a)}
-\put(-60,117){(c)}
-\put(-180,-10){(b)}
-\put(-60,-10){(d)}
+\includegraphics[width=0.23\textwidth]{normal_kernels.pdf}
+\includegraphics[width=0.23\textwidth]{lognormal_kernels.pdf}\\
+\includegraphics[width=0.23\textwidth]{normal_kernels_difference.pdf}
+\includegraphics[width=0.23\textwidth]{lognormal_kernels_difference.pdf}
 \caption{
 \label{fig:consist}
-(a,b) Normal and (c,d) lognormal mixtures with
+Normal \emph{(Left)}
+and lognormal \emph{(Right)} mixtures with
 parameters \eqref{eq:20gauss}. We use different kernels
@@
 is $\approx 0.9$.
-In (c) and (d) we show the difference in accuracy
-between $\mathcal{E}^H$ versus kernel $k$-means and
+The two plots in the \emph{Bottom} show the difference in accuracy
+between $\mathcal{E}^H$-clustering versus kernel $k$-means and
 spectral clustering, with $\widetilde{\rho}_1$.
@@
 We increase the sample size $n \in [10, 400]$ and show
-the accuracy versus $n$ in Fig.~\ref{fig:consist}a, where we compare
+the accuracy versus $n$ in Fig.~\ref{fig:consist} (top-left), where we compare
 $\mathcal{E}^H$-clustering with different
@@
 superior for small $n$. For
-the same setting, in Fig.~\ref{fig:consist}b we
-we show the difference in accuracy provided by $\mathcal{E}^H$ minus
+the same setting, in Fig.~\ref{fig:consist} (bottom-left) we
+we show the difference in accuracy provided by $\mathcal{E}^H$-clustering minus
 kernel $k$-means and spectral clustering when using $\widetilde{\rho}_1$.
@@
 The same experiments are shown in 
-Fig.~\ref{fig:consist}c--d, where
+Fig.~\ref{fig:consist} (top-right and bottom-right), where
 $\mathcal{E}^H$-clustering still performs accurately 
@@
 \begin{figure}[b!]
-\includegraphics[scale=0.21]{complex_data.pdf}
-\put(-210,-10){(a)}
-\put(-150,-10){(b)}
-\put(-90,-10){(c)}
-\put(-35,-10){(d)}
+\centering
+\includegraphics[width=0.15\textwidth]{2cigars.pdf}
+\includegraphics[width=0.15\textwidth]{2circles.pdf}
+\includegraphics[width=0.15\textwidth]{3circles.pdf}
 \caption{\label{fig:other}
-(a) Parallel cigars, $200$ points each. (b,c) Two and three  
-concentric circles with Gaussian noise, $400$ points each.   
-(d) MNIST handwritten digits.
-Clustering results are in Tables~\ref{table:other} and
-\ref{table:mnist}.
+\emph{Left}: Parallel cigars, $200$ points each. \emph{Center and Right:}
+Two and three concentric circles, respectively,
+with Gaussian noise. $400$ points for each class.
 }
@@
 \caption{\label{table:other}
-Clustering data from Fig.~\ref{fig:other}a--c.
+Clustering data from Figure~\ref{fig:other}.
 }
@@
 \toprule[1pt]
- & & \emph{Fig.~\ref{fig:other}a}
- & & \emph{Fig.~\ref{fig:other}b}
- & & \emph{Fig.~\ref{fig:other}c} \\
+ & & \emph{Parallel Cigars}
+ & & \emph{Two Circles}
+ & & \emph{Three Circles} \\
 \midrule[0.5pt]
@@
 
-\begin{table*}
-\caption{\label{table:mnist}
-Clustering MNIST data from Fig.~\ref{fig:other}d.
-}
-\begin{center}
-\footnotesize{
-\begin{tabular}{@{}r@{}l@{}@{}l@{}@{}l@{}@{}l@{}l@{}}
-\toprule[1pt]
-\emph{Class Subset}~~~~ & & 
-%$~\{0,1,2\}$ &
-$~\{0,1,\dotsc,4\}~$ &
-$~\{0,1,\dotsc,6\}~$ &
-$~\{0,1,\dotsc,8\}~$ &
-$~\{0,1,\dotsc,9\}~$ \\
-\emph{parameter}~~~~ & {$\sigma$}
-%& $~10.34~$
-& $~10.41~$
-& $~10.41~$
-& $~10.37~$
-& $~10.19~$ \\
-\midrule[0.5pt]
-\multirow{3}{*}{\emph{$\mathcal{E}^H$-clustering~~~~}}
-& $\rho_{1}$\hspace{1em} 
-%&$~0.937\pm 0.006~$
-&$~0.873\pm 0.025~$
-&$~0.731\pm 0.016~$
-&$~0.687\pm 0.016~$
-&$~0.581\pm 0.011~$
-\\
-& $\rho_{1/2}$ 
-%&$~0.939\pm 0.006~$
-&$~0.874\pm 0.027~$
-&$~0.722\pm 0.017~$
-&$~0.647\pm 0.017~$
-&$~\bm{0.600\pm 0.009}~$
-\\
-& $\widetilde{\rho}_{\sigma}$ 
-%&$~0.939\pm 0.006~$
-&$~0.847\pm 0.031~$
-&$~0.695\pm 0.023~$
-&$~0.657\pm 0.014~$
-&$~0.584\pm 0.013~$
-\\
-& $\widehat{\rho}_{\sigma}$ 
-%&$~0.933\pm 0.005~$
-&$~\bm{0.891\pm 0.009}~$
-&$~\bm{0.759\pm 0.011}~$
-&$~\bm{0.704\pm 0.011}~$
-&$~\bm{0.591\pm 0.012}~$
-\\
-\midrule[.5pt]
-\emph{spectral-clustering~~~~} 
-& $\widehat{\rho}_{\sigma}$ 
-%&$~0.823\pm 0.015~$
-&$~0.769\pm 0.012~$
-&$~0.678\pm 0.014~$
-&$~0.649\pm 0.018~$
-&$~0.565\pm 0.009~$
-\\
-\emph{$k$-means}~~~~ & \xmark
-%&$~0.927\pm 0.004~$
-&$~0.878\pm 0.010~$
-&$~0.744\pm 0.008~$
-&$~0.695\pm 0.012~$
-&$~0.557\pm 0.012~$
-\\
-\emph{GMM}~~~~ & \xmark
-%&$~\bm{0.952\pm 0.005}~$
-&$~0.839\pm 0.015~$
-&$~0.694\pm 0.010~$
-&$~0.621\pm 0.009~$
-&$~0.540\pm 0.009~$
-\\
-\bottomrule[1pt]
-\end{tabular}
-}
-\end{center}
-\end{table*}
-
-In Fig.~\ref{fig:other}a--c we have 
-complex two dimensional data. 
-We apply $\mathcal{E}^H$  with different kernel choices,
+In Fig.~\ref{fig:other} we have examples of
+complex two dimensional datasets. 
+We apply $\mathcal{E}^H$-clustering  with different kernel choices,
 and also spectral clustering with the best kernel choice, besides 
 $k$-means and GMM. Here we perform only $10$ Monte Carlo runs.
-In (a) we initialize
-all algorithms with $k$-means++, and in (b) and (c) we initialize at
-random.
-The results are in Table~\ref{table:other}.
-$\mathcal{E}^H$ has superior performance
+For parallel cigars we initialize
+all algorithms with $k$-means++, and the concentric circles example
+we initialize at random.
+The results are shown in Table~\ref{table:other}.
+$\mathcal{E}^H$-clustering has superior performance
 in every example, in particular better than the
 spectral clustering.
-In (a) the semimetrics $\rho_1$ and $\rho_{1/2}$ still provide
+For parallel cigars the semimetrics $\rho_1$ and $\rho_{1/2}$ still provide
 accurate results, however, for
- (b) and (c) the kernel choice is more sensitive.
-
-Consider 
-the infamous MNIST handwritten digits
-from Fig.~\ref{fig:other}d.
-Each data point is an $8$-bit gray scale
-image forming a $784$-dimensional vector 
-corresponding to the digits $\{0,1,\dotsc,9 \}$.
-We compute the parameter 
-%\begin{equation}
-%\label{eq:sigma}
-$
-\sigma^2 = \tfrac{1}{n^2} \sum_{i,j=1}^n \| x_i - x_j \|^2 
-$
-%\end{equation}
-from a separate
-training set, to be used in the kernels.
-We sample subsets of $\{0,1,\dotsc,9 \}$ with $100$ points 
-for each class, over $10$ Monte Carlo trials.
-The results are shown in Table~\ref{table:mnist}.
-Unsupervised clustering on MNIST without feature extraction
-is not trivial. Similar 
-experiment was performed in \citep{Sapiro} where a low-rank
-transformation is learned then subsequently used in subspace clustering. 
-It would be interesting to explore methods
-for learning a better representation of the data and subsequently apply
-$\mathcal{E}^H$-clustering.
+the concentric circles the kernel choice is more sensitive.
+
+\subsection{Real Data Experiment}
+
+\begin{figure*}
+\includegraphics[width=0.18\textwidth]{eigen_gap.pdf}
+\includegraphics[width=0.2\textwidth]{heat_means_k6_energy.pdf}
+\includegraphics[width=0.2\textwidth]{synapse_cluster_2d_k6_energy.pdf}
+\includegraphics[width=0.2\textwidth]{heat_means_k6_gauss.pdf}
+\includegraphics[width=0.2\textwidth]{synapse_cluster_2d_k6_gauss.pdf}
+\vspace{-0.8cm}
+\caption{\label{fig:synapse}
+Clustering synapse dataset.
+Plots in order from \emph{left} to \emph{right}.
+\emph{First:} Eigenvalues of random walk Laplacian. We choose
+$k=6$ since its the first time we see a peak. We use energy clustering
+with the two semimetrics $\rho_1$ and $\widehat{\rho}_2$; see
+\eqref{eq:test_metrics}.
+\emph{Second:} Heat map of the cluster means versus features
+using energy clustering with standard
+metric $\rho_1$ from energy statistics.
+\emph{Third:} Clustered points projected into the two principal components,
+using $\rho_1$.
+\emph{Fourth:} Heat map of the cluster means versus features
+using energy clustering with Gaussian semimetric 
+$\widehat{\rho}_2$.
+\emph{Fifth:} Clustered points projected into the two principal components,
+using $\widehat{\rho}_2$.
+}
+\end{figure*}
+
+We now illustrate an application to real data of 
+$\mathcal{E}^H$-clustering algorithm based on Hartigan's method.
+This dataset describes the shapes of neural synapses and was collected
+by expert neuroscientists. It consists of $1025$ data points
+with $11$ features. Each point
+corresponds to $(x,y,z)$ location in the brain. We normalize this
+dataset beforehand.
+
+We apply $\mathcal{E}^H$-clustering using the two
+semimetrics $\rho_1$ and $\widehat{\rho}_2$; see \eqref{eq:test_metrics}.
+We compute the difference between eigenvalues of the random walk Laplacian
+obtained from the kernel matrix $G$ \eqref{eq:kernel_matrix}.
+The choice of $k$ corresponds to the first time we see a meaningfull
+peak in the plot, which occurs at $k=6$; first plot in
+Fig.~\ref{fig:synapse}.
+This procedure to find $k$ is commonn in the literature \cite{vonLuxburg2007}.
+Having found $k$ we now cluster the data.
+First we use the semimetric $\rho_1$, which
+is the standard Euclidean choice in energy statistics.
+In the second and third plots of Fig.~\ref{fig:synapse} we show a heat
+map of the cluster means versus features and the clustered data points
+projected into the 2 principal components, respectivelly.
+The fourth and fifth plots in Fig.~\ref{fig:synapse} show the same
+experiment but using the semimetric $\widehat{\rho}_2$ corresponding
+to a Gaussian kernel.
+
+We remark that we also used the gap statistic \cite{Tibshirani2001} based
+on $k$-means to determine the number of clusters as a comparison.
+The results slightly suggest $k=6$, but it does not pass the gap
+statistic test.
 
@@
 statistics, valid for arbitrary spaces of negative type.
-Our mathematical formulation of energy clustering 
-reduces to a QCQP in the associated kernel Hilbert space; 
+Our mathematical formulation yields
+a QCQP in the associated kernel Hilbert space; 
 see  Proposition~\ref{th:qcqp2}.
@@
 
-We proposed the iterative $\mathcal{E}^H$-clustering which can be seen
-as a kernelized version of Hartigan's method; see Algorithm~\ref{algo}. 
-It was compared to kernel $k$-means algorithm
-which is based on Lloyd's heuristic.
+We proposed the iterative $\mathcal{E}^H$-clustering 
+(Algorithm~\ref{algo})
+which is 
+a kernelized version of Hartigan's method. 
+It was compared to kernel $k$-means algorithm,
+based on Lloyd's heuristic.
 Both have the same time complexity, however, numerical and theoretical
@@
 $\mathcal{E}^H$-clustering also surpassed spectral clustering, and in
-others performed similarly but never worse. Moreover, computing eigenvectors
+others performed similarly but never worse. 
+Note that computing eigenvectors
 of large matrices can be unfeasible and iterative methods are preferred.
-Finally, kernel methods can benefit from sparsity and
+We provided an application of $\mathcal{E}^H$-clustering to a real dataset
+describing shapes of neural synapses.
+Finally, we remark that kernel methods can benefit from sparsity and
 fixed-rank approximations of the Gram matrix and there is plenty
@@
 \subsection*{Acknowledgements}
-We would like to thank XXX
-%Carey Priebe 
+We would like to thank Carey Priebe 
 for discussions.
-This work was supported by 
-%NIH TRA 
-XXX
-grant.
-
-%\bibliographystyle{unsrt}
-%\bibliographystyle{plainnat}
-\bibliographystyle{abbrvnat}
+We would like to acknowledge the support of the Transformative
+Research Award (NIH \#R01NS092474) and  the Defense Advanced Research Projects
+Agency’s (DARPA) SIMPLEX program through SPAWAR contract N66001-15-C-4041.
+
+\bibliographystyle{icml2018}
 \bibliography{biblio.bib}

### 2018-02-09 energy_clustering/code/energy/data.py

@@ -245,6 +245,7 @@ def histogram_gauss_loggauss(xlim=None, ylim=None, fname='plot.pdf'):
     ax.fill_between(xs, 0, y2s, color=c, alpha=0.3)
 
     ax.set_xlabel(r'$x$')
+    ax.set_ylabel(r'density')
     if xlim:
         ax.set_xlim(xlim)
     if ylim:
@@ -322,9 +323,9 @@ if __name__ == '__main__':
     #          xlim=[0, 4.5])
     #histogram(X, z, colors=['g', 'k'], fname='hist_lognormal.pdf', bins=300,
     #    xlim=[0, 5])
-    #histogram_gauss_loggauss(fname='hist_normal.pdf', xlim=[-3,3], ylim=[0,1.4])
-    histogram_gauss_loggauss(fname='hist_lognormal.pdf', xlim=[0,8],
-                    ylim=[0,0.85])
+    #histogram_gauss_loggauss(fname='hist_normal.pdf', xlim=[-4,4], ylim=[0,1.4])
+    #histogram_gauss_loggauss(fname='hist_lognormal.pdf', xlim=[0,8],
+    #                ylim=[0,0.85])
 
     """
     # cigars
@@ -341,6 +342,7 @@ if __name__ == '__main__':
     s2 = np.array([[1,0],[0,20]])
     X, z = multivariate_normal([m1, m2], [s1, s2], [200, 200])
     plot(X, z, fname='./ball_cigar.pdf')
+    """
     
     # 2 circles
     X, z = circles([1, 3], [0.2, 0.2], [400, 400])
@@ -350,6 +352,7 @@ if __name__ == '__main__':
     X, z = circles([1, 3, 5], [0.2, 0.2, 0.2], [400, 400, 400])
     plot(X, z, fname='./3circles.pdf')
 
+    """
     # spirals
     X, z = spirals([1,-1], [[0.2,0.0], [-0.2,-0.0]], [400,400], noise=0.2)
     plot(X, z, fname='./2spiral.pdf')

### 2018-02-09 energy_clustering/code/energy/metric.py

@@ -7,6 +7,8 @@ from __future__ import division
 
 import numpy as np
 import scipy.optimize
+from sklearn.metrics.cluster import mutual_info_score
+from sklearn.metrics.cluster import entropy
 
 
 def accuracy(z, zh):
@@ -33,4 +35,9 @@ def accuracy(z, zh):
     row_ind, col_ind = scipy.optimize.linear_sum_assignment(-cost_matrix)
     
     return cost_matrix[row_ind, col_ind].sum()/n
+
+def variation_information(z, zh):
+    """Compute variation of information based on M. Meila (2007)."""
+    return entropy(z) + entropy(zh) - 2*mutual_info_score(z, zh)
+
 

### 2018-02-09 energy_clustering/code/experiment2_2.py

@@ -116,25 +116,25 @@ def make_plot(*data_files):
     ## customize plot below ##
     p = plot.ErrorBar()
     p.xlabel = r'$\#$ dimensions'
-    p.ylabel = 'accuracy'
-    p.legends = [r'$\mathcal{E}^{H}$', 
-                 #r'kernel', 
-                 r'$\mathcal{E}^L$', 
-                 r'spectral', 
+    p.ylabel = r'accuracy'
+    p.legends = [r'$\mathcal{E}^{H}$-clustering', 
+                 r'kernel $k$-means', 
+                 #r'$\mathcal{E}^L$', 
+                 r'spectral clustering', 
                  r'$k$-means', 
                  r'GMM']
     p.loc = 3
     p.colors = ['b', 'r', 'g', 'm', 'c']
     p.symbols = ['o', 's', 'D', '^', 'v']
     p.lines = ['-', '-', '-', '-', '-']
-    #p.output = './experiments_figs2/normal_highdim_mean.pdf'
-    p.output = './experiments_figs/normal_highdim_cov.pdf'
-    #p.bayes = 0.86
-    p.bayes = 0.9537075
-    #p.xlim = [10, 200]
-    p.xlim = [10, 700]
-    #p.ylim = [0.55, 0.87]
-    p.ylim = [0.55, 0.965]
+    p.output = './experiments_figs2/normal_highdim_mean.pdf'
+    #p.output = './experiments_figs2/normal_highdim_cov.pdf'
+    p.bayes = 0.86
+    #p.bayes = 0.9537075
+    p.xlim = [10, 200]
+    #p.xlim = [10, 700]
+    p.ylim = [0.55, 0.87]
+    #p.ylim = [0.55, 0.965]
     p.make_plot(table)
 
 def gen_data(fname):
@@ -165,7 +165,7 @@ def worker(dimensions, fname):
 
 ###############################################################################
 if __name__ == '__main__':
-    #fname = './experiments_data2/experiment_highdim_mean_%i.csv'
-    fname = './experiments_data2/experiment_highdim_cov1_%i.csv'
+    fname = './experiments_data2/experiment_highdim_mean_%i.csv'
+    #fname = './experiments_data2/experiment_highdim_cov1_%i.csv'
     #gen_data(fname)
     make_plot(fname%0, fname%1, fname%2, fname%3)

### 2018-02-09 energy_clustering/code/number_clusters.py

@@ -57,7 +57,6 @@ ax.plot(ks, bic_scores, 'r-o', lw=2, ms=8)
 ax.set_xlabel('number of clusters')
 ax.set_ylabel('gmm BIC')
 
-
 ax = fig.add_subplot(236)
 ax.plot(ks, kmeans_gaps, 'r-o', lw=2, ms=8)
 ax.set_xlabel('number of clusters')

### 2018-02-09 energy_clustering/code/run_clustering.py

@@ -15,6 +15,7 @@ import energy.metric as metric
 import energy.eclust as eclust 
 import energy.energy1d as energy1d
 
+
 def initialize(init, k, G, X):
     """Initializaton function to use in energy Lloyd and Hartigan."""
     if init == "spectral":
@@ -43,7 +44,9 @@ def energy_lloyd(k, X, G, z, run_times=10, init="spectral"):
             best_score = score
             best_z = zh
 
-    return metric.accuracy(z, best_z)
+    a = metric.accuracy(z, best_z)
+    v = metric.variation_information(z, best_z)
+    return a, v
 
 def energy_hartigan(k, X, G, z, run_times=10, init="spectral"):
     """Run few times and pick the best objective function value."""
@@ -58,7 +61,9 @@ def energy_hartigan(k, X, G, z, run_times=10, init="spectral"):
             best_score = score
             best_z = zh
 
-    return metric.accuracy(z, best_z)
+    a = metric.accuracy(z, best_z)
+    v = metric.variation_information(z, best_z)
+    return a, v
 
 def energy_spectral(k, X, G, z, run_times=10, init="random"):
     """Run few times and pick the best objective function value.
@@ -75,7 +80,9 @@ def energy_spectral(k, X, G, z, run_times=10, init="random"):
             best_score = score
             best_z = zh
 
-    return metric.accuracy(z, best_z)
+    a = metric.accuracy(z, best_z)
+    v = metric.variation_information(z, best_z)
+    return a, v
 
 def energy1D(X, z):
     """Energy clustering in 1 dimension. No need to run multiple times since
@@ -92,7 +99,9 @@ def kmeans(k, X, z, run_times=10, init='k-means++'):
     km = KMeans(k, n_init=run_times, init=init)
     km.fit(X)
     zh = km.labels_
-    return metric.accuracy(z, zh)
+    a = metric.accuracy(z, zh)
+    v = metric.variation_information(z, zh)
+    return a, v
 
 def gmm(k, X, z, run_times=10, init='kmeans'):
     """GMM from sklearn library. init = {'kmeans', 'random'}, run_times
@@ -103,7 +112,9 @@ def gmm(k, X, z, run_times=10, init='kmeans'):
     gm = GMM(k, n_init=run_times, init_params=init)
     gm.fit(X)
     zh = gm.predict(X)
-    return metric.accuracy(z, zh)
[…]

### 2018-02-13 energy_clustering/icml18/energy_clustering.tex
icml18 paper

@@
 % Use the following line for the initial blind version submitted for review:
-%\usepackage{icml2018}
+\usepackage{icml2018}
 
 % If accepted, instead use the following line for the camera-ready submission:
-\usepackage[accepted]{icml2018}
+%\usepackage[accepted]{icml2018}
 
@@
 standard $k$-means and Gaussian mixture models in a variety of settings,
-specially in high dimensions. We also employ the method 
-in a real data application describing shapes of neural synapses.
+specially in high dimensions. We employ the method 
+to an important real dataset describing protein expressions
+of neural synapses.
 \end{abstract}
@@
 $\mathcal{E}^H$-clustering, kernel $k$-means and spectral clustering
-with \mbox{generating} kernels to the following semimetrics:
+with \mbox{generating} kernels to the following metrics:
 \begin{subequations}
@@
 spectral clustering.
-For parallel cigars the semimetrics $\rho_1$ and $\rho_{1/2}$ still provide
+For parallel cigars the metrics $\rho_1$ and $\rho_{1/2}$ still provide
 accurate results, however, for
@@
 $k=6$ since its the first time we see a peak. We use energy clustering
-with the two semimetrics $\rho_1$ and $\widehat{\rho}_2$; see
+with the two metrics $\rho_1$ and $\widehat{\rho}_2$; see
 \eqref{eq:test_metrics}.
@@
 \emph{Fourth:} Heat map of the cluster means versus features
-using energy clustering with Gaussian semimetric 
+using energy clustering with Gaussian metric 
 $\widehat{\rho}_2$.
@@
 
-We now illustrate an application to real data of 
-$\mathcal{E}^H$-clustering algorithm based on Hartigan's method.
-This dataset describes the shapes of neural synapses and was collected
-by expert neuroscientists. It consists of $1025$ data points
+The considered dataset 
+describes protein expression of neural synapses (chemical 
+connections between neurons), being
+crucial for understanding the diversity of synapses. This is
+part of a large NIH funded consortium within the BRAIN initiative;
+we refer to \citet{Collman2015} for details.
+We have $1025$ data points
 with $11$ features. Each point
-corresponds to $(x,y,z)$ location in the brain. We normalize this
-dataset beforehand.
-
-We apply $\mathcal{E}^H$-clustering using the two
-semimetrics $\rho_1$ and $\widehat{\rho}_2$; see \eqref{eq:test_metrics}.
+corresponds to a $(x,y,z)$ location in the brain. We normalize this
+dataset before applying $\mathcal{E}^H$-clustering.
+
+We use the two
+metrics $\rho_1$ and $\widehat{\rho}_2$; see \eqref{eq:test_metrics}.
 We compute the difference between eigenvalues of the random walk Laplacian
 obtained from the kernel matrix $G$ \eqref{eq:kernel_matrix}.
-The choice of $k$ corresponds to the first time we see a meaningfull
+The choice of $k$ corresponds to the first time we see a meaningful
 peak in the plot, which occurs at $k=6$; first plot in
 Fig.~\ref{fig:synapse}.
-This procedure to find $k$ is commonn in the literature \cite{vonLuxburg2007}.
+This procedure to find $k$ is common in the literature \cite{vonLuxburg2007}.
 Having found $k$ we now cluster the data.
-First we use the semimetric $\rho_1$, which
+First we use the metric $\rho_1$, which
 is the standard Euclidean choice in energy statistics.
@@
 map of the cluster means versus features and the clustered data points
-projected into the 2 principal components, respectivelly.
+projected into the 2 principal components, respectively.
 The fourth and fifth plots in Fig.~\ref{fig:synapse} show the same
-experiment but using the semimetric $\widehat{\rho}_2$ corresponding
+experiment but using the metric $\widehat{\rho}_2$ corresponding
 to a Gaussian kernel.
@@
 Our mathematical formulation yields
-a QCQP in the associated kernel Hilbert space; 
-see  Proposition~\ref{th:qcqp2}.
-We demonstrated that such QCQP
+a QCQP in the associated RKHS;
+Proposition~\ref{th:qcqp2}.
+We showed that such QCQP
 is equivalent to kernel $k$-means optimization problem, 
-once the kernel is fixed; see
-Proposition~\ref{th:kernel_kmeans}. Energy statistics, however, fixes
+once the kernel is fixed; 
+Proposition~\ref{th:kernel_kmeans}. However, energy statistics fixes
 a family of standard kernels in Euclidean space, and
 more general kernels 
-on spaces of negative type can also be obtained.
+on spaces of negative type as well.
 
@@
 based on Lloyd's heuristic.
-Both have the same time complexity, however, numerical and theoretical
+Both have the same complexity, however, numerical and theoretical
 results provide compelling evidence that $\mathcal{E}^H$-clustering
-is more robust with a superior performance, specially in high
+is more robust with superior performance, specially in high
 dimensions. 
-Furthermore, $\mathcal{E}^H$-clustering with standard kernels from energy
+Moreover, $\mathcal{E}^H$-clustering with standard kernels from energy
 statistics outperformed $k$-means and GMM
@@
 of large matrices can be unfeasible and iterative methods are preferred.
-We provided an application of $\mathcal{E}^H$-clustering to a real dataset
-describing shapes of neural synapses.
+We provided an application of $\mathcal{E}^H$-clustering to an
+important real dataset
+describing protein expression of neural synapses.
 Finally, we remark that kernel methods can benefit from sparsity and
@@
 
-\subsection*{Acknowledgements}
-We would like to thank Carey Priebe 
-for discussions.
-We would like to acknowledge the support of the Transformative
-Research Award (NIH \#R01NS092474) and  the Defense Advanced Research Projects
-Agency’s (DARPA) SIMPLEX program through SPAWAR contract N66001-15-C-4041.
+%\subsection*{Acknowledgements}
+%We would like to thank Carey Priebe 
+%for discussions.
+%We would like to acknowledge the support of the Transformative
+%Research Award (NIH \#R01NS092474) and  the Defense Advanced Research Projects
+%Agency’s (DARPA) SIMPLEX program through SPAWAR contract N66001-15-C-4041.
 

### 2018-06-04 energy_clustering/old_stuff/icml18/rebuttal/rebuttal.txt
energy pami

* R1

We highly appreciate the referee's careful reading of the manuscript 
and comments.

We remark that this paper is the first to relate energy statistics, 
clustering and kernel methods, therefore we believe there is a theoretical 
significance.

The goal of the experiments was never to compete with standard clustering
methods, where a careful tuning of parameters and choice of kernels is
important. Rather, the experiments has the goal of comparing Hartigan's 
method versus Lloyd's method when applied to our QCQP problem, and also
to illustrate the flexibility of energy clustering.
We don't see much relevance in the fact that a scikit-learn implementation of 
spectral clustering, with a careful kernel choice, has a high performance 
in a particular example (we could for instance, introduce a local version
of energy clustering through more sophisticated methods, and we are confident
that this would be better than standard spectral clustering implementations).

* R2

We greatly appreciate the referee's analysis of our paper and comments.

We agree that we should emphasize better the two mentioned contributions.
However, the previous related works cited in the paper, only connects energy
statistics to clustering in a particular case (Euclidean) and there are no
connections to kernels at all, therefore, we believe our paper closes
an important gap and provides a more consistent derivation of clustering 
from energy statistics, in more general settings, and at the same time 
relates this approach to kernel methods.

To the best of our knowledge, Hartigan's method was never applied to
a general kernel method before. We believe this is the advantage of our
approach, where after one chooses a semi-metric, or equivalently, a positive 
definite kernel, one can just run the algorithm proposed in the paper.

We agree that the real data experiment needs a comparison and more detailed
discussion.

* R3

We highly appreciate the referee's careful reading of our manuscript 
and comments.

We kindly disagree with the statement "the contribution of the paper is quite
marginal given the previous work using energy statistics theory for clustering
purpose (e.g, Li (2015) for Euclidean case)". Our paper is the first to 
relate energy statistics, clustering, and kernel methods. Going from Euclidean 
to a general semi-metric with connections with RKHS is not straightforward.
For instance, in going from k-means to kernel k-means there is quite a big
gap and much research involved. The same is true for energy statistics.
It was proposed in the 80's, but only in 2013 connections to kernels were
established, and this was not previously explored in clustering.

We agree that it is desirable to establish the results of Theorems 4 and 5
under general kernels. We believe this is an interesting topic for future
work since it is not straightforward.

The point of Figure 1 is to illustrate that energy clustering attains high 
accuracy, under the same choice of parameters, in different settings. 
Thus, we still believe the comparison in Figure 1 is fair, regardless
if GMM is not the right model to the data.

### 2018-06-04 energy_clustering/old_stuff/aistats18/AISTATS2018PaperPack/Style file AISTATS 2018/sample_paper.tex
energy pami

\documentclass[twoside]{article}
\usepackage{aistats2018}
 % If your paper is accepted, change the options for the package
% aistats2018 as follows:
%
%\usepackage[accepted]{aistats2018}
%
% This option will print headings for the title of your paper and
% headings for the authors names, plus a copyright note at the end of
% the first column of the first page.

\begin{document}

% If your paper is accepted and the title of your paper is very long,
% the style will print as headings an error message. Use the following
% command to supply a shorter title of your paper so that it can be
% used as headings.
%
%\runningtitle{I use this title instead because the last one was very long}

% If your paper is accepted and the number of authors is large, the
% style will print as headings an error message. Use the following
% command to supply a shorter version of the authors names so that
% they can be used as headings (for example, use only the surnames)
%
%\runningauthor{Surname 1, Surname 2, Surname 3, ...., Surname n}

\twocolumn[

\aistatstitle{Instructions for paper submissions to AISTATS 2018}

\aistatsauthor{ Author 1 \And Author 2 \And  Author 3 }

\aistatsaddress{ Institution 1 \And  Institution 2 \And Institution 3 } ]

\begin{abstract}
  The Abstract paragraph should be indented 0.25 inch (1.5 picas) on
  both left and right-hand margins. Use 10~point type, with a vertical
  spacing of 11~points. The {\bf Abstract} heading must be centered,
  bold, and in point size 12. Two line spaces precede the
  Abstract. The Abstract must be limited to one paragraph.
\end{abstract}

\section{GENERAL FORMATTING INSTRUCTIONS}

The camera-ready versions of the accepted papers are 8 pages,
plus any additional pages needed for references.

Papers are in 2 columns with the overall line width of 6.75~inches (41~picas). Each column is 3.25~inches wide (19.5~picas).  The space
between the columns is .25~inches wide (1.5~picas).  The left margin is 1~inch (6~picas).  Use 10~point type with a vertical spacing of
11~points. Please use US Letter size paper instead of A4.

Paper title is 16~point, caps/lc, bold, centered between 2~horizontal rules.  Top rule is 4~points thick and bottom rule is 1~point thick.
Allow 1/4~inch space above and below title to rules.

Author descriptions are center-justified, initial caps.  The lead
author is to be listed first (left-most), and the Co-authors are set
to follow.  If up to three authors, use a single row of author
descriptions, each one center-justified, and all set side by side;
with more authors or unusually long names or institutions, use more
rows.

Use one-half line space between paragraphs, with no indent.

\section{FIRST LEVEL HEADINGS}

First level headings are all caps, flush left, bold, and in point size
12. Use one line space before the first level heading and one-half line space
after the first level heading.

\subsection{Second Level Heading}

Second level headings are initial caps, flush left, bold, and in point
size 10. Use one line space before the second level heading and one-half line
space after the second level heading.

\subsubsection{Third Level Heading}

Third level headings are flush left, initial caps, bold, and in point
size 10. Use one line space before the third level heading and one-half line
space after the third level heading.

\paragraph{Fourth Level Heading}

Fourth level headings must be flush left, initial caps, bold, and
Roman type.  Use one line space before the fourth level heading, and
place the section text immediately after the heading with no line
break, but an 11 point horizontal space.

\subsection{CITATIONS, FIGURES, REFERENCES}

\subsubsection{Citations in Text}

Citations within the text should include the author's last name and
year, e.g., (Cheesman, 1985). References should follow any style that
you are used to using, as long as their style is consistent throughout
the paper.  Be sure that the sentence reads correctly if the citation
is deleted: e.g., instead of ``As described by (Cheesman, 1985), we
first frobulate the widgets,'' write ``As described by Cheesman
(1985), we first frobulate the widgets.''  %Be sure to avoid
%accidentally disclosing author identities through citations.

\subsubsection{Footnotes}

Indicate footnotes with a number\footnote{Sample of the first
  footnote.} in the text. Use 8 point type for footnotes. Place the
footnotes at the bottom of the column in which their markers appear,
continuing to the next column if required. Precede the footnote
section of a column with a 0.5 point horizontal rule 1~inch (6~picas)
long.\footnote{Sample of the second footnote.}

\subsubsection{Figures}

All artwork must be centered, neat, clean, and legible.  All lines
should be very dark for purposes of reproduction, and art work should
not be hand-drawn.  Figures may appear at the top of a column, at the
top of a page spanning multiple columns, inline within a column, or
with text wrapped around them, but the figure number and caption
always appear immediately below the figure.  Leave 2 line spaces
between the figure and the caption. The figure caption is initial caps
and each figure should be numbered consecutively.

Make sure that the figure caption does not get separated from the
figure. Leave extra white space at the bottom of the page rather than
splitting the figure and figure caption.
\begin{figure}[h]
\vspace{.3in}
\centerline{\fbox{This figure intentionally left non-blank}}
\vspace{.3in}
\caption{Sample Figure Caption}
\end{figure}

\subsubsection{Tables}

All tables must be centered, neat, clean, and legible. Do not use hand-drawn tables. Table number and title always appear above the table.
See Table~\ref{sample-table}.

Use one line space before the table title, one line space after the table title, and one line space after the table. The table title must be
initial caps and each table numbered consecutively.

\begin{table}[h]
\caption{Sample Table Title} \label{sample-table}
\begin{center}
\begin{tabular}{ll}
{\bf PART}  &{\bf DESCRIPTION} \\
\hline \\
Dendrite         &Input terminal \\
Axon             &Output terminal \\
Soma             &Cell body (contains cell nucleus) \\
\end{tabular}
\end{center}
\end{table}

\section{SUPPLEMENTARY MATERIAL}

If you need to include additional appendices during submission, you
can include them in the supplementary material file.

\section{INSTRUCTIONS FOR CAMERA-READY PAPERS}

For the camera-ready paper, if you are using \LaTeX, please make sure
that you follow these instructions.  (If you are not using \LaTeX,
please make sure to achieve the same effect using your chosen
typesetting package.)

\begin{enumerate}
    \item Download \texttt{fancyhdr.sty} -- the
    \texttt{aistats2018.sty} file will make use of it.
    \item Begin your document with
    \begin{flushleft}
    \texttt{\textbackslash documentclass[twoside]\{article\}}\\
    \texttt{\textbackslash usepackage[accepted]\{aistats2018\}}
    \end{flushleft}
    The \texttt{twoside} option for the class article allows the
    package \texttt{fancyhdr.sty} to include headings for even and odd
    numbered pages. The option \texttt{accepted} for the package
    \texttt{aistats2018.sty} will write a copyright notice at the end of
    the first column of the first page. This option will also print
    headings for the paper.  For the \emph{even} pages, the title of
    the paper will be used as heading and for \emph{odd} pages the
    author names will be used as heading.  If the title of the paper
    is too long or the number of authors is too large, the style will
    print a warning message as heading. If this happens additional
    commands can be used to place as headings shorter versions of the
    title and the author names. This is explained in the next point.
    \item  If you get warning messages as described above, then
    immediately after $\texttt{\textbackslash
    begin\{document\}}$, write
    \begin{flushleft}
    \texttt{\textbackslash runningtitle\{Provide here an alternative shorter version of the title of your
    paper\}}\\
    \texttt{\textbackslash runningauthor\{Provide here the surnames of the authors of your paper, all separated by
    commas\}}
    \end{flushleft}
    Note that the text that appears as argument in \texttt{\textbackslash
      runningtitle} will be printed as a heading in the \emph{even}
    pages. The text that appears as argument in \texttt{\textbackslash
      runningauthor} will be printed as a heading in the \emph{odd}
    pages.  If even the author surnames do not fit, it is acceptable
    to give a subset of author names followed by ``et al.''

    \item Use the file sample\_paper.tex as an example.

    \item The camera-ready versions of the accepted papers are 8
      pages, plus any additional pages needed for references.

    \item If you need to include additional appendices,
      you can include them in the supplementary
      material file.

    \item Please, don't change the layout given by the above
      instructions and by the style file.

\end{enumerate}

\subsubsection*{Acknowledgements}

Use unnumbered third level headings for the acknowledgements.  All
acknowledgements go at the end of the paper.

\subsubsection*{References}

References follow the acknowledgements.  Use an unnumbered third level
heading for the references section.  Any choice of citation style is
acceptable as long as you are consistent.  Please use the same font
size for references as for the body of the paper---remember that
references do not count against your page length total.

J.~Alspector, B.~Gupta, and R.~B.~Allen (1989). Performance of a
stochastic learning microchip.  In D. S. Touretzky (ed.), {\it
  Advances in Neural Information Processing Systems 1}, 748-760.  San
Mateo, Calif.: Morgan Kaufmann.

F.~Rosenblatt (1962). {\it Principles of Neurodynamics.} Washington,
D.C.: Spartan Books.

G.~Tesauro (1989). Neurogammon wins computer Olympiad.  {\it Neural
  Computation} {\bf 1}(3):321-323.

\end{document}

### 2018-06-04 energy_clustering/old_stuff/aistats18/energy_clustering.tex
energy pami

--- 2017-10-13 energy_clustering/aistats18/energy_clustering.tex
@@
 \subsection*{Acknowledgements}
-We would like to thank XXX
-%Carey Priebe 
+We would like to thank Carey Priebe 
 for discussions.
-This work was supported by 
-%NIH TRA 
-XXX
-grant.
+We would like to acknowledge the support of the Transformative
+Research Award (NIH #R01NS092474) and  the Defense Advanced Research Projects
+Agency’s (DARPA) SIMPLEX program through SPAWAR contract N66001-15-C-4041.
 

### 2018-06-04 energy_clustering/old_stuff/code/synapse_data2/Extras/gsheet.Rmd
energy pami

---
title: ""
date: '`r Sys.Date()`'
output:
  html_document:
    fig_caption: yes
    fig_height: 5
    fig_width: 5
    highlight: pygments
    number_sections: yes
    theme: cerulean
    toc: yes
    toc_depth: 2
---
```{r knitOPTS, include=FALSE}
knitr::opts_chunk$set(cache = TRUE, dev = "png")
```

```{r render, eval=FALSE, echo=FALSE}
require(rmarkdown)
#rmarkdown::render("collman14v2_meda_plots.Rmd")
#system("open collman14v2_meda_plots.html")
```

```{r setup,include=FALSE,results='asis',message=FALSE,warning=FALSE, echo = FALSE}
# Library calls here.
require(rmarkdown)
require(knitr)
require(rhdf5)
require(foreach)
require(googlesheets)
suppressMessages(require(meda))
```

## Make plots

```{r, eval = FALSE}
gs_auth(new_user = TRUE)
gs_ls()

tmp <- gs_title("MNSite3Synaptograms.csv")
#dat <- as.data.frame(gs_read_csv(ss = tmp, ws = 1, skip = 1))
dat <- as.data.table(gs_read_csv(ss = tmp, ws = 1, skip = 1))

dat$cx <- 
  (dat$maxX - dat$minX)/2 + dat$minX

dat$cy <- 
  (dat$maxY - dat$minY)/2 + dat$minY

dat$cz <- 
  (dat$maxZ - dat$minZ)/2 + dat$minZ

dat$atx <- round((dat$cx * 3) / 96)
dat$aty <- round((dat$cy * 3) / 96)
dat$atz <- round(dat$cz)

dat$nmx <- (dat$cx / 3)
dat$nmy <- (dat$cy / 3)
dat$nmz <- dat$cz/50

dx <- 438
dy <- 460
dz <- 49

ind <- 
  (dat$atx - 5) >= 0  &
  (dat$aty - 5) >= 0  &
  (dat$atz - 5) >= 0  &
  (dat$atx + 6) <= dx &
  (dat$aty + 6) <= dy &
  (dat$atz + 6) <= dz

dat$buff <- as.numeric(ind)

dat$gaba <- dat$GABA
newDat <- dat[buff == 1, .(atx,aty,atz, id, gaba, shaft, TdTom, NotSynapse)]

write.csv(newDat, file = "GoogleDocData.csv", row.names = FALSE)
write.csv(newDat[, .(atx,aty,atx,id)], file = "GoogleDocData_loc.csv", row.names = FALSE)

inIDS <- as.numeric(read.csv("rorb_avg_at_orderLocations.csv")[[1]])
IDS123 <- as.numeric(read.csv("rorb_avg_at_IDS123.csv")[[1]])

#L <- foreach(i = inIDS, .combine = 'rbind') %do% {
L <- foreach(i = IDS123, .combine = 'rbind') %do% {
  dat[id == i,]
}

glut <- L$glut
glutIDS <- data.table(id = inIDS, glut = as.numeric(glut))
write.csv(glutIDS, file = "rorb_glut.csv", row.names = FALSE)

gaba <- L$gaba
gabaIDS <- data.table(id = inIDS, gaba = as.numeric(gaba))
write.csv(gabaIDS, file = "rorb_gaba.csv", row.names = FALSE)

shaft <- L$shaft
shaftIDS <- data.table(id = inIDS, shaft = as.numeric(shaft))
write.csv(shaftIDS, file = "rorb_shaft.csv", row.names = FALSE)

TdTom <- L$TdTom
TdIDS <- data.table(id = inIDS, TdTom = as.numeric(TdTom))
write.csv(TdIDS, file = "rorb_TdTom.csv", row.names = FALSE)

write.csv(L, file = "rorb_id123subset.csv", row.names = FALSE)
```

### 2018-06-04 energy_clustering/pami18/energy_clustering.tex
energy pami

\documentclass[10pt,journal,compsoc]{IEEEtran}

\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts,amsthm,amscd,bm,bbm}
\usepackage{mathtools}
%\mathtoolsset{showonlyrefs}
%\usepackage[inline]{enumitem}
\usepackage[pdftex]{graphicx}
\graphicspath{{./figs/}}
\usepackage{algorithmic}
\usepackage{algorithm}
%\usepackage[none]{hyphenat}
\usepackage{multirow}
\usepackage{colortbl}
\usepackage{array,booktabs}
\usepackage{xcolor}
\usepackage{makecell}
\usepackage{pifont}
\newcommand{\cmark}{\ding{51}}
\newcommand{\xmark}{\ding{55}}

\newtheorem{theorem}{Theorem}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{assumption}[theorem]{Assumption}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{example}{Example}

\DeclareMathOperator{\aff}{aff}
\DeclareMathOperator{\st}{s.t.}
\DeclareMathOperator{\affnot}{aff_0}
\DeclareMathOperator{\conv}{conv}
\DeclareMathOperator{\relint}{relint}
\DeclareMathOperator{\vol}{vol}
\DeclareMathOperator{\range}{range}
\DeclareMathOperator{\image}{im}
\DeclareMathOperator{\nullspace}{null}
\DeclareMathOperator{\area}{area}
\DeclareMathOperator{\vspan}{span}
\DeclareMathOperator{\id}{Id}
\DeclareMathOperator{\cond}{cond}
\DeclareMathOperator{\prox}{prox}
\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator*{\minimize}{minimize}
\DeclareMathOperator{\diag}{diag}
\DeclareMathOperator{\Tr}{Tr}

\newcommand\Energy{\mathcal{E}}
\newcommand\EnergyH{\mathcal{E}^{H}}
\newcommand\EnergyL{\mathcal{E}^{L}}
\newcommand\EnergyOne{\mathcal{E}^{1D}}
\newcommand\E{\mathbb{E}}
\newcommand\kk{K}
\newcommand\kkk{h}
\newcommand\Hk{{\mathcal{H}}_{\kk}}
\newcommand\HH{\mathcal{H}}
\newcommand\C{{\mathcal{C}}}
\newcommand\tC{{\widetilde{\C}}}
\newcommand\OO{{\mathcal{O}}}
\newcommand\W{{\mathcal{W}}}
\newcommand\Zt{Y}
\newcommand{\Ind}[1]{\mathbbm{1}_{#1}}
\newcommand\e{e}
\newcommand\om{\omega}

\hyphenation{op-tical net-works semi-conduc-tor}

\begin{document}

\title{Kernel k-Groups and Hartigan's Method}

\author{Guilherme~Fran\c ca,~Maria Rizzo~and~Joshua~Vogelstein
\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem G. Fran\c ca 
and J. Vogelstein are with the Center for Imaging Science, Johns Hopkins
University; J. Vogelstein is also with 
the Department of Biomedical Engineering and Institute for Computational
Medicine, Johns Hopkins University. \protect\\
% note need leading \protect in front of \\ to get a newline within \thanks as
% \\ is fragile and will error, could use \hfil\break instead.
E-mail: guifranca@jhu.edu; jovo@jhu.edu
\IEEEcompsocthanksitem M. Rizzo is with the Department of Mathematics
and Statistics, Bowling Green State University. \protect\\
E-mail: mrizzo@bgsu.edu}% <-this % stops an unwanted space
\thanks{Manuscript received March 15, 2018; revised March 15, 2018.}}

% The paper headers
\markboth{Journal of \LaTeX\ Class Files,~Vol.~14, No.~8, August~2015}%
{Energy Clustering}

\IEEEtitleabstractindextext{%
\begin{abstract}
Energy statistics was proposed by Sz\' ekely in the 80's inspired by 
Newton's gravitational potential in classical mechanics, and it provides
a model-free 
hypothesis test for equality of distributions. In its original form, energy
statistics was formulated in 
Euclidean spaces. More recently, it 
was generalized to metric spaces of negative type. 
In this paper, we consider a formulation for the clustering problem
using a weighted version of 
energy statistics in spaces of negative type.
We show that this approach leads to a
quadratically constrained
quadratic program in the associated kernel space, establishing
connections with graph partitioning problems and
kernel methods in unsupervised machine learning.
To find local solutions of such an optimization problem, 
we propose an extension of Hartigan's method to kernel spaces.
Our method has the same computational cost
as kernel k-means algorithm, which is based on Lloyd's heuristic, but 
our numerical results show an improved performance,
specially in high dimensions.
\end{abstract}

% Note that keywords are not normally used for peerreview papers.
\begin{IEEEkeywords}
Clustering, Energy Statistics, Kernel Methods.
\end{IEEEkeywords}}

% make the title area
\maketitle

% To allow for easy dual compilation without having to reenter the
% abstract/keywords data, the \IEEEtitleabstractindextext text will
% not be used in maketitle, but will appear (i.e., to be "transported")
% here as \IEEEdisplaynontitleabstractindextext when the compsoc 
% or transmag modes are not selected <OR> if conference mode is selected 
% - because all conference papers position the abstract like regular
% papers do.
\IEEEdisplaynontitleabstractindextext
% \IEEEdisplaynontitleabstractindextext has no effect when using
% compsoc or transmag under a non-conference mode.

% For peer review papers, you can put extra information on the cover
% page as needed:
% \ifCLASSOPTIONpeerreview
% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
% \fi
%
% For peerreview papers, this IEEEtran command inserts a page break and
% creates the second title. It will be ignored for other modes.
\IEEEpeerreviewmaketitle

\IEEEraisesectionheading{\section{Introduction}\label{sec:introduction}}
% Computer Society journal (but not conference!) papers do something unusual
% with the very first section heading (almost always called "Introduction").
% They place it ABOVE the main text! IEEEtran.cls does not automatically do
% this for you, but you can achieve this effect with the provided
% \IEEEraisesectionheading{} command. Note the need to keep any \label that
% is to refer to the section immediately after \section in the above as
% \IEEEraisesectionheading puts \section within a raised box.

% The very first letter is a 2 line initial drop letter followed
% by the rest of the first word in caps (small caps for compsoc).
% 
% form to use if the first word consists of a single letter:
% \IEEEPARstart{A}{demo} file is ....
% 
% form to use if you need the single drop letter followed by
% normal text (unknown if ever used by the IEEE):
% \IEEEPARstart{A}{}demo file is ....
% 
% Some journals put the first two words in caps:
% \IEEEPARstart{T}{his demo} file is ....
% 
% Here we have the typical use of a "T" for an initial drop letter
% and "HIS" in caps to complete the first word.

\IEEEPARstart{E}{nergy Statistics} \cite{Szkely2013,Szkely2017}
is based on a 
notion of statistical potential energy between probability distributions,
in close analogy to Newton's gravitational potential in classical mechanics.
When probability distributions are different, the 
``statistical potential energy'' diverges as sample size increases, 
while tends 
to a nondegenerate limit distribution when probability
distributions are equal. 
Thus, it provides a model-free hypothesis test for equality of 
distributions which is achieved under minimum energy.

Energy statistics has been applied to several goodness-of-fit 
hypothesis tests, multi-sample tests of equality of distributions, 
analysis of variance \cite{RizzoVariance}, nonlinear dependence tests through
distance covariance and distance correlation, which generalizes the Pearson
correlation coefficient, and hierarchical clustering 
by extending Ward's method of minimum variance
\cite{RizzoClustering}; 
see \cite{Szkely2013,Szkely2017} for an overview of energy
statistics and its applications.
Moreover, in Euclidean spaces, an application of 
energy statistics to clustering
was recently proposed \cite{Kgroups}, and the method was named 
\emph{k-groups}.

In its original formulation, energy statistics has a compact representation
in terms of expectations of pairwise Euclidean distances, providing
straightforward empirical estimates. 
More recently, the notion of distance covariance was further 
generalized from Euclidean 
spaces to metric spaces of negative type \cite{Lyons}. Furthermore, 
the link between energy distance based tests and kernel 
based tests has 
been recently established \cite{Sejdinovic2013} 
through an equivalence between generalized energy distances and maximum
mean discrepancies (MMD), which are distances between embeddings of 
distributions in reproducing kernel Hilbert spaces (RKHS).
This equivalence immediately relates energy statistics to
kernel methods often used in machine learning, and form the basis 
of our approach in this paper.

Clustering is an important unsupervised learning problem and 
has such a long history in statistics and machine learning, making it
impossible to mention all important contributions in a short space. 
Perhaps, the most used method is k-means \cite{Lloyd,MacQueen,Forgy}, which
is based on Lloyd's heuristic \cite{Lloyd} of iteratively computing
the means of each cluster and then assigning points to
the cluster with closest center. The only statistical 
information about each cluster comes from its mean, making the method sensitive 
to outliers. Nevertheless, k-means works very well when data is 
linearly separable in Euclidean space. Gaussian mixture models (GMM) is 
another very common approach, providing more flexibility than k-means, 
however, it still makes strong assumptions about the distribution of 
the data.

To account for nonlinearities, kernel methods were introduced 
\cite{Smola,Girolami}. A mercer kernel \cite{Mercer} is used to implicitly
map data points to a RKHS, then clustering can be performed in the associated
Hilbert space by using its inner product. However, the kernel choice remains 
the biggest challenge since there is no principled theory to construct a kernel
for a given dataset, and usually a kernel introduces hyperparameters that 
need to be carefully chosen. A well-known kernel based clustering method
is kernel k-means, which is precisely k-means 
formulated in the feature space \cite{Girolami}. 
Furthermore, kernel k-means algorithm
\cite{Dhillon2,Dhillon} is still based on Loyd's heuristic.
We refer the reader to \cite{Filippone} for a survey of clustering
methods.

The predominant approach to clustering %, in Euclidean or kernel spaces,
is based on Lloyd's method.
However, there is an old heuristic
due to Hartigan \cite{Hartigan1975,Hartigan1979}, which apparently did
not receive the deserved attention, 
and goes as follows: for each data point, simply assign it to a cluster
in an optimal way such that a loss function is minimized.
While Lloyd's method only iterates if some cluster contains a point
that is closer to the mean of another cluster, Hartigan's method may iterate
even if that is not the case, and moreover, it takes into account the motion
of the means resulting from the reassignments. In this sense, Hartigan's
method may potentially escape local minima of Lloyd's method.
In the Euclidean case, this was shown to be the case \cite{Telgarsky}.
Moreover, the advantages of Hartigan's over Lloyd's method was verified
empirically \cite{Telgarsky,Slonin}, however, 
although it was observed to be as fast as Lloyd's method, no complexity
analysis was provided.

\subsection*{Contributions}
Although k-groups considers clustering from energy statistics
in the particular Euclidean case \cite{Kgroups},
the precise optimization problem behind this approach
remains obscure, as well as the connection with other methods
in machine learning.
The main theoretical contribution of this paper is to fill these gaps, 
which 
we do so in more generality. For instance, our approach is not limited
to the Euclidean case but holds for general arbitrary spaces
of negative type. 
For generality, we also
consider a weighted version of energy statistics.
Our approach reveals connections
between energy statistics based clustering and existing methods
such as kernel k-means and graph partitioning problems.

Another contribution of this paper is to extend 
Hartigan's method to  kernel spaces. 
To the best of our knowledge, such an extension was not previously considered.
Since this approach was motivated by energy statistics and  
\cite{Kgroups} considered the Euclidean case, we call the 
proposed method \emph{kernel k-groups}.
We show that kernel k-groups 
has the same complexity as kernel k-means algorithm, however,
our numerical results provide compelling evidence that 
kernel k-groups 
is more accurate and robust, specially in high dimensions.

Using the standard kernel defined by energy
statistics, our experiments illustrate 
that kernel k-groups
is able to perform accurately on data coming from 
very different distributions, contrary to k-means and GMM, for instance.
More specifically, our method performs 
closely to k-means and GMM on normally distributed data, however,
it is significantly better on data that 
is not normally distributed. 
Its superiority in high dimensions
is striking, being more accurate than k-means and GMM 
even on Gaussian settings. We also illustrate the advantages of
kernel k-groups on a real dataset.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Review of Energy Statistics and RKHS}
\label{sec:background}

In this section, we introduce the main concepts from energy
statistics and its relation to 
RKHS which form the basis of our work.
For more details we refer 
to \cite{Szkely2013} and \cite{Lyons,Sejdinovic2013}.

Consider random variables in $\mathbb{R}^D$ 
such that $X,X' \stackrel{iid}{\sim} P$ and 
$Y,Y' \stackrel{iid}{\sim} Q$, where $P$ and $Q$ are cumulative
distribution functions with finite first moments. 
The quantity 
\begin{equation}
\label{eq:energy}
\Energy(P, Q) \equiv 2 \E \| X - Y\| - \E \| X - X' \| - \E \| Y - Y' \|,
\end{equation}
called \emph{energy distance} \cite{Szkely2013}, 
is rotationally invariant and nonnegative, $\Energy(P,Q) \ge 0$, where
equality
to zero holds if and only if $P = Q$.
Above, $\| \cdot \|$ denotes the
Euclidean norm in $\mathbb{R}^D$. 
Energy distance
provides a characterization of equality of distributions, and
$\Energy^{1/2}$ is
a metric on the space of distributions.

The energy distance can be generalized as, for instance,
\begin{equation}
\label{eq:energy2}
\Energy_\alpha(P, Q) \equiv 
2 \E \| X - Y\|^{\alpha} - \E \| X - X' \|^{\alpha} - 
\E \| Y - Y' \|^{\alpha}
\end{equation}
where $0<\alpha\le 2$. This quantity is also nonnegative,
$\Energy_\alpha(P,Q) \ge 0$. Furthermore, for $0<\alpha<2$ we have that
$\Energy_\alpha(P,Q) = 0$ if and only if $P=Q$, while for $\alpha=2$ 
we have $\Energy_2(P,Q) = 2\| \E(X) - \E(Y) \|^2$ which shows that
equality to zero only requires
equality of the means, and thus $\Energy_2(P,Q)=0$ does 
not imply equality of distributions.

The energy distance can be even further generalized.
Let $X, Y \in \mathcal{X}$  where $\mathcal{X}$ is an arbitrary space endowed
with a \emph{semimetric of negative type}
$\rho: \mathcal{X}\times\mathcal{X} \to \mathbb{R}$, which is required
to satisfy
\begin{equation}
\label{eq:negative_type}
\sum_{i,j=1}^n c_i c_j \rho(X_i, X_j) \le 0,
\end{equation}
where $X_i \in \mathcal{X}$ and $c_i \in \mathbb{R}$ such that
$\sum_{i=1}^n c_i = 0$. Then, $\mathcal{X}$ is called a \emph{space of
negative type}.
We can thus replace $\mathbb{R}^D$ by $\mathcal{X}$ and 
$\| X - Y \|$ by $\rho(X , Y)$ in the definition \eqref{eq:energy}, obtaining
the \emph{generalized energy distance}
\begin{equation}
\label{eq:energy3}
\Energy(P, Q) \equiv 2 \E \rho(X,Y) - \E \rho(X, X') - \E \rho(Y,Y').
\end{equation}
For spaces of negative type, there exists a Hilbert space $\mathcal{H}$ and
a map $\varphi: \mathcal{X} \to
\mathcal{H}$ such that
$\rho(X, Y) = \| \varphi(X) - \varphi(Y) \|_{\mathcal{H}}^2$. This
allows us to compute quantities related to probability distributions over
$\mathcal{X}$ in the associated Hilbert space $\mathcal{H}$.
Even though the semimetric 
$\rho$ may not satisfy the triangle inequality, 
$\rho^{1/2}$ does since it can be shown to be a proper metric. 
Our energy clustering formulation, proposed in the next section,
will be based on the generalized
energy distance \eqref{eq:energy3}.

There is an equivalence between energy distance, 
commonly used in statistics,
and distances between embeddings of distributions in 
RKHS, commonly used in machine learning. 
This equivalence was established
in \cite{Sejdinovic2013}. Let us first recall the definition of
RKHS. Let $\HH$ be a Hilbert space of real-valued functions
over $\mathcal{X}$. A function 
$\kk : \mathcal{X} \times \mathcal{X} \to 
\mathbb{R}$ is a reproducing kernel of $\HH$ if it satisfies
the following two conditions:
\begin{enumerate}
\item $\kkk_x \equiv \kk(\cdot, x) \in \HH$ 
for all $x \in \mathcal{X}$;
\item $\langle \kkk_x, f \rangle_{\HH} = f(x)$ for
all $x\in\mathcal{X}$ and $f\in \HH$.
\end{enumerate}
In other words, for any $x \in \mathcal{X}$ and any function $f \in \HH$,
there is a unique 
$\kkk_x \in \HH$ that reproduces $f(x)$ through the inner product
of $\HH$.
If such a \emph{kernel} 
function $\kk$ exists, then $\HH$ is called a RKHS. The above two 
properties immediately imply that $\kk$ is symmetric and positive
definite. 
Defining the Gram matrix $G$ with
elements $G_{ij} = \kk(x_i,x_j)$, this is equivalent to $G=G^\top$ being
positive semidefinite, i.e., $v^\top G \, v \ge 0$ for any vector
$v \in \mathbb{R}^n$.

The Moore-Aronszajn theorem 
\cite{Aronszajn}
establishes the converse of the above paragraph.
For every symmetric
and positive definite function $\kk: \mathcal{X}\times \mathcal{X} \to
\mathbb{R}$, there is an associated RKHS, $\Hk$,
with reproducing
kernel $\kk$. The map $\varphi: x \mapsto \kkk_x \in \Hk$ is called
the canonical \emph{feature map}. Given a kernel $\kk$,
this theorem enables us to define an embedding of a probability measure
$P$ into the RKHS as follows: $P \mapsto \kkk_P \in
\Hk$ such that 
$\int f(x) d P(x) = \langle f, \kkk_P \rangle$ for all $f \in \Hk$,
or alternatively, $\kkk_P \equiv \int \kk( \, \cdot \,, x)  d P(x)$. 
We can now  introduce the 
notion of distance between two probability measures using the inner product
of $\Hk$, which is called the maximum mean discrepancy (MMD) and
is given by
\begin{equation}
\label{eq:mmd}
\gamma_\kk(P,Q) \equiv \| \kkk_P - \kkk_Q \|_{\Hk}.
\end{equation}
This can also be written as \cite{Gretton2012}
\begin{equation}\label{eq:mmd2}
\gamma_\kk^2(P,Q) = \E \kk(X,X') + \E \kk(Y,Y') - 2 \E \kk(X, Y)
\end{equation}
where $X,X' \stackrel{iid}{\sim} P$ and $Y,Y'\stackrel{iid}{\sim} Q$.
From the equality between \eqref{eq:mmd} and \eqref{eq:mmd2} we also
have $\langle \kkk_P, \kkk_Q \rangle_{\Hk} = \E \, \kk(X, Y)$.

The following important result shows that semimetrics of negative
type and symmetric positive definite kernels are closely related
\cite{Berg1984}. Let $\rho: \mathcal{X} \times \mathcal{X} \to \mathbb{R}$
and $x_0 \in \mathcal{X}$ an arbitrary but fixed point.
Define
\begin{equation}
\label{eq:kernel_semimetric}
\kk(x,y) \equiv 
\tfrac{1}{2} \left[  \rho(x,x_0) + \rho(y,x_0) - \rho(x,y)\right].
\end{equation}
Then, it can be shown that 
$\kk$ is positive definite if and only if $\rho$ is a semimetric
of negative type.
We have a family of kernels, one for each choice of $x_0$. Conversely,
if $\rho$ is a semimetric of negative type and $\kk$ is a kernel in this
family, then 
\begin{equation}
\label{eq:gen_kernel}
\begin{split}
\rho(x,y) &= \kk(x,x) + \kk(y,y) -2\kk(x,y) \\
&=  \| \kkk_x - \kkk_y \|^2_{\Hk}
\end{split}
\end{equation}
and the canonical feature map 
$\varphi: x \mapsto \kkk_x$ is injective \cite{Sejdinovic2013}.
When these conditions are satisfied, we say that the kernel $\kk$ 
generates the semimetric $\rho$. 
If two different kernels generate the same $\rho$, they are
said to be equivalent kernels.

Now we can state the equivalence between the generalized 
energy distance \eqref{eq:energy3} and
inner products on RKHS, which is one of the main results of
\cite{Sejdinovic2013}. If $\rho$ is a semimetric
of negative type and $\kk$ a kernel that generates $\rho$, then
replacing \eqref{eq:gen_kernel} into
\eqref{eq:energy3}, and using \eqref{eq:mmd2}, yields
\begin{equation} \label{eq:Erho}
\begin{split}
\Energy(P, Q) &= 
2 \left[ \E \, \kk(X, X') + \E \, \kk(Y, Y') - 2\E \, \kk(X, Y)\right]  \\
&= 2 \gamma_\kk^2(P,Q) .
\end{split}
\end{equation}
Due to \eqref{eq:mmd}, we can compute the energy 
distance $\mathcal{E}(P, Q)$ between two probability distributions
using the inner 
product of $\Hk$.

Finally, let us recall the main formulas from generalized energy statistics
for the test statistic of equality of distributions \cite{Szkely2013}. 
Assume that we have data $\mathbb{X} = \{ x_1,\dotsc, x_n \}$, where
$x_i \in \mathcal{X}$, and $\mathcal{X}$ is a space of negative type.
Consider a disjoint partition $\mathbb{X} = \bigcup_{j=1}^k \C_j$, with
$\C_i \cap \C_j = \emptyset$.
Each expectation in the generalized energy distance
\eqref{eq:energy3}
can be computed 
through the function
\begin{equation}
\label{eq:g_def}
g (\C_i, \C_j) \equiv 
\dfrac{1}{n_i n_j}
\sum_{x \in \C_i} 
\sum_{y \in \C_j} \rho(x, y) ,
\end{equation}
where $n_i = |\C_i|$ is the number of elements in partition
$\C_i$. 
The \emph{within energy dispersion} is defined by
\begin{equation}
\label{eq:within}
W \equiv
\sum_{j=1}^{k} \dfrac{n_j}{2} g(\C_j, \C_j),
\end{equation}
and the \emph{between-sample energy statistic} is defined by
\begin{equation}
\label{eq:between}
S \equiv \!\!\!\!
\sum_{1 \le  i < j \le k } \dfrac{n_i n_{j}}{2 n} \left[
2 g(\C_i, \C_j) - 
g(\C_i, \C_i) - 
g(\C_j, \C_j)
\right],
\end{equation}
where $n = \sum_{j=1}^k n_j$.
Given a set of distributions
$\{ P_j\}_{j=1}^k$, where $x \in \C_j$ if and only if $x \sim P_j$, 
the quantity $S$ provides
a test statistic for equality of distributions
\cite{Szkely2013}.
When the sample size is large enough, $n\to \infty$,
under the null hypothesis $H_0: P_1=P_2=\dotsm=P_k$, we have that
$S\to 0$, 
and under
the alternative hypothesis $H_1: P_i \ne P_j$ for at least two $i\ne j$, 
we have that $S \to \infty$.
%Note that this test does not make any assumptions
%about the form of distribution $P_j$, thus it is said to be 
%distribution-free.

%One can make a physical analogy by thinking 
%that points $ x \in \C_j$ form a massive body 
%whose total mass is characterized by the distribution function $P_j$.
%The quantity $S$ is thus a potential
%energy of the from $S(P_1,\dotsc,P_k)$ which measures how different
%the distribution of these masses are,  and achieves the ground state
%$S=0$ when all bodies have the same mass distribution. On the other hand,
%the potential energy
%$S$ increases as bodies have different mass distributions.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The Clustering Problem Formulation}
\label{sec:clustering_theory}

This section contains our main theoretical results. First, we
generalize the previous formulas from energy statistics by introducing
weights associated to data points. 
Second, we formulate an optimization problem for clustering 
in the associated RKHS, making connection with kernel methods in
machine learning.

Let $w(x)$ be a weight function associated to point $x \in \mathcal{X}$
and define
\begin{equation}
\label{eq:g_def2}
g(\C_i, \C_j) \equiv \dfrac{1}{s_i s_j} \sum_{x\in \C_i}\sum_{y\in\C_j}
w(x)w(y) \rho(x,y),
\end{equation}
where
\begin{equation}
s_i \equiv \sum_{x\in\C_i} w(x), \qquad
s \equiv \sum_{j=1}^k s_j.
\end{equation}
The  weighted version of the within energy dispersion 
and between-sample
energy statistic are thus given by
\begin{align}
W &\equiv
\sum_{j=1}^{k} \dfrac{s_j}{2} g(\C_j, \C_j), \label{eq:within2} \\
S &\equiv \!\!\! \sum_{1 \le  i < j \le k } \!\!\! 
\dfrac{s_i s_{j}}{2 s} \left[
2 g(\C_i, \C_j) - 
g(\C_i, \C_i) - 
g(\C_j, \C_j)
\right]. \label{eq:between2}
\end{align}
Note that if $w(x) = 1$ for every $x$ we recover the previous formulas.

Due to the test statistic for equality of distributions,
the obvious criterion for clustering data is to 
maximize $S$ in \eqref{eq:between2},
which makes  each cluster as different as possible from the other ones.
In other words, given a set of points coming from different probability
distributions, the test statistic $S$ should attain a maximum when 
each point is correctly classified as belonging to the cluster associated 
to its probability distribution. The following 
result
shows that maximizing $S$ is, however, equivalent to minimizing
$W$ in \eqref{eq:within2}.% which has a more convenient form.

\begin{lemma}
\label{th:minimize}
Let $\mathbb{X} = \{x_1,\dotsc,x_n\}$ where each data point
$x_i$ lives in a space $\mathcal{X}$ endowed with a semimetric $\rho:
\mathcal{X}\times\mathcal{X} \to \mathbb{R}$ of
negative type. For a fixed integer $k$,
the partition
$\mathbb{X} = \bigcup_{j=1}^k \C^\star_j$, where 
$\C^\star_i \cap C^\star_j = \emptyset$ for
all $i\ne j$, maximizes the between-sample statistic $S$, defined
in equation \eqref{eq:between2}, if and only if
\begin{equation}
\label{eq:minimize}
\{ \C_1^\star,\dotsc,\C_k^\star \} = \argmin_{\C_1,\dotsc,C_k  } W(
\C_1, \dotsc, \C_k) ,
\end{equation}
where the within energy dispersion $W$ is defined by \eqref{eq:within2}.
\end{lemma}
\begin{proof}
From \eqref{eq:within2} and \eqref{eq:between2}
we have that
\begin{align}
%\begin{equation}
%\begin{split}
& S + W  
\nonumber \\
&= \dfrac{1}{2s} \sum_{\substack{i,j=1 \\ i\ne j}}^k s_i s_j g(\C_i, \C_j)
+ \dfrac{1}{2s} \sum_{i=1}^{k} 
\bigg[ s - 
\sum_{\substack{j= 1 \\ j\ne i}}^k s_j \bigg] 
s_i g(\C_i, \C_i) 
\nonumber \\
& = \dfrac{1}{2s} \sum_{i,j=1}^k s_i s_j g(\C_i, \C_j) 
\nonumber \\
& = \dfrac{1}{2s} \sum_{x \in \mathbb{X}} \sum_{y \in \mathbb{X}} 
w(x) w(y)\rho(x,y)  
\nonumber \\
&= \dfrac{s}{2} g(\mathbb{X}, \mathbb{X}).
%\end{split}
%\end{equation}
\end{align}
Since $g(\mathbb{X}, \mathbb{X})$ is independent 
of the choice of partition, $\max_{\{ \C_i \}} S = - \max_{\{\C_i\}} W = 
\min_{\{ \C_i \}}W$, as claimed.
\end{proof}

For a given $k$, the clustering problem amounts to
finding the best partitioning of the data by minimizing $W$.
In the current form of problem \eqref{eq:minimize},
the relationship
with other clustering methods or kernel spaces is totally obscure.
In the following, we demonstrate what is the explicit 
optimization problem behind 
\eqref{eq:minimize} in the corresponding RKHS, 
which establishes the connection with kernel methods.

Based on the relationship between kernels and semimetrics of negative
type, 
assume that the kernel $\kk: \mathcal{X} \times \mathcal{X} \to \mathbb{R}$ 
generates $\rho$.  Define  the Gram matrix
\begin{equation}
\label{eq:kernel_matrix}
G \equiv \begin{pmatrix}
\kk(x_1,x_1)     & \dotsm     & \kk(x_1,x_n) \\
\vdots           & \ddots     & \vdots \\
\kk(x_n,x_1)     & \dotsm     & \kk(x_n,x_n) 
\end{pmatrix} .
\end{equation}
Let $Z \in \{ 0,1 \}^{n\times k}$ be the label matrix, 
with only one nonvanishing entry per row, 
indicating to which cluster (column)
each point (row) belongs to. This matrix satisfies
$Z^\top Z = D$, where the diagonal matrix 
$D = \diag( n_1,\dotsc, n_k )$  contains
the number of points in each cluster. We also introduce the rescaled
matrix  $Y$ below. In component form they are 
\begin{equation}
\label{eq:label_matrix}
Z_{ij} \equiv \begin{cases}
1 & \mbox{if $x_i \in \C_j$ } \\
0 & \mbox{otherwise}
\end{cases}, \qquad
\Zt_{ij} \equiv \begin{cases}
\tfrac{1}{\sqrt{s_j}} & \mbox{if $x_i \in \C_j$ } \\
0 & \mbox{otherwise}
\end{cases} .
\end{equation}
Throughout the paper, we use the notation $M_{i\bullet}$ to denote
the $i$th row of a matrix $M$, and $M_{\bullet j}$ denotes its $j$th column.
We also define the following:
\begin{equation}
\label{eq:weighted_matrices}
\mathcal{W} \equiv \diag(w_1,\dotsc,w_n), \quad
H \equiv \mathcal{W}^{1/2} Y, \quad
\om \equiv \mathcal{W} \e,
\end{equation}
where $w_i = w(x_i)$ is the weight associated to point $x_i$, 
and $\e =(1,\dotsc,1)^\top \in \mathbb{R}^n$ 
is the all-ones vector.

Our next result shows that the optimization problem \eqref{eq:minimize}
is NP-hard since it is a quadratically constrained quadratic program 
(QCQP) in the associated RKHS.

\begin{theorem}
\label{th:qcqp3}
The optimization 
problem \eqref{eq:minimize} is equivalent to
\begin{equation}
\label{eq:qcqp3}
\begin{split}
&\max_H  \Tr \left[ H^\top \left(\mathcal{W}^{1/2} \, G \, 
\mathcal{W}^{1/2}\right) H  \right] \\
&\mbox{such that $H \ge 0$, $H^\top H = I$, $H H^\top \om = \om$,}
\end{split}
\end{equation}
where $G$ is the Gram matrix \eqref{eq:kernel_matrix} 
and the other quantities are defined in \eqref{eq:weighted_matrices}.
\end{theorem}
\begin{proof}
From 
\eqref{eq:gen_kernel},
\eqref{eq:g_def2}, and
\eqref{eq:within2}
we have
\begin{equation}
\label{eq:W2}
\begin{split}
\!\! \!\!     W
&= \sum_{j=1}^k \dfrac{1}{2 s_j} 
\!\sum_{x,y \in \C_j} w(x)w(y)\rho(x,y) \\
&= \! \sum_{j=1}^k   \sum_{x \in \C_j} \!\! \bigg[
w(x)\kk(x,x) \!  - \! \dfrac{1}{s_j} \! \sum_{y \in \C_j} \!w(x)w(y) \kk(x,y) \bigg].
\end{split}
\end{equation}
Note that the first term is global so it does not contribute to the 
optimization problem.
Therefore, problem \eqref{eq:minimize} becomes
\begin{equation}
\label{eq:max_prob}
\max_{ \C_1,\dotsc,\C_k } 
\sum_{j=1}^k \dfrac{1}{s_j} \sum_{x,y\in C_j} w(x) w(y) \kk(x,y) .
\end{equation}
Using the definitions 
\eqref{eq:label_matrix} and 
\eqref{eq:weighted_matrices}, the previous 
objective function can be written as
\begin{equation}
\begin{split}
& \hspace{-2em} \sum_{j=1}^k \dfrac{1}{s_j} 
\sum_{p=1}^n \sum_{q=1}^n 
w_p w_q Z_{pj} Z_{qj} G_{pq}  \\
&= 
\sum_{j=1}^k 
\sum_{p=1}^n \sum_{q=1}^n 
\dfrac{Z^\top_{jp}\sqrt{w_p}}{\sqrt{s_j}} w_p^{1/2} G_{pq} w_q^{1/2} 
\dfrac{\sqrt{w_q} Z_{qj}}{\sqrt{s_j}} \\
&= 
\sum_{j=1}^k \left(H^\top \mathcal{W}^{1/2} G \mathcal{W}^{1/2} H\right)_{jj}
\\
&= \Tr\left[ H^\top \mathcal{W}^{1/2} \, G \, \mathcal{W}^{1/2} H  \right].
\end{split}
\end{equation}

Now it remains to obtain the constraints. 
Note that $H_{ij} \ge 0$ by definition, and
\begin{equation}
\begin{split}
(H^\top H)_{ij} &= \sum_{\ell=1}^n 
Y_{\ell i} \mathcal{W}_{\ell \ell} Y_{\ell j } \\
&= 
\dfrac{1}{\sqrt{s_i}\sqrt{s_j}} \sum_{\ell=1}^n w_\ell Z_{\ell i} Z_{\ell j}
\\
&= \dfrac{\delta_{ij}}{s_i} \sum_{\ell=1}^n w_\ell Z_{\ell i} 
\\
&= \delta_{ij}
\end{split}
\end{equation}
where $\delta_{ij}=1$ if $i=j$ and $\delta_{ij}=0$ if $i\ne j$ is the
Kronecker delta. Therefore, $H^\top H = I$. 
This is a constraint on the rows of $H$.
To obtain a constraint on its columns,
observe that
\begin{equation}
\begin{split}
\left(H^\top H\right)_{pq} &= \sqrt{w_p w_q}\sum_{j=1}^k \dfrac{Z_{pj}
Z_{qj}}{s_j} \\
& = \begin{cases}
\dfrac{\sqrt{w_p w_q}}{s_i} & \mbox{if both $x_p,x_q \in \C_i$,} \\
0 & \mbox{otherwise}.
\end{cases}
\end{split}
\end{equation}
Therefore, $(H^\top H \mathcal{W}^{1/2})_{pq} = \sqrt{w_p} \, w_q s_i^{-1}$
if both points $x_p$ and $x_q$ belong to the same cluster, which
we denote by $\C_i$ for some $i\in\{1,\dotsc,k\}$, and 
$(H^\top H \mathcal{W}^{1/2})_{pq} = 0 $ otherwise. Thus, the $p$th
line of this matrix is nonzero only on entries corresponding to points
that are in the same cluster as $x_p$. If we sum over the columns of this
line we obtain $\sqrt{w_p} s_i^{-1} \sum_{q=1}^n w_q Z_{qi} = \sqrt{w_p}$,
or equivalently 
\begin{equation}
H H^\top \mathcal{W}^{1/2} \e = \mathcal{W}^{1/2} \e.
\end{equation}
From \eqref{eq:weighted_matrices} this
gives $H H^\top \om = \om$, finishing the proof.
\end{proof}

The optimization problem \eqref{eq:qcqp3} is nonconvex, besides
being NP-hard, thus a direct approach 
is computational prohibitive even for small datasets.
However, one can find approximate solutions by relaxing some 
of the constraints. For instance, consider the relaxed problem
\begin{equation}
\label{eq:relaxed}
\max_{H} \Tr \left[ H^\top \widetilde{G} \, H \right] 
\qquad \mbox{such that $H^\top H = I$},
\end{equation}
where $\widetilde{G} \equiv \mathcal{W}^{1/2} \, G \, \mathcal{W}^{1/2}$.
This problem has a well-known closed form solution $H^\star = U R$, where the
columns of $U \in \mathbb{R}^{n\times k}$ 
contain the top $k$ eigenvectors of $\widetilde{G}$ corresponding
to the $k$ largest eigenvalues, $\lambda_1\ge \lambda_2\ge\dotsc\ge\lambda_k$, 
and $R \in \mathbb{R}^{k\times k}$ is an arbitrary orthogonal matrix. 
The resulting optimal objective function assumes the value 
$\max \Tr \big[ {H^\star}^\top \widetilde{G} \, H^\star \big]  = 
\sum_{i=1}^k \lambda_i$. 
Spectral clustering is based on this approach, where
one further normalize the rows of $H^\star$, then cluster
the resulting rows as data points using any clustering method such
as k-means.
A procedure on these lines was proposed in the seminal papers
\cite{Malik,NgJordan}.

\subsection{Connection with Graph Partitioning}

We now show how
graph partitioning problems are related to the energy statistics
formulation leading to problem \eqref{eq:qcqp3}.

Consider a graph $\mathcal{G} = (\mathcal{V}, \mathcal{E}, \mathcal{A})$,
where $\mathcal{V}$ is the set of vertices, $\mathcal{E}$ the set of edges,
and $\mathcal{A}$ is an affinity matrix  
which measures the 
similarities between pairs of nodes. Thus, $\mathcal{A}_{ij} \ne 0$
if $(i,j) \in \mathcal{E}$, and $\mathcal{A}_{ij} = 0$ otherwise.
We also associate weights to every vertex, 
$w_i = w(i)$ for $i \in \mathcal{V}$, and let $s_j = \sum_{ i \in \C_j} w_i$,
where $\C_j \subseteq \mathcal{V}$ is one partition of $\mathcal{V}$.
Let
\begin{equation}
\textnormal{links}(\C_\ell, \C_m) \equiv  
\!\sum_{\substack{i\in \C_\ell\\ j\in \C_m} } A_{ij} .
\end{equation}
Our goal is to partition the set of vertices $\mathcal{V}$ into $k$ disjoint
subsets, $\mathcal{V} = \bigcup_{j=1}^k \C_j $. 
The generalized ratio association problem is given by
\begin{equation}
\label{eq:assoc}
\max_{\C_i,\dots,\C_k} \sum_{j=1}^k \dfrac{\textnormal{links}(\C_j,\C_j)}{s_j}
\end{equation}
and maximizes the within cluster association.
The generalized ratio cut problem
\begin{equation}
\label{eq:cut}
\min_{\C_i,\dots,\C_k} \sum_{j=1}^k
\dfrac{\textnormal{links}(\C_j,\mathcal{V} \setminus \C_j)}{s_j}
\end{equation}
minimizes the cut between clusters. Both problems \eqref{eq:assoc}  and
\eqref{eq:cut} are equivalent,
in analogous way as minimizing \eqref{eq:within2} is equivalent to
maximizing \eqref{eq:between2}, as shown in Lemma~\ref{th:minimize}.
Here, this equivalence is  a consequence of the equality
$\textnormal{links}(\C_j,\mathcal{V} \setminus  \C_j)=
\textnormal{links}(\C_j,\mathcal{V}) - \textnormal{links}(\C_j,\C_j)$.
Several graph partitioning methods 
\cite{Kernighan,Malik,Chan,Yu}
can be seen as a particular case of problems
\eqref{eq:assoc} or \eqref{eq:cut}.

Consider the ratio association problem \eqref{eq:assoc}, 
whose objective function can be written as
\begin{equation}
\begin{split}
\sum_{j=1}^k \dfrac{1}{s_j} \sum_{p \in \C_j} \sum_{q \in \C_j}
\mathcal{A}_{pq} &= \sum_{j=1}^k \sum_{p=1}^n \sum_{q=1}^n 
\dfrac{Z^\top_{jp}}{\sqrt{s_j}} \, \mathcal{A}_{pq} \, 
\dfrac{Z_{qj}}{\sqrt{s_j}} \\
&= \Tr\big[ Y^\top \mathcal{A} \, Y \big] ,
\end{split}
\end{equation}
where we recall that $Z$ is defined in \eqref{eq:label_matrix} and $Y$ is
defined in
\eqref{eq:weighted_matrices}. Therefore, the ratio association problem
can be written in the form \eqref{eq:qcqp3}, i.e.,
\begin{equation}
\begin{split}
&\max_H \Tr \left[
H^\top \mathcal{W}^{-1/2} \mathcal{A} \, \mathcal{W}^{-1/2} H 
\right] \\ &\mbox{such that $H\ge 0$, $H^\top H = I$, $H H^\top
\om=\om$}.
\end{split}
\end{equation}
This is exactly the same as \eqref{eq:qcqp3}
with  $G = \mathcal{W}^{-1} \mathcal{A} \mathcal{W}^{-1}$. Assuming that this
matrix is positive semidefinite, this generates a semimetric
\eqref{eq:gen_kernel} for graphs given by
\begin{equation}
\label{eq:metric_graphs}
\rho(i,j) = 
\dfrac{\mathcal{A}_{ii}}{w_i^{2}}
+\dfrac{\mathcal{A}_{jj}}{w_j^{2}}
-\dfrac{2 \mathcal{A}_{ij}}{w_i w_j}
\end{equation}
for vertices $i,j \in \mathcal{V}$. If
we assume the graph has no self-loops we must replace
$\mathcal{A}_{ii} = 0$ above. 
The weight of node $i\in \mathcal{V}$ can be, for instance,
its degree $w_i = w(i) = d(i)$.

\subsection{Connection with Kernel k-Means}
\label{sec:kernel_kmeans}

We now show that kernel k-means optimization problem \cite{Dhillon2,Dhillon}
is also related 
to the previous energy statistics formulation to clustering. To be precise,
we consider a weighted generalization of kernel k-means.

For a positive semidefinite Gram matrix $G$, as defined in
\eqref{eq:kernel_matrix},
there exists a map
$\varphi: \mathcal{X} \to \HH_\kk$ such that
\begin{equation}
\kk(x,y) = \langle \varphi(x), \varphi(y) \rangle. 
\end{equation}
Define the weighted mean of cluster $\C_j$ as
\begin{equation}
\label{eq:muj}
\mu_j = \dfrac{1}{s_j} \sum_{x \in \C_j} w(x) x.
\end{equation}
Disregarding the first global term in \eqref{eq:W2}, note
that the second term,
$-\tfrac{1}{s_j} \sum_{x,y \in \C_j} w(x) w(y) \kk(x,y)$,
is equal to
\begin{multline}
\dfrac{1}{s_j^2}\sum_{x,y,z\in\C_j} 
\langle w(y) \varphi(y),  w(z) \varphi(z) \rangle \\
-\dfrac{2}{s_j}\sum_{x,y\in\C_j} \langle w(x) \varphi(x),  
w(y) \varphi(y) \rangle ,
\end{multline}
which using 
\eqref{eq:muj}
becomes
\begin{multline}
\sum_{x\in\C_j} 
\big\{ \langle \varphi(\mu_j), \varphi(\mu_j) \rangle - 
2 \langle w(x)\varphi(x),
\varphi(\mu_j) \rangle \big\} \\ = 
\sum_{x\in\C_j}\big\{ \| w(x) \varphi(x) - \varphi(\mu_j) \|^2 -
\|w(x)\varphi(x)\|^2\big\}.
\end{multline}
Therefore, minimizing $W$ in \eqref{eq:W2} is equivalent to
\begin{equation}
\label{eq:kernel_kmeans}
\min_{\C_1,\dotsc,\C_k}\bigg\{ 
J(\{\C_j\}) \equiv  \sum_{j=1}^k
\sum_{x \in \C_j} \| w(x) \varphi(x) - \varphi(\mu_j) \|^2
\bigg\} .
\end{equation}
Problem \eqref{eq:kernel_kmeans} is obviously  equivalent to
problem \eqref{eq:qcqp3}. 
When $w(x) = 1$ for all
$x$, problem \eqref{eq:kernel_kmeans} corresponds to 
kernel k-means problem
\cite{Dhillon2,Dhillon}. Thus, the result \eqref{eq:kernel_kmeans} shows
that the previous energy statistics formulation to clustering
is equivalent to a weighted
modification of kernel k-means\footnote{
One should not confuse kernel k-means optimization problem,
given by \eqref{eq:kernel_kmeans}, with kernel k-means \emph{algorithm}.
We will discuss two approaches to solve \eqref{eq:kernel_kmeans},
or equivalently \eqref{eq:qcqp3}. One is 
based on Lloyd's method, which leads to kernel k-means algorithm, and the other is based on Hartigan's method,
which leads to a new algorithm.
}. One must note, however, that energy statistics fixes the kernel
through \eqref{eq:kernel_semimetric}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Iterative Algorithms}
\label{sec:algo}

%In this section we introduce an iterative algorithm to find a local
%maximizer of the optimization problem \eqref{eq:qcqp3}. Due to 
%the discussion of Section~\ref{sec:kernel_kmeans} 
%we can also find an approximate
%solution by kernel k-means algorithm, which is based
%on Lloyd's heuristic \cite{Dhillon2,Dhillon}. For
%convenience, such an algorithm will also be stated in the present context.

In this section, we introduce two iterative algorithms to solve
the optimization problem \eqref{eq:qcqp3}. The first is based on Lloyd's
method, while the second is based on Hartigan's method.

Consider the optimization problem 
\eqref{eq:max_prob} 
written as
\begin{equation}
\label{eq:maxQ}
\! \max_{\{ \C_1,\dotsc,\C_k \}}  \!
\bigg\{ Q =  \sum_{j=1}^k \dfrac{Q_j}{s_j}  \bigg\},
\ Q_j \equiv \!\!\!\sum_{x,y\in\C_j} \!\!\! w(x) w(y)\kk(x,y),
\end{equation}
where $Q_j$ represents an internal cost of cluster $\C_j$, and
$Q$ is the total cost where each $Q_j$ 
is weighted by the inverse
of the sum of weights of the points in $\C_j$. 
For a data point $x_i$, we denote
its cost
with cluster $\C_\ell$ by
\begin{equation}
\label{eq:costxij}
Q_\ell(x_i) \equiv \sum_{y\in\C_\ell} w(x_i) w(y) \kk(x_i, y) = 
(\mathcal{W} G \mathcal{W})_{i \bullet} \cdot Z_{\bullet \ell},
\end{equation}
where we recall that $M_{i\bullet}$ ($M_{\bullet i}$) denotes
the $i$th row (column) of matrix $M$.

\subsection{Weighted Kernel k-Means Algorithm}

Using the definitions \eqref{eq:maxQ} and \eqref{eq:costxij}, the
the optimization problem \eqref{eq:kernel_kmeans} can be written as 
\begin{equation}
\label{eq:min_lloyd}
\min_{Z} \sum_{i = 1}^{n} \sum_{\ell=1}^{k} Z_{i\ell} J^{(\ell)}(x_i)
\end{equation}
where
\begin{equation}
\label{eq:Jell}
J^{(\ell)}(x_i) \equiv 
\dfrac{1}{s_\ell^2} Q_\ell
-\dfrac{2}{s_\ell} Q_\ell(x_i) .
\end{equation}
A possible strategy to solve \eqref{eq:min_lloyd} is to
assign  $x_i$ to cluster $\C_{j^\star}$ according
to 
\begin{equation}
\label{eq:min_lloyd2}
j^\star = \argmin_{\ell=1,\dotsc,k} J^{(\ell)}(x_i) .
\end{equation}
This should be done for every data point $x_i$ and repeated until
convergence, i.e., until no new assignments are made.
The entire procedure is described in Algorithm~\ref{kmeans_algo}.
It can be shown that this algorithm converges when $G$ is positive
semidefinite.

\begin{algorithm}
\begin{algorithmic}[1]
    \INPUT $k$, $G$, $\mathcal{W}$, $Z \leftarrow Z_0$
    \OUTPUT $Z$ 
  \STATE $q \leftarrow (Q_1, \dotsc, Q_k)^\top$ 
            \hfill (see \eqref{eq:maxQ})
  \STATE $s \leftarrow (s_1,\dotsc,s_k)^\top$ 
  \REPEAT
    \FOR{ $i=1,\dotsc,n$}
        \STATE let $j$ be such that $x_i \in \C_j$
        \STATE $j^\star \leftarrow \argmin_{\ell=1,\dotsc,k} J^{(\ell)}(x_i)$
            \hfill (see \eqref{eq:Jell})
        \IF{ $j^\star \ne j$} 
            \STATE $Z_{ij} \leftarrow 0$ 
            \STATE $Z_{ij^\star} \leftarrow 1$
            \STATE $s_j \leftarrow s_j - \mathcal{W}_{ii}$ 
            \STATE $s_{j^\star} \leftarrow s_{j^\star} + \mathcal{W}_{ii}$
            \STATE $q_j \leftarrow q_j - 2Q_j(x_i)$ 
            \hfill (see \eqref{eq:costxij})
            \STATE $q_{j^\star} \leftarrow q_{j^\star} + 2Q_{j^\star}(x_i)$
            \hfill (see \eqref{eq:costxij})
    %    \ELSE
    %        \STATE Do nothing;
        \ENDIF
    \ENDFOR
  \UNTIL{convergence}
\end{algorithmic}
\caption{\label{kmeans_algo}
Weighted version of kernel k-means algorithm
to find local solutions to the optimization problem 
\eqref{eq:qcqp3}.
}
\end{algorithm}

To see that the above procedure is indeed kernel k-means
\cite{Dhillon2,Dhillon}, based
on Lloyd's heuristic \cite{Lloyd}, note that from \eqref{eq:kernel_kmeans} 
and \eqref{eq:Jell} we have
\begin{equation}
\label{eq:closest_mean}
\min_\ell J^{(\ell)}(x_i) = 
\min_\ell \| w(x_i) \varphi(x_i) - \varphi(\mu_\ell)\|^2.
\end{equation}
Therefore, we are 
assigning $x_i$ to the cluster with closest center, in the
feature space.
When $w(x) = 1$ for all $x$, the above method  is exactly 
kernel k-means algorithm.

To check the complexity of Algorithm~\ref{kmeans_algo}, 
note that the second term %of $J^{(\ell)}(x_i)$
in \eqref{eq:Jell} requires
$\OO(n_\ell)$ operations, and although the first term requires
$\OO(n_\ell^2)$ it only needs to be computed once outside loop through
data points (step 1).
Therefore, the time complexity of Algorithm~\ref{kmeans_algo}
is $\OO(n k \max_\ell n_\ell) = \OO(k n^2)$. For a sparse
Gram matrix $G$, having
$\tilde{n}$ nonzero elements, this can be further reduced
to $\OO(k \tilde{n})$.

\subsection{Kernel k-Groups Algorithm}

We now consider Hartigan's method \cite{Hartigan1975,Hartigan1979} 
applied to the optimization problem in the form \eqref{eq:maxQ}, which gives
a local solution to \eqref{eq:qcqp3}. 
The method is based in computing the maximum change
in the total cost function $Q$ when moving each data point to
another cluster. More specifically, 
suppose that point $x_i$
is currently assigned to  cluster $\C_j$ yielding
a total cost function denoted by $Q^{(j)}$.
Moving $x_i$ to cluster $\C_\ell$ yields another total cost function
denoted by $Q^{(\ell)}$. We are interested in computing the maximum 
change
$\Delta Q^{(j\to \ell)} (x_i) \equiv Q^{(\ell)} - Q^{(j)}$, for $\ell\ne j$. 
From \eqref{eq:maxQ}, by explicitly writing the costs related to these 
two cluster we obtain
\begin{equation}
\label{eq:deltaQ}
\Delta Q^{(j\to \ell)} (x_i) = \dfrac{Q_\ell^{+}}{s_\ell+w_i} + 
\dfrac{Q_j^-}{s_j-w_i} - \dfrac{Q_\ell}{s_\ell} - \dfrac{Q_j}{s_j}  ,
\end{equation}
where $Q^{+}_\ell$ denote the cost of the new $\ell$th cluster
with the point $x_i$ added to it, and $Q^-_j$ is the cost of new 
$j$th cluster with $x_i$ removed from it. Recall also that $w_i=w(x_i)$
is the weight associated to point $x_i$. Noting that 
\begin{align}
Q_\ell^{+} &= Q_\ell + 2 Q_\ell(x_i) + (\W G \W)_{ii}, 
\label{eq:Qplus} \\
Q_j^{-} &= Q_j - 2 Q_j(x_i) + (\W G \W)_{ii}, 
\label{eq:Qminus}
\end{align}
we obtain
\begin{multline}
\label{eq:changeQ}
\Delta Q^{(j \to \ell)}(x_i)  = 
\dfrac{1}{s_j - w_i}  \left[ \dfrac{w_i}{s_j}Q_j 
- 2 Q_j(x_i) + (\W G \W)_{ii} \right] \\
- \dfrac{1}{s_\ell + w_i}\left[ \dfrac{w_i}{s_\ell}Q_\ell - 2 Q_\ell(x_i) 
- (\W G \W )_{ii} \right].
\end{multline}
Therefore, we compute
\begin{equation}
j^\star = \argmax_{\ell=1,\dotsc,k \, | \, \ell\ne j} 
\Delta Q^{(j \to \ell)}(x_i)
\end{equation}
and if $\Delta Q^{j \to j^\star}(x_i) > 0$ 
we move $x_i$ to cluster $\C_{j^\star}$, otherwise 
we keep $x_i$ in its original cluster $\C_j$. 
This process is repeated
until no points are assigned to new clusters. 
The entire procedure is described in Algorithm~\ref{algo}, which we
call kernel k-groups. This method is a generalization of the k-groups
with first variations proposed in \cite{Kgroups}, which only considers
the Euclidean case, and also an extension
of Hartigan's method to kernel spaces.

\begin{algorithm}
\begin{algorithmic}[1]
    \INPUT $k$, $G$, $\W$, $Z \leftarrow Z_0$
    \OUTPUT $Z$
  \STATE $q \leftarrow (Q_1, \dotsc, Q_k)^\top$ 
            \hfill (see \eqref{eq:maxQ})
  \STATE $s \leftarrow (s_1,\dotsc,s_k)^\top$ 
  \REPEAT
    \FOR{ $i=1,\dotsc,n$}
        \STATE let $j$ be such that $x_i \in \C_j$
        \STATE $j^\star \leftarrow \argmax_{\ell=1,\dotsc,k \, | \, \ell\ne j} 
                \Delta Q^{j\to \ell}(x_i)$
            \hfill (see \eqref{eq:changeQ}) \label{stepmove}
        \IF{ $\Delta Q^{j \to j^\star}(x_i) > 0$ }
            \STATE $Z_{ij} \leftarrow 0$
            \STATE $Z_{ij^\star} \leftarrow 1$
            \STATE $s_j \leftarrow s_j - \mathcal{W}_{ii}$ 
            \STATE $s_{j^\star} \leftarrow s_{j^\star} + \mathcal{W}_{ii}$
            \STATE $q_j \leftarrow q_j - 2Q_j(x_i) + (\W G \W)_{ii}$
                \hfill (see \eqref{eq:costxij})
            \STATE 
            $q_{j^\star} \leftarrow q_{j^\star} + 2Q_{j^\star}(x_i)+ 
                (\W G\W)_{ii}$
                \hfill (see \eqref{eq:costxij})
    %    \ELSE
    %        \STATE Do nothing;
        \ENDIF
    \ENDFOR
  \UNTIL{convergence}
\end{algorithmic}
\caption{\label{algo}
Kernel k-groups algorithm, based on Hartigan's method, to find
local solutions to problem \eqref{eq:qcqp3}.
%Steps $6$, $12$ and $13$ are different
%compared to Algorithm~\ref{kmeans_algo}. 
}
\end{algorithm}

Note that Algorithm~\ref{algo} automatically ensures that the 
objective function is
monotonically increasing at each iteration, and consequently the algorithm
converges in a finite number of steps.

The complexity analysis of Algorithm~\ref{algo} is the following.
%Computing the Gram matrix $G$ requires $\OO( D n^2)$ operations, where 
%$D$ is the dimension of each data point and $n$ is the data size. However,
%both algorithms $\mathcal{E}^L$- and $\mathcal{E}^H$-clustering 
%assume that $G$ is given. There are more efficient
%methods to compute $G$, specially if it is sparse, but we will not consider
%this further and just assume that $G$ is given.
The computation of each cluster cost
$Q_j$ has complexity $\OO(n_j^2)$, and overall to compute $q$
we have $\OO(n_1^2+\dots + n_k^2) = \OO(k \max_j n_j^2)$. 
These operations only need to be performed a single time. For
each point $x_i$ we need to compute $Q_j(x_i)$ once, which is
$\OO(n_j)$, and we need to compute $Q_\ell(x_i)$ for each $\ell\ne j$. 
The cost of computing 
$Q_\ell(x_i)$
is $\OO(n_\ell)$, thus the cost of step~$6$ in
Algorithm~\ref{algo} is $\OO(k \max_\ell n_\ell)$ for $\ell=1,\dotsc,k$.
For the 
entire dataset this gives a complexity
of $\OO(n k  \max_\ell n_\ell) =\OO(k n^2)$. Note that this is the same cost as
in kernel k-means algorithm. 
Again, if $G$ is sparse
this can be reduced to $\OO(k \tilde{n})$ where $\tilde{n}$ 
is the number of nonzero
entries of $G$.

%In the following we mention some important known 
%results about Hartigan's method.
%
%\begin{theorem}[Telgarsky-Vattani \cite{Telgarsky}]
%Hartigan's method has the cost function strictly decreasing in each
%iteration. Moreover, if $n > k$ then 
%\begin{enumerate}
%\item \label{noempty} the resulting partition has no empty clusters, and
%\item \label{diffmean} the resulting partition has distinct means.
%\end{enumerate}
%\end{theorem}
%
%Neither of these two conditions are guaranteed to be 
%satisfied  by Lloyd's method,
%and consequently by $\mathcal{E}^L$-clustering 
%algorithm.
%The next result indicates that Hartigan's method can potentially 
%escape local optima of Lloyd's method.
%
%\begin{theorem}[Telgarsky-Vattani \cite{Telgarsky}]
%The set of local optima of Hartigan's method is a (possibly strict) subset
%of local optima of Lloyd's method.
%\end{theorem}
%
%The above theorem implies that $\mathcal{E}^L$-clustering  cannot
%improve on a local optima of $\mathcal{E}^H$-clustering. On the other hand,
%$\mathcal{E}^H$ might improve on a local optima of 
%$\mathcal{E}^L$. Lloyd's method forms Voronoi partitions,
%while Hartigan's method groups data
%in regions formed by the intersection of spheres called circlonoi cells.
%It can be shown that the circlonoi cells are contained within
%a smaller volume of a Voronoi cell, and this excess volume grows
%exponentially with the dimension of $\mathcal{X}$ 
%\cite[Theorems 2.4 and 3.1]{Telgarsky}. 
%Points in this excess volume
%force Hartigan's method to iterate, contrary
%to Lloyd's method. Therefore, Hartigan's 
%can escape local
%optima of Lloyd's. 
%Moreover, this improvement should be more prominent as
%dimension increases. Also, the improvement grows as the number of clusters
%$k$ increases.
%The empirical results of \cite{Telgarsky} show that 
%an implementation of Hartigan's method has comparable execution time 
%to an implementation of
%Lloyd's method,
%but no explicit complexity was provided. We show that both
%$\mathcal{E}^L$- and $\mathcal{E}^H$-clustering
%have the same time complexity. To the best of our knowledge, Hartigan's
%method was not previously considered together with kernels, as we 
%are proposing
%in $\mathcal{E}^{H}$-clustering algorithm.
%
%In \cite{Slonin}, Hartigan's method was applied to k-means problem
%with any Bregman divergence. It was shown that the number of Hartigan's
%local optima is upper bounded by $\mathcal{O}(1/k)$ 
%\cite[Proposition 5.1]{Slonin}. 
%In addition, it was provided examples where
%\emph{any} initial partition correspond to a local optima of Lloyd's 
%method, while  the number of local optima in Hartigan's method is small and 
%correspond to true partitions of the data. Empirically, the number of
%Hartigan's local optima was considerably smaller than the number of Lloyd's
%local optima.
%
%The above results indicate that Hartigan's method
%provides several advantages over Lloyd's method, a fact that will also
%be supported by our numerical experiments in the next section where 
%$\mathcal{E}^H$  outperforms
%of $\mathcal{E}^L$ (kernel k-means) in several settings,
%specially in high dimensions.

\section{Numerical Experiments}
\label{sec:numerics}

The main goal of this section is twofold. 
First, to illustrate that in Euclidean spaces with the standard
metric of energy statistics, as defined by the energy 
distance \eqref{eq:energy}, the
clustering method implemented by kernel k-groups
is more flexible and in general more accurate than
k-means and GMM.
Second, we want to compare kernel k-groups with kernel k-means
and spectral clustering when these methods
operate on the same kernel.

The following experimental setup holds unless specified
otherwise. 
We consider
the metrics
\begin{align}
\rho_{\alpha}(x,y) &= \| x-y \|^{\alpha}, 
%\kk_{\alpha}(x,y) = \tfrac{1}{2} \left( \| x \|^{\alpha} + \| y \|^{\alpha} 
%- \| x-y \|^{\alpha} \right), 
\label{eq:rho_alpha} \\
%
\widetilde{\rho}_{\sigma}(x,y) &= 2 - 2 e^{-\tfrac{\|x-y\|}{2 \sigma}},  
%\widetilde{\kk}_{\sigma}(x,y) = e^{-\tfrac{\|x-y\|}{2\sigma}}, 
\label{eq:rho_tilde}\\
%
\widehat{\rho}_{\sigma}(x,y) &= 2 - 2 e^{-\tfrac{\|x-y\|^2}{2 \sigma^2}},
%\widehat{\kk}_{\sigma}(x,y) = e^{-\tfrac{\|x-y\|^2}{2\sigma^2}}  
\label{eq:rho_hat}
\end{align}
which define the 
corresponding kernels through 
\eqref{eq:kernel_semimetric}, where we always fix $x_0=0$.
%The standard metric from energy statistics is $\rho_1$ and 
%will always be present in the experiments as a reference.
We consider the weights associated to data points to be $w(x) = 1$ for
all $x$, so that $\W = I$ in 
Algorithms~\ref{kmeans_algo}~and~\ref{algo}.
For k-means, GMM and spectral clustering we use the
implementations of \emph{scikit-learn} library \cite{scikit-learn}, where  
k-means is initialized with 
k-means++ \cite{Vassilvitskii}, 
and GMM is initialized with the output of k-means, making 
it more robust and preventing
it from breaking in high dimensions. 
The spectral clustering implementation of scikit-learn 
is based on \cite{Malik}. 
Kernel k-means is implemented according to Algorithm~\ref{kmeans_algo} while
kernel k-groups follows Algorithm~\ref{algo}.
Both will also be initialized with k-means++, unless specified otherwise.
We run every algorithms $5$ times with different initializations and then
we choose the result with the best objective function value. 
We evaluate the clustering quality by
the accuracy defined as
\begin{equation}
\label{eq:accuracy}
\textnormal{accuracy}(\hat{Z}) \equiv \max_\pi
\dfrac{1}{n}\sum_{i=1}^n\sum_{j=1}^k \hat{Z}_{i \pi(j)} Z_{ij} ,
\end{equation}
where $\hat{Z}$ is the predicted label matrix, $Z$
is the ground truth label matrix, 
and $\pi$ is a permutation of $\{1,2,\dotsc,k\}$. 
Thus, the accuracy corresponds to the fraction
of correctly classified data points, and it is always between $[0,1]$.
For each setting, we show the average accuracy over $100$ Monte
Carlo trials, with error bars indicating the standard error.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Synthetic Experiments}

%In the first two experiments 
%we use $\mathcal{E}^H$-clustering with the standard metric from energy
%statistics, i.e. \eqref{eq:rho_alpha} with \mbox{$\alpha=1$}. 
%We consider univariate
%normal and lognormal mixtures according to
%\begin{align}
%x &\stackrel{iid}{\sim} \tfrac{1}{2}
%\mathcal{N}(\mu_1,\sigma_1^2) + 
%\tfrac{1}{2} \mathcal{N}(\mu_2,\sigma_2^2) ,\label{eq:uni_normal} \\
%x & \stackrel{iid}{\sim} \tfrac{1}{2}
%\exp\left\{ \mathcal{N}(\mu_1,\sigma_1^2) \right\} + 
%\tfrac{1}{2} \exp \left\{ \mathcal{N}(\mu_2,\sigma_2^2)\right\} ,
%\label{eq:uni_lognormal} \\
%\mu_1 &= 1.5, \ \sigma_1=0.3, \ \mu_2=0, \ \sigma_2=1.5.
%\label{eq:uni_params}
%\end{align}
%We increase the sample size $n$ and show the clustering accuracy 
%\eqref{eq:accuracy}
%versus $n$ in Fig.~\ref{fig:1D}. 
%In both cases the optimal accuracy 
%obtained from Bayes classification error
%is $ \approx 0.88$, indicated by the dashed line.
%Note that $\mathcal{E}^H$-clustering 
%performs better than k-means for normally distributed data, 
%and worse than GMM,
%as expected. 
%However, for lognormal distributions k-means and GMM are only slightly
%better than chance while $\mathcal{E}^H$-clustering is still 
%accurate. The reason is that k-means and GMM models are inconsistent
%with this lognormal data. Energy statistics on the other hand
%is model-free. We also show 
%a density estimation after clustering $1000$ points from these
%mixtures.
%Note that only $\mathcal{E}^{H}$-clustering
%was able to distinguish between the two classes, while 
%k-means and GMM put most of the
%points in a single cluster and points on the tail of the second component
%of \eqref{eq:uni_lognormal} in the other cluster.
%
%
%\begin{figure}
%\begin{minipage}{0.49\linewidth}
%\begin{center}
%\includegraphics[width=\textwidth]{1D_normal.pdf}\\
%~~~~~\includegraphics[width=.9\textwidth]{normal_density.pdf}
%\end{center}
%\end{minipage}
%\begin{minipage}{0.49\linewidth}
%\begin{center}
%\includegraphics[width=\textwidth]{1D_lognormal.pdf}\\
%~~~~~\includegraphics[width=.9\textwidth]{lognormal_density.pdf}
%\end{center}
%\end{minipage}
%%\put(-140,5){\scriptsize{(a)}}
%%\put(-140,-78){\scriptsize{(c)}}
%%\put(-10,5){\scriptsize{(b)}}
%%\put(-10,-78){\scriptsize{(d)}}
%\caption{
%\label{fig:1D}
%\emph{Top Left:} Clustering data from \eqref{eq:uni_normal}.
%\emph{Top Right:} Clustering data from \eqref{eq:uni_lognormal}.
%The dashed lines show the optimal Bayes accuracy, which in both cases 
%are $\approx 0.88$.
%\emph{Bottom Left:} 
%Density estimation of each component in the mixture \eqref{eq:uni_normal} 
%after clustering $1000$ sampled points.
%\emph{Bottom Right:} 
% The same but for lognormal data \eqref{eq:uni_lognormal}.
%}
%\end{figure}

First, we analyze
how the algorithms degrade as the number of dimensions increase.
Consider data from the Gaussian mixture
\begin{equation}
\label{eq:gauss1}
\begin{split}
x  &\stackrel{iid}{\sim} 
\tfrac{1}{2} \mathcal{N}(\mu_1,\Sigma_1) +
\tfrac{1}{2} \mathcal{N}(\mu_2,\Sigma_2), \quad
\Sigma_1=\Sigma_2 = I_D, \\
\mu_1 &= (\underbrace{0,\dotsc,0}_{\times D})^\top, \qquad
\mu_2 = 0.7 (\underbrace{1,\dots,1}_{\times 10},
\underbrace{0,\dots,0}_{\times (D-10)})^\top.
\end{split}
\end{equation}
The Bayes error is fixed as $D$ increases giving an optimal accuracy
of $\approx 0.86$.
We sample $200$ points on each trial.
The results are shown in Fig.~\ref{fig:gauss1}.
We see that 
kernel k-groups and spectral clustering have close
performance, being superior to kernel k-means, k-means, and GMM.
The improvement is noticeable in 
higher dimensions.

\begin{figure}
\centering
\includegraphics[scale=0.5]{gauss1.pdf}
\caption{
\label{fig:gauss1}
High dimensional Gaussian mixture according to \eqref{eq:gauss1}.
We sample $200$ points total on each trial. The dashed line is
Bayes accuracy $\approx 0.86$. We use the metric $\rho_1$ in
\eqref{eq:rho_alpha}, which is standard in energy statistics.
}
\end{figure}

Still for a two-class Gaussian mixture as in \eqref{eq:gauss1}, we now choose
different numbers for the  diagonal covariance $\Sigma_2$.
We have $\Sigma_1=I_D$, $\mu_1=(0,\dotsc,0)^\top \in \mathbb{R}^D$,
$\mu_2=(1,\dotsc,1,0,\dotsc,0)^T \in \mathbb{R}^D$, 
with signal in the first $10$ dimensions, and
\begin{equation}
\label{eq:gauss2}
\begin{split}
\Sigma_2 &= \left( \begin{array}{c|c}
\widetilde{\Sigma}_{10} & 0 \\ \hline 
0 & I_{D-10} \end{array}\right), \\
\widetilde{\Sigma}_{10} &= \diag(1.367,  3.175,  3.247,  4.403,  1.249,\\
&\hspace{3.6em}1.969, 4.035,   4.237,  2.813,  3.637).
\end{split}
\end{equation}
We simply chose $10$ numbers uniformly at random on the interval
$[1,5]$ and other choice would give analogous results.
Bayes  accuracy is fixed at $\approx 0.95$.
From Fig.~\ref{fig:gauss2} we see that 
all the methods are similarly accurate in low dimensions, but they quickly
degenerate as the number of dimensions increase, except 
kernel k-groups which is much more stable.

\begin{figure}
\centering
\includegraphics[scale=0.5]{gauss2.pdf}
\caption{
\label{fig:gauss2}
High dimensional Gaussian mixture according to \eqref{eq:gauss2}.
We sample $200$ points total on each trial. The dashed line is
Bayes accuracy $\approx 0.95$. We use $\rho_1$ in \eqref{eq:rho_alpha}
which standard metric in energy statistics.
}
\end{figure}

Now, consider
$x \stackrel{iid}{\sim} \tfrac{1}{2} \mathcal{N}(\mu_1,\Sigma_1)+
\tfrac{1}{2} \mathcal{N}(\mu_2,\Sigma_2)$
with 
\begin{equation}
\label{eq:20gauss}
\begin{split}
2\Sigma_1 &= \Sigma_2 = I_{20} \\
\mu_1 &= (\underbrace{0,\dotsc,0}_{\times 20})^\top , 
\quad \mu_2 = \tfrac{1}{2} 
(\underbrace{1,\dotsc,1}_{5},\underbrace{0,\dotsc,0}_{15})^\top.
\end{split}
\end{equation}
Bayes accuracy is $\approx 0.90$. 
We increase the sample size $n \in [10, 400]$ and show
the accuracy versus $n$ in Fig.~\ref{fig:gauss_n}. We compare
kernel k-groups, with different metrics,
to k-means and GMM. We also use
the best metric in this example for spectral clustering.
We notice a superior performance of kernel k-groups 
compared to the other methods.

\begin{figure}
\centering
\includegraphics[scale=0.5]{gauss_n.pdf}
\caption{
\label{fig:gauss_n}
Gaussian mixture with parameters \eqref{eq:20gauss}. We increase
the number of sampled points in each trial. We use different
metrics; see \eqref{eq:rho_alpha}--\eqref{eq:rho_hat}. 
Here, kernel k-groups is more accurate than spectral clustering.
}
\end{figure}

To consider non-normal data, we sample from the lognormal mixture 
$x \stackrel{iid}{\sim} \tfrac{1}{2} 
e^{\mathcal{N}(\mu_1,\Sigma_1)}+
\tfrac{1}{2} e^{\mathcal{N}(\mu_2,\Sigma_2)}$ with the same
parameters as in \eqref{eq:20gauss}. 
The optimal Bayes accuracy is still $\approx 0.9$.
The results are shown in Fig.~\ref{fig:loggauss_n}.
We use exactly the same metrics as in the normal
mixture of Fig.~\ref{fig:gauss_n} to illustrate that the 
proposed method still performs accurately.

\begin{figure}
\centering
\includegraphics[scale=0.5]{loggauss_n.pdf}
\caption{
\label{fig:loggauss_n}
Same experiment as in Fig.~\ref{fig:gauss_n} but with a 
lognormal mixture with parameters \eqref{eq:20gauss}.
Again, kernel k-groups is more accurate than alternatives.
The figure suggests that neither of these methods are
consistent on this example since Bayes accuracy is $\approx 0.90$.
}
\end{figure}

Finally, we show a limitation of kernel k-groups, which is shared
between 
all the other methods except for GMM.
For highly unbalanced clusters, k-means, spectral
clustering, kernel k-means and kernel k-groups all degenerate
more quickly than GMM. This is illustrated in Figure~\ref{fig:gauss_unbal}
where we generate data according to
\begin{equation}
\label{eq:gauss3}
\begin{split}
x &\stackrel{iid} \sim  
\dfrac{n_1}{2N} \mathcal{N}(\mu_1,\Sigma_1)+
\dfrac{n_1}{2N} \mathcal{N}(\mu_1,\Sigma_1), \\
\mu_1 &= (0,0,0,0)^\top , \quad
\mu_2 = 1.5\times (1,1,0,0)^\top, \\
\Sigma_1 &= I_4, \quad
\Sigma_2 = \left( 
\begin{array}{c|c} 
\tfrac{1}{2} I_2 & 0  \\ \hline
0 & I_2 
\end{array}\right), \\
n_1 &= N - m, \quad  n_2 = N + m, \quad N=300.
\end{split}
\end{equation}
We then increase $m \in [0,240]$ making
the clusters progressively more unbalanced.
For highly unbalanced clusters, we see that GMM performs better than
the other methods, which have basically similar performance.
Based on this experiment, an interesting problem would be to
extend kernel k-groups to account for unbalanced clusters.

\begin{figure}
\centering
\includegraphics[scale=0.5]{gauss_unbal.pdf}
\caption{
\label{fig:gauss_unbal}
Comparison between clustering methods
on unbalanced clusters. The data is normally 
distributed as \eqref{eq:gauss3} where
we vary $m \in [0, 240]$. We use the standard metric $\rho_1$ (see
\eqref{eq:rho_alpha}) from energy statistics.
}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Real Data Experiment}

We consider the dermatology dataset \cite{Dua2017,Guvenir1998}
which has 366 data points, each with 34 attributes where 33 are
linear valued and one is categorical. There are 8 data points with
missing entries in the ``age'' column. We complete the missing entries
with the mean of the entire column, and then we normalize the entire
dataset to zero mean and unit variance. There are  a total of $6$ classes,
and this is a challenging clustering problem. We refer the reader
to \cite{Dua2017,Guvenir1998} for a complete description of the dataset, 
and also to \cite{RizzoClustering} where this dataset was previously
analyzed. In Table~\ref{tb:dermatology} we show the results of kernel
k-groups using the metric \eqref{eq:rho_alpha} with $\alpha=1/2$.
We show the number of points assigned to each class while indicating
the actual class that points belong to. In Table~\ref{tb:dermatology_accuracy}
we show this experiment using several clustering methods, and we also
compare with the results from \cite{RizzoClustering} and \cite{Kgroups}
on this same data. Our results provide an improvement in comparison
to all the other methods and also the analysis
of \cite{RizzoClustering}.

\begin{table}
\caption{
\label{tb:dermatology}
Clustering the dermatology 
dataset of \cite{Dua2017,Guvenir1998} with kernel k-groups
using the metric $\rho_{1/2}$ (see \eqref{eq:rho_alpha}) from energy
statistics. The table below should be compared with Table~2 of
\cite{RizzoClustering}, for which our results are slightly more accurate.
See also Table~\ref{tb:dermatology_accuracy} below for clustering metrics.
The classes in the vertical indicates the ground truth and the classes
in the horizontal correspond to the classification
obtained by kernel k-groups. We show the estimated number of points
for each class.
}
\centering
\begin{tabular}{@{}l|llllll|l@{}}
%\toprule[1pt]
class & 1 & 2 & 3 & 4 & 5 & 6 & $\#$ cases \\ \midrule[.5pt]
   1   & 112 & 0  & 0  & 0  & 0  & 0  &  112  \\
   2   &  0  & 50 & 0  & 11 & 0  & 0  &   61  \\
   3   &  0  & 0  & 72 & 0  & 0  & 0  &   72  \\
   4   &  0  & 2  & 0  & 47 & 0  & 0  &   49  \\
   5   &  0  & 0  & 0  & 1  & 51 & 0  &   52  \\
   6   &  0  & 0  & 0  & 0  & 0  & 20 &   20  \\ \midrule[.5pt]
total & 112 & 52 & 72 & 59 & 51 & 20 &  366 %\\ \bottomrule[1pt] 
\end{tabular}
\end{table}

\begin{table}
\caption{
\label{tb:dermatology_accuracy}
For the dataset \cite{Dua2017,Guvenir1998} (see also
Table~\ref{tb:dermatology}) we show the accuracy \eqref{eq:accuracy} and
the adjusted Rand index (aRand) of several methods.
In \cite{RizzoClustering} the authors 
obtained $\textnormal{aRand}=0.9195$ using an energy statistics
based method, while \cite{Kgroups} obtains $\textnormal{aRand}=0.9188$ 
where points with missing
entries are removed. Below we complete the missing entries with the mean.
If we remove the points with missing entries, kernel k-groups provides an
improvement of $\textnormal{accuracy}=0.9637$ and $\textnormal{aRand}=0.9396$.
}
\centering
\begin{tabular}{@{}l|ll@{}}
       Method      & Accuracy & aRand \\ \midrule[.5pt]
        kmeans       & 0.7131 & 0.6897 \\
         GMM         & 0.8770 & 0.8403 \\
 spectral clustering & 0.9536 & 0.9125 \\
    kernel k-means   & 0.7514 & 0.8510 \\
   kernel k-groups   & \textbf{0.9617} & \textbf{0.9356}
\end{tabular}
\end{table}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}
\label{sec:conclusion}

We proposed a formulation to clustering based on a weighted
version of energy statistics, 
valid for arbitrary spaces of negative type.
Our mathematical formulation of energy clustering 
reduces to a QCQP in the associated RKHS, as demonstrated in 
Proposition~\ref{th:qcqp3}.
We showed that the optimization problem
is equivalent
to kernel k-means, once the kernel is fixed, and also
to several graph partitioning problems.
Energy statistics, however, fixes
a family of standard kernels in Euclidean space, and
more general kernels 
on spaces of negative type can also be obtained.

We extended Hartigan's method to kernel spaces and proposed
Algorithm~\ref{algo}, which we call kernel k-groups.
This method was compared to kernel k-means and spectral clustering, besides
k-means and GMM. Our numerical results show a superior performance of the
proposed method, specially in high dimensions.
We stress that kernel k-groups has the same complexity of kernel k-means.

Kernel k-groups suffers a limitation shared by kernel k-means and spectral
clustering which involves highly unbalanced
clusters. An interesting problem  that we leave open 
would is to extend the method to such 
situations, as well as a soft version.
Finally, kernel methods can benefit from sparsity and
fixed-rank approximations of the Gram matrix, and there is plenty
of room to make kernel k-groups more scalable.

% @gui: better, but still important to have another paragraph between those connecting it to the existing literature.  compare this work with 3-4 other works explicitly stating the difference.

% if have a single appendix:
%\appendix[Proof of the Zonklar Equations]
% or
%\appendix  % for no appendix heading
% do not use \section anymore after \appendix, only \section*
% is possibly needed

% use appendices with more than one appendix
% then use \section to start each appendix
% you must declare a \section before using any
% \subsection or using \label (\appendices by itself
% starts a section numbered zero.)
%

%\appendices
%\section{Proof of the First Zonklar Equation}
%Appendix one text goes here.

% you can choose not to have a title for an appendix
% if you want by leaving the argument blank
%\section{}
%Appendix two text goes here.

% use section* for acknowledgment
\ifCLASSOPTIONcompsoc
  % The Computer Society usually uses the plural form
  \section*{Acknowledgments}
\else
  % regular IEEE prefers the singular form
  \section*{Acknowledgment}
\fi

We would like to thank Carey Priebe for discussions.
We would like to acknowledge the support of the Transformative
Research Award (NIH \#R01NS092474) and  the Defense Advanced Research Projects
Agency’s (DARPA) SIMPLEX program through SPAWAR contract N66001-15-C-4041.

% Can use something like this to put references on a page
% by themselves when using endfloat and the captionsoff option.
\ifCLASSOPTIONcaptionsoff
  \newpage
\fi

% trigger a \newpage just before the given reference
% number - used to balance the columns on the last page
% adjust value as needed - may need to be readjusted if
% the document is modified later
%\IEEEtriggeratref{8}
% The "triggered" command can be changed if desired:
%\IEEEtriggercmd{\enlargethispage{-5in}}

% references section

% can use a bibliography generated by BibTeX as a .bbl file
% BibTeX documentation can be easily obtained at:
% http://mirror.ctan.org/biblio/bibtex/contrib/doc/
% The IEEEtran BibTeX style support page is at:
% http://www.michaelshell.org/tex/ieeetran/bibtex/
%\bibliographystyle{IEEEtran}
% argument is your BibTeX string definitions and bibliography database(s)
%\bibliography{IEEEabrv,../bib/paper}
%
% <OR> manually copy in the resultant .bbl file
% set second argument of \begin to the number of references
% (used to reserve space for the reference number labels box)

\bibliographystyle{unsrt}
%\bibliographystyle{abbrvnat}
\bibliography{biblio.bib}

% biography section
% 
% If you have an EPS/PDF photo (graphicx package needed) extra braces are
% needed around the contents of the optional argument to biography to prevent
% the LaTeX parser from getting confused when it sees the complicated
% \includegraphics command within an optional argument. (You could create
% your own custom macro containing the \includegraphics command to make things
% simpler here.)
%\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{mshell}}]{Michael Shell}
% or if you just want to reserve a space for a photo:

\begin{IEEEbiography}{Guilherme Fran\c ca}
Biography text here.
\end{IEEEbiography}

% if you will not have a photo at all:
\begin{IEEEbiographynophoto}{Maria Rizzo}
Biography text here.
\end{IEEEbiographynophoto}

% insert where needed to balance the two columns on the last page with
% biographies
%\newpage

\begin{IEEEbiographynophoto}{Joshua Vogelstein}
Biography text here.
\end{IEEEbiographynophoto}

% You can push biographies down or up by placing
% a \vfill before or after them. The appropriate
% use of \vfill depends on what kind of text is
% on the last page and whether or not the columns
% are being equalized.

%\vfill

% Can be used to pull up biographies so that the bottom of the last one
% is flush with the other column.
%\enlargethispage{-5in}

% that's all folks
\end{document}

### 2018-06-04 energy_clustering/pami18/old_version/energy_clustering.tex
energy pami

--- 2017-10-26 energy_clustering/draft_v2/energy_clustering.tex
@@
-\documentclass[aps,preprint,nofootinbib,floatfix]{revtex4-1}
+%% bare_jrnl_compsoc.tex
+%% V1.4b
+%% 2015/08/26
+%% by Michael Shell
+%% See:
+%% http://www.michaelshell.org/
+%% for current contact information.
+%%
+%% This is a skeleton file demonstrating the use of IEEEtran.cls
+%% (requires IEEEtran.cls version 1.8b or later) with an IEEE
+%% Computer Society journal paper.
+%%
+%% Support sites:
+%% http://www.michaelshell.org/tex/ieeetran/
+%% http://www.ctan.org/pkg/ieeetran
+%% and
+%% http://www.ieee.org/
+
+%%*************************************************************************
+%% Legal Notice:
+%% This code is offered as-is without any warranty either expressed or
+%% implied; without even the implied warranty of MERCHANTABILITY or
+%% FITNESS FOR A PARTICULAR PURPOSE! 
+%% User assumes all risk.
+%% In no event shall the IEEE or any contributor to this code be liable for
+%% any damages or losses, including, but not limited to, incidental,
+%% consequential, or any other damages, resulting from the use or misuse
+%% of any information contained here.
+%%
+%% All comments are the opinions of their respective authors and are not
+%% necessarily endorsed by the IEEE.
+%%
+%% This work is distributed under the LaTeX Project Public License (LPPL)
+%% ( http://www.latex-project.org/ ) version 1.3, and may be freely used,
+%% distributed and modified. A copy of the LPPL, version 1.3, is included
+%% in the base LaTeX documentation of all distributions of LaTeX released
+%% 2003/12/01 or later.
+%% Retain all contribution notices and credits.
+%% ** Modified files should be clearly indicated as such, including  **
+%% ** renaming them and changing author support contact information. **
+%%*************************************************************************
+
+% *** Authors should verify (and, if needed, correct) their LaTeX system  ***
+% *** with the testflow diagnostic prior to trusting their LaTeX platform ***
+% *** with production work. The IEEE's font choices and paper sizes can   ***
+% *** trigger bugs that do not appear when using other class files.       ***                          ***
+% The testflow support page is at:
+% http://www.michaelshell.org/tex/testflow/
+
+\documentclass[10pt,journal,compsoc]{IEEEtran}
+%
+% If IEEEtran.cls has not been installed into the LaTeX system files,
+% manually specify the path to it like:
+% \documentclass[10pt,journal,compsoc]{../sty/IEEEtran}
+
+% Some very useful LaTeX packages include:
+% (uncomment the ones you want to load)
+
+% *** MISC UTILITY PACKAGES ***
+%
+%\usepackage{ifpdf}
+% Heiko Oberdiek's ifpdf.sty is very useful if you need conditional
+% compilation based on whether the output is pdf or dvi.
+% usage:
+% \ifpdf
+%   % pdf code
+% \else
+%   % dvi code
+% \fi
+% The latest version of ifpdf.sty can be obtained from:
+% http://www.ctan.org/pkg/ifpdf
+% Also, note that IEEEtran.cls V1.7 and later provides a builtin
+% \ifCLASSINFOpdf conditional that works the same way.
+% When switching from latex to pdflatex and vice-versa, the compiler may
+% have to be run twice to clear warning/error messages.
 
@@
 %\emergencystretch=10pt
-
 \newtheorem{theorem}{Theorem}
@@
 
+% *** CITATION PACKAGES ***
+%
+\ifCLASSOPTIONcompsoc
+  % IEEE Computer Society needs nocompress option
+  % requires cite.sty v4.0 or later (November 2003)
+  \usepackage[nocompress]{cite}
+\else
+  % normal IEEE
+  \usepackage{cite}
+\fi
+% cite.sty was written by Donald Arseneau
+% V1.6 and later of IEEEtran pre-defines the format of the cite.sty package
+% \cite{} output to follow that of the IEEE. Loading the cite package will
+% result in citation numbers being automatically sorted and properly
+% "compressed/ranged". e.g., [1], [9], [2], [7], [5], [6] without using
+% cite.sty will become [1], [2], [5]--[7], [9] using cite.sty. cite.sty's
+% \cite will automatically add leading space, if needed. Use cite.sty's
+% noadjust option (cite.sty V3.8 and later) if you want to turn this off
+% such as if a citation ever needs to be enclosed in parenthesis.
+% cite.sty is already installed on most LaTeX systems. Be sure and use
+% version 5.0 (2009-03-20) and later if using hyperref.sty.
+% The latest version can be obtained at:
+% http://www.ctan.org/pkg/cite
+% The documentation is contained in the cite.sty file itself.
+%
+% Note that some packages require special options to format as the Computer
+% Society requires. In particular, Computer Society  papers do not use
+% compressed citation ranges as is done in typical IEEE papers
+% (e.g., [1]-[4]). Instead, they list every citation separately in order
+% (e.g., [1], [2], [3], [4]). To get the latter we need to load the cite
+% package with the nocompress option which is supported by cite.sty v4.0
+% and later. Note also the use of a CLASSOPTION conditional provided by
+% IEEEtran.cls V1.7 and later.
+
+% *** GRAPHICS RELATED PACKAGES ***
+%
+\ifCLASSINFOpdf
+  % \usepackage[pdftex]{graphicx}
+  % declare the path(s) where your graphic files are
+  % \graphicspath{{../pdf/}{../jpeg/}}
+  % and their extensions so you won't have to specify these with
+  % every instance of \includegraphics
+  % \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
+\else
+  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
+  % will default to the driver specified in the system graphics.cfg if no
+  % driver is specified.
+  % \usepackage[dvips]{graphicx}
+  % declare the path(s) where your graphic files are
+  % \graphicspath{{../eps/}}
+  % and their extensions so you won't have to specify these with
+  % every instance of \includegraphics
+  % \DeclareGraphicsExtensions{.eps}
+\fi
+% graphicx was written by David Carlisle and Sebastian Rahtz. It is
+% required if you want graphics, photos, etc. graphicx.sty is already
+% installed on most LaTeX systems. The latest version and documentation
+% can be obtained at: 
+% http://www.ctan.org/pkg/graphicx
+% Another good source of documentation is "Using Imported Graphics in
+% LaTeX2e" by Keith Reckdahl which can be found at:
+% http://www.ctan.org/pkg/epslatex
+%
+% latex, and pdflatex in dvi mode, support graphics in encapsulated
+% postscript (.eps) format. pdflatex in pdf mode supports graphics
+% in .pdf, .jpeg, .png and .mps (metapost) formats. Users should ensure
+% that all non-photo figures use a vector format (.eps, .pdf, .mps) and
+% not a bitmapped formats (.jpeg, .png). The IEEE frowns on bitmapped formats
+% which can result in "jaggedy"/blurry rendering of lines and letters as
+% well as large increases in file sizes.
+%
+% You can find documentation about the pdfTeX application at:
+% http://www.tug.org/applications/pdftex
+
+% *** MATH PACKAGES ***
+%
+%\usepackage{amsmath}
+% A popular package from the American Mathematical Society that provides
+% many useful and powerful commands for dealing with mathematics.
+%
+% Note that the amsmath package sets \interdisplaylinepenalty to 10000
+% thus preventing page breaks from occurring within multiline equations. Use:
+%\interdisplaylinepenalty=2500
+% after loading amsmath to restore such page breaks as IEEEtran.cls normally
+% does. amsmath.sty is already installed on most LaTeX systems. The latest
+% version and documentation can be obtained at:
+% http://www.ctan.org/pkg/amsmath
+
+% *** SPECIALIZED LIST PACKAGES ***
+%
+%\usepackage{algorithmic}
+% algorithmic.sty was written by Peter Williams and Rogerio Brito.
+% This package provides an algorithmic environment fo describing algorithms.
+% You can use the algorithmic environment in-text or within a figure
+% environment to provide for a floating algorithm. Do NOT use the algorithm
+% floating environment provided by algorithm.sty (by the same authors) or
+% algorithm2e.sty (by Christophe Fiorio) as the IEEE does not use dedicated
+% algorithm float types and packages that provide these will not provide
+% correct IEEE style captions. The latest version and documentation of
+% algorithmic.sty can be obtained at:
+% http://www.ctan.org/pkg/algorithms
+% Also of interest may be the (relatively newer and more customizable)
+% algorithmicx.sty package by Szasz Janos:
+% http://www.ctan.org/pkg/algorithmicx
+
+% *** ALIGNMENT PACKAGES ***
+%
+%\usepackage{array}
+% Frank Mittelbach's and David Carlisle's array.sty patches and improves
+% the standard LaTeX2e array and tabular environments to provide better
+% appearance and additional user controls. As the default LaTeX2e table
+% generation code is lacking to the point of almost being broken with
+% respect to the quality of the end results, all users are strongly
+% advised to use an enhanced (at the very least that provided by array.sty)
+% set of table tools. array.sty is already installed on most systems. The
+% latest version and documentation can be obtained at:
+% http://www.ctan.org/pkg/array
+
+% IEEEtran contains the IEEEeqnarray family of commands that can be used to
+% generate multiline equations as well as matrices, tables, etc., of high
+% quality.
+
+% *** SUBFIGURE PACKAGES ***
+%\ifCLASSOPTIONcompsoc
+%  \usepackage[caption=false,font=footnotesize,labelfont=sf,textfont=sf]{subfig}
+%\else
+%  \usepackage[caption=false,font=footnotesize]{subfig}
+%\fi
+% subfig.sty, written by Steven Douglas Cochran, is the modern replacement
+% for subfigure.sty, the latter of which is no longer maintained and is
+% incompatible with some LaTeX packages including fixltx2e. However,
+% subfig.sty requires and automatically loads Axel Sommerfeldt's caption.sty
+% which will override IEEEtran.cls' handling of captions and this will result
+% in non-IEEE style figure/table captions. To prevent this problem, be sure
+% and invoke subfig.sty's "caption=false" package option (available since
+% subfig.sty version 1.3, 2005/06/28) as this is will preserve IEEEtran.cls
+% handling of captions.
+% Note that the Computer Society format requires a sans serif font rather
+% than the serif font used in traditional IEEE formatting and thus the need
+% to invoke different subfig.sty package options depending on whether
+% compsoc mode has been enabled.
+%
+% The latest version and documentation of subfig.sty can be obtained at:
+% http://www.ctan.org/pkg/subfig
+
+% *** FLOAT PACKAGES ***
+%
+%\usepackage{fixltx2e}
+% fixltx2e, the successor to the earlier fix2col.sty, was written by
+% Frank Mittelbach and David Carlisle. This package corrects a few problems
+% in the LaTeX2e kernel, the most notable of which is that in current
+% LaTeX2e releases, the ordering of single and double column floats is not
+% guaranteed to be preserved. Thus, an unpatched LaTeX2e can allow a
+% single column figure to be placed prior to an earlier double column
+% figure.
+% Be aware that LaTeX2e kernels dated 2015 and later have fixltx2e.sty's
+% corrections already built into the system in which case a warning will
+% be issued if an attempt is made to load fixltx2e.sty as it is no longer
+% needed.
+% The latest version and documentation can be found at:
+% http://www.ctan.org/pkg/fixltx2e
+
+%\usepackage{stfloats}
+% stfloats.sty was written by Sigitas Tolusis. This package gives LaTeX2e
+% the ability to do double column floats at the bottom of the page as well
+% as the top. (e.g., "\begin{figure*}[!b]" is not normally possible in
+% LaTeX2e). It also provides a command:
+%\fnbelowfloat
+% to enable the placement of footnotes below bottom floats (the standard
+% LaTeX2e kernel puts them above bottom floats). This is an invasive package
+% which rewrites many portions of the LaTeX2e float routines. It may not work
+% with other packages that modify the LaTeX2e float routines. The latest
+% version and documentation can be obtained at:
+% http://www.ctan.org/pkg/stfloats
+% Do not use the stfloats baselinefloat ability as the IEEE does not allow
+% \baselineskip to stretch. Authors submitting work to the IEEE should note
+% that the IEEE rarely uses double column equations and that authors should try
+% to avoid such use. Do not be tempted to use the cuted.sty or midfloat.sty
+% packages (also by Sigitas Tolusis) as the IEEE does not format its papers in
+% such ways.
+% Do not attempt to use stfloats with fixltx2e as they are incompatible.
+% Instead, use Morten Hogholm'a dblfloatfix which combines the features
+% of both fixltx2e and stfloats:
+%
+% \usepackage{dblfloatfix}
+% The latest version can be found at:
+% http://www.ctan.org/pkg/dblfloatfix
+
+%\ifCLASSOPTIONcaptionsoff
+%  \usepackage[nomarkers]{endfloat}
+% \let\MYoriglatexcaption\caption
+% \renewcommand{\caption}[2][\relax]{\MYoriglatexcaption[#2]{#2}}
+%\fi
+% endfloat.sty was written by James Darrell McCauley, Jeff Goldberg and 
+% Axel Sommerfeldt. This package may be useful when used in conjunction with 
+% IEEEtran.cls'  captionsoff option. Some IEEE journals/societies require that
+% submissions have lists of figures/tables at the end of the paper and that
+% figures/tables without any captions are placed on a page by themselves at
+% the end of the document. If needed, the draftcls IEEEtran class option or
+% \CLASSINPUTbaselinestretch interface can be used to increase the line
+% spacing as well. Be sure and use the nomarkers option of endfloat to
+% prevent endfloat from "marking" where the figures would have been placed
+% in the text. The two hack lines of code above are a slight modification of
+% that suggested by in the endfloat docs (section 8.4.1) to ensure that
+% the full captions always appear in the list of figures/tables - even if
+% the user used the short optional argument of \caption[]{}.
+% IEEE papers do not typically make use of \caption[]'s optional argument,
+% so this should not be an issue. A similar trick can be used to disable
+% captions of packages such as subfig.sty that lack options to turn off
+% the subcaptions:
+% For subfig.sty:
+% \let\MYorigsubfloat\subfloat
+% \renewcommand{\subfloat}[2][\relax]{\MYorigsubfloat[]{#2}}
+% However, the above trick will not work if both optional arguments of
+% the \subfloat command are used. Furthermore, there needs to be a
+% description of each subfigure *somewhere* and endfloat does not add
+% subfigure captions to its list of figures. Thus, the best approach is to
+% avoid the use of subfigure captions (many IEEE journals avoid them anyway)
+% and instead reference/explain all the subfigures within the main caption.
+% The latest version of endfloat.sty and its documentation can obtained at:
+% http://www.ctan.org/pkg/endfloat
+%
+% The IEEEtran \ifCLASSOPTIONcaptionsoff conditional can also be used
+% later in the document, say, to conditionally put the References on a 
+% page by themselves.
+
+% *** PDF, URL AND HYPERLINK PACKAGES ***
+%
+%\usepackage{url}
+% url.sty was written by Donald Arseneau. It provides better support for
+% handling and breaking URLs. url.sty is already installed on most LaTeX
+% systems. The latest version and documentation can be obtained at:
+% http://www.ctan.org/pkg/url
+% Basically, \url{my_url_here}.
+
+% *** Do not adjust lengths that control margins, column widths, etc. ***
+% *** Do not use packages that alter fonts (such as pslatex).         ***
+% There should be no need to do such things with IEEEtran.cls V1.6 and later.
+% (Unless specifically asked to do so by the journal or conference you plan
+% to submit to, of course. )
+
+% correct bad hyphenation here
+\hyphenation{op-tical net-works semi-conduc-tor}
+
 \begin{document}
-
+%
+% paper title
+% Titles are generally capitalized except for words such as a, an, and, as,
+% at, but, by, for, in, nor, of, on, or, the, to and up, which are usually
+% not capitalized unless they are the first or last word of the title.
+% Linebreaks \\ can be used within to get better formatting as desired.
+% Do not put math or special symbols in the title.
 \title{Energy Clustering}
-
-\author{Guilherme Fran\c ca}
-\email{guifranca@gmail.com} 
-\author{Joshua T. Vogelstein}
-\email{jovo@jhu.edu}
-\affiliation{Johns Hopkins University}
-
+%
+%
+% author names and IEEE memberships
+% note positions of commas and nonbreaking spaces ( ~ ) LaTeX will not break
+% a structure at a ~ so this keeps an author's name from being broken across
+% two lines.
+% use \thanks{} to gain access to the first footnote area
+% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
+% was not built to handle multiple paragraphs
+%
+%
+%\IEEEcompsocitemizethanks is a special \thanks that produces the bulleted
+% lists the Computer Society journals use for "first footnote" author
+% affiliations. Use \IEEEcompsocthanksitem which works much like \item
+% for each affiliation group. When not in compsoc mode,
+% \IEEEcompsocitemizethanks becomes like \thanks and
+% \IEEEcompsocthanksitem becomes a line break with idention. This
+% facilitates dual compilation, although admittedly the differences in the
+% desired content of \author between the different types of papers makes a
+% one-size-fits-all approach a daunting prospect. For instance, compsoc 
+% journal papers have the author affiliations above the "Manuscript
+% received ..."  text while in non-compsoc journals this is reversed. Sigh.
+
+\author{Guilherme~Fran\c ca,~Maria Rizzo~and~Joshua~Vogelstein
+\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem G. Fran\c ca is
+with the Center for Imaging Science, Johns Hopkins
+University; J. Vogelstein is with the Center for Imaging Science,
+Department of Biomedical Engineering and Institute for Computational
+Medicine, Johns Hopkins University. \protect\\
+% note need leading \protect in front of \\ to get a newline within \thanks as
+% \\ is fragile and will error, could use \hfil\break instead.
+E-mail: guifranca@jhu.edu; jovo@jhu.edu
+\IEEEcompsocthanksitem M. Rizzo is with the Department of Mathematics
+and Statistics, Bowling Green State University. \protect\\
+E-mail: mrizzo@bgsu.edu}% <-this % stops an unwanted space
+\thanks{Manuscript received March 15, 2018; revised March 15, 2018.}}
+
+% note the % following the last \IEEEmembership and also \thanks - 
+% these prevent an unwanted space from occurring between the last author name
+% and the end of the author line. i.e., if you had this:
+% 
+% \author{....lastname \thanks{...} \thanks{...} }
+%                     ^------------^------------^----Do not want these spaces!
+%
+% a space would be appended to the last name and could cause every name on that
+% line to be shifted left slightly. This is one of those "LaTeX things". For
+% instance, "\textbf{A} \textbf{B}" will typeset as "A B" not "AB". To get
+% "AB" then you have to do: "\textbf{A}\textbf{B}"
+% \thanks is no different in this regard, so shield the last } of each \thanks
+% that ends a line with a % and do not let a space in before the next \thanks.
+% Spaces after \IEEEmembership other than the last one are OK (and needed) as
+% you are supposed to have spaces between the names. For what it is worth,
+% this is a minor point as most people would not even notice if the said evil
+% space somehow managed to creep in.
+
+% The paper headers
+\markboth{Journal of \LaTeX\ Class Files,~Vol.~14, No.~8, August~2015}%
+{Energy Clustering}
+% The only time the second header will appear is for the odd numbered pages
+% after the title page when using the twoside option.
+% 
+% *** Note that you probably will NOT want to include the author's ***
+% *** name in the headers of peer review papers.                   ***
+% You can use \ifCLASSOPTIONpeerreview for conditional compilation here if
+% you desire.
+
+% The publisher's ID mark at the bottom of the page is less important with
+% Computer Society journal papers as those publications place the marks
+% outside of the main text columns and, therefore, unlike regular IEEE
+% journals, the available text space is not reduced by their presence.
+% If you want to put a publisher's ID mark on the page you can do it like
+% this:
+%\IEEEpubid{0000--0000/00\$00.00~\copyright~2015 IEEE}
+% or like this to get the Computer Society new two part style.
+%\IEEEpubid{\makebox[\columnwidth]{\hfill 0000--0000/00/\$00.00~\copyright~2015 IEEE}%
+%\hspace{\columnsep}\makebox[\columnwidth]{Published by the IEEE Computer Society\hfill}}
+% Remember, if you use this you must call \IEEEpubidadjcol in the second
+% column for its text to clear the IEEEpubid mark (Computer Society jorunal
+% papers don't need this extra clearance.)
+
+% use for special paper notices
+%\IEEEspecialpapernotice{(Invited Paper)}
+
+% for Computer Society papers, we must declare the abstract and index terms
+% PRIOR to the title within the \IEEEtitleabstractindextext IEEEtran
+% command as these need to go into the title area created by \maketitle.
+% As a general rule, do not put math, special symbols or citations
+% in the abstract or keywords.
+\IEEEtitleabstractindextext{%
 \begin{abstract}
 Energy statistics was proposed by Sz\' ekely in the 80's inspired by the
-Newtonian gravitational potential from classical mechanics, and it provides 
-a 
-hypothesis test for equality of distributions. 
+Newtonian gravitational potential from classical mechanics and it provides
+a hypothesis test for equality of distributions.
 It was further generalized from Euclidean spaces to metric spaces of
@@
 kernel Hilbert spaces (RKHS) was established.
-Here we consider the clustering problem from an 
-energy statistics theory perspective, 
-providing a precise mathematical formulation 
+In this paper, we consider the problem of clustering data 
+from the perspective of energy statistics theory. 
+We provide a precise mathematical formulation 
 yielding a quadratically constrained
 quadratic program (QCQP) in the associated RKHS, thus establishing the
-connection with kernel methods. We show that this QCQP
-is equivalent to kernel $k$-means optimization problem once the kernel
-is fixed.
-These results imply a first principles derivation of kernel $k$-means 
-from energy statistics.
-However, energy statistics fixes a family of standard kernels.
-Furthermore, we also consider a weighted version of energy statistics, 
-making connection to graph partitioning problems.
-To find local optimizers of such QCQP we propose an iterative algorithm based 
+connection with kernel methods. 
+%We show that this QCQP
+%is equivalent to kernel $k$-means optimization problem once the kernel
+%is fixed.
+%These results imply a first principles derivation of kernel $k$-means
+%from energy statistics.
+%However, energy statistics fixes a family of standard kernels.
+Moreover, we also propose a weighted version of energy statistics applied
+to clustering, making connection to graph partitioning problems.
+To find local solutions of such QCQP, we propose an iterative algorithm based
 on Hartigan's method, which in this case has the same computational cost
-as kernel $k$-means algorithm, based on Lloyd's heuristic, but usually 
+as kernel $k$-means algorithm, which is on Lloyd's heuristic, but usually
 with better clustering quality.
@@
 of the proposed method compared to kernel $k$-means, spectral clustering,
-standard $k$-means, and Gaussian mixture models in a variety of settings.
+standard $k$-means, and Gaussian mixture models in a variety of settings,
+specially in high dimensions. We also 
+employ the proposed method to an important
+real dataset describing protein expressions of neural synapses.
 \end{abstract}
 
+% Note that keywords are not normally used for peerreview papers.
+\begin{IEEEkeywords}
+Clustering, Energy Statistics, Kernel Methods.
+\end{IEEEkeywords}}
+
+% make the title area
 \maketitle
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Introduction}
-
-Energy statistics \cite{Szkely2013}
+% To allow for easy dual compilation without having to reenter the
+% abstract/keywords data, the \IEEEtitleabstractindextext text will
+% not be used in maketitle, but will appear (i.e., to be "transported")
+% here as \IEEEdisplaynontitleabstractindextext when the compsoc 
+% or transmag modes are not selected <OR> if conference mode is selected 
+% - because all conference papers position the abstract like regular
+% papers do.
+\IEEEdisplaynontitleabstractindextext
+% \IEEEdisplaynontitleabstractindextext has no effect when using
+% compsoc or transmag under a non-conference mode.
+
+% For peer review papers, you can put extra information on the cover
+% page as needed:
+% \ifCLASSOPTIONpeerreview
+% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
+% \fi
+%
+% For peerreview papers, this IEEEtran command inserts a page break and
+% creates the second title. It will be ignored for other modes.
+\IEEEpeerreviewmaketitle
+
+\IEEEraisesectionheading{\section{Introduction}\label{sec:introduction}}
+% Computer Society journal (but not conference!) papers do something unusual
+% with the very first section heading (almost always called "Introduction").
+% They place it ABOVE the main text! IEEEtran.cls does not automatically do
+% this for you, but you can achieve this effect with the provided
+% \IEEEraisesectionheading{} command. Note the need to keep any \label that
+% is to refer to the section immediately after \section in the above as
+% \IEEEraisesectionheading puts \section within a raised box.
+
+% The very first letter is a 2 line initial drop letter followed
+% by the rest of the first word in caps (small caps for compsoc).
+% 
+% form to use if the first word consists of a single letter:
+% \IEEEPARstart{A}{demo} file is ....
+% 
+% form to use if you need the single drop letter followed by
+% normal text (unknown if ever used by the IEEE):
+% \IEEEPARstart{A}{}demo file is ....
+% 
+% Some journals put the first two words in caps:
+% \IEEEPARstart{T}{his demo} file is ....
+% 
+% Here we have the typical use of a "T" for an initial drop letter
+% and "HIS" in caps to complete the first word.
+
+\IEEEPARstart{E}{nergy Statistics} \cite{Szkely2013}
 is based on a 
@@
 under minimum energy. When probability distributions are different the 
-statistical potential energy diverges as sample size increases, while tends 
+``statistical potential energy'' diverges as sample size increases, 
+while tends 
 to a nondegenerate limit distribution when probability
@@
 spaces to metric spaces of strong negative type \cite{Lyons}. Furthermore, 
-the missing link between energy distance based tests and kernel 
+the link between energy distance based tests and kernel 
 based tests has 
-been recently resolved \cite{Sejdinovic2013}, 
-establishing an equivalence between generalized energy distances to maximum
+been recently established \cite{Sejdinovic2013} 
+through an equivalence between generalized energy distances to maximum
 mean discrepancies (MMD), which are distances between embeddings of 
@@
 the precise optimization problem behind this approach
-remains elusive, as well as the connection with kernel methods.
+remains obscure, as well as the connection with kernel methods.
 The main theoretical contribution of this paper is to fill this gap.
@@
 or \emph{energy clustering} for short, with kernel methods. Moreover,
-our formulation holds for general semimetric spaces of negative type.
+our formulation holds beyond the Euclidean case, more precisely, it holds
+for general semimetric spaces of negative type.
 We also show that such QCQP is equivalent to kernel $k$-means 
-optimization problem, however, the kernel is fixed by 
-energy statistics.
+optimization problem when the kernel is fixed. However, energy statistics
+is able to fix a family of kernels. Therefore, one can see 
+kernel $k$-means as a consequence of energy statistics theory.
 The equivalence between kernel $k$-means, spectral clustering, and graph 
@@
 We make clear the advantages of this proposal versus 
-Lloyd's method, which kernel $k$-means is based upon and will also be used 
+Lloyd's method, which kernel $k$-means algorithm 
+is based upon and will also be used 
 to solve our QCQP. We show that both algorithms  have the same
 time complexity, but Hartigan's method in kernel spaces offer several
-advantages. Furthermore, in the examples considered in this paper, it 
+advantages since it is able to escape local minima of Lloyd's method. 
+Furthermore, in the examples considered in this paper, the proposed method
 also provides superior performance compared spectral clustering, which
@@
 
+%
+%\begin{figure}[!t]
+%\centering
+%\includegraphics[width=2.5in]{myfigure}
+% where an .eps filename suffix will be assumed under latex, 
+% and a .pdf suffix will be assumed for pdflatex; or what has been declared
+% via \DeclareGraphicsExtensions.
+%\caption{Simulation results for the network.}
+%\label{fig_sim}
+%\end{figure}
+
+% Note that the IEEE typically puts floats only at the top, even when this
+% results in a large percentage of a column being occupied by floats.
+% However, the Computer Society has been known to put floats at the bottom.
+
+% An example of a double column floating figure using two subfigures.
+% (The subfig.sty package must be loaded for this to work.)
+% The subfigure \label commands are set within each subfloat command,
+% and the \label for the overall figure must come after \caption.
+% \hfil is used as a separator to get equal spacing.
+% Watch out that the combined width of all the subfigures on a 
+% line do not exceed the text width or a line break will occur.
+%
+%\begin{figure*}[!t]
+%\centering
+%\subfloat[Case I]{\includegraphics[width=2.5in]{box}%
+%\label{fig_first_case}}
+%\hfil
+%\subfloat[Case II]{\includegraphics[width=2.5in]{box}%
+%\label{fig_second_case}}
+%\caption{Simulation results for the network.}
+%\label{fig_sim}
+%\end{figure*}
+%
+% Note that often IEEE papers with subfigures do not employ subfigure
+% captions (using the optional argument to \subfloat[]), but instead will
+% reference/describe all of them (a), (b), etc., within the main caption.
+% Be aware that for subfig.sty to generate the (a), (b), etc., subfigure
+% labels, the optional argument to \subfloat must be present. If a
+% subcaption is not desired, just leave its contents blank,
+% e.g., \subfloat[].
+
+% An example of a floating table. Note that, for IEEE style tables, the
+% \caption command should come BEFORE the table and, given that table
+% captions serve much like titles, are usually capitalized except for words
+% such as a, an, and, as, at, but, by, for, in, nor, of, on, or, the, to
+% and up, which are usually not capitalized unless they are the first or
+% last word of the caption. Table text will default to \footnotesize as
+% the IEEE normally uses this smaller font for tables.
+% The \label must come after \caption as always.
+%
+%\begin{table}[!t]
+%% increase table row spacing, adjust to taste
+%\renewcommand{\arraystretch}{1.3}
+% if using array.sty, it might be a good idea to tweak the value of
+% \extrarowheight as needed to properly center the text within the cells
+%\caption{An Example of a Table}
+%\label{table_example}
+%\centering
+%% Some packages, such as MDW tools, offer better commands for making tables
+%% than the plain LaTeX2e tabular which is used here.
+%\begin{tabular}{|c||c|}
+%\hline
+%One & Two\\
+%\hline
+%Three & Four\\
+%\hline
+%\end{tabular}
+%\end{table}
+
+% Note that the IEEE does not put floats in the very first column
+% - or typically anywhere on the first page for that matter. Also,
+% in-text middle ("here") positioning is typically not used, but it
+% is allowed and encouraged for Computer Society conferences (but
+% not Computer Society journals). Most IEEE journals/conferences use
+% top floats exclusively. 
+% Note that, LaTeX2e, unlike IEEE journals/conferences, places
+% footnotes above bottom floats. This can be corrected via the
+% \fnbelowfloat command of the stfloats package.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Background on Energy Statistics and RKHS}
+\section{Background}
 \label{sec:background}
@@
 $\| X - Y \| \to \rho(X , Y)$ in the definition \eqref{eq:energy}, obtaining
-the generalized energy distance
+the \emph{generalized energy distance}
 \begin{equation}
@@
 properties immediately imply that $\kk$ is symmetric and positive
-definite. Indeed, notice that
-$\langle \kkk_x, \kkk_y \rangle = \kkk_y(x) = \kk(x,y)$, and by 
-definition 
-$\langle \kkk_x, \kkk_y \rangle^* = \langle \kkk_y, \kkk_x \rangle$, but
-since the inner product is real we have 
-$\langle \kkk_y, \kkk_x \rangle = \langle \kkk_x, \kkk_y \rangle$, or
-equivalently 
-$\kk(y,x) = \kk(x,y)$. Moreover, for any $w \in
-\HH$ we can write $w = \sum_{i=1}^n c_i \kkk_{x_i}$ where
-$\{ \kkk_{x_i} \}_{i=1}^n$ is a basis of $\HH$. It follows that
-$\langle w, w \rangle_{\HH}  = \sum_{i,j=1}^n c_i c_j \kk(x_i,x_j) \ge 0$,
-showing that the kernel is positive definite. If $G$ is a matrix with
-elements $G_{ij} = \kk(x_i,x_j)$ this is equivalent to $G$ being
+definite. 
+%Indeed, notice that
+%$\langle \kkk_x, \kkk_y \rangle = \kkk_y(x) = \kk(x,y)$, and by 
+%definition 
+%$\langle \kkk_x, \kkk_y \rangle^* = \langle \kkk_y, \kkk_x \rangle$, but
+%since the inner product is real we have 
+%$\langle \kkk_y, \kkk_x \rangle = \langle \kkk_x, \kkk_y \rangle$, or
+%equivalently 
+%$\kk(y,x) = \kk(x,y)$. Moreover, for any $w \in
+%\HH$ we can write $w = \sum_{i=1}^n c_i \kkk_{x_i}$ where
+%$\{ \kkk_{x_i} \}_{i=1}^n$ is a basis of $\HH$. It follows that
+%$\langle w, w \rangle_{\HH}  = \sum_{i,j=1}^n c_i c_j \kk(x_i,x_j) \ge 0$,
+%showing that the kernel is positive definite. 
+Defining the Gram matrix $G$ with
+elements $G_{ij} = \kk(x_i,x_j)$, this is equivalent to $G=G^\top$ being
 positive semidefinite, i.e. $v^\top G \, v \ge 0$ for any vector
@@
 have 
-\begin{equation}\label{eq:inner_data}
+%\begin{equation}\label{eq:inner_data}
+$
 \langle \kkk_P, \kkk_Q \rangle_{\Hk} = \E \, \kk(X, Y).
-\end{equation}
-Thus, in practice, we can estimate the inner product between  
-embedded distributions 
-by averaging the kernel function over sampled data.
+$
+%\end{equation}
+%Thus, in practice, we can estimate the inner product between  
+%embedded distributions 
+%by averaging the kernel function over sampled data.
 
@@
 \begin{equation} \label{eq:Erho}
-\Energy(P, Q) = 
-2 \left[ \E \, \kk(X, X') + \E \, \kk(Y, Y') - 2\E \, \kk(X, Y)\right] 
-= 2 \gamma_\kk^2(P,Q) .
+\begin{split}
+\Energy(P, Q) &= 
+2 \left[ \E \, \kk(X, X') + \E \, \kk(Y, Y') - 2\E \, \kk(X, Y)\right]  \\
+&= 2 \gamma_\kk^2(P,Q) .
+\end{split}
 \end{equation}
@@
 When the sample size is large enough, $n\to \infty$,
-under the null hypothesis $H_0: P_1=P_2=\dotsm=P_k$ we have that
+under the null hypothesis $H_0: P_1=P_2=\dotsm=P_k$, we have that
 $S\to 0$, 
@@
 Note that this test does not make any assumptions
-about the distributions $P_j$, thus it is said to be non-parametric or
+about the form of distribution $P_j$, thus it is said to be 
 distribution-free.
@@
 we formulate an optimization problem for clustering 
-based on energy statistics and RKHS introduced in the previous section.
+based on energy statistics in the associated RKHS.
 
@@
 criterion for clustering data is to 
-maximize $S$ which makes 
+maximize $S$, given by \eqref{eq:between},
+which makes 
 each cluster as different
@@
 shows that maximizing $S$ is, however, equivalent to minimizing
-$W$ which has a more convenient form.
+$W$, given by \eqref{eq:within}, which has a more convenient form.
 
@@
 \label{eq:minimize}
-\min_{\C_1,\dotsc,C_k  } W(
-\C_1, \dotsc, \C_k),
+\{ \C_1^\star,\dotsc,\C_k^\star \} = \argmin_{\C_1,\dotsc,C_k  } W(
+\C_1, \dotsc, \C_k)
 \end{equation}
@@
 From \eqref{eq:within} and \eqref{eq:between}
-we have
+we have that
 \begin{equation}
 \begin{split}
-S + W &= 
-\dfrac{1}{2n} \sum_{\substack{i,j=1 \\ i\ne j}}^k n_i n_j g(\C_i, \C_j)
+& S + W \\
+&= \dfrac{1}{2n} \sum_{\substack{i,j=1 \\ i\ne j}}^k n_i n_j g(\C_i, \C_j)
 + \dfrac{1}{2n} \sum_{i=1}^{k} 
@@
 Notice that this is a hard clustering problem as partitions
-are disjoint. The optimization problem \eqref{eq:minimize} based on
-energy statistics was already proposed in \cite{Kgroups}. However, it is
-important to note that this is equivalent to maximizing $S$,
-which is the test statistic for equality of distributions. In this current
-form, the relation with kernels and other clustering methods is obscure.
+are disjoint. In the Euclidean case, the optimization 
+problem \eqref{eq:minimize}
+was already proposed in \cite{Kgroups}.
+Thus, Lemma~\ref{th:minimize} provides a generalization to arbirary spaces
+of negative type. Moreover, it is
+important to note that this problem is equivalent to maximizing $S$
+which is the test statistic for equality of distributions. 
+More importantly, in this current
+form, the relation of problem \eqref{eq:minimize} 
+with kernels and other clustering methods is obscure.
 In the following, we show what is the explicit optimization problem behind 
@@
 \label{eq:qcqp2}
-\max_{\Zt} \Tr \left( \Zt^\top G \, \Zt \right)  \qquad
+\max_{\Zt} \Tr \left( \Zt^\top G \, \Zt \right) \ \
 \mbox{s.t. $\Zt \ge 0$, $\Zt^\top \Zt = I$, 
@@
 \label{eq:W2}
-W(\C_1,\dotsc,\C_k  )
-= \dfrac{1}{2} \sum_{j=1}^k \dfrac{1}{n_j} \sum_{x,y \in \C_j} \rho(x,y)
-= \sum_{j=1}^k \sum_{x \in \C_j}  \bigg(
+\begin{split}
+W
+&= \dfrac{1}{2} \sum_{j=1}^k \dfrac{1}{n_j} \sum_{x,y \in \C_j} \rho(x,y) \\
+&= \sum_{j=1}^k \sum_{x \in \C_j}  \bigg(
 \kk(x,x) - \dfrac{1}{n_j} \sum_{y \in \C_j} \kk(x,y) \bigg).
+\end{split}
 \end{equation}
@@
 \eqref{eq:label_matrix}. 
-Notice that $n_j^{-1} = D^{-1}_{jj}$, where the diagonal matrix $D = 
+Notice that $n_j^{-1} = D^{-1}_{jj}$, where $D = 
 \diag(n_1,\dotsc,n_k)$ contains the number of points in each cluster, 
@@
 \begin{equation}\label{eq:qcqp}
-\max_{Z} \Tr\Big( \big( Z D^{-1/2}\big)^\top G 
+\begin{split}
+&\max_{Z} \Tr\Big( \big( Z D^{-1/2}\big)^\top G 
 \big( ZD^{-1/2} \big) 
-\Big) \quad
-\mbox{s.t. $Z_{ij} \in \{0,1\}$, $\sum_{j=1}^k Z_{ij} = 1$, 
+\Big) \\
+&\mbox{s.t. $Z_{ij} \in \{0,1\}$, $\sum_{j=1}^k Z_{ij} = 1$, 
 $\sum_{i=1}^n Z_{ij} = n_j$.}
+\end{split}
 \end{equation}
@@
 that $Y D^{1/2} \e = Z \e = \e$, we finally obtain
-$Y Y^\top \e = \e$. Therefore, the optimization 
+$Y Y^\top \e = \e$. Therefore,  
 problem \eqref{eq:qcqp} is equivalent
@@
 
-Note that the optimization problem \eqref{eq:qcqp2} based on energy
-statistics
-is valid for data living in an \emph{arbitrary} space of negative type, where
-a semimetric $\rho$, and thus the kernel $\kk$, are
-assumed to be known. The standard energy distance \eqref{eq:energy2}
+We remark that problem \eqref{eq:qcqp2} 
+is valid for data living in an \emph{arbitrary} space of negative type.
+The standard energy distance \eqref{eq:energy2}
 fixes a family of choices in Euclidean spaces given by
 \begin{equation}
-\rho_\alpha(x,y) = \| x - y\|^\alpha, \qquad 
-K_\alpha(x,y) = \tfrac{1}{2} \left( \| x \|^\alpha + \| y \|^\alpha - 
+\begin{split}
+\rho_\alpha(x,y) &= \| x - y\|^\alpha,   \\
+K_\alpha(x,y) &= \tfrac{1}{2} \left( \| x \|^\alpha + \| y \|^\alpha - 
 \| x-y\|^\alpha \right) ,
+\end{split}
 \end{equation}
@@
 fixes the kernel.
-In practice, the clustering quality strongly depend on the choice 
+In practice, the clustering quality strongly depends on the choice 
 of a suitable $\rho$. 
 Nevertheless, if prior information is available to make this choice, it
-can be immediately incorporated in the 
+can be immediately incorporated into the 
 optimization problem \eqref{eq:qcqp2}.
-
-\subsection*{Relation to Kernel $\bm{k}$-Means}
 
@@
 relates to the well-known kernel $k$-means problem%
-\footnote{When we refer to kernel $k$-means problem we mean specifically 
+%%%
+\footnote{When we refer to kernel $k$-means \emph{problem} we mean specifically 
 the optimization problem \eqref{eq:kernel_kmeans}, which should not be 
-confused with kernel $k$-means algorithm that is just one possible recipe 
+confused with kernel $k$-means \emph{algorithm} which 
+is just one possible recipe 
 to solve \eqref{eq:kernel_kmeans}. The distinction should also be clear
 from the context.}
+%%%%
 which is extensively used in machine learning.
@@
 $\kk(x,y) = \langle \varphi(x), \varphi(y) \rangle$. 
-Kernel $k$-means optimization
-problem,
-in feature space,
+The kernel $k$-means optimization
+problem
 is defined by
@@
 is strongly tied to the idea of minimizing distances between points
-and cluster centers, which arises from $k$-means objective function based
-on Lloyd's method \cite{Lloyd}.
+and cluster centers, which is
+Lloyd's method \cite{Lloyd}.
 It is known \cite{Dhillon2,Dhillon}
@@
 For a fixed kernel,
-the clustering optimization problem
+the optimization problem
 \eqref{eq:minimize} based on energy statistics 
@@
 \end{equation}
-The first term is global so it does not contribute to the optimization
-problem. Notice that the third term gives
+The first term is global so it does not contribute.
+The third term gives
 $\sum_{x\in\C_j} \tfrac{1}{n_j^2} \sum_{y,z\in\C_j} \kk(y,z) =
@@
 The remaining of the proof proceeds as 
-already shown in the proof of Proposition~\ref{th:qcqp2}, leading to
-the optimization problem \eqref{eq:qcqp2}.
+already shown in the proof of Proposition~\ref{th:qcqp2}.
 \end{proof}
@@
 formulated in the energy statistics framework, when operating on the same
-kernel. Notice, however, that 
-energy statistics is valid for arbitrary semimetric spaces of
-negative type, fixing the kernel function in the associated RKHS, which
-is guaranteed to be positive definite. On the other hand, 
-kernel $k$-means \eqref{eq:kernel_kmeans} by itself is just an heuristic 
-approach that does not make any explicit mention to the kernel. 
-Based on Proposition~\ref{th:kernel_kmeans}
-one may view kernel $k$-means as being derived from the energy statistics
-framework.
+kernel. This result could not have been anticipated a priori.
+It is important to note, however, that kernel $k$-means formulation
+\eqref{eq:kernel_kmeans} is completelly heuristic, while problem
+\eqref{eq:minimize2} was derived from first-principles using energy statistics
+theory, which includes the kernel construction. Therefore, one may view kernel
+$k$-means optimization problem as a consequence of energy statistics.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Clustering Based on Weighted Energy Statistics}
+\section{A Weighted Generalization}
 \label{sec:weighted}
@@
 Define
-\begin{equation}
+\begin{align}
 \label{eq:g_def2}
-g(\C_i, \C_j) \equiv \dfrac{1}{s_i s_j} \sum_{x\in \C_i}\sum_{y\in\C_j}
-w(x)w(y) \rho(x,y), \qquad s_i \equiv \sum_{x\in\C_i} w(x).
-\end{equation}
-Replace this function in the formulas \eqref{eq:within} and
-\eqref{eq:between}, with 
-$n_i \to s_i$ and $n \to s$, where $s = \sum_{j=1}^k s_j$.
+g(\C_i, \C_j) &\equiv \dfrac{1}{s_i s_j} \sum_{x\in \C_i}\sum_{y\in\C_j}
+w(x)w(y) \rho(x,y), \\
+s_i &\equiv \sum_{x\in\C_i} w(x), \qquad
+s \equiv \sum_{j=1}^k s_j.
+\end{align}
+Replace \eqref{eq:g_def2} in the formulas \eqref{eq:within} and
+\eqref{eq:between} with 
+$n_i \to s_i$ and $n \to s$.
 With these
-changes Proposition~\ref{th:minimize} remains the unaltered, so the
+changes, Proposition~\ref{th:minimize} remains the unaltered, so the
 clustering problem becomes
@@
 where now $g$ is given by \eqref{eq:g_def2}. 
-Define the following matrices and vector:
-\begin{equation}
-\label{eq:weighted_matrices}
-Y_{ij} \equiv \begin{cases}
+Define the following:
+\begin{align}
+%\begin{split}
+Y_{ij} &\equiv \begin{cases}
 \tfrac{1}{\sqrt{s_j}} & \mbox{if $x_i \in \C_j$} \\
 0 & \mbox{otherwise}
-\end{cases}, \qquad
-\mathcal{W} \equiv \diag(w_1,\dotsc,w_n), \qquad
-H \equiv \mathcal{W}^{1/2} Y, \qquad
-\om \equiv \mathcal{W} \e,
-\end{equation}
+\end{cases},  &
+\mathcal{W} &\equiv \diag(w_1,\dotsc,w_n), \nonumber \\
+H &\equiv \mathcal{W}^{1/2} Y, &
+ \om &\equiv \mathcal{W} \e,
+\label{eq:weighted_matrices}
+%\end{split}
+\end{align}
 where $w_i = w(x_i)$ and $\e \in \mathbb{R}^n$ is the all-ones
-vector.The analogous of
+vector. The analogous of
 Proposition~\ref{th:qcqp2} is as follows.
@@
 \label{eq:qcqp3}
-\max_H \Tr \left\{ H^\top (\mathcal{W}^{1/2} G \mathcal{W}^{1/2}) H  \right\}
-\qquad \mbox{s.t. $H \ge 0$, $H^\top H = I$, $H H^\top \om =
-\om$,}
+\begin{split}
+&\max_H \Tr \left\{ H^\top (\mathcal{W}^{1/2} G \mathcal{W}^{1/2}) H  \right\}\\
+&\mbox{s.t. $H \ge 0$, $H^\top H = I$, $H H^\top \om = \om$,}
+\end{split}
 \end{equation}
@@
 \begin{split}
-\sum_{j=1}^k \dfrac{1}{s_j} 
+& \hspace{-2em} \sum_{j=1}^k \dfrac{1}{s_j} 
 \sum_{p=1}^n \sum_{q=1}^n 
-w_p w_q Z_{pj} Z_{qj} G_{pq} &= 
+w_p w_q Z_{pj} Z_{qj} G_{pq}  \\
+&= 
 \sum_{j=1}^k 
@@
 \begin{equation}
-(H^\top H)_{ij} = \sum_{\ell=1}^n 
-Y_{\ell i} \mathcal{W}_{\ell \ell} Y_{\ell j } = 
+\begin{split}
+(H^\top H)_{ij} &= \sum_{\ell=1}^n 
+Y_{\ell i} \mathcal{W}_{\ell \ell} Y_{\ell j } \\
+&= 
 \dfrac{1}{\sqrt{s_i}\sqrt{s_j}} \sum_{\ell=1}^n w_\ell Z_{\ell i} Z_{\ell j}
-= \dfrac{\delta_{ij}}{s_i} \sum_{\ell=1}^n w_\ell Z_{\ell i} = \delta_{ij},
+\\
+&= \dfrac{\delta_{ij}}{s_i} \sum_{\ell=1}^n w_\ell Z_{\ell i} 
+\\
+&= \delta_{ij},
+\end{split}
 \end{equation}
@@
 This is a constraint on the rows of $H$.
-To obtain a condition on its columns
+To obtain a condition on its columns,
 observe that
 \begin{equation}
-\left(H^\top H\right)_{pq} = \sqrt{w_p w_q}\sum_{j=1}^k \dfrac{Z_{pj}
-Z_{qj}}{s_j} = \begin{cases}
+\begin{split}
+\left(H^\top H\right)_{pq} &= \sqrt{w_p w_q}\sum_{j=1}^k \dfrac{Z_{pj}
+Z_{qj}}{s_j} \\
+& = \begin{cases}
 \dfrac{\sqrt{w_p w_q}}{s_i} & \mbox{if both $x_p,x_q \in \C_i$} \\
@@
 \end{cases}
+\end{split}
 \end{equation}
@@
 
-\subsection*{Connection with Graph Partitioning}
+\subsection{Connection with Graph Partitioning}
 
@@
 \begin{equation}
+\begin{split}
 \sum_{j=1}^k \dfrac{1}{s_j} \sum_{p \in \C_j} \sum_{q \in \C_j}
-\mathcal{A}_{pq} = \sum_{j=1}^k \sum_{p=1}^n \sum_{q=1}^n 
+\mathcal{A}_{pq} &= \sum_{j=1}^k \sum_{p=1}^n \sum_{q=1}^n 
 \dfrac{Z^\top_{jp}}{\sqrt{s_j}} \, \mathcal{A}_{pq} \, 
-\dfrac{Z_{qj}}{\sqrt{s_j}}
-= \Tr\left( Y^\top \mathcal{A} Y \right) ,
+\dfrac{Z_{qj}}{\sqrt{s_j}} \\
+&= \Tr\left( Y^\top \mathcal{A} Y \right) ,
+\end{split}
 \end{equation}
@@
 \begin{equation}
-\max_H \Tr\left( H^\top \mathcal{W}^{-1/2} \mathcal{A} \mathcal{W}^{-1/2} H 
-\right) \qquad \mbox{s.t. $H\ge 0$, $H^\top H = I$, $H H^\top
+\begin{split}
+&\max_H \Tr\left( H^\top \mathcal{W}^{-1/2} \mathcal{A} \mathcal{W}^{-1/2} H 
+\right) \\ &\mbox{s.t. $H\ge 0$, $H^\top H = I$, $H H^\top
 \om=\om$}.
+\end{split}
 \end{equation}
@@
 +\dfrac{\mathcal{A}_{jj}}{w_j^{2}}
--\dfrac{2 \mathcal{A}_{ij}}{w_i w_j} \qquad\mbox{or}\qquad
-\rho(i,j) = -\dfrac{2 \mathcal{A}_{ij}}{w_i w_j}
-\end{equation}
-for vertices $i,j \in \mathcal{V}$, and 
-where in the second equation we assume the graph has no self-loops,
-i.e. $\mathcal{A}_{ii} = 0$. 
-Using \eqref{eq:metric_graphs} 
-in the energy statistics formulation 
-allows one to make inference
-on graphs.
-Above, the weight $w_i = w(i)$ of node $i\in \mathcal{V}$ can be, for instance,
-its degree $w_i = d(i)$.
+-\dfrac{2 \mathcal{A}_{ij}}{w_i w_j}
+\end{equation}
+for vertices $i,j \in \mathcal{V}$. If
+we assume the graph has no self-loops we must replace
+$\mathcal{A}_{ii} = 0$ above. 
+Above, the weight of node $i\in \mathcal{V}$ can be, for instance,
+its degree, $w_i = w(i) = d(i)$.
 
@@
 
-\begin{figure}
-\begin{flushleft}
-\begin{algorithm}[H]
-\vspace{.5em}
+\begin{algorithm}
 \begin{algorithmic}[1]
-\INPUT data $\mathbb{X}$
-\OUTPUT label matrix $Z$
-\STATE sort $\mathbb{X}$ obtaining 
+\INPUT $\mathbb{X}$
+\OUTPUT $Z$
+\STATE sort $\mathbb{X}$, obtaining 
 $\widetilde{\mathbb{X}}= [ x_1,\dotsc,x_n ]$
     \FOR{$j\in [ 1,\dotsc,n ]$}
-        \STATE $\tC_{1,j} \leftarrow [x_i: i=1,\dotsc,j]$, and 
-                $\tC_{2,j} \leftarrow [x_i : i=j+1,\dotsc,n]$
+        \STATE $\tC_{1,j} \leftarrow [x_i: i=1,\dotsc,j]$
+        \STATE $\tC_{2,j} \leftarrow [x_i : i=j+1,\dotsc,n]$
         \STATE  
-            $W^{(j)} \leftarrow W \big( \tC_{1,j},\tC_{2,j}\big)$, 
-            from \eqref{eq:w1d}
+            $W^{(j)} \leftarrow W \big( \tC_{1,j},\tC_{2,j}\big)$ 
+            (see \eqref{eq:w1d})
     \ENDFOR
     \STATE $j^\star \leftarrow \argmin_j W^{(j)}$ 
-    \STATE $Z_{j\bullet} \leftarrow (1,0) $ if $j\le j^\star$, and
-           $Z_{j\bullet} \leftarrow (0,1)$ otherwise, for $j=1,\dotsc,n$
+    \STATE for $j=1,\dotsc,n$ 
+        assign $Z_{j\bullet} \leftarrow \begin{cases} (1,0) & \mbox{if $j \le
+    j^\star$} \\ (0,1) & \mbox{otherwise}\end{cases}$ 
 \end{algorithmic}
@@
 problem \eqref{eq:minimize} 
-for a two-class problem in one dimension. \hspace{\fill}
+for a two-class problem in one dimension. 
 }
 \end{algorithm}
-\end{flushleft}
-\end{figure}
 
@@
 \label{eq:two_normal}
-x \stackrel{iid}{\sim} 
+\begin{split}
+x &\stackrel{iid}{\sim} 
 \tfrac{1}{2}\mathcal{N}\big(\mu_1,\sigma_1^2\big) 
-+\tfrac{1}{2}\mathcal{N}\big(\mu_2,\sigma_2^2\big), 
-\qquad 
-\mu_1 = 1.5, 
-\sigma_1=0.3, 
-\mu_2 = 0, 
++\tfrac{1}{2}\mathcal{N}\big(\mu_2,\sigma_2^2\big),  \\
+\mu_1 &= 1.5, \quad
+\sigma_1=0.3, \quad
+\mu_2 = 0,  \quad
 \sigma_2 = 1.5.
+\end{split}
 \end{equation}
@@
 \begin{equation}
+\begin{split}
 \label{eq:two_lognormal}
-x \stackrel{iid}{\sim} 
+x &\stackrel{iid}{\sim} 
 \tfrac{1}{2} \exp\left\{\mathcal{N}\big( \mu_1,\sigma_1^2\big)\right\} 
 +\tfrac{1}{2} \exp\left\{\mathcal{N}\big( \mu_2,\sigma_2^2\big)\right\}, 
-\qquad 
-\mu_1 = 1.5, 
-\sigma_1=0.3, 
-\mu_2 = 0, 
+\\
+\mu_1 &= 1.5,  \quad
+\sigma_1=0.3,  \quad
+\mu_2 = 0, \quad
 \sigma_2 = 1.5.
+\end{split}
 \end{equation}
@@
 \begin{figure}
-\begin{minipage}{0.49\textwidth}
-\includegraphics[width=\textwidth]{1D_normal.pdf}\\[-1em](a)\\
-~~~~~\includegraphics[width=.9\textwidth]{normal_density.pdf}\\[-1.5em](c)
+\begin{minipage}{0.49\linewidth}
+\begin{center}
+\includegraphics[width=\textwidth]{1D_normal.pdf}\\
+~~~~~\includegraphics[width=.9\textwidth]{normal_density.pdf}
+\end{center}
 \end{minipage}
-\begin{minipage}{0.49\textwidth}
-\includegraphics[width=\textwidth]{1D_lognormal.pdf}\\[-1em](b)\\
-~~~~~\includegraphics[width=.9\textwidth]{lognormal_density.pdf}\\[-1.5em](d)
+\begin{minipage}{0.49\linewidth}
+\begin{center}
+\includegraphics[width=\textwidth]{1D_lognormal.pdf}\\
+~~~~~\includegraphics[width=.9\textwidth]{lognormal_density.pdf}
+\end{center}
 \end{minipage}
+\put(-140,5){\scriptsize{(a)}}
+\put(-140,-78){\scriptsize{(c)}}
+\put(-10,5){\scriptsize{(b)}}
+\put(-10,-78){\scriptsize{(d)}}
 \caption{
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Iterative Algorithms for Energy Clustering}
+\section{Iterative Algorithms}
 \label{sec:algo}
@@
 
-\subsection*{Lloyd's Method for Energy Clustering}
+\subsection{Lloyd's Method for Energy Clustering}
 
@@
 \begin{equation}
-%\begin{split}
+\begin{split}
 J^{(\ell)}(x_i) 
-%&= \dfrac{1}{n_\ell^2} \sum_{x, y \in \C_\ell} 
-%\langle \varphi(x), \varphi(y) \rangle - \dfrac{2}{n_\ell}
-%\sum_{y\in \C_\ell} \langle \varphi(x_i), \varphi(y)\rangle \\ 
-= 
+&= 
 \langle \varphi(\mu_\ell), \varphi(\mu_\ell) \rangle
 -2 \langle \varphi(x_i), \varphi(\mu_\ell) \rangle 
-= 
+\\
+&= 
 \| \varphi(x_i) - \varphi(\mu_\ell) \|^2 - \| \varphi(x_i) \|^2,
-%\end{split}
+\end{split}
 \end{equation}
@@
 
-\begin{figure}
-\begin{flushleft}
-\begin{algorithm}[H]
-\vspace{.5em}
+\begin{algorithm}
 \begin{algorithmic}[1]
-    \INPUT number of clusters $k$, Gram matrix $G$, initial label
-    matrix $Z \leftarrow Z_0$
-    \OUTPUT label matrix $Z$ 
+    \INPUT $k$, $G$, $Z \leftarrow Z_0$
+    \OUTPUT $Z$ 
   \STATE $q \leftarrow (Q_1, \dotsc, Q_k)^\top$ 
-            have the costs of each cluster, defined in \eqref{eq:maxQ}
+            (see \eqref{eq:maxQ})
   \STATE $n \leftarrow (n_1,\dotsc,n_k)^\top$ 
-        have the number of points in each cluster%, obtained 
-        %from $D = Z^\top Z$
   \REPEAT
@@
         \STATE let $j$ be such that $x_i \in \C_j$
-        \STATE $j^\star \leftarrow \argmin_{\ell=1,\dotsc,k} J^{(\ell)}(x_i)$,
-            where $J^{(\ell)}(x_i)$ is defined in \eqref{eq:Jell}
-            %, for $\ell=1,2,\dots,k$
+        \STATE $j^\star \leftarrow \argmin_{\ell=1,\dotsc,k} J^{(\ell)}(x_i)$
+            (see \eqref{eq:Jell})
         \IF{ $j^\star \ne j$} 
-            \STATE move $x_i$ to $\C_{j^\star}$: $Z_{ij} \leftarrow 0$ and
-            $Z_{ij^\star} \leftarrow 1$
-            \STATE update $n$: $n_j \leftarrow n_j - 1$ and
-                    $n_{j^\star} \leftarrow n_{j^\star} + 1$
-            \STATE update $q$: $q_j \leftarrow q_j - 2Q_j(x_i)$ and
-    $q_{j^\star} \leftarrow q_{j^\star} + 2Q_{j^\star}(x_i)$
+            \STATE $Z_{ij} \leftarrow 0$ 
+            \STATE $Z_{ij^\star} \leftarrow 1$
+            \STATE $n_j \leftarrow n_j - 1$ 
+            \STATE $n_{j^\star} \leftarrow n_{j^\star} + 1$
+            \STATE $q_j \leftarrow q_j - 2Q_j(x_i)$ 
+            \STATE $q_{j^\star} \leftarrow q_{j^\star} + 2Q_{j^\star}(x_i)$
     %    \ELSE
@@
 $\mathcal{E}^{L}$-clustering is Lloyd's method for energy clustering, which
-is precisely~~~~~~~~~ \\
+is precisely
 kernel $k$-means algorithm, with the kernel induced by energy statistics. 
-This procedure~~~~~~~ \\
+This procedure
 finds
-local solutions to the optimization problem \eqref{eq:qcqp2}.\hspace{\fill}
+local solutions to the optimization problem \eqref{eq:qcqp2}.
 }
 \end{algorithm}
-\end{flushleft}
-\end{figure}
-
-\subsection*{Hartigan's Method for Energy Clustering}
+
+\subsection{Hartigan's Method for Energy Clustering}
 
@@
 $Q_j^{-} = Q_j - 2 Q_j(x_i) + G_{ii}$, we get the formula
-\begin{equation}
+\begin{multline}
 \label{eq:changeQ}
 \Delta Q^{j \to \ell}(x_i)  = 
-\dfrac{1}{n_j - 1}\left[ \dfrac{Q_j}{n_j} - 2 Q_j(x_i) + G_{ii} \right]
+\dfrac{1}{n_j - 1}\left[ \dfrac{Q_j}{n_j} - 2 Q_j(x_i) + G_{ii} \right] \\
 - \dfrac{1}{n_\ell + 1}\left[ \dfrac{Q_\ell}{n_\ell} - 2 Q_\ell(x_i) 
 - G_{ii} \right].
-\end{equation}
+\end{multline}
 Therefore, if $\Delta Q^{j\to \ell}(x_i) > 0$ we get closer to a 
@@
 
-\begin{figure}
-\begin{flushleft}
-\begin{algorithm}[H]
-\vspace{.5em}
+\begin{algorithm}
 \begin{algorithmic}[1]
-    \INPUT number of clusters $k$, Gram matrix $G$, 
-                initial label matrix $Z \leftarrow Z_0$
-    \OUTPUT label matrix $Z$
+    \INPUT $k$, $G$, $Z \leftarrow Z_0$
+    \OUTPUT $Z$
   \STATE $q \leftarrow (Q_1, \dotsc, Q_k)^\top$ 
-            have the energy costs of each cluster, defined in \eqref{eq:maxQ}
-  \STATE $n \leftarrow (n_1,\dotsc,n_k)^\top$ have the number of points 
-        in each cluster%, obtained from $D=Z^\top Z$
+            (see \eqref{eq:maxQ})
+  \STATE $n \leftarrow (n_1,\dotsc,n_k)^\top$ 
   \REPEAT
@@
                 \Delta Q^{j\to \ell}(x_i)$
-            %for $\ell=1,2,\dots,k$ and $\ell \ne j$; 
-            using \eqref{eq:changeQ} \label{stepmove}
+            (see \eqref{eq:changeQ}) \label{stepmove}
         \IF{ $\Delta Q^{j \to j^\star}(x_i) > 0$ }
-            \STATE move $x_i$ to $\C_{j^\star}$: $Z_{ij} \leftarrow 0$ and 
-            $Z_{ij^\star} \leftarrow 1$
-            \STATE update $n$: $n_j \leftarrow n_j - 1$ and
-                    $n_{j^\star} \leftarrow n_{j^\star} + 1$
-            \STATE update $q$: $q_j \leftarrow q_j - 2Q_j(x_i) + G_{ii}$ and
+            \STATE $Z_{ij} \leftarrow 0$
+            \STATE $Z_{ij^\star} \leftarrow 1$
+            \STATE $n_j \leftarrow n_j - 1$ 
+            \STATE $n_{j^\star} \leftarrow n_{j^\star} + 1$
+            \STATE $q_j \leftarrow q_j - 2Q_j(x_i) + G_{ii}$
+            \STATE 
             $q_{j^\star} \leftarrow q_{j^\star} + 2Q_{j^\star}(x_i)+ G_{ii}$
@@
 $\mathcal{E}^H$-clustering is Hartigan's method for energy clustering.
-This algorithm~~~~~~ \\ finds local solutions to  
+This algorithm finds local solutions to  
 the optimization problem \eqref{eq:qcqp2}.
-The steps $6$ and $10$ are different~~~~~~~~ \\
+The steps $6$, $12$ and $13$ are different
 than $\mathcal{E}^L$-clustering
-described in Algorithm~\ref{kmeans_algo}. \hspace{\fill}
+described in Algorithm~\ref{kmeans_algo}. 
 }
 \end{algorithm}
-\end{flushleft}
-\end{figure}
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Numerical Experiments}
@@
 $\mathcal{E}^H$-clustering, $\mathcal{E}^L$-clustering 
-and spectral clustering with the following semimetrics and corresponding
+and spectral clustering with the following metrics and corresponding
 generating kernels:
@@
 \rho_{\alpha}(x,y) &= \| x-y \|^{\alpha}, 
-&\kk_{\alpha}(x,y) &= \tfrac{1}{2} \left( \| x \|^{\alpha} + \| y \|^{\alpha} 
-- \| x-y \|^{\alpha} \right), \label{eq:rho_alpha} \\
+%\kk_{\alpha}(x,y) = \tfrac{1}{2} \left( \| x \|^{\alpha} + \| y \|^{\alpha} 
+%- \| x-y \|^{\alpha} \right), 
+\label{eq:rho_alpha} \\
 %
 \widetilde{\rho}_{\sigma}(x,y) &= 2 - 2 e^{-\tfrac{\|x-y\|}{2 \sigma}},  
-&\widetilde{\kk}_{\sigma}(x,y) &= e^{-\tfrac{\|x-y\|}{2\sigma}}, 
+%\widetilde{\kk}_{\sigma}(x,y) = e^{-\tfrac{\|x-y\|}{2\sigma}}, 
 \label{eq:rho_tilde}\\
 %
-\widehat{\rho}_{\sigma}(x,y) &= 2 - 2 e^{-\tfrac{\|x-y\|^2}{2 \sigma^2}}, 
-&\widehat{\kk}_{\sigma}(x,y) &= e^{-\tfrac{\|x-y\|^2}{2\sigma^2}}  .
+\widehat{\rho}_{\sigma}(x,y) &= 2 - 2 e^{-\tfrac{\|x-y\|^2}{2 \sigma^2}}. 
+%\widehat{\kk}_{\sigma}(x,y) = e^{-\tfrac{\|x-y\|^2}{2\sigma^2}}  
 \label{eq:rho_hat}
@@
 For $k$-means, GMM and spectral clustering we use the robust 
-implementations of \emph{scikit-learn} library \citep{scikit-learn}, where  
+implementations of \emph{scikit-learn} library \cite{scikit-learn}, where  
 $k$-means is initialized with 
-$k$-means++ \citep{Vassilvitskii}, 
+$k$-means++ \cite{Vassilvitskii}, 
 and GMM with the output of $k$-means, making it more robust and preventing
 it from breaking in high dimensions. 
-Spectral clustering implementation is based on \citep{Malik}. 
+Spectral clustering implementation is based on \cite{Malik}. 
 We implemented $\mathcal{E}^L$-clustering as described
@@
 \begin{figure}
-\begin{minipage}{0.49\textwidth}
-\includegraphics[width=1\textwidth]{pairsplot1.pdf}\\[-1.0em] (a)
-\end{minipage}
-\begin{minipage}{0.49\textwidth}
-\includegraphics[width=1\textwidth]{pairsplot2.pdf}\\[-1.0em] (a)
-\end{minipage}
+\begin{center}
+\includegraphics[width=.8\linewidth]{pairsplot1.pdf}\\[-1em]
+\includegraphics[width=.8\linewidth]{pairsplot2.pdf}
+\put(-210,5){\scriptsize{(b)}}
+\put(-210,210){\scriptsize{(a)}}
+\end{center}
 \caption{
@@
 \tfrac{1}{2} \mathcal{N}(\mu_1,\Sigma_1) +
-\tfrac{1}{2} \mathcal{N}(\mu_2,\Sigma_2), \\
-\Sigma_1&=\Sigma_2 = I_D, \quad
-\mu_1 = (\underbrace{0,\dotsc,0}_{\times D})^\top, \quad
+\tfrac{1}{2} \mathcal{N}(\mu_2,\Sigma_2), \quad
+\Sigma_1=\Sigma_2 = I_D, \\
+\mu_1 &= (\underbrace{0,\dotsc,0}_{\times D})^\top, \qquad
 \mu_2 = 0.7 (\underbrace{1,\dots,1}_{\times 10},
@@
 \begin{figure}
-\begin{minipage}{0.49\textwidth}
-\includegraphics[width=1\textwidth]{normal_highdim_mean.pdf}\\[-1.0em] (a)
+\begin{minipage}{0.49\linewidth}
+\includegraphics[width=1\textwidth]{normal_highdim_mean.pdf}
 \end{minipage}
-\begin{minipage}{0.49\textwidth}
-\includegraphics[width=1\textwidth]{normal_highdim_cov.pdf}\\[-1.0em] (b)
+\begin{minipage}{0.49\linewidth}
+\includegraphics[width=1\textwidth]{normal_highdim_cov.pdf}
 \end{minipage}
+\put(-245,-35){\scriptsize{(a)}}
+\put(-120,-35){\scriptsize{(b)}}
 \caption{
@@
 \tfrac{1}{2}\mathcal{N}(\mu_2,\Sigma_2),  \\
-\mu_1 &= (\underbrace{0,\dotsc,0}_{\times D})^\top , \quad
+\mu_1 &= (\underbrace{0,\dotsc,0}_{\times D})^\top, \qquad
 \mu_2 = (\underbrace{1,\dots,1}_{\times 10}, 
-\underbrace{0,\dots,0}_{\times (D-10)})^\top, \quad
-\Sigma_1 = I_D, \quad
+\underbrace{0,\dots,0}_{\times (D-10)})^\top, \\
+\Sigma_1 &= I_D, \qquad
 \Sigma_2 = \left( \begin{array}{c|c}
@@
 0 & I_{D-10} \end{array}\right), \\
-\widetilde{\Sigma}_{10} &= \diag(1.367,  3.175,  3.247,  4.403,  1.249,
-1.969, 4.035,   4.237,  2.813,  3.637).
+\widetilde{\Sigma}_{10} &= \diag(1.367,  3.175,  3.247,  4.403,  
+1.249, 1.969, \\
+& \qquad \qquad 4.035,   4.237,  2.813,  3.637).
 \end{split}
@@
 x &\stackrel{iid}{\sim} \tfrac{1}{2} \mathcal{N}(\mu_1,\Sigma_1)+
-\tfrac{1}{2} \mathcal{N}(\mu_2,\Sigma_2) , \\
-\mu_1 &= (\underbrace{0,\dotsc,0}_{\times 20})^\top ,\quad
+\tfrac{1}{2} \mathcal{N}(\mu_2,\Sigma_2) , \quad
+2\Sigma_1 = \Sigma_2 = I_{20} \\
+\mu_1 &= (\underbrace{0,\dotsc,0}_{\times 20})^\top , \quad
 \mu_2 = \tfrac{1}{2} 
-(\underbrace{1,\dotsc,1}_{5},\underbrace{0,\dotsc,0}_{15})^\top, \quad
-\Sigma_1 = \tfrac{1}{2} I_{20},  \quad
-\Sigma_2 = I_{20}.
+(\underbrace{1,\dotsc,1}_{5},\underbrace{0,\dotsc,0}_{15})^\top. 
 \end{split}
@@
 \begin{figure}
-\begin{minipage}{0.49\textwidth}
-\centering
-\includegraphics[width=1\textwidth]{normal_kernels.pdf}\\[-1.0em](a)
+\begin{minipage}{0.49\linewidth}
+\includegraphics[width=1\textwidth]{normal_kernels.pdf}
+\includegraphics[width=1\textwidth]{normal_kernels_difference.pdf}
 \end{minipage}
-\begin{minipage}{0.49\textwidth}
-\centering
-\includegraphics[width=1\textwidth]{lognormal_kernels.pdf}\\[-1.0em](c)
+\begin{minipage}{0.49\linewidth}
+\includegraphics[width=1\textwidth]{lognormal_kernels.pdf}
+\includegraphics[width=1\textwidth]{lognormal_kernels_difference.pdf}
 \end{minipage}
-\begin{minipage}{0.49\textwidth}
-\centering
-\includegraphics[width=1\textwidth]{normal_kernels_difference.pdf}\\[-1.0em](b)
-\end{minipage}
-\begin{minipage}{0.49\textwidth}
-\centering
-\includegraphics[width=1\textwidth]{lognormal_kernels_difference.pdf}\\[-1.0em]
-(d)
-\end{minipage}
+\put(-240,8){\scriptsize{(a)}}
+\put(-240,-75){\scriptsize{(b)}}
+\put(-110,8){\scriptsize{(c)}}
+\put(-110,-75){\scriptsize{(d)}}
 \caption{
 \label{fig:consist}
-$\mathcal{E}^H$-clustering with kernels \eqref{eq:rho_alpha} and 
+$\mathcal{E}^H$-clustering with metrics \eqref{eq:rho_alpha} and 
 \eqref{eq:rho_tilde} versus
@@
 between $\mathcal{E}^H$ versus $\mathcal{E}^L$ (kernel $k$-means) and
-spectral clustering, with the kernel $\widetilde{K}_1$.
+spectral clustering, with metric $\widetilde{\rho}_1$.
 }
@@
 x &\stackrel{iid}{\sim} \tfrac{1}{2} 
-\exp\left\{\mathcal{N}(\mu_1,\Sigma_1)\right\}+
-\tfrac{1}{2} \exp\left\{\mathcal{N}(\mu_2,\Sigma_2)\right\} , \\
+e^{\mathcal{N}(\mu_1,\Sigma_1)}+
+\tfrac{1}{2} e^{\mathcal{N}(\mu_2,\Sigma_2)} , \quad
+2\Sigma_1 = \Sigma_2 = I_{20}, \\
 \mu_1 &= (\underbrace{0,\dotsc,0}_{\times 20})^\top ,\quad
 \mu_2 = \tfrac{1}{2} 
-(\underbrace{1,\dotsc,1}_{5},\underbrace{0,\dotsc,0}_{15})^\top, \quad
-\Sigma_1 = \tfrac{1}{2} I_{20},  \quad
-\Sigma_2 = I_{20}.
+(\underbrace{1,\dotsc,1}_{5},\underbrace{0,\dotsc,0}_{15})^\top.
 \end{split}
@@
 \begin{figure}
-\begin{minipage}{0.24\textwidth}
-\includegraphics[width=1\textwidth]{2cigars.pdf}\\[-1em] (a)
+\begin{minipage}{0.24\linewidth}
+\includegraphics[width=1\textwidth]{2cigars.pdf}\\\scriptsize{(a)}
 \end{minipage}
-\begin{minipage}{0.24\textwidth}
-\includegraphics[width=1\textwidth]{2circles.pdf}\\[-1em] (b)
+\begin{minipage}{0.24\linewidth}
+\includegraphics[width=1\textwidth]{2circles.pdf}\\\scriptsize{ (b)}
 \end{minipage}
-\begin{minipage}{0.24\textwidth}
-\includegraphics[width=1\textwidth]{3circles.pdf}\\[-1em] (c)  
+\begin{minipage}{0.24\linewidth}
+\includegraphics[width=1\textwidth]{3circles.pdf}\\\scriptsize{(c)  }
 \end{minipage}
-\begin{minipage}{0.23\textwidth}
-\vspace{-0.9em}
-\includegraphics[width=1\textwidth]{mnist.pdf}\\[-1.8em] (d)  
+\begin{minipage}{0.23\linewidth}
+\vspace{-.5em}
+\includegraphics[width=1\textwidth]{mnist.pdf}\\[-.6em]\scriptsize{(d)  }
 \end{minipage}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Real Data Experiment}
+
+\begin{table*}
+\caption{\label{table:synapse}
+Features of the synapse dataset according to \cite{Collman2015}.
+}
+\begin{center}
+\begin{tabular}{lllllllllll}
+\toprule[1pt]
+$x_{1}$ &
+$x_{2}$ &
+$x_{3}$ &
+$x_{4}$ &
+$x_{5}$ &
+$x_{6}$ &
+$x_{7}$ &
+$x_{8}$ &
+$x_{9}$ &
+$x_{10}$ &
+$x_{11}$ \\
+\midrule[.5pt]
+bIIItubulin&DAPI-2nd&DAPI-3rd&GABA&GAD2&VGAT&gephyrin&NR1&VGluT1&synapsin&PSDr
+\\
+\bottomrule[1pt]
+\end{tabular}
+\end{center}
+\end{table*}
+
+\begin{figure*}
+\includegraphics[width=0.18\textwidth]{eigen_gap.pdf}
+\includegraphics[width=0.2\textwidth]{heat_means_k6_energy.pdf}
+\includegraphics[width=0.2\textwidth]{synapse_cluster_2d_k6_energy.pdf}
+\includegraphics[width=0.2\textwidth]{heat_means_k6_gauss.pdf}
+\includegraphics[width=0.2\textwidth]{synapse_cluster_2d_k6_gauss.pdf}
+\put(-510,-5){\scriptsize{(a)}}
+\put(-415,-5){\scriptsize{(b)}}
+\put(-310,-5){\scriptsize{(c)}}
+\put(-205,-5){\scriptsize{(d)}}
+\put(-100,-5){\scriptsize{(e)}}
+\caption{\label{fig:synapse}
+Clustering synapse dataset.
+(a) Eigenvalues of random walk Laplacian. We choose
+$k=6$ since its the first time we see a peak. We use energy clustering
+with the two metrics $\rho_1$ and $\widehat{\rho}_2$.
+(b) Heat map of the cluster means versus features
+using energy clustering with standard
+metric $\rho_1$ from energy statistics.
+(c) Clustered points projected into the two principal components,
+using $\rho_1$.
+(d) Heat map of the cluster means versus features
+using energy clustering with Gaussian metric 
+$\widehat{\rho}_2$.
+(e) Clustered points projected into the two principal components,
+using $\widehat{\rho}_2$.
+}
+\end{figure*}
+
+We now consider a dataset  that
+describes protein expression of neural synapses --- chemical 
+connections between neurons --- which is
+crucial for understanding the diversity of synapses. This is
+part of a large NIH funded consortium within the BRAIN initiative.
+The details about this detaset, including experimental methods, are
+fully described in \cite{Collman2015}.
+Here we have $1025$ data points
+with $11$ features, as described in Table~\ref{table:synapse}.
+Each point
+corresponds to a $(x,y,z)$ location in the brain. We normalize this
+dataset before applying $\mathcal{E}^H$-clustering.
+
+We use the two
+metrics $\rho_1$ and $\widehat{\rho}_2$; see \eqref{eq:rho_alpha} and
+\eqref{eq:rho_hat}.
+We compute the difference between eigenvalues of the random walk Laplacian
+obtained from the kernel matrix $G$ constructed from these metrics;
+see \eqref{eq:kernel_matrix} and \eqref{eq:kernel_semimetric}.
+The choice of $k$ corresponds to the first time we see a meaningful
+peak in the plot, which occurs at $k=6$ as illustrated in 
+Fig.~\ref{fig:synapse}a.
+This procedure to find $k$ is common in the literature \cite{vonLuxburg2007}.
+
+Having found $k,$ we now cluster the data.
+First we use the metric $\rho_1$, which
+is the standard Euclidean choice in energy statistics.
+In Fig.~\ref{fig:synapse}b and Fig.~\ref{fig:synapse}c we show a heat
+map of the cluster means versus features and the clustered data points
+projected into the 2 principal components, respectively.
+The plots of Fig.~\ref{fig:synapse}d and Fig.~\ref{fig:synapse}e show the same
+experiment but using the metric $\widehat{\rho}_2$ corresponding
+to a Gaussian kernel.
+
+We remark that we also used the gap statistic \cite{Tibshirani2001} based
+on $k$-means to determine the number of clusters as a comparison.
+The results slightly suggest $k=6$, but it does not pass the gap
+statistic test.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Discussion}
@@
 \dfrac{n_1}{2N} \mathcal{N}(\mu_1,\Sigma_1)+
-\dfrac{n_1}{2N} \mathcal{N}(\mu_1,\Sigma_1), \quad 
-\mu_1 = (0,0,0,0)^\top , \quad
+\dfrac{n_1}{2N} \mathcal{N}(\mu_1,\Sigma_1), \\
+\mu_1 &= (0,0,0,0)^\top , \quad
 \mu_2 = 1.5\times (1,1,0,0)^\top, \\
@@
 0 & I_2 
-\end{array}\right), \quad
-n_1 = N - m, \quad  n_2 = N + m, \quad N=300.
+\end{array}\right), \\
+n_1 &= N - m, \quad  n_2 = N + m, \quad N=300.
 \end{split}
@@
 The results are shown in Fig.~\ref{fig:unbalanced}.
-For highly unbalanced clusters we see that GMM performs better than
+For highly unbalanced clusters, we see that GMM performs better than
 the other methods, which have basically similar performance.
@@
 \begin{figure}
-\includegraphics[width=0.5\textwidth]{normal_unbalanced.pdf}\vspace{-1em}
+\centering
+\includegraphics[width=0.7\linewidth]{normal_unbalanced.pdf}
 \caption{
@@
 
-\begin{acknowledgements}
+% if have a single appendix:
+%\appendix[Proof of the Zonklar Equations]
+% or
+%\appendix  % for no appendix heading
+% do not use \section anymore after \appendix, only \section*
+% is possibly needed
+
+% use appendices with more than one appendix
+% then use \section to start each appendix
+% you must declare a \section before using any
+% \subsection or using \label (\appendices by itself
+% starts a section numbered zero.)
+%
+
+%\appendices
+%\section{Proof of the First Zonklar Equation}
+%Appendix one text goes here.
+
+% you can choose not to have a title for an appendix
+% if you want by leaving the argument blank
+%\section{}
+%Appendix two text goes here.
+
+% use section* for acknowledgment
+\ifCLASSOPTIONcompsoc
+  % The Computer Society usually uses the plural form
+  \section*{Acknowledgments}
+\else
+  % regular IEEE prefers the singular form
+  \section*{Acknowledgment}
+\fi
+
 We would like to thank Carey Priebe for discussions.
@@
 Agency’s (DARPA) SIMPLEX program through SPAWAR contract N66001-15-C-4041.
-\end{acknowledgements}
+
+% Can use something like this to put references on a page
+% by themselves when using endfloat and the captionsoff option.
+\ifCLASSOPTIONcaptionsoff
+  \newpage
+\fi
+
+% trigger a \newpage just before the given reference
+% number - used to balance the columns on the last page
+% adjust value as needed - may need to be readjusted if
+% the document is modified later
+%\IEEEtriggeratref{8}
+% The "triggered" command can be changed if desired:
+%\IEEEtriggercmd{\enlargethispage{-5in}}
+
+% references section
+
+% can use a bibliography generated by BibTeX as a .bbl file
+% BibTeX documentation can be easily obtained at:
+% http://mirror.ctan.org/biblio/bibtex/contrib/doc/
+% The IEEEtran BibTeX style support page is at:
+% http://www.michaelshell.org/tex/ieeetran/bibtex/
+%\bibliographystyle{IEEEtran}
+% argument is your BibTeX string definitions and bibliography database(s)
+%\bibliography{IEEEabrv,../bib/paper}
+%
+% <OR> manually copy in the resultant .bbl file
+% set second argument of \begin to the number of references
+% (used to reserve space for the reference number labels box)
 
@@
 
+% biography section
+% 
+% If you have an EPS/PDF photo (graphicx package needed) extra braces are
+% needed around the contents of the optional argument to biography to prevent
+% the LaTeX parser from getting confused when it sees the complicated
+% \includegraphics command within an optional argument. (You could create
+% your own custom macro containing the \includegraphics command to make things
+% simpler here.)
+%\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{mshell}}]{Michael Shell}
+% or if you just want to reserve a space for a photo:
+
+\begin{IEEEbiography}{Guilherme Fran\c ca}
+Biography text here.
+\end{IEEEbiography}
+
+% if you will not have a photo at all:
+\begin{IEEEbiographynophoto}{Maria Rizzo}
+Biography text here.
+\end{IEEEbiographynophoto}
+
+% insert where needed to balance the two columns on the last page with
+% biographies
+%\newpage
+
+\begin{IEEEbiographynophoto}{Joshua Vogelstein}
+Biography text here.
+\end{IEEEbiographynophoto}
+
+% You can push biographies down or up by placing
+% a \vfill before or after them. The appropriate
+% use of \vfill depends on what kind of text is
+% on the last page and whether or not the columns
+% are being equalized.
+
+%\vfill
+
+% Can be used to pull up biographies so that the bottom of the last one
+% is flush with the other column.
+%\enlargethispage{-5in}
+
+% that's all folks
 \end{document}

### 2018-06-04 energy_clustering/kgroups_code/bayes.py

@@ -1,0 +1,212 @@
+"""Functions to compute Bayes errors."""
+
+# Guilherme S. Franca <guifranca@gmail.com>
+# Johns Hopkins University
+
+
+from __future__ import division
+
+from functools import partial
+from scipy.optimize import minimize
+from scipy.optimize import root
+import numpy as np
+import scipy.stats as stats
+import scipy.integrate as integrate
+import matplotlib.pyplot as plt
+
+import data
+
+
+def bayes_univariate_normal(mu1, mu2, sigma1, sigma2, 
+                            num_times=10, num_points=1000, p=0.5):
+    """Estimate Bayes' error for two-class normal distributions in 1D."""
+    error = np.zeros(num_times)
+    for i in range(num_times):
+        q = 1-p
+        n1, n2 = np.random.multinomial(num_points, [p,q])
+        x1 = np.random.normal(mu1, sigma1, n1)
+        x2 = np.random.normal(mu2, sigma2, n2)
+    
+        error1 = sum([1 for x in x1 
+            if stats.norm.pdf(x,mu1,sigma1)-stats.norm.pdf(x,mu2,sigma2)<=0
+        ])
+        error2 = sum([1 for x in x2 
+            if stats.norm.pdf(x,mu1,sigma1)-stats.norm.pdf(x,mu2,sigma2)>=0
+        ])
+        error[i] = p*error1/n1+q*error2/n2
+    return 1-error.mean()
+
+def integrate_univariate_normal(mu1, mu2, sigma1, sigma2, p=0.5):
+    """Same as above but through numerical integration."""
+    
+    g1 = partial(stats.norm.pdf, loc=mu1, scale=sigma1)
+    g2 = partial(stats.norm.pdf, loc=mu2, scale=sigma2)
+    
+    # make it visual to confirm
+    #xs = np.arange(-10,20, 0.001)
+    #y1s = np.array([g1(x) for x in xs])
+    #y2s = np.array([g2(x) for x in xs])
+    #y3s = np.array([np.abs(g1(x)-g2(x)) for x in xs])
+    #plt.plot(xs, y1s)
+    #plt.plot(xs, y2s)
+    #plt.plot(xs, y3s)
+    #plt.show()
+
+    x0 = root(lambda x: g1(x) - g2(x), 1).x[0]
+    
+    # another way
+    #cons = (
[…]

### 2018-06-04 energy_clustering/kgroups_code/data/gauss2_4_20exp.py

@@ -1,0 +1,2901 @@
+method,dimension,accuracy
+k-means,10,0.91
+gmm,10,0.905
+spectral clustering,10,0.885
+kernel k-means,10,0.955
+kernel k-groups,10,0.95
+k-means,35,0.825
+gmm,35,0.845
+spectral clustering,35,0.875
+kernel k-means,35,0.94
+kernel k-groups,35,0.935
+k-means,60,0.815
+gmm,60,0.795
+spectral clustering,60,0.8
+kernel k-means,60,0.835
+kernel k-groups,60,0.805
+k-means,85,0.82
+gmm,85,0.77
+spectral clustering,85,0.825
+kernel k-means,85,0.825
+kernel k-groups,85,0.855
+k-means,110,0.775
+gmm,110,0.765
+spectral clustering,110,0.775
+kernel k-means,110,0.77
+kernel k-groups,110,0.78
+k-means,135,0.82
+gmm,135,0.57
+spectral clustering,135,0.79
+kernel k-means,135,0.71
+kernel k-groups,135,0.815
+k-means,160,0.82
+gmm,160,0.865
+spectral clustering,160,0.825
+kernel k-means,160,0.775
+kernel k-groups,160,0.83
+k-means,185,0.845
+gmm,185,0.825
+spectral clustering,185,0.855
+kernel k-means,185,0.75
+kernel k-groups,185,0.86
+k-means,210,0.86
+gmm,210,0.81
+spectral clustering,210,0.91
+kernel k-means,210,0.75
+kernel k-groups,210,0.87
+k-means,235,0.625
+gmm,235,0.62
+spectral clustering,235,0.785
+kernel k-means,235,0.615
+kernel k-groups,235,0.795
+k-means,260,0.765
+gmm,260,0.835
+spectral clustering,260,0.84
+kernel k-means,260,0.63
+kernel k-groups,260,0.845
+k-means,285,0.735
+gmm,285,0.555
+spectral clustering,285,0.78
+kernel k-means,285,0.6
+kernel k-groups,285,0.8
+k-means,310,0.705
+gmm,310,0.625
+spectral clustering,310,0.8
+kernel k-means,310,0.62
+kernel k-groups,310,0.82
+k-means,335,0.62
+gmm,335,0.505
+spectral clustering,335,0.685
+kernel k-means,335,0.635
+kernel k-groups,335,0.685
+k-means,360,0.63
+gmm,360,0.625
+spectral clustering,360,0.805
[…]

### 2018-06-04 energy_clustering/kgroups_code/data/gauss2_5_20exp.py

@@ -1,0 +1,2901 @@
+method,dimension,accuracy
+k-means,10,0.84
+gmm,10,0.95
+spectral clustering,10,0.865
+kernel k-means,10,0.92
+kernel k-groups,10,0.925
+k-means,35,0.855
+gmm,35,0.815
+spectral clustering,35,0.865
+kernel k-means,35,0.865
+kernel k-groups,35,0.855
+k-means,60,0.845
+gmm,60,0.83
+spectral clustering,60,0.845
+kernel k-means,60,0.84
+kernel k-groups,60,0.855
+k-means,85,0.79
+gmm,85,0.74
+spectral clustering,85,0.8
+kernel k-means,85,0.84
+kernel k-groups,85,0.82
+k-means,110,0.83
+gmm,110,0.77
+spectral clustering,110,0.84
+kernel k-means,110,0.765
+kernel k-groups,110,0.82
+k-means,135,0.845
+gmm,135,0.82
+spectral clustering,135,0.805
+kernel k-means,135,0.715
+kernel k-groups,135,0.835
+k-means,160,0.7
+gmm,160,0.555
+spectral clustering,160,0.815
+kernel k-means,160,0.645
+kernel k-groups,160,0.815
+k-means,185,0.79
+gmm,185,0.86
+spectral clustering,185,0.85
+kernel k-means,185,0.82
+kernel k-groups,185,0.835
+k-means,210,0.78
+gmm,210,0.515
+spectral clustering,210,0.85
+kernel k-means,210,0.535
+kernel k-groups,210,0.85
+k-means,235,0.715
+gmm,235,0.61
+spectral clustering,235,0.76
+kernel k-means,235,0.505
+kernel k-groups,235,0.815
+k-means,260,0.625
+gmm,260,0.625
+spectral clustering,260,0.775
+kernel k-means,260,0.665
+kernel k-groups,260,0.805
+k-means,285,0.68
+gmm,285,0.77
+spectral clustering,285,0.825
+kernel k-means,285,0.67
+kernel k-groups,285,0.855
+k-means,310,0.605
+gmm,310,0.735
+spectral clustering,310,0.87
+kernel k-means,310,0.595
+kernel k-groups,310,0.86
+k-means,335,0.73
+gmm,335,0.65
+spectral clustering,335,0.85
+kernel k-means,335,0.555
+kernel k-groups,335,0.86
+k-means,360,0.685
+gmm,360,0.615
+spectral clustering,360,0.78
[…]

### 2018-06-04 energy_clustering/kgroups_code/dermatology_table.py

@@ -1,0 +1,121 @@
+"""Construct table based on dermatology clustering.
+
+We obtained the following:
+
++---------------------+----------------+----------------+
+|      Algorithm      |    Accuracy    |     A-Rand     |
++---------------------+----------------+----------------+
+|        kmeans       | 0.724043715847 | 0.69533698756  |
+|         GMM         | 0.893442622951 |  0.8465538079  |
+| spectral clustering | 0.953551912568 | 0.912517180432 |
+|       spectral      | 0.950819672131 | 0.904961836633 |
+|    kernel k-means   | 0.882513661202 | 0.861732888116 |
+|   kernel k-groups   | 0.96174863388  | 0.935633109305 |
++---------------------+----------------+----------------+
+
+kernel k-means and k-groups were initialized with k-means++
+
+
+The number of points are:
+
+[[112   0   0   0   0   0]
+ [  0  61   0   0   0   0]
+ [  0   0  72   0   0   0]
+ [  0   0   0  49   0   0]
+ [  0   0   0   0  52   0]
+ [  0   0   0   0   0  20]]
+
+The estimated number of points are:
+
+[[ 51   0   0   0   0   0]
+ [  0  52   0   0   0   0]
+ [  0   0 112   0   0   0]
+ [  0   0   0  59   0   0]
+ [  0   0   0   0  72   0]
+ [  0   0   0   0   0  20]]
+
+These results are in the appropriate files inside ./data/
+
+"""
+
+# author: Guilherme Franca <guifranca@gmail.com>
+# Johns Hopkinks University
+
+import numpy as np
+import pandas as pd
+from prettytable import PrettyTable
+
+import eclust
+
+Z = pd.read_csv("data/dermatology_true_label_matrix.csv", header=None).values
+Zh = pd.read_csv("data/dermatology_pred_label_matrix.csv", header=None).values
+
+df = pd.read_csv('data/dermatology_data.csv', sep=',', header=None)
+X = df.values[:,:-1]
+missing_id = np.where(X=='?')
+not_missing_id = [i for i in range(len(X)) if i not in missing_id[0]]
+mean_age = np.array(X[not_missing_id,33], dtype=float).mean()
+X[missing_id] = mean_age
+X = np.array(X, dtype=float)
+
+print "True number of points"
+print Z.T.dot(Z)
[…]

### 2018-06-04 energy_clustering/kgroups_code/eclust.py

@@ -1,0 +1,266 @@
+"""Iterative algorithms to optimize QCQP for energy statistics clustering."""
+
+# Guilherme Franca <guifranca@gmail.com>
+# Johns Hopkins University
+
+
+from __future__ import division
+
+import numpy as np
+from sklearn.metrics.pairwise import pairwise_distances
+
+def ztoZ(z):
+    """Convert label vector to label matrix."""
+    n = z.shape[0]
+    k = np.unique(z).shape[0]
+    Z = np.zeros((n,k))
+    for i in range(n):
+        Z[i, int(z[i])] = 1
+    return Z
+
+def Ztoz(Z):
+    """Convert label matrix to label vector."""
+    n, k = Z.shape
+    z = np.zeros(n)
+    points, labels = np.where(Z==1)
+    for i in range(len(points)):
+        z[points[i]] = labels[i]
+    return z
+
+def objective(Z, G, W):
+    """Compute objective function."""
+    W_half = np.sqrt(W)
+    W_half_Z = W_half.dot(Z) # This is the same as Z with 1 -> w_i^{1/2} 
+    Sum_weights = W_half_Z.T.dot(W_half_Z) # diagonal matrix with s_i 
+    Y = Z.dot(np.linalg.inv(np.sqrt(Sum_weights)))
+    G_tilde = W.dot(G.dot(W))
+    return np.trace(Y.T.dot(G_tilde.dot(Y)))
+
+def kernel_function(x, y, rho, x0=None):
+    """Return kernel function based on rho."""
+    if type(x0) == type(None):
+        x0 = np.zeros(x.shape)
+    val = 0.5*(rho(x,x0) + rho(y,x0) - rho(x,y))
+    return val
+
+def kernel_matrix(X, rho, x0=None):
+    """Compute Kernel matrix based on kernel function K(x,y)."""
+    kfunc = lambda x, y: kernel_function(x, y, rho, x0)
+    #G = pairwise_distances(X, metric=kfunc, n_jobs=4)
+    G = pairwise_distances(X, metric=kfunc)
+    return G
+
+def kernel_kgroups(k, G, Z0, W, max_iter=100, tol=1e-4, verbose=False,
+                   return_Z=False):
+    """Optimize the W objective function by considering moving points
+    to different partitions. Compute the change in the cost function by
+    moving a point then decide the best partition to optimize the cost
+    function. 
+    
+    """
+    n = G.shape[0]
+    Z = np.copy(Z0)
+    Zt = Z.T
[…]

### 2018-06-04 energy_clustering/kgroups_code/gauss2.py

@@ -1,0 +1,75 @@
+"""Experiment for high dimensional Gaussians, here we change the covariance."""
+
+# author: Guilherme S. Franca <guifranca@gmail.com>
+# Johns Hopkins University
+
+from __future__ import division
+
+import numpy as np
+import multiprocessing as mp
+import pandas as pd
+
+import argparse
+
+import data
+import eclust
+import metric
+import wrapper
+
+parser = argparse.ArgumentParser(description="High dimensional Gaussian"\
+                                             "changing the covariance.")
+parser.add_argument('-o', dest='output', required=True,
+                    type=str, action='store', help="file name for output")
+parser.add_argument('-n', type=int, required=True,
+                    dest='num_experiments', action='store', 
+                    help="number of experiments")
+
+args = parser.parse_args()
+
+dimensions = range(10, 725, 25)
+k = 2
+total_points = 200
+output = args.output
+num_experiments = args.num_experiments
+
+def generate_data(D):
+    d = 10
+    m1 = np.zeros(D)
+    s1 = np.eye(D)
+    m2 = np.concatenate((np.ones(d), np.zeros(D-d)))
+    s2_1 = np.array([1.367,  3.175,  3.247,  4.403,  1.249,
+                     1.969, 4.035,   4.237,  2.813,  3.637])
+    s2 = np.diag(np.concatenate((s2_1, np.ones(D-d))))
+    n1, n2 = np.random.multinomial(total_points, [0.5, 0.5])
+    X, z = data.multivariate_normal([m1, m2], [s1, s2], [n1, n2])
+    return X, z
+
+r = []
+for _ in range(num_experiments):
+    for dim in dimensions:
+        X, z = generate_data(dim)
+        G = eclust.kernel_matrix(X, lambda x, y: np.linalg.norm(x-y))
+        
+        zh = wrapper.kmeans(k, X)
+        a = metric.accuracy(z, zh)
+        r.append(['k-means', dim, a])
+        
+        zh = wrapper.gmm(k, X)
+        a = metric.accuracy(z, zh)
+        r.append(['gmm', dim, a])
+        
+        zh = wrapper.spectral_clustering(k, X, G)
+        a = metric.accuracy(z, zh)
+        r.append(['spectral clustering', dim, a])
+        
+        zh = wrapper.kernel_kmeans(k, X, G)
+        a = metric.accuracy(z, zh)
+        r.append(['kernel k-means', dim, a])
[…]

### 2018-06-04 energy_clustering/kgroups_code/init.py

@@ -1,0 +1,222 @@
+"""Initialization methods."""
+
+# Guilherme Franca <guifranca@gmail.com>
+# Johns Hopkins University
+
+from __future__ import division
+
+import numpy as np
+from scipy.linalg import eigh
+from sklearn.cluster import KMeans
+from sklearn.cluster import SpectralClustering
+
+from sklearn.metrics.pairwise import pairwise_distances
+
+
+def euclidean_distance(x, y):
+    return np.linalg.norm(x-y)**2
+
+def kmeans_plus(k, X):
+    """This is the k-means++ initialization proposed by Arthur and
+    Vassilvitskii (2007). We label the points according to closest
+    distance to the centers.
+    
+    """
+    K = k
+    N = X.shape[0]
+    distance = euclidean_distance
+    C = [X[np.random.randint(0, N)]] # centers
+    D = np.zeros(N)                  # distances
+    for k in range(1, K):
+        for n in range(N):
+            D[n] = np.array([distance(X[n], c) for c in C]).min()
+        p = D/D.sum()
+        j = discrete_rv(p)
+        C.append(X[j])
+    mus = np.array(C)
+    
+    labels = np.empty(N, dtype=np.int8)
+    for n in range(N):
+        for k in range(K):
+            D = np.array([distance(X[n], mus[k]) for k in range(K)])
+            labels[n] = np.argmin(D)
+    return labels
+
+def kmeans_plus2(k, X):
+    """This is the k-means++ initialization proposed by Arthur and
+    Vassilvitskii (2007). We label the points according to closest
+    distance to the centers.
+
+    Same as above but return labels and means.
+    
+    """
+    K = k
+    N = X.shape[0]
+    distance = euclidean_distance
+    C = [X[np.random.randint(0, N)]] # centers
+    D = np.zeros(N)                  # distances
+    for k in range(1, K):
+        for n in range(N):
+            D[n] = np.array([distance(X[n], c) for c in C]).min()
+        p = D/D.sum()
+        j = discrete_rv(p)
+        C.append(X[j])
+    mus = np.array(C)
+    
+    labels = np.empty(N, dtype=np.int8)
+    for n in range(N):
+        for k in range(K):
+            D = np.array([distance(X[n], mus[k]) for k in range(K)])
[…]

### 2018-06-04 energy_clustering/kgroups_code/plot_gauss2.py

@@ -1,0 +1,48 @@
+# author: Guilherme S. Franca
+# Johns Hopkins University
+
+import matplotlib.pyplot as plt
+import numpy as np
+import pandas as pd
+import argparse
+import sys
+
+parser = argparse.ArgumentParser(description="Plot the high dimensional"\
+                    "Gaussian experiment which changes in the covariance.")
+parser.add_argument('-o', dest='output', required=True,
+                    type=str, action='store', help="file name for output")
+parser.add_argument('-i', dest='input', required=True,
+                    type=str, action='store', nargs='*', 
+                    help="file name for output")
+
+args = parser.parse_args()
+
+fnames = args.input
+output = args.output
+
+df = pd.concat(pd.read_csv(f) for f in fnames)
+
+methods = ['k-means', 'gmm', 'spectral clustering', 'kernel k-means', 
+           'kernel k-groups']
+markers = iter(['^', 'v', 'p', 's', 'o'])
+
+dimensions = np.unique(df['dimension'].values)
+
+fig = plt.figure()
+ax = fig.add_subplot(111)
+for method in methods:
+    r = []
+    for d in dimensions:
+        df2 = df[(df['method']==method) & (df['dimension']==d)]
+        r.append([d, df2['accuracy'].mean(), df2['accuracy'].sem()])
+    r = np.array(r)
+    ax.plot(r[:,0], [0.95]*len(r[:,0]), '--', linewidth=1, color='k', alpha=0.5)
+    mk = next(markers)
+    ax.errorbar(r[:,0], r[:,1], yerr=r[:,2], marker=mk, elinewidth=1,
+                label=method)
+ax.set_xlabel(r'$D$')
+ax.set_ylabel(r'accuracy')
+ax.set_xlim([10,700])
+ax.legend()
+fig.savefig(output, bbox_inches='tight')
+

### 2018-06-04 energy_clustering/kgroups_code/waveform.py

@@ -1,0 +1,76 @@
+"""Clustering using the waveform dataset:
+
+https://archive.ics.uci.edu/ml/datasets/Waveform+Database+Generator+%28Version+1%29
+
+"""
+
+# Guilhere Franca <guifranca@gmail.com>
+# Johns Hopkins University, Neurodata
+
+from __future__ import division
+
+import numpy as np
+import pandas as pd
+import sklearn
+
+from prettytable import PrettyTable
+
+import wrapper
+import eclust
+import metric
+
+import sys
+
+# some semmimetric functions
+rho = lambda x, y: np.power(np.linalg.norm(x-y), 1)
+rho1 = lambda x, y: np.linalg.norm(x-y)
+
+
+# get dermatology data, last column are labels
+df = pd.read_csv('data/waveform-+noise.data', sep=',', header=None)
+
+data = df.values[:,:-1]
+z = df.values[:,-1]  # 3 classes, starting with 0
+
+idx = np.random.choice(range(len(data)), 2000)
+data = data[idx]
+z = z[idx]
+data = (data - data.mean(axis=0))/data.std(axis=0)
+
+sigma2 = sum([np.linalg.norm(x-y)**2 
+                for x in data for y in data])/(len(data)**2)
+sigma = np.sqrt(sigma2)
+
+rho_exp = lambda x, y: 2-2*np.exp(-np.linalg.norm(x-y)/(2*sigma))
+rho_gauss = lambda x, y: 2-2*np.exp(-np.linalg.norm(x-y)**2/(2*(sigma)**2))
+
+# normalize data
+
+G = eclust.kernel_matrix(data, rho_gauss)
+#G = energy.eclust.kernel_matrix(data, rho_gauss)
+#G = energy.eclust.kernel_matrix(data, rho_exp)
+
+r = []
+r.append(wrapper.kmeans(3, data, run_times=5))
+r.append(wrapper.gmm(3, data, run_times=5))
+r.append(wrapper.spectral_clustering(3, data, G, run_times=5))
+r.append(wrapper.spectral(3, data, G, run_times=5))
+#r.append(wrapper.kernel_kmeans(3, data, G, run_times=5, ini='random'))
+#r.append(wrapper.kernel_kmeans(3, data, G, run_times=5, ini='k-means++'))
+r.append(wrapper.kernel_kmeans(3, data, G, run_times=5, ini='spectral'))
[…]

### 2018-06-04 energy_clustering/kgroups_code/wrapper.py

@@ -1,0 +1,138 @@
+"""Wrappers to run clustering algorithms."""
+
+# Guilhere Franca <guifranca@gmail.com>
+# Johns Hopkins University, Neurodata
+
+from __future__ import division
+
+import numpy as np
+from sklearn.cluster import KMeans
+from sklearn.mixture import GaussianMixture as GMM
+from sklearn.cluster import SpectralClustering
+
+import init
+import metric
+import eclust
+
+def initialize(method, k, G, X, W):
+    if method == "spectral":
+        z0 = init.topeigen(k, G, W)
+    elif method == "k-means++":
+        z0 = init.kmeans_plus(k, X)
+    else:
+        z0 = np.random.randint(0, k, len(X))
+    Z0 = eclust.ztoZ(z0)
+    return Z0
+
+def kernel_kmeans(k, X, G, W=None, run_times=5, ini="k-means++"):
+    if type(W) == type(None):
+        W = np.eye(len(X))
+    best_score = -np.inf
+    for _ in range(run_times):
+        Z0 = initialize(ini, k, G, X, W)
+        zh = eclust.kernel_kmeans(k, G, Z0, W, max_iter=300)
+        Zh = eclust.ztoZ(zh)
+        score = eclust.objective(Zh, G, W)
+        if score > best_score:
+            best_score = score
+            best_z = zh
+    return best_z
+
+def kernel_kgroups(k, X, G, W=None, run_times=5, ini="k-means++"):
+    if type(W) == type(None):
+        W = np.eye(len(X))
+    best_score = -np.inf
+    for _ in range(run_times):
+        Z0 = initialize(ini, k, G, X, W)
+        zh = eclust.kernel_kgroups(k, G, Z0, W, max_iter=300)
+        Zh = eclust.ztoZ(zh)
+        score = eclust.objective(Zh, G, W)
+        if score > best_score:
+            best_score = score
+            best_z = zh
+    return best_z
+
+def spectral(k, X, G, W=None, run_times=5):
+    if type(W) == type(None):
+        W = np.eye(len(X))
+    best_score = -np.inf
+    for _ in range(run_times):
+        zh = init.topeigen(k, G, W, run_times=run_times)
+        Zh = eclust.ztoZ(zh)
+        score = eclust.objective(Zh, G, W)
+        if score > best_score:
[…]

### 2018-06-04 energy_clustering/old_stuff/code/energy/eclust.py

@@ -1,0 +1,334 @@
+"""Iterative algorithms to optimize QCQP for energy statistics clustering."""
+
+# Guilherme Franca <guifranca@gmail.com>
+# Johns Hopkins University, Neurodata
+
+
+from __future__ import division
+
+import numpy as np
+from sklearn.metrics.pairwise import pairwise_distances
+from sklearn.base import BaseEstimator, ClusterMixin
+from sklearn.utils import check_random_state
+
+
+def ztoZ(z):
+    """Convert label vector to label matrix."""
+    n = z.shape[0]
+    k = np.unique(z).shape[0]
+    Z = np.zeros((n,k))
+    for i in range(n):
+        Z[i, int(z[i])] = 1
+    return Z
+
+def Ztoz(Z):
+    """Convert label matrix to label vector."""
+    n, k = Z.shape
+    z = np.zeros(n)
+    points, labels = np.where(Z==1)
+    for i in range(len(points)):
+        z[points[i]] = labels[i]
+    return z
+
+def objective(Z, G):
+    """Compute objective function."""
+    Zt = Z.T
+    D = np.dot(Zt, Z)
+    Dinv = np.linalg.inv(D)
+    return np.trace(Dinv.dot(Zt.dot(G.dot(Z))))
+
+def kernel_function(x, y, rho, x0=None):
+    """Return kernel function based on rho."""
+    if x0 == None:
+        x0 = np.zeros(x.shape)
+    val = 0.5*(rho(x,x0) + rho(y,x0) - rho(x,y))
+    return val
+
+def kernel_matrix(X, rho, x0=None):
+    """Compute Kernel matrix based on kernel function K(x,y)."""
+    kfunc = lambda x, y: kernel_function(x, y, rho, x0)
+    #G = pairwise_distances(X, metric=kfunc, n_jobs=4)
+    G = pairwise_distances(X, metric=kfunc)
+    return G
+
+def energy_clustering_brute(k, G, Z0, max_iter=100, tol=1e-4, verbose=False):
+    """Optimize energy QCQP in a brute force way. This is expensive,
+    but it works. We move points to other partitions and compute the change
+    in the cost function.
+
+    """
+    
+    n = G.shape[0]
+    Z = np.copy(Z0)
+    count = 0
+    converged = False
+    while not converged and count < max_iter:
+        
+        n_changed = 0
+        
+        for l in range(n):
+            
+            curr_label = np.where(Z[l]==1)[0][0]
+            Ws = []
+            
[…]

### 2018-06-04 energy_clustering/old_stuff/code2/dermatology.py

@@ -1,0 +1,70 @@
+"""Clustering using the dermatology dataset:
+
+https://archive.ics.uci.edu/ml/datasets/dermatology
+
+"""
+
+# Guilhere Franca <guifranca@gmail.com>
+# Johns Hopkins University, Neurodata
+
+from __future__ import division
+
+import numpy as np
+import pandas as pd
+import sklearn
+
+from prettytable import PrettyTable
+
+import run_clustering as cluster
+import energy
+
+
+# some semmimetric functions
+rho = lambda x, y: np.power(np.linalg.norm(x-y), 0.5)
+rho_gauss = lambda x, y: 2-2*np.exp(-np.linalg.norm(x-y)**2/(2*(3.5)**2))
+rho_exp = lambda x, y: 2-2*np.exp(-np.linalg.norm(x-y)/(2*3.5))
+
+
+# get dermatology data, last column are labels
+df = pd.read_csv('./data/dermatology.data', sep=',', header=None)
+data = df.values[:,:-1]
+true_labels = df.values[:,-1] - 1 # 6 classes, starting with 0
+
+# delete missing entries
+delete_missing = np.where(data=='?')[0]
+data = np.delete(data, delete_missing, axis=0)
+data = np.array(data, dtype=float)
+true_labels = np.delete(true_labels, delete_missing, axis=0)
+
+# normalize data
+data = (data - data.mean(axis=0))/data.std(axis=0)
+
+G = energy.eclust.kernel_matrix(data, rho)
+#G = energy.eclust.kernel_matrix(data, rho_gauss)
+#G = energy.eclust.kernel_matrix(data, rho_exp)
+
+kmeans_labels = cluster.kmeans(6, data, run_times=10, init="k-means++")
+gmm_labels = cluster.gmm(6, data, run_times=10, init="kmeans")
+spectral_labels = cluster.spectral(6, data, G, run_times=10)
+energy_spectral_labels = cluster.energy_spectral(6, data, G, run_times=10)
+lloyd_labels = cluster.energy_lloyd(6, data, G, run_times=10, init="spectral")
+hart_labels = cluster.energy_hartigan(6,data,G,run_times=10,init="spectral")
+
+t = PrettyTable(['Algorithm', 'Accuracy', 'A-Rand', 'Mutual Info', 'V-Measure', 
+                'Fowlkes-Mallows'])
+
+algos = ['kmeans', 'GMM', 'spectral', 'energy_spectral', 'energy_lloyd',
+         'energy_hartigan']
[…]

### 2018-06-04 energy_clustering/old_stuff/code2/energy/bayes.py

@@ -1,0 +1,214 @@
+"""Functions to compute Bayes errors."""
+
+# Guilherme S. Franca <guifranca@gmail.com>
+# Johns Hopkins, Neurodata
+
+
+from __future__ import division
+
+from functools import partial
+from scipy.optimize import minimize
+from scipy.optimize import root
+import numpy as np
+import scipy.stats as stats
+import scipy.integrate as integrate
+import matplotlib.pyplot as plt
+
+import data
+
+
+def bayes_univariate_normal(mu1, mu2, sigma1, sigma2, 
+                            num_times=10, num_points=1000, p=0.5):
+    """Estimate Bayes' error for two-class normal distributions in 1D."""
+    error = np.zeros(num_times)
+    for i in range(num_times):
+        q = 1-p
+        n1, n2 = np.random.multinomial(num_points, [p,q])
+        x1 = np.random.normal(mu1, sigma1, n1)
+        x2 = np.random.normal(mu2, sigma2, n2)
+    
+        error1 = sum([1 for x in x1 
+            if stats.norm.pdf(x,mu1,sigma1)-stats.norm.pdf(x,mu2,sigma2)<=0
+        ])
+        error2 = sum([1 for x in x2 
+            if stats.norm.pdf(x,mu1,sigma1)-stats.norm.pdf(x,mu2,sigma2)>=0
+        ])
+        error[i] = p*error1/n1+q*error2/n2
+    return 1-error.mean()
+
+def integrate_univariate_normal(mu1, mu2, sigma1, sigma2, p=0.5):
+    """Same as above but through numerical integration."""
+    
+    g1 = partial(stats.norm.pdf, loc=mu1, scale=sigma1)
+    g2 = partial(stats.norm.pdf, loc=mu2, scale=sigma2)
+    
+    # make it visual to confirm
+    #xs = np.arange(-10,20, 0.001)
+    #y1s = np.array([g1(x) for x in xs])
+    #y2s = np.array([g2(x) for x in xs])
+    #y3s = np.array([np.abs(g1(x)-g2(x)) for x in xs])
+    #plt.plot(xs, y1s)
+    #plt.plot(xs, y2s)
+    #plt.plot(xs, y3s)
+    #plt.show()
+
+    x0 = root(lambda x: g1(x) - g2(x), 1).x[0]
+    
+    # another way
+    #cons = (
[…]

### 2018-06-04 energy_clustering/old_stuff/code2/energy/eclust.py

@@ -1,0 +1,335 @@
+"""Iterative algorithms to optimize QCQP for energy statistics clustering."""
+
+# Guilherme Franca <guifranca@gmail.com>
+# Johns Hopkins University, Neurodata
+
+
+from __future__ import division
+
+import numpy as np
+from sklearn.metrics.pairwise import pairwise_distances
+from sklearn.base import BaseEstimator, ClusterMixin
+from sklearn.utils import check_random_state
+
+
+def ztoZ(z):
+    """Convert label vector to label matrix."""
+    n = z.shape[0]
+    k = np.unique(z).shape[0]
+    Z = np.zeros((n,k))
+    for i in range(n):
+        Z[i, int(z[i])] = 1
+    return Z
+
+def Ztoz(Z):
+    """Convert label matrix to label vector."""
+    n, k = Z.shape
+    z = np.zeros(n)
+    points, labels = np.where(Z==1)
+    for i in range(len(points)):
+        z[points[i]] = labels[i]
+    return z
+
+def objective(Z, G):
+    """Compute objective function."""
+    Zt = Z.T
+    D = np.dot(Zt, Z)
+    Dinv = np.linalg.inv(D)
+    return np.trace(Dinv.dot(Zt.dot(G.dot(Z))))
+
+def kernel_function(x, y, rho, x0=None):
+    """Return kernel function based on rho."""
+    if x0 == None:
+        x0 = np.zeros(x.shape)
+    val = 0.5*(rho(x,x0) + rho(y,x0) - rho(x,y))
+    return val
+
+def kernel_matrix(X, rho, x0=None):
+    """Compute Kernel matrix based on kernel function K(x,y)."""
+    kfunc = lambda x, y: kernel_function(x, y, rho, x0)
+    #G = pairwise_distances(X, metric=kfunc, n_jobs=4)
+    G = pairwise_distances(X, metric=kfunc)
+    return G
+
+def energy_clustering_brute(k, G, Z0, max_iter=100, tol=1e-4, verbose=False):
+    """Optimize energy QCQP in a brute force way. This is expensive,
+    but it works. We move points to other partitions and compute the change
+    in the cost function.
+
+    """
+    
+    n = G.shape[0]
+    Z = np.copy(Z0)
+    count = 0
+    converged = False
+    while not converged and count < max_iter:
+        
+        n_changed = 0
+        
+        for l in range(n):
+            
+            curr_label = np.where(Z[l]==1)[0][0]
+            Ws = []
+            
[…]

### 2018-06-04 energy_clustering/old_stuff/code2/energy/metric.py

@@ -1,0 +1,43 @@
+"""Metric functions for clustering."""
+
+# Guilherme Franca <guifranca@gmail.com>
+# Johns Hopkins University, Neurodata
+
+from __future__ import division
+
+import numpy as np
+import scipy.optimize
+from sklearn.metrics.cluster import mutual_info_score
+from sklearn.metrics.cluster import entropy
+
+
+def accuracy(z, zh):
+    """Compute misclassification error, or better the accuracy which is
+    1 - error. Use Hungarian algorithm which is O(n^3) instead
+    of O(n!). z and zh are vectors with the dimension
+    being the number of points, and each entry is the cluster label assigned
+    to that point.
+
+    """
+    n, k = len(z), len(np.unique(z))
+    m, q = len(zh), len(np.unique(zh))
+
+    assert m == n
+    
+    Q = np.zeros((n,k))
+    for i in range(n):
+        Q[i, int(z[i])] = 1
+    Qh = np.zeros((n,k))
+    for i in range(m):
+        Qh[i, int(zh[i])] = 1
+    
+    cost_matrix = Qh.T.dot(Q)
+    row_ind, col_ind = scipy.optimize.linear_sum_assignment(-cost_matrix)
+    
+    return cost_matrix[row_ind, col_ind].sum()/n
+
+def variation_information(z, zh):
+    """Compute variation of information based on M. Meila (2007)."""
+    return entropy(z) + entropy(zh) - 2*mutual_info_score(z, zh)
+
+

### 2018-06-04 energy_clustering/old_stuff/code2/normal_umbalanced.py

@@ -1,0 +1,108 @@
+"""Third experiment. Normal distributions for unbalanced clusters.
+Using mixtures now.
+
+"""
+
+# Guilhere Franca <guifranca@gmail.com>
+# Johns Hopkins University, Neurodata
+
+from __future__ import division
+
+import numpy as np
+import multiprocessing as mp
+
+import energy.data as data
+import energy.eclust as eclust
+import energy.initialization as initialization
+import run_clustering
+import plot
+
+def gauss_dimensions_pi(num_points=range(0, 180, 10), num_experiments=10):
+    """Test unbalanced clusters."""
+    k = 2
+    D = 4
+    d = 2
+    N = 300
+    table = np.zeros((num_experiments*len(num_points), 6))
+    count = 0
+
+    for p in num_points:
+        for i in range(num_experiments):
+
+            # generate data
+            m1 = np.zeros(D)
+            s1 = np.eye(D)
+            m2 = np.concatenate((1.5*np.ones(d), np.zeros(D-d)))
+            s2 = np.diag(np.concatenate((.5*np.ones(d), np.ones(D-d))))
+            pi1 = (N-p)/N/2.
+            pi2 = (N+p)/N/2.
+            n1, n2 = np.random.multinomial(N, [pi1,pi2])
+            X, z = data.multivariate_normal([m1, m2], [s1, s2], [n1, n2])
+            rho = lambda x, y: np.linalg.norm(x-y)
+            G = eclust.kernel_matrix(X, rho)
+
+            table[count, 0] = p
+            table[count, 1] = run_clustering.energy_hartigan(k, X, G, z,
+                                                init="k-means++", run_times=5)
+            table[count, 2] = run_clustering.energy_lloyd(k, X, G, z,
+                                                init="k-means++", run_times=5)
+            table[count, 3] = run_clustering.spectral(k, X, G, z,
+                                                run_times=5)
+            table[count, 4] = run_clustering.kmeans(k, X, z,
+                                                init="k-means++", run_times=5)
+            table[count, 5] = run_clustering.gmm(k, X, z,
+                                                init="kmeans", run_times=5)
+
+            count += 1
+
+    return table
+
+def make_plot(*data_files):
+    table = []
+    for f in data_files:
+        t = np.loadtxt(f, delimiter=',')
+        t[:,0] = np.array([int(D) for D in t[:,0]])
+        table.append(t)
[…]

### 2018-06-04 energy_clustering/old_stuff/code3/oldstuff/normal_high_dim.py

@@ -1,0 +1,171 @@
+"""Second experiment. Normal distributions in high dimensions.
+The difference here is only that we use bernoulli coins to get
+number of points for each cluster, and also we are using spectral
+clustering from scikit-learn.
+
+"""
+
+# Guilhere Franca <guifranca@gmail.com>
+# Johns Hopkins University, Neurodata
+
+import numpy as np
+import multiprocessing as mp
+
+import energy.data as data
+import energy.eclust as eclust
+import energy.initialization as initialization
+import run_clustering
+import plot
+
+def gauss_dimensions_mean(dimensions=range(2,100,20), total_points=200,
+                          num_experiments=100, d=None):
+    # data distribution
+    k = 2
+    delta = 0.7
+    if not d:
+        d = dimensions[0]
+    table = np.zeros((num_experiments*len(dimensions), 6))
+    count = 0
+
+    for D in dimensions:
+        for i in range(num_experiments):
+
+            ### generate data
+            m1 = np.zeros(D)
+            s1 = np.eye(D)
+            m2 = np.concatenate((delta*np.ones(d), np.zeros(D-d)))
+            s2 = np.eye(D)
+            # flip Bernoulli coins to get number of points in each cluster
+            n1, n2 = np.random.multinomial(total_points, [0.5,0.5])
+            # get data, construct gram Matrix
+            X, z = data.multivariate_normal([m1, m2], [s1, s2], [n1, n2])
+            rho = lambda x, y: np.linalg.norm(x-y)
+            G = eclust.kernel_matrix(X, rho)
+            ##################
+
+            ### cluster with different algorithms
+            # can change number of times we execute each experiment
+            # and initialization as well
+            table[count, 0] = D
+            table[count, 1] = run_clustering.energy_hartigan(k, X, G, z, 
+                                                init="k-means++", run_times=5)
+            table[count, 2] = run_clustering.energy_lloyd(k, X, G, z, 
+                                                init="k-means++", run_times=5)
+            table[count, 3] = run_clustering.spectral(k, X, G, z, 
+                                                run_times=5)
+            table[count, 4] = run_clustering.kmeans(k, X, z, 
+                                                init="k-means++", run_times=5)
+            table[count, 5] = run_clustering.gmm(k, X, z, 
[…]

### 2018-06-04 energy_clustering/old_stuff/code3/oldstuff/normal_lognormal_highd.py

@@ -1,0 +1,200 @@
+"""Fourth experiment. Normal and lognormal distributions in high dimensions,
+using different kernels.
+Using spectral clustering from sklearn and mixtures.
+
+"""
+
+# Guilhere Franca <guifranca@gmail.com>
+# Johns Hopkins University, Neurodata
+
+import numpy as np
+import multiprocessing as mp
+
+import energy.data as data
+import energy.eclust as eclust
+import energy.initialization as initialization
+import run_clustering
+import plot
+
+def normal_or_lognormal(numpoints=range(10,100,10), num_experiments=100,
+                        kind='normal'):
+    table = np.zeros((num_experiments*len(numpoints), 6))
+    count = 0
+    k = 2
+
+    for n in numpoints:
+        for i in range(num_experiments):
+
+            # generate data
+            D = 20
+            d = 5
+            m1 = np.zeros(D)
+            s1 = 0.5*np.eye(D)
+            m2 = 0.5*np.concatenate((np.ones(d), np.zeros(D-d)))
+            s2 = np.eye(D)
+            n1, n2 = np.random.multinomial(n, [0.5,0.5])
+
+            if kind == 'normal':
+                X, z = data.multivariate_normal([m1,m2], [s1,s2], [n1,n2])
+            else:
+                X, z = data.multivariate_lognormal([m1,m2], [s1,s2], [n1,n2])
+
+            rho = lambda x, y: np.linalg.norm(x-y)
+            G = eclust.kernel_matrix(X, rho)
+
+            rho2 = lambda x, y: np.power(np.linalg.norm(x-y), 0.5)
+            G2 = eclust.kernel_matrix(X, rho2)
+
+            rho3 = lambda x, y: 2-2*np.exp(-np.linalg.norm(x-y)/2)
+            G3 = eclust.kernel_matrix(X, rho3)
+
+            table[count, 0] = n
+            table[count, 1] = run_clustering.energy_hartigan(k, X, G, z, 
+                                                init="k-means++", run_times=5)
+            table[count, 2] = run_clustering.energy_hartigan(k, X, G2, z, 
+                                                init="k-means++", run_times=5)
+            table[count, 3] = run_clustering.energy_hartigan(k, X, G3, z, 
+                                                init="k-means++", run_times=5)
+            table[count, 4] = run_clustering.kmeans(k, X, z, 
[…]

### 2018-06-04 energy_clustering/old_stuff/code3/oldstuff/one_dimensional.py

@@ -1,0 +1,126 @@
+"""First experiment. 1D data."""
+
+# Guilhere Franca <guifranca@gmail.com>
+# Johns Hopkins University, Neurodata
+
+import numpy as np
+import multiprocessing as mp
+
+import energy.data as data
+import energy.eclust as eclust
+import energy.initialization as initialization
+import run_clustering
+import plot
+
+def sample_normal(m1, m2, s1, s2, n1, n2):
+    return data.univariate_normal([m1, m2], [s1, s2], [n1, n2])
+
+def sample_lognormal(m1, m2, s1, s2, n1, n2):
+    return data.univariate_lognormal([m1, m2], [s1, s2], [n1, n2])
+
+
+class Clustering1D:
+
+    def __init__(self):
+        # parameters for distributions
+        self.m1 = 1.5
+        self.s1 = 0.3
+        self.m2 = 0
+        self.s2 = 1.5 
+        self.distr = 'normal' # should be normal or lognormal
+        self.numpoints = range(50, 100, 20)
+        self.experiments = 100 # number of experiments
+
+    def get_sample(self, n):
+        n1, n2 = np.random.multinomial(n, [0.5, 0.5])
+        if self.distr == 'normal':
+            X, z = sample_normal(self.m1, self.m2, self.s1, self.s2, n1, n2)
+        elif self.distr == 'lognormal':
+            X, z = sample_lognormal(self.m1, self.m2, self.s1, self.s2, n1, n2)
+        else:
+            raise ValueError("Clustering 1D, unknown distribution to sample.")
+        Y = np.array([[x] for x in X])
+        return Y, z
+
+    def run(self):
+        k = 2
+        ncols = 3 + 1 # number of clustering methods + 1
+        table = np.zeros((self.experiments*len(self.numpoints), ncols))
+        count = 0
+        init = 'k-means++'
+        for i, n in enumerate(self.numpoints):
+            for ne in range(self.experiments):
+                Y, z = self.get_sample(n)
+                G = eclust.kernel_matrix(Y, lambda x, y: np.linalg.norm(x-y))
+                table[count, 0] = n
+                table[count, 1] = run_clustering.energy_hartigan(k, Y, G, z,
+                                                    init=init, run_times=5)
+                table[count, 2] = run_clustering.kmeans(k, Y, z, init=init,
+                                                    run_times=5)
[…]

### 2018-06-04 energy_clustering/old_stuff/code3/oldstuff/plot.py

@@ -1,0 +1,88 @@
+"""Plot accuracy results."""
+
+# Guilhere Franca <guifranca@gmail.com>
+# Johns Hopkins University, Neurodata
+
+from __future__ import division
+
+import numpy as np
+import scipy.stats
+import matplotlib as mpl
+import matplotlib.pyplot as plt
+
+from customize_plots import *
+
+
+class ErrorBar:
+
+    def __init__(self):
+        self.xlabel = 'dimension'
+        self.ylabel = 'accuracy'
+        self.symbols = ['o', 's', 'D', 'v']
+        self.legends = ['energy', r'$k$-means', 'GMM', r'kernel $k$-means']
+        self.colors = ['b', 'r', 'g', 'c']
+        self.lines = ['-', '-', '-', '-', '-', '-']
+        self.xlim = None
+        self.ylim = None
+        self.bayes = None
+        self.loc = 0
+        self.doublex = False
+        self.legcols = 1
+        self.output = 'plot.pdf'
+
+    def _set_params(self, table):
+        self.col0 = col0 = np.array([int(x) for x in table[:,0]])
+        self.xs = np.unique(col0)
+        
+        self.n = len(table[0][1:])
+        #self.colors = iter(plt.cm.brg(np.linspace(0,1,self.n+1)))
+        self.colors = iter(self.colors)
+        self.legends = iter(self.legends)
+        self.symbols = iter(self.symbols)
+        self.lines = iter(self.lines)
+    
+        if self.doublex:
+            self.xs2 = 2*self.xs
+        else:
+            self.xs2 = self.xs
+            
+        if self.bayes is not None:
+            if not isinstance(self.bayes, list):
+                self.bayes = [self.bayes]*len(self.xs)
+
+    def make_plot(self, table):
+        self._set_params(table)
+
+        fig = plt.figure()
+        ax = fig.add_subplot(111)
+    
+        for i in range(1, self.n+1):
+            
+            c = next(self.colors)
+            l = next(self.legends)
+            m = next(self.symbols)
+            t = next(self.lines)
+
+            mean_ = np.array([table[np.where(self.col0==x)[0],i].mean() 
+                              for x in self.xs])
+            stderr_ = np.array([scipy.stats.sem(
+                table[np.where(self.col0==x)[0],i]) for x in self.xs])
+            if self.bayes:
+                ax.plot(self.xs2, self.bayes, '--', color='black', 
[…]

### 2018-06-04 energy_clustering/old_stuff/code3/oldstuff/print_digits.py

@@ -1,0 +1,42 @@
+"""Make a figure of some digits from MNIST."""
+
+import matplotlib.pyplot as plt
+import cPickle, gzip
+import matplotlib.pyplot as plt
+import numpy as np
+
+from customize_plots import *
+
+
+f = gzip.open('data/mnist.pkl.gz', 'rb')
+train_set, valid_set, test_set = cPickle.load(f)
+f.close()
+images, labels = train_set
+
+X = []
+for d in range(0,10):
+    x = np.where(labels==d)[0]
+    js = np.random.choice(x, 2, replace=False)
+    for j in js:
+        im = np.array(images[j])
+        X.append(im)
+X = np.array(X)
+idx = range(20)
+np.random.shuffle(idx)
+X = X[idx]
+fig, axes = plt.subplots(nrows=5, ncols=4,
+                        #figsize=(3.85,4.69))
+                        figsize=(4,5))
+i = 0
+for row in axes:
+    for ax in row:
+        ax.imshow(X[i].reshape((28,28)), cmap=plt.cm.gray)
+        ax.axis('off')
+        ax.set_aspect('equal')
+        #ax.set_xticklabels([])
+        #ax.set_yticklabels([])
+        i += 1
+fig.subplots_adjust(wspace=-0.1, hspace=-0.1)
+#fig.tight_layout()
+fig.savefig('figs/mnist.pdf')
+

### 2018-06-16 energy_clustering/pami18/energy_clustering.tex
pami

@@
 
-\begin{algorithm}
+\begin{algorithm}[t]
 \begin{algorithmic}[1]
@@
 
-We now consider Hartigan's method \cite{Hartigan1975,Hartigan1979} 
-applied to the optimization problem in the form \eqref{eq:maxQ}, which gives
-a local solution to \eqref{eq:qcqp3}. 
-The method is based in computing the maximum change
-in the total cost function $Q$ when moving each data point to
-another cluster. More specifically, 
-suppose that point $x_i$
-is currently assigned to  cluster $\C_j$ yielding
-a total cost function denoted by $Q^{(j)}$.
-Moving $x_i$ to cluster $\C_\ell$ yields another total cost function
-denoted by $Q^{(\ell)}$. We are interested in computing the maximum 
-change
-$\Delta Q^{(j\to \ell)} (x_i) \equiv Q^{(\ell)} - Q^{(j)}$, for $\ell\ne j$. 
-From \eqref{eq:maxQ}, by explicitly writing the costs related to these 
-two cluster we obtain
-\begin{equation}
-\label{eq:deltaQ}
-\Delta Q^{(j\to \ell)} (x_i) = \dfrac{Q_\ell^{+}}{s_\ell+w_i} + 
-\dfrac{Q_j^-}{s_j-w_i} - \dfrac{Q_\ell}{s_\ell} - \dfrac{Q_j}{s_j}  ,
-\end{equation}
-where $Q^{+}_\ell$ denote the cost of the new $\ell$th cluster
-with the point $x_i$ added to it, and $Q^-_j$ is the cost of new 
-$j$th cluster with $x_i$ removed from it. Recall also that $w_i=w(x_i)$
-is the weight associated to point $x_i$. Noting that 
-\begin{align}
-Q_\ell^{+} &= Q_\ell + 2 Q_\ell(x_i) + (\W G \W)_{ii}, 
-\label{eq:Qplus} \\
-Q_j^{-} &= Q_j - 2 Q_j(x_i) + (\W G \W)_{ii}, 
-\label{eq:Qminus}
-\end{align}
-we obtain
-\begin{multline}
-\label{eq:changeQ}
-\Delta Q^{(j \to \ell)}(x_i)  = 
-\dfrac{1}{s_j - w_i}  \left[ \dfrac{w_i}{s_j}Q_j 
-- 2 Q_j(x_i) + (\W G \W)_{ii} \right] \\
-- \dfrac{1}{s_\ell + w_i}\left[ \dfrac{w_i}{s_\ell}Q_\ell - 2 Q_\ell(x_i) 
-- (\W G \W )_{ii} \right].
-\end{multline}
-Therefore, we compute
-\begin{equation}
-j^\star = \argmax_{\ell=1,\dotsc,k \, | \, \ell\ne j} 
-\Delta Q^{(j \to \ell)}(x_i)
-\end{equation}
-and if $\Delta Q^{j \to j^\star}(x_i) > 0$ 
-we move $x_i$ to cluster $\C_{j^\star}$, otherwise 
-we keep $x_i$ in its original cluster $\C_j$. 
-This process is repeated
-until no points are assigned to new clusters. 
-The entire procedure is described in Algorithm~\ref{algo}, which we
-call kernel k-groups. This method is a generalization of the k-groups
-with first variations proposed in \cite{Kgroups}, which only considers
-the Euclidean case, and also an extension
-of Hartigan's method to kernel spaces.
-
-\begin{algorithm}
+\begin{algorithm}[t]
 \begin{algorithmic}[1]
@@
 \end{algorithm}
+
+We now consider Hartigan's method \cite{Hartigan1975,Hartigan1979} 
+applied to the optimization problem in the form \eqref{eq:maxQ}, which gives
+a local solution to \eqref{eq:qcqp3}. 
+The method is based in computing the maximum change
+in the total cost function $Q$ when moving each data point to
+another cluster. More specifically, 
+suppose that point $x_i$
+is currently assigned to  cluster $\C_j$ yielding
+a total cost function denoted by $Q^{(j)}$.
+Moving $x_i$ to cluster $\C_\ell$ yields another total cost function
+denoted by $Q^{(\ell)}$. We are interested in computing the maximum 
+change
+$\Delta Q^{(j\to \ell)} (x_i) \equiv Q^{(\ell)} - Q^{(j)}$, for $\ell\ne j$. 
+From \eqref{eq:maxQ}, by explicitly writing the costs related to these 
+two cluster we obtain
+\begin{equation}
+\label{eq:deltaQ}
+\Delta Q^{(j\to \ell)} (x_i) = \dfrac{Q_\ell^{+}}{s_\ell+w_i} + 
+\dfrac{Q_j^-}{s_j-w_i} - \dfrac{Q_\ell}{s_\ell} - \dfrac{Q_j}{s_j}  ,
+\end{equation}
+where $Q^{+}_\ell$ denote the cost of the new $\ell$th cluster
+with the point $x_i$ added to it, and $Q^-_j$ is the cost of new 
+$j$th cluster with $x_i$ removed from it. Recall also that $w_i=w(x_i)$
+is the weight associated to point $x_i$. Noting that 
+\begin{align}
+Q_\ell^{+} &= Q_\ell + 2 Q_\ell(x_i) + (\W G \W)_{ii}, 
+\label{eq:Qplus} \\
+Q_j^{-} &= Q_j - 2 Q_j(x_i) + (\W G \W)_{ii}, 
+\label{eq:Qminus}
+\end{align}
+we obtain
+\begin{multline}
+\label{eq:changeQ}
+\Delta Q^{(j \to \ell)}(x_i)  = 
+\dfrac{1}{s_j - w_i}  \left[ \dfrac{w_i}{s_j}Q_j 
+- 2 Q_j(x_i) + (\W G \W)_{ii} \right] \\
+- \dfrac{1}{s_\ell + w_i}\left[ \dfrac{w_i}{s_\ell}Q_\ell - 2 Q_\ell(x_i) 
+- (\W G \W )_{ii} \right].
+\end{multline}
+Therefore, we compute
+\begin{equation}
+j^\star = \argmax_{\ell=1,\dotsc,k \, | \, \ell\ne j} 
+\Delta Q^{(j \to \ell)}(x_i)
+\end{equation}
+and if $\Delta Q^{j \to j^\star}(x_i) > 0$ 
+we move $x_i$ to cluster $\C_{j^\star}$, otherwise 
+we keep $x_i$ in its original cluster $\C_j$. 
+This process is repeated
+until no points are assigned to new clusters. 
+The entire procedure is described in Algorithm~\ref{algo}, which we
+call kernel k-groups. This method is a generalization of the k-groups
+with first variations proposed in \cite{Kgroups}, which only considers
+the Euclidean case, and also an extension
+of Hartigan's method to kernel spaces.
 
@@
 
-%In the first two experiments 
-%we use $\mathcal{E}^H$-clustering with the standard metric from energy
-%statistics, i.e. \eqref{eq:rho_alpha} with \mbox{$\alpha=1$}. 
-%We consider univariate
-%normal and lognormal mixtures according to
-%\begin{align}
-%x &\stackrel{iid}{\sim} \tfrac{1}{2}
-%\mathcal{N}(\mu_1,\sigma_1^2) + 
-%\tfrac{1}{2} \mathcal{N}(\mu_2,\sigma_2^2) ,\label{eq:uni_normal} \\
-%x & \stackrel{iid}{\sim} \tfrac{1}{2}
-%\exp\left\{ \mathcal{N}(\mu_1,\sigma_1^2) \right\} + 
-%\tfrac{1}{2} \exp \left\{ \mathcal{N}(\mu_2,\sigma_2^2)\right\} ,
-%\label{eq:uni_lognormal} \\
-%\mu_1 &= 1.5, \ \sigma_1=0.3, \ \mu_2=0, \ \sigma_2=1.5.
-%\label{eq:uni_params}
-%\end{align}
-%We increase the sample size $n$ and show the clustering accuracy 
-%\eqref{eq:accuracy}
-%versus $n$ in Fig.~\ref{fig:1D}. 
-%In both cases the optimal accuracy 
-%obtained from Bayes classification error
-%is $ \approx 0.88$, indicated by the dashed line.
-%Note that $\mathcal{E}^H$-clustering 
-%performs better than k-means for normally distributed data, 
-%and worse than GMM,
-%as expected. 
-%However, for lognormal distributions k-means and GMM are only slightly
-%better than chance while $\mathcal{E}^H$-clustering is still 
-%accurate. The reason is that k-means and GMM models are inconsistent
-%with this lognormal data. Energy statistics on the other hand
-%is model-free. We also show 
-%a density estimation after clustering $1000$ points from these
-%mixtures.
-%Note that only $\mathcal{E}^{H}$-clustering
-%was able to distinguish between the two classes, while 
-%k-means and GMM put most of the
-%points in a single cluster and points on the tail of the second component
-%of \eqref{eq:uni_lognormal} in the other cluster.
-%
-%
-%\begin{figure}
-%\begin{minipage}{0.49\linewidth}
-%\begin{center}
-%\includegraphics[width=\textwidth]{1D_normal.pdf}\\
-%~~~~~\includegraphics[width=.9\textwidth]{normal_density.pdf}
-%\end{center}
-%\end{minipage}
-%\begin{minipage}{0.49\linewidth}
-%\begin{center}
-%\includegraphics[width=\textwidth]{1D_lognormal.pdf}\\
-%~~~~~\includegraphics[width=.9\textwidth]{lognormal_density.pdf}
-%\end{center}
-%\end{minipage}
-%%\put(-140,5){\scriptsize{(a)}}
-%%\put(-140,-78){\scriptsize{(c)}}
-%%\put(-10,5){\scriptsize{(b)}}
-%%\put(-10,-78){\scriptsize{(d)}}
-%\caption{
-%\label{fig:1D}
-%\emph{Top Left:} Clustering data from \eqref{eq:uni_normal}.
-%\emph{Top Right:} Clustering data from \eqref{eq:uni_lognormal}.
-%The dashed lines show the optimal Bayes accuracy, which in both cases 
-%are $\approx 0.88$.
-%\emph{Bottom Left:} 
-%Density estimation of each component in the mixture \eqref{eq:uni_normal} 
-%after clustering $1000$ sampled points.
-%\emph{Bottom Right:} 
-% The same but for lognormal data \eqref{eq:uni_lognormal}.
-%}
-%\end{figure}
-
-First, we analyze
+\begin{figure*}
+\centering
+\begin{minipage}{.4\textwidth}
+\includegraphics[width=1\textwidth]{gauss_1d.pdf}\\[-2em]
+(a)
+\end{minipage}
+\begin{minipage}{.4\textwidth}
+\includegraphics[width=1\textwidth]{loggauss_1d.pdf} \\[-2em]
+(b)
+\end{minipage}\\[1em]
+\begin{minipage}{.37\textwidth}
+\includegraphics[width=1\textwidth]{normal_density.pdf}\\[-2em]
+(c)
+\end{minipage}\hspace{2em}
+\begin{minipage}{.37\textwidth}
+\includegraphics[width=1\textwidth]{lognormal_density.pdf}\\[-2em]
+(d)
+\end{minipage}
+\caption{
+\label{fig:1dexperiment}
+Clustering one-dimensional data for a two-class problem.
+(a) $x \stackrel{iid}{\sim} \tfrac{1}{2} \mathcal{N}(0, 1.5) +
+\tfrac{1}{2}\mathcal{N}(1.5,0.3)$. (b) $x \stackrel{iid}{\sim}
+\tfrac{1}{2}e^{\mathcal{N}(0, 1.5)} + \tfrac{1}{2} e^{\mathcal{N}(1.5, 0.3)}$.
+In both cases we plot the accuracy versus the total number of points.
+(c) Kernel density estimation for a single trial with 2000 sampled
+points from the
+distribution in the case (a). (d) Kernel density estimation in a single
+trial with 
+2000 points sampled from the distribution in the case (b).
+}
+\end{figure*}
+
+We first consider one-dimensional data for a two-class problem. We compare
+kernel k-groups with k-means and GMM, as illustrated in 
+Fig.~\ref{fig:1dexperiment}. We have two mixtures of normal and 
+lognormal distributions. Notice that in the kernel density estimation plots
+for lognormal distribution, only kernel k-groups was able to distinguish
+between the two classes. The accuracy results for both density estimation
+cases are in Table~\ref{tb:dens}. We remark that kernel k-groups in
+this one-dimensional example performed the same as the deterministic
+Algorithm~\ref{algo1d} introduced in the Appendix.
+
+\begin{table}[h]
+\centering
+\caption{
+\label{tb:dens} Accuracy results for the density
+estimation of Fig.~\ref{fig:1dexperiment}d--e.
+}
+\begin{tabular}{@{}l|ll@{}}
+       Method        & normal & lognormal \\ \midrule[.5pt]
+        kmeans       & 0.773  & 0.502     \\
+         GMM         & 0.881  & 0.513     \\
+   kernel k-groups   & 0.802  & 0.847  
+\end{tabular}
+\end{table}
+
+Next, we analyze
 how the algorithms degrade as the number of dimensions increase.
@@
 of $\approx 0.86$.
-We sample $200$ points on each trial.
-The results are shown in Fig.~\ref{fig:gauss1}.
+We sample $200$ points on each trial. A scatter plot of the last two
+dimensions that contains signal in $\mu_2$ is shown in
+Fig.~\ref{fig:scatter}a.
+The clustering results are shown in Fig.~\ref{fig:plots}a.
 We see that 
@@
 
-\begin{figure}
+\begin{figure*}
 \centering
-\includegraphics[scale=0.5]{gauss1.pdf}
+\includegraphics[width=0.19\textwidth]{gauss_means_scatter.pdf}
+\includegraphics[width=0.19\textwidth]{gauss_cov_scatter.pdf}
+\includegraphics[width=0.19\textwidth]{gauss_diff_metrics.pdf}
+\includegraphics[width=0.19\textwidth]{loggauss_diff_metrics.pdf}
+\includegraphics[width=0.19\textwidth]{unbalanced_scatter.pdf}
+\put(-500,-8){(a)}
+\put(-398,-8){(b)}
+\put(-298,-8){(c)}
+\put(-195,-8){(d)}
+\put(-95,-8){(e)}
 \caption{
-\label{fig:gauss1}
+\label{fig:scatter}
+Scatter plot of the last two dimensions where $\mu_2$ has signal.
+Each plot has 200 points total.
+(a) Data distributed as in \eqref{eq:gauss1}.
+(b) Data distributed as in \eqref{eq:gauss2}.
+(c) Data distributed as in \eqref{eq:20gauss}.
+(d) Parameters as in \eqref{eq:20gauss} but for lognormal mixture.
+(e) Data from \eqref{eq:gauss3} with $N = 300$ and $m=200$.
+}
+\end{figure*}
+
+\begin{figure*}
+\centering
+\includegraphics[width=0.33\textwidth]{gauss1.pdf}
+\includegraphics[width=0.33\textwidth]{gauss2.pdf}
+\includegraphics[width=0.33\textwidth]{gauss_n.pdf}
+\includegraphics[width=0.33\textwidth]{loggauss_n.pdf}
+\includegraphics[width=0.315\textwidth]{gauss_unbal.pdf}
+\put(-400,130){(a)}
+\put(-230,130){(b)}
+\put(-50,130){(c)}
+\put(-320,-8){(d)}
+\put(-140,-8){(e)}
+\caption{
+\label{fig:plots}
+Clustering results associated to the data illustrated in 
+Fig.~\ref{fig:scatter}. For each experiment we
+perform 100 Monte Carlo runs. Error bars indicate standard error.
+(a)
 High dimensional Gaussian mixture according to \eqref{eq:gauss1}.
-We sample $200$ points total on each trial. The dashed line is
+The dashed line is
 Bayes accuracy $\approx 0.86$. We use the metric $\rho_1$ in
 \eqref{eq:rho_alpha}, which is standard in energy statistics.
+(b)
+High dimensional Gaussian mixture according to \eqref{eq:gauss2}.
+Bayes accuracy $\approx 0.95$. We use $\rho_1$ in \eqref{eq:rho_alpha}.
+(c)
+Gaussian mixture with parameters \eqref{eq:20gauss}. We increase
+the number of sampled points in each trial. We use different
+metrics; see \eqref{eq:rho_alpha}--\eqref{eq:rho_hat}. 
+Here, kernel k-groups is more accurate than spectral clustering.
+(d)
+Same experiment as in Fig.~\ref{fig:plots}c but with a 
+lognormal mixture with parameters \eqref{eq:20gauss}.
+Again, kernel k-groups is more accurate than alternatives.
+The plot suggests that neither of these methods are
+consistent on this example since Bayes accuracy is $\approx 0.90$.
+(e)
+Comparison between clustering methods
+on unbalanced clusters. The data is normally 
+distributed as \eqref{eq:gauss3} where
+we vary $m \in [0, 240]$. We use the standard metric $\rho_1$ (see
+\eqref{eq:rho_alpha}) from energy statistics.
 }
-\end{figure}
+\end{figure*}
 
@@
 $[1,5]$ and other choice would give analogous results.
-Bayes  accuracy is fixed at $\approx 0.95$.
-From Fig.~\ref{fig:gauss2} we see that 
+Bayes  accuracy is fixed at $\approx 0.95$. In Fig~\ref{fig:scatter}b
+we show a scatter plot of the 9th and 10th dimension.
+From Fig.~\ref{fig:plots}b we see that 
 all the methods are similarly accurate in low dimensions, but they quickly
@@
 kernel k-groups which is much more stable.
-
-\begin{figure}
-\centering
-\includegraphics[scale=0.5]{gauss2.pdf}
-\caption{
-\label{fig:gauss2}
-High dimensional Gaussian mixture according to \eqref{eq:gauss2}.
-We sample $200$ points total on each trial. The dashed line is
-Bayes accuracy $\approx 0.95$. We use $\rho_1$ in \eqref{eq:rho_alpha}
-which standard metric in energy statistics.
-}
-\end{figure}
 
@@
 \end{equation}
-Bayes accuracy is $\approx 0.90$. 
+Bayes accuracy is $\approx 0.90$.  A scatter plot of the 4th and 5th
+dimensions is shown in Fig.~\ref{fig:scatter}c.
 We increase the sample size $n \in [10, 400]$ and show
-the accuracy versus $n$ in Fig.~\ref{fig:gauss_n}. We compare
+the accuracy versus $n$ in Fig.~\ref{fig:plots}c. We compare
 kernel k-groups, with different metrics,
@@
 compared to the other methods.
-
-\begin{figure}
-\centering
-\includegraphics[scale=0.5]{gauss_n.pdf}
-\caption{
-\label{fig:gauss_n}
-Gaussian mixture with parameters \eqref{eq:20gauss}. We increase
-the number of sampled points in each trial. We use different
-metrics; see \eqref{eq:rho_alpha}--\eqref{eq:rho_hat}. 
-Here, kernel k-groups is more accurate than spectral clustering.
-}
-\end{figure}
 
@@
 The optimal Bayes accuracy is still $\approx 0.9$.
-The results are shown in Fig.~\ref{fig:loggauss_n}.
+A scatter plot is in Fig.~\ref{fig:scatter}d and the
+results are shown in Fig.~\ref{fig:plots}d.
 We use exactly the same metrics as in the normal
-mixture of Fig.~\ref{fig:gauss_n} to illustrate that the 
+mixture of Fig.~\ref{fig:plots}c to illustrate that the 
 proposed method still performs accurately.
-
-\begin{figure}
-\centering
-\includegraphics[scale=0.5]{loggauss_n.pdf}
-\caption{
-\label{fig:loggauss_n}
-Same experiment as in Fig.~\ref{fig:gauss_n} but with a 
-lognormal mixture with parameters \eqref{eq:20gauss}.
-Again, kernel k-groups is more accurate than alternatives.
-The figure suggests that neither of these methods are
-consistent on this example since Bayes accuracy is $\approx 0.90$.
-}
-\end{figure}
 
@@
 clustering, kernel k-means and kernel k-groups all degenerate
-more quickly than GMM. This is illustrated in Figure~\ref{fig:gauss_unbal}
+more quickly than GMM. A scatter plot of the first two dimensions
+is shown in Fig.~\ref{fig:scatter}e and the clustering results are
+in Fig.~\ref{fig:plots}e,
 where we generate data according to
@@
 
+In Fig.~\ref{fig:2dscatter} we show examples of two-dimensional
+datasets whose clustering results are shown in Table~\ref{tb:cigar_circle}.
+For kernel k-means and kernel k-groups we initialize at random.
+We see that both methods perform closely, with higher accuracy
+than the other ones.
+
 \begin{figure}
-\centering
-\includegraphics[scale=0.5]{gauss_unbal.pdf}
+\includegraphics[width=.24\textwidth]{2cigars.pdf}
+\includegraphics[width=.24\textwidth]{2circles.pdf}
+\put(-240,-8){(a)}
+\put(-110,-8){(b)}
 \caption{
-\label{fig:gauss_unbal}
-Comparison between clustering methods
-on unbalanced clusters. The data is normally 
-distributed as \eqref{eq:gauss3} where
-we vary $m \in [0, 240]$. We use the standard metric $\rho_1$ (see
-\eqref{eq:rho_alpha}) from energy statistics.
+\label{fig:2dscatter}
+(a) Data distributed as 
+$x \stackrel{iid}{\sim}
+\tfrac{1}{2}\mathcal{N}(\mu_1,\Sigma_1)+
+\tfrac{1}{2}\mathcal{N}(\mu_1,\Sigma_1)$, 
+$\mu_1= (0,0)^\top$, 
+$\mu_2 = (6.5, 0)^\top$ and 
+$\Sigma_1 = \Sigma_2 = \diag(1, 20)$. We sample $800$ points.
+(b) Concentric circles with radius $r_1 = 1$ and $r_2 = 3$, with 
+noise $0.2 \cdot \mathcal{N}(0, I_2)$. We sample 
+800 points with probability $1/2$ for each class.
 } 
 \end{figure}
+
+\begin{table}
+\caption{
+\label{tb:cigar_circle}
+Clustering the data from Fig.~\ref{fig:2dscatter}.
+For spcetral clustering, kernel k-means and kernel k-groups we use
+the metric $\widetilde{\rho}_2$ (see \eqref{eq:rho_tilde}) for the data
+in Fig.~\ref{fig:2dscatter}a, while $\widehat{\rho}_1$ (see \eqref{eq:rho_hat})
+for the data in Fig.~\ref{fig:2dscatter}b. We have 800
+points on each trial and 30 Monte Carlo runs for both datasets.
+}
+\centering
+\begin{tabular}{@{}l|ll|ll@{}}
+                   & \multicolumn{2}{c|}{Fig.~\ref{fig:2dscatter}a}  &
+                     \multicolumn{2}{c}{Fig.~\ref{fig:2dscatter}b}  \\
+       Method      & Accuracy & SEM     & Accuracy & SEM \\ \midrule[.5pt]
+        kmeans       & 0.5330 & 0.0048  & 0.5210   & 0.0027 \\
+         GMM         & 0.9292 & 0.0295  & 0.5330   & 0.0039 \\
+ spectral clustering & 0.5775 & 0.0096  & 0.7253   & 0.0028 \\
+    kernel k-means   & 0.9999 & 0.0002  & 0.9999   & $4.16\, 10^{-5}$ \\
+   kernel k-groups   & 0.9987 & 0.0003  & 0.9999   & $4.16\, 10^{-5}$  
+\end{tabular}
+\end{table}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Discussion}
+\section{Conclusion}
 \label{sec:conclusion}
@@
 to several graph partitioning problems.
-Energy statistics, however, fixes
-a family of standard kernels in Euclidean space, and
-more general kernels 
-on spaces of negative type can also be obtained.
+%Energy statistics, however, fixes
+%a family of standard kernels in Euclidean space, and
+%more general kernels 
+%on spaces of negative type can also be obtained.
 
 We extended Hartigan's method to kernel spaces and proposed
-Algorithm~\ref{algo}, which we call kernel k-groups.
+Algorithm~\ref{algo}, which we called kernel k-groups.
 This method was compared to kernel k-means and spectral clustering, besides
@@
 clusters. An interesting problem  that we leave open 
-would is to extend the method to such 
-situations, as well as a soft version.
+is to extend the method to such 
+situations.
 Finally, kernel methods can benefit from sparsity and
@@
 % or
-%\appendix  % for no appendix heading
+\appendix  % for no appendix heading
 % do not use \section anymore after \appendix, only \section*
@@
 %Appendix two text goes here.
+
+\section*{Two-Class Problem in One Dimension}
+\label{sec:twoclass}
+
+Here we consider the simplest possible case which
+is one-dimensional data and a two-class problem. We propose
+an algorithm that does not depend on initialization. We used this simple
+scheme to compare with kernel k-groups given
+in algorithm \ref{algo}. Both algorithms have the same clustering
+performance in the one-dimensional examples that we tested.
+
+Let us fix 
+$\rho(x,y) = |x - y|$ according to the standard energy distance. We also
+fix the weights $w(x) = 1$ for every data point $x$. We can
+thus 
+compute the function
+\eqref{eq:g_def} in $\OO(n \log n)$ and minimize
+$W$ directly.
+This is done by noting that
+\begin{equation}
+\begin{aligned}
+|x - y|  &= (x-y)\Ind{x \ge y} -
+(x-y) \Ind{x < y}  \\
+&= 
+x \left( \Ind{x \ge y} - \Ind{x < y} \right)  + 
+y \left( \Ind{y > x} - \Ind{y \le x} \right)  
+\end{aligned}
+\end{equation}
+where we have the indicator function defined by
+$\Ind{A}=1$ if $A$ is true, and $\Ind{A}=0$ otherwise.
+Let $\C$ be a partition with
+$n$ elements. Using the above distance we have
+\begin{equation}
+\label{eq:g_ind}
+ g\left(\C,\C\right) =  \dfrac{1}{n^2}  
+\sum_{x \in \C} 
+\sum_{y \in \C} 
+x \left(
+\Ind{x \ge y} + \Ind{y > x} - 
+\Ind{x \ge y}-\Ind{x < y} \right) .
+\end{equation}
+The sum over $y$ can be eliminated since each term in
+the parenthesis is simply counting the number of elements in $\C$ that satisfy
+the condition of the indicator function. Assuming
+that we first order the data in $\C$, obtaining
+$\tC = [ x_j \in \C: x_1 \le x_2 \le \dotsm \le x_{n}]$, we
+get
+\begin{equation}
+\label{eq:g1d}
+g\big(\tC, \tC \big) = 
+\dfrac{2}{n^2} \sum_{\ell=1}^n (2\ell - 1 - n) x_\ell .
+\end{equation}
+Note that the cost of computing
+$g\big( \tC, \tC \big)$
+is $\OO(n)$ and the cost of
+sorting the data
+is at the most $\OO(n\log n)$.
+Assuming that each partition is ordered,
+$\mathbb{X} = \bigcup_{j=1}^k \tC_j$,
+the within energy dispersion
+can be written explicitly as
+\begin{equation}
+\label{eq:w1d}
+W\big( \tC_1,\dotsc,\tC_k \big) = 
+\sum_{j=1}^k \sum_{\ell=1}^{n_j} \dfrac{2\ell - 1 - n_j}{n_j} \, x_\ell.
+\end{equation}
+
+For a two-class problem we can use the formula
+\eqref{eq:w1d} to cluster the data
+through a simple algorithm
+as follows. We first order
+the entire dataset, $\mathbb{X} \to \widetilde{\mathbb{X}}$. Then
+we compute \eqref{eq:w1d} for each possible split of $\widetilde{\mathbb{X}}$
+and pick the point which gives the minimum value of $W$.
+This procedure is described in Algorithm~\ref{algo1d}.
+Note that this algorithm is deterministic,
+however,
+it only works for one-dimensional data with Euclidean distance. Its total
+complexity is $\OO(n\log n + n^2) = \OO(n^2)$.
+
+\begin{algorithm}[h]
+\vspace{.5em}
+\begin{algorithmic}[1]
+\INPUT data $\mathbb{X}$
+\OUTPUT label matrix $Z$
+\STATE sort $\mathbb{X}$ obtaining
+$\widetilde{\mathbb{X}}= [ x_1,\dotsc,x_n ]$
+    \FOR{$j\in [ 1,\dotsc,n ]$}
+        \STATE $\tC_{1,j} \leftarrow [x_i: i=1,\dotsc,j]$
+        \STATE $\tC_{2,j} \leftarrow [x_i : i=j+1,\dotsc,n]$
+        \STATE
+            $W^{(j)} \leftarrow W \big( \tC_{1,j},\tC_{2,j}\big)$
+            \hfill (see \eqref{eq:w1d})
+    \ENDFOR
+    \STATE $j^\star \leftarrow \argmin_j W^{(j)}$
+	\FOR{$j \in [1,\dotsc,n]$}
+		\IF{ $j\le j^\star$ }
+    		\STATE $Z_{j\bullet} \leftarrow (1,0) $
+        \ELSE
+			\STATE $Z_{j\bullet} \leftarrow (0,1)$
+		\ENDIF
+	\ENDFOR
+\end{algorithmic}
+\caption{
+\label{algo1d}
+Clustering algorithm to
+find local solutions to the optimization
+problem \eqref{eq:minimize}
+for a two-class problem in one dimension. 
+}
+\end{algorithm}
 

### 2018-06-16 energy_clustering/kgroups_code/data.py

@@ -165,13 +165,14 @@ def plot(X, z, fname='plot.pdf'):
     fig = plt.figure()
     ax = fig.add_subplot(111)
 
-    colors = iter(['b', 'r', 'g'])
+    #colors = iter(['#1b9e77', '#d95f02', '#7570b3'])
+    colors = iter(['#e41a1c', '#377eb8', '#4daf4a'])
     
     for k in z_unique:
         idx = np.where(z==k)
         x = X_new[idx][:,0]
         y = X_new[idx][:,1]
-        ax.plot(x, y, 'bo', markersize=4, alpha=.7, color=next(colors))
+        ax.plot(x, y, 'bo', markersize=4, alpha=.5, color=next(colors))
     
     ax.set_xticks([])
     ax.set_yticks([])
@@ -311,6 +312,63 @@ def from_sets_to_labels(A):
 ###############################################################################
 if __name__ == '__main__':
 
+    # some components of first gaussian experiment
+    n = 200
+    n1, n2 = np.random.multinomial(n, [0.5, 0.5])
+    m1 = np.zeros(2)
+    s1 = s2 = np.eye(2)
+    m2 = np.array([0.7, 0.7])
+    X, z = multivariate_normal([m1, m2], [s1, s2], [n1, n2])
+    plot(X, z, fname='./figs/gauss_means_scatter.pdf')
+    
+    # some components of second gaussian experiment
+    n = 200
+    n1, n2 = np.random.multinomial(n, [0.5, 0.5])
+    m1 = np.zeros(2)
+    m2 = np.array([1, 1])
+    s1 = np.eye(2)
+    s2 = np.array([[2.813, 0], [0, 3.637]])
+    X, z = multivariate_normal([m1, m2], [s1, s2], [n1, n2])
+    plot(X, z, fname='./figs/gauss_cov_scatter.pdf')
+    
+    # some components of third gaussian/loggaussian experiment
+    n = 200
+    n1, n2 = np.random.multinomial(n, [0.5, 0.5])
+    m1 = np.zeros(2)
+    m2 = 0.5*np.array([1, 1])
+    s1 = 0.5*np.eye(2)
+    s2 = np.eye(2)
+    X, z = multivariate_normal([m1, m2], [s1, s2], [n1, n2])
[…]

### 2018-06-16 energy_clustering/kgroups_code/gauss_loggauss_1d.py

@@ -1,0 +1,70 @@
+"""Experiments for one dimensional data."""
+
+# author: Guilherme S. Franca <guifranca@gmail.com>
+# Johns Hopkins University
+
+from __future__ import division
+
+import numpy as np
+import multiprocessing as mp
+import pandas as pd
+
+import argparse
+
+import data
+import eclust
+import metric
+import wrapper
+
+parser = argparse.ArgumentParser(description="One dimensional data.")
+parser.add_argument('-o', dest='output', required=True,
+                    type=str, action='store', help="file name for output")
+parser.add_argument('-t', dest='type', required=True, 
+                    choices=['gauss', 'loggauss'],
+                    type=str, action='store', help="can be gauss or loggauss")
+parser.add_argument('-n', type=int, required=True,
+                    dest='num_experiments', action='store', 
+                    help="number of experiments")
+
+args = parser.parse_args()
+
+number_points = range(10, 800, 20)
+k = 2
+output = args.output
+num_experiments = args.num_experiments
+type_dist = args.type
+
+def generate_data(total_points):
+    m1 = 0
+    s1 = 1.5
+    m2 = 1.5
+    s2 = 0.3
+    n1, n2 = np.random.multinomial(total_points, [0.5, 0.5])
+    if type_dist == 'gauss':
+        X, z = data.univariate_normal([m1, m2], [s1, s2], [n1, n2])
+    elif type_dist == 'loggauss':
+        X, z = data.univariate_lognormal([m1, m2], [s1, s2], [n1, n2])
+    Y = np.array([[x] for x in X])
+    return Y, z
+
+r = []
+for _ in range(num_experiments):
+    for n in number_points:
+        X, z = generate_data(n)
+        G = eclust.kernel_matrix(X, lambda x, y: np.linalg.norm(x-y))
+        
+        zh = wrapper.kmeans(k, X)
+        a = metric.accuracy(z, zh)
+        r.append(['k-means', n, a])
+        
+        zh = wrapper.gmm(k, X)
+        a = metric.accuracy(z, zh)
+        r.append(['gmm', n, a])
+        
+        zh = wrapper.kernel_kgroups(k, X, G)
+        a = metric.accuracy(z, zh)
+        r.append(['kernel k-groups', n, a])
+
+df = pd.DataFrame(np.array(r), columns=['method', 'num_points', 'accuracy'])
+df.to_csv(output, index=False)
+

### 2018-06-16 energy_clustering/kgroups_code/kernel_density.py

@@ -1,0 +1,84 @@
+"""Density estimation for the algorithms in 1D."""
+
+# Guilherme Franca <guifranca@gmail.com>
+# Johns Hopkins University
+
+import numpy as np
+#import seaborn.apionly as sns
+import seaborn as sns
+import matplotlib.pyplot as plt
+from prettytable import PrettyTable
+
+import wrapper
+import data
+import eclust
+import metric
+
+
+### generate data
+k = 2
+n = 2000
+n1, n2 = np.random.multinomial(n, [0.5, 0.5])
+m1 = 0
+s1 = 1.5
+m2 = 1.5
+s2 = 0.3
+#X, z = data.univariate_normal([m1, m2], [s1, s2], [n1, n2])
+X, z = data.univariate_lognormal([m1, m2], [s1, s2], [n1, n2])
+Y = np.array([[x] for x in X])
+
+### clustering
+t = PrettyTable(['Method', 'Accuracy'])
+G = eclust.kernel_matrix(Y, lambda x, y: np.linalg.norm(x-y))
+zh_kmeans = wrapper.kmeans(k, Y)
+t.add_row(['k-means', metric.accuracy(z, zh_kmeans)])
+zh_gmm = wrapper.gmm(k, Y)
+t.add_row(['gmm', metric.accuracy(z, zh_gmm)])
+zh_kgroups = wrapper.kernel_kgroups(k, Y, G)
+t.add_row(['kernel k-groups', metric.accuracy(z, zh_kgroups)])
+print t
+
+### estimated classes
+x1_true = X[np.where(z==0)]
+x2_true = X[np.where(z==1)]
+
+x1_kmeans = X[np.where(zh_kmeans==0)]
+x2_kmeans = X[np.where(zh_kmeans==1)]
+
+x1_gmm = X[np.where(zh_gmm==0)]
+x2_gmm = X[np.where(zh_gmm==1)]
+
+x1_kgroups = X[np.where(zh_kgroups==0)]
+x2_kgroups = X[np.where(zh_kgroups==1)]
+
+#c = ['#e41a1c', '#377eb8', '#4daf4a', '#984ea3']
+c = ['#1F77B4', '#FF7F0E', '#2CA02C']
+s = ['^', 'v', 'o']
+me = 10
+
+### doing density estimation and ploting
+ax = sns.kdeplot(x1_true, shade=True, label=r"truth", color='k')
[…]

### 2018-06-16 energy_clustering/kgroups_code/kernel_density2.py

@@ -1,0 +1,96 @@
+"""Density estimation for the algorithms in 1D."""
+
+# Guilherme Franca <guifranca@gmail.com>
+# Johns Hopkins University
+
+import numpy as np
+import matplotlib.pyplot as plt
+from sklearn.neighbors import KernelDensity
+from prettytable import PrettyTable
+
+import sys
+
+import wrapper
+import data
+import eclust
+import metric
+
+
+### generate data
+k = 2
+n = 2000
+n1, n2 = np.random.multinomial(n, [0.5, 0.5])
+m1 = 0
+s1 = 1.5
+m2 = 1.5
+s2 = 0.3
+#X, z = data.univariate_normal([m1, m2], [s1, s2], [n1, n2])
+X, z = data.univariate_lognormal([m1, m2], [s1, s2], [n1, n2])
+Y = np.array([[x] for x in X])
+bw = 0.5 # bandwidth
+num_points = 1500 # number points for linspace
+#low = -6
+#high = 6
+low = -2
+high = 20 
+
+### clustering
+t = PrettyTable(['Method', 'Accuracy'])
+G = eclust.kernel_matrix(Y, lambda x, y: np.linalg.norm(x-y))
+zh_kmeans = wrapper.kmeans(k, Y)
+t.add_row(['k-means', metric.accuracy(z, zh_kmeans)])
+zh_gmm = wrapper.gmm(k, Y)
+t.add_row(['gmm', metric.accuracy(z, zh_gmm)])
+zh_kgroups = wrapper.kernel_kgroups(k, Y, G)
+t.add_row(['kernel k-groups', metric.accuracy(z, zh_kgroups)])
+print t
+
+X_plot = np.linspace(low,high,num_points)[:,np.newaxis]
+
+### kernel density estimation
+x1_true = X[np.where(z==0)][:, np.newaxis]
+x2_true = X[np.where(z==1)][:, np.newaxis]
+
+fig = plt.figure()
+ax = fig.add_subplot(111)
+
+kde1 = KernelDensity(kernel='gaussian', bandwidth=bw).fit(x1_true)
+log_dens1 = kde1.score_samples(X_plot)
+kde2 = KernelDensity(kernel='gaussian', bandwidth=bw).fit(x2_true)
+log_dens2 = kde2.score_samples(X_plot)
[…]

### 2018-06-23 energy_clustering/pami18/energy_clustering.tex
maria comments

@@
 our numerical results show an improved performance,
-specially in high dimensions.
+especially in high dimensions.
 \end{abstract}
@@
 Clustering is an important unsupervised learning problem and 
-has such a long history in statistics and machine learning, making it
+has a long history in statistics and machine learning, making it
 impossible to mention all important contributions in a short space. 
@@
 linearly separable in Euclidean space. Gaussian mixture models (GMM) is 
-another very common approach, providing more flexibility than k-means, 
+another very common approach, providing more flexibility than k-means; 
 however, it still makes strong assumptions about the distribution of 
@@
 To account for nonlinearities, kernel methods were introduced 
-\cite{Smola,Girolami}. A mercer kernel \cite{Mercer} is used to implicitly
+\cite{Smola,Girolami}. A Mercer kernel \cite{Mercer} is used to implicitly
 map data points to a RKHS, then clustering can be performed in the associated
@@
 
-The predominant approach to clustering %, in Euclidean or kernel spaces,
-is based on Lloyd's method.
-However, there is an old heuristic
-due to Hartigan \cite{Hartigan1975,Hartigan1979}, which apparently did
-not receive the deserved attention, 
-and goes as follows: for each data point, simply assign it to a cluster
+Besides Lloyd's approach to clustering %, in Euclidean or kernel spaces,
+there is an old heuristic
+due to Hartigan \cite{Hartigan1975,Hartigan1979} that
+goes as follows: for each data point, simply assign it to a cluster
 in an optimal way such that a loss function is minimized.
@@
 Moreover, the advantages of Hartigan's over Lloyd's method was verified
-empirically \cite{Telgarsky,Slonin}, however, 
+empirically \cite{Telgarsky,Slonin}. However, 
 although it was observed to be as fast as Lloyd's method, no complexity
@@
 kernel k-groups 
-is more accurate and robust, specially in high dimensions.
+is more accurate and robust, especially in high dimensions.
 
@@
 More specifically, our method performs 
-closely to k-means and GMM on normally distributed data, however,
+closely to k-means and GMM on normally distributed data, while
 it is significantly better on data that 
@@
 is striking, being more accurate than k-means and GMM 
-even on Gaussian settings. We also illustrate the advantages of
-kernel k-groups on a real dataset.
+even in Gaussian settings. We also illustrate the advantages of
+kernel k-groups on real data.
 

### 2018-08-10 energy_clustering/pami18/energy_clustering.tex
minor updates

@@
 
-\title{Kernel k-Groups and Hartigan's Method}
-
-\author{Guilherme~Fran\c ca,~Maria Rizzo~and~Joshua~Vogelstein
+\title{Kernel k-Groups via Hartigan's Method}
+
+\author{Guilherme~Fran\c ca,~Maria Rizzo~and~Joshua T.~Vogelstein
 \IEEEcompsocitemizethanks{\IEEEcompsocthanksitem G. Fran\c ca 
@@
 % The paper headers
-\markboth{Journal of \LaTeX\ Class Files,~Vol.~14, No.~8, August~2015}%
-{Energy Clustering}
+\markboth{Energy Clustering}{}
+%{Journal of \LaTeX\ Class Files,~Vol.~14, No.~8, August~2015}%
 
@@
 analysis of variance \cite{RizzoVariance}, nonlinear dependence tests through
-distance covariance and distance correlation, which generalizes the Pearson
+distance covariance and distance correlation~\cite{Szekely2007-mm}, which generalizes the Pearson
 correlation coefficient, and hierarchical clustering 
@@
 been recently established \cite{Sejdinovic2013} 
-through an equivalence between generalized energy distances and maximum
+through an asymptotic equivalence between generalized energy distances and maximum
 mean discrepancies (MMD), which are distances between embeddings of 
 distributions in reproducing kernel Hilbert spaces (RKHS).
+Even more recently, generalized energy distances and kernel methods have been demonstrated to be exactly equivalent, for all finite samples~\cite{Shen2018-st}.
 This equivalence immediately relates energy statistics to
@@
 which 
-we do so in more generality. For instance, our approach is not limited
+we do   more generality. For instance, our approach is not limited
 to the Euclidean case but holds for general arbitrary spaces
@@
 being NP-hard, thus a direct approach 
-is computational prohibitive even for small datasets.
+is computationally prohibitive even for small datasets.
 However, one can find approximate solutions by relaxing some 

### 2018-08-10 energy_clustering/pami18/energy_clustering.tex
fixed # of sigdigs

@@
 
-The following experimental setup holds unless specified
-otherwise. 
+%The following experimental setup holds unless specified otherwise. 
 We consider
@@
 \eqref{eq:kernel_semimetric}, where we always fix $x_0=0$.
-%The standard metric from energy statistics is $\rho_1$ and 
+We use $\rho_\alpha$ by default, unless otherwise specified.
 %will always be present in the experiments as a reference.
@@
 kernel k-groups with k-means and GMM, as illustrated in 
-Fig.~\ref{fig:1dexperiment}. We have two mixtures of normal and 
-lognormal distributions. Notice that in the kernel density estimation plots
+Fig.~\ref{fig:1dexperiment}. The left panels show a mixture of Gaussians, and the right panels show a mixture of log Gaussians (see caption for details).
+% have two mixtures of normal and lognormal distributions. 
+Notice that in the kernel density estimation plots
 for lognormal distribution, only kernel k-groups was able to distinguish
@@
 cases are in Table~\ref{tb:dens}. We remark that kernel k-groups in
-this one-dimensional example performed the same as the deterministic
+this one-dimensional example performed the same as the exact deterministic
 Algorithm~\ref{algo1d} introduced in the Appendix.
@@
         kmeans       & 0.773  & 0.502     \\
-         GMM         & 0.881  & 0.513     \\
-   kernel k-groups   & 0.802  & 0.847  
+         GMM         & \textbf{0.881}  & 0.513     \\
+   kernel k-groups   & 0.802  & \textbf{0.847}  
 \end{tabular}
@@
        Method      & Accuracy & SEM     & Accuracy & SEM \\ \midrule[.5pt]
-        kmeans       & 0.5330 & 0.0048  & 0.5210   & 0.0027 \\
-         GMM         & 0.9292 & 0.0295  & 0.5330   & 0.0039 \\
- spectral clustering & 0.5775 & 0.0096  & 0.7253   & 0.0028 \\
-    kernel k-means   & 0.9999 & 0.0002  & 0.9999   & $4.16\, 10^{-5}$ \\
-   kernel k-groups   & 0.9987 & 0.0003  & 0.9999   & $4.16\, 10^{-5}$  
+        kmeans       & 0.533 & 0.005  & 0.521   & 0.003 \\
+         GMM         & 0.929 & 0.029  & 0.533   & 0.004 \\
+ spectral clustering & 0.577 & 0.010  & 0.725   & 0.003 \\
+    kernel k-means   & \textbf{1.000} & 0.000  & \textbf{1.000}   & 0.000 \\
+   kernel k-groups   & \textbf{1.000} & 0.000  & \textbf{1.000}  & 0.000  
 \end{tabular}
@@
        Method      & Accuracy & aRand \\ \midrule[.5pt]
-        kmeans       & 0.7131 & 0.6897 \\
-         GMM         & 0.8770 & 0.8403 \\
- spectral clustering & 0.9536 & 0.9125 \\
-    kernel k-means   & 0.7514 & 0.8510 \\
-   kernel k-groups   & \textbf{0.9617} & \textbf{0.9356}
+        kmeans       & 0.713 & 0.690 \\
+         GMM         & 0.877 & 0.840 \\
+ spectral clustering & 0.954 & 0.912 \\
+    kernel k-means   & 0.751 & 0.851 \\
+   kernel k-groups   & \textbf{0.962} & \textbf{0.936}
 \end{tabular}

### 2018-08-10 energy_clustering/pami18/energy_clustering.tex
added jovo bio

@@
 
-\begin{IEEEbiographynophoto}{Joshua Vogelstein}
-Biography text here.
+\begin{IEEEbiographynophoto}{Joshua T.~Vogelstein}
+I received a B.S degree from the Department of Biomedical Engineering (BME) at Washington University in St.
+Louis, MO in 2002, a M.S.
+degree from the Department of Applied Mathematics and Statistics (AMS) at Johns Hopkins University (JHU) in Baltimore, MD in 2009, and a Ph.D.
+degree from the Department of Neuroscience at JHU in 2009.
+I was a Postdoctoral Fellow in AMS@JHU from 2009 until 2011, at which time I was appointed an Assistant Research Scientist, and became a member of the Institute for Data Intensive Science and Engineering.
+I spent 2 years at Information Initiative at Duke University, before coming home to my current appointment as Assistant Professor in BME@JHU, and core faculty in both the Institute for Computational Medicine and the Center for Imaging Science, as well as a member of the Kavli Neuroscience Discovery Institute.
+I married my kindergarten sweetheart in the summer of 2014.
+
+My primary research interest is to extend and fuse statistical machine learning and big data science to address the most important brain science and mental health questions of our time, particularly  connectomics.
+Our group’s research has been featured in a number of prominent scientific and engineering journals and conferences including Nature, Nature Methods, Science, Cell, PNAS, Science Translational Medicine, JMLR, PAMI, Annals of Applied Statistics,  NIPS, and SIAM Journal of Matrix Analysis and Applications.
+In 2011, I co-founded the Open Connectome Project which expanded in 2015 to be NeuroData, whose mission is to enable terascale neuroscience for everyone.
+All our work is conducted according to the highest standards of open science.
 \end{IEEEbiographynophoto}

### 2018-08-10 energy_clustering/pami18/energy_clustering.tex
pami update

@@
 \tfrac{1}{2}e^{\mathcal{N}(0, 1.5)} + \tfrac{1}{2} e^{\mathcal{N}(1.5, 0.3)}$.
-In both cases we plot the accuracy versus the total number of points.
+In both cases we plot the average accuracy versus the total number of points.
+We ommit error bars since they are too small.
 (c) Kernel density estimation for a single trial with 2000 sampled
@@
 Fig.~\ref{fig:scatter}. For each experiment we
-perform 100 Monte Carlo runs. Error bars indicate standard error.
+perform 100 Monte Carlo runs and show the average accuracy. 
+We ommit error bars since they are too small to be visible.
 (a)
@@
 \begin{figure}
-\includegraphics[width=.24\textwidth]{2cigars.pdf}
-\includegraphics[width=.24\textwidth]{2circles.pdf}
-\put(-240,-8){(a)}
-\put(-110,-8){(b)}
+\centering
+\begin{minipage}{0.2\textwidth}
+\includegraphics[width=1\textwidth]{2cigars.pdf}\\[-.2em](a)
+\end{minipage}\hspace{2em}
+\begin{minipage}{0.2\textwidth}
+\includegraphics[width=1\textwidth]{2circles.pdf}\\[-.2em](b)
+\end{minipage}
 \caption{
@@
 
-\begin{IEEEbiography}{Guilherme Fran\c ca}
-Biography text here.
+\begin{IEEEbiography}[{\includegraphics[width=1in,clip,keepaspectratio]{gui2.png}}]{Guilherme Fran\c ca} 
+has a BSc in physics, and received both his MSc and 
+PhD in Theoretical Physics from the Institute of 
+Theoretical Physics IFT-UNESP/ICTP-SAIFR (2012). He was a Postdoctoral Fellow
+with the High Energy Theory group in the Physics Department 
+at Cornell University (2013--2015), working on topics of mathematical physics
+and non-perturbative methods in quantum field theory 
+and statistical mechanics.
+He was a Postdoctoral Fellow in
+the Computer Science Department at Boston College (2015), working in
+optimization theory.
+Since 2016 he is a Postdoctoral Fellow at Johns Hopkins University, currently
+with the Mathematical Institute for Data Science (MINDS). 
+His current research interests are in theoretical methods in the broad
+area of machine learning 
+and optimization.
 \end{IEEEbiography}
@@
 
-\begin{IEEEbiographynophoto}{Joshua T.~Vogelstein}
+\begin{IEEEbiography}[{\includegraphics[width=1in,clip,keepaspectratio]{jovo.png}}]{Joshua~T.~Vogelstein}
 I received a B.S degree from the Department of Biomedical Engineering (BME) at Washington University in St.
@@
 All our work is conducted according to the highest standards of open science.
-\end{IEEEbiographynophoto}
+\end{IEEEbiography}
 

### 2018-08-10 energy_clustering/kgroups_code/kernel_density2.py

@@ -24,15 +24,15 @@ m1 = 0
 s1 = 1.5
 m2 = 1.5
 s2 = 0.3
-#X, z = data.univariate_normal([m1, m2], [s1, s2], [n1, n2])
-X, z = data.univariate_lognormal([m1, m2], [s1, s2], [n1, n2])
+X, z = data.univariate_normal([m1, m2], [s1, s2], [n1, n2])
+#X, z = data.univariate_lognormal([m1, m2], [s1, s2], [n1, n2])
 Y = np.array([[x] for x in X])
 bw = 0.5 # bandwidth
 num_points = 1500 # number points for linspace
-#low = -6
-#high = 6
-low = -2
-high = 20 
+low = -6
+high = 6
+#low = -2
+#high = 20 
 
 ### clustering
 t = PrettyTable(['Method', 'Accuracy'])
@@ -59,7 +59,9 @@ log_dens1 = kde1.score_samples(X_plot)
 kde2 = KernelDensity(kernel='gaussian', bandwidth=bw).fit(x2_true)
 log_dens2 = kde2.score_samples(X_plot)
 ax.fill_between(X_plot[:,0], np.exp(log_dens1), alpha=.3, color='k')
+ax.plot(X_plot[:,0], np.exp(log_dens1), color='k', label='truth')
 ax.fill_between(X_plot[:,0], np.exp(log_dens2), alpha=.3, color='k')
+ax.plot(X_plot[:,0], np.exp(log_dens2), color='k')
 
 r = []
 x1_kmeans = X[np.where(zh_kmeans==0)][:, np.newaxis]
@@ -73,8 +75,7 @@ x2_kgroups = X[np.where(zh_kgroups==1)][:, np.newaxis]
 r.append([x1_kgroups, x2_kgroups])
 
 methods = ['k-means', 'GMM', 'kernel k-groups']
-#c = ['#e41a1c', '#377eb8', '#4daf4a', '#984ea3']
-colors = ['#1F77B4', '#FF7F0E', '#2CA02C']
+colors = ["#54278f", "#5F8793", "#F41711"]
 s = ['^', 'v', 'o']
 
 for i, (d, c, m) in enumerate(zip(r, colors, methods)):
@@ -82,15 +83,16 @@ for i, (d, c, m) in enumerate(zip(r, colors, methods)):
     log_dens1 = kde1.score_samples(X_plot)
[…]

### 2018-08-10 energy_clustering/kgroups_code/plot_gauss2.py

@@ -25,6 +25,7 @@ df = pd.concat(pd.read_csv(f) for f in fnames)
 methods = ['k-means', 'gmm', 'spectral clustering', 'kernel k-means', 
            'kernel k-groups']
 markers = iter(['^', 'v', 'p', 's', 'o'])
+colors = iter(["#54278f", "#5F8793", "#99B4C6", "#318dde", "#F41711"])
 
 dimensions = np.unique(df['dimension'].values)
 
@@ -36,10 +37,12 @@ for method in methods:
         df2 = df[(df['method']==method) & (df['dimension']==d)]
         r.append([d, df2['accuracy'].mean(), df2['accuracy'].sem()])
     r = np.array(r)
-    ax.plot(r[:,0], [0.95]*len(r[:,0]), '--', linewidth=1, color='k', alpha=0.5)
+    ax.plot(r[:,0], [0.95]*len(r[:,0]), '--', linewidth=1, color='k')
     mk = next(markers)
-    ax.errorbar(r[:,0], r[:,1], yerr=r[:,2], marker=mk, elinewidth=1,
-                label=method)
+    cl = next(colors)
+    #ax.errorbar(r[:,0], r[:,1], yerr=r[:,2], marker=mk, elinewidth=1,
+    #            label=method)
+    ax.errorbar(r[:,0], r[:,1], marker=mk, color=cl, label=method)
 ax.set_xlabel(r'$D$')
 ax.set_ylabel(r'accuracy')
 ax.set_xlim([10,700])

### 2018-08-10 energy_clustering/kgroups_code/plot_gauss_loggaus_n.py

@@ -28,6 +28,10 @@ methods = ['k-means', 'gmm',
             r'kernel k-groups $\rho_{1/2}$',
             r'kernel k-groups $\widetilde{\rho}_{1}$']
 markers = iter(['^', 'v', 'p', 'o', 'o', 'o'])
+#colors = iter(["#54278f", "#5F8793", "#99B4C6", "#318dde", "#F41711"])
+colors = iter(["#54278f", "#5F8793", "#99B4C6", 
+               "#F41711", "#F41711", "#F41711"])
+lines = iter(["-", "-", "-", "-", "--", ":"])
 
 points = np.unique(df['points'].values)
 
@@ -42,7 +46,11 @@ for method in methods:
     #ax.plot(r[:,0], [0.9]*len(r[:,0]), '--', linewidth=1, color='k',
     #        alpha=0.5)
     mk = next(markers)
-    ax.errorbar(r[:,0], r[:,1], yerr=r[:,2], marker=mk, elinewidth=1,
+    cl = next(colors)
+    ln = next(lines)
+    #ax.errorbar(r[:,0], r[:,1], yerr=r[:,2], marker=mk, elinewidth=1,
+    #            label=method)
+    ax.errorbar(r[:,0], r[:,1], marker=mk, linestyle=ln, color=cl, 
                 label=method)
 ax.set_xlabel(r'$n$')
 ax.set_ylabel(r'accuracy')

### 2018-08-10 energy_clustering/kgroups_code/plot_gauss_loggauss_1d.py

@@ -23,6 +23,7 @@ df = pd.concat(pd.read_csv(f) for f in fnames)
 
 methods = ['k-means', 'gmm', 'kernel k-groups']
 markers = iter(['^', 'v', 'o'])
+colors = iter(["#54278f", "#5F8793", "#F41711"])
 
 num_points = np.unique(df['num_points'].values)
 
@@ -36,8 +37,10 @@ for method in methods:
     r = np.array(r)
     ax.plot(r[:,0], [0.88]*len(r[:,0]), '--', linewidth=1, color='k', alpha=0.5)
     mk = next(markers)
-    ax.errorbar(r[:,0], r[:,1], yerr=r[:,2], marker=mk, elinewidth=1,
-                label=method)
+    cl = next(colors)
+    #ax.errorbar(r[:,0], r[:,1], yerr=r[:,2], marker=mk, elinewidth=1,
+    #            label=method)
+    ax.plot(r[:,0], r[:,1], marker=mk, label=method, color=cl)
 ax.set_xlabel(r'$n$')
 ax.set_ylabel(r'accuracy')
 ax.set_xlim([10,800])

### 2018-08-10 energy_clustering/kgroups_code/plot_gauss_unbal.py

@@ -24,6 +24,7 @@ df = pd.concat(pd.read_csv(f) for f in fnames)
 methods = ['k-means', 'gmm', 'spectral clustering', 'kernel k-means', 
            'kernel k-groups']
 markers = iter(['^', 'v', 'p', 's', 'o'])
+colors = iter(["#54278f", "#5F8793", "#99B4C6", "#318dde", "#F41711"])
 
 points = np.unique(df['points'].values)
 
@@ -36,8 +37,10 @@ for method in methods:
         r.append([p, df2['accuracy'].mean(), df2['accuracy'].sem()])
     r = np.array(r)
     mk = next(markers)
-    ax.errorbar(r[:,0], r[:,1], yerr=r[:,2], marker=mk, elinewidth=1,
-                label=method)
+    cl = next(colors)
+    #ax.errorbar(r[:,0], r[:,1], yerr=r[:,2], marker=mk, elinewidth=1,
+    #            label=method)
+    ax.plot(r[:,0], r[:,1], marker=mk, color=cl, label=method)
 ax.set_xlabel(r'$\#$ unbalanced points')
 ax.set_ylabel(r'accuracy')
 ax.set_xlim([0,240])

### 2018-08-13 energy_clustering/pami18/energy_clustering.tex
pami update
pami

@@
 with the Mathematical Institute for Data Science (MINDS). 
-His current research interests are in theoretical methods in the broad
-area of machine learning 
+His current research interests are in theoretical methods in 
+machine learning 
 and optimization.
@@
 
-% if you will not have a photo at all:
-\begin{IEEEbiographynophoto}{Maria Rizzo}
-Biography text here.
-\end{IEEEbiographynophoto}
+\begin{IEEEbiography}[{\includegraphics[width=1in,clip,keepaspectratio,height=1.25in]{rizzo.jpg}}]{Maria Rizzo}
+I hold a BS and MA degree in Mathematics from the University of Toledo, a MS in
+Applied Statistics
+from BGSU, and a PhD from BGSU in 2002. After four years as Assistant Professor
+at Ohio University
+Department of Mathematics, I returned to BGSU in 2006, where I am currently
+Professor of Statistics in
+the Department of Mathematics and Statistics. Primary research interests are
+centered on energy statistics,
+distance correlation and its applications, computational statistics, and
+related topics. In addition
+to teaching statistics, data science, and actuarial science, I enjoy writing
+textbooks about
+statistical computing and using R software, a forthcoming research monograph on
+energy statistics
+(joint with G. Szekely) and have other book writing projects under contract as
+well.
+\end{IEEEbiography}
 

### 2018-08-13 energy_clustering/pami18/paper-insert.tex
pami

--- 2018-08-13 energy_clustering/pami18/energy_clustering.tex
@@
-\documentclass[10pt,journal,compsoc]{IEEEtran}
+%%%%%%%% insert above Conclusion???
 
-\usepackage{cite}
-\usepackage{amsmath,amssymb,amsfonts,amsthm,amscd,bm,bbm}
-\usepackage{mathtools}
-%\mathtoolsset{showonlyrefs}
-%\usepackage[inline]{enumitem}
-\usepackage[pdftex]{graphicx}
-\graphicspath{{./figs/}}
-\usepackage{algorithmic}
-\usepackage{algorithm}
-%\usepackage[none]{hyphenat}
-\usepackage{multirow}
-\usepackage{colortbl}
-\usepackage{array,booktabs}
-\usepackage{xcolor}
-\usepackage{makecell}
-\usepackage{pifont}
-\newcommand{\cmark}{\ding{51}}
-\newcommand{\xmark}{\ding{55}}
+An implementation of unweighted $k$-groups is available in the
+energy package for R \cite{energy}.
 
-\newtheorem{theorem}{Theorem}
-\newtheorem{definition}[theorem]{Definition}
-\newtheorem{assumption}[theorem]{Assumption}
-\newtheorem{lemma}[theorem]{Lemma}
-\newtheorem{corollary}[theorem]{Corollary}
-\newtheorem{proposition}[theorem]{Proposition}
-\newtheorem{conjecture}[theorem]{Conjecture}
-\newtheorem{remark}[theorem]{Remark}
-\newtheorem{example}{Example}
+%%%%%%%%%  here is the citation for the package, update submitted to CRAN on Aug. 11
 
-\DeclareMathOperator{\aff}{aff}
-\DeclareMathOperator{\st}{s.t.}
-\DeclareMathOperator{\affnot}{aff_0}
-\DeclareMathOperator{\conv}{conv}
-\DeclareMathOperator{\relint}{relint}
-\DeclareMathOperator{\vol}{vol}
-\DeclareMathOperator{\range}{range}
-\DeclareMathOperator{\image}{im}
-\DeclareMathOperator{\nullspace}{null}
-\DeclareMathOperator{\area}{area}
-\DeclareMathOperator{\vspan}{span}
-\DeclareMathOperator{\id}{Id}
-\DeclareMathOperator{\cond}{cond}
-\DeclareMathOperator{\prox}{prox}
-\DeclareMathOperator*{\argmax}{arg\,max}
-\DeclareMathOperator*{\argmin}{arg\,min}
-\DeclareMathOperator*{\minimize}{minimize}
-\DeclareMathOperator{\diag}{diag}
-\DeclareMathOperator{\Tr}{Tr}
+  Maria Rizzo and Gabor Szekely (2018). energy: E-Statistics: Multivariate Inference via the Energy of
+  Data. R package version 1.7-5. https://CRAN.R-project.org/package=energy
 
-\newcommand\Energy{\mathcal{E}}
-\newcommand\EnergyH{\mathcal{E}^{H}}
-\newcommand\EnergyL{\mathcal{E}^{L}}
-\newcommand\EnergyOne{\mathcal{E}^{1D}}
-\newcommand\E{\mathbb{E}}
-\newcommand\kk{K}
-\newcommand\kkk{h}
-\newcommand\Hk{{\mathcal{H}}_{\kk}}
-\newcommand\HH{\mathcal{H}}
-\newcommand\C{{\mathcal{C}}}
-\newcommand\tC{{\widetilde{\C}}}
-\newcommand\OO{{\mathcal{O}}}
-\newcommand\W{{\mathcal{W}}}
-\newcommand\Zt{Y}
-\newcommand{\Ind}[1]{\mathbbm{1}_{#1}}
-\newcommand\e{e}
-\newcommand\om{\omega}
+or
 
-\hyphenation{op-tical net-works semi-conduc-tor}
+@Manual{energy,
+    title = {energy: E-Statistics: Multivariate Inference via the Energy of Data},
+    author = {Maria Rizzo and Gabor Szekely},
+    year = {2018},
+    note = {R package version 1.7-5},
+    url = {https://CRAN.R-project.org/package=energy},
+  }
 
-\begin{document}
-
-\title{Kernel k-Groups via Hartigan's Method}
-
-\author{Guilherme~Fran\c ca,~Maria Rizzo~and~Joshua T.~Vogelstein
-\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem G. Fran\c ca 
-and J. Vogelstein are with the Center for Imaging Science, Johns Hopkins
-University; J. Vogelstein is also with 
-the Department of Biomedical Engineering and Institute for Computational
-Medicine, Johns Hopkins University. \protect\\
-% note need leading \protect in front of \\ to get a newline within \thanks as
-% \\ is fragile and will error, could use \hfil\break instead.
-E-mail: guifranca@jhu.edu; jovo@jhu.edu
-\IEEEcompsocthanksitem M. Rizzo is with the Department of Mathematics
-and Statistics, Bowling Green State University. \protect\\
-E-mail: mrizzo@bgsu.edu}% <-this % stops an unwanted space
-\thanks{Manuscript received March 15, 2018; revised March 15, 2018.}}
-
-% The paper headers
-\markboth{Energy Clustering}{}
-%{Journal of \LaTeX\ Class Files,~Vol.~14, No.~8, August~2015}%
-
-\IEEEtitleabstractindextext{%
-\begin{abstract}
-Energy statistics was proposed by Sz\' ekely in the 80's inspired by 
-Newton's gravitational potential in classical mechanics, and it provides
-a model-free 
-hypothesis test for equality of distributions. In its original form, energy
-statistics was formulated in 
-Euclidean spaces. More recently, it 
-was generalized to metric spaces of negative type. 
-In this paper, we consider a formulation for the clustering problem
-using a weighted version of 
-energy statistics in spaces of negative type.
-We show that this approach leads to a
-quadratically constrained
-quadratic program in the associated kernel space, establishing
-connections with graph partitioning problems and
-kernel methods in unsupervised machine learning.
-To find local solutions of such an optimization problem, 
-we propose an extension of Hartigan's method to kernel spaces.
-Our method has the same computational cost
-as kernel k-means algorithm, which is based on Lloyd's heuristic, but 
-our numerical results show an improved performance,
-especially in high dimensions.
-\end{abstract}
-
-% Note that keywords are not normally used for peerreview papers.
-\begin{IEEEkeywords}
-Clustering, Energy Statistics, Kernel Methods.
-\end{IEEEkeywords}}
-
-% make the title area
-\maketitle
-
-% To allow for easy dual compilation without having to reenter the
-% abstract/keywords data, the \IEEEtitleabstractindextext text will
-% not be used in maketitle, but will appear (i.e., to be "transported")
-% here as \IEEEdisplaynontitleabstractindextext when the compsoc 
-% or transmag modes are not selected <OR> if conference mode is selected 
-% - because all conference papers position the abstract like regular
-% papers do.
-\IEEEdisplaynontitleabstractindextext
-% \IEEEdisplaynontitleabstractindextext has no effect when using
-% compsoc or transmag under a non-conference mode.
-
-% For peer review papers, you can put extra information on the cover
-% page as needed:
-% \ifCLASSOPTIONpeerreview
-% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
-% \fi
-%
-% For peerreview papers, this IEEEtran command inserts a page break and
-% creates the second title. It will be ignored for other modes.
-\IEEEpeerreviewmaketitle
-
-\IEEEraisesectionheading{\section{Introduction}\label{sec:introduction}}
-% Computer Society journal (but not conference!) papers do something unusual
-% with the very first section heading (almost always called "Introduction").
-% They place it ABOVE the main text! IEEEtran.cls does not automatically do
-% this for you, but you can achieve this effect with the provided
-% \IEEEraisesectionheading{} command. Note the need to keep any \label that
-% is to refer to the section immediately after \section in the above as
-% \IEEEraisesectionheading puts \section within a raised box.
-
-% The very first letter is a 2 line initial drop letter followed
-% by the rest of the first word in caps (small caps for compsoc).
-% 
-% form to use if the first word consists of a single letter:
-% \IEEEPARstart{A}{demo} file is ....
-% 
-% form to use if you need the single drop letter followed by
-% normal text (unknown if ever used by the IEEE):
-% \IEEEPARstart{A}{}demo file is ....
-% 
-% Some journals put the first two words in caps:
-% \IEEEPARstart{T}{his demo} file is ....
-% 
-% Here we have the typical use of a "T" for an initial drop letter
-% and "HIS" in caps to complete the first word.
-
-\IEEEPARstart{E}{nergy Statistics} \cite{Szkely2013,Szkely2017}
-is based on a 
-notion of statistical potential energy between probability distributions,
-in close analogy to Newton's gravitational potential in classical mechanics.
-When probability distributions are different, the 
-``statistical potential energy'' diverges as sample size increases, 
-while tends 
-to a nondegenerate limit distribution when probability
-distributions are equal. 
-Thus, it provides a model-free hypothesis test for equality of 
-distributions which is achieved under minimum energy.
-
-Energy statistics has been applied to several goodness-of-fit 
-hypothesis tests, multi-sample tests of equality of distributions, 
-analysis of variance \cite{RizzoVariance}, nonlinear dependence tests through
-distance covariance and distance correlation~\cite{Szekely2007-mm}, which generalizes the Pearson
-correlation coefficient, and hierarchical clustering 
-by extending Ward's method of minimum variance
-\cite{RizzoClustering}; 
-see \cite{Szkely2013,Szkely2017} for an overview of energy
-statistics and its applications.
-Moreover, in Euclidean spaces, an application of 
-energy statistics to clustering
-was recently proposed \cite{Kgroups}, and the method was named 
-\emph{k-groups}.
-
-In its original formulation, energy statistics has a compact representation
-in terms of expectations of pairwise Euclidean distances, providing
-straightforward empirical estimates. 
-More recently, the notion of distance covariance was further 
-generalized from Euclidean 
-spaces to metric spaces of negative type \cite{Lyons}. Furthermore, 
-the link between energy distance based tests and kernel 
-based tests has 
-been recently established \cite{Sejdinovic2013} 
-through an asymptotic equivalence between generalized energy distances and maximum
-mean discrepancies (MMD), which are distances between embeddings of 
-distributions in reproducing kernel Hilbert spaces (RKHS).
-Even more recently, generalized energy distances and kernel methods have been demonstrated to be exactly equivalent, for all finite samples~\cite{Shen2018-st}.
-This equivalence immediately relates energy statistics to
-kernel methods often used in machine learning, and form the basis 
-of our approach in this paper.
-
-Clustering is an important unsupervised learning problem and 
-has a long history in statistics and machine learning, making it
-impossible to mention all important contributions in a short space. 
-Perhaps, the most used method is k-means \cite{Lloyd,MacQueen,Forgy}, which
-is based on Lloyd's heuristic \cite{Lloyd} of iteratively computing
-the means of each cluster and then assigning points to
-the cluster with closest center. The only statistical 
-information about each cluster comes from its mean, making the method sensitive 
-to outliers. Nevertheless, k-means works very well when data is 
-linearly separable in Euclidean space. Gaussian mixture models (GMM) is 
-another very common approach, providing more flexibility than k-means; 
-however, it still makes strong assumptions about the distribution of 
-the data.
-
-To account for nonlinearities, kernel methods were introduced 
-\cite{Smola,Girolami}. A Mercer kernel \cite{Mercer} is used to implicitly
-map data points to a RKHS, then clustering can be performed in the associated
-Hilbert space by using its inner product. However, the kernel choice remains 
-the biggest challenge since there is no principled theory to construct a kernel
-for a given dataset, and usually a kernel introduces hyperparameters that 
-need to be carefully chosen. A well-known kernel based clustering method
-is kernel k-means, which is precisely k-means 
-formulated in the feature space \cite{Girolami}. 
-Furthermore, kernel k-means algorithm
-\cite{Dhillon2,Dhillon} is still based on Loyd's heuristic.
-We refer the reader to \cite{Filippone} for a survey of clustering
-methods.
-
-Besides Lloyd's approach to clustering %, in Euclidean or kernel spaces,
-there is an old heuristic
-due to Hartigan \cite{Hartigan1975,Hartigan1979} that
-goes as follows: for each data point, simply assign it to a cluster
-in an optimal way such that a loss function is minimized.
-While Lloyd's method only iterates if some cluster contains a point
-that is closer to the mean of another cluster, Hartigan's method may iterate
-even if that is not the case, and moreover, it takes into account the motion
-of the means resulting from the reassignments. In this sense, Hartigan's
-method may potentially escape local minima of Lloyd's method.
-In the Euclidean case, this was shown to be the case \cite{Telgarsky}.
-Moreover, the advantages of Hartigan's over Lloyd's method was verified
-empirically \cite{Telgarsky,Slonin}. However, 
-although it was observed to be as fast as Lloyd's method, no complexity
-analysis was provided.
-
-\subsection*{Contributions}
-Although k-groups considers clustering from energy statistics
-in the particular Euclidean case \cite{Kgroups},
-the precise optimization problem behind this approach
-remains obscure, as well as the connection with other methods
-in machine learning.
-The main theoretical contribution of this paper is to fill these gaps, 
-which 
-we do   more generality. For instance, our approach is not limited
-to the Euclidean case but holds for general arbitrary spaces
-of negative type. 
-For generality, we also
-consider a weighted version of energy statistics.
-Our approach reveals connections
-between energy statistics based clustering and existing methods
-such as kernel k-means and graph partitioning problems.
-
-Another contribution of this paper is to extend 
-Hartigan's method to  kernel spaces. 
-To the best of our knowledge, such an extension was not previously considered.
-Since this approach was motivated by energy statistics and  
-\cite{Kgroups} considered the Euclidean case, we call the 
-proposed method \emph{kernel k-groups}.
-We show that kernel k-groups 
-has the same complexity as kernel k-means algorithm, however,
-our numerical results provide compelling evidence that 
-kernel k-groups 
-is more accurate and robust, especially in high dimensions.
-
-Using the standard kernel defined by energy
-statistics, our experiments illustrate 
-that kernel k-groups
-is able to perform accurately on data coming from 
-very different distributions, contrary to k-means and GMM, for instance.
-More specifically, our method performs 
-closely to k-means and GMM on normally distributed data, while
-it is significantly better on data that 
-is not normally distributed. 
-Its superiority in high dimensions
-is striking, being more accurate than k-means and GMM 
-even in Gaussian settings. We also illustrate the advantages of
-kernel k-groups on real data.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Review of Energy Statistics and RKHS}
-\label{sec:background}
-
-In this section, we introduce the main concepts from energy
-statistics and its relation to 
-RKHS which form the basis of our work.
-For more details we refer 
-to \cite{Szkely2013} and \cite{Lyons,Sejdinovic2013}.
-
-Consider random variables in $\mathbb{R}^D$ 
-such that $X,X' \stackrel{iid}{\sim} P$ and 
-$Y,Y' \stackrel{iid}{\sim} Q$, where $P$ and $Q$ are cumulative
-distribution functions with finite first moments. 
-The quantity 
-\begin{equation}
-\label{eq:energy}
-\Energy(P, Q) \equiv 2 \E \| X - Y\| - \E \| X - X' \| - \E \| Y - Y' \|,
-\end{equation}
-called \emph{energy distance} \cite{Szkely2013}, 
-is rotationally invariant and nonnegative, $\Energy(P,Q) \ge 0$, where
-equality
-to zero holds if and only if $P = Q$.
-Above, $\| \cdot \|$ denotes the
-Euclidean norm in $\mathbb{R}^D$. 
-Energy distance
-provides a characterization of equality of distributions, and
-$\Energy^{1/2}$ is
-a metric on the space of distributions.
-
-The energy distance can be generalized as, for instance,
-\begin{equation}
-\label{eq:energy2}
-\Energy_\alpha(P, Q) \equiv 
-2 \E \| X - Y\|^{\alpha} - \E \| X - X' \|^{\alpha} - 
-\E \| Y - Y' \|^{\alpha}
-\end{equation}
-where $0<\alpha\le 2$. This quantity is also nonnegative,
-$\Energy_\alpha(P,Q) \ge 0$. Furthermore, for $0<\alpha<2$ we have that
-$\Energy_\alpha(P,Q) = 0$ if and only if $P=Q$, while for $\alpha=2$ 
-we have $\Energy_2(P,Q) = 2\| \E(X) - \E(Y) \|^2$ which shows that
-equality to zero only requires
-equality of the means, and thus $\Energy_2(P,Q)=0$ does 
-not imply equality of distributions.
-
-The energy distance can be even further generalized.
-Let $X, Y \in \mathcal{X}$  where $\mathcal{X}$ is an arbitrary space endowed
-with a \emph{semimetric of negative type}
-$\rho: \mathcal{X}\times\mathcal{X} \to \mathbb{R}$, which is required
-to satisfy
-\begin{equation}
-\label{eq:negative_type}
-\sum_{i,j=1}^n c_i c_j \rho(X_i, X_j) \le 0,
-\end{equation}
-where $X_i \in \mathcal{X}$ and $c_i \in \mathbb{R}$ such that
-$\sum_{i=1}^n c_i = 0$. Then, $\mathcal{X}$ is called a \emph{space of
-negative type}.
-We can thus replace $\mathbb{R}^D$ by $\mathcal{X}$ and 
-$\| X - Y \|$ by $\rho(X , Y)$ in the definition \eqref{eq:energy}, obtaining
-the \emph{generalized energy distance}
-\begin{equation}
-\label{eq:energy3}
-\Energy(P, Q) \equiv 2 \E \rho(X,Y) - \E \rho(X, X') - \E \rho(Y,Y').
-\end{equation}
-For spaces of negative type, there exists a Hilbert space $\mathcal{H}$ and
-a map $\varphi: \mathcal{X} \to
-\mathcal{H}$ such that
-$\rho(X, Y) = \| \varphi(X) - \varphi(Y) \|_{\mathcal{H}}^2$. This
-allows us to compute quantities related to probability distributions over
-$\mathcal{X}$ in the associated Hilbert space $\mathcal{H}$.
-Even though the semimetric 
-$\rho$ may not satisfy the triangle inequality, 
-$\rho^{1/2}$ does since it can be shown to be a proper metric. 
-Our energy clustering formulation, proposed in the next section,
-will be based on the generalized
-energy distance \eqref{eq:energy3}.
-
-There is an equivalence between energy distance, 
-commonly used in statistics,
-and distances between embeddings of distributions in 
-RKHS, commonly used in machine learning. 
-This equivalence was established
-in \cite{Sejdinovic2013}. Let us first recall the definition of
-RKHS. Let $\HH$ be a Hilbert space of real-valued functions
-over $\mathcal{X}$. A function 
-$\kk : \mathcal{X} \times \mathcal{X} \to 
-\mathbb{R}$ is a reproducing kernel of $\HH$ if it satisfies
-the following two conditions:
-\begin{enumerate}
-\item $\kkk_x \equiv \kk(\cdot, x) \in \HH$ 
-for all $x \in \mathcal{X}$;
-\item $\langle \kkk_x, f \rangle_{\HH} = f(x)$ for
-all $x\in\mathcal{X}$ and $f\in \HH$.
-\end{enumerate}
-In other words, for any $x \in \mathcal{X}$ and any function $f \in \HH$,
-there is a unique 
-$\kkk_x \in \HH$ that reproduces $f(x)$ through the inner product
-of $\HH$.
-If such a \emph{kernel} 
-function $\kk$ exists, then $\HH$ is called a RKHS. The above two 
-properties immediately imply that $\kk$ is symmetric and positive
-definite. 
-Defining the Gram matrix $G$ with
-elements $G_{ij} = \kk(x_i,x_j)$, this is equivalent to $G=G^\top$ being
-positive semidefinite, i.e., $v^\top G \, v \ge 0$ for any vector
-$v \in \mathbb{R}^n$.
-
-The Moore-Aronszajn theorem 
-\cite{Aronszajn}
-establishes the converse of the above paragraph.
-For every symmetric
-and positive definite function $\kk: \mathcal{X}\times \mathcal{X} \to
-\mathbb{R}$, there is an associated RKHS, $\Hk$,
-with reproducing
-kernel $\kk$. The map $\varphi: x \mapsto \kkk_x \in \Hk$ is called
-the canonical \emph{feature map}. Given a kernel $\kk$,
-this theorem enables us to define an embedding of a probability measure
-$P$ into the RKHS as follows: $P \mapsto \kkk_P \in
-\Hk$ such that 
-$\int f(x) d P(x) = \langle f, \kkk_P \rangle$ for all $f \in \Hk$,
-or alternatively, $\kkk_P \equiv \int \kk( \, \cdot \,, x)  d P(x)$. 
-We can now  introduce the 
-notion of distance between two probability measures using the inner product
-of $\Hk$, which is called the maximum mean discrepancy (MMD) and
-is given by
-\begin{equation}
-\label{eq:mmd}
-\gamma_\kk(P,Q) \equiv \| \kkk_P - \kkk_Q \|_{\Hk}.
-\end{equation}
-This can also be written as \cite{Gretton2012}
-\begin{equation}\label{eq:mmd2}
-\gamma_\kk^2(P,Q) = \E \kk(X,X') + \E \kk(Y,Y') - 2 \E \kk(X, Y)
-\end{equation}
-where $X,X' \stackrel{iid}{\sim} P$ and $Y,Y'\stackrel{iid}{\sim} Q$.
-From the equality between \eqref{eq:mmd} and \eqref{eq:mmd2} we also
-have $\langle \kkk_P, \kkk_Q \rangle_{\Hk} = \E \, \kk(X, Y)$.
-
-The following important result shows that semimetrics of negative
-type and symmetric positive definite kernels are closely related
-\cite{Berg1984}. Let $\rho: \mathcal{X} \times \mathcal{X} \to \mathbb{R}$
-and $x_0 \in \mathcal{X}$ an arbitrary but fixed point.
-Define
-\begin{equation}
-\label{eq:kernel_semimetric}
-\kk(x,y) \equiv 
-\tfrac{1}{2} \left[  \rho(x,x_0) + \rho(y,x_0) - \rho(x,y)\right].
-\end{equation}
-Then, it can be shown that 
-$\kk$ is positive definite if and only if $\rho$ is a semimetric
-of negative type.
-We have a family of kernels, one for each choice of $x_0$. Conversely,
-if $\rho$ is a semimetric of negative type and $\kk$ is a kernel in this
-family, then 
-\begin{equation}
-\label{eq:gen_kernel}
-\begin{split}
-\rho(x,y) &= \kk(x,x) + \kk(y,y) -2\kk(x,y) \\
-&=  \| \kkk_x - \kkk_y \|^2_{\Hk}
-\end{split}
-\end{equation}
-and the canonical feature map 
-$\varphi: x \mapsto \kkk_x$ is injective \cite{Sejdinovic2013}.
-When these conditions are satisfied, we say that the kernel $\kk$ 
-generates the semimetric $\rho$. 
-If two different kernels generate the same $\rho$, they are
-said to be equivalent kernels.
-
-Now we can state the equivalence between the generalized 
-energy distance \eqref{eq:energy3} and
-inner products on RKHS, which is one of the main results of
-\cite{Sejdinovic2013}. If $\rho$ is a semimetric
-of negative type and $\kk$ a kernel that generates $\rho$, then
-replacing \eqref{eq:gen_kernel} into
-\eqref{eq:energy3}, and using \eqref{eq:mmd2}, yields
-\begin{equation} \label{eq:Erho}
-\begin{split}
-\Energy(P, Q) &= 
-2 \left[ \E \, \kk(X, X') + \E \, \kk(Y, Y') - 2\E \, \kk(X, Y)\right]  \\
-&= 2 \gamma_\kk^2(P,Q) .
-\end{split}
-\end{equation}
-Due to \eqref{eq:mmd}, we can compute the energy 
-distance $\mathcal{E}(P, Q)$ between two probability distributions
-using the inner 
-product of $\Hk$.
-
-Finally, let us recall the main formulas from generalized energy statistics
-for the test statistic of equality of distributions \cite{Szkely2013}. 
-Assume that we have data $\mathbb{X} = \{ x_1,\dotsc, x_n \}$, where
-$x_i \in \mathcal{X}$, and $\mathcal{X}$ is a space of negative type.
-Consider a disjoint partition $\mathbb{X} = \bigcup_{j=1}^k \C_j$, with
-$\C_i \cap \C_j = \emptyset$.
-Each expectation in the generalized energy distance
-\eqref{eq:energy3}
-can be computed 
-through the function
-\begin{equation}
-\label{eq:g_def}
-g (\C_i, \C_j) \equiv 
-\dfrac{1}{n_i n_j}
-\sum_{x \in \C_i} 
-\sum_{y \in \C_j} \rho(x, y) ,
-\end{equation}
-where $n_i = |\C_i|$ is the number of elements in partition
-$\C_i$. 
-The \emph{within energy dispersion} is defined by
-\begin{equation}
-\label{eq:within}
-W \equiv
-\sum_{j=1}^{k} \dfrac{n_j}{2} g(\C_j, \C_j),
-\end{equation}
-and the \emph{between-sample energy statistic} is defined by
-\begin{equation}
-\label{eq:between}
-S \equiv \!\!\!\!
-\sum_{1 \le  i < j \le k } \dfrac{n_i n_{j}}{2 n} \left[
-2 g(\C_i, \C_j) - 
-g(\C_i, \C_i) - 
-g(\C_j, \C_j)
-\right],
-\end{equation}
-where $n = \sum_{j=1}^k n_j$.
-Given a set of distributions
-$\{ P_j\}_{j=1}^k$, where $x \in \C_j$ if and only if $x \sim P_j$, 
-the quantity $S$ provides
-a test statistic for equality of distributions
-\cite{Szkely2013}.
-When the sample size is large enough, $n\to \infty$,
-under the null hypothesis $H_0: P_1=P_2=\dotsm=P_k$, we have that
-$S\to 0$, 
-and under
-the alternative hypothesis $H_1: P_i \ne P_j$ for at least two $i\ne j$, 
-we have that $S \to \infty$.
-%Note that this test does not make any assumptions
-%about the form of distribution $P_j$, thus it is said to be 
-%distribution-free.
-
-%One can make a physical analogy by thinking 
-%that points $ x \in \C_j$ form a massive body 
-%whose total mass is characterized by the distribution function $P_j$.
-%The quantity $S$ is thus a potential
-%energy of the from $S(P_1,\dotsc,P_k)$ which measures how different
-%the distribution of these masses are,  and achieves the ground state
-%$S=0$ when all bodies have the same mass distribution. On the other hand,
-%the potential energy
-%$S$ increases as bodies have different mass distributions.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{The Clustering Problem Formulation}
-\label{sec:clustering_theory}
-
-This section contains our main theoretical results. First, we
-generalize the previous formulas from energy statistics by introducing
-weights associated to data points. 
-Second, we formulate an optimization problem for clustering 
-in the associated RKHS, making connection with kernel methods in
-machine learning.
-
-Let $w(x)$ be a weight function associated to point $x \in \mathcal{X}$
-and define
-\begin{equation}
-\label{eq:g_def2}
-g(\C_i, \C_j) \equiv \dfrac{1}{s_i s_j} \sum_{x\in \C_i}\sum_{y\in\C_j}
-w(x)w(y) \rho(x,y),
-\end{equation}
-where
-\begin{equation}
-s_i \equiv \sum_{x\in\C_i} w(x), \qquad
-s \equiv \sum_{j=1}^k s_j.
-\end{equation}
-The  weighted version of the within energy dispersion 
-and between-sample
-energy statistic are thus given by
-\begin{align}
-W &\equiv
-\sum_{j=1}^{k} \dfrac{s_j}{2} g(\C_j, \C_j), \label{eq:within2} \\
-S &\equiv \!\!\! \sum_{1 \le  i < j \le k } \!\!\! 
-\dfrac{s_i s_{j}}{2 s} \left[
-2 g(\C_i, \C_j) - 
-g(\C_i, \C_i) - 
-g(\C_j, \C_j)
-\right]. \label{eq:between2}
-\end{align}
-Note that if $w(x) = 1$ for every $x$ we recover the previous formulas.
-
-Due to the test statistic for equality of distributions,
-the obvious criterion for clustering data is to 
-maximize $S$ in \eqref{eq:between2},
-which makes  each cluster as different as possible from the other ones.
-In other words, given a set of points coming from different probability
-distributions, the test statistic $S$ should attain a maximum when 
-each point is correctly classified as belonging to the cluster associated 
-to its probability distribution. The following 
-result
-shows that maximizing $S$ is, however, equivalent to minimizing
-$W$ in \eqref{eq:within2}.% which has a more convenient form.
-
-\begin{lemma}
-\label{th:minimize}
-Let $\mathbb{X} = \{x_1,\dotsc,x_n\}$ where each data point
-$x_i$ lives in a space $\mathcal{X}$ endowed with a semimetric $\rho:
-\mathcal{X}\times\mathcal{X} \to \mathbb{R}$ of
-negative type. For a fixed integer $k$,
-the partition
-$\mathbb{X} = \bigcup_{j=1}^k \C^\star_j$, where 
-$\C^\star_i \cap C^\star_j = \emptyset$ for
-all $i\ne j$, maximizes the between-sample statistic $S$, defined
-in equation \eqref{eq:between2}, if and only if
-\begin{equation}
-\label{eq:minimize}
-\{ \C_1^\star,\dotsc,\C_k^\star \} = \argmin_{\C_1,\dotsc,C_k  } W(
-\C_1, \dotsc, \C_k) ,
-\end{equation}
-where the within energy dispersion $W$ is defined by \eqref{eq:within2}.
-\end{lemma}
-\begin{proof}
-From \eqref{eq:within2} and \eqref{eq:between2}
-we have that
-\begin{align}
-%\begin{equation}
-%\begin{split}
-& S + W  
-\nonumber \\
-&= \dfrac{1}{2s} \sum_{\substack{i,j=1 \\ i\ne j}}^k s_i s_j g(\C_i, \C_j)
-+ \dfrac{1}{2s} \sum_{i=1}^{k} 
-\bigg[ s - 
-\sum_{\substack{j= 1 \\ j\ne i}}^k s_j \bigg] 
-s_i g(\C_i, \C_i) 
-\nonumber \\
-& = \dfrac{1}{2s} \sum_{i,j=1}^k s_i s_j g(\C_i, \C_j) 
-\nonumber \\
-& = \dfrac{1}{2s} \sum_{x \in \mathbb{X}} \sum_{y \in \mathbb{X}} 
-w(x) w(y)\rho(x,y)  
-\nonumber \\
-&= \dfrac{s}{2} g(\mathbb{X}, \mathbb{X}).
-%\end{split}
-%\end{equation}
-\end{align}
-Since $g(\mathbb{X}, \mathbb{X})$ is independent 
-of the choice of partition, $\max_{\{ \C_i \}} S = - \max_{\{\C_i\}} W = 
-\min_{\{ \C_i \}}W$, as claimed.
-\end{proof}
-
-For a given $k$, the clustering problem amounts to
-finding the best partitioning of the data by minimizing $W$.
-In the current form of problem \eqref{eq:minimize},
-the relationship
-with other clustering methods or kernel spaces is totally obscure.
-In the following, we demonstrate what is the explicit 
-optimization problem behind 
-\eqref{eq:minimize} in the corresponding RKHS, 
-which establishes the connection with kernel methods.
-
-Based on the relationship between kernels and semimetrics of negative
-type, 
-assume that the kernel $\kk: \mathcal{X} \times \mathcal{X} \to \mathbb{R}$ 
-generates $\rho$.  Define  the Gram matrix
-\begin{equation}
-\label{eq:kernel_matrix}
-G \equiv \begin{pmatrix}
-\kk(x_1,x_1)     & \dotsm     & \kk(x_1,x_n) \\
-\vdots           & \ddots     & \vdots \\
-\kk(x_n,x_1)     & \dotsm     & \kk(x_n,x_n) 
-\end{pmatrix} .
-\end{equation}
-Let $Z \in \{ 0,1 \}^{n\times k}$ be the label matrix, 
-with only one nonvanishing entry per row, 
-indicating to which cluster (column)
-each point (row) belongs to. This matrix satisfies
-$Z^\top Z = D$, where the diagonal matrix 
-$D = \diag( n_1,\dotsc, n_k )$  contains
-the number of points in each cluster. We also introduce the rescaled
-matrix  $Y$ below. In component form they are 
-\begin{equation}
-\label{eq:label_matrix}
-Z_{ij} \equiv \begin{cases}
-1 & \mbox{if $x_i \in \C_j$ } \\
-0 & \mbox{otherwise}
-\end{cases}, \qquad
-\Zt_{ij} \equiv \begin{cases}
-\tfrac{1}{\sqrt{s_j}} & \mbox{if $x_i \in \C_j$ } \\
-0 & \mbox{otherwise}
-\end{cases} .
-\end{equation}
-Throughout the paper, we use the notation $M_{i\bullet}$ to denote
-the $i$th row of a matrix $M$, and $M_{\bullet j}$ denotes its $j$th column.
-We also define the following:
-\begin{equation}
-\label{eq:weighted_matrices}
-\mathcal{W} \equiv \diag(w_1,\dotsc,w_n), \quad
-H \equiv \mathcal{W}^{1/2} Y, \quad
-\om \equiv \mathcal{W} \e,
-\end{equation}
-where $w_i = w(x_i)$ is the weight associated to point $x_i$, 
-and $\e =(1,\dotsc,1)^\top \in \mathbb{R}^n$ 
-is the all-ones vector.
-
-Our next result shows that the optimization problem \eqref{eq:minimize}
-is NP-hard since it is a quadratically constrained quadratic program 
-(QCQP) in the associated RKHS.
-
-\begin{theorem}
-\label{th:qcqp3}
-The optimization 
-problem \eqref{eq:minimize} is equivalent to
-\begin{equation}
-\label{eq:qcqp3}
-\begin{split}
-&\max_H  \Tr \left[ H^\top \left(\mathcal{W}^{1/2} \, G \, 
-\mathcal{W}^{1/2}\right) H  \right] \\
-&\mbox{such that $H \ge 0$, $H^\top H = I$, $H H^\top \om = \om$,}
-\end{split}
-\end{equation}
-where $G$ is the Gram matrix \eqref{eq:kernel_matrix} 
-and the other quantities are defined in \eqref{eq:weighted_matrices}.
-\end{theorem}
-\begin{proof}
-From 
-\eqref{eq:gen_kernel},
-\eqref{eq:g_def2}, and
-\eqref{eq:within2}
-we have
-\begin{equation}
-\label{eq:W2}
-\begin{split}
-\!\! \!\!     W
-&= \sum_{j=1}^k \dfrac{1}{2 s_j} 
-\!\sum_{x,y \in \C_j} w(x)w(y)\rho(x,y) \\
-&= \! \sum_{j=1}^k   \sum_{x \in \C_j} \!\! \bigg[
-w(x)\kk(x,x) \!  - \! \dfrac{1}{s_j} \! \sum_{y \in \C_j} \!w(x)w(y) \kk(x,y) \bigg].
-\end{split}
-\end{equation}
-Note that the first term is global so it does not contribute to the 
-optimization problem.
-Therefore, problem \eqref{eq:minimize} becomes
-\begin{equation}
-\label{eq:max_prob}
-\max_{ \C_1,\dotsc,\C_k } 
-\sum_{j=1}^k \dfrac{1}{s_j} \sum_{x,y\in C_j} w(x) w(y) \kk(x,y) .
-\end{equation}
-Using the definitions 
-\eqref{eq:label_matrix} and 
-\eqref{eq:weighted_matrices}, the previous 
-objective function can be written as
-\begin{equation}
-\begin{split}
-& \hspace{-2em} \sum_{j=1}^k \dfrac{1}{s_j} 
-\sum_{p=1}^n \sum_{q=1}^n 
-w_p w_q Z_{pj} Z_{qj} G_{pq}  \\
-&= 
-\sum_{j=1}^k 
-\sum_{p=1}^n \sum_{q=1}^n 
-\dfrac{Z^\top_{jp}\sqrt{w_p}}{\sqrt{s_j}} w_p^{1/2} G_{pq} w_q^{1/2} 
-\dfrac{\sqrt{w_q} Z_{qj}}{\sqrt{s_j}} \\
-&= 
-\sum_{j=1}^k \left(H^\top \mathcal{W}^{1/2} G \mathcal{W}^{1/2} H\right)_{jj}
-\\
-&= \Tr\left[ H^\top \mathcal{W}^{1/2} \, G \, \mathcal{W}^{1/2} H  \right].
-\end{split}
-\end{equation}
-
-Now it remains to obtain the constraints. 
-Note that $H_{ij} \ge 0$ by definition, and
-\begin{equation}
-\begin{split}
-(H^\top H)_{ij} &= \sum_{\ell=1}^n 
-Y_{\ell i} \mathcal{W}_{\ell \ell} Y_{\ell j } \\
-&= 
-\dfrac{1}{\sqrt{s_i}\sqrt{s_j}} \sum_{\ell=1}^n w_\ell Z_{\ell i} Z_{\ell j}
-\\
-&= \dfrac{\delta_{ij}}{s_i} \sum_{\ell=1}^n w_\ell Z_{\ell i} 
-\\
-&= \delta_{ij}
-\end{split}
-\end{equation}
-where $\delta_{ij}=1$ if $i=j$ and $\delta_{ij}=0$ if $i\ne j$ is the
-Kronecker delta. Therefore, $H^\top H = I$. 
-This is a constraint on the rows of $H$.
-To obtain a constraint on its columns,
-observe that
-\begin{equation}
-\begin{split}
-\left(H^\top H\right)_{pq} &= \sqrt{w_p w_q}\sum_{j=1}^k \dfrac{Z_{pj}
-Z_{qj}}{s_j} \\
-& = \begin{cases}
-\dfrac{\sqrt{w_p w_q}}{s_i} & \mbox{if both $x_p,x_q \in \C_i$,} \\
-0 & \mbox{otherwise}.
-\end{cases}
-\end{split}
-\end{equation}
-Therefore, $(H^\top H \mathcal{W}^{1/2})_{pq} = \sqrt{w_p} \, w_q s_i^{-1}$
-if both points $x_p$ and $x_q$ belong to the same cluster, which
-we denote by $\C_i$ for some $i\in\{1,\dotsc,k\}$, and 
-$(H^\top H \mathcal{W}^{1/2})_{pq} = 0 $ otherwise. Thus, the $p$th
-line of this matrix is nonzero only on entries corresponding to points
-that are in the same cluster as $x_p$. If we sum over the columns of this
-line we obtain $\sqrt{w_p} s_i^{-1} \sum_{q=1}^n w_q Z_{qi} = \sqrt{w_p}$,
-or equivalently 
-\begin{equation}
-H H^\top \mathcal{W}^{1/2} \e = \mathcal{W}^{1/2} \e.
-\end{equation}
-From \eqref{eq:weighted_matrices} this
-gives $H H^\top \om = \om$, finishing the proof.
-\end{proof}
-
-The optimization problem \eqref{eq:qcqp3} is nonconvex, besides
-being NP-hard, thus a direct approach 
-is computationally prohibitive even for small datasets.
-However, one can find approximate solutions by relaxing some 
-of the constraints. For instance, consider the relaxed problem
-\begin{equation}
-\label{eq:relaxed}
-\max_{H} \Tr \left[ H^\top \widetilde{G} \, H \right] 
-\qquad \mbox{such that $H^\top H = I$},
-\end{equation}
-where $\widetilde{G} \equiv \mathcal{W}^{1/2} \, G \, \mathcal{W}^{1/2}$.
-This problem has a well-known closed form solution $H^\star = U R$, where the
-columns of $U \in \mathbb{R}^{n\times k}$ 
-contain the top $k$ eigenvectors of $\widetilde{G}$ corresponding
-to the $k$ largest eigenvalues, $\lambda_1\ge \lambda_2\ge\dotsc\ge\lambda_k$, 
-and $R \in \mathbb{R}^{k\times k}$ is an arbitrary orthogonal matrix. 
-The resulting optimal objective function assumes the value 
-$\max \Tr \big[ {H^\star}^\top \widetilde{G} \, H^\star \big]  = 
-\sum_{i=1}^k \lambda_i$. 
-Spectral clustering is based on this approach, where
-one further normalize the rows of $H^\star$, then cluster
-the resulting rows as data points using any clustering method such
-as k-means.
-A procedure on these lines was proposed in the seminal papers
-\cite{Malik,NgJordan}.
-
-\subsection{Connection with Graph Partitioning}
-
-We now show how
-graph partitioning problems are related to the energy statistics
-formulation leading to problem \eqref{eq:qcqp3}.
-
-Consider a graph $\mathcal{G} = (\mathcal{V}, \mathcal{E}, \mathcal{A})$,
-where $\mathcal{V}$ is the set of vertices, $\mathcal{E}$ the set of edges,
-and $\mathcal{A}$ is an affinity matrix  
-which measures the 
-similarities between pairs of nodes. Thus, $\mathcal{A}_{ij} \ne 0$
-if $(i,j) \in \mathcal{E}$, and $\mathcal{A}_{ij} = 0$ otherwise.
-We also associate weights to every vertex, 
-$w_i = w(i)$ for $i \in \mathcal{V}$, and let $s_j = \sum_{ i \in \C_j} w_i$,
-where $\C_j \subseteq \mathcal{V}$ is one partition of $\mathcal{V}$.
-Let
-\begin{equation}
-\textnormal{links}(\C_\ell, \C_m) \equiv  
-\!\sum_{\substack{i\in \C_\ell\\ j\in \C_m} } A_{ij} .
-\end{equation}
-Our goal is to partition the set of vertices $\mathcal{V}$ into $k$ disjoint
-subsets, $\mathcal{V} = \bigcup_{j=1}^k \C_j $. 
-The generalized ratio association problem is given by
-\begin{equation}
-\label{eq:assoc}
-\max_{\C_i,\dots,\C_k} \sum_{j=1}^k \dfrac{\textnormal{links}(\C_j,\C_j)}{s_j}
-\end{equation}
-and maximizes the within cluster association.
-The generalized ratio cut problem
-\begin{equation}
-\label{eq:cut}
-\min_{\C_i,\dots,\C_k} \sum_{j=1}^k
-\dfrac{\textnormal{links}(\C_j,\mathcal{V} \setminus \C_j)}{s_j}
-\end{equation}
-minimizes the cut between clusters. Both problems \eqref{eq:assoc}  and
-\eqref{eq:cut} are equivalent,
-in analogous way as minimizing \eqref{eq:within2} is equivalent to
-maximizing \eqref{eq:between2}, as shown in Lemma~\ref{th:minimize}.
-Here, this equivalence is  a consequence of the equality
-$\textnormal{links}(\C_j,\mathcal{V} \setminus  \C_j)=
-\textnormal{links}(\C_j,\mathcal{V}) - \textnormal{links}(\C_j,\C_j)$.
-Several graph partitioning methods 
-\cite{Kernighan,Malik,Chan,Yu}
-can be seen as a particular case of problems
-\eqref{eq:assoc} or \eqref{eq:cut}.
-
-Consider the ratio association problem \eqref{eq:assoc}, 
-whose objective function can be written as
-\begin{equation}
-\begin{split}
-\sum_{j=1}^k \dfrac{1}{s_j} \sum_{p \in \C_j} \sum_{q \in \C_j}
-\mathcal{A}_{pq} &= \sum_{j=1}^k \sum_{p=1}^n \sum_{q=1}^n 
-\dfrac{Z^\top_{jp}}{\sqrt{s_j}} \, \mathcal{A}_{pq} \, 
-\dfrac{Z_{qj}}{\sqrt{s_j}} \\
-&= \Tr\big[ Y^\top \mathcal{A} \, Y \big] ,
-\end{split}
-\end{equation}
-where we recall that $Z$ is defined in \eqref{eq:label_matrix} and $Y$ is
-defined in
-\eqref{eq:weighted_matrices}. Therefore, the ratio association problem
-can be written in the form \eqref{eq:qcqp3}, i.e.,
-\begin{equation}
-\begin{split}
-&\max_H \Tr \left[
-H^\top \mathcal{W}^{-1/2} \mathcal{A} \, \mathcal{W}^{-1/2} H 
-\right] \\ &\mbox{such that $H\ge 0$, $H^\top H = I$, $H H^\top
-\om=\om$}.
-\end{split}
-\end{equation}
-This is exactly the same as \eqref{eq:qcqp3}
-with  $G = \mathcal{W}^{-1} \mathcal{A} \mathcal{W}^{-1}$. Assuming that this
-matrix is positive semidefinite, this generates a semimetric
-\eqref{eq:gen_kernel} for graphs given by
-\begin{equation}
-\label{eq:metric_graphs}
-\rho(i,j) = 
-\dfrac{\mathcal{A}_{ii}}{w_i^{2}}
-+\dfrac{\mathcal{A}_{jj}}{w_j^{2}}
--\dfrac{2 \mathcal{A}_{ij}}{w_i w_j}
-\end{equation}
-for vertices $i,j \in \mathcal{V}$. If
-we assume the graph has no self-loops we must replace
-$\mathcal{A}_{ii} = 0$ above. 
-The weight of node $i\in \mathcal{V}$ can be, for instance,
-its degree $w_i = w(i) = d(i)$.
-
-\subsection{Connection with Kernel k-Means}
-\label{sec:kernel_kmeans}
-
-We now show that kernel k-means optimization problem \cite{Dhillon2,Dhillon}
-is also related 
-to the previous energy statistics formulation to clustering. To be precise,
-we consider a weighted generalization of kernel k-means.
-
-For a positive semidefinite Gram matrix $G$, as defined in
-\eqref{eq:kernel_matrix},
-there exists a map
-$\varphi: \mathcal{X} \to \HH_\kk$ such that
-\begin{equation}
-\kk(x,y) = \langle \varphi(x), \varphi(y) \rangle. 
-\end{equation}
-Define the weighted mean of cluster $\C_j$ as
-\begin{equation}
-\label{eq:muj}
-\mu_j = \dfrac{1}{s_j} \sum_{x \in \C_j} w(x) x.
-\end{equation}
-Disregarding the first global term in \eqref{eq:W2}, note
-that the second term,
-$-\tfrac{1}{s_j} \sum_{x,y \in \C_j} w(x) w(y) \kk(x,y)$,
-is equal to
-\begin{multline}
-\dfrac{1}{s_j^2}\sum_{x,y,z\in\C_j} 
-\langle w(y) \varphi(y),  w(z) \varphi(z) \rangle \\
--\dfrac{2}{s_j}\sum_{x,y\in\C_j} \langle w(x) \varphi(x),  
-w(y) \varphi(y) \rangle ,
-\end{multline}
-which using 
-\eqref{eq:muj}
-becomes
-\begin{multline}
-\sum_{x\in\C_j} 
-\big\{ \langle \varphi(\mu_j), \varphi(\mu_j) \rangle - 
-2 \langle w(x)\varphi(x),
-\varphi(\mu_j) \rangle \big\} \\ = 
-\sum_{x\in\C_j}\big\{ \| w(x) \varphi(x) - \varphi(\mu_j) \|^2 -
-\|w(x)\varphi(x)\|^2\big\}.
-\end{multline}
-Therefore, minimizing $W$ in \eqref{eq:W2} is equivalent to
-\begin{equation}
-\label{eq:kernel_kmeans}
-\min_{\C_1,\dotsc,\C_k}\bigg\{ 
-J(\{\C_j\}) \equiv  \sum_{j=1}^k
-\sum_{x \in \C_j} \| w(x) \varphi(x) - \varphi(\mu_j) \|^2
-\bigg\} .
-\end{equation}
-Problem \eqref{eq:kernel_kmeans} is obviously  equivalent to
-problem \eqref{eq:qcqp3}. 
-When $w(x) = 1$ for all
-$x$, problem \eqref{eq:kernel_kmeans} corresponds to 
-kernel k-means problem
-\cite{Dhillon2,Dhillon}. Thus, the result \eqref{eq:kernel_kmeans} shows
-that the previous energy statistics formulation to clustering
-is equivalent to a weighted
-modification of kernel k-means\footnote{
-One should not confuse kernel k-means optimization problem,
-given by \eqref{eq:kernel_kmeans}, with kernel k-means \emph{algorithm}.
-We will discuss two approaches to solve \eqref{eq:kernel_kmeans},
-or equivalently \eqref{eq:qcqp3}. One is 
-based on Lloyd's method, which leads to kernel k-means algorithm, and the other is based on Hartigan's method,
-which leads to a new algorithm.
-}. One must note, however, that energy statistics fixes the kernel
-through \eqref{eq:kernel_semimetric}.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Iterative Algorithms}
-\label{sec:algo}
-
-%In this section we introduce an iterative algorithm to find a local
-%maximizer of the optimization problem \eqref{eq:qcqp3}. Due to 
-%the discussion of Section~\ref{sec:kernel_kmeans} 
-%we can also find an approximate
-%solution by kernel k-means algorithm, which is based
-%on Lloyd's heuristic \cite{Dhillon2,Dhillon}. For
-%convenience, such an algorithm will also be stated in the present context.
-
-In this section, we introduce two iterative algorithms to solve
-the optimization problem \eqref{eq:qcqp3}. The first is based on Lloyd's
-method, while the second is based on Hartigan's method.
-
-Consider the optimization problem 
-\eqref{eq:max_prob} 
-written as
-\begin{equation}
-\label{eq:maxQ}
-\! \max_{\{ \C_1,\dotsc,\C_k \}}  \!
-\bigg\{ Q =  \sum_{j=1}^k \dfrac{Q_j}{s_j}  \bigg\},
-\ Q_j \equiv \!\!\!\sum_{x,y\in\C_j} \!\!\! w(x) w(y)\kk(x,y),
-\end{equation}
-where $Q_j$ represents an internal cost of cluster $\C_j$, and
-$Q$ is the total cost where each $Q_j$ 
-is weighted by the inverse
-of the sum of weights of the points in $\C_j$. 
-For a data point $x_i$, we denote
-its cost
-with cluster $\C_\ell$ by
-\begin{equation}
-\label{eq:costxij}
-Q_\ell(x_i) \equiv \sum_{y\in\C_\ell} w(x_i) w(y) \kk(x_i, y) = 
-(\mathcal{W} G \mathcal{W})_{i \bullet} \cdot Z_{\bullet \ell},
-\end{equation}
-where we recall that $M_{i\bullet}$ ($M_{\bullet i}$) denotes
-the $i$th row (column) of matrix $M$.
-
-\subsection{Weighted Kernel k-Means Algorithm}
-
-Using the definitions \eqref{eq:maxQ} and \eqref{eq:costxij}, the
-the optimization problem \eqref{eq:kernel_kmeans} can be written as 
-\begin{equation}
-\label{eq:min_lloyd}
-\min_{Z} \sum_{i = 1}^{n} \sum_{\ell=1}^{k} Z_{i\ell} J^{(\ell)}(x_i)
-\end{equation}
-where
-\begin{equation}
-\label{eq:Jell}
-J^{(\ell)}(x_i) \equiv 
-\dfrac{1}{s_\ell^2} Q_\ell
--\dfrac{2}{s_\ell} Q_\ell(x_i) .
-\end{equation}
-A possible strategy to solve \eqref{eq:min_lloyd} is to
-assign  $x_i$ to cluster $\C_{j^\star}$ according
-to 
-\begin{equation}
-\label{eq:min_lloyd2}
-j^\star = \argmin_{\ell=1,\dotsc,k} J^{(\ell)}(x_i) .
-\end{equation}
-This should be done for every data point $x_i$ and repeated until
-convergence, i.e., until no new assignments are made.
-The entire procedure is described in Algorithm~\ref{kmeans_algo}.
-It can be shown that this algorithm converges when $G$ is positive
-semidefinite.
-
-\begin{algorithm}[t]
-\begin{algorithmic}[1]
-    \INPUT $k$, $G$, $\mathcal{W}$, $Z \leftarrow Z_0$
-    \OUTPUT $Z$ 
-  \STATE $q \leftarrow (Q_1, \dotsc, Q_k)^\top$ 
-            \hfill (see \eqref{eq:maxQ})
-  \STATE $s \leftarrow (s_1,\dotsc,s_k)^\top$ 
-  \REPEAT
-    \FOR{ $i=1,\dotsc,n$}
-        \STATE let $j$ be such that $x_i \in \C_j$
-        \STATE $j^\star \leftarrow \argmin_{\ell=1,\dotsc,k} J^{(\ell)}(x_i)$
-            \hfill (see \eqref{eq:Jell})
-        \IF{ $j^\star \ne j$} 
-            \STATE $Z_{ij} \leftarrow 0$ 
-            \STATE $Z_{ij^\star} \leftarrow 1$
-            \STATE $s_j \leftarrow s_j - \mathcal{W}_{ii}$ 
-            \STATE $s_{j^\star} \leftarrow s_{j^\star} + \mathcal{W}_{ii}$
-            \STATE $q_j \leftarrow q_j - 2Q_j(x_i)$ 
-            \hfill (see \eqref{eq:costxij})
-            \STATE $q_{j^\star} \leftarrow q_{j^\star} + 2Q_{j^\star}(x_i)$
-            \hfill (see \eqref{eq:costxij})
-    %    \ELSE
-    %        \STATE Do nothing;
-        \ENDIF
-    \ENDFOR
-  \UNTIL{convergence}
-\end{algorithmic}
-\caption{\label{kmeans_algo}
-Weighted version of kernel k-means algorithm
-to find local solutions to the optimization problem 
-\eqref{eq:qcqp3}.
-}
-\end{algorithm}
-
-To see that the above procedure is indeed kernel k-means
-\cite{Dhillon2,Dhillon}, based
-on Lloyd's heuristic \cite{Lloyd}, note that from \eqref{eq:kernel_kmeans} 
-and \eqref{eq:Jell} we have
-\begin{equation}
-\label{eq:closest_mean}
-\min_\ell J^{(\ell)}(x_i) = 
-\min_\ell \| w(x_i) \varphi(x_i) - \varphi(\mu_\ell)\|^2.
-\end{equation}
-Therefore, we are 
-assigning $x_i$ to the cluster with closest center, in the
-feature space.
-When $w(x) = 1$ for all $x$, the above method  is exactly 
-kernel k-means algorithm.
-
-To check the complexity of Algorithm~\ref{kmeans_algo}, 
-note that the second term %of $J^{(\ell)}(x_i)$
-in \eqref{eq:Jell} requires
-$\OO(n_\ell)$ operations, and although the first term requires
-$\OO(n_\ell^2)$ it only needs to be computed once outside loop through
-data points (step 1).
-Therefore, the time complexity of Algorithm~\ref{kmeans_algo}
-is $\OO(n k \max_\ell n_\ell) = \OO(k n^2)$. For a sparse
-Gram matrix $G$, having
-$\tilde{n}$ nonzero elements, this can be further reduced
-to $\OO(k \tilde{n})$.
-
-\subsection{Kernel k-Groups Algorithm}
-
-\begin{algorithm}[t]
-\begin{algorithmic}[1]
-    \INPUT $k$, $G$, $\W$, $Z \leftarrow Z_0$
-    \OUTPUT $Z$
-  \STATE $q \leftarrow (Q_1, \dotsc, Q_k)^\top$ 
-            \hfill (see \eqref{eq:maxQ})
-  \STATE $s \leftarrow (s_1,\dotsc,s_k)^\top$ 
-  \REPEAT
-    \FOR{ $i=1,\dotsc,n$}
-        \STATE let $j$ be such that $x_i \in \C_j$
-        \STATE $j^\star \leftarrow \argmax_{\ell=1,\dotsc,k \, | \, \ell\ne j} 
-                \Delta Q^{j\to \ell}(x_i)$
-            \hfill (see \eqref{eq:changeQ}) \label{stepmove}
-        \IF{ $\Delta Q^{j \to j^\star}(x_i) > 0$ }
-            \STATE $Z_{ij} \leftarrow 0$
-            \STATE $Z_{ij^\star} \leftarrow 1$
-            \STATE $s_j \leftarrow s_j - \mathcal{W}_{ii}$ 
-            \STATE $s_{j^\star} \leftarrow s_{j^\star} + \mathcal{W}_{ii}$
-            \STATE $q_j \leftarrow q_j - 2Q_j(x_i) + (\W G \W)_{ii}$
-                \hfill (see \eqref{eq:costxij})
-            \STATE 
-            $q_{j^\star} \leftarrow q_{j^\star} + 2Q_{j^\star}(x_i)+ 
-                (\W G\W)_{ii}$
-                \hfill (see \eqref{eq:costxij})
-    %    \ELSE
-    %        \STATE Do nothing;
-        \ENDIF
-    \ENDFOR
-  \UNTIL{convergence}
-\end{algorithmic}
-\caption{\label{algo}
-Kernel k-groups algorithm, based on Hartigan's method, to find
-local solutions to problem \eqref{eq:qcqp3}.
-%Steps $6$, $12$ and $13$ are different
-%compared to Algorithm~\ref{kmeans_algo}. 
-}
-\end{algorithm}
-
-We now consider Hartigan's method \cite{Hartigan1975,Hartigan1979} 
-applied to the optimization problem in the form \eqref{eq:maxQ}, which gives
-a local solution to \eqref{eq:qcqp3}. 
-The method is based in computing the maximum change
-in the total cost function $Q$ when moving each data point to
-another cluster. More specifically, 
-suppose that point $x_i$
-is currently assigned to  cluster $\C_j$ yielding
-a total cost function denoted by $Q^{(j)}$.
-Moving $x_i$ to cluster $\C_\ell$ yields another total cost function
-denoted by $Q^{(\ell)}$. We are interested in computing the maximum 
-change
-$\Delta Q^{(j\to \ell)} (x_i) \equiv Q^{(\ell)} - Q^{(j)}$, for $\ell\ne j$. 
-From \eqref{eq:maxQ}, by explicitly writing the costs related to these 
-two cluster we obtain
-\begin{equation}
-\label{eq:deltaQ}
-\Delta Q^{(j\to \ell)} (x_i) = \dfrac{Q_\ell^{+}}{s_\ell+w_i} + 
-\dfrac{Q_j^-}{s_j-w_i} - \dfrac{Q_\ell}{s_\ell} - \dfrac{Q_j}{s_j}  ,
-\end{equation}
-where $Q^{+}_\ell$ denote the cost of the new $\ell$th cluster
-with the point $x_i$ added to it, and $Q^-_j$ is the cost of new 
-$j$th cluster with $x_i$ removed from it. Recall also that $w_i=w(x_i)$
-is the weight associated to point $x_i$. Noting that 
-\begin{align}
-Q_\ell^{+} &= Q_\ell + 2 Q_\ell(x_i) + (\W G \W)_{ii}, 
-\label{eq:Qplus} \\
-Q_j^{-} &= Q_j - 2 Q_j(x_i) + (\W G \W)_{ii}, 
-\label{eq:Qminus}
-\end{align}
-we obtain
-\begin{multline}
-\label{eq:changeQ}
-\Delta Q^{(j \to \ell)}(x_i)  = 
-\dfrac{1}{s_j - w_i}  \left[ \dfrac{w_i}{s_j}Q_j 
-- 2 Q_j(x_i) + (\W G \W)_{ii} \right] \\
-- \dfrac{1}{s_\ell + w_i}\left[ \dfrac{w_i}{s_\ell}Q_\ell - 2 Q_\ell(x_i) 
-- (\W G \W )_{ii} \right].
-\end{multline}
-Therefore, we compute
-\begin{equation}
-j^\star = \argmax_{\ell=1,\dotsc,k \, | \, \ell\ne j} 
-\Delta Q^{(j \to \ell)}(x_i)
-\end{equation}
-and if $\Delta Q^{j \to j^\star}(x_i) > 0$ 
-we move $x_i$ to cluster $\C_{j^\star}$, otherwise 
-we keep $x_i$ in its original cluster $\C_j$. 
-This process is repeated
-until no points are assigned to new clusters. 
-The entire procedure is described in Algorithm~\ref{algo}, which we
-call kernel k-groups. This method is a generalization of the k-groups
-with first variations proposed in \cite{Kgroups}, which only considers
-the Euclidean case, and also an extension
-of Hartigan's method to kernel spaces.
-
-Note that Algorithm~\ref{algo} automatically ensures that the 
-objective function is
-monotonically increasing at each iteration, and consequently the algorithm
-converges in a finite number of steps.
-
-The complexity analysis of Algorithm~\ref{algo} is the following.
-%Computing the Gram matrix $G$ requires $\OO( D n^2)$ operations, where 
-%$D$ is the dimension of each data point and $n$ is the data size. However,
-%both algorithms $\mathcal{E}^L$- and $\mathcal{E}^H$-clustering 
-%assume that $G$ is given. There are more efficient
-%methods to compute $G$, specially if it is sparse, but we will not consider
-%this further and just assume that $G$ is given.
-The computation of each cluster cost
-$Q_j$ has complexity $\OO(n_j^2)$, and overall to compute $q$
-we have $\OO(n_1^2+\dots + n_k^2) = \OO(k \max_j n_j^2)$. 
-These operations only need to be performed a single time. For
-each point $x_i$ we need to compute $Q_j(x_i)$ once, which is
-$\OO(n_j)$, and we need to compute $Q_\ell(x_i)$ for each $\ell\ne j$. 
-The cost of computing 
-$Q_\ell(x_i)$
-is $\OO(n_\ell)$, thus the cost of step~$6$ in
-Algorithm~\ref{algo} is $\OO(k \max_\ell n_\ell)$ for $\ell=1,\dotsc,k$.
-For the 
-entire dataset this gives a complexity
-of $\OO(n k  \max_\ell n_\ell) =\OO(k n^2)$. Note that this is the same cost as
-in kernel k-means algorithm. 
-Again, if $G$ is sparse
-this can be reduced to $\OO(k \tilde{n})$ where $\tilde{n}$ 
-is the number of nonzero
-entries of $G$.
-
-%In the following we mention some important known 
-%results about Hartigan's method.
-%
-%\begin{theorem}[Telgarsky-Vattani \cite{Telgarsky}]
-%Hartigan's method has the cost function strictly decreasing in each
-%iteration. Moreover, if $n > k$ then 
-%\begin{enumerate}
-%\item \label{noempty} the resulting partition has no empty clusters, and
-%\item \label{diffmean} the resulting partition has distinct means.
-%\end{enumerate}
-%\end{theorem}
-%
-%Neither of these two conditions are guaranteed to be 
-%satisfied  by Lloyd's method,
-%and consequently by $\mathcal{E}^L$-clustering 
-%algorithm.
-%The next result indicates that Hartigan's method can potentially 
-%escape local optima of Lloyd's method.
-%
-%\begin{theorem}[Telgarsky-Vattani \cite{Telgarsky}]
-%The set of local optima of Hartigan's method is a (possibly strict) subset
-%of local optima of Lloyd's method.
-%\end{theorem}
-%
-%The above theorem implies that $\mathcal{E}^L$-clustering  cannot
-%improve on a local optima of $\mathcal{E}^H$-clustering. On the other hand,
-%$\mathcal{E}^H$ might improve on a local optima of 
-%$\mathcal{E}^L$. Lloyd's method forms Voronoi partitions,
-%while Hartigan's method groups data
-%in regions formed by the intersection of spheres called circlonoi cells.
-%It can be shown that the circlonoi cells are contained within
-%a smaller volume of a Voronoi cell, and this excess volume grows
-%exponentially with the dimension of $\mathcal{X}$ 
-%\cite[Theorems 2.4 and 3.1]{Telgarsky}. 
-%Points in this excess volume
-%force Hartigan's method to iterate, contrary
-%to Lloyd's method. Therefore, Hartigan's 
-%can escape local
-%optima of Lloyd's. 
-%Moreover, this improvement should be more prominent as
-%dimension increases. Also, the improvement grows as the number of clusters
-%$k$ increases.
-%The empirical results of \cite{Telgarsky} show that 
-%an implementation of Hartigan's method has comparable execution time 
-%to an implementation of
-%Lloyd's method,
-%but no explicit complexity was provided. We show that both
-%$\mathcal{E}^L$- and $\mathcal{E}^H$-clustering
-%have the same time complexity. To the best of our knowledge, Hartigan's
-%method was not previously considered together with kernels, as we 
-%are proposing
-%in $\mathcal{E}^{H}$-clustering algorithm.
-%
-%In \cite{Slonin}, Hartigan's method was applied to k-means problem
-%with any Bregman divergence. It was shown that the number of Hartigan's
-%local optima is upper bounded by $\mathcal{O}(1/k)$ 
-%\cite[Proposition 5.1]{Slonin}. 
-%In addition, it was provided examples where
-%\emph{any} initial partition correspond to a local optima of Lloyd's 
-%method, while  the number of local optima in Hartigan's method is small and 
-%correspond to true partitions of the data. Empirically, the number of
-%Hartigan's local optima was considerably smaller than the number of Lloyd's
-%local optima.
-%
-%The above results indicate that Hartigan's method
-%provides several advantages over Lloyd's method, a fact that will also
-%be supported by our numerical experiments in the next section where 
-%$\mathcal{E}^H$  outperforms
-%of $\mathcal{E}^L$ (kernel k-means) in several settings,
-%specially in high dimensions.
-
-\section{Numerical Experiments}
-\label{sec:numerics}
-
-The main goal of this section is twofold. 
-First, to illustrate that in Euclidean spaces with the standard
-metric of energy statistics, as defined by the energy 
-distance \eqref{eq:energy}, the
-clustering method implemented by kernel k-groups
-is more flexible and in general more accurate than
-k-means and GMM.
-Second, we want to compare kernel k-groups with kernel k-means
-and spectral clustering when these methods
-operate on the same kernel.
-
-%The following experimental setup holds unless specified otherwise. 
-We consider
-the metrics
-\begin{align}
-\rho_{\alpha}(x,y) &= \| x-y \|^{\alpha}, 
-%\kk_{\alpha}(x,y) = \tfrac{1}{2} \left( \| x \|^{\alpha} + \| y \|^{\alpha} 
-%- \| x-y \|^{\alpha} \right), 
-\label{eq:rho_alpha} \\
-%
-\widetilde{\rho}_{\sigma}(x,y) &= 2 - 2 e^{-\tfrac{\|x-y\|}{2 \sigma}},  
-%\widetilde{\kk}_{\sigma}(x,y) = e^{-\tfrac{\|x-y\|}{2\sigma}}, 
-\label{eq:rho_tilde}\\
-%
-\widehat{\rho}_{\sigma}(x,y) &= 2 - 2 e^{-\tfrac{\|x-y\|^2}{2 \sigma^2}},
-%\widehat{\kk}_{\sigma}(x,y) = e^{-\tfrac{\|x-y\|^2}{2\sigma^2}}  
-\label{eq:rho_hat}
-\end{align}
-which define the 
-corresponding kernels through 
-\eqref{eq:kernel_semimetric}, where we always fix $x_0=0$.
-We use $\rho_\alpha$ by default, unless otherwise specified.
-%will always be present in the experiments as a reference.
-We consider the weights associated to data points to be $w(x) = 1$ for
-all $x$, so that $\W = I$ in 
-Algorithms~\ref{kmeans_algo}~and~\ref{algo}.
-For k-means, GMM and spectral clustering we use the
-implementations of \emph{scikit-learn} library \cite{scikit-learn}, where  
-k-means is initialized with 
-k-means++ \cite{Vassilvitskii}, 
-and GMM is initialized with the output of k-means, making 
-it more robust and preventing
-it from breaking in high dimensions. 
-The spectral clustering implementation of scikit-learn 
-is based on \cite{Malik}. 
-Kernel k-means is implemented according to Algorithm~\ref{kmeans_algo} while
-kernel k-groups follows Algorithm~\ref{algo}.
-Both will also be initialized with k-means++, unless specified otherwise.
-We run every algorithms $5$ times with different initializations and then
-we choose the result with the best objective function value. 
-We evaluate the clustering quality by
-the accuracy defined as
-\begin{equation}
-\label{eq:accuracy}
-\textnormal{accuracy}(\hat{Z}) \equiv \max_\pi
-\dfrac{1}{n}\sum_{i=1}^n\sum_{j=1}^k \hat{Z}_{i \pi(j)} Z_{ij} ,
-\end{equation}
-where $\hat{Z}$ is the predicted label matrix, $Z$
-is the ground truth label matrix, 
-and $\pi$ is a permutation of $\{1,2,\dotsc,k\}$. 
-Thus, the accuracy corresponds to the fraction
-of correctly classified data points, and it is always between $[0,1]$.
-For each setting, we show the average accuracy over $100$ Monte
-Carlo trials, with error bars indicating the standard error.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Synthetic Experiments}
-
-\begin{figure*}
-\centering
-\begin{minipage}{.4\textwidth}
-\includegraphics[width=1\textwidth]{gauss_1d.pdf}\\[-2em]
-(a)
-\end{minipage}
-\begin{minipage}{.4\textwidth}
-\includegraphics[width=1\textwidth]{loggauss_1d.pdf} \\[-2em]
-(b)
-\end{minipage}\\[1em]
-\begin{minipage}{.37\textwidth}
-\includegraphics[width=1\textwidth]{normal_density.pdf}\\[-2em]
-(c)
-\end{minipage}\hspace{2em}
-\begin{minipage}{.37\textwidth}
-\includegraphics[width=1\textwidth]{lognormal_density.pdf}\\[-2em]
-(d)
-\end{minipage}
-\caption{
-\label{fig:1dexperiment}
-Clustering one-dimensional data for a two-class problem.
-(a) $x \stackrel{iid}{\sim} \tfrac{1}{2} \mathcal{N}(0, 1.5) +
-\tfrac{1}{2}\mathcal{N}(1.5,0.3)$. (b) $x \stackrel{iid}{\sim}
-\tfrac{1}{2}e^{\mathcal{N}(0, 1.5)} + \tfrac{1}{2} e^{\mathcal{N}(1.5, 0.3)}$.
-In both cases we plot the average accuracy versus the total number of points.
-We ommit error bars since they are too small.
-(c) Kernel density estimation for a single trial with 2000 sampled
-points from the
-distribution in the case (a). (d) Kernel density estimation in a single
-trial with 
-2000 points sampled from the distribution in the case (b).
-}
-\end{figure*}
-
-We first consider one-dimensional data for a two-class problem. We compare
-kernel k-groups with k-means and GMM, as illustrated in 
-Fig.~\ref{fig:1dexperiment}. The left panels show a mixture of Gaussians, and the right panels show a mixture of log Gaussians (see caption for details).
-% have two mixtures of normal and lognormal distributions. 
-Notice that in the kernel density estimation plots
-for lognormal distribution, only kernel k-groups was able to distinguish
-between the two classes. The accuracy results for both density estimation
-cases are in Table~\ref{tb:dens}. We remark that kernel k-groups in
-this one-dimensional example performed the same as the exact deterministic
-Algorithm~\ref{algo1d} introduced in the Appendix.
-
-\begin{table}[h]
-\centering
-\caption{
-\label{tb:dens} Accuracy results for the density
-estimation of Fig.~\ref{fig:1dexperiment}d--e.
-}
-\begin{tabular}{@{}l|ll@{}}
-       Method        & normal & lognormal \\ \midrule[.5pt]
-        kmeans       & 0.773  & 0.502     \\
-         GMM         & \textbf{0.881}  & 0.513     \\
-   kernel k-groups   & 0.802  & \textbf{0.847}  
-\end{tabular}
-\end{table}
-
-Next, we analyze
-how the algorithms degrade as the number of dimensions increase.
-Consider data from the Gaussian mixture
-\begin{equation}
-\label{eq:gauss1}
-\begin{split}
-x  &\stackrel{iid}{\sim} 
-\tfrac{1}{2} \mathcal{N}(\mu_1,\Sigma_1) +
-\tfrac{1}{2} \mathcal{N}(\mu_2,\Sigma_2), \quad
-\Sigma_1=\Sigma_2 = I_D, \\
-\mu_1 &= (\underbrace{0,\dotsc,0}_{\times D})^\top, \qquad
-\mu_2 = 0.7 (\underbrace{1,\dots,1}_{\times 10},
-\underbrace{0,\dots,0}_{\times (D-10)})^\top.
-\end{split}
-\end{equation}
-The Bayes error is fixed as $D$ increases giving an optimal accuracy
-of $\approx 0.86$.
-We sample $200$ points on each trial. A scatter plot of the last two
-dimensions that contains signal in $\mu_2$ is shown in
-Fig.~\ref{fig:scatter}a.
-The clustering results are shown in Fig.~\ref{fig:plots}a.
-We see that 
-kernel k-groups and spectral clustering have close
-performance, being superior to kernel k-means, k-means, and GMM.
-The improvement is noticeable in 
-higher dimensions.
-
-\begin{figure*}
-\centering
-\includegraphics[width=0.19\textwidth]{gauss_means_scatter.pdf}
-\includegraphics[width=0.19\textwidth]{gauss_cov_scatter.pdf}
-\includegraphics[width=0.19\textwidth]{gauss_diff_metrics.pdf}
-\includegraphics[width=0.19\textwidth]{loggauss_diff_metrics.pdf}
-\includegraphics[width=0.19\textwidth]{unbalanced_scatter.pdf}
-\put(-500,-8){(a)}
-\put(-398,-8){(b)}
-\put(-298,-8){(c)}
-\put(-195,-8){(d)}
-\put(-95,-8){(e)}
-\caption{
-\label{fig:scatter}
-Scatter plot of the last two dimensions where $\mu_2$ has signal.
-Each plot has 200 points total.
-(a) Data distributed as in \eqref{eq:gauss1}.
-(b) Data distributed as in \eqref{eq:gauss2}.
-(c) Data distributed as in \eqref{eq:20gauss}.
-(d) Parameters as in \eqref{eq:20gauss} but for lognormal mixture.
-(e) Data from \eqref{eq:gauss3} with $N = 300$ and $m=200$.
-}
-\end{figure*}
-
-\begin{figure*}
-\centering
-\includegraphics[width=0.33\textwidth]{gauss1.pdf}
-\includegraphics[width=0.33\textwidth]{gauss2.pdf}
-\includegraphics[width=0.33\textwidth]{gauss_n.pdf}
-\includegraphics[width=0.33\textwidth]{loggauss_n.pdf}
-\includegraphics[width=0.315\textwidth]{gauss_unbal.pdf}
-\put(-400,130){(a)}
-\put(-230,130){(b)}
-\put(-50,130){(c)}
-\put(-320,-8){(d)}
-\put(-140,-8){(e)}
-\caption{
-\label{fig:plots}
-Clustering results associated to the data illustrated in 
-Fig.~\ref{fig:scatter}. For each experiment we
-perform 100 Monte Carlo runs and show the average accuracy. 
-We ommit error bars since they are too small to be visible.
-(a)
-High dimensional Gaussian mixture according to \eqref{eq:gauss1}.
-The dashed line is
-Bayes accuracy $\approx 0.86$. We use the metric $\rho_1$ in
-\eqref{eq:rho_alpha}, which is standard in energy statistics.
-(b)
-High dimensional Gaussian mixture according to \eqref{eq:gauss2}.
-Bayes accuracy $\approx 0.95$. We use $\rho_1$ in \eqref{eq:rho_alpha}.
-(c)
-Gaussian mixture with parameters \eqref{eq:20gauss}. We increase
-the number of sampled points in each trial. We use different
-metrics; see \eqref{eq:rho_alpha}--\eqref{eq:rho_hat}. 
-Here, kernel k-groups is more accurate than spectral clustering.
-(d)
-Same experiment as in Fig.~\ref{fig:plots}c but with a 
-lognormal mixture with parameters \eqref{eq:20gauss}.
-Again, kernel k-groups is more accurate than alternatives.
-The plot suggests that neither of these methods are
-consistent on this example since Bayes accuracy is $\approx 0.90$.
-(e)
-Comparison between clustering methods
-on unbalanced clusters. The data is normally 
-distributed as \eqref{eq:gauss3} where
-we vary $m \in [0, 240]$. We use the standard metric $\rho_1$ (see
-\eqref{eq:rho_alpha}) from energy statistics.
-}
-\end{figure*}
-
-Still for a two-class Gaussian mixture as in \eqref{eq:gauss1}, we now choose
-different numbers for the  diagonal covariance $\Sigma_2$.
-We have $\Sigma_1=I_D$, $\mu_1=(0,\dotsc,0)^\top \in \mathbb{R}^D$,
-$\mu_2=(1,\dotsc,1,0,\dotsc,0)^T \in \mathbb{R}^D$, 
-with signal in the first $10$ dimensions, and
-\begin{equation}
-\label{eq:gauss2}
-\begin{split}
-\Sigma_2 &= \left( \begin{array}{c|c}
-\widetilde{\Sigma}_{10} & 0 \\ \hline 
-0 & I_{D-10} \end{array}\right), \\
-\widetilde{\Sigma}_{10} &= \diag(1.367,  3.175,  3.247,  4.403,  1.249,\\
-&\hspace{3.6em}1.969, 4.035,   4.237,  2.813,  3.637).
-\end{split}
-\end{equation}
-We simply chose $10$ numbers uniformly at random on the interval
-$[1,5]$ and other choice would give analogous results.
-Bayes  accuracy is fixed at $\approx 0.95$. In Fig~\ref{fig:scatter}b
-we show a scatter plot of the 9th and 10th dimension.
-From Fig.~\ref{fig:plots}b we see that 
-all the methods are similarly accurate in low dimensions, but they quickly
-degenerate as the number of dimensions increase, except 
-kernel k-groups which is much more stable.
-
-Now, consider
-$x \stackrel{iid}{\sim} \tfrac{1}{2} \mathcal{N}(\mu_1,\Sigma_1)+
-\tfrac{1}{2} \mathcal{N}(\mu_2,\Sigma_2)$
-with 
-\begin{equation}
-\label{eq:20gauss}
-\begin{split}
-2\Sigma_1 &= \Sigma_2 = I_{20} \\
-\mu_1 &= (\underbrace{0,\dotsc,0}_{\times 20})^\top , 
-\quad \mu_2 = \tfrac{1}{2} 
-(\underbrace{1,\dotsc,1}_{5},\underbrace{0,\dotsc,0}_{15})^\top.
-\end{split}
-\end{equation}
-Bayes accuracy is $\approx 0.90$.  A scatter plot of the 4th and 5th
-dimensions is shown in Fig.~\ref{fig:scatter}c.
-We increase the sample size $n \in [10, 400]$ and show
-the accuracy versus $n$ in Fig.~\ref{fig:plots}c. We compare
-kernel k-groups, with different metrics,
-to k-means and GMM. We also use
-the best metric in this example for spectral clustering.
-We notice a superior performance of kernel k-groups 
-compared to the other methods.
-
-To consider non-normal data, we sample from the lognormal mixture 
-$x \stackrel{iid}{\sim} \tfrac{1}{2} 
-e^{\mathcal{N}(\mu_1,\Sigma_1)}+
-\tfrac{1}{2} e^{\mathcal{N}(\mu_2,\Sigma_2)}$ with the same
-parameters as in \eqref{eq:20gauss}. 
-The optimal Bayes accuracy is still $\approx 0.9$.
-A scatter plot is in Fig.~\ref{fig:scatter}d and the
-results are shown in Fig.~\ref{fig:plots}d.
-We use exactly the same metrics as in the normal
-mixture of Fig.~\ref{fig:plots}c to illustrate that the 
-proposed method still performs accurately.
-
-Finally, we show a limitation of kernel k-groups, which is shared
-between 
-all the other methods except for GMM.
-For highly unbalanced clusters, k-means, spectral
-clustering, kernel k-means and kernel k-groups all degenerate
-more quickly than GMM. A scatter plot of the first two dimensions
-is shown in Fig.~\ref{fig:scatter}e and the clustering results are
-in Fig.~\ref{fig:plots}e,
-where we generate data according to
-\begin{equation}
-\label{eq:gauss3}
-\begin{split}
-x &\stackrel{iid} \sim  
-\dfrac{n_1}{2N} \mathcal{N}(\mu_1,\Sigma_1)+
-\dfrac{n_1}{2N} \mathcal{N}(\mu_1,\Sigma_1), \\
-\mu_1 &= (0,0,0,0)^\top , \quad
-\mu_2 = 1.5\times (1,1,0,0)^\top, \\
-\Sigma_1 &= I_4, \quad
-\Sigma_2 = \left( 
-\begin{array}{c|c} 
-\tfrac{1}{2} I_2 & 0  \\ \hline
-0 & I_2 
-\end{array}\right), \\
-n_1 &= N - m, \quad  n_2 = N + m, \quad N=300.
-\end{split}
-\end{equation}
-We then increase $m \in [0,240]$ making
-the clusters progressively more unbalanced.
-For highly unbalanced clusters, we see that GMM performs better than
-the other methods, which have basically similar performance.
-Based on this experiment, an interesting problem would be to
-extend kernel k-groups to account for unbalanced clusters.
-
-In Fig.~\ref{fig:2dscatter} we show examples of two-dimensional
-datasets whose clustering results are shown in Table~\ref{tb:cigar_circle}.
-For kernel k-means and kernel k-groups we initialize at random.
-We see that both methods perform closely, with higher accuracy
-than the other ones.
-
-\begin{figure}
-\centering
-\begin{minipage}{0.2\textwidth}
-\includegraphics[width=1\textwidth]{2cigars.pdf}\\[-.2em](a)
-\end{minipage}\hspace{2em}
-\begin{minipage}{0.2\textwidth}
-\includegraphics[width=1\textwidth]{2circles.pdf}\\[-.2em](b)
-\end{minipage}
-\caption{
-\label{fig:2dscatter}
-(a) Data distributed as 
-$x \stackrel{iid}{\sim}
-\tfrac{1}{2}\mathcal{N}(\mu_1,\Sigma_1)+
-\tfrac{1}{2}\mathcal{N}(\mu_1,\Sigma_1)$, 
-$\mu_1= (0,0)^\top$, 
-$\mu_2 = (6.5, 0)^\top$ and 
-$\Sigma_1 = \Sigma_2 = \diag(1, 20)$. We sample $800$ points.
-(b) Concentric circles with radius $r_1 = 1$ and $r_2 = 3$, with 
-noise $0.2 \cdot \mathcal{N}(0, I_2)$. We sample 
-800 points with probability $1/2$ for each class.
-} 
-\end{figure}
-
-\begin{table}
-\caption{
-\label{tb:cigar_circle}
-Clustering the data from Fig.~\ref{fig:2dscatter}.
-For spcetral clustering, kernel k-means and kernel k-groups we use
-the metric $\widetilde{\rho}_2$ (see \eqref{eq:rho_tilde}) for the data
-in Fig.~\ref{fig:2dscatter}a, while $\widehat{\rho}_1$ (see \eqref{eq:rho_hat})
-for the data in Fig.~\ref{fig:2dscatter}b. We have 800
-points on each trial and 30 Monte Carlo runs for both datasets.
-}
-\centering
-\begin{tabular}{@{}l|ll|ll@{}}
-                   & \multicolumn{2}{c|}{Fig.~\ref{fig:2dscatter}a}  &
-                     \multicolumn{2}{c}{Fig.~\ref{fig:2dscatter}b}  \\
-       Method      & Accuracy & SEM     & Accuracy & SEM \\ \midrule[.5pt]
-        kmeans       & 0.533 & 0.005  & 0.521   & 0.003 \\
-         GMM         & 0.929 & 0.029  & 0.533   & 0.004 \\
- spectral clustering & 0.577 & 0.010  & 0.725   & 0.003 \\
-    kernel k-means   & \textbf{1.000} & 0.000  & \textbf{1.000}   & 0.000 \\
-   kernel k-groups   & \textbf{1.000} & 0.000  & \textbf{1.000}  & 0.000  
-\end{tabular}
-\end{table}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Real Data Experiment}
-
-We consider the dermatology dataset \cite{Dua2017,Guvenir1998}
-which has 366 data points, each with 34 attributes where 33 are
-linear valued and one is categorical. There are 8 data points with
-missing entries in the ``age'' column. We complete the missing entries
-with the mean of the entire column, and then we normalize the entire
-dataset to zero mean and unit variance. There are  a total of $6$ classes,
-and this is a challenging clustering problem. We refer the reader
-to \cite{Dua2017,Guvenir1998} for a complete description of the dataset, 
-and also to \cite{RizzoClustering} where this dataset was previously
-analyzed. In Table~\ref{tb:dermatology} we show the results of kernel
-k-groups using the metric \eqref{eq:rho_alpha} with $\alpha=1/2$.
-We show the number of points assigned to each class while indicating
-the actual class that points belong to. In Table~\ref{tb:dermatology_accuracy}
-we show this experiment using several clustering methods, and we also
-compare with the results from \cite{RizzoClustering} and \cite{Kgroups}
-on this same data. Our results provide an improvement in comparison
-to all the other methods and also the analysis
-of \cite{RizzoClustering}.
-
-\begin{table}
-\caption{
-\label{tb:dermatology}
-Clustering the dermatology 
-dataset of \cite{Dua2017,Guvenir1998} with kernel k-groups
-using the metric $\rho_{1/2}$ (see \eqref{eq:rho_alpha}) from energy
-statistics. The table below should be compared with Table~2 of
-\cite{RizzoClustering}, for which our results are slightly more accurate.
-See also Table~\ref{tb:dermatology_accuracy} below for clustering metrics.
-The classes in the vertical indicates the ground truth and the classes
-in the horizontal correspond to the classification
-obtained by kernel k-groups. We show the estimated number of points
-for each class.
-}
-\centering
-\begin{tabular}{@{}l|llllll|l@{}}
-%\toprule[1pt]
-class & 1 & 2 & 3 & 4 & 5 & 6 & $\#$ cases \\ \midrule[.5pt]
-   1   & 112 & 0  & 0  & 0  & 0  & 0  &  112  \\
-   2   &  0  & 50 & 0  & 11 & 0  & 0  &   61  \\
-   3   &  0  & 0  & 72 & 0  & 0  & 0  &   72  \\
-   4   &  0  & 2  & 0  & 47 & 0  & 0  &   49  \\
-   5   &  0  & 0  & 0  & 1  & 51 & 0  &   52  \\
-   6   &  0  & 0  & 0  & 0  & 0  & 20 &   20  \\ \midrule[.5pt]
-total & 112 & 52 & 72 & 59 & 51 & 20 &  366 %\\ \bottomrule[1pt] 
-\end{tabular}
-\end{table}
-
-\begin{table}
-\caption{
-\label{tb:dermatology_accuracy}
-For the dataset \cite{Dua2017,Guvenir1998} (see also
-Table~\ref{tb:dermatology}) we show the accuracy \eqref{eq:accuracy} and
-the adjusted Rand index (aRand) of several methods.
-In \cite{RizzoClustering} the authors 
-obtained $\textnormal{aRand}=0.9195$ using an energy statistics
-based method, while \cite{Kgroups} obtains $\textnormal{aRand}=0.9188$ 
-where points with missing
-entries are removed. Below we complete the missing entries with the mean.
-If we remove the points with missing entries, kernel k-groups provides an
-improvement of $\textnormal{accuracy}=0.9637$ and $\textnormal{aRand}=0.9396$.
-}
-\centering
-\begin{tabular}{@{}l|ll@{}}
-       Method      & Accuracy & aRand \\ \midrule[.5pt]
-        kmeans       & 0.713 & 0.690 \\
-         GMM         & 0.877 & 0.840 \\
- spectral clustering & 0.954 & 0.912 \\
-    kernel k-means   & 0.751 & 0.851 \\
-   kernel k-groups   & \textbf{0.962} & \textbf{0.936}
-\end{tabular}
-\end{table}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Conclusion}
-\label{sec:conclusion}
-
-We proposed a formulation to clustering based on a weighted
-version of energy statistics, 
-valid for arbitrary spaces of negative type.
-Our mathematical formulation of energy clustering 
-reduces to a QCQP in the associated RKHS, as demonstrated in 
-Proposition~\ref{th:qcqp3}.
-We showed that the optimization problem
-is equivalent
-to kernel k-means, once the kernel is fixed, and also
-to several graph partitioning problems.
-%Energy statistics, however, fixes
-%a family of standard kernels in Euclidean space, and
-%more general kernels 
-%on spaces of negative type can also be obtained.
-
-We extended Hartigan's method to kernel spaces and proposed
-Algorithm~\ref{algo}, which we called kernel k-groups.
-This method was compared to kernel k-means and spectral clustering, besides
-k-means and GMM. Our numerical results show a superior performance of the
-proposed method, specially in high dimensions.
-We stress that kernel k-groups has the same complexity of kernel k-means.
-
-Kernel k-groups suffers a limitation shared by kernel k-means and spectral
-clustering which involves highly unbalanced
-clusters. An interesting problem  that we leave open 
-is to extend the method to such 
-situations.
-Finally, kernel methods can benefit from sparsity and
-fixed-rank approximations of the Gram matrix, and there is plenty
-of room to make kernel k-groups more scalable.
-
-% @gui: better, but still important to have another paragraph between those connecting it to the existing literature.  compare this work with 3-4 other works explicitly stating the difference.
-
-% if have a single appendix:
-%\appendix[Proof of the Zonklar Equations]
-% or
-\appendix  % for no appendix heading
-% do not use \section anymore after \appendix, only \section*
-% is possibly needed
-
-% use appendices with more than one appendix
-% then use \section to start each appendix
-% you must declare a \section before using any
-% \subsection or using \label (\appendices by itself
-% starts a section numbered zero.)
-%
-
-%\appendices
-%\section{Proof of the First Zonklar Equation}
-%Appendix one text goes here.
-
-% you can choose not to have a title for an appendix
-% if you want by leaving the argument blank
-%\section{}
-%Appendix two text goes here.
-
-\section*{Two-Class Problem in One Dimension}
-\label{sec:twoclass}
-
-Here we consider the simplest possible case which
-is one-dimensional data and a two-class problem. We propose
-an algorithm that does not depend on initialization. We used this simple
-scheme to compare with kernel k-groups given
-in algorithm \ref{algo}. Both algorithms have the same clustering
-performance in the one-dimensional examples that we tested.
-
-Let us fix 
-$\rho(x,y) = |x - y|$ according to the standard energy distance. We also
-fix the weights $w(x) = 1$ for every data point $x$. We can
-thus 
-compute the function
-\eqref{eq:g_def} in $\OO(n \log n)$ and minimize
-$W$ directly.
-This is done by noting that
-\begin{equation}
-\begin{aligned}
-|x - y|  &= (x-y)\Ind{x \ge y} -
-(x-y) \Ind{x < y}  \\
-&= 
-x \left( \Ind{x \ge y} - \Ind{x < y} \right)  + 
-y \left( \Ind{y > x} - \Ind{y \le x} \right)  
-\end{aligned}
-\end{equation}
-where we have the indicator function defined by
-$\Ind{A}=1$ if $A$ is true, and $\Ind{A}=0$ otherwise.
-Let $\C$ be a partition with
-$n$ elements. Using the above distance we have
-\begin{equation}
-\label{eq:g_ind}
- g\left(\C,\C\right) =  \dfrac{1}{n^2}  
-\sum_{x \in \C} 
-\sum_{y \in \C} 
-x \left(
-\Ind{x \ge y} + \Ind{y > x} - 
-\Ind{x \ge y}-\Ind{x < y} \right) .
-\end{equation}
-The sum over $y$ can be eliminated since each term in
-the parenthesis is simply counting the number of elements in $\C$ that satisfy
-the condition of the indicator function. Assuming
-that we first order the data in $\C$, obtaining
-$\tC = [ x_j \in \C: x_1 \le x_2 \le \dotsm \le x_{n}]$, we
-get
-\begin{equation}
-\label{eq:g1d}
-g\big(\tC, \tC \big) = 
-\dfrac{2}{n^2} \sum_{\ell=1}^n (2\ell - 1 - n) x_\ell .
-\end{equation}
-Note that the cost of computing
-$g\big( \tC, \tC \big)$
-is $\OO(n)$ and the cost of
-sorting the data
-is at the most $\OO(n\log n)$.
-Assuming that each partition is ordered,
-$\mathbb{X} = \bigcup_{j=1}^k \tC_j$,
-the within energy dispersion
-can be written explicitly as
-\begin{equation}
-\label{eq:w1d}
-W\big( \tC_1,\dotsc,\tC_k \big) = 
-\sum_{j=1}^k \sum_{\ell=1}^{n_j} \dfrac{2\ell - 1 - n_j}{n_j} \, x_\ell.
-\end{equation}
-
-For a two-class problem we can use the formula
-\eqref{eq:w1d} to cluster the data
-through a simple algorithm
-as follows. We first order
-the entire dataset, $\mathbb{X} \to \widetilde{\mathbb{X}}$. Then
-we compute \eqref{eq:w1d} for each possible split of $\widetilde{\mathbb{X}}$
-and pick the point which gives the minimum value of $W$.
-This procedure is described in Algorithm~\ref{algo1d}.
-Note that this algorithm is deterministic,
-however,
-it only works for one-dimensional data with Euclidean distance. Its total
-complexity is $\OO(n\log n + n^2) = \OO(n^2)$.
-
-\begin{algorithm}[h]
-\vspace{.5em}
-\begin{algorithmic}[1]
-\INPUT data $\mathbb{X}$
-\OUTPUT label matrix $Z$
-\STATE sort $\mathbb{X}$ obtaining
-$\widetilde{\mathbb{X}}= [ x_1,\dotsc,x_n ]$
-    \FOR{$j\in [ 1,\dotsc,n ]$}
-        \STATE $\tC_{1,j} \leftarrow [x_i: i=1,\dotsc,j]$
-        \STATE $\tC_{2,j} \leftarrow [x_i : i=j+1,\dotsc,n]$
-        \STATE
-            $W^{(j)} \leftarrow W \big( \tC_{1,j},\tC_{2,j}\big)$
-            \hfill (see \eqref{eq:w1d})
-    \ENDFOR
-    \STATE $j^\star \leftarrow \argmin_j W^{(j)}$
-	\FOR{$j \in [1,\dotsc,n]$}
-		\IF{ $j\le j^\star$ }
-    		\STATE $Z_{j\bullet} \leftarrow (1,0) $
-        \ELSE
-			\STATE $Z_{j\bullet} \leftarrow (0,1)$
-		\ENDIF
-	\ENDFOR
-\end{algorithmic}
-\caption{
-\label{algo1d}
-Clustering algorithm to
-find local solutions to the optimization
-problem \eqref{eq:minimize}
-for a two-class problem in one dimension. 
-}
-\end{algorithm}
-
-% use section* for acknowledgment
-\ifCLASSOPTIONcompsoc
-  % The Computer Society usually uses the plural form
-  \section*{Acknowledgments}
-\else
-  % regular IEEE prefers the singular form
-  \section*{Acknowledgment}
-\fi
-
-We would like to thank Carey Priebe for discussions.
-We would like to acknowledge the support of the Transformative
-Research Award (NIH \#R01NS092474) and  the Defense Advanced Research Projects
-Agency’s (DARPA) SIMPLEX program through SPAWAR contract N66001-15-C-4041.
-
-% Can use something like this to put references on a page
-% by themselves when using endfloat and the captionsoff option.
-\ifCLASSOPTIONcaptionsoff
-  \newpage
-\fi
-
-% trigger a \newpage just before the given reference
-% number - used to balance the columns on the last page
-% adjust value as needed - may need to be readjusted if
-% the document is modified later
-%\IEEEtriggeratref{8}
-% The "triggered" command can be changed if desired:
-%\IEEEtriggercmd{\enlargethispage{-5in}}
-
-% references section
-
-% can use a bibliography generated by BibTeX as a .bbl file
-% BibTeX documentation can be easily obtained at:
-% http://mirror.ctan.org/biblio/bibtex/contrib/doc/
-% The IEEEtran BibTeX style support page is at:
-% http://www.michaelshell.org/tex/ieeetran/bibtex/
-%\bibliographystyle{IEEEtran}
-% argument is your BibTeX string definitions and bibliography database(s)
-%\bibliography{IEEEabrv,../bib/paper}
-%
-% <OR> manually copy in the resultant .bbl file
-% set second argument of \begin to the number of references
-% (used to reserve space for the reference number labels box)
-
-\bibliographystyle{unsrt}
-%\bibliographystyle{abbrvnat}
-\bibliography{biblio.bib}
+%%%%%%%%%%% bio sketch for Maria inserted below %%%%%%%%%%%%
 
@@
 
-\begin{IEEEbiography}[{\includegraphics[width=1in,clip,keepaspectratio,height=1.25in]{rizzo.jpg}}]{Maria Rizzo}
-I hold a BS and MA degree in Mathematics from the University of Toledo, a MS in
-Applied Statistics
-from BGSU, and a PhD from BGSU in 2002. After four years as Assistant Professor
-at Ohio University
-Department of Mathematics, I returned to BGSU in 2006, where I am currently
-Professor of Statistics in
-the Department of Mathematics and Statistics. Primary research interests are
-centered on energy statistics,
-distance correlation and its applications, computational statistics, and
-related topics. In addition
-to teaching statistics, data science, and actuarial science, I enjoy writing
-textbooks about
-statistical computing and using R software, a forthcoming research monograph on
-energy statistics
-(joint with G. Szekely) and have other book writing projects under contract as
-well.
+\begin{IEEEbiography}[{\includegraphics[width=1in,clip,keepaspectratio]{Maria_Rizzo.jpg}}]{Maria Rizzo}
+I hold a BS and MA degree in Mathematics from the University of Toledo, a MS in Applied Statistics
+from BGSU, and a PhD from BGSU in 2002. After four years as Assistant Professor at Ohio University
+Department of Mathematics, I returned to BGSU in 2006, where I am currently Professor of Statistics in
+the Department of Mathematics and Statistics. Primary research interests are centered on energy statistics,
+distance correlation and its applications, computational statistics, and related topics. In addition
+to teaching statistics, data science, and actuarial science, I enjoy writing textbooks about
+statistical computing and using R software, a forthcoming research monograph on energy statistics 
+(joint with G. Szekely) and have other book writing projects under contract as well. 
 \end{IEEEbiography}

### 2018-08-13 energy_clustering/pami18/energy_clustering.tex
pami

--- 2018-08-10 energy_clustering/pami18/energy_clustering.tex
@@
 \author{Guilherme~Fran\c ca,~Maria Rizzo~and~Joshua T.~Vogelstein
-\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem G. Fran\c ca 
-and J. Vogelstein are with the Center for Imaging Science, Johns Hopkins
-University; J. Vogelstein is also with 
+\IEEEcompsocitemizethanks{
+\IEEEcompsocthanksitem G. Fran\c ca
+is with the Mathematical Institute for Data Science (MINDS), Johns Hopkins
+University.\protect \\
+E-mail: guifranca@jhu.edu
+\IEEEcompsocthanksitem J. Vogelstein is 
+with the Center for Imaging Science, 
 the Department of Biomedical Engineering and Institute for Computational
@@
 % \\ is fragile and will error, could use \hfil\break instead.
-E-mail: guifranca@jhu.edu; jovo@jhu.edu
+E-mail: jovo@jhu.edu
 \IEEEcompsocthanksitem M. Rizzo is with the Department of Mathematics
@@
 E-mail: mrizzo@bgsu.edu}% <-this % stops an unwanted space
-\thanks{Manuscript received March 15, 2018; revised March 15, 2018.}}
+\thanks{Manuscript received August 14, 2018; revised August 14, 2018.}}
 
@@
 analysis of variance \cite{RizzoVariance}, nonlinear dependence tests through
-distance covariance and distance correlation~\cite{Szekely2007-mm}, which generalizes the Pearson
+distance covariance and distance correlation~\cite{Szekely2007-mm}, 
+which generalizes the Pearson
 correlation coefficient, and hierarchical clustering 
@@
 energy statistics to clustering
-was recently proposed \cite{Kgroups}, and the method was named 
+was recently proposed \cite{Kgroups} and the method was named 
 \emph{k-groups}.
@@
 This equivalence immediately relates energy statistics to
-kernel methods often used in machine learning, and form the basis 
+kernel methods often used in machine learning and form the basis 
 of our approach in this paper.
@@
 which 
-we do   more generality. For instance, our approach is not limited
+we do in  more generality. For instance, our approach is not limited
 to the Euclidean case but holds for general arbitrary spaces
 of negative type. 
-For generality, we also
+Moreover, we also
 consider a weighted version of energy statistics.
@@
 For each setting, we show the average accuracy over $100$ Monte
-Carlo trials, with error bars indicating the standard error.
+Carlo trials (we ommit the error bars since they are too small to be
+visible in our experiments).
 
@@
 \subsection{Synthetic Experiments}
+
+%\begin{figure*}
+%\centering
+%\begin{minipage}{.4\textwidth}
+%\includegraphics[width=1\textwidth]{gauss_1d.pdf}\\[-2em]
+%(a)
+%\end{minipage}
+%\begin{minipage}{.4\textwidth}
+%\includegraphics[width=1\textwidth]{loggauss_1d.pdf} \\[-2em]
+%(b)
+%\end{minipage}\\[1em]
+%\begin{minipage}{.37\textwidth}
+%\includegraphics[width=1\textwidth]{normal_density.pdf}\\[-2em]
+%(c)
+%\end{minipage}\hspace{2em}
+%\begin{minipage}{.37\textwidth}
+%\includegraphics[width=1\textwidth]{lognormal_density.pdf}\\[-2em]
+%(d)
+%\end{minipage}
+%\caption{
+%\label{fig:1dexperiment}
+%Clustering one-dimensional data for a two-class problem.
+%(a) $x \stackrel{iid}{\sim} \tfrac{1}{2} \mathcal{N}(0, 1.5) +
+%\tfrac{1}{2}\mathcal{N}(1.5,0.3)$. (b) $x \stackrel{iid}{\sim}
+%\tfrac{1}{2}e^{\mathcal{N}(0, 1.5)} + \tfrac{1}{2} e^{\mathcal{N}(1.5, 0.3)}$.
+%In both cases we plot the average accuracy versus the total number of points.
+%We ommit error bars since they are too small.
+%(c) Kernel density estimation for a single trial with 2000 sampled
+%points from the
+%distribution in the case (a). (d) Kernel density estimation in a single
+%trial with 
+%2000 points sampled from the distribution in the case (b).
+%}
+%\end{figure*}
 
@@
 \begin{minipage}{.4\textwidth}
-\includegraphics[width=1\textwidth]{gauss_1d.pdf}\\[-2em]
-(a)
+\includegraphics[width=1\textwidth]{gauss_1d.pdf}\\[-2em](a)
 \end{minipage}
 \begin{minipage}{.4\textwidth}
-\includegraphics[width=1\textwidth]{loggauss_1d.pdf} \\[-2em]
-(b)
+\includegraphics[width=1\textwidth]{loggauss_1d.pdf} \\[-2em](b)
 \end{minipage}\\[1em]
 \begin{minipage}{.37\textwidth}
-\includegraphics[width=1\textwidth]{normal_density.pdf}\\[-2em]
-(c)
+\includegraphics[width=1\textwidth]{normal_density2.pdf}\\[-2em](c)
 \end{minipage}\hspace{2em}
 \begin{minipage}{.37\textwidth}
-\includegraphics[width=1\textwidth]{lognormal_density.pdf}\\[-2em]
-(d)
+\includegraphics[width=1\textwidth]{lognormal_density2.pdf}\\[-2em](d)
 \end{minipage}
@@
 Clustering one-dimensional data for a two-class problem.
-(a) $x \stackrel{iid}{\sim} \tfrac{1}{2} \mathcal{N}(0, 1.5) +
-\tfrac{1}{2}\mathcal{N}(1.5,0.3)$. (b) $x \stackrel{iid}{\sim}
-\tfrac{1}{2}e^{\mathcal{N}(0, 1.5)} + \tfrac{1}{2} e^{\mathcal{N}(1.5, 0.3)}$.
-In both cases we plot the average accuracy versus the total number of points.
-We ommit error bars since they are too small.
-(c) Kernel density estimation for a single trial with 2000 sampled
-points from the
-distribution in the case (a). (d) Kernel density estimation in a single
-trial with 
-2000 points sampled from the distribution in the case (b).
+(a) Data normally distributed as
+$x \stackrel{iid}{\sim} (1/2) \mathcal{N}(0, 1.5) +
+(1/2) \mathcal{N}(1.5,0.3)$. (b) Data following lognormal distributions
+as $x \stackrel{iid}{\sim}
+(1/2)e^{\mathcal{N}(0, 1.5)} + (1/2) e^{\mathcal{N}(1.5, 0.3)}$.
+In both cases we plot the average accuracy versus the total number of points
+(error bars are too small to be visible).
+(c) For the same distribution as in item (a), we sample 2000
+points, cluster them with the three methods, then perform a kernel density 
+estimation for kernel k-groups, since this is a model-free method,
+while for k-means and GMM we show the estimated Gaussian for each
+class.
+The clustering accuracy for each method is in Table~\ref{tb:dens}.
+(d) Exactly the same experiment but for the distribution in item (b). 
+Note that only kernel k-groups is able to distinguish between the two classes
+in this example.
 }
@@
        Method        & normal & lognormal \\ \midrule[.5pt]
-        kmeans       & 0.773  & 0.502     \\
-         GMM         & \textbf{0.881}  & 0.513     \\
-   kernel k-groups   & 0.802  & \textbf{0.847}  
+        kmeans       & 0.778  & 0.520     \\
+         GMM         & \textbf{0.887}  & 0.542     \\
+   kernel k-groups   & 0.807  & \textbf{0.846}  
 \end{tabular}
@@
 To consider non-normal data, we sample from the lognormal mixture 
-$x \stackrel{iid}{\sim} \tfrac{1}{2} 
-e^{\mathcal{N}(\mu_1,\Sigma_1)}+
-\tfrac{1}{2} e^{\mathcal{N}(\mu_2,\Sigma_2)}$ with the same
+$x \stackrel{iid}{\sim}  
+(1/2) e^{\mathcal{N}(\mu_1,\Sigma_1)}+
+(1/2) e^{\mathcal{N}(\mu_2,\Sigma_2)}$ with the same
 parameters as in \eqref{eq:20gauss}. 
@@
 \begin{minipage}{0.2\textwidth}
-\includegraphics[width=1\textwidth]{2cigars.pdf}\\[-.2em](a)
+(a)~\includegraphics[width=.8\textwidth,trim=0 .8cm 0 0]{2cigars.pdf}%\\[-.2em](a)
 \end{minipage}\hspace{2em}
 \begin{minipage}{0.2\textwidth}
-\includegraphics[width=1\textwidth]{2circles.pdf}\\[-.2em](b)
+(b)~\includegraphics[width=.8\textwidth,trim=0 .8cm 0 0]{2circles.pdf}%\\[-.2em](b)
 \end{minipage}
@@
 $x \stackrel{iid}{\sim}
-\tfrac{1}{2}\mathcal{N}(\mu_1,\Sigma_1)+
-\tfrac{1}{2}\mathcal{N}(\mu_1,\Sigma_1)$, 
+(1/2)\mathcal{N}(\mu_1,\Sigma_1)+
+(1/2)\mathcal{N}(\mu_1,\Sigma_1)$, 
 $\mu_1= (0,0)^\top$, 
@@
 
-\begin{table}
+\begin{table}[t]
 \caption{
@@
 Clustering the data from Fig.~\ref{fig:2dscatter}.
-For spcetral clustering, kernel k-means and kernel k-groups we use
+For spectral clustering, kernel k-means and kernel k-groups we use
 the metric $\widetilde{\rho}_2$ (see \eqref{eq:rho_tilde}) for the data
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Real Data Experiment}
-
-We consider the dermatology dataset \cite{Dua2017,Guvenir1998}
-which has 366 data points, each with 34 attributes where 33 are
-linear valued and one is categorical. There are 8 data points with
-missing entries in the ``age'' column. We complete the missing entries
-with the mean of the entire column, and then we normalize the entire
-dataset to zero mean and unit variance. There are  a total of $6$ classes,
-and this is a challenging clustering problem. We refer the reader
-to \cite{Dua2017,Guvenir1998} for a complete description of the dataset, 
-and also to \cite{RizzoClustering} where this dataset was previously
-analyzed. In Table~\ref{tb:dermatology} we show the results of kernel
-k-groups using the metric \eqref{eq:rho_alpha} with $\alpha=1/2$.
-We show the number of points assigned to each class while indicating
-the actual class that points belong to. In Table~\ref{tb:dermatology_accuracy}
-we show this experiment using several clustering methods, and we also
-compare with the results from \cite{RizzoClustering} and \cite{Kgroups}
-on this same data. Our results provide an improvement in comparison
-to all the other methods and also the analysis
-of \cite{RizzoClustering}.
-
-\begin{table}
+\begin{table}[t]
 \caption{
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Real Data Experiment}
+
+We consider the dermatology dataset \cite{Dua2017,Guvenir1998}
+which has 366 data points, each with 34 attributes where 33 are
+linear valued and one is categorical. There are 8 data points with
+missing entries in the ``age'' column. We complete the missing entries
+with the mean of the entire column, and then we normalize the entire
+dataset to zero mean and unit variance. There are  a total of $6$ classes,
+and this is a challenging clustering problem. We refer the reader
+to \cite{Dua2017,Guvenir1998} for a complete description of the dataset, 
+and also to \cite{RizzoClustering} where this dataset was previously
+analyzed. In Table~\ref{tb:dermatology} we show the results of kernel
+k-groups using the metric \eqref{eq:rho_alpha} with $\alpha=1/2$.
+We show the number of points assigned to each class while indicating
+the actual class that points belong to. In Table~\ref{tb:dermatology_accuracy}
+we show this experiment using several clustering methods, and we also
+compare with the results from \cite{RizzoClustering} and \cite{Kgroups}
+on this same data. Our results provide an improvement in comparison
+to all the other methods and also the analysis
+of \cite{RizzoClustering}.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Conclusion}
@@
 proposed method, specially in high dimensions.
-We stress that kernel k-groups has the same complexity of kernel k-means.
+We stress that kernel k-groups has the same complexity of kernel
+k-means\footnote{
+An implementation of unweighted $k$-groups is publicly available in the
+energy package for R \cite{energy}.}.
 
@@
 \begin{IEEEbiography}[{\includegraphics[width=1in,clip,keepaspectratio]{gui2.png}}]{Guilherme Fran\c ca} 
-has a BSc in physics, and received both his MSc and 
+has a BS degree in physics and received both his MS and 
 PhD in Theoretical Physics from the Institute of 
@@
 He was a Postdoctoral Fellow in
-the Computer Science Department at Boston College (2015), working in
-optimization theory.
+the Computer Science Department at Boston College (2015), working
+in optimization methods.
 Since 2016 he is a Postdoctoral Fellow at Johns Hopkins University, currently
 with the Mathematical Institute for Data Science (MINDS). 
-His current research interests are in theoretical methods in the broad
-area of machine learning 
-and optimization.
+His current research interests involve  
+machine learning and optimization.
 \end{IEEEbiography}
 
-% if you will not have a photo at all:
-\begin{IEEEbiographynophoto}{Maria Rizzo}
-Biography text here.
-\end{IEEEbiographynophoto}
+\begin{IEEEbiography}[{\includegraphics[width=1in,clip,keepaspectratio,height=1.30in]{rizzo.jpg}}]{Maria Rizzo}
+holds a BS and MA degree in Mathematics from the University of Toledo, a MS in
+Applied Statistics from BGSU, and a PhD from BGSU in 2002. After four years as
+Assistant Professor at Ohio University Department of Mathematics, she 
+returned to BGSU in 2006, where she is currently Professor of Statistics 
+in the Department of
+Mathematics and Statistics. Her primary research interests are 
+centered on energy
+statistics, distance correlation and its applications, computational
+statistics, and related topics. In addition to teaching statistics, data
+science, and actuarial science, she enjoys writing textbooks about statistical
+computing and using R software. She has a forthcoming research 
+monograph on energy
+statistics (joint with G. Szekely) and also other book writing projects under
+contract as well.
+\end{IEEEbiography}
 
@@
 \begin{IEEEbiography}[{\includegraphics[width=1in,clip,keepaspectratio]{jovo.png}}]{Joshua~T.~Vogelstein}
-I received a B.S degree from the Department of Biomedical Engineering (BME) at Washington University in St.
-Louis, MO in 2002, a M.S.
-degree from the Department of Applied Mathematics and Statistics (AMS) at Johns Hopkins University (JHU) in Baltimore, MD in 2009, and a Ph.D.
-degree from the Department of Neuroscience at JHU in 2009.
-I was a Postdoctoral Fellow in AMS@JHU from 2009 until 2011, at which time I was appointed an Assistant Research Scientist, and became a member of the Institute for Data Intensive Science and Engineering.
-I spent 2 years at Information Initiative at Duke University, before coming home to my current appointment as Assistant Professor in BME@JHU, and core faculty in both the Institute for Computational Medicine and the Center for Imaging Science, as well as a member of the Kavli Neuroscience Discovery Institute.
-I married my kindergarten sweetheart in the summer of 2014.
-
-My primary research interest is to extend and fuse statistical machine learning and big data science to address the most important brain science and mental health questions of our time, particularly  connectomics.
-Our group’s research has been featured in a number of prominent scientific and engineering journals and conferences including Nature, Nature Methods, Science, Cell, PNAS, Science Translational Medicine, JMLR, PAMI, Annals of Applied Statistics,  NIPS, and SIAM Journal of Matrix Analysis and Applications.
-In 2011, I co-founded the Open Connectome Project which expanded in 2015 to be NeuroData, whose mission is to enable terascale neuroscience for everyone.
-All our work is conducted according to the highest standards of open science.
+received a BS degree from the Department of Biomedical Engineering (BME) at
+Washington University in St. Louis, MO in 2002, a MS degree from the
+Department of Applied Mathematics and Statistics (AMS) at Johns Hopkins
+University (JHU) in Baltimore, MD in 2009, and a PhD degree from the
+Department of Neuroscience at JHU in 2009. He was a Postdoctoral Fellow in
+AMS/JHU from 2009 until 2011, at which time he was appointed an Assistant
+Research Scientist, and became a member of the Institute for Data Intensive
+Science and Engineering. He spent 2 years at Information Initiative at Duke
+University, before coming home to his current appointment as 
+Assistant Professor
+in BME/JHU, and core faculty in both the Institute for Computational Medicine
+and the Center for Imaging Science, as well as a member of the Kavli
+Neuroscience Discovery Institute. He married his kindergarten sweetheart in the
+summer of 2014.
+
+His primary research interest is to extend and fuse statistical 
+machine learning
+and big data science to address the most important brain science and mental
+health questions of our time, particularly  connectomics. 
+His group’s research
+has been featured in a number of prominent scientific and engineering journals
+and conferences including Nature, Nature Methods, Science, Cell, PNAS, Science
+Translational Medicine, JMLR, PAMI, Annals of Applied Statistics,  NIPS, and
+SIAM Journal of Matrix Analysis and Applications. In 2011, he co-founded the
+Open Connectome Project which expanded in 2015 to be NeuroData, whose mission
+is to enable terascale neuroscience for everyone. All his works are conducted
+according to the highest standards of open science.
 \end{IEEEbiography}

### 2018-08-13 energy_clustering/kgroups_code/kernel_density2.py

@@ -7,6 +7,9 @@ import numpy as np
 import matplotlib.pyplot as plt
 from sklearn.neighbors import KernelDensity
 from prettytable import PrettyTable
+from sklearn.cluster import KMeans
+from sklearn.mixture import GaussianMixture as GMM
+import scipy.stats
 
 import sys
 
@@ -16,7 +19,8 @@ import eclust
 import metric
 
 
-### generate data
+### parameters ################################################################
+
 k = 2
 n = 2000
 n1, n2 = np.random.multinomial(n, [0.5, 0.5])
@@ -24,30 +28,65 @@ m1 = 0
 s1 = 1.5
 m2 = 1.5
 s2 = 0.3
-X, z = data.univariate_normal([m1, m2], [s1, s2], [n1, n2])
-#X, z = data.univariate_lognormal([m1, m2], [s1, s2], [n1, n2])
+
+#X, z = data.univariate_normal([m1, m2], [s1, s2], [n1, n2])
+X, z = data.univariate_lognormal([m1, m2], [s1, s2], [n1, n2])
+
 Y = np.array([[x] for x in X])
 bw = 0.5 # bandwidth
 num_points = 1500 # number points for linspace
-low = -6
-high = 6
-#low = -2
-#high = 20 
 
-### clustering
+#low = -6
+#high = 6
+#ymax = 1.1
+low = -2
+high = 20 
+ymax = 0.5
+
+methods = ['k-means', 'GMM', 'kernel k-groups']
+colors = ["#54278f", "#5F8793", "#F41711"]
+
+#fname = "normal_density2.pdf"
+fname = "lognormal_density2.pdf"
+
+###############################################################################
+
 t = PrettyTable(['Method', 'Accuracy'])
+
+km = KMeans(k, n_init=5)
+km.fit(Y)
+zh_kmeans = km.labels_
+x1_kmeans = X[np.where(zh_kmeans==0)][:, np.newaxis]
+x2_kmeans = X[np.where(zh_kmeans==1)][:, np.newaxis]
+x1_mu_kmeans, x2_mu_kmeans = km.cluster_centers_
+x1_mu_kmeans, x2_mu_kmeans = x1_mu_kmeans[0], x2_mu_kmeans[0]
[…]

### 2018-08-13 energy_clustering/kgroups_code/wrapper.py

@@ -69,14 +69,17 @@ def kmeans(k, X, run_times=5):
     km = KMeans(k, n_init=run_times)
     km.fit(X)
     zh = km.labels_
-    return zh
+    mu = km.cluster_centers_
+    return zh, mu
 
 def gmm(k, X, run_times=5):
     gm = GMM(k, n_init=run_times, init_params='kmeans')
     #gm = GMM(k)
     gm.fit(X)
     zh = gm.predict(X)
-    return zh
+    mu = gm.means_
+    cov = gm.covariances_
+    return zh, mu, cov
 
 def spectral_clustering(k, X, G, W=None, run_times=5):
     if type(W) == type(None):

### 2018-08-13 energy_clustering/pami18/energy_clustering.tex
pami

@@
 
-\author{Guilherme~Fran\c ca,~Maria Rizzo~and~Joshua T.~Vogelstein
+\author{Guilherme~Fran\c ca,~Maria L.~Rizzo~and~Joshua T.~Vogelstein
 \IEEEcompsocitemizethanks{
@@
 E-mail: guifranca@jhu.edu
-\IEEEcompsocthanksitem J. Vogelstein is 
+\IEEEcompsocthanksitem J. T. Vogelstein is 
 with the Center for Imaging Science, 
@@
 E-mail: jovo@jhu.edu
-\IEEEcompsocthanksitem M. Rizzo is with the Department of Mathematics
+\IEEEcompsocthanksitem M. L. Rizzo is with the Department of Mathematics
 and Statistics, Bowling Green State University. \protect\\
@@
 
-\begin{IEEEbiography}[{\includegraphics[width=1in,clip,keepaspectratio,height=1.30in]{rizzo.jpg}}]{Maria Rizzo}
+\begin{IEEEbiography}[{\includegraphics[width=1in,clip,keepaspectratio,height=1.30in]{rizzo.jpg}}]{Maria~L.~Rizzo}
 holds a BS and MA degree in Mathematics from the University of Toledo, a MS in

### 2018-08-14 energy_clustering/arxiv/energy_clustering.tex
pami/arxiv

--- 2018-08-13 energy_clustering/pami18/energy_clustering.tex
@@
 E-mail: mrizzo@bgsu.edu}% <-this % stops an unwanted space
-\thanks{Manuscript received August 14, 2018; revised August 14, 2018.}}
+%\thanks{Manuscript received August 14, 2018; revised August 14, 2018.}
+}
 
 % The paper headers
-\markboth{Energy Clustering}{}
+%\markboth{Energy Clustering}{}
 %{Journal of \LaTeX\ Class Files,~Vol.~14, No.~8, August~2015}%
@@
 
-\begin{IEEEbiography}[{\includegraphics[width=1in,clip,keepaspectratio]{gui2.png}}]{Guilherme Fran\c ca} 
-has a BS degree in physics and received both his MS and 
-PhD in Theoretical Physics from the Institute of 
-Theoretical Physics IFT-UNESP/ICTP-SAIFR (2012). He was a Postdoctoral Fellow
-with the High Energy Theory group in the Physics Department 
-at Cornell University (2013--2015), working on topics of mathematical physics
-and non-perturbative methods in quantum field theory 
-and statistical mechanics.
-He was a Postdoctoral Fellow in
-the Computer Science Department at Boston College (2015), working
-in optimization methods.
-Since 2016 he is a Postdoctoral Fellow at Johns Hopkins University, currently
-with the Mathematical Institute for Data Science (MINDS). 
-His current research interests involve  
-machine learning and optimization.
-\end{IEEEbiography}
-
-\begin{IEEEbiography}[{\includegraphics[width=1in,clip,keepaspectratio,height=1.30in]{rizzo.jpg}}]{Maria~L.~Rizzo}
-holds a BS and MA degree in Mathematics from the University of Toledo, a MS in
-Applied Statistics from BGSU, and a PhD from BGSU in 2002. After four years as
-Assistant Professor at Ohio University Department of Mathematics, she 
-returned to BGSU in 2006, where she is currently Professor of Statistics 
-in the Department of
-Mathematics and Statistics. Her primary research interests are 
-centered on energy
-statistics, distance correlation and its applications, computational
-statistics, and related topics. In addition to teaching statistics, data
-science, and actuarial science, she enjoys writing textbooks about statistical
-computing and using R software. She has a forthcoming research 
-monograph on energy
-statistics (joint with G. Szekely) and also other book writing projects under
-contract as well.
-\end{IEEEbiography}
-
-% insert where needed to balance the two columns on the last page with
-% biographies
-%\newpage
-
-\begin{IEEEbiography}[{\includegraphics[width=1in,clip,keepaspectratio]{jovo.png}}]{Joshua~T.~Vogelstein}
-received a BS degree from the Department of Biomedical Engineering (BME) at
-Washington University in St. Louis, MO in 2002, a MS degree from the
-Department of Applied Mathematics and Statistics (AMS) at Johns Hopkins
-University (JHU) in Baltimore, MD in 2009, and a PhD degree from the
-Department of Neuroscience at JHU in 2009. He was a Postdoctoral Fellow in
-AMS/JHU from 2009 until 2011, at which time he was appointed an Assistant
-Research Scientist, and became a member of the Institute for Data Intensive
-Science and Engineering. He spent 2 years at Information Initiative at Duke
-University, before coming home to his current appointment as 
-Assistant Professor
-in BME/JHU, and core faculty in both the Institute for Computational Medicine
-and the Center for Imaging Science, as well as a member of the Kavli
-Neuroscience Discovery Institute. He married his kindergarten sweetheart in the
-summer of 2014.
-
-His primary research interest is to extend and fuse statistical 
-machine learning
-and big data science to address the most important brain science and mental
-health questions of our time, particularly  connectomics. 
-His group’s research
-has been featured in a number of prominent scientific and engineering journals
-and conferences including Nature, Nature Methods, Science, Cell, PNAS, Science
-Translational Medicine, JMLR, PAMI, Annals of Applied Statistics,  NIPS, and
-SIAM Journal of Matrix Analysis and Applications. In 2011, he co-founded the
-Open Connectome Project which expanded in 2015 to be NeuroData, whose mission
-is to enable terascale neuroscience for everyone. All his works are conducted
-according to the highest standards of open science.
-\end{IEEEbiography}
+%\begin{IEEEbiography}[{\includegraphics[width=1in,clip,keepaspectratio]{gui2.png}}]{Guilherme Fran\c ca} 
+%has a BS degree in physics and received both his MS and 
+%PhD in Theoretical Physics from the Institute of 
+%Theoretical Physics IFT-UNESP/ICTP-SAIFR (2012). He was a Postdoctoral Fellow
+%with the High Energy Theory group in the Physics Department 
+%at Cornell University (2013--2015), working on topics of mathematical physics
+%and non-perturbative methods in quantum field theory 
+%and statistical mechanics.
+%He was a Postdoctoral Fellow in
+%the Computer Science Department at Boston College (2015), working
+%in optimization methods.
+%Since 2016 he is a Postdoctoral Fellow at Johns Hopkins University, currently
+%with the Mathematical Institute for Data Science (MINDS). 
+%His current research interests involve  
+%machine learning and optimization.
+%\end{IEEEbiography}
+%
+%\begin{IEEEbiography}[{\includegraphics[width=1in,clip,keepaspectratio,height=1.30in]{rizzo.jpg}}]{Maria~L.~Rizzo}
+%holds a BS and MA degree in Mathematics from the University of Toledo, a MS in
+%Applied Statistics from BGSU, and a PhD from BGSU in 2002. After four years as
+%Assistant Professor at Ohio University Department of Mathematics, she 
+%returned to BGSU in 2006, where she is currently Professor of Statistics 
+%in the Department of
+%Mathematics and Statistics. Her primary research interests are 
+%centered on energy
+%statistics, distance correlation and its applications, computational
+%statistics, and related topics. In addition to teaching statistics, data
+%science, and actuarial science, she enjoys writing textbooks about statistical
+%computing and using R software. She has a forthcoming research 
+%monograph on energy
+%statistics (joint with G. Szekely) and also other book writing projects under
+%contract as well.
+%\end{IEEEbiography}
+%
+%
+%% insert where needed to balance the two columns on the last page with
+%% biographies
+%%\newpage
+%
+%\begin{IEEEbiography}[{\includegraphics[width=1in,clip,keepaspectratio]{jovo.png}}]{Joshua~T.~Vogelstein}
+%received a BS degree from the Department of Biomedical Engineering (BME) at
+%Washington University in St. Louis, MO in 2002, a MS degree from the
+%Department of Applied Mathematics and Statistics (AMS) at Johns Hopkins
+%University (JHU) in Baltimore, MD in 2009, and a PhD degree from the
+%Department of Neuroscience at JHU in 2009. He was a Postdoctoral Fellow in
+%AMS/JHU from 2009 until 2011, at which time he was appointed an Assistant
+%Research Scientist, and became a member of the Institute for Data Intensive
+%Science and Engineering. He spent 2 years at Information Initiative at Duke
+%University, before coming home to his current appointment as 
+%Assistant Professor
+%in BME/JHU, and core faculty in both the Institute for Computational Medicine
+%and the Center for Imaging Science, as well as a member of the Kavli
+%Neuroscience Discovery Institute. He married his kindergarten sweetheart in the
+%summer of 2014. 
+%
+%His primary research interest is to extend and fuse statistical 
+%machine learning
+%and big data science to address the most important brain science and mental
+%health questions of our time, particularly  connectomics. 
+%His group’s research
+%has been featured in a number of prominent scientific and engineering journals
+%and conferences including Nature, Nature Methods, Science, Cell, PNAS, Science
+%Translational Medicine, JMLR, PAMI, Annals of Applied Statistics,  NIPS, and
+%SIAM Journal of Matrix Analysis and Applications. In 2011, he co-founded the
+%Open Connectome Project which expanded in 2015 to be NeuroData, whose mission
+%is to enable terascale neuroscience for everyone. All his works are conducted
+%according to the highest standards of open science.
+%\end{IEEEbiography}
 

### 2018-10-23 energy_clustering/arxiv/energy_clustering.tex
acknowledgements

@@
 Research Award (NIH \#R01NS092474) and  the Defense Advanced Research Projects
-Agency’s (DARPA) SIMPLEX program through SPAWAR contract N66001-15-C-4041.
+Agency’s (DARPA) SIMPLEX program through SPAWAR contract N66001-15-C-4041,
+and also the 
+DARPA Lifelong Learning Machines program through contract FA8650-18-2-7834.
 

### 2018-10-23 energy_clustering/pami18/energy_clustering.tex
acknowledgements

@@
 % The paper headers
-\markboth{Energy Clustering}{}
+%\markboth{Energy Clustering}{}
 %{Journal of \LaTeX\ Class Files,~Vol.~14, No.~8, August~2015}%
@@
 Research Award (NIH \#R01NS092474) and  the Defense Advanced Research Projects
-Agency’s (DARPA) SIMPLEX program through SPAWAR contract N66001-15-C-4041.
+Agency’s (DARPA) SIMPLEX program through SPAWAR contract N66001-15-C-4041,
+and also the DARPA Lifelong Learning Machines program through contract
+FA8650-18-2-7834.
 

### 2019-09-26 energy_clustering/pami18/energy_clustering.tex
does this work?

@@
 with the Center for Imaging Science, 
-the Department of Biomedical Engineering and Institute for Computational
-Medicine, Johns Hopkins University. \protect\\
+the Department of Biomedical Engineering, Institute for Computational
+Medicine, and Kavli Neuroscience Discovery Institute, Johns Hopkins University. \protect\\
 % note need leading \protect in front of \\ to get a newline within \thanks as
@@
 }
+% todo i don't understand what the filled in content is? is that showing the truth? gotta state it whatever it is.  i would remove the black lines labeled truth if you keep the filled.
 \end{figure*}
@@
 (e) Data from \eqref{eq:gauss3} with $N = 300$ and $m=200$.
+% i'd make the points bigger.
 }
